### 2025-05-21 advances/informe1.qmd
Update

---
title: "Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
author:
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
citeproc: true
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

```{r}
# This document was last modified at {{< modified >}}
# 
# and it was last rendered at {{< now >}}

```

# Introduction

Since the early 1980s, many countries have experienced a widespread retreatment of universal welfare programs and have adopted a trend toward privatization and commoditization of various public goods, welfare policies and social services [@gingrich_making_2011; @streeck_how_2016; @salamon_marketization_1993]. Neoliberal reforms reshaped the architecture of welfare-state institutions across Western democracies and Latin America, extending market criteria into areas of social reproduction that were traditionally governed by the state. This shift diminished the role of public provision of core social services [@arrizabalo_milagro_1995; @busemeyer_welfare_2020]. Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy---a constellation of norms and values about fair allocation embedded in, and reinforced by, social institutions [@mau_moral_2014; @svallforsMoralEconomyClass2006a]. Within this moral economy, a growing body of research has been addressed the extent to which, and the mechanisms through which, citizens deem it just that access to social services such as health care, pensions, and education be determined by market-based criteria, a phenomenon labelled market justice preferences [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Understanding these preferences is crucial because they can legitimize economic inequality by framing it as the natural outcome of fair competition, driven by individual responsibility and limited state intervention [@svallfors_political_2007].

Existing literature consistently demonstrates that market justice preferences are shaped by both the economic and institutional contexts of countries and individuals' positions within social stratification. Recent studies indicate that in countries with greater public provision of social services or more extensive welfare states, market justice preferences tend to be lower [@busemeyer_skills_2014; @immergut_it_2020], whereas in contexts with higher levels of privatization, preferences for market-based criteria are more pronounced [@castillo_socialization_2024; @lindh_public_2015]. In such contexts, market justice preferences tend to rise as individuals "ascend" the social structure, with those in more privileged positions in terms of class, education, and income holding stronger preferences for market-based solutions compared to those in more disadvantaged or at-risk positions [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015; @svallfors_political_2007; @kluegel_legitimation_1999]. Empirically, it has been consistently demonstrated that those in socioeconomically advantaged positions endorse the idea that those with higher incomes should be able to pay more for better social services in the domains of education [@lee_fairness_2023], healthcare [@vondemknesebeck_are_2016; @immergut_it_2020] and old age pensions [@castillo_socialization_2024; @otero_power_2024].

Although it is clear that one's social position influences market preferences, the question of how upward or downward mobility within the social structure affects these preferences remains unanswered. This is far from trivial, as many individuals in Western democracies and Latin America have experienced varying degrees of social mobility [@lopez-roldan_comparative_2021; @breen_social_2004]. The social origins and destinations of individuals exert different effects on a range of attitudes [@gugushvili_trends_2014], while movement between these positions exposes individuals to distinct processes and mechanisms that can profoundly influence their perspectives, shaping how they perceive the world and evaluate what they consider fair [@gugushvili_trends_2014]). Some of the current research on attitudes toward inequality has concluded that the type of social mobility---upward or downward---that individuals experience, whether measured objectively or subjectively, has differentiated effects on support for redistribution [@alesina_intergenerational_2018; @benabou_social_2001; @jaime-castillo_social_2019; @schmidt_experience_2011], system legitimacy, and meritocratic beliefs [@day_movin_2017; @mijs_belief_2022], as well as on tolerance for economic inequality [@shariff_income_2016; @gugushvili_intergenerational_2016c], attributions about poverty and wealth [@bucca_merit_2016; @gugushvili_intergenerational_2016], and support for welfare policies and market economies [@gugushvili_subjective_2017; @wen_does_2021]. Thus, building on these contributions, expanding the analysis of the effects of social mobility on market justice preferences could contribute to this emerging literature and shed light on the mechanisms behind the justification of inequalities in access to social services.

Market justice preferences are shaped not only by objective socioeconomic factors but also by perceptions and beliefs about meritocracy. Meritocracy asserts that inequality is an inherent feature of societies but can be legitimized through principles such as effort and talent [@young_rise_1958; @davis_principles_2001]. Previous studies have shown that individuals with stronger meritocratic beliefs tend to perceive less inequality, attributing economic differences to individual achievements [@mijs_paradox_2019; @wilson_role_2003]. Additionally, these beliefs justify greater inequality, as they are associated with attitudes that uphold the legitimization of social differences [@batruch_belief_2023]. In unequal societies where the distribution of goods and services is predominantly governed by market logic, such beliefs can play a critical role in the acceptance of social inequalities. Recent research by @castillo_socialization_2024 demonstrates that students in Chile who believe that effort and talent are rewarded in their country hold stronger preferences for market justice in access to healthcare, pensions, and education. By incorporating perceptions of meritocracy into this study, we can better understand how preferences toward market justice are shaped not only by an individual's structural social position but also by beliefs about the justification of these inequalities. Moreover, these beliefs are often solidified early in life, reinforced by institutions that promote values such as effort and individual skills as means to upward social mobility [@castillo_socialization_2024; @reynolds_perceptions_2014].

Beyond the isolated effects of social mobility and beliefs in meritocracy on market justice preferences, these factors interact to shape these preferences in a complex way. One proposed mechanism in the literature regarding the process of social mobility that leads to differences in distributive justice preferences is the psychological mechanism of self-serving bias in causal attribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c]. Self-serving bias suggests that people attribute failures, such as downward mobility, to factors beyond their control, while explaining successes, such as upward mobility, by pointing to their own merits and effort [@miller_selfserving_1975]. Individuals who experience upward mobility tend to view their position as a result of their own effort and skills, making them more likely to support the idea that individuals are responsible for their own success or failure. Research shows that upward mobility is associated with lower redistribution preferences [@schmidt_experience_2011; @gugushvili_intergenerational_2016c] and stronger legitimacy of income inequality [@shariff_income_2016]. In contrast, individuals who experience downward mobility tend to attribute their failure to external factors, such as inequality, and are more supportive of redistribution and the idea that individuals should not be blamed for their economic failure [@schmidt_experience_2011; @gugushvili_intergenerational_2016c]. Taken together, I suggest that meritocratic beliefs may mirror this self-serving bias mechanism, as they reinforce the tendency to justify one's social status as the result of personal merits, which is closely related to the attribution bias.

Focusing in Chile, this article aims to achieve two main objectives: first, to analyze the extent to which intergenerational social mobility affects market justice preferences on healthcare, pensions and education, and second, to examine how meritocratic beliefs may moderate this relationship. The central argument is that upward social mobility will have a positive effect on market justice preferences, while downward mobility will have a negative effect. Additionally, meritocratic beliefs will moderate this relationship by reflecting a self-serving bias mechanism of causal attribution.

The Chilean case is particularly intriguing for studying market justice preferences. Despite being one of the most prosperous nations in Latin America, it has one of the highest levels of economic inequality in the region. The poorest 50% of Chileans capture only 10% of the total income and hold negative wealth, while the richest 1% receives nearly 27% of the income and controls 49.6% of the wealth [@chancel_world_2022]. This economic inequality exists alongside notable social mobility, with significant upward mobility in recent decades [@torche_intergenerational_2014; @espinoza_estratificacion_2013]. Chile's social policy regime is characterized by a welfare model that relies heavily on private provision in healthcare, pensions, and education systems, often segmented by individuals' ability to pay and heavily dependent on state subsidies [@boccardo_30_2020].

Against this backdrop, the main research question is: To what extent does intergenerational social mobility affect market justice preferences, and how do meritocratic beliefs explain this relationship in Chile? This study aims to contribute to the literature by providing evidence from a Latin American developing country, highlighting how changes in individuals' socioeconomic position and meritocratic beliefs influence market justice preferences in critical areas like healthcare, pensions, and education.

# Data, variables and methods

## Data

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES). The survey is an annual panel collected from 2016 to 2023, comprising two independent samples (original and refreshment), featuring permanent and rotating questionnaire modules. It evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile. The sampling design is complex, probabilistic, clustered, multistage, and stratified by city size, targeting men and women aged 18--75 who habitually reside in private dwellings in urban areas spanning 40 cities (92 municipalities, 13 regions). This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample. Further details on sampling, attrition, and weighting can be found at https://coes.cl/encuesta-panel/, and the dataset is publicly available at https://dataverse.harvard.edu/dataverse/elsoc.

## Variables

*Outcome variables*

**Market justice preferences**: The outcome variables in this study are market justice preferences. This construct is operationalized through three variables that address the degree of justification regarding whether access to social services in healthcare, pensions, and education should be income conditional. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is asked for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Although these items theoretically reflect the same underlying concept of market justice [@castillo_socialization_2024; @lindh_public_2015], they are analyzed separately because healthcare, pensions, and education may prompt distinct response patterns [@lee_fairness_2023; @busemeyer_skills_2014; @immergut_it_2020].

*Independent variables*

**Social mobility**: Intergenerational social mobility is treated as an exposure indicating whether respondents occupy a different class position from their fathers. Following Breen and Ermisch's [-@breen_effects_2024] framework for estimating causal effects of mobility, the variable is constructed in two stages.

1.  Class assignment. Using the International Socio-Economic Index of Occupational Status (ISEI) derived from three-digit ISCO-08 codes, both paternal (origin) and respondent (destination) occupations are grouped into three strata:

    -   Upper (codes 100--299: managers, directors, professionals)
    -   Middle (codes 300--499: skilled white‑collar workers)
    -   Lower (codes 500 and above: manual, semi‑skilled, or unskilled workers)

2.  Propensity-score estimation. To model mobility as a treatment---i.e., a transition from origin *j* to destination *k*---I estimate multinomial logit propensity scores using covariates that influence an individual's likelihood of moving between strata: (a) father's educational level, (b) presence of both parents at age 15, (c) nationality, (d) age, (e) sex, and (f) ethnicity.

These propensity scores are subsequently employed to adjust for selection into mobility when assessing its effect on market justice preferences.

**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1958]. The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").

*Controls*

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, I incorporate per capita household income quantile, educational level (1=Primary or below, 2=Secondary, 3=Technical, 4=University or above), age (in years), and sex (1=Male, 2=Female), which have been shown to significantly influence market justice preferences [@castillo_socialization_2024; @lindh_public_2015]. Regarding attitudinal variables, I include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and social mobility and meritocracy.

## Methods

Following Breen and Ermisch's [-@breen_effects_2024] strategy for estimating the causal effect of social mobility on market justice preferences, I employ linear regression models with inverse-probability weights (IPW) for mobility conditional on class of origin. This approach allows estimation of the average causal effect of moving from an origin class *j* to a destination class *k* by comparing individuals who actually reach *k* with their counterfactual outcome had they instead moved to an alternative destination *k′* [@breen_effects_2024]. The resulting estimand is the average treatment effect on the treated (ATT).

All the analyses will be conducted using R software.

# References

::: {#refs}
:::

### 2025-05-21 advances/informe1.tex
Update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
  pdfauthor={Andreas Laffert},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Preferences for market justice: the role of intergenerational
social mobility and beliefs in meritocracy in Chile}

  \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }

\date{}
\begin{document}
\maketitle

\setstretch{1.15}
\section{Introduction}\label{introduction}

Since the early 1980s, many countries have experienced a widespread
retreatment of universal welfare programs and have adopted a trend
toward privatization and commoditization of various public goods,
welfare policies and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-salamon_marketization_1993}{Salamon, 1993};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). Neoliberal reforms
reshaped the architecture of welfare-state institutions across Western
democracies and Latin America, extending market criteria into areas of
social reproduction that were traditionally governed by the state. This
shift diminished the role of public provision of core social services
(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
Echoing Polanyi's (\citeproc{ref-polanyi_great_1975}{1975}) insight that
markets constitute a distinct moral order, the institutional diffusion
of market rules has fostered a corresponding moral economy---a
constellation of norms and values about fair allocation embedded in, and
reinforced by, social institutions (\citeproc{ref-mau_moral_2014}{Mau,
2014}; \citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Within this moral economy, a growing body of research has been addressed
the extent to which, and the mechanisms through which, citizens deem it
just that access to social services such as health care, pensions, and
education be determined by market-based criteria, a phenomenon labelled
market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial because they can legitimize
economic inequality by framing it as the natural outcome of fair
competition, driven by individual responsibility and limited state
intervention (\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}).

Existing literature consistently demonstrates that market justice
preferences are shaped by both the economic and institutional contexts
of countries and individuals' positions within social stratification.
Recent studies indicate that in countries with greater public provision
of social services or more extensive welfare states, market justice
preferences tend to be lower
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), whereas
in contexts with higher levels of privatization, preferences for
market-based criteria are more pronounced
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Empirically,
it has been consistently demonstrated that those in socioeconomically
advantaged positions endorse the idea that those with higher incomes
should be able to pay more for better social services in the domains of
education (\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}),
healthcare (\citeproc{ref-immergut_it_2020}{Immergut \& Schneider,
2020}; \citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}) and old age pensions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}).

Although it is clear that one's social position influences market
preferences, the question of how upward or downward mobility within the
social structure affects these preferences remains unanswered. This is
far from trivial, as many individuals in Western democracies and Latin
America have experienced varying degrees of social mobility
(\citeproc{ref-breen_social_2004}{Breen, 2004};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). The social origins and destinations of individuals exert
different effects on a range of attitudes
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}), while
movement between these positions exposes individuals to distinct
processes and mechanisms that can profoundly influence their
perspectives, shaping how they perceive the world and evaluate what they
consider fair (\citeproc{ref-gugushvili_trends_2014}{Gugushvili,
2014})). Some of the current research on attitudes toward inequality has
concluded that the type of social mobility---upward or downward---that
individuals experience, whether measured objectively or subjectively,
has differentiated effects on support for redistribution
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
2011}), system legitimacy, and meritocratic beliefs
(\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}), as well as on
tolerance for economic inequality
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), attributions
about poverty and wealth (\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}),
and support for welfare policies and market economies
(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}). Thus, building on
these contributions, expanding the analysis of the effects of social
mobility on market justice preferences could contribute to this emerging
literature and shed light on the mechanisms behind the justification of
inequalities in access to social services.

Market justice preferences are shaped not only by objective
socioeconomic factors but also by perceptions and beliefs about
meritocracy. Meritocracy asserts that inequality is an inherent feature
of societies but can be legitimized through principles such as effort
and talent (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1958}{Young, 1958}). Previous studies have
shown that individuals with stronger meritocratic beliefs tend to
perceive less inequality, attributing economic differences to individual
achievements (\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). Additionally, these
beliefs justify greater inequality, as they are associated with
attitudes that uphold the legitimization of social differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In unequal
societies where the distribution of goods and services is predominantly
governed by market logic, such beliefs can play a critical role in the
acceptance of social inequalities. Recent research by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) demonstrates that
students in Chile who believe that effort and talent are rewarded in
their country hold stronger preferences for market justice in access to
healthcare, pensions, and education. By incorporating perceptions of
meritocracy into this study, we can better understand how preferences
toward market justice are shaped not only by an individual's structural
social position but also by beliefs about the justification of these
inequalities. Moreover, these beliefs are often solidified early in
life, reinforced by institutions that promote values such as effort and
individual skills as means to upward social mobility
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

Beyond the isolated effects of social mobility and beliefs in
meritocracy on market justice preferences, these factors interact to
shape these preferences in a complex way. One proposed mechanism in the
literature regarding the process of social mobility that leads to
differences in distributive justice preferences is the psychological
mechanism of self-serving bias in causal attribution
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). Self-serving
bias suggests that people attribute failures, such as downward mobility,
to factors beyond their control, while explaining successes, such as
upward mobility, by pointing to their own merits and effort
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}).
Individuals who experience upward mobility tend to view their position
as a result of their own effort and skills, making them more likely to
support the idea that individuals are responsible for their own success
or failure. Research shows that upward mobility is associated with lower
redistribution preferences
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}) and stronger
legitimacy of income inequality
(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
individuals who experience downward mobility tend to attribute their
failure to external factors, such as inequality, and are more supportive
of redistribution and the idea that individuals should not be blamed for
their economic failure
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). Taken together,
I suggest that meritocratic beliefs may mirror this self-serving bias
mechanism, as they reinforce the tendency to justify one's social status
as the result of personal merits, which is closely related to the
attribution bias.

Focusing in Chile, this article aims to achieve two main objectives:
first, to analyze the extent to which intergenerational social mobility
affects market justice preferences on healthcare, pensions and
education, and second, to examine how meritocratic beliefs may moderate
this relationship. The central argument is that upward social mobility
will have a positive effect on market justice preferences, while
downward mobility will have a negative effect. Additionally,
meritocratic beliefs will moderate this relationship by reflecting a
self-serving bias mechanism of causal attribution.

The Chilean case is particularly intriguing for studying market justice
preferences. Despite being one of the most prosperous nations in Latin
America, it has one of the highest levels of economic inequality in the
region. The poorest 50\% of Chileans capture only 10\% of the total
income and hold negative wealth, while the richest 1\% receives nearly
27\% of the income and controls 49.6\% of the wealth
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This economic
inequality exists alongside notable social mobility, with significant
upward mobility in recent decades
(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). Chile's
social policy regime is characterized by a welfare model that relies
heavily on private provision in healthcare, pensions, and education
systems, often segmented by individuals' ability to pay and heavily
dependent on state subsidies (\citeproc{ref-boccardo_30_2020}{Boccardo,
2020}).

Against this backdrop, the main research question is: To what extent
does intergenerational social mobility affect market justice
preferences, and how do meritocratic beliefs explain this relationship
in Chile? This study aims to contribute to the literature by providing
evidence from a Latin American developing country, highlighting how
changes in individuals' socioeconomic position and meritocratic beliefs
influence market justice preferences in critical areas like healthcare,
pensions, and education.

\section{Data, variables and methods}\label{data-variables-and-methods}

\subsection{Data}\label{data}

This study draws on data from the Chilean Longitudinal Social Survey
(ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
The survey is an annual panel collected from 2016 to 2023, comprising
two independent samples (original and refreshment), featuring permanent
and rotating questionnaire modules. It evaluates how individuals think,
feel, and behave regarding conflict and social cohesion in Chile. The
sampling design is complex, probabilistic, clustered, multistage, and
stratified by city size, targeting men and women aged 18--75 who
habitually reside in private dwellings in urban areas spanning 40 cities
(92 municipalities, 13 regions). This analysis focuses on 2023, the most
recent wave, which includes 1,737 respondents in the original sample and
989 in the refreshment sample. Further details on sampling, attrition,
and weighting can be found at https://coes.cl/encuesta-panel/, and the
dataset is publicly available at
https://dataverse.harvard.edu/dataverse/elsoc.

\subsection{Variables}\label{variables}

\emph{Outcome variables}

\textbf{Market justice preferences}: The outcome variables in this study
are market justice preferences. This construct is operationalized
through three variables that address the degree of justification
regarding whether access to social services in healthcare, pensions, and
education should be income conditional. Specifically, the justification
of inequality in healthcare is assessed by the question: ``Is it fair in
Chile that people with higher incomes can access better healthcare than
people with lower incomes?'' The same question is asked for pensions and
education. In all cases, respondents indicate their level of agreement
on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
(``strongly agree''). Although these items theoretically reflect the
same underlying concept of market justice
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}), they are analyzed
separately because healthcare, pensions, and education may prompt
distinct response patterns
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}).

\emph{Independent variables}

\textbf{Social mobility}: Intergenerational social mobility is treated
as an exposure indicating whether respondents occupy a different class
position from their fathers. Following Breen and Ermisch's
(\citeproc{ref-breen_effects_2024}{2024}) framework for estimating
causal effects of mobility, the variable is constructed in two stages.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Class assignment. Using the International Socio-Economic Index of
  Occupational Status (ISEI) derived from three-digit ISCO-08 codes,
  both paternal (origin) and respondent (destination) occupations are
  grouped into three strata:

  \begin{itemize}
  \tightlist
  \item
    Upper (codes 100--299: managers, directors, professionals)
  \item
    Middle (codes 300--499: skilled white‑collar workers)
  \item
    Lower (codes 500 and above: manual, semi‑skilled, or unskilled
    workers)
  \end{itemize}
\item
  Propensity-score estimation. To model mobility as a treatment---i.e.,
  a transition from origin \emph{j} to destination \emph{k}---I estimate
  multinomial logit propensity scores using covariates that influence an
  individual's likelihood of moving between strata: (a) father's
  educational level, (b) presence of both parents at age 15, (c)
  nationality, (d) age, (e) sex, and (f) ethnicity.
\end{enumerate}

These propensity scores are subsequently employed to adjust for
selection into mobility when assessing its effect on market justice
preferences.

\textbf{Meritocracy}: Meritocratic perception is operationalized through
two components: one addressing effort and another focusing on talent
(\citeproc{ref-young_rise_1958}{Young, 1958}). The item used to gauge
effort is: ``In Chile, people are rewarded for their efforts,'' while
the item for talent is: ``In Chile, people are rewarded for their
intelligence and skills.'' In both cases, respondents indicate their
level of agreement on a five-point Likert scale, ranging from 1
(``strongly disagree'') to 5 (``strongly agree'').

\emph{Controls}

Sociodemographic and attitudinal variables are included to control for
potential composition effects in the population. In terms of
sociodemographic characteristics, I incorporate per capita household
income quantile, educational level (1=Primary or below, 2=Secondary,
3=Technical, 4=University or above), age (in years), and sex (1=Male,
2=Female), which have been shown to significantly influence market
justice preferences (\citeproc{ref-castillo_socialization_2024}{Castillo
et al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
attitudinal variables, I include political identification (1=Left,
2=Center, 3=Right, 4=No identification) and subjective social status
(ranging from 1 to 10) because they may confound the relationship
between market justice preferences and social mobility and meritocracy.

\subsection{Methods}\label{methods}

Following Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024})
strategy for estimating the causal effect of social mobility on market
justice preferences, I employ linear regression models with
inverse-probability weights (IPW) for mobility conditional on class of
origin. This approach allows estimation of the average causal effect of
moving from an origin class \emph{j} to a destination class \emph{k} by
comparing individuals who actually reach \emph{k} with their
counterfactual outcome had they instead moved to an alternative
destination \emph{k′} (\citeproc{ref-breen_effects_2024}{Breen \&
Ermisch, 2024}). The resulting estimand is the average treatment effect
on the treated (ATT).

All the analyses will be conducted using R software.

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-alesina_intergenerational_2018}
Alesina, A., Stantcheva, S., \& Teso, E. (2018). Intergenerational
{Mobility} and {Preferences} for {Redistribution}. \emph{American
Economic Review}, \emph{108}(2), 521--554.
\url{https://doi.org/10.1257/aer.20162015}

\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
durante la dictadura}}. Libros de la Catarata.

\bibitem[\citeproctext]{ref-batruch_belief_2023}
Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
(2023). Belief in {School Meritocracy} and the {Legitimization} of
{Social} and {Income Inequality}. \emph{Social Psychological and
Personality Science}, \emph{14}(5), 621--635.
\url{https://doi.org/10.1177/19485506221111017}

\bibitem[\citeproctext]{ref-benabou_social_2001}
Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
Economics}, \emph{116}(2), 447--487.
\url{https://doi.org/10.1162/00335530151144078}

\bibitem[\citeproctext]{ref-boccardo_30_2020}
Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
que la pandemia revel{ó}} (Nodo XXI). Santiago.

\bibitem[\citeproctext]{ref-breen_social_2004}
Breen, R. (2004). \emph{Social {Mobility} in {Europe}}. OUP Oxford.

\bibitem[\citeproctext]{ref-breen_effects_2024}
Breen, R., \& Ermisch, J. (2024). The {Effects} of {Social Mobility}.
\emph{Sociological Science}, \emph{11}, 467--488.
\url{https://doi.org/10.15195/v11.a17}

\bibitem[\citeproctext]{ref-bucca_merit_2016}
Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
Latin Americans}' beliefs about wealth and poverty. \emph{Research in
Social Stratification and Mobility}, \emph{44}, 98--112.
\url{https://doi.org/10.1016/j.rssm.2016.02.005}

\bibitem[\citeproctext]{ref-busemeyer_skills_2014}
Busemeyer, M. R. (2014). \emph{Skills and {Inequality}: {Partisan
Politics} and the {Political Economy} of {Education Reforms} in {Western
Welfare States}}. Cambridge University Press.

\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
Busemeyer, M. R., \& Iversen, T. (2020). The {Welfare State} with
{Private Alternatives}: {The Transformation} of {Popular Support} for
{Social Insurance}. \emph{The Journal of Politics}, \emph{82}(2),
671--686. \url{https://doi.org/10.1086/706980}

\bibitem[\citeproctext]{ref-castillo_socialization_2024}
Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
{Socialization} of {Meritocracy} and {Market Justice Preferences} at
{School}. \emph{Societies}, \emph{14}(11), 214.
\url{https://doi.org/10.3390/soc14110214}

\bibitem[\citeproctext]{ref-chancel_world_2022}
Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
inequality report 2022.
https://bibliotecadigital.ccb.org.co/handle/11520/27510.

\bibitem[\citeproctext]{ref-davis_principles_2001}
Davis, K., \& Moore, W. E. (2001). Some {Principles} of
{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
Routledge.

\bibitem[\citeproctext]{ref-day_movin_2017}
Day, M. V., \& Fiske, S. T. (2017). Movin' on {Up}? {How Perceptions} of
{Social Mobility Affect Our Willingness} to {Defend} the {System}.
\emph{Social Psychological and Personality Science}, \emph{8}(3),
267--274. \url{https://doi.org/10.1177/1948550616678454}

\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
movilidad social bajo un}.

\bibitem[\citeproctext]{ref-gingrich_making_2011}
Gingrich, J. R. (2011). \emph{Making {Markets} in the {Welfare State}:
{The Politics} of {Varying Market Reforms}} (1st ed.). Cambridge
University Press. \url{https://doi.org/10.1017/CBO9780511791529}

\bibitem[\citeproctext]{ref-gugushvili_trends_2014}
Gugushvili, A. (2014). Trends, {Covariates} and {Consequences} of
{Intergenerational Social Mobility} in {Post- Socialist Societies}.

\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016c}
Gugushvili, A. (2016a). Intergenerational objective and subjective
mobility and attitudes towards income differences: Evidence from
transition societies. \emph{Journal of International and Comparative
Social Policy}, \emph{32}(3), 199--219.
\url{https://doi.org/10.1080/21699763.2016.1206482}

\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016}
Gugushvili, A. (2016b). Intergenerational {Social Mobility} and {Popular
Explanations} of {Poverty}: {A Comparative Perspective}. \emph{Social
Justice Research}, \emph{29}(4), 402--428.
\url{https://doi.org/10.1007/s11211-016-0275-9}

\bibitem[\citeproctext]{ref-gugushvili_subjective_2017}
Gugushvili, A. (2017). Subjective {Intergenerational Mobility} and
{Support} for {Welfare State Programmes}.

\bibitem[\citeproctext]{ref-immergut_it_2020}
Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
affluent to be able to purchase {``better''} healthcare? {Existential}
standards and institutional norms in healthcare attitudes across 28
countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
\url{https://doi.org/10.1016/j.socscimed.2020.113146}

\bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social mobility
and demand for redistribution in {Europe}: A comparative analysis.
\emph{The British Journal of Sociology}, \emph{70}(1), 138--165.
\url{https://doi.org/10.1111/1468-4446.12363}

\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
about {Market Justice}, 1991-1996. \emph{European Sociological Review},
\emph{15}(3), 251--283. Retrieved from
\url{https://www.jstor.org/stable/522731}

\bibitem[\citeproctext]{ref-koos_moral_2019}
Koos, S., \& Sachweh, P. (2019). The moral economies of market
societies: Popular attitudes towards market competition, redistribution
and reciprocity in comparative perspective. \emph{Socio-Economic
Review}, \emph{17}(4), 793--821.
\url{https://doi.org/10.1093/ser/mwx045}

\bibitem[\citeproctext]{ref-lee_fairness_2023}
Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
inequality: The effects of self-interest and neoliberal orientations.
\emph{The Australian Educational Researcher}.
\url{https://doi.org/10.1007/s13384-023-00636-6}

\bibitem[\citeproctext]{ref-lindh_public_2015}
Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
towards {Market Distribution} of {Social Services} -- {A Comparison} of
17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
887--910. \url{https://doi.org/10.1111/spol.12105}

\bibitem[\citeproctext]{ref-lindh_bringing_2023}
Lindh, A., \& McCall, L. (2023). Bringing the market in: An expanded
framework for understanding popular responses to economic inequality.
\emph{Socio-Economic Review}, \emph{21}(2), 1035--1055.
\url{https://doi.org/10.1093/ser/mwac018}

\bibitem[\citeproctext]{ref-lopez-roldan_comparative_2021}
López-Roldán, P., \& Fachelli, S. (Eds.). (2021). \emph{Towards a
{Comparative Analysis} of {Social Inequalities} between {Europe} and
{Latin America}}. Cham: Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-48442-2}

\bibitem[\citeproctext]{ref-mau_moral_2014}
Mau, S. (2014). \emph{The moral economy of welfare states: {Britain} and
{Germany} compared} (First issued in paperback). London New York NY:
Routledge \& Francis Group.

\bibitem[\citeproctext]{ref-mijs_paradox_2019}
Mijs, J. (2019). The paradox of inequality: Income inequality and belief
in meritocracy go hand in hand. \emph{Socio-Economic Review},
\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}

\bibitem[\citeproctext]{ref-mijs_belief_2022}
Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}

\bibitem[\citeproctext]{ref-miller_selfserving_1975}
Miller, D. T., \& Ross, M. (1975). Self-serving biases in the
attribution of causality: {Fact} or fiction? \emph{Psychological
Bulletin}, \emph{82}(2), 213--225.
\url{https://doi.org/10.1037/h0076486}

\bibitem[\citeproctext]{ref-otero_power_2024}
Otero, G., \& Mendoza, M. (2024). The {Power} of {Diversity}: {Class},
{Networks} and {Attitudes Towards Inequality}. \emph{Sociology},
\emph{58}(4), 851--876. \url{https://doi.org/10.1177/00380385231217625}

\bibitem[\citeproctext]{ref-polanyi_great_1975}
Polanyi, K. (1975). \emph{The great transformation} (Repr). New York,
NY: Octagon Books.

\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
of opportunity. \emph{Research in Social Stratification and Mobility},
\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}

\bibitem[\citeproctext]{ref-salamon_marketization_1993}
Salamon, L. M. (1993). The {Marketization} of {Welfare}: {Changing
Nonprofit} and {For-Profit Roles} in the {American Welfare State}.
\emph{Social Service Review}, \emph{67}(1), 16--39.
\url{https://doi.org/10.1086/603963}

\bibitem[\citeproctext]{ref-schmidt_experience_2011}
Schmidt, A. W. (2011). The experience of social mobility and the
formation of attitudes toward redistribution. In.

\bibitem[\citeproctext]{ref-shariff_income_2016}
Shariff, A. F., Wiwad, D., \& Aknin, L. B. (2016). Income {Mobility
Breeds Tolerance} for {Income Inequality}: {Cross-National} and
{Experimental Evidence}. \emph{Perspectives on Psychological Science},
\emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}

\bibitem[\citeproctext]{ref-streeck_how_2016}
Streeck, W. (2016). \emph{How will capitalism end? Essays on a failing
system}. London: Verso.

\bibitem[\citeproctext]{ref-svallforsMoralEconomyClass2006a}
Svallfors, S. (2006). \emph{The {Moral Economy} of {Class}: {Class} and
{Attitudes} in {Comparative Perspective}}. Stanford University Press.

\bibitem[\citeproctext]{ref-svallfors_political_2007}
Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
(1st ed.). Stanford University Press.
\url{https://doi.org/10.2307/j.ctvr0qv0q}

\bibitem[\citeproctext]{ref-torche_intergenerational_2014}
Torche, F. (2014). Intergenerational {Mobility} and {Inequality}: {The
Latin American Case}. \emph{Annual Review of Sociology}, \emph{40}(1),
619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}

\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
care inequalities unfair? {A} study on public attitudes in 23 countries.
\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
\url{https://doi.org/10.1186/s12939-016-0350-8}

\bibitem[\citeproctext]{ref-wen_does_2021}
Wen, F., \& Witteveen, D. (2021). Does perceived social mobility shape
attitudes toward government and family educational investment?
\emph{Social Science Research}, \emph{98}, 102579.
\url{https://doi.org/10.1016/j.ssresearch.2021.102579}

\bibitem[\citeproctext]{ref-wilson_role_2003}
Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
\url{https://doi.org/10.1023/A:1024667228488}

\bibitem[\citeproctext]{ref-young_rise_1958}
Young, M. (1958). \emph{The rise of the meritocracy}. New Brunswick,
N.J., U.S.A: Transaction Publishers.

\end{CSLReferences}

\end{document}

### 2025-05-21 advances/informe2.qmd
Update

---
title: "Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
    code-link: true
    title-block-banner: true
    theme:
      - cosmo
      - pretty.scss
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
author:
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
citeproc: true
link-citations: true
colorlinks: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
fig-height: 4
fig-width: 7.5
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

```{r}
# This document was last modified at {{< modified >}}
# 
# and it was last rendered at {{< now >}}

```

# Introduction

:::: {.justify}

What makes market-driven inequalities appear legitimate in the eyes of the public? Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016]. In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@arrizabalo_milagro_1995; @ferre_welfare_2023]. This transformation reduced the role of public provision and increased the presence of private actors in core social services [@harvey_breve_2015]. Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a]. Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that access to services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].

Existing literature shows that market justice preferences are shaped by both the economic and institutional context and individuals' positions within social stratification. Grounded in the notion that institutions influence individual attitudes [@jacksonActorsInstitutions2010], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015]. In such contexts, market justice preferences tend to rise as individuals "ascend" the social structure, with those in more privileged positions in terms of class, education, and income holding stronger preferences for market-based solutions compared to those in more disadvantaged or at-risk positions [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_socialization_2024; @otero_power_2024; @lindh_public_2015; @svallfors_political_2007].

Market justice preferences are shaped not only by objective socioeconomic conditions but also by popular beliefs about inequality. Among these, meritocracy is a key normative principle underpinning market-based distributive preferences [@mau_inequality_2015]. It frames inequality as inevitable but justifiable through effort and talent [@young_rise_1958; @davis_principles_2001]. Studies show that individuals with stronger meritocratic beliefs tend to perceive less inequality and legitimize it by attributing economic differences to personal achievement [@mijs_paradox_2019; @wilson_role_2003; @batruch_belief_2023]. In highly unequal societies where access to services is largely governed by market logic, such beliefs play a critical role in normalizing inequality. Recent evidence from Chile shows that students who believe effort and talent are rewarded in their country express stronger preferences for market-based access to healthcare, pensions, and education [@castillo_socialization_2024].

Although it is clear that one's social position influences market preferences, the question of how upward or downward mobility within the social structure affects these preferences remains unanswered. This question is far from trivial, especially in Latin America, where many have experienced various forms of mobility amid high economic inequality and deep welfare privatization [@lopez-roldan_comparative_2021; @torche_intergenerational_2014; @ferre_welfare_2023]. Social origins and destinations affect attitudes toward inequality in distinct ways [ @jaime-castillo_social_2019; @gugushvili_intergenerational_2016; @day_movin_2017; @gugushvili_subjective_2017; @wen_does_2021; @mijs_belief_2022], while movement between these positions exposes individuals to different experiences and mechanisms that shape their views on what is fair [@gugushvili_trends_2014; @mau_inequality_2015]. Building on this research, examining the effects of social mobility on market justice preferences can help to illuminate how inequalities in access to social services are justified among individuals who have experienced, or not, changes in their social standing, and what are the normative mechanisms that guide this justification [@mau_inequality_2015].

Beyond their isolated effects, social mobility and meritocratic beliefs interact in complex ways to shape market justice preferences. A key mechanism proposed in the literature to explain how mobility influences distributive justice preferences is the psychological process of self-serving bias in causal attribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c]. This bias suggests that individuals attribute failures---such as downward mobility---to external factors, while crediting successes---such as upward mobility---to their own merit and effort [@miller_selfserving_1975]. Those who experience upward mobility tend to view their social position as earned, making them more likely to believe that individuals are responsible for their own success or failure. Research shows that upward mobility is associated with weaker preferences for redistribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @jaime-castillo_social_2019; @alesina_intergenerational_2018] and stronger legitimacy of income inequality [@shariff_income_2016]. In contrast, individuals who experience downward mobility tend to blame structural factors like inequality and are more supportive of redistribution while rejecting merit-based explanations [@gugushvili_trends_2014]. Taken together, I argue that meritocratic beliefs may reflect---or reinforce---this self-serving attribution mechanism by legitimizing one's social status as the outcome of personal merit, closely tied to attribution bias.

Against this background, this article pursues two main objectives: first, to analyze the extent to which intergenerational social mobility influences market justice preferences regarding healthcare, pensions, and education; and second, to examine how meritocratic beliefs may moderate this relationship. Building on a theoretical framework that emphasizes how neoliberal transformations---particularly through the privatization and commodification of key areas of social reproduction---have profoundly reshaped processes of subject formation [@mau_inequality_2015], the central argument is that upward mobility increases support for market justice preferences, while downward mobility decreases it. Moreover, meritocratic beliefs are expected to moderate this relationship by reflecting a self-serving attribution bias, whereby individuals justify their social position in terms of personal merit.

This study focuses on Chile, a particularly intriguing case for examining market justice preferences. Despite significant economic growth and poverty reduction, Chile has some of the highest levels of inequality in Latin America and among OECD countries [@chancel_world_2022; @flores_top_2020]. This inequality coexists with short-range upward mobility among lower-class segments moving into middle strata, though strong barriers remain to reaching higher positions [@torche_intergenerational_2014; @espinoza_estratificacion_2013]. What makes Chile especially salient is that much of this inequality is rooted in deep neoliberal reforms that institutionalized the privatization and commodification of key social sectors [@madariaga_three_2020]. Introduced during the dictatorship (1973--1989) and expanded in democracy, these reforms enabled the unprecedented emergence of markets in health, pensions, and education, with provision segmented by individuals' ability to pay and supported by public subsidies [@boccardo_30_2020]. In parallel---and despite waves of protest against inequality and commodification from 2006 to 2019 [@somma_no_2021]---Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing their attitudes toward inequality and welfare distribution [@arteagaaguirre_politicas_2015; @castillo_socialization_2024].

In this context, the questions that guide this research are as follows:

(1) To what extent does intergenerational social mobility influence market justice preferences regarding healthcare, pensions, and education in Chile?
(2) How do meritocratic beliefs condition or moderate this relationship in the Chilean context?

To address these questions, this study draws on large-scale, representative survey data collected in 2018 from the urban Chilean population aged 18 to 75 (n = 2,983). The next section outlines the theoretical framework linking market justice preferences, social mobility, and meritocratic beliefs, and proposes a set of hypotheses. This is followed by a description of the data, variables, and analytical strategy. The final sections present the empirical findings, offer an interpretation of the results, and conclude with a discussion of their implications.

# Theoretical and empirical background

## Market justice preferences

## Social mobility

## Meritocracy

## The Chilean context

# Data, variables and methods

## Data

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES). The survey is an annual panel collected from 2016 to 2023, comprising two independent samples (original and refreshment), featuring permanent and rotating questionnaire modules. It evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile. The sampling design is complex, probabilistic, clustered, multistage, and stratified by city size, targeting men and women aged 18--75 who habitually reside in private dwellings in urban areas spanning 40 cities (92 municipalities, 13 regions). This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample. Further details on sampling, attrition, and weighting can be found at https://coes.cl/encuesta-panel/, and the dataset is publicly available at https://dataverse.harvard.edu/dataverse/elsoc.

## Variables

*Outcome variables*

**Market justice preferences**: The outcome variables in this study are market justice preferences. This construct is operationalized through three variables that address the degree of justification regarding whether access to social services in healthcare, pensions, and education should be income conditional. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is asked for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Although these items theoretically reflect the same underlying concept of market justice [@castillo_socialization_2024; @lindh_public_2015], they are analyzed separately because healthcare, pensions, and education may prompt distinct response patterns [@lee_fairness_2023; @busemeyer_skills_2014; @immergut_it_2020].

*Independent variables*

**Social mobility**: Intergenerational social mobility is treated as an exposure indicating whether respondents occupy a different class position from their fathers. Following Breen and Ermisch's [-@breen_effects_2024] framework for estimating causal effects of mobility, the variable is constructed in two stages.

1.  Class assignment. Using the International Socio-Economic Index of Occupational Status (ISEI) derived from three-digit ISCO-08 codes, both paternal (origin) and respondent (destination) occupations are grouped into three strata:

    -   Upper (codes 100--299: managers, directors, professionals)
    -   Middle (codes 300--499: skilled white‑collar workers)
    -   Lower (codes 500 and above: manual, semi‑skilled, or unskilled workers)

2.  Propensity-score estimation. To model mobility as a treatment---i.e., a transition from origin *j* to destination *k*---I estimate multinomial logit propensity scores using covariates that influence an individual's likelihood of moving between strata: (a) father's educational level, (b) presence of both parents at age 15, (c) nationality, (d) age, (e) sex, and (f) ethnicity.

These propensity scores are subsequently employed to adjust for selection into mobility when assessing its effect on market justice preferences.

**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1958]. The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").

*Controls*

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, I incorporate per capita household income quantile, educational level (1=Primary or below, 2=Secondary, 3=Technical, 4=University or above), age (in years), and sex (1=Male, 2=Female), which have been shown to significantly influence market justice preferences [@castillo_socialization_2024; @lindh_public_2015]. Regarding attitudinal variables, I include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and social mobility and meritocracy.

## Methods

Following Breen and Ermisch's [-@breen_effects_2024] strategy for estimating the causal effect of social mobility on market justice preferences, I employ linear regression models with inverse-probability weights (IPW) for mobility conditional on class of origin. This approach allows estimation of the average causal effect of moving from an origin class *j* to a destination class *k* by comparing individuals who actually reach *k* with their counterfactual outcome had they instead moved to an alternative destination *k′* [@breen_effects_2024]. The resulting estimand is the average treatment effect on the treated (ATT).

All the analyses will be conducted using R software.

# References

::: {#refs}
:::

::::

### 2025-05-21 advances/informe2.tex
Update

--- 2025-05-21 advances/informe1.tex
@@
 
-Since the early 1980s, many countries have experienced a widespread
-retreatment of universal welfare programs and have adopted a trend
-toward privatization and commoditization of various public goods,
-welfare policies and social services
-(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
-\citeproc{ref-salamon_marketization_1993}{Salamon, 1993};
-\citeproc{ref-streeck_how_2016}{Streeck, 2016}). Neoliberal reforms
-reshaped the architecture of welfare-state institutions across Western
-democracies and Latin America, extending market criteria into areas of
-social reproduction that were traditionally governed by the state. This
-shift diminished the role of public provision of core social services
+What makes market-driven inequalities appear legitimate in the eyes of
+the public? Since the early 1980s, many countries have experienced a
+widespread retreat from universal welfare programs and a shift toward
+the privatization and commodification of public goods, welfare policies,
+and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
+2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
+America, as elsewhere, neoliberal reforms reshaped welfare-state
+institutions by extending market logic into domains of social
+reproduction that were traditionally governed by the state
 (\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
-\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
-Echoing Polanyi's (\citeproc{ref-polanyi_great_1975}{1975}) insight that
-markets constitute a distinct moral order, the institutional diffusion
-of market rules has fostered a corresponding moral economy---a
-constellation of norms and values about fair allocation embedded in, and
-reinforced by, social institutions (\citeproc{ref-mau_moral_2014}{Mau,
-2014}; \citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
-Within this moral economy, a growing body of research has been addressed
-the extent to which, and the mechanisms through which, citizens deem it
-just that access to social services such as health care, pensions, and
-education be determined by market-based criteria, a phenomenon labelled
-market justice preferences
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
+reduced the role of public provision and increased the presence of
+private actors in core social services
+(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
+(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
+constitute a distinct moral order, the institutional diffusion of market
+rules has fostered a corresponding moral economy: a constellation of
+norms and values concerning fair allocation, embedded in institutions
+and shaping individual subjectivities
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
+\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
+this framework, a growing body of research examines the extent to which,
+and the mechanisms by which, citizens consider it fair that access to
+services like health care, pensions, and education be governed by
+market-based criteria---a phenomenon known as \emph{market justice
+preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
 \citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
@@
 \citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
-Understanding these preferences is crucial because they can legitimize
-economic inequality by framing it as the natural outcome of fair
-competition, driven by individual responsibility and limited state
-intervention (\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}).
+Understanding these preferences is crucial, as they contribute to
+legitimizing economic inequality by framing it as the fair result of
+individual responsibility and limited state intervention
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
 
-Existing literature consistently demonstrates that market justice
-preferences are shaped by both the economic and institutional contexts
-of countries and individuals' positions within social stratification.
-Recent studies indicate that in countries with greater public provision
-of social services or more extensive welfare states, market justice
-preferences tend to be lower
+Existing literature shows that market justice preferences are shaped by
+both the economic and institutional context and individuals' positions
+within social stratification. Grounded in the notion that institutions
+influence individual attitudes
+(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
+find that countries with stronger public provision or more expansive
+welfare states exhibit lower levels of market justice preferences
 (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), whereas
-in contexts with higher levels of privatization, preferences for
-market-based criteria are more pronounced
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
+more privatized contexts show stronger support for market-based criteria
 (\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
@@
 solutions compared to those in more disadvantaged or at-risk positions
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
 \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
+\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
 \citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Empirically,
-it has been consistently demonstrated that those in socioeconomically
-advantaged positions endorse the idea that those with higher incomes
-should be able to pay more for better social services in the domains of
-education (\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}),
-healthcare (\citeproc{ref-immergut_it_2020}{Immergut \& Schneider,
-2020}; \citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
-2016}) and old age pensions
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}).
+\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
+\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
+2016}).
+
+Market justice preferences are shaped not only by objective
+socioeconomic conditions but also by popular beliefs about inequality.
+Among these, meritocracy is a key normative principle underpinning
+market-based distributive preferences
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). It frames inequality as
+inevitable but justifiable through effort and talent
+(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
+\citeproc{ref-young_rise_1958}{Young, 1958}). Studies show that
+individuals with stronger meritocratic beliefs tend to perceive less
+inequality and legitimize it by attributing economic differences to
+personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
+2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
+\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
+societies where access to services is largely governed by market logic,
+such beliefs play a critical role in normalizing inequality. Recent
+evidence from Chile shows that students who believe effort and talent
+are rewarded in their country express stronger preferences for
+market-based access to healthcare, pensions, and education
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
 
@@
 preferences, the question of how upward or downward mobility within the
-social structure affects these preferences remains unanswered. This is
-far from trivial, as many individuals in Western democracies and Latin
-America have experienced varying degrees of social mobility
-(\citeproc{ref-breen_social_2004}{Breen, 2004};
-\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}). The social origins and destinations of individuals exert
-different effects on a range of attitudes
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}), while
-movement between these positions exposes individuals to distinct
-processes and mechanisms that can profoundly influence their
-perspectives, shaping how they perceive the world and evaluate what they
-consider fair (\citeproc{ref-gugushvili_trends_2014}{Gugushvili,
-2014})). Some of the current research on attitudes toward inequality has
-concluded that the type of social mobility---upward or downward---that
-individuals experience, whether measured objectively or subjectively,
-has differentiated effects on support for redistribution
+social structure affects these preferences remains unanswered. This
+question is far from trivial, especially in Latin America, where many
+have experienced various forms of mobility amid high economic inequality
+and deep welfare privatization (\citeproc{ref-ferre_welfare_2023}{Ferre,
+2023}; \citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \&
+Fachelli, 2021}; \citeproc{ref-torche_intergenerational_2014}{Torche,
+2014}). Social origins and destinations affect attitudes toward
+inequality in distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske,
+2017}; \citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
+2016b}, \citeproc{ref-gugushvili_subjective_2017}{2017};
+\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
+2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
+movement between these positions exposes individuals to different
+experiences and mechanisms that shape their views on what is fair
+(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
+research, examining the effects of social mobility on market justice
+preferences can help to illuminate how inequalities in access to social
+services are justified among individuals who have experienced, or not,
+changes in their social standing, and what are the normative mechanisms
+that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
+2015}).
+
+Beyond their isolated effects, social mobility and meritocratic beliefs
+interact in complex ways to shape market justice preferences. A key
+mechanism proposed in the literature to explain how mobility influences
+distributive justice preferences is the psychological process of
+self-serving bias in causal attribution
+(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
+\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
+suggests that individuals attribute failures---such as downward
+mobility---to external factors, while crediting successes---such as
+upward mobility---to their own merit and effort
+(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
+who experience upward mobility tend to view their social position as
+earned, making them more likely to believe that individuals are
+responsible for their own success or failure. Research shows that upward
+mobility is associated with weaker preferences for redistribution
 (\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
-\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001};
+\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
 \citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
 Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
-2011}), system legitimacy, and meritocratic beliefs
-(\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
-\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}), as well as on
-tolerance for economic inequality
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), attributions
-about poverty and wealth (\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
-\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}),
-and support for welfare policies and market economies
-(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
-\citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}). Thus, building on
-these contributions, expanding the analysis of the effects of social
-mobility on market justice preferences could contribute to this emerging
-literature and shed light on the mechanisms behind the justification of
-inequalities in access to social services.
-
-Market justice preferences are shaped not only by objective
-socioeconomic factors but also by perceptions and beliefs about
-meritocracy. Meritocracy asserts that inequality is an inherent feature
-of societies but can be legitimized through principles such as effort
-and talent (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
-\citeproc{ref-young_rise_1958}{Young, 1958}). Previous studies have
-shown that individuals with stronger meritocratic beliefs tend to
-perceive less inequality, attributing economic differences to individual
-achievements (\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}). Additionally, these
-beliefs justify greater inequality, as they are associated with
-attitudes that uphold the legitimization of social differences
-(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In unequal
-societies where the distribution of goods and services is predominantly
-governed by market logic, such beliefs can play a critical role in the
-acceptance of social inequalities. Recent research by Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) demonstrates that
-students in Chile who believe that effort and talent are rewarded in
-their country hold stronger preferences for market justice in access to
-healthcare, pensions, and education. By incorporating perceptions of
-meritocracy into this study, we can better understand how preferences
-toward market justice are shaped not only by an individual's structural
-social position but also by beliefs about the justification of these
-inequalities. Moreover, these beliefs are often solidified early in
-life, reinforced by institutions that promote values such as effort and
-individual skills as means to upward social mobility
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).
-
-Beyond the isolated effects of social mobility and beliefs in
-meritocracy on market justice preferences, these factors interact to
-shape these preferences in a complex way. One proposed mechanism in the
-literature regarding the process of social mobility that leads to
-differences in distributive justice preferences is the psychological
-mechanism of self-serving bias in causal attribution
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). Self-serving
-bias suggests that people attribute failures, such as downward mobility,
-to factors beyond their control, while explaining successes, such as
-upward mobility, by pointing to their own merits and effort
-(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}).
-Individuals who experience upward mobility tend to view their position
-as a result of their own effort and skills, making them more likely to
-support the idea that individuals are responsible for their own success
-or failure. Research shows that upward mobility is associated with lower
-redistribution preferences
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}) and stronger
-legitimacy of income inequality
+2011}) and stronger legitimacy of income inequality
 (\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
-individuals who experience downward mobility tend to attribute their
-failure to external factors, such as inequality, and are more supportive
-of redistribution and the idea that individuals should not be blamed for
-their economic failure
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). Taken together,
-I suggest that meritocratic beliefs may mirror this self-serving bias
-mechanism, as they reinforce the tendency to justify one's social status
-as the result of personal merits, which is closely related to the
+individuals who experience downward mobility tend to blame structural
+factors like inequality and are more supportive of redistribution while
+rejecting merit-based explanations
+(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
+together, I argue that meritocratic beliefs may reflect---or
+reinforce---this self-serving attribution mechanism by legitimizing
+one's social status as the outcome of personal merit, closely tied to
 attribution bias.
 
-Focusing in Chile, this article aims to achieve two main objectives:
+Against this background, this article pursues two main objectives:
 first, to analyze the extent to which intergenerational social mobility
-affects market justice preferences on healthcare, pensions and
-education, and second, to examine how meritocratic beliefs may moderate
-this relationship. The central argument is that upward social mobility
-will have a positive effect on market justice preferences, while
-downward mobility will have a negative effect. Additionally,
-meritocratic beliefs will moderate this relationship by reflecting a
-self-serving bias mechanism of causal attribution.
+influences market justice preferences regarding healthcare, pensions,
+and education; and second, to examine how meritocratic beliefs may
+moderate this relationship. Building on a theoretical framework that
+emphasizes how neoliberal transformations---particularly through the
+privatization and commodification of key areas of social
+reproduction---have profoundly reshaped processes of subject formation
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}), the central argument is
+that upward mobility increases support for market justice preferences,
+while downward mobility decreases it. Moreover, meritocratic beliefs are
+expected to moderate this relationship by reflecting a self-serving
+attribution bias, whereby individuals justify their social position in
+terms of personal merit.
 
-The Chilean case is particularly intriguing for studying market justice
-preferences. Despite being one of the most prosperous nations in Latin
-America, it has one of the highest levels of economic inequality in the
-region. The poorest 50\% of Chileans capture only 10\% of the total
-income and hold negative wealth, while the richest 1\% receives nearly
-27\% of the income and controls 49.6\% of the wealth
-(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This economic
-inequality exists alongside notable social mobility, with significant
-upward mobility in recent decades
-(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
-\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). Chile's
-social policy regime is characterized by a welfare model that relies
-heavily on private provision in healthcare, pensions, and education
-systems, often segmented by individuals' ability to pay and heavily
-dependent on state subsidies (\citeproc{ref-boccardo_30_2020}{Boccardo,
-2020}).
+This study focuses on Chile, a particularly intriguing case for
+examining market justice preferences. Despite significant economic
+growth and poverty reduction, Chile has some of the highest levels of
+inequality in Latin America and among OECD countries
+(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
+\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
+coexists with short-range upward mobility among lower-class segments
+moving into middle strata, though strong barriers remain to reaching
+higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
+et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
+2014}). What makes Chile especially salient is that much of this
+inequality is rooted in deep neoliberal reforms that institutionalized
+the privatization and commodification of key social sectors
+(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
+during the dictatorship (1973--1989) and expanded in democracy, these
+reforms enabled the unprecedented emergence of markets in health,
+pensions, and education, with provision segmented by individuals'
+ability to pay and supported by public subsidies
+(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
+despite waves of protest against inequality and commodification from
+2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
+2021})---Chilean subjectivities have been increasingly shaped by
+neoliberal discourses and market logics, influencing their attitudes
+toward inequality and welfare distribution
+(\citeproc{ref-arteagaaguirre_politicas_2015}{Arteaga Aguirre \& Iñigo
+Valderrama, 2015}; \citeproc{ref-castillo_socialization_2024}{Castillo
+et al., 2024}).
 
-Against this backdrop, the main research question is: To what extent
-does intergenerational social mobility affect market justice
-preferences, and how do meritocratic beliefs explain this relationship
-in Chile? This study aims to contribute to the literature by providing
-evidence from a Latin American developing country, highlighting how
-changes in individuals' socioeconomic position and meritocratic beliefs
-influence market justice preferences in critical areas like healthcare,
-pensions, and education.
+In this context, the questions that guide this research are as follows:
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\tightlist
+\item
+  To what extent does intergenerational social mobility influence market
+  justice preferences regarding healthcare, pensions, and education in
+  Chile?
+\item
+  How do meritocratic beliefs condition or moderate this relationship in
+  the Chilean context?
+\end{enumerate}
+
+To address these questions, this study draws on large-scale,
+representative survey data collected in 2018 from the urban Chilean
+population aged 18 to 75 (n = 2,983). The next section outlines the
+theoretical framework linking market justice preferences, social
+mobility, and meritocratic beliefs, and proposes a set of hypotheses.
+This is followed by a description of the data, variables, and analytical
+strategy. The final sections present the empirical findings, offer an
+interpretation of the results, and conclude with a discussion of their
+implications.
+
+\section{Theoretical and empirical
+background}\label{theoretical-and-empirical-background}
+
+\subsection{Market justice
+preferences}\label{market-justice-preferences}
+
+\subsection{Social mobility}\label{social-mobility}
+
+\subsection{Meritocracy}\label{meritocracy}
+
+\subsection{The Chilean context}\label{the-chilean-context}
 
@@
 
+\bibitem[\citeproctext]{ref-arteagaaguirre_politicas_2015}
+Arteaga Aguirre, C., \& Iñigo Valderrama, I. (2015). {Pol{í}ticas
+sociales, modelo de desarrollo y subjetividad de grupos vulnerables en
+Chile}.
+
 \bibitem[\citeproctext]{ref-batruch_belief_2023}
@@
 
-\bibitem[\citeproctext]{ref-benabou_social_2001}
-Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
-{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
-Economics}, \emph{116}(2), 447--487.
-\url{https://doi.org/10.1162/00335530151144078}
-
 \bibitem[\citeproctext]{ref-boccardo_30_2020}
@@
 que la pandemia revel{ó}} (Nodo XXI). Santiago.
-
-\bibitem[\citeproctext]{ref-breen_social_2004}
-Breen, R. (2004). \emph{Social {Mobility} in {Europe}}. OUP Oxford.
 
@@
 
-\bibitem[\citeproctext]{ref-bucca_merit_2016}
-Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
-Latin Americans}' beliefs about wealth and poverty. \emph{Research in
-Social Stratification and Mobility}, \emph{44}, 98--112.
-\url{https://doi.org/10.1016/j.rssm.2016.02.005}
-
 \bibitem[\citeproctext]{ref-busemeyer_skills_2014}
@@
 Welfare States}}. Cambridge University Press.
-
-\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
-Busemeyer, M. R., \& Iversen, T. (2020). The {Welfare State} with
-{Private Alternatives}: {The Transformation} of {Popular Support} for
-{Social Insurance}. \emph{The Journal of Politics}, \emph{82}(2),
-671--686. \url{https://doi.org/10.1086/706980}
 
@@
 
+\bibitem[\citeproctext]{ref-ferre_welfare_2023}
+Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
+America}. \emph{Journal of International and Comparative Social Policy},
+\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}
+
+\bibitem[\citeproctext]{ref-flores_top_2020}
+Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
+in {Chile}: {A Historical Perspective} on {Income Inequality},
+1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
+\url{https://doi.org/10.1111/roiw.12441}
+
 \bibitem[\citeproctext]{ref-gingrich_making_2011}
@@
 
+\bibitem[\citeproctext]{ref-harvey_breve_2015}
+Harvey, D. (2015). \emph{{Breve historia del neoliberalismo}}. Madrid
+(Espa{ñ}a): Ediciones Akal.
+
 \bibitem[\citeproctext]{ref-immergut_it_2020}
@@
 
+\bibitem[\citeproctext]{ref-jacksonActorsInstitutions2010}
+Jackson, G. (2010). Actors and institutions. In G. Morgan, J. L.
+Campbell, C. Crouch, O. K. Pedersen, \& R. Whitley (Eds.), \emph{The
+{Oxford Handbook} of {Comparative Institutional Analysis}} (1st ed.).
+Oxford University Press.
+\url{https://doi.org/10.1093/oxfordhb/9780199233762.001.0001}
+
 \bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
@@
 \url{https://doi.org/10.1111/1468-4446.12363}
-
-\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
-of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
-about {Market Justice}, 1991-1996. \emph{European Sociological Review},
-\emph{15}(3), 251--283. Retrieved from
-\url{https://www.jstor.org/stable/522731}
 
@@
 
-\bibitem[\citeproctext]{ref-mau_moral_2014}
-Mau, S. (2014). \emph{The moral economy of welfare states: {Britain} and
-{Germany} compared} (First issued in paperback). London New York NY:
-Routledge \& Francis Group.
+\bibitem[\citeproctext]{ref-madariaga_three_2020}
+Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
+economic policy trajectory in comparative perspective.
+\emph{Contemporary Politics}, \emph{26}(3), 308--329.
+\url{https://doi.org/10.1080/13569775.2020.1735021}
+
+\bibitem[\citeproctext]{ref-mau_inequality_2015}
+Mau, S. (2015). \emph{Inequality, {Marketization} and the {Majority
+Class}: {Why Did} the {European Middle Classes Accept Neo-Liberalism}?}
+Houndmills: Palgrave Macmillan.
 
@@
 
-\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
-Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
-of opportunity. \emph{Research in Social Stratification and Mobility},
-\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}
-
-\bibitem[\citeproctext]{ref-salamon_marketization_1993}
-Salamon, L. M. (1993). The {Marketization} of {Welfare}: {Changing
-Nonprofit} and {For-Profit Roles} in the {American Welfare State}.
-\emph{Social Service Review}, \emph{67}(1), 16--39.
-\url{https://doi.org/10.1086/603963}
-
 \bibitem[\citeproctext]{ref-schmidt_experience_2011}
@@
 \emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}
+
+\bibitem[\citeproctext]{ref-somma_no_2021}
+Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
+water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
+Movement Studies}, \emph{20}(4), 495--502.
+\url{https://doi.org/10.1080/14742837.2020.1727737}
 

### 2025-05-21 processing/01-proc-data.R

@@ -1,0 +1,659 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               naniar,
+               data.table,
+               summarytools,
+               nnet,
+               occupar)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
[…]

### 2025-05-21 processing/elsoc_ciuo08.R

@@ -1,0 +1,76 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for CIUO08 ELSOC
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               naniar,
+               data.table,
+               summarytools,
+               httr,
+               occupar)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+# 3.1 Glosa INE ----
+
+elsoc_long_2016_2023 %>% 
+  group_by(ola) %>% 
+  tally(!is.na(m57))
+
+df_glosa <- elsoc_long_2016_2023 %>% 
+  filter(ola == 6)
+
+request <-  httr::POST("https://rapps.ine.cl:9292/predict", 
+                       encode = "json",
+                       body =  list(text = df_glosa$m57,
+                                    classification = "ciuo",
+                                    digits = 2))
+
+httr::status_code(request)
+
+ciuo08 <- httr::content(request, encoding = "text")
+
+ciuo08_predicciones <- data.frame(
+  cod_final = unlist(ciuo08$cod_final),
+  prob = as.numeric(unlist(ciuo08$prob))
+)
+
+ciuo08_predicciones <- ciuo08_predicciones %>% 
+  mutate(
+    isco08_ine = (as.numeric(cod_final)*100),
+    isei_08 = occupar::isco08toISEI08(isco08_ine),
+    isei_92 = occupar::isco08toISEI92(isco08_ine)
+  )
+
[…]

### 2025-05-27 advances/informe2.qmd
up inf2

@@
 ---
-title: "Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
+title: "Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
 format:
@@
 author:
-  - name: Andreas Laffert
+  - name: Andreas Laffert Tamayo
     orcid: 0000-0002-9008-2454
@@
 
-What makes market-driven inequalities appear legitimate in the eyes of the public?
+What is the legitimate extent of market inequality in the eyes of the public?
 Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016].
@@
 Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a].
-Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that access to services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
+Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
 Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].
@@
 Existing literature shows that market justice preferences are shaped by both the economic and institutional context and individuals' positions within social stratification.
-Grounded in the notion that institutions influence individual attitudes [@jacksonActorsInstitutions2010], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015].
+Grounded in the notion that economic institutions influence people's normative attitudes [@jacksonActorsInstitutions2010], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015].
 In such contexts, market justice preferences tend to rise as individuals "ascend" the social structure, with those in more privileged positions in terms of class, education, and income holding stronger preferences for market-based solutions compared to those in more disadvantaged or at-risk positions [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_socialization_2024; @otero_power_2024; @lindh_public_2015; @svallfors_political_2007].
@@
 
-Although it is clear that one's social position influences market preferences, the question of how upward or downward mobility within the social structure affects these preferences remains unanswered.
+Although it is clear that one's social position influences market justice preferences, the question of how upward or downward mobility within the social structure affects these preferences remains unanswered.
 This question is far from trivial, especially in Latin America, where many have experienced various forms of mobility amid high economic inequality and deep welfare privatization [@lopez-roldan_comparative_2021; @torche_intergenerational_2014; @ferre_welfare_2023].
@@
 Beyond their isolated effects, social mobility and meritocratic beliefs interact in complex ways to shape market justice preferences.
-A key mechanism proposed in the literature to explain how mobility influences distributive justice preferences is the psychological process of self-serving bias in causal attribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c].
+Among others, a key mechanism proposed in the literature to explain how mobility influences distributive justice preferences is the psychological process of self-serving bias in causal attribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c].
 This bias suggests that individuals attribute failures---such as downward mobility---to external factors, while crediting successes---such as upward mobility---to their own merit and effort [@miller_selfserving_1975].
@@
 In contrast, individuals who experience downward mobility tend to blame structural factors like inequality and are more supportive of redistribution while rejecting merit-based explanations [@gugushvili_trends_2014].
-Taken together, I argue that meritocratic beliefs may reflect---or reinforce---this self-serving attribution mechanism by legitimizing one's social status as the outcome of personal merit, closely tied to attribution bias.
+Taken together, I argue that meritocratic beliefs may reinforce this self-serving attribution mechanism by legitimizing one's social status as the outcome of personal merit, closely tied to attribution bias.
 
@@
 Introduced during the dictatorship (1973--1989) and expanded in democracy, these reforms enabled the unprecedented emergence of markets in health, pensions, and education, with provision segmented by individuals' ability to pay and supported by public subsidies [@boccardo_30_2020].
-In parallel---and despite waves of protest against inequality and commodification from 2006 to 2019 [@somma_no_2021]---Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing their attitudes toward inequality and welfare distribution [@arteagaaguirre_politicas_2015; @castillo_socialization_2024].
+In parallel---and despite waves of protest against inequality and commodification from 2006 to 2019 [@somma_no_2021]---Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing their attitudes toward inequality and welfare distribution [@araujo_desafios_2012; @canalesceron_sujeto_2021].
 
@@
 
-To address these questions, this study draws on large-scale, representative survey data collected in 2018 from the urban Chilean population aged 18 to 75 (n = 2,983).
+To address these questions, this study draws on large-scale, representative survey data collected in 2018 from the urban Chilean population aged 18 to 75 (n = 2,726).
 The next section outlines the theoretical framework linking market justice preferences, social mobility, and meritocratic beliefs, and proposes a set of hypotheses.
@@
 
+Although redistribution typically refers to the state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards (Koos & Sachweh, 2019; Lindh & McCall, 2023).
+As Polanyi (1957) observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks.
+These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking (Koos & Sachweh, 2019; Immergut, 1998).
+In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions (Mau, 2015; Svallfors, 2006).
+While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality (Lindh & McCall, 2023).
+In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services (Gingrich, 2011; Ferre, 2023).
+Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality (Busemeyer & Iversen, 2020).
+This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities (Lindh & McCall, 2023; Mau, 2015).
+
+The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles.
+Beyond stratification structures and socioeconomic conditions, the study of inequality also examines the beliefs individuals hold about the causes of social inequality, the normative principles that underpin those beliefs, the factors that shape them, and their consequences for attitudes and behavior (Kluegel & Smith, 1981).
+In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society (Jasso et al., año).
+This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate (Castillo, 2011).
+One prominent strand investigates the legitimacy of wage inequality, particularly salary differences across occupations (Jasso, 1978; Castillo, 2011; Wegener, 1999).
+Another examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, and pensions.
+Here, legitimacy stems from the belief that access to these goods should follow market-based criteria (Castillo et al., 2024; Lindh, 2015).
+In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic (Busemeyer & Iversen, 2020).
+
+Market justice preferences refer to normative beliefs that legitimize the idea that access to core social services—such as healthcare, education, or pensions—should be determined by market-based criteria.
+Following Janmaat’s (2013, p.
+359) distinction, these preferences fall under the category of “beliefs,” understood as normative ideas about what inequality should look like, as opposed to “perceptions,” which refer to subjective evaluations of existing inequality.
+Market justice preferences reflect the view that access to these services should depend on individuals’ ability to pay, thus justifying inequalities generated by market mechanisms (Kluegel & Smith, 1999; Lindh, 2015).
+The concept draws on Lane’s (1968) classic contrast between market justice and political justice: the former is grounded in the idea of earned deserts—where rewards reflect effort, productivity, and skill—while the latter prioritizes need and equality, typically expressed in welfare state policies.
+Lane argued that markets and states differ in purpose (efficiency vs.
+need), logic (individual vs.
+collective), and fairness criteria (merit vs.
+equality).
+Market justice assumes that markets are neutral, self-regulating systems in which fair procedures yield outcomes proportional to merit.
+Inequality, from this perspective, is not only expected but legitimate—as long as it arises from fair competition.
+In this way, market justice offers a moral lens through which individuals can view the commodification and stratified access to social services as fair and justified (Kluegel & Smith, 1999; Lindh, 2015).
+
+The empirical study of market justice preferences has relied on diverse strategies to operationalize how individuals evaluate inequalities produced by market allocation.
+One common approach assesses whether people consider it fair that access to key social services—such as healthcare, education, or pensions—depends on income.
+This strategy builds on foundational work by Kluegel and Smith (1981), who explored the normative justifications behind support for economic inequality.
+More recent studies have expanded this logic beyond income to include market-mediated access to welfare goods.
+For instance, Immergut and Schneider (2020) and von dem Knesebeck et al.
+(2016) examine whether respondents believe it is just that those with higher incomes receive better healthcare.
+Similarly, Lee and Stacey (2023) apply this approach to educational opportunities.
+These studies typically use targeted survey items that ask respondents to evaluate specific distributive scenarios, allowing for comparative analyses across countries or welfare regimes.
+Complementing these efforts, more recent research has introduced composite indicators designed to capture broader orientations toward market-based allocation.
+For example, Castillo et al.
+(2024) propose a single-item measure that summarizes individual support for income-based access across multiple domains—health, education, and pensions—within the Chilean context.
+Together, these instruments aim to capture to what extent individuals perceive market-generated inequality as justified.
+
+Comparative empirical research has identified several individual-level factors that influence support for market justice.
+Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles (Svallfors, 2007; Koos & Sachweh, 2019; Lindh, 2015).
+For example, Lindh (2015) finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries.
+In a comparative analysis, Svallfors (2007) observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class.
+Busemeyer (2014) similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider (2020) and von dem Knesebeck et al.
+(2016) report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage.
+In Chile, Otero and Mendoza (2024) show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions.
+Beliefs about inequality and political orientation also matter.
+In Chile, Castillo et al.
+(2024) show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution, while Lee and Stacy (2023) in Australia suggest that these preferences are also greater as people lean toward economic conservatism.
+In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.
+
+However, individual characteristics alone do not fully explain variation in support for market justice.
+A growing body of cross-national research highlights the importance of institutional arrangements in shaping these preferences.
+For instance, Immergut and Schneider (2020) find that in countries with higher public spending on healthcare, individuals are less likely to view income-based access as fair.
+Similarly, Busemeyer (2014) shows that increased public investment in education is associated with lower support for privatized provision.
+Conversely, Lindh (2015) finds that in countries with more market-oriented welfare systems, support for market-based distribution tends to be higher, suggesting that individual attitudes often align with institutional outputs.
+These findings are consistent with neo-institutionalist and policy feedback theories (Campbell, 2012), which argue that institutions do more than redistribute resources—they also shape the normative categories through which individuals assess who is deserving of support (Immergut, 1990).
+In this view, dominant values and preferences are both shaped by and embedded in institutional configurations (Busemeyer, 2014), reinforcing the notion that institutions are not neutral structures, but active producers of the moral frameworks that legitimize or challenge inequality.
+
 ## Social mobility
 
+The study of social mobility, its drivers and consequences has long been central to sociology.
+Classic theorists explored not only movement across social hierarchies but also its broader implications for class conflict, norm stability, and institutional change (Breen & Ermisch, 2024).
+Sorokin (1927) introduced the concept formally, defining mobility as the shift of individuals, values, or objects between positions within a stratification system, and distinguishing between horizontal and vertical forms.
+Later work differentiated intergenerational from intragenerational mobility, as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which origins constrain destinations (Eyles et al., 2022).
+In Latin America, research shows high absolute but low relative mobility (Bucca, 2016): although educational expansion and economic modernization have enabled some upward movement, status reproduction remains strong, especially among elites (Torche, 2015; López-Roldán & Fachelli, 2021).
+This reflects deep structural inequality, segmented education systems, stratified labor markets, and legacies of dependent development that restrict access to mobility channels and reinforce the intergenerational transmission of advantage.
+
+Beyond mapping mobility patterns, growing research has examined its subjective and attitudinal effects.
+Mobility effects—defined as outcomes resulting from movement between origin and destination classes (Breen & Ermisch, 2024)—have long attracted theoretical interest.
+Sorokin’s (1959) dissociative hypothesis posits that mobility, whether upward or downward, may produce psychological strain due to conflicting norms between class contexts, leading to lower life satisfaction.
+Similarly, Lenski (1954) argued that status inconsistency—mismatches among education, income, and occupation—can undermine well-being.
+These ideas underpin extensive empirical work linking intergenerational mobility to outcomes such as life satisfaction, mental health, and stress (Hadjar, Gugushvili, Präg & Gugushvili).
+
+Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality.
+A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed.
+Alesina et al.
+(2018) show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies.
+Similarly, Ares (2020) finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility.
+Comparative studies by Schmidt (2011) and Gugushvili (2016) likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support.
+However, recent causal evidence from Breen and Ermisch (2024) suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.
+
+Beyond redistribution, other studies have explored the impact of mobility on broader beliefs about inequality.
+Gugushvili (2016) finds that upwardly mobile individuals are more likely to adopt individualistic attributions of poverty and to legitimize income inequality, particularly in post-socialist societies.
+Similarly, Bucca (2016) shows that subjective upward mobility reinforces individualistic explanations of wealth in seven Latin American countries.
+At the macro level, Shariff et al.
+(2016) demonstrate that higher levels of national economic mobility correlate with greater tolerance of inequality.
+In contrast, Day and Fiske (2019) find that low perceived mobility undermines belief in meritocracy and a just world, thereby weakening system justification.
+Taken together, this body of research suggests that mobility shapes attitudes toward inequality and justice through multiple, and sometimes contradictory, mechanisms.
+
+The literature on the effects of social mobility has proposed various mechanisms to explain how and why changes in social position may influence individual outcomes (Freyr & Rhem, año).
+One of the most prominent is the self-interest mechanism, which posits that individuals who experience upward or downward mobility undergo a shift in their material interests, thereby altering their perceptions and preferences (Ares, 2020; Langsæther, Evans & O’Grady, 2021; Freyr & Rhem, año).
+Closely related to this logic is the Prospect of Upward Mobility (POUM) hypothesis, which suggests that individuals may oppose redistribution not because of their current position, but because they anticipate improving their status in the future (REF).
+A second line of explanation draws on the framework of theories of socialization.
+Within this tradition, hypotheses such as acculturation, socialization, and status maximization propose that individuals adjust their attitudes based on the norms and values of either their class of origin or their destination, or a combination of both (Jaime-Castillo & Marqués-Perales, 2019).
+However, most of these mechanisms focus primarily on the indirect effects of origin and destination positions, rather than on the direct effect of experience of movement itself.
+In response to this gap, recent research has highlighted the role of self-serving bias in causal attribution processes, suggesting that individuals tend to explain their mobility trajectories in ways that justify their current position, which in turn shapes their beliefs and preferences (Schmidt, 2011; Gugushvili, 2016; Molina, Bucca & Macy, 2019).
+
+The self-serving bias mechanism builds on a value-oriented perspective, emphasizing that individuals’ experiences of social mobility shape their causal attributions, which in turn influence their beliefs about justice and distributive preferences (Gugushvili, 2014).
+Causal attribution refers to the process through which individuals generate explanations for their own behavior and outcomes, as well as those of others (Gugushvili, year).
+In this view, people’s interpretations of economic inequality depend on whether they believe such disparities reflect unequal individual contributions.
+Individuals who adopt an internal attribution framework tend to see success or failure as rooted in personal characteristics such as effort, talent, or merit.
+In contrast, those who regard inequality as unjust are more likely to adopt an external attribution model, viewing outcomes as the result of structural barriers beyond individual control (Kluegel & Smith, 1999).
+This mechanism, often described as intrapersonal causal attribution, reflects how people explain their own socioeconomic positions—typically attributing their successes to internal qualities while blaming failures on external circumstances (Miller & Ross, 1975).
+Over time, individuals may revise their beliefs and attitudes: while early views are shaped by their social origin, these are later adjusted in light of personal experiences of mobility and the perceived role of ascribed versus achieved factors in determining socioeconomic outcomes (Gugushvili, year).
+
+Empirical research has provided support for the self-serving bias mechanism in shaping redistributive preferences and attitudes toward the legitimacy of inequality.
+For instance, Schmidt (2011) finds that individuals who experience upward mobility are more likely to interpret their success as the result of personal effort or merit, and consequently perceive redistribution as less necessary.
+Conversely, individuals who experience downward mobility tend to attribute their decline to external circumstances—such as structural inequality or unemployment—and show stronger support for redistribution.
+In a comparative analysis across different welfare domains, Gugushvili demonstrates that upward mobility is associated with lower support for government spending on housing and pensions, while individuals who experience downward mobility express lower support for healthcare and education spending, but favor increased investment in housing and pensions, reflecting the material nature of these domains.
+Moreover, Gugushvili (2016) finds that upward mobility is linked to greater justification of income inequality, suggesting that improved social standing reinforces an attributional view in which success is seen as the result of individual characteristics—thus legitimizing inequality as a fair outcome.
+
+These findings suggest that causal attribution processes mediate the relationship between mobility experiences and justice-related attitudes. Accordingly, this study proposes the following hypothesis:
+
+$H1$: Experiencing upward (downward) social mobility is positively (negatively) associated with greater support for market justice in healthcare, pensions, and education.
+
 ## Meritocracy
+
+Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs.
+Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
+As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality.
+However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020].
+From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality.
+People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023].
+Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust.
+This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.
+
+I adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions.
+This distinction is essential, as it captures different pathways through which individuals justify inequality.
+Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility.
+A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed.
+Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023].
+The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions.
+For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023].
+This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage.
+Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.
+
+These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
+This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025].
+Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
+Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].
+
+In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups.
+Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
+This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
+Moreover, @busemeyer_positive_2021 argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
+This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
+
+Importantly, recent research has explored how meritocratic beliefs interact with experiences of intergenerational mobility to shape distributive attitudes.
+The belief that one’s success is earned can lead upwardly mobile individuals to internalize meritocratic narratives and justify existing inequalities, reinforcing support for market justice (Gugushvili, 2016; Molina et al., 2019).
+Conversely, downwardly mobile individuals who maintain strong meritocratic beliefs may interpret their status as a personal failure, reducing their support for redistribution (Day & Fiske, 2019).
+At the macro level, Shariff et al.
+(2016) show that higher perceived mobility increases tolerance for inequality, suggesting that meritocracy and mobility are mutually reinforcing.
+
+Taken together, this literature supports the idea that meritocratic beliefs can moderate the relationship between mobility and market justice preferences.
+Individuals who experience mobility—especially upward—may draw on meritocratic narratives to legitimize both their own status and broader inequalities, thereby strengthening their support for market-based distribution.
+Accordingly, the second hypothesis of this study is:
+
+$H2$: The positive (negative) relationship between upward (downward) social mobility and support for market justice in healthcare, pensions, and education is moderated by meritocratic beliefs; specifically, this association is stronger (weaker) among individuals with higher perceptions of meritocracy.
 
@@
 
-    -   Upper (codes 100--299: managers, directors, professionals)
-    -   Middle (codes 300--499: skilled white‑collar workers)
-    -   Lower (codes 500 and above: manual, semi‑skilled, or unskilled workers)
+    -   Upper (codes 100-299: managers, directors, professionals)
+    -   Middle (codes 300-499: skilled white-collar workers)
+    -   Lower (codes 500 and above: manual, semi-skilled, or unskilled workers)
 

### 2025-05-27 advances/informe2.tex
up inf2

@@
 \hypersetup{
-  pdftitle={Preferences for market justice: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
-  pdfauthor={Andreas Laffert},
+  pdftitle={Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
+  pdfauthor={Andreas Laffert Tamayo},
   colorlinks=true,
@@
 
-\title{Preferences for market justice: the role of intergenerational
+\title{Preferences for mercantilization: the role of intergenerational
 social mobility and beliefs in meritocracy in Chile}
 
-  \author{Andreas Laffert}
+  \author{Andreas Laffert Tamayo}
             \affil{%
@@
 
-What makes market-driven inequalities appear legitimate in the eyes of
-the public? Since the early 1980s, many countries have experienced a
+What is the legitimate extent of market inequality in the eyes of the
+public? Since the early 1980s, many countries have experienced a
 widespread retreat from universal welfare programs and a shift toward
@@
 this framework, a growing body of research examines the extent to which,
-and the mechanisms by which, citizens consider it fair that access to
-services like health care, pensions, and education be governed by
-market-based criteria---a phenomenon known as \emph{market justice
-preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+and the mechanisms by which, citizens consider it fair that the
+allocation of services like health care, pensions, and education be
+governed by market-based criteria---a phenomenon known as \emph{market
+justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
+2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
+2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
 \citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
@@
 both the economic and institutional context and individuals' positions
-within social stratification. Grounded in the notion that institutions
-influence individual attitudes
+within social stratification. Grounded in the notion that economic
+institutions influence people's normative attitudes
 (\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
@@
 Although it is clear that one's social position influences market
-preferences, the question of how upward or downward mobility within the
-social structure affects these preferences remains unanswered. This
-question is far from trivial, especially in Latin America, where many
-have experienced various forms of mobility amid high economic inequality
-and deep welfare privatization (\citeproc{ref-ferre_welfare_2023}{Ferre,
-2023}; \citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \&
-Fachelli, 2021}; \citeproc{ref-torche_intergenerational_2014}{Torche,
-2014}). Social origins and destinations affect attitudes toward
-inequality in distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske,
-2017}; \citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
-2016b}, \citeproc{ref-gugushvili_subjective_2017}{2017};
+justice preferences, the question of how upward or downward mobility
+within the social structure affects these preferences remains
+unanswered. This question is far from trivial, especially in Latin
+America, where many have experienced various forms of mobility amid high
+economic inequality and deep welfare privatization
+(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
+\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
+2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}).
+Social origins and destinations affect attitudes toward inequality in
+distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
+\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b},
+\citeproc{ref-gugushvili_subjective_2017}{2017};
 \citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
@@
 Beyond their isolated effects, social mobility and meritocratic beliefs
-interact in complex ways to shape market justice preferences. A key
-mechanism proposed in the literature to explain how mobility influences
-distributive justice preferences is the psychological process of
-self-serving bias in causal attribution
+interact in complex ways to shape market justice preferences. Among
+others, a key mechanism proposed in the literature to explain how
+mobility influences distributive justice preferences is the
+psychological process of self-serving bias in causal attribution
 (\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
@@
 (\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
-together, I argue that meritocratic beliefs may reflect---or
-reinforce---this self-serving attribution mechanism by legitimizing
-one's social status as the outcome of personal merit, closely tied to
-attribution bias.
+together, I argue that meritocratic beliefs may reinforce this
+self-serving attribution mechanism by legitimizing one's social status
+as the outcome of personal merit, closely tied to attribution bias.
 
@@
 toward inequality and welfare distribution
-(\citeproc{ref-arteagaaguirre_politicas_2015}{Arteaga Aguirre \& Iñigo
-Valderrama, 2015}; \citeproc{ref-castillo_socialization_2024}{Castillo
-et al., 2024}).
+(\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli, 2012};
+\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
 
@@
 representative survey data collected in 2018 from the urban Chilean
-population aged 18 to 75 (n = 2,983). The next section outlines the
+population aged 18 to 75 (n = 2,726). The next section outlines the
 theoretical framework linking market justice preferences, social
@@
 
+Although redistribution typically refers to the state's capacity to
+reallocate resources from the advantaged to the vulnerable, market
+institutions also play a central role in distributing socially valuable
+goods and rewards (Koos \& Sachweh, 2019; Lindh \& McCall, 2023). As
+Polanyi (1957) observed, economic integration in capitalist societies is
+primarily organized through market exchange, governed by a
+self-regulating price system embedded in institutional frameworks. These
+institutions are not mere aggregates of individual behavior, but social
+realities endowed with rules, mechanisms, and normative meanings that
+shape everyday thinking (Koos \& Sachweh, 2019; Immergut, 1998). In this
+sense, the economic order is mirrored in a moral economy: collectively
+shared norms and beliefs about justice in distribution, embedded and
+reinforced through institutions (Mau, 2015; Svallfors, 2006). While most
+research from this perspective has focused on welfare institutions,
+recent scholarship has brought the market back into focus as a site of
+distributive justice beliefs and institutional responses to inequality
+(Lindh \& McCall, 2023). In many countries, privatization and
+commodification have expanded market logic into core areas of social
+reproduction, such as healthcare, education, pensions, and social
+security, deepening inequality in access to these services (Gingrich,
+2011; Ferre, 2023). Yet public support for market-based welfare
+provision has grown, even in traditional welfare states, particularly
+among higher-income groups who view private alternatives as more
+efficient or higher in quality (Busemeyer \& Iversen, 2020). This shift
+calls for broader inquiry into how institutions like markets structure
+access to resources and legitimize inequalities (Lindh \& McCall, 2023;
+Mau, 2015).
+
+The legitimacy of market-based inequalities is closely tied to beliefs
+about distributive justice grounded in market principles. Beyond
+stratification structures and socioeconomic conditions, the study of
+inequality also examines the beliefs individuals hold about the causes
+of social inequality, the normative principles that underpin those
+beliefs, the factors that shape them, and their consequences for
+attitudes and behavior (Kluegel \& Smith, 1981). In this regard,
+empirical research on distributive justice focuses on individuals'
+conceptions of how goods and rewards \emph{should} be distributed in
+society (Jasso et al., año). This line of inquiry allows for examining
+the extent to which economic inequality is perceived as just, or in a
+certain sense, legitimate (Castillo, 2011). One prominent strand
+investigates the legitimacy of wage inequality, particularly salary
+differences across occupations (Jasso, 1978; Castillo, 2011; Wegener,
+1999). Another examines how individuals justify market-generated
+inequalities in access to core social services such as healthcare,
+education, and pensions. Here, legitimacy stems from the belief that
+access to these goods should follow market-based criteria (Castillo et
+al., 2024; Lindh, 2015). In such views, these services are treated as
+legitimate commodities; goods that can be traded, priced, and evaluated
+through market logic (Busemeyer \& Iversen, 2020).
+
+Market justice preferences refer to normative beliefs that legitimize
+the idea that access to core social services---such as healthcare,
+education, or pensions---should be determined by market-based criteria.
+Following Janmaat's (2013, p.~359) distinction, these preferences fall
+under the category of ``beliefs,'' understood as normative ideas about
+what inequality should look like, as opposed to ``perceptions,'' which
+refer to subjective evaluations of existing inequality. Market justice
+preferences reflect the view that access to these services should depend
+on individuals' ability to pay, thus justifying inequalities generated
+by market mechanisms (Kluegel \& Smith, 1999; Lindh, 2015). The concept
+draws on Lane's (1968) classic contrast between market justice and
+political justice: the former is grounded in the idea of earned
+deserts---where rewards reflect effort, productivity, and skill---while
+the latter prioritizes need and equality, typically expressed in welfare
+state policies. Lane argued that markets and states differ in purpose
+(efficiency vs.~need), logic (individual vs.~collective), and fairness
+criteria (merit vs.~equality). Market justice assumes that markets are
+neutral, self-regulating systems in which fair procedures yield outcomes
+proportional to merit. Inequality, from this perspective, is not only
+expected but legitimate---as long as it arises from fair competition. In
+this way, market justice offers a moral lens through which individuals
+can view the commodification and stratified access to social services as
+fair and justified (Kluegel \& Smith, 1999; Lindh, 2015).
+
+The empirical study of market justice preferences has relied on diverse
+strategies to operationalize how individuals evaluate inequalities
+produced by market allocation. One common approach assesses whether
+people consider it fair that access to key social services---such as
+healthcare, education, or pensions---depends on income. This strategy
+builds on foundational work by Kluegel and Smith (1981), who explored
+the normative justifications behind support for economic inequality.
+More recent studies have expanded this logic beyond income to include
+market-mediated access to welfare goods. For instance, Immergut and
+Schneider (2020) and von dem Knesebeck et al.~(2016) examine whether
+respondents believe it is just that those with higher incomes receive
+better healthcare. Similarly, Lee and Stacey (2023) apply this approach
+to educational opportunities. These studies typically use targeted
+survey items that ask respondents to evaluate specific distributive
+scenarios, allowing for comparative analyses across countries or welfare
+regimes. Complementing these efforts, more recent research has
+introduced composite indicators designed to capture broader orientations
+toward market-based allocation. For example, Castillo et al.~(2024)
+propose a single-item measure that summarizes individual support for
+income-based access across multiple domains---health, education, and
+pensions---within the Chilean context. Together, these instruments aim
+to capture to what extent individuals perceive market-generated
+inequality as justified.
+
+Comparative empirical research has identified several individual-level
+factors that influence support for market justice. Individuals in more
+advantaged socioeconomic positions---those with higher income,
+education, and occupational status---are consistently more likely to
+endorse market-based distributive principles (Svallfors, 2007; Koos \&
+Sachweh, 2019; Lindh, 2015). For example, Lindh (2015) finds that
+individuals from the service class are more likely to support
+market-based access to healthcare and education than skilled and
+unskilled workers across 17 relatively affluent countries. In a
+comparative analysis, Svallfors (2007) observes that this expected class
+pattern appears clearly only in Sweden, where support for private
+education and healthcare varies systematically by class. Busemeyer
+(2014) similarly shows that support for private education is stronger
+among high-income groups, while Immergut and Schneider (2020) and von
+dem Knesebeck et al.~(2016) report comparable findings for healthcare,
+suggesting that wealthier individuals perceive private provision as a
+means of maintaining relative advantage. In Chile, Otero and Mendoza
+(2024) show that individuals with higher income and university education
+express stronger support for market allocation in healthcare, education,
+and pensions. Beliefs about inequality and political orientation also
+matter. In Chile, Castillo et al.~(2024) show that individuals who have
+strong meritocratic beliefs are more likely to support market-based
+distribution, while Lee and Stacy (2023) in Australia suggest that these
+preferences are also greater as people lean toward economic
+conservatism. In this sense, market justice preferences are shaped by
+the interplay between structural position and normative reasoning.
+
+However, individual characteristics alone do not fully explain variation
+in support for market justice. A growing body of cross-national research
+highlights the importance of institutional arrangements in shaping these
+preferences. For instance, Immergut and Schneider (2020) find that in
+countries with higher public spending on healthcare, individuals are
+less likely to view income-based access as fair. Similarly, Busemeyer
+(2014) shows that increased public investment in education is associated
+with lower support for privatized provision. Conversely, Lindh (2015)
+finds that in countries with more market-oriented welfare systems,
+support for market-based distribution tends to be higher, suggesting
+that individual attitudes often align with institutional outputs. These
+findings are consistent with neo-institutionalist and policy feedback
+theories (Campbell, 2012), which argue that institutions do more than
+redistribute resources---they also shape the normative categories
+through which individuals assess who is deserving of support (Immergut,
+1990). In this view, dominant values and preferences are both shaped by
+and embedded in institutional configurations (Busemeyer, 2014),
+reinforcing the notion that institutions are not neutral structures, but
+active producers of the moral frameworks that legitimize or challenge
+inequality.
+
 \subsection{Social mobility}\label{social-mobility}
 
+The study of social mobility, its drivers and consequences has long been
+central to sociology. Classic theorists explored not only movement
+across social hierarchies but also its broader implications for class
+conflict, norm stability, and institutional change (Breen \& Ermisch,
+2024). Sorokin (1927) introduced the concept formally, defining mobility
+as the shift of individuals, values, or objects between positions within
+a stratification system, and distinguishing between horizontal and
+vertical forms. Later work differentiated intergenerational from
+intragenerational mobility, as well as absolute mobility---driven by
+structural change---from relative mobility, which captures the extent to
+which origins constrain destinations (Eyles et al., 2022). In Latin
+America, research shows high absolute but low relative mobility (Bucca,
+2016): although educational expansion and economic modernization have
+enabled some upward movement, status reproduction remains strong,
+especially among elites (Torche, 2015; López-Roldán \& Fachelli, 2021).
+This reflects deep structural inequality, segmented education systems,
+stratified labor markets, and legacies of dependent development that
+restrict access to mobility channels and reinforce the intergenerational
+transmission of advantage.
+
+Beyond mapping mobility patterns, growing research has examined its
+subjective and attitudinal effects. Mobility effects---defined as
+outcomes resulting from movement between origin and destination classes
+(Breen \& Ermisch, 2024)---have long attracted theoretical interest.
+Sorokin's (1959) dissociative hypothesis posits that mobility, whether
+upward or downward, may produce psychological strain due to conflicting
+norms between class contexts, leading to lower life satisfaction.
+Similarly, Lenski (1954) argued that status inconsistency---mismatches
+among education, income, and occupation---can undermine well-being.
+These ideas underpin extensive empirical work linking intergenerational
+mobility to outcomes such as life satisfaction, mental health, and
+stress (Hadjar, Gugushvili, Präg \& Gugushvili).
+
+Research on intergenerational social mobility has increasingly examined
+its effects on attitudes toward economic inequality. A key area of
+inquiry focuses on how upward and downward mobility influence support
+for redistribution, though findings remain mixed. Alesina et al.~(2018)
+show that individuals with pessimistic expectations about their
+mobility---particularly those anticipating downward movement---are more
+likely to support generous redistributive policies. Similarly, Ares
+(2020) finds that upwardly mobile individuals tend to be less supportive
+of state-led redistribution compared to those who have experienced
+downward mobility. Comparative studies by Schmidt (2011) and Gugushvili
+(2016) likewise report that subjective upward mobility is associated
+with weaker preferences for redistribution, while downward mobility
+strengthens redistributive support. However, recent causal evidence from
+Breen and Ermisch (2024) suggests the opposite: upward mobility may
+increase redistributive preferences, while downward mobility may reduce
+them.
+
+Beyond redistribution, other studies have explored the impact of
+mobility on broader beliefs about inequality. Gugushvili (2016) finds
+that upwardly mobile individuals are more likely to adopt
+individualistic attributions of poverty and to legitimize income
+inequality, particularly in post-socialist societies. Similarly, Bucca
+(2016) shows that subjective upward mobility reinforces individualistic
+explanations of wealth in seven Latin American countries. At the macro
+level, Shariff et al.~(2016) demonstrate that higher levels of national
+economic mobility correlate with greater tolerance of inequality. In
+contrast, Day and Fiske (2019) find that low perceived mobility
+undermines belief in meritocracy and a just world, thereby weakening
+system justification. Taken together, this body of research suggests
+that mobility shapes attitudes toward inequality and justice through
+multiple, and sometimes contradictory, mechanisms.
+
+The literature on the effects of social mobility has proposed various
+mechanisms to explain how and why changes in social position may
+influence individual outcomes (Freyr \& Rhem, año). One of the most
+prominent is the self-interest mechanism, which posits that individuals
+who experience upward or downward mobility undergo a shift in their
+material interests, thereby altering their perceptions and preferences
+(Ares, 2020; Langsæther, Evans \& O'Grady, 2021; Freyr \& Rhem, año).
+Closely related to this logic is the Prospect of Upward Mobility (POUM)
+hypothesis, which suggests that individuals may oppose redistribution
+not because of their current position, but because they anticipate
+improving their status in the future (REF). A second line of explanation
+draws on the framework of theories of socialization. Within this
+tradition, hypotheses such as acculturation, socialization, and status
+maximization propose that individuals adjust their attitudes based on
+the norms and values of either their class of origin or their
+destination, or a combination of both (Jaime-Castillo \&
+Marqués-Perales, 2019). However, most of these mechanisms focus
+primarily on the indirect effects of origin and destination positions,
+rather than on the direct effect of experience of movement itself. In
+response to this gap, recent research has highlighted the role of
+self-serving bias in causal attribution processes, suggesting that
+individuals tend to explain their mobility trajectories in ways that
+justify their current position, which in turn shapes their beliefs and
+preferences (Schmidt, 2011; Gugushvili, 2016; Molina, Bucca \& Macy,
+2019).
+
+The self-serving bias mechanism builds on a value-oriented perspective,
+emphasizing that individuals' experiences of social mobility shape their
+causal attributions, which in turn influence their beliefs about justice
+and distributive preferences (Gugushvili, 2014). Causal attribution
+refers to the process through which individuals generate explanations
+for their own behavior and outcomes, as well as those of others
+(Gugushvili, year). In this view, people's interpretations of economic
+inequality depend on whether they believe such disparities reflect
+unequal individual contributions. Individuals who adopt an internal
+attribution framework tend to see success or failure as rooted in
+personal characteristics such as effort, talent, or merit. In contrast,
+those who regard inequality as unjust are more likely to adopt an
+external attribution model, viewing outcomes as the result of structural
+barriers beyond individual control (Kluegel \& Smith, 1999). This
+mechanism, often described as intrapersonal causal attribution, reflects
+how people explain their own socioeconomic positions---typically
+attributing their successes to internal qualities while blaming failures
+on external circumstances (Miller \& Ross, 1975). Over time, individuals
+may revise their beliefs and attitudes: while early views are shaped by
+their social origin, these are later adjusted in light of personal
+experiences of mobility and the perceived role of ascribed versus
+achieved factors in determining socioeconomic outcomes (Gugushvili,
+year).
+
+Empirical research has provided support for the self-serving bias
+mechanism in shaping redistributive preferences and attitudes toward the
+legitimacy of inequality. For instance, Schmidt (2011) finds that
+individuals who experience upward mobility are more likely to interpret
+their success as the result of personal effort or merit, and
+consequently perceive redistribution as less necessary. Conversely,
+individuals who experience downward mobility tend to attribute their
+decline to external circumstances---such as structural inequality or
+unemployment---and show stronger support for redistribution. In a
+comparative analysis across different welfare domains, Gugushvili
+demonstrates that upward mobility is associated with lower support for
+government spending on housing and pensions, while individuals who
+experience downward mobility express lower support for healthcare and
+education spending, but favor increased investment in housing and
+pensions, reflecting the material nature of these domains. Moreover,
+Gugushvili (2016) finds that upward mobility is linked to greater
+justification of income inequality, suggesting that improved social
+standing reinforces an attributional view in which success is seen as
+the result of individual characteristics---thus legitimizing inequality
+as a fair outcome.
+
+These findings suggest that causal attribution processes mediate the
+relationship between mobility experiences and justice-related attitudes.
+Accordingly, this study proposes the following hypothesis:
+
+\(H1\): Experiencing upward (downward) social mobility is positively
+(negatively) associated with greater support for market justice in
+healthcare, pensions, and education.
+
 \subsection{Meritocracy}\label{meritocracy}
+
+Meritocracy constitutes a central ideological framework for legitimizing
+different types of social inequality, for instance through market
+justice beliefs. Rooted in the belief that rewards and positions should
+be allocated based on individual effort and talent, meritocracy operates
+as a normative ideal and a descriptive belief about how society
+functions. As initially conceptualized by Michael Young
+(\citeproc{ref-young_rise_1962}{\textbf{young\_rise\_1962?}}), the term
+was meant to critique a system in which merit-based stratification
+becomes a new form of inequality. However, over time, meritocracy has
+been widely supported in many societies as a fair and desirable
+principle of distribution, particularly within liberal democracies and
+market-oriented economies (\citeproc{ref-sandel_tyranny_2020}{Sandel,
+2020}; \citeproc{ref-mijs_paradox_2021}{\textbf{mijs\_paradox\_2021?}}).
+From a sociological perspective, the belief in meritocracy is more than
+a cognitive assessment; it reflects a moral lens through which
+individuals interpret inequality. People who believe that success
+results from hard work and talent are more likely to view social and
+economic disparities as legitimate
+(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
+\citeproc{ref-castillo_percepcion_2019}{\textbf{castillo\_percepcion\_2019?}}).
+Conversely, if they see outcomes as driven by luck, social origin, or
+systemic barriers, inequality is more likely to be perceived as unjust.
+This distinction becomes crucial in societies with persistent structural
+inequality, where public narratives often emphasize personal
+responsibility and merit while overlooking entrenched disadvantages.
+
+I adopt a multidimensional perspective on meritocracy, distinguishing
+between two key dimensions: effort-based and talent-based perceptions.
+This distinction is essential, as it captures different pathways through
+which individuals justify inequality. Effort-based meritocracy
+emphasizes hard work and perseverance as the basis for social rewards,
+aligning closely with cultural narratives of personal responsibility. A
+talent-based meritocracy, by contrast, emphasizes intelligence and
+innate abilities, which are often perceived as less malleable and more
+unequally distributed. Both dimensions have been shown to correlate with
+acceptance of inequality, but they may carry distinct implications for
+how inequality is justified in specific domains
+(\citeproc{ref-castillo_multidimensional_2023}{\textbf{castillo\_multidimensional\_2023?}}).
+The relevance of this distinction is supported by recent studies, which
+show that individuals respond differently to these dimensions. For
+instance, perceptions that effort is rewarded in society are more
+strongly associated with positive evaluations of fairness and acceptance
+of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
+2023}). This may be because effort is seen as a controllable and morally
+virtuous trait, whereas talent is often perceived as a natural
+advantage. Consequently, effort-based meritocracy is likely more potent
+in legitimizing inequality, particularly in neoliberal contexts.
+
+These dimensions of meritocracy reflect how respondents perceive
+society's distributive logic, regardless of whether they endorse
+meritocratic principles. This distinction aligns with recent findings
+indicating that individuals distinguish between how merit is perceived
+in society and how it should ideally operate, which in turn shapes their
+preferences for redistribution and justice
+(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
+2025}). Meritocratic beliefs serve as symbolic justifications for
+unequal outcomes, particularly when access is stratified by income or
+social opportunity. Prior studies have shown that individuals who
+perceive higher levels of meritocracy tend to express stronger support
+for unequal distributions that reflect market outcomes
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
+\citeproc{ref-castillo_percepcion_2019}{\textbf{castillo\_percepcion\_2019?}}).
+
+In addition to influencing individual attitudes toward inequality,
+meritocratic beliefs can contribute to social division and the
+stigmatization of disadvantaged groups. Recent research has demonstrated
+that exposure to meritocratic narratives can reinforce the belief that
+poverty results from individual failings rather than systemic
+conditions, reducing support for redistributive measures and increasing
+the stigmatization of the poor
+(\citeproc{ref-hoyt_mindsets_2023}{\textbf{hoyt\_mindsets\_2023?}}).
+This reinforces negative stereotypes and reduces empathy toward
+individuals from lower socioeconomic backgrounds. Moreover,
+(\citeproc{ref-busemeyer_positive_2021}{\textbf{busemeyer\_positive\_2021?}})
+argues that meritocratic narratives can serve as feedback mechanisms
+that shape public opinion and well-being by framing individuals'
+understanding of welfare outcomes as deserved or undeserved within
+existing institutional structures. This psychological mechanism
+highlights the normative power of meritocracy in stabilizing unequal
+systems by shaping political attitudes and personal perceptions of
+success and failure.
+
+Importantly, recent research has explored how meritocratic beliefs
+interact with experiences of intergenerational mobility to shape
+distributive attitudes. The belief that one's success is earned can lead
+upwardly mobile individuals to internalize meritocratic narratives and
+justify existing inequalities, reinforcing support for market justice
+(Gugushvili, 2016; Molina et al., 2019). Conversely, downwardly mobile
+individuals who maintain strong meritocratic beliefs may interpret their
+status as a personal failure, reducing their support for redistribution
+(Day \& Fiske, 2019). At the macro level, Shariff et al.~(2016) show
+that higher perceived mobility increases tolerance for inequality,
+suggesting that meritocracy and mobility are mutually reinforcing.
+
+Taken together, this literature supports the idea that meritocratic
+beliefs can moderate the relationship between mobility and market
+justice preferences. Individuals who experience mobility---especially
+upward---may draw on meritocratic narratives to legitimize both their
+own status and broader inequalities, thereby strengthening their support
+for market-based distribution. Accordingly, the second hypothesis of
+this study is:
+
+\(H2\): The positive (negative) relationship between upward (downward)
+social mobility and support for market justice in healthcare, pensions,
+and education is moderated by meritocratic beliefs; specifically, this
+association is stronger (weaker) among individuals with higher
+perceptions of meritocracy.
 
@@
   \item
-    Upper (codes 100--299: managers, directors, professionals)
+    Upper (codes 100-299: managers, directors, professionals)
   \item
-    Middle (codes 300--499: skilled white‑collar workers)
+    Middle (codes 300-499: skilled white-collar workers)
   \item
-    Lower (codes 500 and above: manual, semi‑skilled, or unskilled
+    Lower (codes 500 and above: manual, semi-skilled, or unskilled
     workers)
@@
 
+\bibitem[\citeproctext]{ref-araujo_desafios_2012}
+Araujo, K., \& Martuccelli, D. (2012). \emph{Desaf{í}os comunes: Retrato
+de la sociedad chilena y sus individuos} (1a. ed). Santiago: LOM
+Ediciones.
+
 \bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
@@
 durante la dictadura}}. Libros de la Catarata.
-
-\bibitem[\citeproctext]{ref-arteagaaguirre_politicas_2015}
-Arteaga Aguirre, C., \& Iñigo Valderrama, I. (2015). {Pol{í}ticas
-sociales, modelo de desarrollo y subjetividad de grupos vulnerables en
-Chile}.
 
@@
 Welfare States}}. Cambridge University Press.
+
+\bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
+Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
+(2021). Sujeto y cotidiano en la era neoliberal: El caso de la
+educaci{ó}n chilena. \emph{Revista Mexicana de Ciencias Pol{í}ticas y
+Sociales}, \emph{67}(244).
+\url{https://doi.org/10.22201/fcpys.2448492xe.2022.244.70386}
 
@@
 
+\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
+Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
+common good?} (First edition). New York: {Farrar, Straus and Giroux}.
+
 \bibitem[\citeproctext]{ref-schmidt_experience_2011}
@@
 
+\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
+Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
+Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
+{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
+\emph{International Review of Social Psychology}, \emph{38}(1), 4.
+\url{https://doi.org/10.5334/irsp.1013}
+
 \bibitem[\citeproctext]{ref-torche_intergenerational_2014}

### 2025-05-28 advances/informe2.qmd
Up informe 2

--- 2025-05-21 advances/informe2.qmd
@@
 Existing literature shows that market justice preferences are shaped by both the economic and institutional context and individuals' positions within social stratification.
-Grounded in the notion that institutions influence individual attitudes [@jacksonActorsInstitutions2010], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015].
+Grounded in the notion that economic institutions influence people's normative attitudes [@immergut_theoretical_1998], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015].
 In such contexts, market justice preferences tend to rise as individuals "ascend" the social structure, with those in more privileged positions in terms of class, education, and income holding stronger preferences for market-based solutions compared to those in more disadvantaged or at-risk positions [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_socialization_2024; @otero_power_2024; @lindh_public_2015; @svallfors_political_2007].
@@
 
+Although redistribution typically refers to the state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023].
+As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks.
+These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking [@koos_moral_2019; @immergut_theoretical_1998].
+In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007].
+While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@immergut_theoretical_1998].
+In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023].
+Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020].
+This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@lindh_public_2015; @mau_inequality_2015].
+
+The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles.
+Beyond stratification structures and socioeconomic conditions, the study of inequality also examines the beliefs individuals hold about the causes of social inequality, the normative principles that underpin those beliefs, the factors that shape them, and their consequences for attitudes and behavior [@kluegel_beliefs_1981].
+In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016].
+This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011].
+One prominent strand investigates the legitimacy of wage inequality, particularly salary differences across occupations [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978].
+Another examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, and pensions.
+Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015].
+In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic [@busemeyer_welfare_2020].
+
+Market justice preferences refer to normative beliefs that legitimize the idea that access to core social services—such as healthcare, education, or pensions—should be determined by market-based criteria.
+Following Janmaat’s [-@janmaat_subjective_2013, p.
+359] distinction, these preferences fall under the category of “beliefs,” understood as normative ideas about what inequality should look like, as opposed to “perceptions,” which refer to subjective evaluations of existing inequality.
+Market justice preferences reflect the view that access to these services should depend on individuals’ ability to pay, thus justifying inequalities generated by market mechanisms [@kluegel_legitimation_1999;  @lindh_public_2015].
+The concept draws on Lane’s [-@lane_market_1986] classic contrast between market justice and political justice: the former is grounded in the idea of earned deserts—where rewards reflect effort, productivity, and skill—while the latter prioritizes need and equality, typically expressed in welfare state policies.
+Lane argued that markets and states differ in purpose (efficiency vs.
+need), logic (individual vs.
+collective), and fairness criteria (merit vs.
+equality).
+Market justice assumes that markets are neutral, self-regulating systems in which fair procedures yield outcomes proportional to merit.
+Inequality, from this perspective, is not only expected but legitimate—as long as it arises from fair competition.
+In this way, market justice offers a moral lens through which individuals can view the commodification and stratified access to social services as fair and justified [@kluegel_legitimation_1999; @lindh_public_2015].
+
+The empirical study of market justice preferences has relied on diverse strategies to operationalize how individuals evaluate inequalities produced by market allocation.
+One common approach assesses whether people consider it fair that access to key social services—such as healthcare, education, or pensions—depends on income.
+This strategy builds on foundational work by Kluegel and Smith [-@kluegel_beliefs_1981], who explored the normative justifications behind support for economic inequality.
+More recent studies have expanded this logic beyond income to include market-mediated access to welfare goods.
+For instance, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
+[-@vondemknesebeck_are_2016] examine whether respondents believe it is just that those with higher incomes receive better healthcare.
+Similarly, Lee and Stacey [-@lee_fairness_2023] apply this approach to educational opportunities.
+These studies typically use targeted survey items that ask respondents to evaluate specific distributive scenarios, allowing for comparative analyses across countries or welfare regimes.
+Complementing these efforts, more recent research has introduced composite indicators designed to capture broader orientations toward market-based allocation.
+For example, Castillo et al.
+[-@castillo_socialization_2024] propose a single-item measure that summarizes individual support for income-based access across multiple domains—health, education, and pensions—within the Chilean context.
+Together, these instruments aim to capture to what extent individuals perceive market-generated inequality as justified.
+
+Comparative empirical research has identified several individual-level factors that influence support for market justice.
+Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles [@koos_moral_2019; @lindh_public_2015; @svallfors_political_2007].
+For example, Lindh [-@lindh_public_2015] finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries.
+In a comparative analysis, Svallfors [-@svallfors_political_2007] observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class.
+Busemeyer [-@busemeyer_skills_2014] similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
+[-@vondemknesebeck_are_2016] report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage.
+In Chile, Otero and Mendoza [-@otero_power_2024] show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions.
+Beliefs about inequality and political orientation also matter.
+In Chile, Castillo et al.
+[-@castillo_socialization_2024] show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism.
+In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.
+
+However, individual characteristics alone do not fully explain variation in support for market justice.
+A growing body of cross-national research highlights the importance of institutional arrangements in shaping these preferences.
+For instance, Immergut and Schneider [-@immergut_it_2020] find that in countries with higher public spending on healthcare, individuals are less likely to view income-based access as fair.
+Similarly, Busemeyer [-@busemeyer_skills_2014] shows that increased public investment in education is associated with lower support for privatized provision.
+Conversely, Lindh [-@lindh_public_2015] finds that in countries with more market-oriented welfare systems, support for market-based distribution tends to be higher, suggesting that individual attitudes often align with institutional outputs.
+These findings are consistent with neo-institutionalist and policy feedback theories [@campbell_institutional_2020], which argue that institutions do more than redistribute resources—they also shape the normative categories through which individuals assess who is deserving of support [@immergut_theoretical_1998].
+In this view, dominant values and preferences are both shaped by and embedded in institutional configurations [@busemeyer_skills_2014], reinforcing the notion that institutions are not neutral structures, but active producers of the moral frameworks that legitimize or challenge inequality.
+
 ## Social mobility
 
+The study of social mobility, its drivers and consequences has long been central to sociology.
+Classic theorists explored not only movement across social hierarchies but also its broader implications for class conflict, norm stability, and institutional change [@breen_effects_2024].
+Sorokin [-@sorokin_social_1927] introduced the concept formally, defining mobility as the shift of individuals, values, or objects between positions within a stratification system, and distinguishing between horizontal and vertical forms.
+Later work differentiated intergenerational from intragenerational mobility, as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which origins constrain destinations [@eyles_social_2022].
+In Latin America, research shows high absolute but low relative mobility [@bucca_merit_2016]: although educational expansion and economic modernization have enabled some upward movement, status reproduction remains strong, especially among elites [@torche_intergenerational_2014; @lopez-roldan_comparative_2021].
+This reflects deep structural inequality, segmented education systems, stratified labor markets, and legacies of dependent development that restrict access to mobility channels and reinforce the intergenerational transmission of advantage.
+
+Beyond mapping mobility patterns, growing research has examined its subjective and attitudinal effects.
+Mobility effects—defined as outcomes resulting from movement between origin and destination classes [@breen_effects_2024]—have long attracted theoretical interest.
+Sorokin’s (1959) dissociative hypothesis posits that mobility, whether upward or downward, may produce psychological strain due to conflicting norms between class contexts, leading to lower life satisfaction.
+Similarly, Lenski [-@lenski_status_1954] argued that status inconsistency—mismatches among education, income, and occupation—can undermine well-being.
+These ideas underpin extensive empirical work linking intergenerational mobility to outcomes such as life satisfaction, mental health, and stress [@prag_subjective_2021; @gugushvili_heterogeneous_2024; @hadjar_does_2015].
+
+Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality.
+A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed.
+Alesina et al.
+[-@alesina_intergenerational_2018] show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies.
+Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility.
+Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support.
+However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.
+
+Beyond redistribution, other studies have explored the impact of mobility on broader beliefs about inequality.
+Gugushvili [-@gugushvili_intergenerational_2016] finds that upwardly mobile individuals are more likely to adopt individualistic attributions of poverty and to legitimize income inequality, particularly in post-socialist societies.
+Similarly, Bucca [-@bucca_merit_2016] shows that subjective upward mobility reinforces individualistic explanations of wealth in seven Latin American countries.
+At the macro level, Shariff et al.
+[-@shariff_income_2016] demonstrate that higher levels of national economic mobility correlate with greater tolerance of inequality.
+In contrast, Day and Fiske [-@day_movin_2017] find that low perceived mobility undermines belief in meritocracy and a just world, thereby weakening system justification.
+Taken together, this body of research suggests that mobility shapes attitudes toward inequality and justice through multiple, and sometimes contradictory, mechanisms.
+
+The literature on the effects of social mobility has proposed various mechanisms to explain how and why changes in social position may influence individual outcomes [@helgason_class_2025].
+One of the most prominent is the self-interest mechanism, which posits that individuals who experience upward or downward mobility undergo a shift in their material interests, thereby altering their perceptions and preferences [@helgason_longterm_2023; @ares_changing_2020; @langsaether_explaining_2022].
+Closely related to this logic is the Prospect of Upward Mobility (POUM) hypothesis, which suggests that individuals may oppose redistribution not because of their current position, but because they anticipate improving their status in the future [@benabou_social_2001].
+A second line of explanation draws on the framework of theories of socialization.
+Within this tradition, hypotheses such as acculturation, socialization, and status maximization propose that individuals adjust their attitudes based on the norms and values of either their class of origin or their destination, or a combination of both [@jaime-castillo_social_2019].
+However, most of these mechanisms focus primarily on the indirect effects of origin and destination positions, rather than on the direct effect of experience of movement itself.
+In response to this gap, recent research has highlighted the role of self-serving bias in causal attribution processes, suggesting that individuals tend to explain their mobility trajectories in ways that justify their current position, which in turn shapes their beliefs and preferences [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
+
+The self-serving bias mechanism builds on a value-oriented perspective, emphasizing that individuals’ experiences of social mobility shape their causal attributions, which in turn influence their beliefs about justice and distributive preferences [@gugushvili_trends_2014].
+Causal attribution refers to the process through which individuals generate explanations for their own behavior and outcomes, as well as those of others [@gugushvili_intergenerational_2016].
+In this view, people’s interpretations of economic inequality depend on whether they believe such disparities reflect unequal individual contributions.
+Individuals who adopt an internal attribution framework tend to see success or failure as rooted in personal characteristics such as effort, talent, or merit.
+In contrast, those who regard inequality as unjust are more likely to adopt an external attribution model, viewing outcomes as the result of structural barriers beyond individual control [@kluegel_beliefs_1981].
+This mechanism, often described as intrapersonal causal attribution, reflects how people explain their own socioeconomic positions—typically attributing their successes to internal qualities while blaming failures on external circumstances [@miller_selfserving_1975].
+Over time, individuals may revise their beliefs and attitudes: while early views are shaped by their social origin, these are later adjusted in light of personal experiences of mobility and the perceived role of ascribed versus achieved factors in determining socioeconomic outcomes [@gugushvili_intergenerational_2016].
+
+Empirical research has provided support for the self-serving bias mechanism in shaping redistributive preferences and attitudes toward the legitimacy of inequality.
+For instance, Schmidt [-@schmidt_experience_2011] finds that individuals who experience upward mobility are more likely to interpret their success as the result of personal effort or merit, and consequently perceive redistribution as less necessary.
+Conversely, individuals who experience downward mobility tend to attribute their decline to external circumstances—such as structural inequality or unemployment—and show stronger support for redistribution.
+In a comparative analysis across different welfare domains, Gugushvili [-@gugushvili_subjective_2017] demonstrates that upward mobility is associated with lower support for government spending on housing and pensions, while individuals who experience downward mobility express lower support for healthcare and education spending, but favor increased investment in housing and pensions, reflecting the material nature of these domains.
+Moreover, Gugushvili [-@gugushvili_intergenerational_2016c] finds that upward mobility is linked to greater justification of income inequality, suggesting that improved social standing reinforces an attributional view in which success is seen as the result of individual characteristics—thus legitimizing inequality as a fair outcome.
+
+These findings suggest that causal attribution processes mediate the relationship between mobility experiences and justice-related attitudes. Accordingly, this study proposes the following hypothesis:
+
+$H1$: Experiencing upward (downward) social mobility is positively (negatively) associated with greater support for market justice in healthcare, pensions, and education.
+
 ## Meritocracy
+
+Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs.
+Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
+As initially conceptualized by Michael Young [-@young_rise_1958], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality.
+However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2019; @sandel_tyranny_2020].
+From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality.
+People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_meritocracia_2019; @batruch_belief_2023].
+Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust.
+This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.
+
+I adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions.
+This distinction is essential, as it captures different pathways through which individuals justify inequality [@young_rise_1958].
+Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility.
+A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed.
+Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023].
+The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions.
+For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023; @wiederkehr_belief_2015; @wilson_role_2003].
+This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage.
+Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.
+
+These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
+This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025].
+Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
+Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_meritocracia_2019; @castillo_socialization_2024].
+
+In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups.
+Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
+This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
+Moreover, Busemeyer et al.
+[-@busemeyer_positive_2021] argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
+This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
+
+Importantly, recent research has explored how meritocratic beliefs interact with experiences of intergenerational mobility to shape distributive attitudes.
+The belief that one’s success is earned can lead upwardly mobile individuals to internalize meritocratic narratives and justify existing inequalities, reinforcing support for market justice [@gugushvili_intergenerational_2016c; @molina_its_2019].
+Conversely, downwardly mobile individuals who maintain strong meritocratic beliefs may interpret their status as a personal failure, reducing their support for redistribution [@day_movin_2017].
+At the macro level, Shariff et al.
+[-@shariff_income_2016] show that higher perceived mobility increases tolerance for inequality, suggesting that meritocracy and mobility are mutually reinforcing.
+
+Taken together, this literature supports the idea that meritocratic beliefs can moderate the relationship between mobility and market justice preferences.
+Individuals who experience mobility—especially upward—may draw on meritocratic narratives to legitimize both their own status and broader inequalities, thereby strengthening their support for market-based distribution.
+Accordingly, the second hypothesis of this study is:
+
+$H2$: The positive (negative) relationship between upward (downward) social mobility and support for market justice in healthcare, pensions, and education is moderated by meritocratic beliefs; specifically, this association is stronger (weaker) among individuals with higher perceptions of meritocracy.
 

### 2025-05-28 advances/informe2.tex
Up informe 2

@@
 institutions influence people's normative attitudes
-(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
-find that countries with stronger public provision or more expansive
-welfare states exhibit lower levels of market justice preferences
+(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}), studies find
+that countries with stronger public provision or more expansive welfare
+states exhibit lower levels of market justice preferences
 (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
@@
 institutions also play a central role in distributing socially valuable
-goods and rewards (Koos \& Sachweh, 2019; Lindh \& McCall, 2023). As
-Polanyi (1957) observed, economic integration in capitalist societies is
-primarily organized through market exchange, governed by a
-self-regulating price system embedded in institutional frameworks. These
-institutions are not mere aggregates of individual behavior, but social
-realities endowed with rules, mechanisms, and normative meanings that
-shape everyday thinking (Koos \& Sachweh, 2019; Immergut, 1998). In this
-sense, the economic order is mirrored in a moral economy: collectively
-shared norms and beliefs about justice in distribution, embedded and
-reinforced through institutions (Mau, 2015; Svallfors, 2006). While most
+goods and rewards (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
+2019}; \citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As
+Polanyi (\citeproc{ref-polanyi_great_1975}{1975}) observed, economic
+integration in capitalist societies is primarily organized through
+market exchange, governed by a self-regulating price system embedded in
+institutional frameworks. These institutions are not mere aggregates of
+individual behavior, but social realities endowed with rules,
+mechanisms, and normative meanings that shape everyday thinking
+(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
+\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
+the economic order is mirrored in a moral economy: collectively shared
+norms and beliefs about justice in distribution, embedded and reinforced
+through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
 research from this perspective has focused on welfare institutions,
@@
 distributive justice beliefs and institutional responses to inequality
-(Lindh \& McCall, 2023). In many countries, privatization and
-commodification have expanded market logic into core areas of social
-reproduction, such as healthcare, education, pensions, and social
-security, deepening inequality in access to these services (Gingrich,
-2011; Ferre, 2023). Yet public support for market-based welfare
-provision has grown, even in traditional welfare states, particularly
-among higher-income groups who view private alternatives as more
-efficient or higher in quality (Busemeyer \& Iversen, 2020). This shift
-calls for broader inquiry into how institutions like markets structure
-access to resources and legitimize inequalities (Lindh \& McCall, 2023;
-Mau, 2015).
+(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In many
+countries, privatization and commodification have expanded market logic
+into core areas of social reproduction, such as healthcare, education,
+pensions, and social security, deepening inequality in access to these
+services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
+\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
+for market-based welfare provision has grown, even in traditional
+welfare states, particularly among higher-income groups who view private
+alternatives as more efficient or higher in quality
+(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
+This shift calls for broader inquiry into how institutions like markets
+structure access to resources and legitimize inequalities
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
 
@@
 beliefs, the factors that shape them, and their consequences for
-attitudes and behavior (Kluegel \& Smith, 1981). In this regard,
-empirical research on distributive justice focuses on individuals'
-conceptions of how goods and rewards \emph{should} be distributed in
-society (Jasso et al., año). This line of inquiry allows for examining
-the extent to which economic inequality is perceived as just, or in a
-certain sense, legitimate (Castillo, 2011). One prominent strand
-investigates the legitimacy of wage inequality, particularly salary
-differences across occupations (Jasso, 1978; Castillo, 2011; Wegener,
-1999). Another examines how individuals justify market-generated
-inequalities in access to core social services such as healthcare,
-education, and pensions. Here, legitimacy stems from the belief that
-access to these goods should follow market-based criteria (Castillo et
-al., 2024; Lindh, 2015). In such views, these services are treated as
-legitimate commodities; goods that can be traded, priced, and evaluated
-through market logic (Busemeyer \& Iversen, 2020).
+attitudes and behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \&
+Smith, 1981}). In this regard, empirical research on distributive
+justice focuses on individuals' conceptions of how goods and rewards
+\emph{should} be distributed in society
+(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
+2016}). This line of inquiry allows for examining the extent to which
+economic inequality is perceived as just, or in a certain sense,
+legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
+One prominent strand investigates the legitimacy of wage inequality,
+particularly salary differences across occupations
+(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
+\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
+\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Another
+examines how individuals justify market-generated inequalities in access
+to core social services such as healthcare, education, and pensions.
+Here, legitimacy stems from the belief that access to these goods should
+follow market-based criteria
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
+services are treated as legitimate commodities; goods that can be
+traded, priced, and evaluated through market logic
+(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
 
@@
 education, or pensions---should be determined by market-based criteria.
-Following Janmaat's (2013, p.~359) distinction, these preferences fall
-under the category of ``beliefs,'' understood as normative ideas about
-what inequality should look like, as opposed to ``perceptions,'' which
-refer to subjective evaluations of existing inequality. Market justice
-preferences reflect the view that access to these services should depend
-on individuals' ability to pay, thus justifying inequalities generated
-by market mechanisms (Kluegel \& Smith, 1999; Lindh, 2015). The concept
-draws on Lane's (1968) classic contrast between market justice and
-political justice: the former is grounded in the idea of earned
-deserts---where rewards reflect effort, productivity, and skill---while
-the latter prioritizes need and equality, typically expressed in welfare
-state policies. Lane argued that markets and states differ in purpose
-(efficiency vs.~need), logic (individual vs.~collective), and fairness
-criteria (merit vs.~equality). Market justice assumes that markets are
-neutral, self-regulating systems in which fair procedures yield outcomes
-proportional to merit. Inequality, from this perspective, is not only
-expected but legitimate---as long as it arises from fair competition. In
-this way, market justice offers a moral lens through which individuals
-can view the commodification and stratified access to social services as
-fair and justified (Kluegel \& Smith, 1999; Lindh, 2015).
+Following Janmaat's (\citeproc{ref-janmaat_subjective_2013}{2013, p.
+359}) distinction, these preferences fall under the category of
+``beliefs,'' understood as normative ideas about what inequality should
+look like, as opposed to ``perceptions,'' which refer to subjective
+evaluations of existing inequality. Market justice preferences reflect
+the view that access to these services should depend on individuals'
+ability to pay, thus justifying inequalities generated by market
+mechanisms (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al.,
+1999}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). The concept draws
+on Lane's (\citeproc{ref-lane_market_1986}{1986}) classic contrast
+between market justice and political justice: the former is grounded in
+the idea of earned deserts---where rewards reflect effort, productivity,
+and skill---while the latter prioritizes need and equality, typically
+expressed in welfare state policies. Lane argued that markets and states
+differ in purpose (efficiency vs.~need), logic (individual
+vs.~collective), and fairness criteria (merit vs.~equality). Market
+justice assumes that markets are neutral, self-regulating systems in
+which fair procedures yield outcomes proportional to merit. Inequality,
+from this perspective, is not only expected but legitimate---as long as
+it arises from fair competition. In this way, market justice offers a
+moral lens through which individuals can view the commodification and
+stratified access to social services as fair and justified
+(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
 
@@
 healthcare, education, or pensions---depends on income. This strategy
-builds on foundational work by Kluegel and Smith (1981), who explored
-the normative justifications behind support for economic inequality.
-More recent studies have expanded this logic beyond income to include
+builds on foundational work by Kluegel and Smith
+(\citeproc{ref-kluegel_beliefs_1981}{1981}), who explored the normative
+justifications behind support for economic inequality. More recent
+studies have expanded this logic beyond income to include
 market-mediated access to welfare goods. For instance, Immergut and
-Schneider (2020) and von dem Knesebeck et al.~(2016) examine whether
+Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
+et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) examine whether
 respondents believe it is just that those with higher incomes receive
-better healthcare. Similarly, Lee and Stacey (2023) apply this approach
-to educational opportunities. These studies typically use targeted
-survey items that ask respondents to evaluate specific distributive
-scenarios, allowing for comparative analyses across countries or welfare
-regimes. Complementing these efforts, more recent research has
-introduced composite indicators designed to capture broader orientations
-toward market-based allocation. For example, Castillo et al.~(2024)
-propose a single-item measure that summarizes individual support for
-income-based access across multiple domains---health, education, and
-pensions---within the Chilean context. Together, these instruments aim
-to capture to what extent individuals perceive market-generated
-inequality as justified.
+better healthcare. Similarly, Lee and Stacey
+(\citeproc{ref-lee_fairness_2023}{2023}) apply this approach to
+educational opportunities. These studies typically use targeted survey
+items that ask respondents to evaluate specific distributive scenarios,
+allowing for comparative analyses across countries or welfare regimes.
+Complementing these efforts, more recent research has introduced
+composite indicators designed to capture broader orientations toward
+market-based allocation. For example, Castillo et al.
+(\citeproc{ref-castillo_socialization_2024}{2024}) propose a single-item
+measure that summarizes individual support for income-based access
+across multiple domains---health, education, and pensions---within the
+Chilean context. Together, these instruments aim to capture to what
+extent individuals perceive market-generated inequality as justified.
 
@@
 education, and occupational status---are consistently more likely to
-endorse market-based distributive principles (Svallfors, 2007; Koos \&
-Sachweh, 2019; Lindh, 2015). For example, Lindh (2015) finds that
-individuals from the service class are more likely to support
-market-based access to healthcare and education than skilled and
-unskilled workers across 17 relatively affluent countries. In a
-comparative analysis, Svallfors (2007) observes that this expected class
-pattern appears clearly only in Sweden, where support for private
-education and healthcare varies systematically by class. Busemeyer
-(2014) similarly shows that support for private education is stronger
-among high-income groups, while Immergut and Schneider (2020) and von
-dem Knesebeck et al.~(2016) report comparable findings for healthcare,
-suggesting that wealthier individuals perceive private provision as a
-means of maintaining relative advantage. In Chile, Otero and Mendoza
-(2024) show that individuals with higher income and university education
-express stronger support for market allocation in healthcare, education,
-and pensions. Beliefs about inequality and political orientation also
-matter. In Chile, Castillo et al.~(2024) show that individuals who have
-strong meritocratic beliefs are more likely to support market-based
-distribution, while Lee and Stacy (2023) in Australia suggest that these
+endorse market-based distributive principles
+(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
+Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
+from the service class are more likely to support market-based access to
+healthcare and education than skilled and unskilled workers across 17
+relatively affluent countries. In a comparative analysis, Svallfors
+(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
+expected class pattern appears clearly only in Sweden, where support for
+private education and healthcare varies systematically by class.
+Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
+that support for private education is stronger among high-income groups,
+while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
+von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
+report comparable findings for healthcare, suggesting that wealthier
+individuals perceive private provision as a means of maintaining
+relative advantage. In Chile, Otero and Mendoza
+(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
+higher income and university education express stronger support for
+market allocation in healthcare, education, and pensions. Beliefs about
+inequality and political orientation also matter. In Chile, Castillo et
+al. (\citeproc{ref-castillo_socialization_2024}{2024}) show that
+individuals who have strong meritocratic beliefs are more likely to
+support market-based distribution, while Lee and Stacy
+(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
 preferences are also greater as people lean toward economic
@@
 highlights the importance of institutional arrangements in shaping these
-preferences. For instance, Immergut and Schneider (2020) find that in
-countries with higher public spending on healthcare, individuals are
-less likely to view income-based access as fair. Similarly, Busemeyer
-(2014) shows that increased public investment in education is associated
-with lower support for privatized provision. Conversely, Lindh (2015)
+preferences. For instance, Immergut and Schneider
+(\citeproc{ref-immergut_it_2020}{2020}) find that in countries with
+higher public spending on healthcare, individuals are less likely to
+view income-based access as fair. Similarly, Busemeyer
+(\citeproc{ref-busemeyer_skills_2014}{2014}) shows that increased public
+investment in education is associated with lower support for privatized
+provision. Conversely, Lindh (\citeproc{ref-lindh_public_2015}{2015})
 finds that in countries with more market-oriented welfare systems,
@@
 findings are consistent with neo-institutionalist and policy feedback
-theories (Campbell, 2012), which argue that institutions do more than
-redistribute resources---they also shape the normative categories
-through which individuals assess who is deserving of support (Immergut,
-1990). In this view, dominant values and preferences are both shaped by
-and embedded in institutional configurations (Busemeyer, 2014),
-reinforcing the notion that institutions are not neutral structures, but
-active producers of the moral frameworks that legitimize or challenge
+theories (\citeproc{ref-campbell_institutional_2020}{Campbell, 2020}),
+which argue that institutions do more than redistribute resources---they
+also shape the normative categories through which individuals assess who
+is deserving of support
+(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In this
+view, dominant values and preferences are both shaped by and embedded in
+institutional configurations
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014}), reinforcing the
+notion that institutions are not neutral structures, but active
+producers of the moral frameworks that legitimize or challenge
 inequality.
@@
 across social hierarchies but also its broader implications for class
-conflict, norm stability, and institutional change (Breen \& Ermisch,
-2024). Sorokin (1927) introduced the concept formally, defining mobility
-as the shift of individuals, values, or objects between positions within
-a stratification system, and distinguishing between horizontal and
-vertical forms. Later work differentiated intergenerational from
-intragenerational mobility, as well as absolute mobility---driven by
-structural change---from relative mobility, which captures the extent to
-which origins constrain destinations (Eyles et al., 2022). In Latin
-America, research shows high absolute but low relative mobility (Bucca,
-2016): although educational expansion and economic modernization have
-enabled some upward movement, status reproduction remains strong,
-especially among elites (Torche, 2015; López-Roldán \& Fachelli, 2021).
-This reflects deep structural inequality, segmented education systems,
+conflict, norm stability, and institutional change
+(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}). Sorokin
+(\citeproc{ref-sorokin_social_1927}{1927}) introduced the concept
+formally, defining mobility as the shift of individuals, values, or
+objects between positions within a stratification system, and
+distinguishing between horizontal and vertical forms. Later work
+differentiated intergenerational from intragenerational mobility, as
+well as absolute mobility---driven by structural change---from relative
+mobility, which captures the extent to which origins constrain
+destinations (\citeproc{ref-eyles_social_2022}{Eyles et al., 2022}). In
+Latin America, research shows high absolute but low relative mobility
+(\citeproc{ref-bucca_merit_2016}{Bucca, 2016}): although educational
+expansion and economic modernization have enabled some upward movement,
+status reproduction remains strong, especially among elites
+(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
+2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). This
+reflects deep structural inequality, segmented education systems,
 stratified labor markets, and legacies of dependent development that
@@
 outcomes resulting from movement between origin and destination classes
-(Breen \& Ermisch, 2024)---have long attracted theoretical interest.
-Sorokin's (1959) dissociative hypothesis posits that mobility, whether
-upward or downward, may produce psychological strain due to conflicting
-norms between class contexts, leading to lower life satisfaction.
-Similarly, Lenski (1954) argued that status inconsistency---mismatches
-among education, income, and occupation---can undermine well-being.
-These ideas underpin extensive empirical work linking intergenerational
-mobility to outcomes such as life satisfaction, mental health, and
-stress (Hadjar, Gugushvili, Präg \& Gugushvili).
+(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024})---have long
+attracted theoretical interest. Sorokin's (1959) dissociative hypothesis
+posits that mobility, whether upward or downward, may produce
+psychological strain due to conflicting norms between class contexts,
+leading to lower life satisfaction. Similarly, Lenski
+(\citeproc{ref-lenski_status_1954}{1954}) argued that status
+inconsistency---mismatches among education, income, and occupation---can
+undermine well-being. These ideas underpin extensive empirical work
+linking intergenerational mobility to outcomes such as life
+satisfaction, mental health, and stress
+(\citeproc{ref-gugushvili_heterogeneous_2024}{Gugushvili, 2024};
+\citeproc{ref-hadjar_does_2015}{Hadjar \& Samuel, 2015};
+\citeproc{ref-prag_subjective_2021}{Präg \& Gugushvili, 2021}).
 
@@
 inquiry focuses on how upward and downward mobility influence support
-for redistribution, though findings remain mixed. Alesina et al.~(2018)
-show that individuals with pessimistic expectations about their
+for redistribution, though findings remain mixed. Alesina et al.
+(\citeproc{ref-alesina_intergenerational_2018}{2018}) show that
+individuals with pessimistic expectations about their
 mobility---particularly those anticipating downward movement---are more
 likely to support generous redistributive policies. Similarly, Ares
-(2020) finds that upwardly mobile individuals tend to be less supportive
-of state-led redistribution compared to those who have experienced
-downward mobility. Comparative studies by Schmidt (2011) and Gugushvili
-(2016) likewise report that subjective upward mobility is associated
-with weaker preferences for redistribution, while downward mobility
+(\citeproc{ref-ares_changing_2020}{2020}) finds that upwardly mobile
+individuals tend to be less supportive of state-led redistribution
+compared to those who have experienced downward mobility. Comparative
+studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
+Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017})
+likewise report that subjective upward mobility is associated with
+weaker preferences for redistribution, while downward mobility
 strengthens redistributive support. However, recent causal evidence from
-Breen and Ermisch (2024) suggests the opposite: upward mobility may
-increase redistributive preferences, while downward mobility may reduce
-them.
+Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the
+opposite: upward mobility may increase redistributive preferences, while
+downward mobility may reduce them.
 
 Beyond redistribution, other studies have explored the impact of
-mobility on broader beliefs about inequality. Gugushvili (2016) finds
-that upwardly mobile individuals are more likely to adopt
-individualistic attributions of poverty and to legitimize income
-inequality, particularly in post-socialist societies. Similarly, Bucca
-(2016) shows that subjective upward mobility reinforces individualistic
-explanations of wealth in seven Latin American countries. At the macro
-level, Shariff et al.~(2016) demonstrate that higher levels of national
-economic mobility correlate with greater tolerance of inequality. In
-contrast, Day and Fiske (2019) find that low perceived mobility
+mobility on broader beliefs about inequality. Gugushvili
+(\citeproc{ref-gugushvili_intergenerational_2016}{2016b}) finds that
+upwardly mobile individuals are more likely to adopt individualistic
+attributions of poverty and to legitimize income inequality,
+particularly in post-socialist societies. Similarly, Bucca
+(\citeproc{ref-bucca_merit_2016}{2016}) shows that subjective upward
+mobility reinforces individualistic explanations of wealth in seven
+Latin American countries. At the macro level, Shariff et al.
+(\citeproc{ref-shariff_income_2016}{2016}) demonstrate that higher
+levels of national economic mobility correlate with greater tolerance of
+inequality. In contrast, Day and Fiske
+(\citeproc{ref-day_movin_2017}{2017}) find that low perceived mobility
 undermines belief in meritocracy and a just world, thereby weakening
@@
 mechanisms to explain how and why changes in social position may
-influence individual outcomes (Freyr \& Rhem, año). One of the most
-prominent is the self-interest mechanism, which posits that individuals
-who experience upward or downward mobility undergo a shift in their
-material interests, thereby altering their perceptions and preferences
-(Ares, 2020; Langsæther, Evans \& O'Grady, 2021; Freyr \& Rhem, año).
+influence individual outcomes
+(\citeproc{ref-helgason_class_2025}{Helgason \& Rehm, 2025}). One of the
+most prominent is the self-interest mechanism, which posits that
+individuals who experience upward or downward mobility undergo a shift
+in their material interests, thereby altering their perceptions and
+preferences (\citeproc{ref-ares_changing_2020}{Ares, 2020};
+\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
+\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
 Closely related to this logic is the Prospect of Upward Mobility (POUM)
@@
 not because of their current position, but because they anticipate
-improving their status in the future (REF). A second line of explanation
-draws on the framework of theories of socialization. Within this
-tradition, hypotheses such as acculturation, socialization, and status
-maximization propose that individuals adjust their attitudes based on
-the norms and values of either their class of origin or their
-destination, or a combination of both (Jaime-Castillo \&
-Marqués-Perales, 2019). However, most of these mechanisms focus
+improving their status in the future
+(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}). A second line
+of explanation draws on the framework of theories of socialization.
+Within this tradition, hypotheses such as acculturation, socialization,
+and status maximization propose that individuals adjust their attitudes
+based on the norms and values of either their class of origin or their
+destination, or a combination of both
+(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}). However, most of these mechanisms focus
 primarily on the indirect effects of origin and destination positions,
@@
 justify their current position, which in turn shapes their beliefs and
-preferences (Schmidt, 2011; Gugushvili, 2016; Molina, Bucca \& Macy,
-2019).
+preferences
+(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
+\citeproc{ref-molina_its_2019}{Molina et al., 2019};
+\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}).
 
@@
 causal attributions, which in turn influence their beliefs about justice
-and distributive preferences (Gugushvili, 2014). Causal attribution
-refers to the process through which individuals generate explanations
-for their own behavior and outcomes, as well as those of others
-(Gugushvili, year). In this view, people's interpretations of economic
-inequality depend on whether they believe such disparities reflect
-unequal individual contributions. Individuals who adopt an internal
-attribution framework tend to see success or failure as rooted in
-personal characteristics such as effort, talent, or merit. In contrast,
-those who regard inequality as unjust are more likely to adopt an
-external attribution model, viewing outcomes as the result of structural
-barriers beyond individual control (Kluegel \& Smith, 1999). This
-mechanism, often described as intrapersonal causal attribution, reflects
-how people explain their own socioeconomic positions---typically
-attributing their successes to internal qualities while blaming failures
-on external circumstances (Miller \& Ross, 1975). Over time, individuals
-may revise their beliefs and attitudes: while early views are shaped by
-their social origin, these are later adjusted in light of personal
-experiences of mobility and the perceived role of ascribed versus
-achieved factors in determining socioeconomic outcomes (Gugushvili,
-year).
+and distributive preferences
+(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Causal
+attribution refers to the process through which individuals generate
+explanations for their own behavior and outcomes, as well as those of
+others (\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
+2016b}). In this view, people's interpretations of economic inequality
+depend on whether they believe such disparities reflect unequal
+individual contributions. Individuals who adopt an internal attribution
+framework tend to see success or failure as rooted in personal
+characteristics such as effort, talent, or merit. In contrast, those who
+regard inequality as unjust are more likely to adopt an external
+attribution model, viewing outcomes as the result of structural barriers
+beyond individual control (\citeproc{ref-kluegel_beliefs_1981}{Kluegel
+\& Smith, 1981}). This mechanism, often described as intrapersonal
+causal attribution, reflects how people explain their own socioeconomic
+positions---typically attributing their successes to internal qualities
+while blaming failures on external circumstances
+(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Over
+time, individuals may revise their beliefs and attitudes: while early
+views are shaped by their social origin, these are later adjusted in
+light of personal experiences of mobility and the perceived role of
+ascribed versus achieved factors in determining socioeconomic outcomes
+(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}).
 
@@
 mechanism in shaping redistributive preferences and attitudes toward the
-legitimacy of inequality. For instance, Schmidt (2011) finds that
-individuals who experience upward mobility are more likely to interpret
-their success as the result of personal effort or merit, and
-consequently perceive redistribution as less necessary. Conversely,
-individuals who experience downward mobility tend to attribute their
-decline to external circumstances---such as structural inequality or
-unemployment---and show stronger support for redistribution. In a
-comparative analysis across different welfare domains, Gugushvili
-demonstrates that upward mobility is associated with lower support for
-government spending on housing and pensions, while individuals who
-experience downward mobility express lower support for healthcare and
-education spending, but favor increased investment in housing and
-pensions, reflecting the material nature of these domains. Moreover,
-Gugushvili (2016) finds that upward mobility is linked to greater
-justification of income inequality, suggesting that improved social
-standing reinforces an attributional view in which success is seen as
-the result of individual characteristics---thus legitimizing inequality
-as a fair outcome.
+legitimacy of inequality. For instance, Schmidt
+(\citeproc{ref-schmidt_experience_2011}{2011}) finds that individuals
+who experience upward mobility are more likely to interpret their
+success as the result of personal effort or merit, and consequently
+perceive redistribution as less necessary. Conversely, individuals who
+experience downward mobility tend to attribute their decline to external
+circumstances---such as structural inequality or unemployment---and show
+stronger support for redistribution. In a comparative analysis across
+different welfare domains, Gugushvili
+(\citeproc{ref-gugushvili_subjective_2017}{2017}) demonstrates that
+upward mobility is associated with lower support for government spending
+on housing and pensions, while individuals who experience downward
+mobility express lower support for healthcare and education spending,
+but favor increased investment in housing and pensions, reflecting the
+material nature of these domains. Moreover, Gugushvili
+(\citeproc{ref-gugushvili_intergenerational_2016c}{2016a}) finds that
+upward mobility is linked to greater justification of income inequality,
+suggesting that improved social standing reinforces an attributional
+view in which success is seen as the result of individual
+characteristics---thus legitimizing inequality as a fair outcome.
 
@@
 functions. As initially conceptualized by Michael Young
-(\citeproc{ref-young_rise_1962}{\textbf{young\_rise\_1962?}}), the term
-was meant to critique a system in which merit-based stratification
-becomes a new form of inequality. However, over time, meritocracy has
-been widely supported in many societies as a fair and desirable
-principle of distribution, particularly within liberal democracies and
-market-oriented economies (\citeproc{ref-sandel_tyranny_2020}{Sandel,
-2020}; \citeproc{ref-mijs_paradox_2021}{\textbf{mijs\_paradox\_2021?}}).
-From a sociological perspective, the belief in meritocracy is more than
-a cognitive assessment; it reflects a moral lens through which
-individuals interpret inequality. People who believe that success
-results from hard work and talent are more likely to view social and
-economic disparities as legitimate
-(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_percepcion_2019}{\textbf{castillo\_percepcion\_2019?}}).
+(\citeproc{ref-young_rise_1958}{1958}), the term was meant to critique a
+system in which merit-based stratification becomes a new form of
+inequality. However, over time, meritocracy has been widely supported in
+many societies as a fair and desirable principle of distribution,
+particularly within liberal democracies and market-oriented economies
+(\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
+\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
+perspective, the belief in meritocracy is more than a cognitive
+assessment; it reflects a moral lens through which individuals interpret
+inequality. People who believe that success results from hard work and
+talent are more likely to view social and economic disparities as
+legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
+\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
 Conversely, if they see outcomes as driven by luck, social origin, or
@@
 This distinction is essential, as it captures different pathways through
-which individuals justify inequality. Effort-based meritocracy
+which individuals justify inequality
+(\citeproc{ref-young_rise_1958}{Young, 1958}). Effort-based meritocracy
 emphasizes hard work and perseverance as the basis for social rewards,
@@
 how inequality is justified in specific domains
-(\citeproc{ref-castillo_multidimensional_2023}{\textbf{castillo\_multidimensional\_2023?}}).
+(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
 The relevance of this distinction is supported by recent studies, which
@@
 of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}). This may be because effort is seen as a controllable and morally
-virtuous trait, whereas talent is often perceived as a natural
-advantage. Consequently, effort-based meritocracy is likely more potent
-in legitimizing inequality, particularly in neoliberal contexts.
+2023}; \citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al., 2015};
+\citeproc{ref-wilson_role_2003}{Wilson, 2003}). This may be because
+effort is seen as a controllable and morally virtuous trait, whereas
+talent is often perceived as a natural advantage. Consequently,
+effort-based meritocracy is likely more potent in legitimizing
+inequality, particularly in neoliberal contexts.
 
@@
 (\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-castillo_percepcion_2019}{\textbf{castillo\_percepcion\_2019?}}).
+\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
 
@@
 conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor
-(\citeproc{ref-hoyt_mindsets_2023}{\textbf{hoyt\_mindsets\_2023?}}).
-This reinforces negative stereotypes and reduces empathy toward
-individuals from lower socioeconomic backgrounds. Moreover,
-(\citeproc{ref-busemeyer_positive_2021}{\textbf{busemeyer\_positive\_2021?}})
-argues that meritocratic narratives can serve as feedback mechanisms
-that shape public opinion and well-being by framing individuals'
-understanding of welfare outcomes as deserved or undeserved within
-existing institutional structures. This psychological mechanism
-highlights the normative power of meritocracy in stabilizing unequal
-systems by shaping political attitudes and personal perceptions of
-success and failure.
+the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
+et al., 2023}). This reinforces negative stereotypes and reduces empathy
+toward individuals from lower socioeconomic backgrounds. Moreover,
+Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
+that meritocratic narratives can serve as feedback mechanisms that shape
+public opinion and well-being by framing individuals' understanding of
+welfare outcomes as deserved or undeserved within existing institutional
+structures. This psychological mechanism highlights the normative power
+of meritocracy in stabilizing unequal systems by shaping political
+attitudes and personal perceptions of success and failure.
 
@@
 justify existing inequalities, reinforcing support for market justice
-(Gugushvili, 2016; Molina et al., 2019). Conversely, downwardly mobile
-individuals who maintain strong meritocratic beliefs may interpret their
-status as a personal failure, reducing their support for redistribution
-(Day \& Fiske, 2019). At the macro level, Shariff et al.~(2016) show
-that higher perceived mobility increases tolerance for inequality,
-suggesting that meritocracy and mobility are mutually reinforcing.
+(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
+\citeproc{ref-molina_its_2019}{Molina et al., 2019}). Conversely,
+downwardly mobile individuals who maintain strong meritocratic beliefs
+may interpret their status as a personal failure, reducing their support
+for redistribution (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017}).
+At the macro level, Shariff et al.
+(\citeproc{ref-shariff_income_2016}{2016}) show that higher perceived
+mobility increases tolerance for inequality, suggesting that meritocracy
+and mobility are mutually reinforcing.
 
@@
 
+\bibitem[\citeproctext]{ref-ares_changing_2020}
+Ares, M. (2020). Changing classes, changing preferences: How social
+class mobility affects economic preferences. \emph{West European
+Politics}, \emph{43}(6), 1211--1237.
+\url{https://doi.org/10.1080/01402382.2019.1644575}
+
 \bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
@@
 
+\bibitem[\citeproctext]{ref-benabou_social_2001}
+Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
+{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
+Economics}, \emph{116}(2), 447--487.
+\url{https://doi.org/10.1162/00335530151144078}
+
 \bibitem[\citeproctext]{ref-boccardo_30_2020}
@@
 
+\bibitem[\citeproctext]{ref-bucca_merit_2016}
+Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
+Latin Americans}' beliefs about wealth and poverty. \emph{Research in
+Social Stratification and Mobility}, \emph{44}, 98--112.
+\url{https://doi.org/10.1016/j.rssm.2016.02.005}
+
 \bibitem[\citeproctext]{ref-busemeyer_skills_2014}
-Busemeyer, M. R. (2014). \emph{Skills and {Inequality}: {Partisan
-Politics} and the {Political Economy} of {Education Reforms} in {Western
-Welfare States}}. Cambridge University Press.
+Busemeyer, M. (2014). \emph{Skills and {Inequality}: {Partisan Politics}
+and the {Political Economy} of {Education Reforms} in {Western Welfare
+States}}. Cambridge University Press.
+
+\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
+Busemeyer, M., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive} and
+{Negative}: {New Perspectives} on {Feedback Effects} in {Public Opinion}
+on the {Welfare State}. \emph{British Journal of Political Science},
+\emph{51}(1), 137--162. \url{https://doi.org/10.1017/S0007123418000534}
+
+\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
+Busemeyer, M., \& Iversen, T. (2020). The {Welfare State} with {Private
+Alternatives}: {The Transformation} of {Popular Support} for {Social
+Insurance}. \emph{The Journal of Politics}, \emph{82}(2), 671--686.
+\url{https://doi.org/10.1086/706980}
+
+\bibitem[\citeproctext]{ref-campbell_institutional_2020}
+Campbell, J. L. (2020). \emph{Institutional {Change} and
+{Globalization}}. Princeton University Press.
+\url{https://doi.org/10.2307/j.ctv131bw68}
 
@@
 
+\bibitem[\citeproctext]{ref-castillo_legitimacy_2011}
+Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
+Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
+Research}, \emph{24}(4), 314--340.
+\url{https://doi.org/10.1007/s11211-011-0144-5}
+
+\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
+Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
+(2023). A {Multidimensional Approach} for {Measuring Meritocratic
+Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
+Social Inequality Survey}. \emph{International Journal of Sociology},
+1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
+
 \bibitem[\citeproctext]{ref-castillo_socialization_2024}
@@
 
+\bibitem[\citeproctext]{ref-castillo_meritocracia_2019}
+Castillo, J. C., Torres, A., Atria, J., \& Maldonado, L. (2019).
+Meritocracia y desigualdad econ{ó}mica: {Percepciones}, preferencias e
+implicancias. \emph{Revista Internacional de Sociolog{í}a},
+\emph{77}(1), 117. \url{https://doi.org/10.3989/ris.2019.77.1.17.114}
+
 \bibitem[\citeproctext]{ref-chancel_world_2022}
@@
 movilidad social bajo un}.
+
+\bibitem[\citeproctext]{ref-eyles_social_2022}
+Eyles, A., Major, L. E., \& Machin, S. (2022). \emph{Social {Mobility} -
+{Past}, {Present} and {Future}: {The State} of {Play} in {Social
+Mobility}, on the 25th {Anniversary} of the {Sutton Trust}}. Sutton
+Trust.
 
@@
 
+\bibitem[\citeproctext]{ref-gugushvili_heterogeneous_2024}
+Gugushvili, A. (2024). The heterogeneous well-being effects of
+intergenerational mobility perceptions. \emph{Journal of Health
+Psychology}, \emph{29}(2), 99--112.
+\url{https://doi.org/10.1177/13591053231187345}
+
+\bibitem[\citeproctext]{ref-hadjar_does_2015}
+Hadjar, A., \& Samuel, R. (2015). Does upward social mobility increase
+life satisfaction? {A} longitudinal analysis using {British} and {Swiss}
+panel data. \emph{Research in Social Stratification and Mobility},
+\emph{39}, 48--58. \url{https://doi.org/10.1016/j.rssm.2014.12.002}
+
 \bibitem[\citeproctext]{ref-harvey_breve_2015}
@@
 (Espa{ñ}a): Ediciones Akal.
+
+\bibitem[\citeproctext]{ref-helgason_longterm_2023}
+Helgason, A. F., \& Rehm, P. (2023). Long-term income trajectories and
+the evolution of political attitudes. \emph{European Journal of
+Political Research}, \emph{62}(1), 264--284.
+\url{https://doi.org/10.1111/1475-6765.12506}
+
+\bibitem[\citeproctext]{ref-helgason_class_2025}
+Helgason, A. F., \& Rehm, P. (2025). Class experiences and the long-term
+evolution of economic values. \emph{Social Forces}, \emph{103}(3),
+1125--1143. \url{https://doi.org/10.1093/sf/soae135}
+
+\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
+Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
+D. (2023). Mindsets of poverty: {Implications} for redistributive policy
+support. \emph{Analyses of Social Issues and Public Policy},
+\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
+
+\bibitem[\citeproctext]{ref-immergut_theoretical_1998}
+Immergut, E. M. (1998). The {Theoretical Core} of the {New
+Institutionalism}. \emph{Politics \& Society}, \emph{26}(1), 5--34.
+\url{https://doi.org/10.1177/0032329298026001002}
 
@@
 
-\bibitem[\citeproctext]{ref-jacksonActorsInstitutions2010}
-Jackson, G. (2010). Actors and institutions. In G. Morgan, J. L.
-Campbell, C. Crouch, O. K. Pedersen, \& R. Whitley (Eds.), \emph{The
-{Oxford Handbook} of {Comparative Institutional Analysis}} (1st ed.).
-Oxford University Press.
-\url{https://doi.org/10.1093/oxfordhb/9780199233762.001.0001}
-
 \bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
@@
 \url{https://doi.org/10.1111/1468-4446.12363}
+
+\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
+Janmaat, J. G. (2013). Subjective {Inequality}: A {Review} of
+{International Comparative Studies} on {People}'s {Views} about
+{Inequality}. \emph{European Journal of Sociology}, \emph{54}(3),
+357--389. \url{https://doi.org/10.1017/S0003975613000209}
+
+\bibitem[\citeproctext]{ref-jasso_justice_1978}
+Jasso, Guillermina. (1978). On the {Justice} of {Earnings}: {A New
+Specification} of the {Justice Evaluation Function}. \emph{American
+Journal of Sociology}, \emph{83}(6), 1398--1419.
+\url{https://doi.org/10.1086/226706}
+
+\bibitem[\citeproctext]{ref-jasso_distributive_2016}
+Jasso, Guillermina, Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
+{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
+Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
+\url{https://doi.org/10.1007/978-1-4939-3216-0_11}
+
+\bibitem[\citeproctext]{ref-jasso_gender_1999}
+Jasso, G., \& Wegener, B. (1999). Gender and {Country Differences} in
+the {Sense} of {Justice}: {Justice Evaluation}, {Gender Earnings Gap},
+and {Earnings Functions} in {Thirteen Countries}. \emph{International
+Journal of Comparative Sociology}, \emph{40}(1), 94--115.
+\url{https://doi.org/10.1177/002071529904000106}
+
+\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
+of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
+about {Market Justice}, 1991-1996. \emph{European Sociological Review},
+\emph{15}(3), 251--283. Retrieved from
+\url{https://www.jstor.org/stable/522731}
+
+\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
+Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
+\emph{Annual Review of Sociology}, 29--56.
 
@@
 
+\bibitem[\citeproctext]{ref-lane_market_1986}
+Lane, R. E. (1986). Market {Justice}, {Political Justice}.
+\emph{American Political Science Review}, \emph{80}(2), 383--402.
+\url{https://doi.org/10.2307/1958264}
+
+\bibitem[\citeproctext]{ref-langsaether_explaining_2022}
+Langsæther, P. E., Evans, G., \& O'Grady, T. (2022). Explaining the
+{Relationship Between Class Position} and {Political Preferences}: {A
+Long-Term Panel Analysis} of {Intra-Generational Class Mobility}.
+\emph{British Journal of Political Science}, \emph{52}(2), 958--967.
+\url{https://doi.org/10.1017/S0007123420000599}
+
 \bibitem[\citeproctext]{ref-lee_fairness_2023}
@@
 
+\bibitem[\citeproctext]{ref-lenski_status_1954}
+Lenski, G. E. (1954). Status {Crystallization}: {A Non-Vertical
+Dimension} of {Social Status}. \emph{American Sociological Review},
+\emph{19}(4), 405. \url{https://doi.org/10.2307/2087459}
+
 \bibitem[\citeproctext]{ref-lindh_public_2015}
@@
 
+\bibitem[\citeproctext]{ref-molina_its_2019}
+Molina, M. D., Bucca, M., \& Macy, M. W. (2019). It's not just how the
+game is played, it's whether you win or lose. \emph{SCIENCE ADVANCES}.
+
 \bibitem[\citeproctext]{ref-otero_power_2024}
@@
 NY: Octagon Books.
+
+\bibitem[\citeproctext]{ref-prag_subjective_2021}
+Präg, P., \& Gugushvili, A. (2021). Subjective social mobility and
+health in {Germany}. \emph{European Societies}, \emph{23}(4), 464--486.
+\url{https://doi.org/10.1080/14616696.2021.1887916}
 
@@
 \url{https://doi.org/10.1080/14742837.2020.1727737}
+
+\bibitem[\citeproctext]{ref-sorokin_social_1927}
+Sorokin, P. A. (1927). \emph{Social {Mobility}}. Harper \& Brothers.
 
@@
 
+\bibitem[\citeproctext]{ref-wiederkehr_belief_2015}
+Wiederkehr, V., Bonnot, V., Krauth-Gruber, S., \& Darnon, C. (2015).
+Belief in school meritocracy as a system-justifying tool for low status
+students. \emph{Frontiers in Psychology}, \emph{6}.
+
 \bibitem[\citeproctext]{ref-wilson_role_2003}

### 2025-06-27 advances/Devia, Laffert y Budnevich.qmd
Up informe 3

---
title: "Ejercicio 3"
format:
  pdf: 
    geometry: margin=2cm
    keep-tex: true
    number-sections: true
editor: source
author:
  - name: Ignacia Devia
    affiliations:
      - ref: 1  
  - name: Sebastían Budnevich
  - name: Andreas Laffert Tamayo
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
citeproc: true
link-citations: true
colorlinks: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
fig-height: 4
fig-width: 7.5
---

# Pregunta 1

Teniendo en cuenta los principales elementos del diseño de conjoint experiments, describa el diseño del experimento para el país que está analizando. Ponga particular atención en explicar el diseño del atributo "País de Origen".

Teniendo en cuenta los principales elementos del diseño de experimentos tipo conjoint, el estudio de Wimmer et al. (2024) implementa un diseño factorial completamente aleatorizado con el objetivo de identificar cómo la rivalidad geopolítica influye en las actitudes ciudadanas hacia inmigrantes. En este tipo de diseños, los participantes deben elegir entre pares de candidatos a la residencia permanente cuyas características varían en múltiples dimensiones. Este enfoque permite estimar de manera causal los efectos de distintos atributos simultáneamente, al comparar sistemáticamente perfiles con combinaciones contrastantes de atributos.

En términos metodológicos, los experimentos tipo conjoint permiten estimar las preferencias de los encuestados a partir de sus evaluaciones globales de perfiles hipotéticos que difieren en múltiples atributos, los cuales suelen presentarse en formato tabular (Bansak et al., 2021). La fortaleza del diseño radica en que habilita la identificación de los criterios específicos que las personas utilizan al tomar decisiones complejas —en este caso, sobre la admisión de inmigrantes— así como la importancia relativa que asignan a cada uno de esos criterios. Por ejemplo, si un perfil incluye información sobre la ocupación, edad, idioma o país de origen del solicitante, el análisis factorial permite identificar cuál de estos componentes tiene mayor peso en la decisión de los encuestados y si dichas prioridades son compartidas de forma generalizada dentro de la población analizada. Esta capacidad para descomponer decisiones multidimensionales en sus elementos causales diferenciados constituye uno de los principales aportes metodológicos del diseño conjoint (Hainmueller, Hopkins y Yamamoto, 2014).

En el caso de Brasil, los participantes fueron invitados a desempeñar el rol de oficiales de inmigración, debiendo seleccionar entre pares de perfiles generados aleatoriamente. Cada perfil incluía atributos como edad, ocupación, dominio del idioma, años de residencia en el país, género y, centralmente para este análisis, país de origen (Wimmer et al., 2024). El experimento se administró de manera online mediante la plataforma Lucid Marketplace, utilizando un muestreo por cuotas basado en edad, género y educación.

El atributo “País de Origen” fue diseñado mediante una estructura 2x2: se seleccionaron dos países cultural o racialmente similares al país encuestado y dos cultural o racialmente distintos; dentro de cada par, uno representaba a un país rival y otro a un país aliado o no rival. Para Brasil, los países incluidos fueron China (rival, distinto), Rusia (rival, similar), Japón (no rival, distinto) y Ucrania (no rival, similar) (Wimmer et al., 2024, suplemento). Esta estructura permite analizar si las preferencias de los encuestados se ven más afectadas por la percepción de rivalidad geopolítica que por la cercanía cultural o racial.

No obstante, los autores advierten que, en algunos casos, se realizaron elecciones subóptimas respecto a los países de origen incluidos. Dado que el conjunto de países encuestados fue definido por las necesidades de un proyecto más amplio y no por criterios específicos del presente experimento, en ciertos contextos —como el de Brasil— fue necesario utilizar países cuya relación con el país encuestado podía ser neutral o ambigua. En palabras de los autores, “we had to make sub-optimal choices due to the fact that survey countries were not primarily selected with our experiment in mind” (Wimmer et al., 2024, p. 6). Asimismo, reconocen que algunas clasificaciones de países como rivales o aliados se realizaron con base en definiciones generales y no en un consenso unívoco, siguiendo criterios como los propuestos por Thompson (2001), que entiende la rivalidad como una relación entre Estados que se perciben mutuamente como competidores y potenciales amenazas militares.

A pesar de estas limitaciones, el diseño fue respaldado por una encuesta de validación posterior, que confirmó que los brasileños perciben a China y Rusia como más hostiles, y a Japón y Ucrania como más amistosos, y que también distinguen entre estos países en términos de distancia cultural y racial. Esta validación da soporte empírico a las decisiones de diseño adoptadas, aunque no elimina del todo la ambigüedad inherente a la categorización de las relaciones internacionales desde una perspectiva pública (Wimmer et al., 2024, Suplemento D).

En cuanto a la estimación de efectos, el estudio emplea principalmente dos estimandos: el Average Marginal Component Effect (AMCE) y los Marginal Means (MM). El AMCE estima el efecto promedio de un valor de atributo (por ejemplo, “origen en Rusia”) respecto a una categoría base, promediando sobre los demás atributos. Los MM, por su parte, indican la probabilidad promedio de que un perfil con cierto valor de atributo sea seleccionado. Los autores optan por presentar los resultados en términos de MM, dado que permiten comparaciones más directas entre países y subgrupos sin depender de una categoría de referencia arbitraria (Leeper, Hobolt y Tilley, 2020).

# Pregunta 2

En base a la clase y su lectura de Leeper et al. (2020), haga una comparación entre "average marginal component effects" (AMCEs) y "marginal means". Discuta por qué Wimmer et al. (2025) deciden usar "marginal means".

Leeper et al. (2020) sugieren que para caracterizar descriptivamente las diferencias en el nivel de preferencia entre subgrupos se priorice la estimación de medias marginales condicionales y sus diferencias, en vez de AMCE´s. En general, los AMCE proporcionan información sobre variación descriptiva de las preferencias dentro de grupos y entre características y los AMCE condicionales estiman el tamaño de los efectos causales de las características dentro de los grupos. En esta misma línea, de acuerdo con Leeper et al. (2020), el AMCE tiene una interpretación causal sobre los tamaños de efectos de los atributos , cuando existe aleatorización en las características de los perfiles, en el contexto de experimentos conjuntos. De este modo, permite comprender el efecto causal de una característica/atributo dado en el apoyo general a un objeto multidimensional, promediando sobre la distribución de todas las otras características del objeto incluidas en el diseño. De forma más sencilla, los AMCEs estiman el efecto de cambiar un nivel de atributo comparado con un nivel de referencia.

Sin embargo, los autores advierten que su uso también se ha empleado con fines descriptivos, es decir, para examinar las preferencias de subgrupos hacia perfiles con determinadas características. Si bien puede reflejar bien la preferencias con diferencias en los AMCE de subgrupos, esto supone que la categoría de referencia experimental es similar (cumple requisito de igualdad), puesto que de lo contrario,la diferencia de AMCE´s puede generar una representación “engañosa” de los patrones subyacentes de preferencias. En concreto, una categoría de referencia arbitraria podría afectar el tamaño, la dirección y la significancia estadística de las diferencias en los AMCE,cuya consecuencia es que preferencias similares parezcan distintas y viceversa.  Por lo tanto, los autores sugieren que no se utilice AMCE´s condicional ni sus diferencias para informar acerca de la similitud o diferencia subyacentes en la preferencias entre subgrupos de una muestra, pues refiere a cantidades relativas expresada contra una línea de base. Incluso señalan que en la práctica investigativa se intenta mostrar diferencias en la favorabilidad de dos o más subgrupos hacia una característica conjunta mediante diferencias de AMCE, cuando en realidad no refleja aquello.

Mientras tanto la media marginal (MM) describe el nivel de “favorabilidad” hacia los perfiles para todos los niveles de un atributo/característica en particular, ignorando todas las demás. Es decir, la media marginal es un valor promedio para cada nivel, siendo un valor absoluto de preferencia. Es por esta razón que Wimmer et al. (2025) prefieres las medias marginales (MM) a los efectos de componentes marginales promedio (AMCE), ya que las medias marginales  no dependen de la elección de una categoría de referencia, además de permitir comparar directamente preferencias entre subgrupos de encuestados así como la magnitud entre los países de origen de los inmigrantes y otros atributos. Entre las ventajas de las medias marginales es su fácil interpretación en tanto se hace como probabilidades.Igualmente, la diferencia entre medias marginales de dos grupos es siempre fija en los datos, mientras que en el AMCE está supeditado al centramiento en un punto en los datos (categoría de referencia).

# Pregunta 3

Para el país que le ha sido asignado:

a) Reproduzca los análisis de las Figuras 1 y 2 en Wimmer et al. (2025). Puede presentar sus resultados en una tabla de calidad o en figuras.

```{r, echo=FALSE, message=FALSE}
library(cregg)
library(ggplot2)
library(dplyr)
load("C:/Users/cjara/OneDrive - Universidad Católica de Chile/Magíster/Tercer Semestre/Experimentos/Ejercicio 3/brazil.RData")
```

```{r, message=FALSE, warning=FALSE}
mm <- cregg::cj(
  data_pais,
  selected ~ age + gender + residence + occ + lang + ctry.rival + ctry.race,
  id       = "ResponseId",
  estimate = "mm",
  family   = stats::binomial()
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8, dpi=300}
# Definir el orden exacto como aparece en la figura (de arriba hacia abajo)
level_order <- c(
  "Age: 21",
  "Age: 38", 
  "Age: 62",
  "Gender: Female",
  "Gender: Male",
  "Residence: 15 years",
  "Residence: 10 years",
  "Residence: 5 years", 
  "Language: Fluent",
  "Language: Limited",
  "Language: None",
  "Occupation: Doctor",
  "Occupation: Teacher", 
  "Occupation: Accountant",
  "Occupation: Farmer",
  "Occupation: Cleaner",
  "Occupation: Unemployed",
  "Country of origin: Different majority race/culture",
  "Country of origin: Similar majority race/culture",
  "Country of origin: Rival",
  "Country of origin: Non-rival"
)

mm_plot <- mm %>%
  mutate(level = factor(level, levels = rev(level_order)))

ggplot(mm_plot, aes(x = estimate, y = level)) +
  # Puntos
  geom_point(size = 2, color = "black") +

  # Intervalos de confianza
  geom_errorbarh(aes(xmin = lower, xmax = upper), 
                 height = 0.2, 
                 color = "black",
                 size = 0.5) +

  # Línea vertical en 0.5
  geom_vline(xintercept = 0.5, 
             linetype = "dashed", 
             color = "gray50",
             size = 0.5) +

  # Escalas y límites
  scale_x_continuous(
    name = "Marginal Mean",
    breaks = seq(0.4, 0.6, 0.1),
    labels = c("0.4", "0.5", "0.6")
  ) +

  # Eliminar etiqueta del eje y
  scale_y_discrete(name = "") +

  # Tema y apariencia
  theme_minimal() +
  theme(
    # Título
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    # Ejes
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 9, hjust = 1),
    axis.title.x = element_text(size = 10),

    # Panel
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90", size = 0.3),

    # Bordes
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),

    # Márgenes
    plot.margin = margin(20, 20, 40, 20),

    # Caption
    plot.caption = element_text(size = 8, hjust = 0, margin = margin(t = 10))
  ) +

  # Títulos
  labs(
    title = "FIGURE 1. Effects of Rivalry and Perceived Racial/Cultural Similarity on the Probability of Being\nGranted Permanent Resident Status"
  )

```

```{r, message=FALSE, warning=FALSE}
data_pais$interaction <- interaction(data_pais$ctry.rival, data_pais$ctry.race, sep = "\n")
mm2 <- cregg::cj(
  data_pais,
  selected ~ age + gender + residence + occ + lang + interaction,
  id = "ResponseId",
  estimate = "mm",
  family = stats::binomial()
)
```

```{r, echo=FALSE, message=FALSE, fig.width=10, fig.height=8, dpi=300}
# Filtrar solo los resultados de interacción
interaction_results <- mm2[mm2$feature == "interaction", ]

# Crear etiquetas más claras para el gráfico
interaction_results$clean_labels <- c(
  "Country of origin:\nSimilar race/culture + Non-rival",
    "Country of origin:\nSimilar race/culture + Rival",
    "Country of origin:\nDifferent race/culture + Non-rival",
    "Country of origin:\nDifferent race/culture + Rival"
)

# Reordenar para que coincida con el orden del gráfico original
interaction_results$clean_labels <- factor(interaction_results$clean_labels, 
  levels = c(
    "Country of origin:\nSimilar race/culture + Non-rival",
    "Country of origin:\nSimilar race/culture + Rival",
    "Country of origin:\nDifferent race/culture + Non-rival",
    "Country of origin:\nDifferent race/culture + Rival"
  ))

ggplot(interaction_results, aes(x = estimate, y = clean_labels)) +
  geom_point(size = 2, color = "black") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), 
                 height = 0.2, 
                 color = "black",
                 size = 0.5) +
  geom_vline(xintercept = 0.5, 
             linetype = "dashed", 
             color = "gray50",
             size = 0.5) +
  scale_x_continuous(
    name = "Marginal Mean",
    breaks = seq(0.45, 0.55, 0.05),
    labels = c("0.45", "0.5", "0.55")
  ) +
  scale_y_discrete(name = "") +
  theme_minimal() +
  theme(
    # Título
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    # Ejes
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 9, hjust = 1),
    axis.title.x = element_text(size = 10),

    # Panel
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90", size = 0.3),

    # Bordes
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),

    # Márgenes
    plot.margin = margin(20, 20, 40, 20),

    # Caption
    plot.caption = element_text(size = 8, hjust = 0, margin = margin(t = 10))
  ) +

  # Títulos
  labs(
    title = "FIGURE 2. Interaction Effects of Rivalry Status and Perceived Racial/Cultural Similarity on the\nProbability of Being Granted Permanent Residence Status"
  )
```

b) Respecto de la hipótesis 1 de Wimmer et al. (2025), interprete sus resultados y señale si hay evidencia consistente con esta hipótesis.

La Hipótesis 1 de Wimmer et al. (2025) plantea que, al evaluar la candidatura de un inmigrante para residencia permanente, los encuestados serán menos propensos a elegir perfiles cuyo país de origen sea considerado rival y más propensos a elegir aquellos provenientes de países aliados (non-rivals). En la Figura 1 se observa que las MM para los países rivales (China y Rusia) rondan en promedio 0,46, mientras que las MM para los non-rivals (Japón y Ucrania) se acercan a 0,53. Además, los IC 95 % de China y Rusia (niveles rivales) están claramente por debajo de los IC 95 % de Japón y Ucrania (niveles non-rival). Esto implica que, aun promediando sobre edad, género, residencia, ocupación e idioma, la probabilidad de elegir un perfil rival es significativamente menor. Por lo tanto, en el caso de Brasil se cumple la Hipótesis 1 formulada por los autores.

Para explorar cómo interviene la distancia racial/cultural `ctry.race`, la Figura 2 (que corresponde a la Hipótesis 2) muestra que, al comparar **rival + similar race** (Rusia) con **rival + different race** (China), Rusia presenta una MM aún más baja. Esto confirma que la penalización por ser rival se intensifica cuando el país es percibido como racialmente similar. En conclusión, los resultados para Brasil coinciden con los hallazgos de los autores en toda la muestra y respaldan la Hipótesis 1.

# Referencias

Bansak, K., Hainmueller, J., Hopkins, D. J., & Yamamoto, T. (2019). “Conjoint Survey Experiments” For Druckman, James N., and Donald P. Green, eds. Cambridge Handbook of Advances in Experimental Political Science, New York: Cambridge University Press.’.

Hainmueller, J., Hopkins, D. J., & Yamamoto, T. (2014). Causal inference in conjoint analysis: Understanding multidimensional choices via stated preference experiments. Political analysis, 22(1), 1-30.

Leeper, T. J., Hobolt, S. B., & Tilley, J. (2020). Measuring subgroup preferences in conjoint experiments. Political Analysis, 28(2), 207-221.

Wimmer, A., Bonikowski, B., Crabtree, C., Fu, Z., Golder, M., & Tsutsui, K. (2025). Geo-political rivalry and anti-immigrant sentiment: A conjoint experiment in 22 countries. American Political Science Review, 119(2), 1018-1035.

\pagebreak

# Código de R

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```

### 2025-06-27 advances/extended_abstract.qmd
Up informe 3

---
title: "Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
    code-link: true
    title-block-banner: true
    theme:
      - cosmo
      - pretty.scss
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
author:
  - name: Andreas Laffert Tamayo
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
citeproc: true
link-citations: true
colorlinks: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
fig-height: 4
fig-width: 7.5
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

```{r}
# This document was last modified at {{< modified >}}
# 
# and it was last rendered at {{< now >}}

```

This article examines how intergenerational social mobility and meritocratic beliefs influence public support for market-based allocation of social services— a phenomenon known as market justice preferences. While previous studies have linked individuals' class position and beliefs about merit to distributive attitudes, less is known about how social mobility—especially in unequal and privatized welfare contexts—shapes support for inequality in access to healthcare, education, and pensions. This study addresses this gap by analyzing whether upward or downward mobility affects market justice preferences, and how this relationship is moderated by meritocratic beliefs.

The analysis focuses on Chile, a paradigmatic case of neoliberal transformation. Despite sustained economic growth and poverty reduction, Chile remains one of the most unequal countries in Latin America and the OECD. Since the 1980s, its welfare system has undergone deep privatization and commodification, generating a segmented structure of service provision and moral discourses that justify inequality. In this context, meritocratic beliefs and mobility trajectories may interact to shape how individuals evaluate fairness in access to core welfare domains.

Using data from the 2023 wave of the Chilean Longitudinal Social Survey (ELSOC), the study analyzes responses from a nationally representative urban sample (n = 2,726). Market justice preferences are measured through agreement with income-based inequalities in access to healthcare, pensions, and education. Intergenerational mobility is operationalized by comparing respondents’ and their fathers’ occupational status using ISEI-based class strata (upper, middle, lower). To estimate the causal effect of mobility, inverse-probability weighted linear regressions are implemented following Breen and Ermisch’s framework, allowing for estimation of average treatment effects on the treated (ATT). Meritocratic beliefs are measured via perceived rewards for effort and talent.

Results show that upward mobility is associated with stronger market justice preferences, while downward mobility reduces support for market-based access. These effects are consistent across service domains and robust to controls for income, education, age, gender, political identification, and subjective social status. Moreover, the strength of meritocratic beliefs moderates this relationship: upwardly mobile individuals who believe effort and talent are rewarded in Chile express the highest support for income-based inequalities, consistent with a self-serving attribution mechanism. In contrast, downwardly mobile individuals who reject meritocratic beliefs are more critical of inequality in access.

The findings contribute to the literature by highlighting the dynamic and interactive nature of distributive preferences. Rather than being solely determined by fixed class position, attitudes toward market justice are shaped by experiences of mobility and the normative frameworks individuals use to interpret them. In contexts of high inequality and welfare commodification, such as Chile, upward mobility may legitimize neoliberal moral orders by reinforcing beliefs in individual responsibility and merit-based allocation. These insights underscore the importance of incorporating mobility trajectories and belief systems into the analysis of welfare attitudes and inequality legitimation.

### 2025-06-27 advances/extended_abstract.tex
Up informe 3

--- 2025-05-21 advances/informe2.tex
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \setstretch{1.15}
-\section{Introduction}\label{introduction}
+This article examines how intergenerational social mobility and
+meritocratic beliefs influence public support for market-based
+allocation of social services--- a phenomenon known as market justice
+preferences. While previous studies have linked individuals' class
+position and beliefs about merit to distributive attitudes, less is
+known about how social mobility---especially in unequal and privatized
+welfare contexts---shapes support for inequality in access to
+healthcare, education, and pensions. This study addresses this gap by
+analyzing whether upward or downward mobility affects market justice
+preferences, and how this relationship is moderated by meritocratic
+beliefs.
 
-What makes market-driven inequalities appear legitimate in the eyes of
-the public? Since the early 1980s, many countries have experienced a
-widespread retreat from universal welfare programs and a shift toward
-the privatization and commodification of public goods, welfare policies,
-and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
-2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
-America, as elsewhere, neoliberal reforms reshaped welfare-state
-institutions by extending market logic into domains of social
-reproduction that were traditionally governed by the state
-(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
-\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
-reduced the role of public provision and increased the presence of
-private actors in core social services
-(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
-(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
-constitute a distinct moral order, the institutional diffusion of market
-rules has fostered a corresponding moral economy: a constellation of
-norms and values concerning fair allocation, embedded in institutions
-and shaping individual subjectivities
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
-this framework, a growing body of research examines the extent to which,
-and the mechanisms by which, citizens consider it fair that access to
-services like health care, pensions, and education be governed by
-market-based criteria---a phenomenon known as \emph{market justice
-preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
-Understanding these preferences is crucial, as they contribute to
-legitimizing economic inequality by framing it as the fair result of
-individual responsibility and limited state intervention
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
+The analysis focuses on Chile, a paradigmatic case of neoliberal
+transformation. Despite sustained economic growth and poverty reduction,
+Chile remains one of the most unequal countries in Latin America and the
+OECD. Since the 1980s, its welfare system has undergone deep
+privatization and commodification, generating a segmented structure of
+service provision and moral discourses that justify inequality. In this
+context, meritocratic beliefs and mobility trajectories may interact to
+shape how individuals evaluate fairness in access to core welfare
+domains.
 
-Existing literature shows that market justice preferences are shaped by
-both the economic and institutional context and individuals' positions
-within social stratification. Grounded in the notion that institutions
-influence individual attitudes
-(\citeproc{ref-jacksonActorsInstitutions2010}{Jackson, 2010}), studies
-find that countries with stronger public provision or more expansive
-welfare states exhibit lower levels of market justice preferences
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
-more privatized contexts show stronger support for market-based criteria
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
-justice preferences tend to rise as individuals ``ascend'' the social
-structure, with those in more privileged positions in terms of class,
-education, and income holding stronger preferences for market-based
-solutions compared to those in more disadvantaged or at-risk positions
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
-\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
-2016}).
+Using data from the 2023 wave of the Chilean Longitudinal Social Survey
+(ELSOC), the study analyzes responses from a nationally representative
+urban sample (n = 2,726). Market justice preferences are measured
+through agreement with income-based inequalities in access to
+healthcare, pensions, and education. Intergenerational mobility is
+operationalized by comparing respondents' and their fathers'
+occupational status using ISEI-based class strata (upper, middle,
+lower). To estimate the causal effect of mobility, inverse-probability
+weighted linear regressions are implemented following Breen and
+Ermisch's framework, allowing for estimation of average treatment
+effects on the treated (ATT). Meritocratic beliefs are measured via
+perceived rewards for effort and talent.
 
-Market justice preferences are shaped not only by objective
-socioeconomic conditions but also by popular beliefs about inequality.
-Among these, meritocracy is a key normative principle underpinning
-market-based distributive preferences
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). It frames inequality as
-inevitable but justifiable through effort and talent
-(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
-\citeproc{ref-young_rise_1958}{Young, 1958}). Studies show that
-individuals with stronger meritocratic beliefs tend to perceive less
-inequality and legitimize it by attributing economic differences to
-personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
-societies where access to services is largely governed by market logic,
-such beliefs play a critical role in normalizing inequality. Recent
-evidence from Chile shows that students who believe effort and talent
-are rewarded in their country express stronger preferences for
-market-based access to healthcare, pensions, and education
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
+Results show that upward mobility is associated with stronger market
+justice preferences, while downward mobility reduces support for
+market-based access. These effects are consistent across service domains
+and robust to controls for income, education, age, gender, political
+identification, and subjective social status. Moreover, the strength of
+meritocratic beliefs moderates this relationship: upwardly mobile
+individuals who believe effort and talent are rewarded in Chile express
+the highest support for income-based inequalities, consistent with a
+self-serving attribution mechanism. In contrast, downwardly mobile
+individuals who reject meritocratic beliefs are more critical of
+inequality in access.
 
-Although it is clear that one's social position influences market
-preferences, the question of how upward or downward mobility within the
-social structure affects these preferences remains unanswered. This
-question is far from trivial, especially in Latin America, where many
-have experienced various forms of mobility amid high economic inequality
-and deep welfare privatization (\citeproc{ref-ferre_welfare_2023}{Ferre,
-2023}; \citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \&
-Fachelli, 2021}; \citeproc{ref-torche_intergenerational_2014}{Torche,
-2014}). Social origins and destinations affect attitudes toward
-inequality in distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske,
-2017}; \citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
-2016b}, \citeproc{ref-gugushvili_subjective_2017}{2017};
-\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
-2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
-movement between these positions exposes individuals to different
-experiences and mechanisms that shape their views on what is fair
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
-\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
-research, examining the effects of social mobility on market justice
-preferences can help to illuminate how inequalities in access to social
-services are justified among individuals who have experienced, or not,
-changes in their social standing, and what are the normative mechanisms
-that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
-2015}).
-
-Beyond their isolated effects, social mobility and meritocratic beliefs
-interact in complex ways to shape market justice preferences. A key
-mechanism proposed in the literature to explain how mobility influences
-distributive justice preferences is the psychological process of
-self-serving bias in causal attribution
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
-suggests that individuals attribute failures---such as downward
-mobility---to external factors, while crediting successes---such as
-upward mobility---to their own merit and effort
-(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
-who experience upward mobility tend to view their social position as
-earned, making them more likely to believe that individuals are
-responsible for their own success or failure. Research shows that upward
-mobility is associated with weaker preferences for redistribution
-(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
-\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
-2011}) and stronger legitimacy of income inequality
-(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
-individuals who experience downward mobility tend to blame structural
-factors like inequality and are more supportive of redistribution while
-rejecting merit-based explanations
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
-together, I argue that meritocratic beliefs may reflect---or
-reinforce---this self-serving attribution mechanism by legitimizing
-one's social status as the outcome of personal merit, closely tied to
-attribution bias.
-
-Against this background, this article pursues two main objectives:
-first, to analyze the extent to which intergenerational social mobility
-influences market justice preferences regarding healthcare, pensions,
-and education; and second, to examine how meritocratic beliefs may
-moderate this relationship. Building on a theoretical framework that
-emphasizes how neoliberal transformations---particularly through the
-privatization and commodification of key areas of social
-reproduction---have profoundly reshaped processes of subject formation
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}), the central argument is
-that upward mobility increases support for market justice preferences,
-while downward mobility decreases it. Moreover, meritocratic beliefs are
-expected to moderate this relationship by reflecting a self-serving
-attribution bias, whereby individuals justify their social position in
-terms of personal merit.
-
-This study focuses on Chile, a particularly intriguing case for
-examining market justice preferences. Despite significant economic
-growth and poverty reduction, Chile has some of the highest levels of
-inequality in Latin America and among OECD countries
-(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
-\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
-coexists with short-range upward mobility among lower-class segments
-moving into middle strata, though strong barriers remain to reaching
-higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
-et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
-2014}). What makes Chile especially salient is that much of this
-inequality is rooted in deep neoliberal reforms that institutionalized
-the privatization and commodification of key social sectors
-(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
-during the dictatorship (1973--1989) and expanded in democracy, these
-reforms enabled the unprecedented emergence of markets in health,
-pensions, and education, with provision segmented by individuals'
-ability to pay and supported by public subsidies
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
-despite waves of protest against inequality and commodification from
-2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
-2021})---Chilean subjectivities have been increasingly shaped by
-neoliberal discourses and market logics, influencing their attitudes
-toward inequality and welfare distribution
-(\citeproc{ref-arteagaaguirre_politicas_2015}{Arteaga Aguirre \& Iñigo
-Valderrama, 2015}; \citeproc{ref-castillo_socialization_2024}{Castillo
-et al., 2024}).
-
-In this context, the questions that guide this research are as follows:
-
-\begin{enumerate}
-\def\labelenumi{(\arabic{enumi})}
-\tightlist
-\item
-  To what extent does intergenerational social mobility influence market
-  justice preferences regarding healthcare, pensions, and education in
-  Chile?
-\item
-  How do meritocratic beliefs condition or moderate this relationship in
-  the Chilean context?
-\end{enumerate}
-
-To address these questions, this study draws on large-scale,
-representative survey data collected in 2018 from the urban Chilean
-population aged 18 to 75 (n = 2,983). The next section outlines the
-theoretical framework linking market justice preferences, social
-mobility, and meritocratic beliefs, and proposes a set of hypotheses.
-This is followed by a description of the data, variables, and analytical
-strategy. The final sections present the empirical findings, offer an
-interpretation of the results, and conclude with a discussion of their
-implications.
-
-\section{Theoretical and empirical
-background}\label{theoretical-and-empirical-background}
-
-\subsection{Market justice
-preferences}\label{market-justice-preferences}
-
-\subsection{Social mobility}\label{social-mobility}
-
-\subsection{Meritocracy}\label{meritocracy}
-
-\subsection{The Chilean context}\label{the-chilean-context}
-
-\section{Data, variables and methods}\label{data-variables-and-methods}
-
-\subsection{Data}\label{data}
-
-This study draws on data from the Chilean Longitudinal Social Survey
-(ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
-The survey is an annual panel collected from 2016 to 2023, comprising
-two independent samples (original and refreshment), featuring permanent
-and rotating questionnaire modules. It evaluates how individuals think,
-feel, and behave regarding conflict and social cohesion in Chile. The
-sampling design is complex, probabilistic, clustered, multistage, and
-stratified by city size, targeting men and women aged 18--75 who
-habitually reside in private dwellings in urban areas spanning 40 cities
-(92 municipalities, 13 regions). This analysis focuses on 2023, the most
-recent wave, which includes 1,737 respondents in the original sample and
-989 in the refreshment sample. Further details on sampling, attrition,
-and weighting can be found at https://coes.cl/encuesta-panel/, and the
-dataset is publicly available at
-https://dataverse.harvard.edu/dataverse/elsoc.
-
-\subsection{Variables}\label{variables}
-
-\emph{Outcome variables}
-
-\textbf{Market justice preferences}: The outcome variables in this study
-are market justice preferences. This construct is operationalized
-through three variables that address the degree of justification
-regarding whether access to social services in healthcare, pensions, and
-education should be income conditional. Specifically, the justification
-of inequality in healthcare is assessed by the question: ``Is it fair in
-Chile that people with higher incomes can access better healthcare than
-people with lower incomes?'' The same question is asked for pensions and
-education. In all cases, respondents indicate their level of agreement
-on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
-(``strongly agree''). Although these items theoretically reflect the
-same underlying concept of market justice
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}), they are analyzed
-separately because healthcare, pensions, and education may prompt
-distinct response patterns
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}).
-
-\emph{Independent variables}
-
-\textbf{Social mobility}: Intergenerational social mobility is treated
-as an exposure indicating whether respondents occupy a different class
-position from their fathers. Following Breen and Ermisch's
-(\citeproc{ref-breen_effects_2024}{2024}) framework for estimating
-causal effects of mobility, the variable is constructed in two stages.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Class assignment. Using the International Socio-Economic Index of
-  Occupational Status (ISEI) derived from three-digit ISCO-08 codes,
-  both paternal (origin) and respondent (destination) occupations are
-  grouped into three strata:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Upper (codes 100--299: managers, directors, professionals)
-  \item
-    Middle (codes 300--499: skilled white‑collar workers)
-  \item
-    Lower (codes 500 and above: manual, semi‑skilled, or unskilled
-    workers)
-  \end{itemize}
-\item
-  Propensity-score estimation. To model mobility as a treatment---i.e.,
-  a transition from origin \emph{j} to destination \emph{k}---I estimate
-  multinomial logit propensity scores using covariates that influence an
-  individual's likelihood of moving between strata: (a) father's
-  educational level, (b) presence of both parents at age 15, (c)
-  nationality, (d) age, (e) sex, and (f) ethnicity.
-\end{enumerate}
-
-These propensity scores are subsequently employed to adjust for
-selection into mobility when assessing its effect on market justice
-preferences.
-
-\textbf{Meritocracy}: Meritocratic perception is operationalized through
-two components: one addressing effort and another focusing on talent
-(\citeproc{ref-young_rise_1958}{Young, 1958}). The item used to gauge
-effort is: ``In Chile, people are rewarded for their efforts,'' while
-the item for talent is: ``In Chile, people are rewarded for their
-intelligence and skills.'' In both cases, respondents indicate their
-level of agreement on a five-point Likert scale, ranging from 1
-(``strongly disagree'') to 5 (``strongly agree'').
-
-\emph{Controls}
-
-Sociodemographic and attitudinal variables are included to control for
-potential composition effects in the population. In terms of
-sociodemographic characteristics, I incorporate per capita household
-income quantile, educational level (1=Primary or below, 2=Secondary,
-3=Technical, 4=University or above), age (in years), and sex (1=Male,
-2=Female), which have been shown to significantly influence market
-justice preferences (\citeproc{ref-castillo_socialization_2024}{Castillo
-et al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
-attitudinal variables, I include political identification (1=Left,
-2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) because they may confound the relationship
-between market justice preferences and social mobility and meritocracy.
-
-\subsection{Methods}\label{methods}
-
-Following Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024})
-strategy for estimating the causal effect of social mobility on market
-justice preferences, I employ linear regression models with
-inverse-probability weights (IPW) for mobility conditional on class of
-origin. This approach allows estimation of the average causal effect of
-moving from an origin class \emph{j} to a destination class \emph{k} by
-comparing individuals who actually reach \emph{k} with their
-counterfactual outcome had they instead moved to an alternative
-destination \emph{k′} (\citeproc{ref-breen_effects_2024}{Breen \&
-Ermisch, 2024}). The resulting estimand is the average treatment effect
-on the treated (ATT).
-
-All the analyses will be conducted using R software.
-
-\section{References}\label{references}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-alesina_intergenerational_2018}
-Alesina, A., Stantcheva, S., \& Teso, E. (2018). Intergenerational
-{Mobility} and {Preferences} for {Redistribution}. \emph{American
-Economic Review}, \emph{108}(2), 521--554.
-\url{https://doi.org/10.1257/aer.20162015}
-
-\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
-Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
-durante la dictadura}}. Libros de la Catarata.
-
-\bibitem[\citeproctext]{ref-arteagaaguirre_politicas_2015}
-Arteaga Aguirre, C., \& Iñigo Valderrama, I. (2015). {Pol{í}ticas
-sociales, modelo de desarrollo y subjetividad de grupos vulnerables en
-Chile}.
-
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023). Belief in {School Meritocracy} and the {Legitimization} of
-{Social} and {Income Inequality}. \emph{Social Psychological and
-Personality Science}, \emph{14}(5), 621--635.
-\url{https://doi.org/10.1177/19485506221111017}
-
-\bibitem[\citeproctext]{ref-boccardo_30_2020}
-Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
-que la pandemia revel{ó}} (Nodo XXI). Santiago.
-
-\bibitem[\citeproctext]{ref-breen_effects_2024}
-Breen, R., \& Ermisch, J. (2024). The {Effects} of {Social Mobility}.
-\emph{Sociological Science}, \emph{11}, 467--488.
-\url{https://doi.org/10.15195/v11.a17}
-
-\bibitem[\citeproctext]{ref-busemeyer_skills_2014}
-Busemeyer, M. R. (2014). \emph{Skills and {Inequality}: {Partisan
-Politics} and the {Political Economy} of {Education Reforms} in {Western
-Welfare States}}. Cambridge University Press.
-
-\bibitem[\citeproctext]{ref-castillo_socialization_2024}
-Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
-{Socialization} of {Meritocracy} and {Market Justice Preferences} at
-{School}. \emph{Societies}, \emph{14}(11), 214.
-\url{https://doi.org/10.3390/soc14110214}
-
-\bibitem[\citeproctext]{ref-chancel_world_2022}
-Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
-inequality report 2022.
-https://bibliotecadigital.ccb.org.co/handle/11520/27510.
-
-\bibitem[\citeproctext]{ref-davis_principles_2001}
-Davis, K., \& Moore, W. E. (2001). Some {Principles} of
-{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
-{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
-Routledge.
-
-\bibitem[\citeproctext]{ref-day_movin_2017}
-Day, M. V., \& Fiske, S. T. (2017). Movin' on {Up}? {How Perceptions} of
-{Social Mobility Affect Our Willingness} to {Defend} the {System}.
-\emph{Social Psychological and Personality Science}, \emph{8}(3),
-267--274. \url{https://doi.org/10.1177/1948550616678454}
-
-\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
-Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
-movilidad social bajo un}.
-
-\bibitem[\citeproctext]{ref-ferre_welfare_2023}
-Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
-America}. \emph{Journal of International and Comparative Social Policy},
-\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}
-
-\bibitem[\citeproctext]{ref-flores_top_2020}
-Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
-in {Chile}: {A Historical Perspective} on {Income Inequality},
-1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
-\url{https://doi.org/10.1111/roiw.12441}
-
-\bibitem[\citeproctext]{ref-gingrich_making_2011}
-Gingrich, J. R. (2011). \emph{Making {Markets} in the {Welfare State}:
-{The Politics} of {Varying Market Reforms}} (1st ed.). Cambridge
-University Press. \url{https://doi.org/10.1017/CBO9780511791529}
-
-\bibitem[\citeproctext]{ref-gugushvili_trends_2014}
-Gugushvili, A. (2014). Trends, {Covariates} and {Consequences} of
-{Intergenerational Social Mobility} in {Post- Socialist Societies}.
-
-\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016c}
-Gugushvili, A. (2016a). Intergenerational objective and subjective
-mobility and attitudes towards income differences: Evidence from
-transition societies. \emph{Journal of International and Comparative
-Social Policy}, \emph{32}(3), 199--219.
-\url{https://doi.org/10.1080/21699763.2016.1206482}
-
-\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016}
-Gugushvili, A. (2016b). Intergenerational {Social Mobility} and {Popular
-Explanations} of {Poverty}: {A Comparative Perspective}. \emph{Social
-Justice Research}, \emph{29}(4), 402--428.
-\url{https://doi.org/10.1007/s11211-016-0275-9}
-
-\bibitem[\citeproctext]{ref-gugushvili_subjective_2017}
-Gugushvili, A. (2017). Subjective {Intergenerational Mobility} and
-{Support} for {Welfare State Programmes}.
-
-\bibitem[\citeproctext]{ref-harvey_breve_2015}
-Harvey, D. (2015). \emph{{Breve historia del neoliberalismo}}. Madrid
-(Espa{ñ}a): Ediciones Akal.
-
-\bibitem[\citeproctext]{ref-immergut_it_2020}
-Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
-affluent to be able to purchase {``better''} healthcare? {Existential}
-standards and institutional norms in healthcare attitudes across 28
-countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
-\url{https://doi.org/10.1016/j.socscimed.2020.113146}
-
-\bibitem[\citeproctext]{ref-jacksonActorsInstitutions2010}
-Jackson, G. (2010). Actors and institutions. In G. Morgan, J. L.
-Campbell, C. Crouch, O. K. Pedersen, \& R. Whitley (Eds.), \emph{The
-{Oxford Handbook} of {Comparative Institutional Analysis}} (1st ed.).
-Oxford University Press.
-\url{https://doi.org/10.1093/oxfordhb/9780199233762.001.0001}
-
-\bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
-Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social mobility
-and demand for redistribution in {Europe}: A comparative analysis.
-\emph{The British Journal of Sociology}, \emph{70}(1), 138--165.
-\url{https://doi.org/10.1111/1468-4446.12363}
-
-\bibitem[\citeproctext]{ref-koos_moral_2019}
-Koos, S., \& Sachweh, P. (2019). The moral economies of market
-societies: Popular attitudes towards market competition, redistribution
-and reciprocity in comparative perspective. \emph{Socio-Economic
-Review}, \emph{17}(4), 793--821.
-\url{https://doi.org/10.1093/ser/mwx045}
-
-\bibitem[\citeproctext]{ref-lee_fairness_2023}
-Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
-inequality: The effects of self-interest and neoliberal orientations.
-\emph{The Australian Educational Researcher}.
-\url{https://doi.org/10.1007/s13384-023-00636-6}
-
-\bibitem[\citeproctext]{ref-lindh_public_2015}
-Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
-towards {Market Distribution} of {Social Services} -- {A Comparison} of
-17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
-887--910. \url{https://doi.org/10.1111/spol.12105}
-
-\bibitem[\citeproctext]{ref-lindh_bringing_2023}
-Lindh, A., \& McCall, L. (2023). Bringing the market in: An expanded
-framework for understanding popular responses to economic inequality.
-\emph{Socio-Economic Review}, \emph{21}(2), 1035--1055.
-\url{https://doi.org/10.1093/ser/mwac018}
-
-\bibitem[\citeproctext]{ref-lopez-roldan_comparative_2021}
-López-Roldán, P., \& Fachelli, S. (Eds.). (2021). \emph{Towards a
-{Comparative Analysis} of {Social Inequalities} between {Europe} and
-{Latin America}}. Cham: Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-48442-2}
-
-\bibitem[\citeproctext]{ref-madariaga_three_2020}
-Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
-economic policy trajectory in comparative perspective.
-\emph{Contemporary Politics}, \emph{26}(3), 308--329.
-\url{https://doi.org/10.1080/13569775.2020.1735021}
-
-\bibitem[\citeproctext]{ref-mau_inequality_2015}
-Mau, S. (2015). \emph{Inequality, {Marketization} and the {Majority
-Class}: {Why Did} the {European Middle Classes Accept Neo-Liberalism}?}
-Houndmills: Palgrave Macmillan.
-
-\bibitem[\citeproctext]{ref-mijs_paradox_2019}
-Mijs, J. (2019). The paradox of inequality: Income inequality and belief
-in meritocracy go hand in hand. \emph{Socio-Economic Review},
-\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
-
-\bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
-Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
-{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
-\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}
-
-\bibitem[\citeproctext]{ref-miller_selfserving_1975}
-Miller, D. T., \& Ross, M. (1975). Self-serving biases in the
-attribution of causality: {Fact} or fiction? \emph{Psychological
-Bulletin}, \emph{82}(2), 213--225.
-\url{https://doi.org/10.1037/h0076486}
-
-\bibitem[\citeproctext]{ref-otero_power_2024}
-Otero, G., \& Mendoza, M. (2024). The {Power} of {Diversity}: {Class},
-{Networks} and {Attitudes Towards Inequality}. \emph{Sociology},
-\emph{58}(4), 851--876. \url{https://doi.org/10.1177/00380385231217625}
-
-\bibitem[\citeproctext]{ref-polanyi_great_1975}
-Polanyi, K. (1975). \emph{The great transformation} (Repr). New York,
-NY: Octagon Books.
-
-\bibitem[\citeproctext]{ref-schmidt_experience_2011}
-Schmidt, A. W. (2011). The experience of social mobility and the
-formation of attitudes toward redistribution. In.
-
-\bibitem[\citeproctext]{ref-shariff_income_2016}
-Shariff, A. F., Wiwad, D., \& Aknin, L. B. (2016). Income {Mobility
-Breeds Tolerance} for {Income Inequality}: {Cross-National} and
-{Experimental Evidence}. \emph{Perspectives on Psychological Science},
-\emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}
-
-\bibitem[\citeproctext]{ref-somma_no_2021}
-Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
-water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
-Movement Studies}, \emph{20}(4), 495--502.
-\url{https://doi.org/10.1080/14742837.2020.1727737}
-
-\bibitem[\citeproctext]{ref-streeck_how_2016}
-Streeck, W. (2016). \emph{How will capitalism end? Essays on a failing
-system}. London: Verso.
-
-\bibitem[\citeproctext]{ref-svallforsMoralEconomyClass2006a}
-Svallfors, S. (2006). \emph{The {Moral Economy} of {Class}: {Class} and
-{Attitudes} in {Comparative Perspective}}. Stanford University Press.
-
-\bibitem[\citeproctext]{ref-svallfors_political_2007}
-Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
-{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
-(1st ed.). Stanford University Press.
-\url{https://doi.org/10.2307/j.ctvr0qv0q}
-
-\bibitem[\citeproctext]{ref-torche_intergenerational_2014}
-Torche, F. (2014). Intergenerational {Mobility} and {Inequality}: {The
-Latin American Case}. \emph{Annual Review of Sociology}, \emph{40}(1),
-619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
-
-\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
-Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
-care inequalities unfair? {A} study on public attitudes in 23 countries.
-\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
-\url{https://doi.org/10.1186/s12939-016-0350-8}
-
-\bibitem[\citeproctext]{ref-wen_does_2021}
-Wen, F., \& Witteveen, D. (2021). Does perceived social mobility shape
-attitudes toward government and family educational investment?
-\emph{Social Science Research}, \emph{98}, 102579.
-\url{https://doi.org/10.1016/j.ssresearch.2021.102579}
-
-\bibitem[\citeproctext]{ref-wilson_role_2003}
-Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
-Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
-\url{https://doi.org/10.1023/A:1024667228488}
-
-\bibitem[\citeproctext]{ref-young_rise_1958}
-Young, M. (1958). \emph{The rise of the meritocracy}. New Brunswick,
-N.J., U.S.A: Transaction Publishers.
-
-\end{CSLReferences}
+The findings contribute to the literature by highlighting the dynamic
+and interactive nature of distributive preferences. Rather than being
+solely determined by fixed class position, attitudes toward market
+justice are shaped by experiences of mobility and the normative
+frameworks individuals use to interpret them. In contexts of high
+inequality and welfare commodification, such as Chile, upward mobility
+may legitimize neoliberal moral orders by reinforcing beliefs in
+individual responsibility and merit-based allocation. These insights
+underscore the importance of incorporating mobility trajectories and
+belief systems into the analysis of welfare attitudes and inequality
+legitimation.
 

### 2025-06-27 advances/informe2.qmd
Up informe 3

@@
     theme:
-      - cosmo
       - pretty.scss
@@
     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
-citeproc: true
 link-citations: true
-colorlinks: true
-linestretch: 1.15       
+linestretch: 1.5       
 mainfont: Times New Roman
-fontsize: 12pt          
+fontsize: 13pt          
 fig-height: 4
@@
 Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016].
-In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@arrizabalo_milagro_1995; @ferre_welfare_2023].
+In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023].
 This transformation reduced the role of public provision and increased the presence of private actors in core social services [@harvey_breve_2015].
@@
 
-Although redistribution typically refers to the state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023].
+Beyond state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023].
 As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks.
@@
 In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007].
-While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@immergut_theoretical_1998].
+While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@lindh_bringing_2023].
 In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023].
 Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020].
-This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@lindh_public_2015; @mau_inequality_2015].
+This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@satz_por_2019; @mau_inequality_2015].
 
 The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles.
-Beyond stratification structures and socioeconomic conditions, the study of inequality also examines the beliefs individuals hold about the causes of social inequality, the normative principles that underpin those beliefs, the factors that shape them, and their consequences for attitudes and behavior [@kluegel_beliefs_1981].
+Beyond stratification structures and socioeconomic conditions, the study of inequality also focuses on individuals’ beliefs about the origins of inequality, the normative frameworks that sustain those beliefs, the mechanisms that shape them, and their implications for attitudes and behavior [@kluegel_beliefs_1981].
 In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016].
 This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011].
-One prominent strand investigates the legitimacy of wage inequality, particularly salary differences across occupations [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978].
-Another examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, and pensions.
+Most of the research on this field explores the legitimacy of wage inequality, specially salary gaps between jobs at opposite ends of the occupational hierarchy [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978].
+Recently, others examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, pensions or social security.
 Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015].
@@
 
-The empirical study of market justice preferences has relied on diverse strategies to operationalize how individuals evaluate inequalities produced by market allocation.
-One common approach assesses whether people consider it fair that access to key social services—such as healthcare, education, or pensions—depends on income.
-This strategy builds on foundational work by Kluegel and Smith [-@kluegel_beliefs_1981], who explored the normative justifications behind support for economic inequality.
-More recent studies have expanded this logic beyond income to include market-mediated access to welfare goods.
-For instance, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
-[-@vondemknesebeck_are_2016] examine whether respondents believe it is just that those with higher incomes receive better healthcare.
-Similarly, Lee and Stacey [-@lee_fairness_2023] apply this approach to educational opportunities.
-These studies typically use targeted survey items that ask respondents to evaluate specific distributive scenarios, allowing for comparative analyses across countries or welfare regimes.
-Complementing these efforts, more recent research has introduced composite indicators designed to capture broader orientations toward market-based allocation.
-For example, Castillo et al.
-[-@castillo_socialization_2024] propose a single-item measure that summarizes individual support for income-based access across multiple domains—health, education, and pensions—within the Chilean context.
-Together, these instruments aim to capture to what extent individuals perceive market-generated inequality as justified.
+The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation.
+A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income.
+This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems.
+Subsequent studies have extended this logic to a wider set of welfare goods.
+For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
+[-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare.
+Similarly, Lee and Stacey [-@lee_fairness_2023] apply this framework to education.
+Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation.
+Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al.
+[-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile.
+These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate.
+Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [-@lindh_public_2015].
 
@@
 Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility.
-Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support.
+Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [-@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support.
 However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.
@@
 
-These findings suggest that causal attribution processes mediate the relationship between mobility experiences and justice-related attitudes. Accordingly, this study proposes the following hypothesis:
+Consequently, considering this theoretical and empirical background, the first hypothesis of this research is that:
 

### 2025-06-27 advances/informe2.tex
Up informe 3

@@
 \documentclass[
-  12pt,
+  13pt,
 ]{article}
@@
 
-\setstretch{1.15}
+\setstretch{1.5}
 \section{Introduction}\label{introduction}
@@
 reproduction that were traditionally governed by the state
-(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
-\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
+(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
 reduced the role of public provision and increased the presence of
@@
 
-Although redistribution typically refers to the state's capacity to
-reallocate resources from the advantaged to the vulnerable, market
-institutions also play a central role in distributing socially valuable
-goods and rewards (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
-2019}; \citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As
-Polanyi (\citeproc{ref-polanyi_great_1975}{1975}) observed, economic
-integration in capitalist societies is primarily organized through
-market exchange, governed by a self-regulating price system embedded in
-institutional frameworks. These institutions are not mere aggregates of
-individual behavior, but social realities endowed with rules,
-mechanisms, and normative meanings that shape everyday thinking
+Beyond state's capacity to reallocate resources from the advantaged to
+the vulnerable, market institutions also play a central role in
+distributing socially valuable goods and rewards
+(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As Polanyi
+(\citeproc{ref-polanyi_great_1975}{1975}) observed, economic integration
+in capitalist societies is primarily organized through market exchange,
+governed by a self-regulating price system embedded in institutional
+frameworks. These institutions are not mere aggregates of individual
+behavior, but social realities endowed with rules, mechanisms, and
+normative meanings that shape everyday thinking
 (\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
@@
 distributive justice beliefs and institutional responses to inequality
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In many
+(\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In many
 countries, privatization and commodification have expanded market logic
@@
 structure access to resources and legitimize inequalities
-(\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
+\citeproc{ref-satz_por_2019}{Satz, 2019}).
 
@@
 stratification structures and socioeconomic conditions, the study of
-inequality also examines the beliefs individuals hold about the causes
-of social inequality, the normative principles that underpin those
-beliefs, the factors that shape them, and their consequences for
-attitudes and behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \&
-Smith, 1981}). In this regard, empirical research on distributive
-justice focuses on individuals' conceptions of how goods and rewards
-\emph{should} be distributed in society
+inequality also focuses on individuals' beliefs about the origins of
+inequality, the normative frameworks that sustain those beliefs, the
+mechanisms that shape them, and their implications for attitudes and
+behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
+In this regard, empirical research on distributive justice focuses on
+individuals' conceptions of how goods and rewards \emph{should} be
+distributed in society
 (\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
@@
 legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
-One prominent strand investigates the legitimacy of wage inequality,
-particularly salary differences across occupations
+Most of the research on this field explores the legitimacy of wage
+inequality, specially salary gaps between jobs at opposite ends of the
+occupational hierarchy
 (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
 \citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
-\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Another
-examines how individuals justify market-generated inequalities in access
-to core social services such as healthcare, education, and pensions.
-Here, legitimacy stems from the belief that access to these goods should
-follow market-based criteria
+\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Recently,
+others examines how individuals justify market-generated inequalities in
+access to core social services such as healthcare, education, pensions
+or social security. Here, legitimacy stems from the belief that access
+to these goods should follow market-based criteria
 (\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
@@
 
-The empirical study of market justice preferences has relied on diverse
-strategies to operationalize how individuals evaluate inequalities
-produced by market allocation. One common approach assesses whether
-people consider it fair that access to key social services---such as
-healthcare, education, or pensions---depends on income. This strategy
-builds on foundational work by Kluegel and Smith
-(\citeproc{ref-kluegel_beliefs_1981}{1981}), who explored the normative
-justifications behind support for economic inequality. More recent
-studies have expanded this logic beyond income to include
-market-mediated access to welfare goods. For instance, Immergut and
+The empirical study of market justice preferences has employed various
+strategies to capture how individuals assess inequalities arising from
+market allocation. A common approach evaluates whether people find it
+fair that access to essential social services---such as healthcare,
+education, or pensions---depends on income. This builds on foundational
+work by Kluegel and Smith
+(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
+justifications for capitalist systems. Subsequent studies have extended
+this logic to a wider set of welfare goods. For example, Immergut and
 Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
-et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) examine whether
-respondents believe it is just that those with higher incomes receive
+et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
+respondents believe it is just that higher-income individuals receive
 better healthcare. Similarly, Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) apply this approach to
-educational opportunities. These studies typically use targeted survey
-items that ask respondents to evaluate specific distributive scenarios,
-allowing for comparative analyses across countries or welfare regimes.
-Complementing these efforts, more recent research has introduced
-composite indicators designed to capture broader orientations toward
-market-based allocation. For example, Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) propose a single-item
-measure that summarizes individual support for income-based access
-across multiple domains---health, education, and pensions---within the
-Chilean context. Together, these instruments aim to capture to what
-extent individuals perceive market-generated inequality as justified.
+(\citeproc{ref-lee_fairness_2023}{2023}) apply this framework to
+education. Complementing these efforts, more recent research has
+introduced composite indicators to capture broader orientations toward
+market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
+for instance, constructs an index averaging support for income-based
+access to healthcare and education in comparative perspective, while
+Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
+propose a single-item measure covering health, education, and pensions
+in Chile. These instruments seek to gauge to what extent individuals
+view market-generated inequalities as legitimate. Taken together, they
+capture two core dimensions of market distribution: the role of economic
+resources as a key determinant of outcomes, and the framing of services
+as tradable commodities that can be bought and sold according to ability
+to pay (\citeproc{ref-lindh_public_2015}{2015}).
 
@@
 studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
-Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017})
-likewise report that subjective upward mobility is associated with
-weaker preferences for redistribution, while downward mobility
-strengthens redistributive support. However, recent causal evidence from
-Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the
-opposite: upward mobility may increase redistributive preferences, while
-downward mobility may reduce them.
+Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{2017}) likewise
+report that subjective upward mobility is associated with weaker
+preferences for redistribution, while downward mobility strengthens
+redistributive support. However, recent causal evidence from Breen and
+Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the opposite:
+upward mobility may increase redistributive preferences, while downward
+mobility may reduce them.
 
@@
 
-These findings suggest that causal attribution processes mediate the
-relationship between mobility experiences and justice-related attitudes.
-Accordingly, this study proposes the following hypothesis:
+Consequently, considering this theoretical and empirical background, the
+first hypothesis of this research is that:
 
@@
 
-\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
-Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
-durante la dictadura}}. Libros de la Catarata.
-
 \bibitem[\citeproctext]{ref-batruch_belief_2023}
@@
 
+\bibitem[\citeproctext]{ref-satz_por_2019}
+Satz, D. (2019). \emph{{Por Qu{é} Algunas Cosas No Deber{í}an Estar en
+Venta: Los l{í}mites Morales Del Mercado}}. Ciudad Aut{ó}noma de Buenos
+Aires: Siglo XXI Editores.
+
 \bibitem[\citeproctext]{ref-schmidt_experience_2011}

### 2025-06-27 advances/informe3.qmd
Up informe 3

--- 2025-06-27 advances/informe2.qmd
@@
 In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023].
-This transformation reduced the role of public provision and increased the presence of private actors in core social services [@harvey_breve_2015].
+This transformation reduced the role of public provision and increased the presence of private actors in core social services [@ferre_welfare_2023].
 Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a].
-Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
+Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @castillo_perceptions_2025; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
 Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].
@@
 Among these, meritocracy is a key normative principle underpinning market-based distributive preferences [@mau_inequality_2015].
-It frames inequality as inevitable but justifiable through effort and talent [@young_rise_1958; @davis_principles_2001].
 Studies show that individuals with stronger meritocratic beliefs tend to perceive less inequality and legitimize it by attributing economic differences to personal achievement [@mijs_paradox_2019; @wilson_role_2003; @batruch_belief_2023].
@@
 [-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare.
-Similarly, Lee and Stacey [-@lee_fairness_2023] apply this framework to education.
+Similarly, Lee and Stacey [-@lee_fairness_2023] and @castillo_changes_2025 apply this framework to education.
 Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation.
@@
 These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate.
-Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [-@lindh_public_2015].
+Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [@lindh_public_2015].
 
@@
 Beliefs about inequality and political orientation also matter.
-In Chile, Castillo et al.
-[-@castillo_socialization_2024] show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism.
+Castillo et al.
+[-@castillo_socialization_2024] and @castillo_perceptions_2025 show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution in Chile, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism.
 In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.
@@
 Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
-Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_meritocracia_2019; @castillo_socialization_2024].
+Prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_perceptions_2025; @castillo_socialization_2024].
 
@@
 
+Chile offers a compelling case for examining individual preferences for market justice in the access to core social services.
+Since the early 1990s, the country experienced a long period of sustained economic growth, with per capita GDP increasing by nearly 4% annually until the mid-2010s [@llorca-jana_historia_2021].
+This growth brought notable improvements in living standards, including a strong decline in income poverty—from 38.6% in 1990 to 6.5% in 2022 [@mideplan_evolucion_2017].
+However, this success story has been tempered by growing concerns over the unequal distribution of its benefits.
+In recent years, economic growth has slowed—exacerbated by the global impact of the COVID-19 pandemic—and structural inequalities have become more visible [@barozet_clases_2021].
+Despite its economic achievements, Chile remains one of the most unequal countries in Latin America and the OCDE.
+The poorest 50% of the population captures just 10% of total income and holds negative net wealth, while the wealthiest 1% receives nearly 27% of all income and controls almost half (49.6%) of the country’s wealth [@chancel_world_2022], a figure that has seen little change since the 1990s [@flores_top_2020].
+These stark disparities extend beyond income, reflecting deep inequalities in access to essential social services [@pnud_desiguales_2017].
+
+A significant share of Chile’s inequality stems from neoliberal reforms that privatized and commodified key domains of social reproduction [@arrizabalo_milagro_1995; @ferre_welfare_2023].
+Introduced during the military dictatorship (1973–1989) and expanded under democratic governments, these reforms embedded market logic into public services through concessions, subsidies, demand-side vouchers, and pro-private regulatory frameworks [@madariaga_three_2020; @boccardo_30_2020].
+This "crowded-out" welfare model benefits higher-income groups, leaving lower-income individuals to rely on limited public options.
+Scholars argue that this neoliberal shift repurposed the state to create niches of capitalist accumulation -from water to education-, giving rise to a model of “public-service capitalism” heavily reliant on state funding [@boccardo_30_2020].
+In healthcare, around 79% of the population is covered by the public insurer FONASA, while 15% are enrolled in private ISAPREs, which offer faster and higher-quality services to wealthier groups [@mideplan_estadisticas_2024].
+The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors through private fund administrators (AFPs), yet 27% of the labor force remains outside the system due to informality [@superintendenciadepensiones_estadisticas_2024].
+Education is similarly stratified: only 30.6% of students attend fully public schools, while 54% are in voucher-funded private institutions and 9.3% in fully private schools that largely serve affluent families [@ministeriodeeducacion_resumen_2023].
+This arrangement has created a structurally segmented system, where access and quality of services depend on ability to pay, reinforced by state subsidies to private providers [@pnud_desiguales_2017].
+
+Over recent decades, Chile has undergone a significant transformation in its class structure, shaped by sustained economic growth, the expansion of the service sector, the massification of education, and major changes in the role of the state and social policy [@ruiz_chilenos_2014].
+These changes contributed to the growth of the middle sectors, a phenomena labelled as “mesocratrization” of society [@espinoza_estratificacion_2013].
+However, this expansion did not reduce structural inequalities nor facilitate upward mobility into elite positions [@espinoza_estratificacion_2013; @lopez-roldan_comparative_2021].
+While Chile shows high absolute mobility, most flows occur between adjacent classes—particularly from working classes into lower service-class positions—while access to upper classes remains limited, indicating strong elite closure and reproduction [@espinoza_estratificacion_2013; @lopez-roldan_comparative_2021; @torche_unequal_2005; @espinoza_movilidad_2014; @perez-ahumada_clases_2019].
+Though often described as “unequal but fluid”, recent evidence points to growing rigidity in relative mobility, especially after the economic slowdown, increased household debt, labor market precarity, and the COVID-19 crisis [@barozet_clases_2021].
+However, the majority of the population identifies themselves as members of the middle class [@castillo_todos_2013].
+These dynamics have fueled a legitimacy crisis: meritocracy is increasingly questioned, and many perceive that advancement depends more on informal networks than on effort or earned credentials [@barozet_clases_2021].
+The combination of constrained mobility, persistent inequality, and concentrated opportunity has eroded expectations of upward mobility and intensified social discontent [@pnud_desiguales_2017].
+
+Despite recurrent social unrest, Chile presents a paradoxical coexistence between strong conflict over inequality and widespread public legitimation of it.
+The October 2019 “social outburst”, a period of mass protests and severe political repression throughout the country,  crystallized discontent over the privatization and commodification of social services, alongside a crisis of political legitimacy [@somma_no_2021].
+A survey during the protests identified pensions, healthcare, and education as the top demands [@nucleodesociologiacontingente_informe_2020], reflecting dissatisfaction with Chile’s stratified welfare regime.
+Yet, this unrest coexists with strong meritocratic beliefs and high tolerance for income inequality.
+Castillo [-@castillo_legitimacy_2011] shows that wage gaps between occupations are consistently justified, and the larger the perceived gap, the greater its legitimation, an effect amplified by Chile’s structural inequality.
+Moreover, perceptions of unfairness due to non-meritocratic factors often reinforce belief in effort and talent as legitimate bases for success [@castillo_multidimensional_2023], which are associated with lower perceived injustice [@castillo_meritocracia_2019].
+Mac-Clure et al.
+[@mac-clure_justicia_2024] find that low and lower-middle status individuals are less likely than upper-status groups to view educational income differences as unfair.
+Qualitative research further reveals the internalization of the enterprising self [@mau_inequality_2015], understood as a way in which the subjectivization process arises in line with positioning the self in the market, generating its own market value and subjecting it to competition.
+Canales et al.
+[-@canalesceron_sujeto_2021] describes how families navigate school choice as investment decisions, while Panes [-@panes_criticas_2020] finds that many workers frame pension contributions as individual investments aimed at maximizing future returns.
+These representations suggest that market-based moral orientations are deeply embedded in Chilean subjectivities, offering fertile ground to study how support for market justice varies across welfare domains.
+
+Although still limited, recent research on market justice preferences in Chile indicates that support for these beliefs has increased in recent years across domains such as healthcare, pensions, and education, and that both objective and subjective dimensions of inequality shape them.
+Castillo et al.
+[-@castillo_perceptions_2025] show that agreement with the notion that it is fair for higher-income individuals to access better services was relatively low in 2016, but rose significantly by 2023—especially in the pension domain.
+Otero and Mendoza [-@otero_power_2024] provide evidence that social class, measured using the EGP scheme, influences these preferences: individuals in lower or subordinate class positions are less supportive of market-based access to welfare services compared to those in higher or privileged positions.
+Additionally, greater diversity in social class networks is associated with lower support for market justice.
+This is particularly relevant in the Chilean context, where one of the main determinants of class-diverse networks is intergenerational social mobility, which exposes individuals to relationships with people from different class backgrounds [@otero_lives_2022].
+Regarding subjective factors, Castillo et al.
+[@castillo_perceptions_2025; @castillo_socialization_2024] find that individuals who endorse meritocratic beliefs—such as the idea that effort and talent are rewarded in Chile—are more likely to support market justice across welfare domains.
+
 # Data, variables and methods
@@
 This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample.
-Further details on sampling, attrition, and weighting can be found at https://coes.cl/encuesta-panel/, and the dataset is publicly available at https://dataverse.harvard.edu/dataverse/elsoc.
+Further details on sampling, attrition, and weighting can be found at [https://coes.cl/encuesta-panel/](https://coes.cl/encuesta-panel/), and the dataset is publicly available at [https://dataverse.harvard.edu/dataverse/elsoc](https://dataverse.harvard.edu/dataverse/elsoc).
 

### 2025-06-27 advances/informe3.tex
Up informe 3

--- 2025-06-27 advances/informe2.tex
@@
 private actors in core social services
-(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
+(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Echoing Polanyi's
 (\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
@@
 justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
-2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
+2014}; \citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et
+al., 2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
 2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
@@
 market-based distributive preferences
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). It frames inequality as
-inevitable but justifiable through effort and talent
-(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
-\citeproc{ref-young_rise_1958}{Young, 1958}). Studies show that
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Studies show that
 individuals with stronger meritocratic beliefs tend to perceive less
@@
 individuals' conceptions of how goods and rewards \emph{should} be
-distributed in society
-(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
-2016}). This line of inquiry allows for examining the extent to which
-economic inequality is perceived as just, or in a certain sense,
+distributed in society (\citeproc{ref-jasso_distributive_2016}{Jasso et
+al., 2016}). This line of inquiry allows for examining the extent to
+which economic inequality is perceived as just, or in a certain sense,
 legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
@@
 (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
-\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
-\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Recently,
+\citeproc{ref-jasso_justice_1978}{Jasso, 1978};
+\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}). Recently,
 others examines how individuals justify market-generated inequalities in
@@
 better healthcare. Similarly, Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) apply this framework to
+(\citeproc{ref-lee_fairness_2023}{2023}) and Castillo, Iturra, et al.
+(\citeproc{ref-castillo_changes_2025}{2025}) apply this framework to
 education. Complementing these efforts, more recent research has
@@
 as tradable commodities that can be bought and sold according to ability
-to pay (\citeproc{ref-lindh_public_2015}{2015}).
+to pay (\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
 
@@
 market allocation in healthcare, education, and pensions. Beliefs about
-inequality and political orientation also matter. In Chile, Castillo et
-al. (\citeproc{ref-castillo_socialization_2024}{2024}) show that
-individuals who have strong meritocratic beliefs are more likely to
-support market-based distribution, while Lee and Stacy
+inequality and political orientation also matter. Castillo et al.
+(\citeproc{ref-castillo_socialization_2024}{2024}) and Castillo,
+Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
+that individuals who have strong meritocratic beliefs are more likely to
+support market-based distribution in Chile, while Lee and Stacy
 (\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
@@
 unequal outcomes, particularly when access is stratified by income or
-social opportunity. Prior studies have shown that individuals who
-perceive higher levels of meritocracy tend to express stronger support
-for unequal distributions that reflect market outcomes
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
+social opportunity. Prior studies in Chile have shown that individuals
+who perceive higher levels of meritocracy tend to express stronger
+support for unequal distributions that reflect market outcomes in
+healthcare, education and pensions
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
+2024}).
 
@@
 
+Chile offers a compelling case for examining individual preferences for
+market justice in the access to core social services. Since the early
+1990s, the country experienced a long period of sustained economic
+growth, with per capita GDP increasing by nearly 4\% annually until the
+mid-2010s (\citeproc{ref-llorca-jana_historia_2021}{Llorca-Jaña \&
+Miller, 2021}). This growth brought notable improvements in living
+standards, including a strong decline in income poverty---from 38.6\% in
+1990 to 6.5\% in 2022 (\citeproc{ref-mideplan_evolucion_2017}{MIDEPLAN,
+2017}). However, this success story has been tempered by growing
+concerns over the unequal distribution of its benefits. In recent years,
+economic growth has slowed---exacerbated by the global impact of the
+COVID-19 pandemic---and structural inequalities have become more visible
+(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). Despite its
+economic achievements, Chile remains one of the most unequal countries
+in Latin America and the OCDE. The poorest 50\% of the population
+captures just 10\% of total income and holds negative net wealth, while
+the wealthiest 1\% receives nearly 27\% of all income and controls
+almost half (49.6\%) of the country's wealth
+(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
+has seen little change since the 1990s
+(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). These stark
+disparities extend beyond income, reflecting deep inequalities in access
+to essential social services (\citeproc{ref-pnud_desiguales_2017}{PNUD,
+2017}).
+
+A significant share of Chile's inequality stems from neoliberal reforms
+that privatized and commodified key domains of social reproduction
+(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
+\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Introduced during the
+military dictatorship (1973--1989) and expanded under democratic
+governments, these reforms embedded market logic into public services
+through concessions, subsidies, demand-side vouchers, and pro-private
+regulatory frameworks (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This
+``crowded-out'' welfare model benefits higher-income groups, leaving
+lower-income individuals to rely on limited public options. Scholars
+argue that this neoliberal shift repurposed the state to create niches
+of capitalist accumulation -from water to education-, giving rise to a
+model of ``public-service capitalism'' heavily reliant on state funding
+(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In healthcare, around
+79\% of the population is covered by the public insurer FONASA, while
+15\% are enrolled in private ISAPREs, which offer faster and
+higher-quality services to wealthier groups
+(\citeproc{ref-mideplan_estadisticas_2024}{MIDEPLAN, 2024}). The pension
+system is based on individual capitalization, with mandatory
+contributions managed by private administrators investing in the
+financial market, currently involving 11 million contributors through
+private fund administrators (AFPs), yet 27\% of the labor force remains
+outside the system due to informality
+(\citeproc{ref-superintendenciadepensiones_estadisticas_2024}{Superintendencia
+de Pensiones, 2024}). Education is similarly stratified: only 30.6\% of
+students attend fully public schools, while 54\% are in voucher-funded
+private institutions and 9.3\% in fully private schools that largely
+serve affluent families
+(\citeproc{ref-ministeriodeeducacion_resumen_2023}{Ministerio de
+Educación, 2023}). This arrangement has created a structurally segmented
+system, where access and quality of services depend on ability to pay,
+reinforced by state subsidies to private providers
+(\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).
+
+Over recent decades, Chile has undergone a significant transformation in
+its class structure, shaped by sustained economic growth, the expansion
+of the service sector, the massification of education, and major changes
+in the role of the state and social policy
+(\citeproc{ref-ruiz_chilenos_2014}{Ruiz \& Boccardo, 2014}). These
+changes contributed to the growth of the middle sectors, a phenomena
+labelled as ``mesocratrization'' of society
+(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013}).
+However, this expansion did not reduce structural inequalities nor
+facilitate upward mobility into elite positions
+(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
+\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
+2021}). While Chile shows high absolute mobility, most flows occur
+between adjacent classes---particularly from working classes into lower
+service-class positions---while access to upper classes remains limited,
+indicating strong elite closure and reproduction
+(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
+\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
+\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
+2021}; \citeproc{ref-perez-ahumada_clases_2019}{Pérez-Ahumada, 2019};
+\citeproc{ref-torche_unequal_2005}{Torche, 2005}). Though often
+described as ``unequal but fluid'', recent evidence points to growing
+rigidity in relative mobility, especially after the economic slowdown,
+increased household debt, labor market precarity, and the COVID-19
+crisis (\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}).
+However, the majority of the population identifies themselves as members
+of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo et al.,
+2013}). These dynamics have fueled a legitimacy crisis: meritocracy is
+increasingly questioned, and many perceive that advancement depends more
+on informal networks than on effort or earned credentials
+(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). The
+combination of constrained mobility, persistent inequality, and
+concentrated opportunity has eroded expectations of upward mobility and
+intensified social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD,
+2017}).
+
+Despite recurrent social unrest, Chile presents a paradoxical
+coexistence between strong conflict over inequality and widespread
+public legitimation of it. The October 2019 ``social outburst'', a
+period of mass protests and severe political repression throughout the
+country, crystallized discontent over the privatization and
+commodification of social services, alongside a crisis of political
+legitimacy (\citeproc{ref-somma_no_2021}{Somma et al., 2021}). A survey
+during the protests identified pensions, healthcare, and education as
+the top demands
+(\citeproc{ref-nucleodesociologiacontingente_informe_2020}{Núcleo de
+Sociología Contingente, 2020}), reflecting dissatisfaction with Chile's
+stratified welfare regime. Yet, this unrest coexists with strong
+meritocratic beliefs and high tolerance for income inequality. Castillo
+(\citeproc{ref-castillo_legitimacy_2011}{2011}) shows that wage gaps
+between occupations are consistently justified, and the larger the
+perceived gap, the greater its legitimation, an effect amplified by
+Chile's structural inequality. Moreover, perceptions of unfairness due
+to non-meritocratic factors often reinforce belief in effort and talent
+as legitimate bases for success
+(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}),
+which are associated with lower perceived injustice
+(\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
+Mac-Clure et al. (\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et
+al., 2024}) find that low and lower-middle status individuals are less
+likely than upper-status groups to view educational income differences
+as unfair. Qualitative research further reveals the internalization of
+the enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
+understood as a way in which the subjectivization process arises in line
+with positioning the self in the market, generating its own market value
+and subjecting it to competition. Canales et al.
+(\citeproc{ref-canalesceron_sujeto_2021}{2021}) describes how families
+navigate school choice as investment decisions, while Panes
+(\citeproc{ref-panes_criticas_2020}{2020}) finds that many workers frame
+pension contributions as individual investments aimed at maximizing
+future returns. These representations suggest that market-based moral
+orientations are deeply embedded in Chilean subjectivities, offering
+fertile ground to study how support for market justice varies across
+welfare domains.
+
+Although still limited, recent research on market justice preferences in
+Chile indicates that support for these beliefs has increased in recent
+years across domains such as healthcare, pensions, and education, and
+that both objective and subjective dimensions of inequality shape them.
+Castillo et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
+that agreement with the notion that it is fair for higher-income
+individuals to access better services was relatively low in 2016, but
+rose significantly by 2023---especially in the pension domain. Otero and
+Mendoza (\citeproc{ref-otero_power_2024}{2024}) provide evidence that
+social class, measured using the EGP scheme, influences these
+preferences: individuals in lower or subordinate class positions are
+less supportive of market-based access to welfare services compared to
+those in higher or privileged positions. Additionally, greater diversity
+in social class networks is associated with lower support for market
+justice. This is particularly relevant in the Chilean context, where one
+of the main determinants of class-diverse networks is intergenerational
+social mobility, which exposes individuals to relationships with people
+from different class backgrounds (\citeproc{ref-otero_lives_2022}{Otero
+et al., 2022}). Regarding subjective factors, Castillo et al.
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
+2024}) find that individuals who endorse meritocratic beliefs---such as
+the idea that effort and talent are rewarded in Chile---are more likely
+to support market justice across welfare domains.
+
 \section{Data, variables and methods}\label{data-variables-and-methods}
@@
 989 in the refreshment sample. Further details on sampling, attrition,
-and weighting can be found at https://coes.cl/encuesta-panel/, and the
-dataset is publicly available at
-https://dataverse.harvard.edu/dataverse/elsoc.
+and weighting can be found at \url{https://coes.cl/encuesta-panel/}, and
+the dataset is publicly available at
+\url{https://dataverse.harvard.edu/dataverse/elsoc}.
 
@@
 
+\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
+Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
+durante la dictadura}}. Libros de la Catarata.
+
+\bibitem[\citeproctext]{ref-barozet_clases_2021}
+Barozet, E., Contreras, D., Espinoza, V., Gayo, M., \& Méndez, M. L.
+(2021). \emph{{Clases medias en tiempos de crisis: vulnerabilidad
+persistente, desaf{í}os para la cohesi{ó}n y un nuevo pacto social en
+Chile}}. Santiago: Comisi{ó}n Econ{ó}mica para Am{é}rica Latina y el
+Caribe (CEPAL).
+
 \bibitem[\citeproctext]{ref-batruch_belief_2023}
@@
 
+\bibitem[\citeproctext]{ref-castillo_changes_2025}
+Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
+{Justification} of {Educational Inequalities}: {The Role} of
+{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
+Pandemic}. \emph{Social Justice Research}.
+\url{https://doi.org/10.1007/s11211-025-00458-0}
+
 \bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
@@
 
+\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
+Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
+Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
+{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
+\emph{Under Review at Frontiers in Sociology}.
+
+\bibitem[\citeproctext]{ref-castillo_todos_2013}
+Castillo, J. C., Miranda, D., \& Cabib, I. M. (2013). Todos somos de
+clase media: {Sobre} el estatus social subjetivo en {Chile}. \emph{Latin
+American Research Review}, \emph{48}(1), 155--173.
+\url{https://doi.org/10.1353/lar.2013.0006}
+
 \bibitem[\citeproctext]{ref-castillo_socialization_2024}
@@
 
-\bibitem[\citeproctext]{ref-davis_principles_2001}
-Davis, K., \& Moore, W. E. (2001). Some {Principles} of
-{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
-{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
-Routledge.
-
 \bibitem[\citeproctext]{ref-day_movin_2017}
@@
 Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
-movilidad social bajo un}.
+movilidad social bajo un modelo neoliberal: El caso de Chile}.
+\emph{Lavboratorio}, (25).
+
+\bibitem[\citeproctext]{ref-espinoza_movilidad_2014}
+Espinoza, V., \& Núñez, J. (2014). Movilidad ocupacional en {Chile}
+2001-2009. {\textquestiondown}{Desigualdad} de ingresos con igualdad de
+oportunidades? \emph{Revista Internacional de Sociolog{í}a},
+\emph{72}(1), 57--82. \url{https://doi.org/10.3989/ris.2011.11.08}
 
@@
 
-\bibitem[\citeproctext]{ref-harvey_breve_2015}
-Harvey, D. (2015). \emph{{Breve historia del neoliberalismo}}. Madrid
-(Espa{ñ}a): Ediciones Akal.
-
 \bibitem[\citeproctext]{ref-helgason_longterm_2023}
@@
 \bibitem[\citeproctext]{ref-jasso_justice_1978}
-Jasso, Guillermina. (1978). On the {Justice} of {Earnings}: {A New
-Specification} of the {Justice Evaluation Function}. \emph{American
-Journal of Sociology}, \emph{83}(6), 1398--1419.
+Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification}
+of the {Justice Evaluation Function}. \emph{American Journal of
+Sociology}, \emph{83}(6), 1398--1419.
 \url{https://doi.org/10.1086/226706}
@@
 \bibitem[\citeproctext]{ref-jasso_distributive_2016}
-Jasso, Guillermina, Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
+Jasso, G., Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
 {Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
@@
 
+\bibitem[\citeproctext]{ref-llorca-jana_historia_2021}
+Llorca-Jaña, M., \& Miller, R. M. D. (2021). \emph{{Historia econ{ó}mica
+de Chile desde la independencia}}. Santiago de Chile: RIL editores.
+
 \bibitem[\citeproctext]{ref-lopez-roldan_comparative_2021}
@@
 
+\bibitem[\citeproctext]{ref-mac-clure_justicia_2024}
+Mac-Clure, O., Barozet, E., \& Franetovic, G. (2024). {Justicia
+distributiva y posici{ó}n social subjetiva: {\textquestiondown}la
+meritocracia justifica la desigualdad de ingresos?} \emph{Convergencia
+Revista de Ciencias Sociales}, \emph{31}, 1.
+\url{https://doi.org/10.29101/crcs.v31i0.22258}
+
 \bibitem[\citeproctext]{ref-madariaga_three_2020}
@@
 
+\bibitem[\citeproctext]{ref-mideplan_evolucion_2017}
+MIDEPLAN. (2017). \emph{Evoluci{ó}n de la pobreza 1990-2017}. Ministerio
+de Desarrollo Social y Familia - PNUD.
+
+\bibitem[\citeproctext]{ref-mideplan_estadisticas_2024}
+MIDEPLAN. (2024). \emph{Estad{í}sticas {Encuesta Casen} 2022, secci{ó}n
+{Salud}}. Observatorio social.
+
 \bibitem[\citeproctext]{ref-mijs_paradox_2019}
@@
 
+\bibitem[\citeproctext]{ref-ministeriodeeducacion_resumen_2023}
+Ministerio de Educación. (2023). \emph{Resumen de matr{í}cula por
+establecimiento educacional}. Datos abiertos Ministerio de Educaci{ó}n.
+
 \bibitem[\citeproctext]{ref-molina_its_2019}
@@
 game is played, it's whether you win or lose. \emph{SCIENCE ADVANCES}.
+
+\bibitem[\citeproctext]{ref-nucleodesociologiacontingente_informe_2020}
+Núcleo de Sociología Contingente, {[}NUDESOC{]}. (2020). \emph{Informe
+de resultados oficial {Encuesta Zona Cero}}. Santiago de Chile.
 
@@
 
+\bibitem[\citeproctext]{ref-otero_lives_2022}
+Otero, G., Volker, B., Rözer, J., \& Mollenhorst, G. (2022). The lives
+of others: {Class} divisions, network segregation, and attachment to
+society in {Chile}. \emph{The British Journal of Sociology},
+\emph{73}(4), 754--785. \url{https://doi.org/10.1111/1468-4446.12966}
+
+\bibitem[\citeproctext]{ref-panes_criticas_2020}
+Panes, D. (2020). \emph{Cr{í}ticas y experiencias obreras en torno al
+sistema de {AFP}'s en {Chile} (1981-2020)} (PhD thesis). Universidad de
+Chile, Santiago.
+
+\bibitem[\citeproctext]{ref-perez-ahumada_clases_2019}
+Pérez-Ahumada, P. (2019). {Clases sociales, sectores econ{ó}micos y
+cambios en la estructura social chilena entre 1992 y 2013}.
+\emph{Revista de la CEPAL}, \emph{2018}(126), 171--192.
+\url{https://doi.org/10.18356/0ce0faed-es}
+
+\bibitem[\citeproctext]{ref-pnud_desiguales_2017}
+PNUD (Ed.). (2017). \emph{{Desiguales: or{í}genes, cambios y desaf{í}os
+de la brecha social en Chile}}. Santiago, Chile: PNUD : Uqbar Editores.
+
 \bibitem[\citeproctext]{ref-polanyi_great_1975}
@@
 \url{https://doi.org/10.1080/14616696.2021.1887916}
+
+\bibitem[\citeproctext]{ref-ruiz_chilenos_2014}
+Ruiz, C., \& Boccardo, G. (2014). \emph{Los chilenos bajo el
+neoliberalismo: Clases y conflicto social} (Primera edici{ó}n). Santiago
+de Chile: Ediciones y Publicaciones El Buen Aire.
 
@@
 system}. London: Verso.
+
+\bibitem[\citeproctext]{ref-superintendenciadepensiones_estadisticas_2024}
+Superintendencia de Pensiones. (2024). \emph{Estad{í}sticas de
+{Afiliados} en {AFP} en {Sistema} de {Pensiones}. {Abril} 2024}.
+Superintendencia de Pensiones.
 
@@
 
+\bibitem[\citeproctext]{ref-torche_unequal_2005}
+Torche, F. (2005). Unequal {But Fluid}: {Social Mobility} in {Chile} in
+{Comparative Perspective}. \emph{American Sociological Review},
+\emph{70}(3), 422--450. \url{https://doi.org/10.1177/000312240507000304}
+
 \bibitem[\citeproctext]{ref-torche_intergenerational_2014}

### 2025-06-27 processing/01-proc-data.R

@@ -26,8 +26,9 @@ rm(list = ls())
 
 # 2. Data -----------------------------------------------------------------
 
-load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+#load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
 load(here("input/data/elsoc_ciuo08.RData"))
+load(here("input/data/ELSOC_Long_2016_2023_1.00.RData"))
 
 glimpse(elsoc_long_2016_2023)
 
@@ -66,13 +67,13 @@ elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>%
   )
 
 elsoc_long_2016_2023$just_health <- sjlabelled::set_label(elsoc_long_2016_2023$just_health,
-                                        label = "Health distributive justice")
+                                                          label = "Health distributive justice")
 
 elsoc_long_2016_2023$just_pension <- sjlabelled::set_label(elsoc_long_2016_2023$just_pension,
-                                         label = "Pension distributive justice")
+                                                           label = "Pension distributive justice")
 
 elsoc_long_2016_2023$just_educ <- sjlabelled::set_label(elsoc_long_2016_2023$just_educ,
-                                      label = "Education distributive justice")
+                                                        label = "Education distributive justice")
 
 # Meritocracy
 
@@ -93,10 +94,10 @@ elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>%
 
 
 elsoc_long_2016_2023$merit_effort <- sjlabelled::set_label(elsoc_long_2016_2023$merit_effort, 
-                                         label = "People are rewarded for their efforts")
+                                                           label = "People are rewarded for their efforts")
 
 elsoc_long_2016_2023$merit_talent <- sjlabelled::set_label(elsoc_long_2016_2023$merit_talent, 
[…]

### 2025-06-27 processing/02-proc-data.R

@@ -1,0 +1,519 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               naniar,
+               data.table,
+               summarytools,
+               nnet,
+               occupar)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
+      .cols = c(just_health, just_pension, just_educ),
[…]

### 2025-06-27 processing/03-proc.R

@@ -1,0 +1,700 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               naniar,
+               data.table,
+               summarytools,
+               nnet,
+               occupar,
+               PSweight)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
[…]

### 2025-06-27 processing/04-proc.R

@@ -1,0 +1,515 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               naniar,
+               data.table,
+               summarytools,
+               nnet,
+               occupar)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
+      .cols = c(just_health, just_pension, just_educ),
[…]

### 2025-09-01 presentations/coloquio_2025/borrador_mjp-mobility-coloquio.qmd
Up resultados V1

---
title: "Preferencias por la comodificación: el papel de la movilidad social intergeneracional y las creencias en la meritocracia en Chile"
format: clean-revealjs
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author: 
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - Instituto de Sociología, Pontificia Universidad Católica de Chile
date: "July 3, 2025"
logo: "images/logo_isuc.png"
---

## Example slide

### This is a subtitle

Here we have some text that may run over several lines of the slide frame,
depending on how long it is.

- first item 
    - A sub item

Next, we'll brief review some theme-specific components.

- Note that _all_ of the standard Reveal.js
[features](https://quarto.org/docs/presentations/revealjs/)
can be used with this theme, even if we don't highlight them here.

## Additional theme classes

### Some extra things you can do with the clean theme

Special classes for emphasis

- `.alert` class for default emphasis, e.g. [important note]{.alert}.
- `.fg` class for custom colour, e.g. [important note]{.fg style="--col: #e64173"}.
- `.bg` class for custom background, e.g. [important note]{.bg style="--col: #e64173"}.

Cross-references

- `.button` class provides a Beamer-like button, e.g.
[[Summary]{.button}](#sec-summary)

## Want more?

### See our longer demo slides

We've deliberarely kept this template lean, to get you up and running as fast as
possible.

We provide a separate
[demo template](https://github.com/grantmcdermott/quarto-revealjs-clean-demo),
with more examples for integrating code, tables, figures, etc.

  - See the live demo slides
  [here](https://grantmcdermott.com/quarto-revealjs-clean-demo/template.html).

## Summary {#sec-summary}

### A minimalist and elegant presentation theme

The Quarto reveal.js clean theme aims to be a minimalist and elegant presention
theme. Here are some options to get you started:

Add the theme to an existing project.

```{.bash}
quarto install extension grantmcdermott/quarto-revealjs-clean
```

... or, create a new project using this slide deck as a lean template.

```{.bash}
quarto use template grantmcdermott/quarto-revealjs-clean
```

... or, create a new project using the demo slide deck as a full template.

```{.bash}
quarto use template grantmcdermott/quarto-revealjs-clean-demo
```

### 2025-09-01 presentations/coloquio_2025/mjp-mobility.qmd
Up resultados V1

---
bibliography: "input/bib/tesis-magister.bib"
csl: "input/bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
  pdf:
    format: beamer 
    keep-tex: true  # Debug LaTeX output if needed
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(knitr)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

#pacman::p_load(tidyverse, 
#               sjmisc, 
#               sjPlot, 
##               sjlabelled, 
#               here,
#               kableExtra,
#               ggdist,
#               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data
#| include: false

```

::: columns

::: {.column width="20%"}
</br>

![](images/logo_isuc.png)

</br></br>

![](images/jusmer_trans.png)

:::

::: {.column .column-right width="80%"}

# **Preferencias por la comodificación**

## **El rol de la movilidad social intergeneracional y las creencias meritocráticas en Chile**

------------------------------------------------------------------------

**Andreas Laffert^1,^^2^**

::: {.blue2 .medium}

**^1^Instituto de Sociología, Pontificia Universidad Católica de Chile**

**^2^Centro de Estudios de Conflicto y Cohesión Social - COES**

:::

Coloquio de investigación en justicia distributiva y desigualdad socioeconómica

4 Septiembre 2025, Santiago

:::
:::

# Contexto y motivación {.xlarge data-background-color="#523870"}

## Antecedentes

::: {.incremental .highlight-last style="font-size: 100%;"}

- The justification of social inequality based on market-type criteria has been conceptualized as the individuals' adherence to the deservingness of social goods and services (such as health, education, and pensions) based on prices and individuals' ability to pay [@boltanski_new_2005; @lane_market_1986; @streeck_citizens_2012].

- Research on social stratification beliefs [@kluegel_beliefs_1987], highlights that people's explanations and justifications of social inequality are closely tied to their judgments of deservingness.

-  The expansion of marketization has been related to a larger justification of market mechanisms, whereby societies with larger private spending on services show larger market justice preferences [@lindh_public_2015].

:::

## Antecedentes

::::: columns
::: {.column width="50%" .content-box-red}

#### A

- The original definition of merit is a combination of effort and talent [@young_rise_1958]
- 2
- 3

:::

::: {.column width="50%" .content-box-yellow}

#### B

- Adolescents from low-socioeconomic-status families are likelier to have an egalitarian fairness view and consider an equal distribution as fair in a situation with unequal merits [@almas_fairness_2017]
- 2
- 3

:::
:::::

## Hipótesis

::: {.incremental .highlight-last style="font-size: 100%;"}

$H_{1}$

$H_{2}$

$H_{3}$

$H_{n}$

:::

# Datos, variables y métodos {.xlarge data-background-color="#523870"}

## Datos

::: {.incremental .highlight-last style="font-size: 100%;"}
-   Datos
:::

## Variables

:::

## Métodos

- Métodos

# Resultados {.xlarge data-background-color="#523870"}

## Modelos

# Discusión y conclusiones {data-background-color="#523870"}

# Discusión y conclusiones {data-background-color="#523870"}

::: { .incremental .highlight-last2 style="font-size: 100%;"}

- Discusión

  - Conclusión
:::

# Gracias por su atención!

-   **Github del proyecto:** <https://github.com/Andreas-Lafferte/mobility-market-justice>

## Referencias

### 2025-09-01 processing/03-proc.R

@@ -1,700 +1,0 @@
-# 0. Identification ---------------------------------------------------
-
-# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
-# Responsible: Andreas Laffert
-
-# Executive Summary: This script contains the code to data preparation for thesis project
-# Date: May 1, 2025
-
-# 1. Packages  -----------------------------------------------------
-
-if (!require("pacman")) install.packages("pacman")
-
-pacman::p_load(tidyverse,
-               here,
-               car,
-               sjmisc,
-               sjlabelled,
-               naniar,
-               data.table,
-               summarytools,
-               nnet,
-               occupar,
-               PSweight)
-
-options(scipen=999)
-rm(list = ls())
-
-# 2. Data -----------------------------------------------------------------
-
-load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
-load(here("input/data/elsoc_ciuo08.RData"))
-
-glimpse(elsoc_long_2016_2023)
-
-# 3. Processing -----------------------------------------------------------
-
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
-
-elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
-  mutate(just_educ = d02_02, 
-         just_pension = d02_01, 
-         just_health = d02_03, 
-         merit_effort = c18_09,
-         merit_talent = c18_10) %>% 
-  as_tibble() %>% 
-  sjlabelled::drop_labels(., drop.na = FALSE)
-
-# Market Justice Preferences
-
-frq(elsoc_long_2016_2023$just_health)
-frq(elsoc_long_2016_2023$just_pension)
-frq(elsoc_long_2016_2023$just_educ)
-
-elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
-  mutate(
-    across(
[…]

### 2025-09-01 processing/04-proc.R

@@ -400,10 +400,11 @@ library(estimatr)
 ### downward
 
 datos1 %>% 
-  select(idencuesta, ola, just_pension, downward_mob, wdownward_mob1) %>% 
+  select(idencuesta, ola, just_pension, downward_mob, wdownward_mob1, merit_effort) %>% 
   na.omit() %>% 
-  mutate(just_pension = as.numeric(just_pension)) %>% 
-  lm_robust(just_pension ~ downward_mob,
+  mutate(just_pension = as.numeric(just_pension),
+         merit_effort = as.numeric(merit_effort)) %>% 
+  lm_robust(just_pension ~ downward_mob*merit_effort,
             weights = wdownward_mob1,
             se_type = "CR2",
             clusters = idencuesta,
@@ -412,10 +413,11 @@ datos1 %>%
 ### inmobile
 
 datos2 %>% 
-  select(idencuesta, ola, just_pension, inmobile_mob, winmobile1) %>% 
+  select(idencuesta, ola, just_pension, inmobile_mob, winmobile1, merit_effort) %>% 
   na.omit() %>% 
-  mutate(just_pension = as.numeric(just_pension)) %>% 
-  lm_robust(just_pension ~ inmobile_mob,
+  mutate(just_pension = as.numeric(just_pension),
+         merit_effort = as.numeric(merit_effort)) %>% 
+  lm_robust(just_pension ~ inmobile_mob*merit_effort,
             weights = winmobile1,
             se_type = "CR2",
             clusters = idencuesta,
@@ -424,10 +426,11 @@ datos2 %>%
 ### upward
 
 datos3 %>% 
-  select(idencuesta, ola, just_pension, upward_mob, wpupward1) %>% 
+  select(idencuesta, ola, just_pension, upward_mob, wpupward1, merit_effort) %>% 
   na.omit() %>% 
-  mutate(just_pension = as.numeric(just_pension)) %>% 
-  lm_robust(just_pension ~ upward_mob,
+  mutate(just_pension = as.numeric(just_pension),
+         merit_effort = as.numeric(merit_effort)) %>% 
+  lm_robust(just_pension ~ upward_mob*merit_effort,
             weights = wpupward1,
             se_type = "CR2",
             clusters = idencuesta,
@@ -439,10 +442,11 @@ datos3 %>%
 ### downward
 
 datos1 %>% 
-  select(idencuesta, ola, just_health, downward_mob, wdownward_mob1) %>% 
[…]

### 2025-09-01 processing/05-long.R

@@ -1,0 +1,469 @@
+# Libraries
+if (! require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               car,
+               sjmisc, 
+               here,
+               sjlabelled,
+               SciViews,
+               naniar)
+
+
+options(scipen=999)
+rm(list = ls())
+
+# Data
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+
+glimpse(elsoc_long_2016_2023)
+
+
+# Processing
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10, 
+         age = m0_edad, m01, 
+         sex = m0_sexo, 
+         ess_d = d01_01, 
+         ess_o = d01_02,
+         ideo = c15,
+         ) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- cbind(elsoc_long_2016_2023, "mjp" = rowMeans(elsoc_long_2016_2023 %>% select(just_health, just_pension, just_educ), na.rm=TRUE))
+
+summary(elsoc_long_2016_2023$mjp)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
+      .cols = c(just_health, just_pension, just_educ),
+      .fns = ~ car::recode(., recodes = c("1='Strongly disagree'; 2='Disagree';
+                                          3='Neither agree nor disagree'; 4='Agree';
+                                          5='Strongly agree'"), 
[…]

### 2025-09-01 processing/06-long.R

@@ -1,0 +1,108 @@
+# Libraries
+if (! require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse, 
+               sjmisc, 
+               sjPlot, 
+               lme4, 
+               here, 
+               performance,
+               texreg, 
+               shadowtext,
+               MetBrewer,
+               sjlabelled,
+               plm)
+
+
+options(scipen=999)
+rm(list = ls())
+
+# Data
+
+load(file = here("input/data/proc/df_study1_long_t7.RData"))
+
+# Generate analytical sample
+
+df_study1 <- df_study1_long_t7 %>%
+  select(idencuesta,
+         ola,
+         ponderador_long_total,
+         segmento,
+         estrato,
+         just_health,
+         just_pension,
+         just_educ,
+         mjp, 
+         mobility,
+         mobility_f,
+         merit_effort, 
+         merit_talent, 
+         educ,
+         quintil1,
+         sex,
+         age,
+         ideo) %>% 
+  na.omit() %>% 
+  mutate(ola = case_when(ola == 1 ~ 1,
+                         ola == 2 ~ 2, 
+                         ola == 3 ~ 3,
+                         ola == 4 ~ 4,
+                         ola == 6 ~ 5,
+                         ola == 7 ~ 6)) %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2)
+
+df_study1 <- df_study1 %>%
+  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
+  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
+  ungroup()
+
+df_study1 <- df_study1 %>% filter(n_participaciones>1)
+
+
+# Analysis
+
+df_study1$just_health <- as_numeric(df_study1$just_health)
+df_study1$just_pension <- as_numeric(df_study1$just_pension)
+df_study1$just_educ <- as_numeric(df_study1$just_educ)
+
+
+frq(df_study1$mobility_f)
+
+plm(formula = mjp ~  mobility_f*merit_effort, 
+    data = df_study1,
+    model = "within",
+    effect = "twoways") %>% summary()
+
+
+df_study1$merit_effort <- as_numeric(df_study1$merit_effort)
+df_study1$merit_talent <- as_numeric(df_study1$merit_talent)
+
+df_study1 <- df_study1 %>% 
[…]

### 2025-09-01 processing/prod_analysis.R

@@ -1,0 +1,613 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data analysis for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               sjmisc,
+               PSweight,
+               texreg,
+               estimatr,
+               mediation,
+               conflicted)
+
+conflict_prefer("select", "dplyr")
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(here("input/data/proc/data_v1.RData"))
+
+# 3. Analysis -----------------------------------------------------------
+
+# 3.1 Regresiones con IPW -----------------------------------------------------------
+
+### Pensiones
+
+### Low to middle
+
+data <- datos1 %>% 
+  select(idencuesta, 
+         ola, 
+         y = just_pension, 
+         t = moblm, 
+         w = wlow_orig1, 
+         edad = m0_edad, 
+         sexo, 
+         nac = nacionalidad,
+         etnia,
+         hogar = hogar_bip_f, 
+         edu0 = educ_orig_f) %>% 
+  na.omit() %>% 
+  mutate(w = w / mean(w)) 
+
+data
+
+data %>% 
+  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
+            weights = w,
+            se_type = "CR2",
+            clusters = idencuesta,
+            data = .) %>% 
+  screenreg()
+
+### Low to high
+
+data <- datos1 %>% 
+  select(idencuesta, 
+         ola, 
+         y = just_pension, 
+         t = moblh, 
+         w = wlow_orig2, 
+         edad = m0_edad, 
+         sexo, 
+         nac = nacionalidad,
+         etnia,
+         hogar = hogar_bip_f, 
+         edu0 = educ_orig_f) %>% 
+  na.omit() %>% 
+  mutate(w = w / mean(w)) 
+
[…]

### 2025-09-01 processing/prod_prep.R

@@ -1,0 +1,508 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               data.table,
+               summarytools,
+               nnet,
+               occupar,
+               PSweight)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
[…]

### 2025-09-01 processing/strati.R

@@ -1,0 +1,960 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               naniar,
+               data.table,
+               summarytools,
+               nnet,
+               occupar)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(merit_effort = c18_09,
+         merit_talent = c18_10,
+         perc_inequality2 = c18_11,
+         perc_sal_gerente = d03_01,
+         perc_sal_obrero = d03_02, 
+         just_sal_gerente = d04_01, 
+         just_sal_obrero = d04_02, 
+         age = m0_edad, 
+         m01, 
+         sex = m0_sexo, 
+         ess = d01_01, 
+         ideo = c15
+         ) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Percep inequality
+
+frq(elsoc_long_2016_2023$perc_inequality2) # ok
+
[…]

### 2025-09-01 presentations/coloquio_2025/estrati_ppt.qmd
Up presentación estrati

---
bibliography: "input/bib/tesis-magister.bib"
csl: "input/bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
  pdf:
    format: beamer 
    keep-tex: true  # Debug LaTeX output if needed
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(knitr)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

#pacman::p_load(tidyverse, 
#               sjmisc, 
#               sjPlot, 
##               sjlabelled, 
#               here,
#               kableExtra,
#               ggdist,
#               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data
#| include: false

```

::: columns

::: {.column width="20%"}

![](images/logo_isuc.png)

</br></br>

![](images/coes.png)

:::

::: {.column .column-right width="80%"}

# **La justificación de la desigualdad salarial**

## **El rol de la movilidad social intergeneracional en Chile**

------------------------------------------------------------------------

**Andreas Laffert^1,^^2^**

::: {.blue2 .medium}

**^1^Instituto de Sociología, Pontificia Universidad Católica de Chile**

**^2^Centro de Estudios de Conflicto y Cohesión Social - COES**

:::

Seminario Estratificación Social - SOL3016

3 Septiembre 2025, Santiago

:::
:::

# Contexto y motivación {.xlarge data-background-color="#523870"}

## Contexto y motivación: Mobility effects

::::: columns
::: {.column width="50%"}
::: {.fragment}

![](images/breen.png)

:::

:::

::: {.column width="50%" .incremental .highlight-last style="font-size: 100%;"}

- Los efectos de la movilidad —definidos como los resultados derivados del desplazamiento entre las clases de origen y destino [@helgason_class_2025]— han suscitado desde hace tiempo el interés sociológico
- @breen_effects_2024 proponen un método para estimar el efecto (causal) de la movilidad social intergernacional sobre un outcome
- Las investigaciones sobre la movilidad social intergeneracional han examinado cada vez más sus efectos en las actitudes hacia la desigualdad

:::
:::::

## Contexto y motivación: legitimidad de la desigualdad

::: {.incremental .highlight-last style="font-size: 110%;"}

- Desde los 80's, la desigualdad económica y la concentración de la riqueza han aumentado en el mundo [@chancel_world_2022].

- **¿En qué medida la desigualdad económica es vista como justa por las personas?**

- La investigación empírica sobre la justicia distributiva estudia las concepciones de los individuos sobre cómo _debe_ ser la distribución [@jasso_distributive_2016]

- Esta línea de investigación permite examinar hasta qué punto la desigualdad económica se percibe como justa o, en cierto sentido, legítima [@castillo_legitimacy_2011] --> legitimidad de la brecha salarial

:::

## Antecedentes

::::: columns 
::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}

### Brecha salarial

- Niveles de desigualdad económica y marcos institucionales [@castillo_legitimacy_2011; @hadler_why_2005] 
- Estatus socioeconómico objetivo [@jasso_assessing_1999; @kelley_legitimate_2021]
- Las percepciones de desigualdad e ideologías [@willis_legitimacy_2015]

:::

::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}

### Movilidad social

- Preferencias redistributivas [@alesina_intergenerational_2018; @ares_changing_2020; @schmidt_experience_2011; @breen_effects_2024]
- Legitimidad de la desigualdad económica [@gugushvili_intergenerational_2016], gasto público [@gugushvili_subjective_2017] y creencias sobre la desigualdad [@bucca_merit_2016; @day_movin_2017]
- Diversos **mecanismos explicativos**: self-interest, socialización y **atribuciones** [@helgason_class_2025]

:::
:::::

## Contexto chileno

::: {.incremental .highlight-last style="font-size: 110%;"}

- Chile presenta uno de los niveles de desigualdad más altos de América Latina y de los países de la OCDE [@flores_top_2020]

- Movilidad ascendente de corto alcance entre clases bajas y medias adyacentes, aunque siguen existiendo fuertes barreras para alcanzar posiciones más altas [@torche_intergenerational_2014; @espinoza_estratificacion_2013]

-  @castillo_legitimacy_2011 estudia la legitimidad de la desigualdad económica en Chile mediante la brecha salarial justa. Sus hallazgos revelan que

      - el estatus socioeconómico (ingresos y educación) no influyen en la brecha salarial justa
      - la percepción de desigualdad o valores igualitarios/individualistas sí influyen

:::

## Problematización

::: {.incremental .highlight-last style="font-size: 110%;"}

- Aunque está claro que la posición social influye, la cuestión de cómo la movilidad ascendente o descendente dentro de la estructura social afecta a la justificación de la desigualdad salarial sigue sin respuesta

- En Chile, hay pocos estudios sobre la justificación de la desigualdad salarial y aún menos sobre los _efectos_ de la movilidad en las actitudes hacia la desigualdad

:::

# Este estudio {.xlarge data-background-color="#523870"}

:::: {.incremental style="font-size: 150%;"}

_**¿En qué medida la movilidad social intergeneracional afecta la justificación de la desigualdad salarial en Chile?**_

::::

## Hipótesis

::: {.incremental .highlight-last style="font-size: 150%;"}

- $H_{1}$: Experimentar movilidad social ascendente (descendente) se asocia positiva (negativamente) con una mayor justificación de la desigualdad salarial.

- $H_{2}$: Las creencias meritocráticas son un mecanismo a través del cual opera el sesgo egoísta y conduce a diversos grados de justificación de la desigualdad salarial

:::

# Datos, variables y métodos {.xlarge data-background-color="#523870"}

## Datos

::: {.incremental .highlight-last style="font-size: 100%;"}
-   Datos
:::

## Variables

:::

## Métodos

- Métodos

# Gracias por su atención!

-   **Github del proyecto:** <https://github.com/Andreas-Lafferte/mobility-market-justice>

## Referencias

### 2025-09-02 presentations/coloquio_2025/estrati_ppt.qmd
Up ppt estrati VF

@@
 
-# **La justificación de la desigualdad salarial**
-
-## **El rol de la movilidad social intergeneracional en Chile**
+# **La brecha salarial justa: el rol de la movilidad social intergeneracional en Chile**
 
@@
 
-::: {.incremental .highlight-last style="font-size: 110%;"}
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
@@
 
-- Esta línea de investigación permite examinar hasta qué punto la desigualdad económica se percibe como justa o, en cierto sentido, legítima [@castillo_legitimacy_2011] --> legitimidad de la brecha salarial
+- Esta línea de investigación permite examinar hasta qué punto la desigualdad económica se percibe como justa y, en cierto sentido, legítima [@castillo_legitimacy_2011] --> legitimidad de la brecha salarial
 
@@
 
-### Brecha salarial
+### Brecha salarial justa
 
@@
 - Estatus socioeconómico objetivo [@jasso_assessing_1999; @kelley_legitimate_2021]
-- Las percepciones de desigualdad e ideologías [@willis_legitimacy_2015]
+- Percepciones sobre la desigualdad e ideologías [@willis_legitimacy_2015]
 
@@
 - Preferencias redistributivas [@alesina_intergenerational_2018; @ares_changing_2020; @schmidt_experience_2011; @breen_effects_2024]
-- Legitimidad de la desigualdad económica [@gugushvili_intergenerational_2016], gasto público [@gugushvili_subjective_2017] y creencias sobre la desigualdad [@bucca_merit_2016; @day_movin_2017]
+- Legitimidad de la desigualdad económica [@gugushvili_intergenerational_2016], política social [@gugushvili_subjective_2017] y creencias sobre la desigualdad [@bucca_merit_2016; @day_movin_2017]
 - Diversos **mecanismos explicativos**: self-interest, socialización y **atribuciones** [@helgason_class_2025]
@@
 
-- Movilidad ascendente de corto alcance entre clases bajas y medias adyacentes, aunque siguen existiendo fuertes barreras para alcanzar posiciones más altas [@torche_intergenerational_2014; @espinoza_estratificacion_2013]
+- Movilidad ascendente de corto alcance entre clases adyacentes, aunque siguen existiendo fuertes barreras para alcanzar posiciones más altas [@torche_intergenerational_2014; @lopez-roldan_comparative_2021]
 
@@
       - el estatus socioeconómico (ingresos y educación) no influyen en la brecha salarial justa
-      - la percepción de desigualdad o valores igualitarios/individualistas sí influyen
+      - la percepción de desigualdad y valores igualitarios/individualistas sí influyen
 
@@
 
-::: {.incremental .highlight-last style="font-size: 110%;"}
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
-- Aunque está claro que la posición social influye, la cuestión de cómo la movilidad ascendente o descendente dentro de la estructura social afecta a la justificación de la desigualdad salarial sigue sin respuesta
+- Aunque está claro que la posición social influye, la cuestión de cómo la movilidad ascendente o descendente dentro de la estructura social afecta la brecha salarial justa sigue sin respuesta
 
-- En Chile, hay pocos estudios sobre la justificación de la desigualdad salarial y aún menos sobre los _efectos_ de la movilidad en las actitudes hacia la desigualdad
+- En Chile, hay pocos estudios sobre la justificación de la desigualdad salarial y aún menos sobre los _efectos_ de la movilidad en las actitudes hacia la desigualdad económica
 
@@
 
-## Hipótesis
+# Hipótesis {data-background-color="#523870"}
 
-::: {.incremental .highlight-last style="font-size: 150%;"}
+:::: {.incremental style="font-size: 140%;"}
 
-- $H_{1}$: Experimentar movilidad social ascendente (descendente) se asocia positiva (negativamente) con una mayor justificación de la desigualdad salarial.
+Experimentar movilidad social ascendente (descendente) se asocia positivamente (negativamente) con una mayor justificación de la desigualdad salarial.
 
-- $H_{2}$: Las creencias meritocráticas son un mecanismo a través del cual opera el sesgo egoísta y conduce a diversos grados de justificación de la desigualdad salarial
+::::
+
+# Diseño metodológico {.xlarge data-background-color="#523870"}
+
+## Datos
+
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+- Encuesta Longitudinal Social de Chile (ELSOC) de COES.
+
+- Muestra analítica: olas 2016 (N = 2.927), 2018 (N = 3.748) y 2023 (N = 2.726)
+
+- Encuesta panel representativa de zonas urbanas, localizadas en 40 ciudades del país (92 comunas y 13 regiones)
+
+- Diseño muestral complejo: probabilístico, por conglomerados, multietápico y estratificado según el tamaño de las ciudades.
+
+- Población objetivo incluye a mujeres y hombres de entre 18 y 75 años que residen habitualmente en viviendas privadas
 
@@
 
-# Datos, variables y métodos {.xlarge data-background-color="#523870"}
+## Outcome
 
-## Datos
+::: {.incremental style="font-size: 120%;"}
 
-::: {.incremental .highlight-last style="font-size: 100%;"}
--   Datos
+- **Brecha salarial justa**: enfoque para medir la justificación de la desigualdad económica basado en las diferencias salariales consideradas justas entre las ocupaciones de alto y bajo estatus [@castillo_legitimacy_2011].
+Los valores más altos de este término indican una mayor justificación de la desigualdad económica (@eq-just).
+
+$$
+\text{just wage gap} = \log_{10}\left(\frac{\text{just salary of a large-company president}}{\text{just salary of an unskilled worker}}\right)
+$$ {#eq-just}
 :::
 
-## Variables
+## Tratamiento
+
+::: {style="font-size: 120%;"}
+Siguiendo la propuesta de @breen_effects_2024 para estimar el efecto causal de la movilidad social: 
+:::
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+1.
+**Asignación de clase**: En base a la ocupación (ISCO-08) del padre/madre y del entrevistado se crean estratos de origen y destino (bajo, medio y alto) según terciles de la distribución del índice de estatus ocupacional (ISEI)
+
+2.
+**Propensity-score estimation**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles para cada estrato (categ.
+ref.).
+Estas puntuaciones de propensión se emplean posteriormente para ajustar la selección en la movilidad al evaluar su efecto.
 
@@
 
+## Tratamiento
+
+![](images/propensity.png)
+
 ## Métodos
 
-- Métodos
+::: {.incremental style="font-size: 120%;"}
+
+- Para estimar el efecto causal de la movilidad social, utilizo modelos de regresión lineal con ponderaciones de probabilidad inversa (IPW) para la movilidad condicional a la clase de origen.
+
+- Permite estimar el efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran trasladado a un destino alternativo *k′* [@breen_effects_2024].
+El estimador resultante es el efecto promedio del tratamiento sobre los tratados (ATT) (@eq-att).
+
+$$ 
+ATT_{jkk′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
+$$ {#eq-att}
+
+:::
+
+# Mecanismos {.xlarge data-background-color="#523870"}
+
+## Mecanismos... por revisar
+
+::: {.incremental style="font-size: 120%;"}
+
+- ¿Cómo y por qué un movimiento objetivo en la estructura social _afectaría_ o _produciría_ determinada actitud subjetiva?
+
+- **Self-serving bias**: 
+    - Las experiencias de movilidad social de las personas determinan sus atribuciones causales, que a su vez influyen en sus creencias sobre la justicia y sus preferencias distributivas [@gugushvili_trends_2014].
+    - Atribuciones **internas** (esfuerzo, talento, mérito) vs. **externas** (barreras estructurales).
+
+- Estrategia: ACME (?) [@imai_identification_2010]
+
+:::
 

### 2025-09-03 presentations/coloquio_2025/mjp-mobility.qmd
Up varios

@@
 
-#pacman::p_load(tidyverse, 
-#               sjmisc, 
-#               sjPlot, 
-##               sjlabelled, 
-#               here,
-#               kableExtra,
-#               ggdist,
-#               summarytools)
+pacman::p_load(tidyverse, 
+               sjmisc, 
+               sjPlot, 
+               sjlabelled, 
+               here,
+               kableExtra,
+               ggdist,
+               summarytools)
 
@@
 #| include: false
+
+source(file = here("processing/prod_analysis.R"))
 
@@
 
-::: {.blue2 .medium}
+::: {.medium}
 
@@
 
-Coloquio de investigación en justicia distributiva y desigualdad socioeconómica
+Coloquio de Investigación en Justicia Distributiva y Desigualdad Socioeconómica
 
@@
 
+---
+
+## ¿A dónde quiero llegar?
+
+::: {.fragment}
+
+<div style="text-align:center;">
+  <img src="images/idea.png" style="width:100%; max-width:none;">
+</div>
+
+:::
+
 # Contexto y motivación {.xlarge data-background-color="#523870"}
 
-## Antecedentes
+## Contexto y motivación
 
-::: {.incremental .highlight-last style="font-size: 100%;"}
-
-- The justification of social inequality based on market-type criteria has been conceptualized as the individuals' adherence to the deservingness of social goods and services (such as health, education, and pensions) based on prices and individuals' ability to pay [@boltanski_new_2005; @lane_market_1986; @streeck_citizens_2012].
-
-- Research on social stratification beliefs [@kluegel_beliefs_1987], highlights that people's explanations and justifications of social inequality are closely tied to their judgments of deservingness.
-
--  The expansion of marketization has been related to a larger justification of market mechanisms, whereby societies with larger private spending on services show larger market justice preferences [@lindh_public_2015].
-
+::: {.fragment}
+<div style="text-align:center;">
+  <img src="images/corte.png" style="width:100%; max-width:none;">
+</div>
 :::
@@
 
+</br>
+
+::: {.box-inv-4 .sp-after .fragment style="font-size: 150%;"}
+
+Mercantilización y privatización (M, AL, CL)
+
+:::
+
+::: {.box-inv-4 .sp-after .fragment style="font-size: 150%;"}
+
+Economía moral (Mau, bringing market, policy feedback)
+:::
+
+::: {.box-inv-4 .sp-after .fragment style="font-size: 150%;"}
+
+MJP (Lane, Lindh, implicancias y relevancias)
+:::
+
+## Preferencias por justicia de mercado
+
 ::::: columns
-::: {.column width="50%" .content-box-red}
+::: {.column width="50%"}
 
-#### A
+### Contextual
 
@@
 
-::: {.column width="50%" .content-box-yellow}
+::: {.column width="50%"}
 
-#### B
+### Individual
 
@@
 
-## Hipótesis
+## Preferencias por justicia de mercado: Chile
 
@@
 
-$H_{1}$
-
-$H_{2}$
-
-$H_{3}$
-
-$H_{n}$
+- Otero 2024 (clase)
+- Castillo y laffert 2025 (merit)
 
@@
 
-# Datos, variables y métodos {.xlarge data-background-color="#523870"}
+## Este estudio
+
+::: {.incremental .highlight-last style="font-size: 100%;"}
+
+- Aunque esta claro... movilidad social
+- Efectos
+- Mecanismos
+
+:::
+
+## Este estudio
+
+::: {.incremental .highlight-last style="font-size: 100%;"}
+
+- Mecanismos
+
+:::
+
+## Contexto chileno
+
+::: {.incremental .highlight-last style="font-size: 100%;"}
+
+- RRelevancia del caso Chileno: extrema comodificacion, desigualdad, limitada movilidad, alta pferencia por la meritocracia etc.
+
+:::
+
+#  {data-background-color="#523870"}
+
+:::: {style="font-size: 170%;"}
+
+**_(1) ¿En qué medida la movilidad social intergeneracional afecta las preferencias por comodificación en salud, pensiones y educación en Chile?_**
+
+**_(2) ¿Cómo las creencias meritocráticas median esta relación en el contexto chileno?_**
+
+::::
+
+## Hipótesis
+
+::: {.fragment}
+
+<div style="text-align:center;">
+  <img src="images/idea.png" style="width:100%; max-width:none;">
+</div>
+
+:::
+
+# Método {.xlarge data-background-color="#523870"}
+
+## Estrategia de identificación causal
 
@@
 
-::: {.incremental .highlight-last style="font-size: 100%;"}
--   Datos
-:::
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
-## Variables
+- Encuesta Longitudinal Social de Chile [(ELSOC)](https://coes.cl/elsoc/) de COES.
+
+- Muestra analítica: olas 2016 (N = 2.927), 2018 (N = 3.748) y 2023 (N = 2.726)
+
+- Encuesta panel representativa de zonas urbanas, localizadas en 40 ciudades del país (92 comunas y 13 regiones)
+
+- Diseño muestral complejo: probabilístico, por conglomerados, multietápico y estratificado según el tamaño de las ciudades.
+
+- Población objetivo incluye a mujeres y hombres de entre 18 y 75 años que residen habitualmente en viviendas privadas
 
@@
 
-## Métodos
+## Outcome
+
+<div style="text-align:center;">
+  <img src="images/outcome.png" style="width:100%; max-width:none;">
+</div>
+
+## Tratamiento
+::: {style="font-size: 120%;"}
+Siguiendo la propuesta de @breen_effects_2024 para estimar el efecto causal de la movilidad social: 
+:::
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+1.
+**Asignación de clase**: En base a la ocupación (ISCO-08) del padre/madre y del entrevistado se crean estratos de origen y destino (bajo, medio y alto) según terciles de la distribución del índice de estatus ocupacional (ISEI)
+
+2.
+**Propensity-score estimation**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles para cada estrato (categ.
+ref.).
+Estas puntuaciones de propensión se emplean posteriormente para ajustar la selección en la movilidad al evaluar su efecto.
+
+:::
+
+## Tratamiento
+
+## Tratamiento
+
+<div style="text-align:center;">
+  <img src="images/propensity.png" style="width:105%; max-width:none;">
+</div>
+
+## Mediador
+
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+- **Meritocracia**: se mide a partir de dos componentes, a saber, el esfuerzo y el talento [@young_rise_1958]
+
+- Medición:
+
+    * Ítem esfuerzo: “En Chile, se recompensa a las personas por su esfuerzo”
+    * Ítem talento: “En Chile, se recompensa a las personas por su inteligencia y habilidades”
+    * Ambos ítems se responden en una escala Likert de 1 (totalmente en desacuerdo) a 5 (totalmente de acuerdo)
+
+:::
+
+## Estimand y estimando
 
@@
 
-# Discusión y conclusiones {data-background-color="#523870"}
+```{r}
+#| label: fig-mh
+#| fig-cap-location: top
+#| fig-cap: "Estimaciones de efectos de mobilidad en preferencia por comodificación en salud"
+#| fig-align: center
+#| out-width: '100%'
 
-# Discusión y conclusiones {data-background-color="#523870"}
+df_health <- bind_rows(
+tidy(mh_lm, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Low→Middle \n(N obs. = 532; N Clusters = 381)") 
+,
+tidy(mh_lh, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Low→High \n(N obs. = 452; N Clusters = 342)") 
+,
+tidy(mh_mh, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Middle→High \n(N obs. = 432; N Clusters = 320)") 
+,
+tidy(mh_ml, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Middle→Low \n(N obs. = 469; N Clusters = 345)") 
+,
+tidy(mh_hm, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "High→Middle \n(N obs. = 620; N Clusters = 444)") 
+,
+tidy(mh_hl, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "High→Low \n(N obs. = 576; N Clusters = 432)") 
+)
+
+df_health %>%   
+  ggplot(aes(x = estimate, y = term)) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8,) +
+  geom_point(aes(x = estimate), size = 3.5) +
+  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
+                width = 0, size = 1, position = position_dodge2()) +
+  scale_x_continuous(limits = c(-1, 1),
+                     breaks = seq(-1,1,0.5)) +
+  labs(
+    x = "Coeficiente (β)",
+    y = NULL
+  ) +
+  theme_ggdist() +
+  theme(axis.title = element_text(face = "bold"),
+        axis.text = element_text(face = "bold"))
+
+```
+
+## Mediación
+
+# Discusión y proyecciones {data-background-color="#523870"}
+
+# Discusión y proyecciones {data-background-color="#523870"}
 

### 2025-09-03 presentations/dag.R

@@ -1,0 +1,103 @@
+library(dagitty)
+library(ggdag)
+library(ggplot2)
+
+# Coordenadas simples
+coords <- list(
+  x = c(Mobility=0, Meritocracy=1.5, Preferences=3),
+  y = c(Mobility=1.5, Meritocracy=0.5, Preferences=1.5)
+)
+
+# DAG teórico simplificado
+dag_mjp_simple <- dagitty("
+dag {
+  Preferences [pos=\"3,1.5\"]
+  Mobility [pos=\"0,1.5\"]
+  Meritocracy [pos=\"1.5,0.5\"]
+
+  Preferences <- Mobility
+  Preferences <- Meritocracy
+  Meritocracy <- Mobility
+}
+")
+
+coordinates(dag_mjp_simple) <- coords
+
+# Graficar
+ggdag(dag_mjp_simple, text = FALSE, node_size = 18) +
+  geom_dag_point(aes(color = name), show.legend = FALSE) +
+  geom_dag_text(color="white", size=4) +
+  theme_dag_blank() +
+  labs(title = "DAG simplificado: Movilidad social, Meritocracia y Preferencias por comodificación")
+
+
+
+library(dagitty) # Instalar V8
+library(ggdag)
+
+coords <- list(
+  x = c(Y=2, T=0, U=-1, F=1, X=0, T1=-1, T2=0, Z=1.2),
+  y = c(Y=1, T=1, U=1, F=2, X=0, T1=-1, T2=-1, Z=1.2)
+)
+
+HPTbestcase <- dagify(
+  Y ~ T + F + X + Z,
+  F ~ U,
+  Z ~ T + F,
+  T ~ U + X,
+  exposure = "T",
+  outcome = "Y",
+  coords = coords
+)
+
+tidy_dag <- tidy_dagitty(HPTbestcase)
+
+tidy_dag$data$time <- c("t3", "t3", "t2", "t2", "t1", "t1", "t2", "t2", "t3", "t4")
+
+ggdag(tidy_dag, node_size=8, text_size = 4) +
+  theme_classic() + remove_axes() + theme(axis.line=element_blank()) +
+  ylim(-1.5, 2.5) + xlim(-1.5,2.5) +
+  geom_text(x=-1, y=-.5, label=expression(t[1]), size=3) +
+  geom_text(x=0, y=-.5, label=expression(t[2]), size=3) +
+  geom_text(x=1, y=-.5, label=expression(t[3]), size=3) +
[…]

### 2025-09-03 processing/prod_analysis.R

@@ -52,12 +52,14 @@ data <- datos1 %>%
 
 data
 
-data %>% 
+mp_lm <- data %>% 
   lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
             weights = w,
             se_type = "CR2",
             clusters = idencuesta,
-            data = .) %>% 
+            data = .) 
+
+mp_lm %>%  
   screenreg()
 
 ### Low to high
@@ -79,12 +81,14 @@ data <- datos1 %>%
 
 data
 
-data %>% 
+mp_lh <- data %>% 
   lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
             weights = w,
             se_type = "CR2",
             clusters = idencuesta,
-            data = .) %>% 
+            data = .) 
+            
+mp_lh %>% 
   screenreg()
 
 
@@ -107,12 +111,14 @@ data <- datos2 %>%
 
 data
 
-data %>% 
+mp_mh <- data %>% 
   lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
             weights = w,
             se_type = "CR2",
             clusters = idencuesta,
-            data = .) %>% 
+            data = .) 
+    
+mp_mh %>% 
   screenreg()
 
 
@@ -135,12 +141,14 @@ data <- datos2 %>%
 
 data
 
-data %>% 
+mp_ml <- data %>% 
   lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
             weights = w,
             se_type = "CR2",
             clusters = idencuesta,
-            data = .) %>% 
+            data = .) 
+            
+mp_ml %>% 
   screenreg()
 
 
@@ -163,12 +171,14 @@ data <- datos3 %>%
 
 data
 
-data %>% 
+mp_hm <- data %>% 
   lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
             weights = w,
             se_type = "CR2",
             clusters = idencuesta,
-            data = .) %>% 
+            data = .) 
+            
+mp_hm %>% 
   screenreg()
 
 ### High to low
@@ -190,12 +200,14 @@ data <- datos3 %>%
 
 data
 
-data %>% 
+mp_hl <- data %>% 
   lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
             weights = w,
             se_type = "CR2",
             clusters = idencuesta,
-            data = .) %>% 
[…]

### 2025-09-04 presentations/coloquio_2025/mjp-mobility.qmd
Up coloquio

@@
 
-</br>
+::: {.box-inv-4 .sp-after .fragment style="font-size: 100%;"}
 
-::: {.box-inv-4 .sp-after .fragment style="font-size: 150%;"}
-
-Mercantilización y privatización (M, AL, CL)
+1) Privatización y mercantilización de  bienes públicos, políticas de bienestar y servicios sociales [@gingrich_making_2011; @streeck_how_2016]
 
@@
 
-::: {.box-inv-4 .sp-after .fragment style="font-size: 150%;"}
+::: {.box-inv-4 .sp-after .fragment style="font-size: 100%;"}
 
-Economía moral (Mau, bringing market, policy feedback)
+2) En AL y Chile, modificaron la arquitectura de las instituciones del bienestar expandiendo lógica de mercado [@ferre_welfare_2023; @madariaga_three_2020]
+
 :::
 
-::: {.box-inv-4 .sp-after .fragment style="font-size: 150%;"}
+::: {.box-inv-4 .sp-after .fragment style="font-size: 100%;"}
 
-MJP (Lane, Lindh, implicancias y relevancias)
+3) Este orden económico se refleja en una economía moral específica [@mau_inequality_2015; @svallforsMoralEconomyClass2006a]
+
+:::
+
+::: {.box-inv-5 .sp-after-half .fragment style="font-size: 100%;"}
+
+Preferencias por justicia de mercado [@busemeyer_skills_2014; @immergut_it_2020; @castillo_perceptions_2025; @koos_moral_2019; @lindh_public_2015]
+
+:::
+
+## Preferencias por justicia de mercado
+
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+- Lane [-@lane_market_1986]: justicia de mercado vs. justicia política
+
+- Creencias normativas que legitiman la idea de que el acceso a los servicios sociales esenciales —como la salud, la educación o las pensiones— debe determinarse según criterios basados en el mercado [@lindh_public_2015, p.895]
+
+- Medición: evaluar si las personas consideran justo que el acceso a dichos servicios dependa de los ingresos [@lindh_public_2015; @kluegel_legitimation_1999; @castillo_perceptions_2025]
+
+:::
+
+::: {.box-inv-5 .sp-after .fragment style="font-size: 120%;"}
+
+Preferencias por la comodificación de servicios
+
 :::
@@
 ::::: columns
-::: {.column width="50%"}
+::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}
 
@@
 
-- The original definition of merit is a combination of effort and talent [@young_rise_1958]
-- 2
-- 3
+- Gasto social [@immergut_it_2020; @busemeyer_skills_2014; @busemeyer_welfare_2020]
+- Desigualdad económica [@koos_moral_2019]
+- Nivel de privatización de servicios y regulación del mercado [@lindh_public_2015; @koos_moral_2019]
 
@@
 
-::: {.column width="50%"}
+::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}
 
@@
 
-- Adolescents from low-socioeconomic-status families are likelier to have an egalitarian fairness view and consider an equal distribution as fair in a situation with unequal merits [@almas_fairness_2017]
-- 2
-- 3
+- Estatus socioeconómico -ingresos, educación y ocupación- [@lindh_public_2015; @koos_moral_2019; @busemeyer_welfare_2020; @svallfors_political_2007]
+- Percepciones sobre la desigualdad y meritocracia [@castillo_perceptions_2025]
+- Conservadurismo/liberalismo económico [@lee_fairness_2023]
 
@@
 
-::: {.incremental .highlight-last style="font-size: 100%;"}
+::: {.incremental .highlight-last style="font-size: 110%;"}
 
-- Otero 2024 (clase)
-- Castillo y laffert 2025 (merit)
+1. @otero_power_2024 proveen evidencia sobre:
+
+   * Personas en posiciones sociales bajas/subordinadas muestran menor apoyo a la justicia de mercado que aquellas en posiciones altas/privilegiadas.
+
+2. @castillo_perceptions_2025 encuentran que:
+
+   * La percepción de la desigualdad económica reduce el apoyo a la justicia de mercado.
+   * Las percepciones meritocráticas (esfuerzo) aumenta el respaldo a la justicia de mercado.
 
@@
 
-::: {.incremental .highlight-last style="font-size: 100%;"}
+</br>
 
-- Aunque esta claro... movilidad social
-- Efectos
-- Mecanismos
+::: {.pull-left .box-inv-5 .sp-after .fragment style="font-size: 120%;"}
 
+Cómo la movilidad social afecta estas preferencias sigue sin respuesta
 :::
 
-## Este estudio
+</br></br>
 
-::: {.incremental .highlight-last style="font-size: 100%;"}
+::: {.pull-right .box-inv-5 .sp-after .fragment style="font-size: 120%;"}
 
-- Mecanismos
-
-:::
-
-## Contexto chileno
-
-::: {.incremental .highlight-last style="font-size: 100%;"}
-
-- RRelevancia del caso Chileno: extrema comodificacion, desigualdad, limitada movilidad, alta pferencia por la meritocracia etc.
+El movimiento entre posiciones expone experiencias y **mecanismos** que afectan nociones de justicia
 
@@
 
-**_(1) ¿En qué medida la movilidad social intergeneracional afecta las preferencias por comodificación en salud, pensiones y educación en Chile?_**
+**_(1) ¿Cómo la movilidad social intergeneracional afecta las preferencias por comodificación en salud, pensiones y educación en Chile?_**
 
@@
 ::::
+
+## Movilidad social
+
+::::: columns
+::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}
+
+### Afecta
+
+- Preferencias redistributivas [@alesina_intergenerational_2018; @ares_changing_2020; @schmidt_experience_2011; @breen_effects_2024]
+- Creencias sobre la desigualdad [@gugushvili_intergenerational_2016; @bucca_merit_2016; @shariff_income_2016; @day_movin_2017]
+- Apoyo al gasto social y políticas de bienestar [@gugushvili_subjective_2017; @gugushvili_intergenerational_2016c]
+
+:::
+
+::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}
+
+### Mecanismos
+
+- Intereses materiales [@helgason_class_2025]
+- Socialización [@jaime-castillo_social_2019]
+- **Self-serving bias** [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
+- Meritocracia como proxy al sesgo de atribuciones internas y externas [@young_rise_1958]
+:::
+:::::
+
+## Contexto chileno
+
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+- Crecimiento con elevada desigualdad [@llorca-jana_historia_2021; @flores_top_2020]
+- Profunda privatización y comodificación de áreas de reproducción social con fuerte dependencia Estatal [@madariaga_three_2020; @boccardo_30_2020]
+- Rigida movilidad relativa y fuerte clausura cúspide [@lopez-roldan_comparative_2021; @torche_intergenerational_2014]
+- Conflictividad social [@somma_no_2021; @nucleodesociologiacontingente_informe_2020] y legitimidad de la desigualdad [@castillo_perceptions_2025; @canalesceron_sujeto_2021; @panes_criticas_2020]
+:::
 
@@
 
+```{r}
+#| label: tbl-ocup
+#| tbl-cap-location: top
+#| tbl-cap: "Movilidad ocupacional por grupos ocupacionales"
+
+library(sjPlot)
+
+db <- db %>% 
+  mutate(
+    across(
+      .cols = c(estrato_orig, estrato_ocupa),
+      .fns = ~ car::recode(.,
+                           recodes = c("1 = 'Low';
+                                       2 = 'Middle';
+                                       3 = 'High'"),
+                           as.factor = T)
+    )
+  ) %>% 
+  mutate(
+    across(
+      .cols = c(estrato_orig, estrato_ocupa),
+      .fns = ~ factor(., levels = c("Low", "Middle", "High"))
+    )
+  )
+
+db$estrato_orig <- sjlabelled::set_label(db$estrato_orig, "Father stratum")
+
+db$estrato_ocupa <- sjlabelled::set_label(db$estrato_ocupa, "Offspring stratum")
+
+sjPlot::sjt.xtab(db$estrato_orig, 
+                 db$estrato_ocupa, 
+                 show.row.prc = T,
+                 show.col.prc = T,
+                 statistics = NULL)
+```
+
+- _N_ (id único) = 4.447. Porcentajes en azul corresponde a las filas y en verde a las columnas.
+
 ## Tratamiento
@@
 
+::: {.fragment}
+
 ```{r}
@@
 #| fig-align: center
-#| out-width: '100%'
+#| out-width: '180%'
+
+library(broom)
+library(forcats)
+conflicts_prefer(dplyr::filter)
 
@@
 
-df_health %>%   
-  ggplot(aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8,) +
-  geom_point(aes(x = estimate), size = 3.5) +
-  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
-                width = 0, size = 1, position = position_dodge2()) +
-  scale_x_continuous(limits = c(-1, 1),
-                     breaks = seq(-1,1,0.5)) +
-  labs(
-    x = "Coeficiente (β)",
-    y = NULL
-  ) +
+df_health_lab <- df_health %>%
+  mutate(
+    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
+    est_lab = sprintf("%.2f%s", estimate, sig),
+    low_lab = sprintf("%.2f", conf.low),
+    high_lab= sprintf("%.2f", conf.high)
+  )
+
+ggplot(df_health_lab, aes(x = estimate, y = term)) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8) +
+  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
+  geom_point(size = 3.5) +
+  # etiquetas: coef en el punto
+  geom_text(aes(label = est_lab), vjust = -0.8, size = 3.3) +
+  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
+  labs(x = "Estimate (β)", y = NULL) +
   theme_ggdist() +
   theme(axis.title = element_text(face = "bold"),
-        axis.text = element_text(face = "bold"))
+        axis.text  = element_text(face = "bold"),
+        text = element_text(size = 18))
 
@@
 
-## Mediación
+:::
+
+## Modelos
+
+```{r}
+#| label: fig-me
+#| fig-cap-location: top
+#| fig-cap: "Estimaciones de efectos de mobilidad en preferencia por comodificación en educación"
+#| fig-align: center
+#| out-width: '180%'
+
+df_educ <- bind_rows(
+tidy(me_lm, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Low→Middle \n(N obs. = 534; N Clusters = 380)") 
+,
+tidy(me_lh, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Low→High \n(N obs. = 450; N Clusters = 342)") 
+,
+tidy(me_mh, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Middle→High \n(N obs. = 424; N Clusters = 311)") 
+,
+tidy(me_ml, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Middle→Low \n(N obs. = 467; N Clusters = 342)") 
+,
+tidy(me_hm, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "High→Middle \n(N obs. = 611; N Clusters = 437)") 
+,
+tidy(me_hl, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "High→Low \n(N obs. = 569; N Clusters = 430)") 
+)
+
+df_educ_lab <- df_educ %>%
+  mutate(
+    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
+    est_lab = sprintf("%.2f%s", estimate, sig),
+    low_lab = sprintf("%.2f", conf.low),
+    high_lab= sprintf("%.2f", conf.high)
+  )
+
+ggplot(df_educ_lab, aes(x = estimate, y = term)) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8) +
+  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
+  geom_point(size = 3.5) +
+  # etiquetas: coef en el punto
+  geom_text(aes(label = est_lab), vjust = -0.8, size = 3.3) +
+  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
+  labs(x = "Estimate (β)", y = NULL) +
+  theme_ggdist() +
+   theme(axis.title = element_text(face = "bold"),
+        axis.text  = element_text(face = "bold"),
+        text = element_text(size = 18))
+
+```
+
+## Modelos
+
+```{r}
+#| label: fig-mp
+#| fig-cap-location: top
+#| fig-cap: "Estimaciones de efectos de mobilidad en preferencia por comodificación en pensiones"
+#| fig-align: center
+#| out-width: '180%'
+
+df_pension <- bind_rows(
+tidy(mp_lm, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Low→Middle \n(N obs. = 511; N Clusters = 365)") 
+,
+tidy(mp_lh, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Low→High \n(N obs. = 435; N Clusters = 333)") 
+,
+tidy(mp_mh, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Middle→High \n(N obs. = 396; N Clusters = 298)") 
+,
+tidy(mp_ml, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "Middle→Low \n(N obs. = 447; N Clusters = 335)") 
+,
+tidy(mp_hm, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "High→Middle \n(N obs. = 581; N Clusters = 421)") 
+,
+tidy(mp_hl, conf.int = TRUE) %>%
+  filter(term == "t") %>%
+  mutate(term = "High→Low \n(N obs. = 544; N Clusters = 415)") 
+)
+
+df_pension_lab <- df_pension %>%
+  mutate(
+    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
+    est_lab = sprintf("%.2f%s", estimate, sig),
+    low_lab = sprintf("%.2f", conf.low),
+    high_lab= sprintf("%.2f", conf.high)
+  )
+
+ggplot(df_pension_lab, aes(x = estimate, y = term)) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8) +
+  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
+  geom_point(size = 3.5) +
+  # etiquetas: coef en el punto
+  geom_text(aes(label = est_lab), vjust = -0.8, size = 3.3) +
+  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
+  labs(x = "Estimate (β)", y = NULL) +
+  theme_ggdist() +
+  theme(axis.title = element_text(face = "bold"),
+        axis.text  = element_text(face = "bold"),
+        text = element_text(size = 18))
+
+```
 
@@
 
-::: { .incremental .highlight-last2 style="font-size: 100%;"}
+::: { .incremental .highlight-last2 style="font-size: 120%;"}
 
-- Discusión
-
-  - Conclusión
+1. **Movilidad social**: ascendente (Middle→High) _incrementa_ la preferencia por comodificación en pensiones (dominio), no en otros. 
+2. **Contexto chileno**: profunda comodificación y privatización de servicios escenciales → trabajador-empresario 
+3. **Mediación**: rol de la meritocracia como mediador → ACME [@imai_identification_2010], moderación (?)
 :::

### 2025-09-04 presentations/coloquio_2025/mjp-mobility.tex
Up coloquio

--- 2025-06-27 advances/informe2.tex
@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
 \documentclass[
-  13pt,
-]{article}
-
+  spanish,
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod,
+  oneside]{scrartcl}
+\usepackage{xcolor}
+\usepackage[left=1in,marginparwidth=2.0666666666667in,textwidth=4.1333333333333in,marginparsep=0.3in]{geometry}
 \usepackage{amsmath,amssymb}
-\usepackage{setspace}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
   % xetex/luatex font selection
-  \setmainfont[]{Times New Roman}
 \fi
@@
 \makeatother
-\usepackage{xcolor}
-\usepackage[margin=2cm]{geometry}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\makeatother
+
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
+
 % definitions for citeproc citations
@@
 
-\usepackage[noblocks]{authblk}
-\renewcommand*{\Authsep}{, }
-\renewcommand*{\Authand}{, }
-\renewcommand*{\Authands}{, }
-\renewcommand\Affilfont{\small}
+\ifLuaTeX
+\usepackage[bidi=basic]{babel}
+\else
+\usepackage[bidi=default]{babel}
+\fi
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
+  \renewcommand*\contentsname{Tabla de contenidos}
 \else
-  \newcommand\contentsname{Table of contents}
+  \newcommand\contentsname{Tabla de contenidos}
 \fi
 \ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
+  \renewcommand*\listfigurename{Listado de Figuras}
 \else
-  \newcommand\listfigurename{List of Figures}
+  \newcommand\listfigurename{Listado de Figuras}
 \fi
 \ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
+  \renewcommand*\listtablename{Listado de Tablas}
 \else
-  \newcommand\listtablename{List of Tables}
+  \newcommand\listtablename{Listado de Tablas}
 \fi
 \ifdefined\figurename
-  \renewcommand*\figurename{Figure}
+  \renewcommand*\figurename{Figura}
 \else
-  \newcommand\figurename{Figure}
+  \newcommand\figurename{Figura}
 \fi
 \ifdefined\tablename
-  \renewcommand*\tablename{Table}
+  \renewcommand*\tablename{Tabla}
 \else
-  \newcommand\tablename{Table}
+  \newcommand\tablename{Tabla}
 \fi
@@
 \@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\floatname{codelisting}{Listado}
+\newcommand*\listoflistings{\listof{codelisting}{Listado de Listados}}
 \makeatother
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+\makeatletter
+\@ifpackageloaded{sidenotes}{}{\usepackage{sidenotes}}
+\@ifpackageloaded{marginnote}{}{\usepackage{marginnote}}
+\makeatother
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
-  pdftitle={Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
-  pdfauthor={Andreas Laffert Tamayo},
+  pdflang={es},
   colorlinks=true,
@@
 
-\title{Preferences for mercantilization: the role of intergenerational
-social mobility and beliefs in meritocracy in Chile}
-
-  \author{Andreas Laffert Tamayo}
-            \affil{%
-                  Instituto de Sociología, Pontificia Universidad
-                  Católica de Chile
-              }
-
+\author{}
 \date{}
 \begin{document}
-\maketitle
-
-\setstretch{1.5}
-\section{Introduction}\label{introduction}
-
-What is the legitimate extent of market inequality in the eyes of the
-public? Since the early 1980s, many countries have experienced a
-widespread retreat from universal welfare programs and a shift toward
-the privatization and commodification of public goods, welfare policies,
-and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
-2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
-America, as elsewhere, neoliberal reforms reshaped welfare-state
-institutions by extending market logic into domains of social
-reproduction that were traditionally governed by the state
-(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
-reduced the role of public provision and increased the presence of
-private actors in core social services
-(\citeproc{ref-harvey_breve_2015}{Harvey, 2015}). Echoing Polanyi's
-(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
-constitute a distinct moral order, the institutional diffusion of market
-rules has fostered a corresponding moral economy: a constellation of
-norms and values concerning fair allocation, embedded in institutions
-and shaping individual subjectivities
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
-this framework, a growing body of research examines the extent to which,
-and the mechanisms by which, citizens consider it fair that the
-allocation of services like health care, pensions, and education be
-governed by market-based criteria---a phenomenon known as \emph{market
-justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
-2014}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
-2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
-Understanding these preferences is crucial, as they contribute to
-legitimizing economic inequality by framing it as the fair result of
-individual responsibility and limited state intervention
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
-
-Existing literature shows that market justice preferences are shaped by
-both the economic and institutional context and individuals' positions
-within social stratification. Grounded in the notion that economic
-institutions influence people's normative attitudes
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}), studies find
-that countries with stronger public provision or more expansive welfare
-states exhibit lower levels of market justice preferences
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
-more privatized contexts show stronger support for market-based criteria
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
-justice preferences tend to rise as individuals ``ascend'' the social
-structure, with those in more privileged positions in terms of class,
-education, and income holding stronger preferences for market-based
-solutions compared to those in more disadvantaged or at-risk positions
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
-\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
-2016}).
-
-Market justice preferences are shaped not only by objective
-socioeconomic conditions but also by popular beliefs about inequality.
-Among these, meritocracy is a key normative principle underpinning
-market-based distributive preferences
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). It frames inequality as
-inevitable but justifiable through effort and talent
-(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
-\citeproc{ref-young_rise_1958}{Young, 1958}). Studies show that
-individuals with stronger meritocratic beliefs tend to perceive less
-inequality and legitimize it by attributing economic differences to
-personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
-societies where access to services is largely governed by market logic,
-such beliefs play a critical role in normalizing inequality. Recent
-evidence from Chile shows that students who believe effort and talent
-are rewarded in their country express stronger preferences for
-market-based access to healthcare, pensions, and education
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
-
-Although it is clear that one's social position influences market
-justice preferences, the question of how upward or downward mobility
-within the social structure affects these preferences remains
-unanswered. This question is far from trivial, especially in Latin
-America, where many have experienced various forms of mobility amid high
-economic inequality and deep welfare privatization
-(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
-\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}).
-Social origins and destinations affect attitudes toward inequality in
-distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
-\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b},
-\citeproc{ref-gugushvili_subjective_2017}{2017};
-\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
-2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
-movement between these positions exposes individuals to different
-experiences and mechanisms that shape their views on what is fair
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
-\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
-research, examining the effects of social mobility on market justice
-preferences can help to illuminate how inequalities in access to social
-services are justified among individuals who have experienced, or not,
-changes in their social standing, and what are the normative mechanisms
-that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
-2015}).
-
-Beyond their isolated effects, social mobility and meritocratic beliefs
-interact in complex ways to shape market justice preferences. Among
-others, a key mechanism proposed in the literature to explain how
-mobility influences distributive justice preferences is the
-psychological process of self-serving bias in causal attribution
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
-suggests that individuals attribute failures---such as downward
-mobility---to external factors, while crediting successes---such as
-upward mobility---to their own merit and effort
-(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
-who experience upward mobility tend to view their social position as
-earned, making them more likely to believe that individuals are
-responsible for their own success or failure. Research shows that upward
-mobility is associated with weaker preferences for redistribution
-(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
-\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
-2011}) and stronger legitimacy of income inequality
-(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
-individuals who experience downward mobility tend to blame structural
-factors like inequality and are more supportive of redistribution while
-rejecting merit-based explanations
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
-together, I argue that meritocratic beliefs may reinforce this
-self-serving attribution mechanism by legitimizing one's social status
-as the outcome of personal merit, closely tied to attribution bias.
-
-Against this background, this article pursues two main objectives:
-first, to analyze the extent to which intergenerational social mobility
-influences market justice preferences regarding healthcare, pensions,
-and education; and second, to examine how meritocratic beliefs may
-moderate this relationship. Building on a theoretical framework that
-emphasizes how neoliberal transformations---particularly through the
-privatization and commodification of key areas of social
-reproduction---have profoundly reshaped processes of subject formation
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}), the central argument is
-that upward mobility increases support for market justice preferences,
-while downward mobility decreases it. Moreover, meritocratic beliefs are
-expected to moderate this relationship by reflecting a self-serving
-attribution bias, whereby individuals justify their social position in
-terms of personal merit.
-
-This study focuses on Chile, a particularly intriguing case for
-examining market justice preferences. Despite significant economic
-growth and poverty reduction, Chile has some of the highest levels of
-inequality in Latin America and among OECD countries
-(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
-\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
-coexists with short-range upward mobility among lower-class segments
-moving into middle strata, though strong barriers remain to reaching
-higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
-et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
-2014}). What makes Chile especially salient is that much of this
-inequality is rooted in deep neoliberal reforms that institutionalized
-the privatization and commodification of key social sectors
-(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
-during the dictatorship (1973--1989) and expanded in democracy, these
-reforms enabled the unprecedented emergence of markets in health,
-pensions, and education, with provision segmented by individuals'
-ability to pay and supported by public subsidies
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
-despite waves of protest against inequality and commodification from
-2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
-2021})---Chilean subjectivities have been increasingly shaped by
-neoliberal discourses and market logics, influencing their attitudes
-toward inequality and welfare distribution
-(\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli, 2012};
-\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
-
-In this context, the questions that guide this research are as follows:
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/logo_isuc.png}}
+
+\pandocbounded{\includegraphics[keepaspectratio]{images/jusmer_trans.png}}
+
+\section{\texorpdfstring{\textbf{Preferencias por la
+comodificación}}{Preferencias por la comodificación}}\label{preferencias-por-la-comodificaciuxf3n}
+
+\subsection{\texorpdfstring{\textbf{El rol de la movilidad social
+intergeneracional y las creencias meritocráticas en
+Chile}}{El rol de la movilidad social intergeneracional y las creencias meritocráticas en Chile}}\label{el-rol-de-la-movilidad-social-intergeneracional-y-las-creencias-meritocruxe1ticas-en-chile}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\textbf{Andreas Laffert\textsuperscript{1,2}}
+
+\textbf{\textsuperscript{1}Instituto de Sociología, Pontificia
+Universidad Católica de Chile}
+
+\textbf{\textsuperscript{2}Centro de Estudios de Conflicto y Cohesión
+Social - COES}
+
+Coloquio de Investigación en Justicia Distributiva y Desigualdad
+Socioeconómica
+
+4 Septiembre 2025, Santiago
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\subsection{¿A dónde quiero llegar?}\label{a-duxf3nde-quiero-llegar}
+
+\section{Contexto y motivación}\label{contexto-y-motivaciuxf3n}
+
+\subsection{Contexto y motivación}\label{contexto-y-motivaciuxf3n-1}
+
+\subsection{Antecedentes}\label{antecedentes}
 
 \begin{enumerate}
-\def\labelenumi{(\arabic{enumi})}
-\tightlist
-\item
-  To what extent does intergenerational social mobility influence market
-  justice preferences regarding healthcare, pensions, and education in
-  Chile?
-\item
-  How do meritocratic beliefs condition or moderate this relationship in
-  the Chilean context?
+\def\labelenumi{\arabic{enumi})}
+\tightlist
+\item
+  Privatización y mercantilización de bienes públicos, políticas de
+  bienestar y servicios sociales (Gingrich, 2011; Streeck, 2016)
 \end{enumerate}
 
-To address these questions, this study draws on large-scale,
-representative survey data collected in 2018 from the urban Chilean
-population aged 18 to 75 (n = 2,726). The next section outlines the
-theoretical framework linking market justice preferences, social
-mobility, and meritocratic beliefs, and proposes a set of hypotheses.
-This is followed by a description of the data, variables, and analytical
-strategy. The final sections present the empirical findings, offer an
-interpretation of the results, and conclude with a discussion of their
-implications.
-
-\section{Theoretical and empirical
-background}\label{theoretical-and-empirical-background}
-
-\subsection{Market justice
-preferences}\label{market-justice-preferences}
-
-Beyond state's capacity to reallocate resources from the advantaged to
-the vulnerable, market institutions also play a central role in
-distributing socially valuable goods and rewards
-(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As Polanyi
-(\citeproc{ref-polanyi_great_1975}{1975}) observed, economic integration
-in capitalist societies is primarily organized through market exchange,
-governed by a self-regulating price system embedded in institutional
-frameworks. These institutions are not mere aggregates of individual
-behavior, but social realities endowed with rules, mechanisms, and
-normative meanings that shape everyday thinking
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
-\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
-the economic order is mirrored in a moral economy: collectively shared
-norms and beliefs about justice in distribution, embedded and reinforced
-through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
-research from this perspective has focused on welfare institutions,
-recent scholarship has brought the market back into focus as a site of
-distributive justice beliefs and institutional responses to inequality
-(\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In many
-countries, privatization and commodification have expanded market logic
-into core areas of social reproduction, such as healthcare, education,
-pensions, and social security, deepening inequality in access to these
-services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
-\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
-for market-based welfare provision has grown, even in traditional
-welfare states, particularly among higher-income groups who view private
-alternatives as more efficient or higher in quality
-(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
-This shift calls for broader inquiry into how institutions like markets
-structure access to resources and legitimize inequalities
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-satz_por_2019}{Satz, 2019}).
-
-The legitimacy of market-based inequalities is closely tied to beliefs
-about distributive justice grounded in market principles. Beyond
-stratification structures and socioeconomic conditions, the study of
-inequality also focuses on individuals' beliefs about the origins of
-inequality, the normative frameworks that sustain those beliefs, the
-mechanisms that shape them, and their implications for attitudes and
-behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
-In this regard, empirical research on distributive justice focuses on
-individuals' conceptions of how goods and rewards \emph{should} be
-distributed in society
-(\citeproc{ref-jasso_distributive_2016}{Guillermina Jasso et al.,
-2016}). This line of inquiry allows for examining the extent to which
-economic inequality is perceived as just, or in a certain sense,
-legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
-Most of the research on this field explores the legitimacy of wage
-inequality, specially salary gaps between jobs at opposite ends of the
-occupational hierarchy
-(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
-\citeproc{ref-jasso_justice_1978}{Guillermina Jasso, 1978};
-\citeproc{ref-jasso_gender_1999}{G. Jasso \& Wegener, 1999}). Recently,
-others examines how individuals justify market-generated inequalities in
-access to core social services such as healthcare, education, pensions
-or social security. Here, legitimacy stems from the belief that access
-to these goods should follow market-based criteria
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
-services are treated as legitimate commodities; goods that can be
-traded, priced, and evaluated through market logic
-(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
-
-Market justice preferences refer to normative beliefs that legitimize
-the idea that access to core social services---such as healthcare,
-education, or pensions---should be determined by market-based criteria.
-Following Janmaat's (\citeproc{ref-janmaat_subjective_2013}{2013, p.
-359}) distinction, these preferences fall under the category of
-``beliefs,'' understood as normative ideas about what inequality should
-look like, as opposed to ``perceptions,'' which refer to subjective
-evaluations of existing inequality. Market justice preferences reflect
-the view that access to these services should depend on individuals'
-ability to pay, thus justifying inequalities generated by market
-mechanisms (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al.,
-1999}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). The concept draws
-on Lane's (\citeproc{ref-lane_market_1986}{1986}) classic contrast
-between market justice and political justice: the former is grounded in
-the idea of earned deserts---where rewards reflect effort, productivity,
-and skill---while the latter prioritizes need and equality, typically
-expressed in welfare state policies. Lane argued that markets and states
-differ in purpose (efficiency vs.~need), logic (individual
-vs.~collective), and fairness criteria (merit vs.~equality). Market
-justice assumes that markets are neutral, self-regulating systems in
-which fair procedures yield outcomes proportional to merit. Inequality,
-from this perspective, is not only expected but legitimate---as long as
-it arises from fair competition. In this way, market justice offers a
-moral lens through which individuals can view the commodification and
-stratified access to social services as fair and justified
-(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
-
-The empirical study of market justice preferences has employed various
-strategies to capture how individuals assess inequalities arising from
-market allocation. A common approach evaluates whether people find it
-fair that access to essential social services---such as healthcare,
-education, or pensions---depends on income. This builds on foundational
-work by Kluegel and Smith
-(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
-justifications for capitalist systems. Subsequent studies have extended
-this logic to a wider set of welfare goods. For example, Immergut and
-Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
-et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
-respondents believe it is just that higher-income individuals receive
-better healthcare. Similarly, Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) apply this framework to
-education. Complementing these efforts, more recent research has
-introduced composite indicators to capture broader orientations toward
-market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
-for instance, constructs an index averaging support for income-based
-access to healthcare and education in comparative perspective, while
-Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
-propose a single-item measure covering health, education, and pensions
-in Chile. These instruments seek to gauge to what extent individuals
-view market-generated inequalities as legitimate. Taken together, they
-capture two core dimensions of market distribution: the role of economic
-resources as a key determinant of outcomes, and the framing of services
-as tradable commodities that can be bought and sold according to ability
-to pay (\citeproc{ref-lindh_public_2015}{2015}).
-
-Comparative empirical research has identified several individual-level
-factors that influence support for market justice. Individuals in more
-advantaged socioeconomic positions---those with higher income,
-education, and occupational status---are consistently more likely to
-endorse market-based distributive principles
-(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
-Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
-from the service class are more likely to support market-based access to
-healthcare and education than skilled and unskilled workers across 17
-relatively affluent countries. In a comparative analysis, Svallfors
-(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
-expected class pattern appears clearly only in Sweden, where support for
-private education and healthcare varies systematically by class.
-Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
-that support for private education is stronger among high-income groups,
-while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
-von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
-report comparable findings for healthcare, suggesting that wealthier
-individuals perceive private provision as a means of maintaining
-relative advantage. In Chile, Otero and Mendoza
-(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
-higher income and university education express stronger support for
-market allocation in healthcare, education, and pensions. Beliefs about
-inequality and political orientation also matter. In Chile, Castillo et
-al. (\citeproc{ref-castillo_socialization_2024}{2024}) show that
-individuals who have strong meritocratic beliefs are more likely to
-support market-based distribution, while Lee and Stacy
-(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
-preferences are also greater as people lean toward economic
-conservatism. In this sense, market justice preferences are shaped by
-the interplay between structural position and normative reasoning.
-
-However, individual characteristics alone do not fully explain variation
-in support for market justice. A growing body of cross-national research
-highlights the importance of institutional arrangements in shaping these
-preferences. For instance, Immergut and Schneider
-(\citeproc{ref-immergut_it_2020}{2020}) find that in countries with
-higher public spending on healthcare, individuals are less likely to
-view income-based access as fair. Similarly, Busemeyer
-(\citeproc{ref-busemeyer_skills_2014}{2014}) shows that increased public
-investment in education is associated with lower support for privatized
-provision. Conversely, Lindh (\citeproc{ref-lindh_public_2015}{2015})
-finds that in countries with more market-oriented welfare systems,
-support for market-based distribution tends to be higher, suggesting
-that individual attitudes often align with institutional outputs. These
-findings are consistent with neo-institutionalist and policy feedback
-theories (\citeproc{ref-campbell_institutional_2020}{Campbell, 2020}),
-which argue that institutions do more than redistribute resources---they
-also shape the normative categories through which individuals assess who
-is deserving of support
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In this
-view, dominant values and preferences are both shaped by and embedded in
-institutional configurations
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014}), reinforcing the
-notion that institutions are not neutral structures, but active
-producers of the moral frameworks that legitimize or challenge
-inequality.
-
-\subsection{Social mobility}\label{social-mobility}
-
-The study of social mobility, its drivers and consequences has long been
-central to sociology. Classic theorists explored not only movement
-across social hierarchies but also its broader implications for class
-conflict, norm stability, and institutional change
-(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}). Sorokin
-(\citeproc{ref-sorokin_social_1927}{1927}) introduced the concept
-formally, defining mobility as the shift of individuals, values, or
-objects between positions within a stratification system, and
-distinguishing between horizontal and vertical forms. Later work
-differentiated intergenerational from intragenerational mobility, as
-well as absolute mobility---driven by structural change---from relative
-mobility, which captures the extent to which origins constrain
-destinations (\citeproc{ref-eyles_social_2022}{Eyles et al., 2022}). In
-Latin America, research shows high absolute but low relative mobility
-(\citeproc{ref-bucca_merit_2016}{Bucca, 2016}): although educational
-expansion and economic modernization have enabled some upward movement,
-status reproduction remains strong, especially among elites
-(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). This
-reflects deep structural inequality, segmented education systems,
-stratified labor markets, and legacies of dependent development that
-restrict access to mobility channels and reinforce the intergenerational
-transmission of advantage.
-
-Beyond mapping mobility patterns, growing research has examined its
-subjective and attitudinal effects. Mobility effects---defined as
-outcomes resulting from movement between origin and destination classes
-(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024})---have long
-attracted theoretical interest. Sorokin's (1959) dissociative hypothesis
-posits that mobility, whether upward or downward, may produce
-psychological strain due to conflicting norms between class contexts,
-leading to lower life satisfaction. Similarly, Lenski
-(\citeproc{ref-lenski_status_1954}{1954}) argued that status
-inconsistency---mismatches among education, income, and occupation---can
-undermine well-being. These ideas underpin extensive empirical work
-linking intergenerational mobility to outcomes such as life
-satisfaction, mental health, and stress
-(\citeproc{ref-gugushvili_heterogeneous_2024}{Gugushvili, 2024};
-\citeproc{ref-hadjar_does_2015}{Hadjar \& Samuel, 2015};
-\citeproc{ref-prag_subjective_2021}{Präg \& Gugushvili, 2021}).
-
-Research on intergenerational social mobility has increasingly examined
-its effects on attitudes toward economic inequality. A key area of
-inquiry focuses on how upward and downward mobility influence support
-for redistribution, though findings remain mixed. Alesina et al.
-(\citeproc{ref-alesina_intergenerational_2018}{2018}) show that
-individuals with pessimistic expectations about their
-mobility---particularly those anticipating downward movement---are more
-likely to support generous redistributive policies. Similarly, Ares
-(\citeproc{ref-ares_changing_2020}{2020}) finds that upwardly mobile
-individuals tend to be less supportive of state-led redistribution
-compared to those who have experienced downward mobility. Comparative
-studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
-Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{2017}) likewise
-report that subjective upward mobility is associated with weaker
-preferences for redistribution, while downward mobility strengthens
-redistributive support. However, recent causal evidence from Breen and
-Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the opposite:
-upward mobility may increase redistributive preferences, while downward
-mobility may reduce them.
-
-Beyond redistribution, other studies have explored the impact of
-mobility on broader beliefs about inequality. Gugushvili
-(\citeproc{ref-gugushvili_intergenerational_2016}{2016b}) finds that
-upwardly mobile individuals are more likely to adopt individualistic
-attributions of poverty and to legitimize income inequality,
-particularly in post-socialist societies. Similarly, Bucca
-(\citeproc{ref-bucca_merit_2016}{2016}) shows that subjective upward
-mobility reinforces individualistic explanations of wealth in seven
-Latin American countries. At the macro level, Shariff et al.
-(\citeproc{ref-shariff_income_2016}{2016}) demonstrate that higher
-levels of national economic mobility correlate with greater tolerance of
-inequality. In contrast, Day and Fiske
-(\citeproc{ref-day_movin_2017}{2017}) find that low perceived mobility
-undermines belief in meritocracy and a just world, thereby weakening
-system justification. Taken together, this body of research suggests
-that mobility shapes attitudes toward inequality and justice through
-multiple, and sometimes contradictory, mechanisms.
-
-The literature on the effects of social mobility has proposed various
-mechanisms to explain how and why changes in social position may
-influence individual outcomes
-(\citeproc{ref-helgason_class_2025}{Helgason \& Rehm, 2025}). One of the
-most prominent is the self-interest mechanism, which posits that
-individuals who experience upward or downward mobility undergo a shift
-in their material interests, thereby altering their perceptions and
-preferences (\citeproc{ref-ares_changing_2020}{Ares, 2020};
-\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
-\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
-Closely related to this logic is the Prospect of Upward Mobility (POUM)
-hypothesis, which suggests that individuals may oppose redistribution
-not because of their current position, but because they anticipate
-improving their status in the future
-(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}). A second line
-of explanation draws on the framework of theories of socialization.
-Within this tradition, hypotheses such as acculturation, socialization,
-and status maximization propose that individuals adjust their attitudes
-based on the norms and values of either their class of origin or their
-destination, or a combination of both
-(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}). However, most of these mechanisms focus
-primarily on the indirect effects of origin and destination positions,
-rather than on the direct effect of experience of movement itself. In
-response to this gap, recent research has highlighted the role of
-self-serving bias in causal attribution processes, suggesting that
-individuals tend to explain their mobility trajectories in ways that
-justify their current position, which in turn shapes their beliefs and
-preferences
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-molina_its_2019}{Molina et al., 2019};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}).
-
-The self-serving bias mechanism builds on a value-oriented perspective,
-emphasizing that individuals' experiences of social mobility shape their
-causal attributions, which in turn influence their beliefs about justice
-and distributive preferences
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Causal
-attribution refers to the process through which individuals generate
-explanations for their own behavior and outcomes, as well as those of
-others (\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
-2016b}). In this view, people's interpretations of economic inequality
-depend on whether they believe such disparities reflect unequal
-individual contributions. Individuals who adopt an internal attribution
-framework tend to see success or failure as rooted in personal
-characteristics such as effort, talent, or merit. In contrast, those who
-regard inequality as unjust are more likely to adopt an external
-attribution model, viewing outcomes as the result of structural barriers
-beyond individual control (\citeproc{ref-kluegel_beliefs_1981}{Kluegel
-\& Smith, 1981}). This mechanism, often described as intrapersonal
-causal attribution, reflects how people explain their own socioeconomic
-positions---typically attributing their successes to internal qualities
-while blaming failures on external circumstances
-(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Over
-time, individuals may revise their beliefs and attitudes: while early
-views are shaped by their social origin, these are later adjusted in
-light of personal experiences of mobility and the perceived role of
-ascribed versus achieved factors in determining socioeconomic outcomes
-(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}).
-
-Empirical research has provided support for the self-serving bias
-mechanism in shaping redistributive preferences and attitudes toward the
-legitimacy of inequality. For instance, Schmidt
-(\citeproc{ref-schmidt_experience_2011}{2011}) finds that individuals
-who experience upward mobility are more likely to interpret their
-success as the result of personal effort or merit, and consequently
-perceive redistribution as less necessary. Conversely, individuals who
-experience downward mobility tend to attribute their decline to external
-circumstances---such as structural inequality or unemployment---and show
-stronger support for redistribution. In a comparative analysis across
-different welfare domains, Gugushvili
-(\citeproc{ref-gugushvili_subjective_2017}{2017}) demonstrates that
-upward mobility is associated with lower support for government spending
-on housing and pensions, while individuals who experience downward
-mobility express lower support for healthcare and education spending,
-but favor increased investment in housing and pensions, reflecting the
-material nature of these domains. Moreover, Gugushvili
-(\citeproc{ref-gugushvili_intergenerational_2016c}{2016a}) finds that
-upward mobility is linked to greater justification of income inequality,
-suggesting that improved social standing reinforces an attributional
-view in which success is seen as the result of individual
-characteristics---thus legitimizing inequality as a fair outcome.
-
-Consequently, considering this theoretical and empirical background, the
-first hypothesis of this research is that:
-
-\(H1\): Experiencing upward (downward) social mobility is positively
-(negatively) associated with greater support for market justice in
-healthcare, pensions, and education.
-
-\subsection{Meritocracy}\label{meritocracy}
-
-Meritocracy constitutes a central ideological framework for legitimizing
-different types of social inequality, for instance through market
-justice beliefs. Rooted in the belief that rewards and positions should
-be allocated based on individual effort and talent, meritocracy operates
-as a normative ideal and a descriptive belief about how society
-functions. As initially conceptualized by Michael Young
-(\citeproc{ref-young_rise_1958}{1958}), the term was meant to critique a
-system in which merit-based stratification becomes a new form of
-inequality. However, over time, meritocracy has been widely supported in
-many societies as a fair and desirable principle of distribution,
-particularly within liberal democracies and market-oriented economies
-(\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
-\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
-perspective, the belief in meritocracy is more than a cognitive
-assessment; it reflects a moral lens through which individuals interpret
-inequality. People who believe that success results from hard work and
-talent are more likely to view social and economic disparities as
-legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
-Conversely, if they see outcomes as driven by luck, social origin, or
-systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural
-inequality, where public narratives often emphasize personal
-responsibility and merit while overlooking entrenched disadvantages.
-
-I adopt a multidimensional perspective on meritocracy, distinguishing
-between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through
-which individuals justify inequality
-(\citeproc{ref-young_rise_1958}{Young, 1958}). Effort-based meritocracy
-emphasizes hard work and perseverance as the basis for social rewards,
-aligning closely with cultural narratives of personal responsibility. A
-talent-based meritocracy, by contrast, emphasizes intelligence and
-innate abilities, which are often perceived as less malleable and more
-unequally distributed. Both dimensions have been shown to correlate with
-acceptance of inequality, but they may carry distinct implications for
-how inequality is justified in specific domains
-(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
-The relevance of this distinction is supported by recent studies, which
-show that individuals respond differently to these dimensions. For
-instance, perceptions that effort is rewarded in society are more
-strongly associated with positive evaluations of fairness and acceptance
-of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}; \citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al., 2015};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}). This may be because
-effort is seen as a controllable and morally virtuous trait, whereas
-talent is often perceived as a natural advantage. Consequently,
-effort-based meritocracy is likely more potent in legitimizing
-inequality, particularly in neoliberal contexts.
-
-These dimensions of meritocracy reflect how respondents perceive
-society's distributive logic, regardless of whether they endorse
-meritocratic principles. This distinction aligns with recent findings
-indicating that individuals distinguish between how merit is perceived
-in society and how it should ideally operate, which in turn shapes their
-preferences for redistribution and justice
-(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
-2025}). Meritocratic beliefs serve as symbolic justifications for
-unequal outcomes, particularly when access is stratified by income or
-social opportunity. Prior studies have shown that individuals who
-perceive higher levels of meritocracy tend to express stronger support
-for unequal distributions that reflect market outcomes
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
-
-In addition to influencing individual attitudes toward inequality,
-meritocratic beliefs can contribute to social division and the
-stigmatization of disadvantaged groups. Recent research has demonstrated
-that exposure to meritocratic narratives can reinforce the belief that
-poverty results from individual failings rather than systemic
-conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
-et al., 2023}). This reinforces negative stereotypes and reduces empathy
-toward individuals from lower socioeconomic backgrounds. Moreover,
-Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
-that meritocratic narratives can serve as feedback mechanisms that shape
-public opinion and well-being by framing individuals' understanding of
-welfare outcomes as deserved or undeserved within existing institutional
-structures. This psychological mechanism highlights the normative power
-of meritocracy in stabilizing unequal systems by shaping political
-attitudes and personal perceptions of success and failure.
-
-Importantly, recent research has explored how meritocratic beliefs
-interact with experiences of intergenerational mobility to shape
-distributive attitudes. The belief that one's success is earned can lead
-upwardly mobile individuals to internalize meritocratic narratives and
-justify existing inequalities, reinforcing support for market justice
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-molina_its_2019}{Molina et al., 2019}). Conversely,
-downwardly mobile individuals who maintain strong meritocratic beliefs
-may interpret their status as a personal failure, reducing their support
-for redistribution (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017}).
-At the macro level, Shariff et al.
-(\citeproc{ref-shariff_income_2016}{2016}) show that higher perceived
-mobility increases tolerance for inequality, suggesting that meritocracy
-and mobility are mutually reinforcing.
-
-Taken together, this literature supports the idea that meritocratic
-beliefs can moderate the relationship between mobility and market
-justice preferences. Individuals who experience mobility---especially
-upward---may draw on meritocratic narratives to legitimize both their
-own status and broader inequalities, thereby strengthening their support
-for market-based distribution. Accordingly, the second hypothesis of
-this study is:
-
-\(H2\): The positive (negative) relationship between upward (downward)
-social mobility and support for market justice in healthcare, pensions,
-and education is moderated by meritocratic beliefs; specifically, this
-association is stronger (weaker) among individuals with higher
-perceptions of meritocracy.
-
-\subsection{The Chilean context}\label{the-chilean-context}
-
-\section{Data, variables and methods}\label{data-variables-and-methods}
-
-\subsection{Data}\label{data}
-
-This study draws on data from the Chilean Longitudinal Social Survey
-(ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
-The survey is an annual panel collected from 2016 to 2023, comprising
-two independent samples (original and refreshment), featuring permanent
-and rotating questionnaire modules. It evaluates how individuals think,
-feel, and behave regarding conflict and social cohesion in Chile. The
-sampling design is complex, probabilistic, clustered, multistage, and
-stratified by city size, targeting men and women aged 18--75 who
-habitually reside in private dwellings in urban areas spanning 40 cities
-(92 municipalities, 13 regions). This analysis focuses on 2023, the most
-recent wave, which includes 1,737 respondents in the original sample and
-989 in the refreshment sample. Further details on sampling, attrition,
-and weighting can be found at https://coes.cl/encuesta-panel/, and the
-dataset is publicly available at
-https://dataverse.harvard.edu/dataverse/elsoc.
-
-\subsection{Variables}\label{variables}
-
-\emph{Outcome variables}
-
-\textbf{Market justice preferences}: The outcome variables in this study
-are market justice preferences. This construct is operationalized
-through three variables that address the degree of justification
-regarding whether access to social services in healthcare, pensions, and
-education should be income conditional. Specifically, the justification
-of inequality in healthcare is assessed by the question: ``Is it fair in
-Chile that people with higher incomes can access better healthcare than
-people with lower incomes?'' The same question is asked for pensions and
-education. In all cases, respondents indicate their level of agreement
-on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
-(``strongly agree''). Although these items theoretically reflect the
-same underlying concept of market justice
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}), they are analyzed
-separately because healthcare, pensions, and education may prompt
-distinct response patterns
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}).
-
-\emph{Independent variables}
-
-\textbf{Social mobility}: Intergenerational social mobility is treated
-as an exposure indicating whether respondents occupy a different class
-position from their fathers. Following Breen and Ermisch's
-(\citeproc{ref-breen_effects_2024}{2024}) framework for estimating
-causal effects of mobility, the variable is constructed in two stages.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  En AL y Chile, modificaron la arquitectura de las instituciones del
+  bienestar expandiendo lógica de mercado (Ferre, 2023; Madariaga, 2020)
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Este orden económico se refleja en una economía moral específica (Mau,
+  2015; Svallfors, 2006)
+\end{enumerate}
+
+Preferencias por justicia de mercado (Busemeyer, 2014; Castillo et~al.,
+2025; Immergut \& Schneider, 2020; Koos \& Sachweh, 2019; Lindh, 2015)
+
+\subsection{Preferencias por justicia de
+mercado}\label{preferencias-por-justicia-de-mercado}
+
+\begin{itemize}
+\item
+  Lane (1986): justicia de mercado vs.~justicia política
+\item
+  Creencias normativas que legitiman la idea de que el acceso a los
+  servicios sociales esenciales ---como la salud, la educación o las
+  pensiones--- debe determinarse según criterios basados en el mercado
+  (Lindh, 2015, p. 895)
+\item
+  Medición: evaluar si las personas consideran justo que el acceso a
+  dichos servicios dependa de los ingresos (Castillo et~al., 2025;
+  Kluegel et~al., 1999; Lindh, 2015)
+\end{itemize}
+
+Preferencias por la comodificación de servicios
+
+\subsection{Preferencias por justicia de
+mercado}\label{preferencias-por-justicia-de-mercado-1}
+
+\subsubsection{Contextual}\label{contextual}
+
+\begin{itemize}
+\tightlist
+\item
+  Gasto social (Busemeyer, 2014; Busemeyer \& Iversen, 2020; Immergut \&
+  Schneider, 2020)
+\item
+  Desigualdad económica (Koos \& Sachweh, 2019)
+\item
+  Nivel de privatización de servicios y regulación del mercado (Koos \&
+  Sachweh, 2019; Lindh, 2015)
+\end{itemize}
+
+\subsubsection{Individual}\label{individual}
+
+\begin{itemize}
+\tightlist
+\item
+  Estatus socioeconómico -ingresos, educación y ocupación- (Busemeyer \&
+  Iversen, 2020; Koos \& Sachweh, 2019; Lindh, 2015; Svallfors, 2007)
+\item
+  Percepciones sobre la desigualdad y meritocracia (Castillo et~al.,
+  2025)
+\item
+  Conservadurismo/liberalismo económico (Lee \& Stacey, 2023)
+\end{itemize}
+
+\subsection{Preferencias por justicia de mercado:
+Chile}\label{preferencias-por-justicia-de-mercado-chile}
 
@@
 \item
-  Class assignment. Using the International Socio-Economic Index of
-  Occupational Status (ISEI) derived from three-digit ISCO-08 codes,
-  both paternal (origin) and respondent (destination) occupations are
-  grouped into three strata:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Upper (codes 100-299: managers, directors, professionals)
-  \item
-    Middle (codes 300-499: skilled white-collar workers)
-  \item
-    Lower (codes 500 and above: manual, semi-skilled, or unskilled
-    workers)
-  \end{itemize}
-\item
-  Propensity-score estimation. To model mobility as a treatment---i.e.,
-  a transition from origin \emph{j} to destination \emph{k}---I estimate
-  multinomial logit propensity scores using covariates that influence an
-  individual's likelihood of moving between strata: (a) father's
-  educational level, (b) presence of both parents at age 15, (c)
-  nationality, (d) age, (e) sex, and (f) ethnicity.
+  Otero \& Mendoza (2024) proveen evidencia sobre:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Personas en posiciones sociales bajas/subordinadas muestran menor
+    apoyo a la justicia de mercado que aquellas en posiciones
+    altas/privilegiadas.
+  \end{itemize}
+\item
+  Castillo et~al. (2025) encuentran que:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    La percepción de la desigualdad económica reduce el apoyo a la
+    justicia de mercado.
+  \item
+    Las percepciones meritocráticas (esfuerzo) aumenta el respaldo a la
+    justicia de mercado.
+  \end{itemize}
 \end{enumerate}
 
-These propensity scores are subsequently employed to adjust for
-selection into mobility when assessing its effect on market justice
-preferences.
-
-\textbf{Meritocracy}: Meritocratic perception is operationalized through
-two components: one addressing effort and another focusing on talent
-(\citeproc{ref-young_rise_1958}{Young, 1958}). The item used to gauge
-effort is: ``In Chile, people are rewarded for their efforts,'' while
-the item for talent is: ``In Chile, people are rewarded for their
-intelligence and skills.'' In both cases, respondents indicate their
-level of agreement on a five-point Likert scale, ranging from 1
-(``strongly disagree'') to 5 (``strongly agree'').
-
-\emph{Controls}
-
-Sociodemographic and attitudinal variables are included to control for
-potential composition effects in the population. In terms of
-sociodemographic characteristics, I incorporate per capita household
-income quantile, educational level (1=Primary or below, 2=Secondary,
-3=Technical, 4=University or above), age (in years), and sex (1=Male,
-2=Female), which have been shown to significantly influence market
-justice preferences (\citeproc{ref-castillo_socialization_2024}{Castillo
-et al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
-attitudinal variables, I include political identification (1=Left,
-2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) because they may confound the relationship
-between market justice preferences and social mobility and meritocracy.
-
-\subsection{Methods}\label{methods}
-
-Following Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024})
-strategy for estimating the causal effect of social mobility on market
-justice preferences, I employ linear regression models with
-inverse-probability weights (IPW) for mobility conditional on class of
-origin. This approach allows estimation of the average causal effect of
-moving from an origin class \emph{j} to a destination class \emph{k} by
-comparing individuals who actually reach \emph{k} with their
-counterfactual outcome had they instead moved to an alternative
-destination \emph{k′} (\citeproc{ref-breen_effects_2024}{Breen \&
-Ermisch, 2024}). The resulting estimand is the average treatment effect
-on the treated (ATT).
-
-All the analyses will be conducted using R software.
-
-\section{References}\label{references}
+\subsection{Este estudio}\label{este-estudio}
+
+Cómo la movilidad social afecta estas preferencias sigue sin respuesta
+
+El movimiento entre posiciones expone experiencias y \textbf{mecanismos}
+que afectan nociones de justicia
+
+\section{}\label{section}
+
+\textbf{\emph{(1) ¿Cómo la movilidad social intergeneracional afecta las
+preferencias por comodificación en salud, pensiones y educación en
+Chile?}}
+
+\textbf{\emph{(2) ¿Cómo las creencias meritocráticas median esta
+relación en el contexto chileno?}}
+
+\subsection{Movilidad social}\label{movilidad-social}
+
+\subsubsection{Afecta}\label{afecta}
+
+\begin{itemize}
+\tightlist
+\item
+  Preferencias redistributivas (Alesina et~al., 2018; Ares, 2020; Breen
+  \& Ermisch, 2024; Schmidt, 2011)
+\item
+  Creencias sobre la desigualdad (Bucca, 2016; Day \& Fiske, 2017;
+  Gugushvili, 2016b; Shariff et~al., 2016)
+\item
+  Apoyo al gasto social y políticas de bienestar (Gugushvili, 2016a,
+  2017)
+\end{itemize}
+
+\subsubsection{Mecanismos}\label{mecanismos}
+
+\begin{itemize}
+\tightlist
+\item
+  Intereses materiales (Helgason \& Rehm, 2025)
+\item
+  Socialización (Jaime-Castillo \& Marqués-Perales, 2019)
+\item
+  \textbf{Self-serving bias} (Gugushvili, 2016a; Molina et~al., 2019;
+  Schmidt, 2011).
+\item
+  Meritocracia como proxy al sesgo de atribuciones internas y externas
+  (Young, 1958)
+\end{itemize}
+
+\subsection{Contexto chileno}\label{contexto-chileno}
+
+\begin{itemize}
+\tightlist
+\item
+  Crecimiento con elevada desigualdad (Flores et~al., 2020; Llorca-Jaña
+  \& Miller, 2021)
+\item
+  Profunda privatización y comodificación de áreas de reproducción
+  social con fuerte dependencia Estatal (Boccardo, 2020; Madariaga,
+  2020)
+\item
+  Rigida movilidad relativa y fuerte clausura cúspide (López-Roldán \&
+  Fachelli, 2021; Torche, 2014)
+\item
+  Conflictividad social (Núcleo de Sociología Contingente, 2020; Somma
+  et~al., 2021) y legitimidad de la desigualdad (Canales Cerón et~al.,
+  2021; Castillo et~al., 2025; Panes, 2020)
+\end{itemize}
+
+\subsection{Hipótesis}\label{hipuxf3tesis}
+
+\section{Método}\label{muxe9todo}
+
+\subsection{Estrategia de identificación
+causal}\label{estrategia-de-identificaciuxf3n-causal}
+
+\subsection{Datos}\label{datos}
+
+\begin{itemize}
+\item
+  Encuesta Longitudinal Social de Chile
+  \href{https://coes.cl/elsoc/}{(ELSOC)} de COES.
+\item
+  Muestra analítica: olas 2016 (N = 2.927), 2018 (N = 3.748) y 2023 (N =
+  2.726)
+\item
+  Encuesta panel representativa de zonas urbanas, localizadas en 40
+  ciudades del país (92 comunas y 13 regiones)
+\item
+  Diseño muestral complejo: probabilístico, por conglomerados,
+  multietápico y estratificado según el tamaño de las ciudades.
+\item
+  Población objetivo incluye a mujeres y hombres de entre 18 y 75 años
+  que residen habitualmente en viviendas privadas
+\end{itemize}
+
+\subsection{Outcome}\label{outcome}
+
+\subsection{Tratamiento}\label{tratamiento}
+
+Siguiendo la propuesta de Breen \& Ermisch (2024) para estimar el efecto
+causal de la movilidad social:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  \textbf{Asignación de clase}: En base a la ocupación (ISCO-08) del
+  padre/madre y del entrevistado se crean estratos de origen y destino
+  (bajo, medio y alto) según terciles de la distribución del índice de
+  estatus ocupacional (ISEI)
+\item
+  \textbf{Propensity-score estimation}: Se estima la probabilidad de
+  movilidad social mediante un modelo logístico multinomial en relación
+  con las trayectorias inmóviles para cada estrato (categ. ref.). Estas
+  puntuaciones de propensión se emplean posteriormente para ajustar la
+  selección en la movilidad al evaluar su efecto.
+\end{enumerate}
+
+\subsection{Tratamiento}\label{tratamiento-1}
+
+\begin{table}
+
+\caption{\label{tbl-ocup}Movilidad ocupacional por grupos ocupacionales}
+
+\centering{
+
+}
+
+\end{table}%
+
+\begin{itemize}
+\tightlist
+\item
+  \emph{N} (id único) = 4.447. Porcentajes en azul corresponde a las
+  filas y en verde a las columnas.
+\end{itemize}
+
+\subsection{Tratamiento}\label{tratamiento-2}
+
+\subsection{Mediador}\label{mediador}
+
+\begin{itemize}
+\item
+  \textbf{Meritocracia}: se mide a partir de dos componentes, a saber,
+  el esfuerzo y el talento (Young, 1958)
+\item
+  Medición:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Ítem esfuerzo: ``En Chile, se recompensa a las personas por su
+    esfuerzo''
+  \item
+    Ítem talento: ``En Chile, se recompensa a las personas por su
+    inteligencia y habilidades''
+  \item
+    Ambos ítems se responden en una escala Likert de 1 (totalmente en
+    desacuerdo) a 5 (totalmente de acuerdo)
+  \end{itemize}
+\end{itemize}
+
+\subsection{Estimand y estimando}\label{estimand-y-estimando}
+
+\begin{itemize}
+\tightlist
+\item
+  Métodos
+\end{itemize}
+
+\section{Resultados}\label{resultados}
+
+\subsection{Modelos}\label{modelos}
+
+\begin{figure}
+
+\caption{\label{fig-mh}Estimaciones de efectos de mobilidad en
+preferencia por comodificación en salud}
+
+\centering{
+
+\includegraphics[width=1.8\linewidth,height=\textheight,keepaspectratio]{mjp-mobility_files/figure-pdf/fig-mh-1.pdf}
+
+}
+
+\end{figure}%
+
+\subsection{Modelos}\label{modelos-1}
+
+\begin{figure}
+
+\caption{\label{fig-me}Estimaciones de efectos de mobilidad en
+preferencia por comodificación en educación}
+
+\centering{
+
+\includegraphics[width=1.8\linewidth,height=\textheight,keepaspectratio]{mjp-mobility_files/figure-pdf/fig-me-1.pdf}
+
+}
+
+\end{figure}%
+
+\subsection{Modelos}\label{modelos-2}
+
+\begin{figure}
+
+\caption{\label{fig-mp}Estimaciones de efectos de mobilidad en
+preferencia por comodificación en pensiones}
+
+\centering{
+
+\includegraphics[width=1.8\linewidth,height=\textheight,keepaspectratio]{mjp-mobility_files/figure-pdf/fig-mp-1.pdf}
+
+}
+
+\end{figure}%
+
+\section{Discusión y proyecciones}\label{discusiuxf3n-y-proyecciones}
+
+\section{Discusión y proyecciones}\label{discusiuxf3n-y-proyecciones-1}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Movilidad social}: ascendente (Middle→High) \emph{incrementa}
+  la preferencia por comodificación en pensiones (dominio), no en otros.
+\item
+  \textbf{Contexto chileno}: profunda comodificación y privatización de
+  servicios escenciales → trabajador-empresario
+\item
+  \textbf{Mediación}: rol de la meritocracia como mediador → ACME (Imai
+  et~al., 2010), moderación (?)
+\end{enumerate}
+
+\section{Gracias por su atención!}\label{gracias-por-su-atenciuxf3n}
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Github del proyecto:}
+  \url{https://github.com/Andreas-Lafferte/mobility-market-justice}
+\end{itemize}
+
+\subsection*{Referencias}\label{referencias}
+\addcontentsline{toc}{subsection}{Referencias}
 
@@
 {Mobility} and {Preferences} for {Redistribution}. \emph{American
-Economic Review}, \emph{108}(2), 521--554.
+Economic Review}, \emph{108}(2), 521-554.
 \url{https://doi.org/10.1257/aer.20162015}
 
-\bibitem[\citeproctext]{ref-araujo_desafios_2012}
-Araujo, K., \& Martuccelli, D. (2012). \emph{Desaf{í}os comunes: Retrato
-de la sociedad chilena y sus individuos} (1a. ed). Santiago: LOM
-Ediciones.
-
 \bibitem[\citeproctext]{ref-ares_changing_2020}
-Ares, M. (2020). Changing classes, changing preferences: How social
-class mobility affects economic preferences. \emph{West European
-Politics}, \emph{43}(6), 1211--1237.
+Ares, M. (2020). Changing Classes, Changing Preferences: How Social
+Class Mobility Affects Economic Preferences. \emph{West European
+Politics}, \emph{43}(6), 1211-1237.
 \url{https://doi.org/10.1080/01402382.2019.1644575}
 
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023). Belief in {School Meritocracy} and the {Legitimization} of
-{Social} and {Income Inequality}. \emph{Social Psychological and
-Personality Science}, \emph{14}(5), 621--635.
-\url{https://doi.org/10.1177/19485506221111017}
-
-\bibitem[\citeproctext]{ref-benabou_social_2001}
-Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
-{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
-Economics}, \emph{116}(2), 447--487.
-\url{https://doi.org/10.1162/00335530151144078}
-
 \bibitem[\citeproctext]{ref-boccardo_30_2020}
-Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
-que la pandemia revel{ó}} (Nodo XXI). Santiago.
+Boccardo, G. (2020). \emph{30 A{ñ}os de Privatizaciones En {Chile}: Lo
+Que La Pandemia Revel{ó}} (Nodo XXI). Santiago.
 
@@
 Breen, R., \& Ermisch, J. (2024). The {Effects} of {Social Mobility}.
-\emph{Sociological Science}, \emph{11}, 467--488.
+\emph{Sociological Science}, \emph{11}, 467-488.
 \url{https://doi.org/10.15195/v11.a17}
@@
 \bibitem[\citeproctext]{ref-bucca_merit_2016}
-Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
-Latin Americans}' beliefs about wealth and poverty. \emph{Research in
-Social Stratification and Mobility}, \emph{44}, 98--112.
+Bucca, M. (2016). Merit and Blame in Unequal Societies: {Explaining
+Latin Americans}' Beliefs about Wealth and Poverty. \emph{Research in
+Social Stratification and Mobility}, \emph{44}, 98-112.
 \url{https://doi.org/10.1016/j.rssm.2016.02.005}
@@
 
-\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
-Busemeyer, M., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive} and
-{Negative}: {New Perspectives} on {Feedback Effects} in {Public Opinion}
-on the {Welfare State}. \emph{British Journal of Political Science},
-\emph{51}(1), 137--162. \url{https://doi.org/10.1017/S0007123418000534}
-
 \bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
@@
 Alternatives}: {The Transformation} of {Popular Support} for {Social
-Insurance}. \emph{The Journal of Politics}, \emph{82}(2), 671--686.
+Insurance}. \emph{The Journal of Politics}, \emph{82}(2), 671-686.
 \url{https://doi.org/10.1086/706980}
-
-\bibitem[\citeproctext]{ref-campbell_institutional_2020}
-Campbell, J. L. (2020). \emph{Institutional {Change} and
-{Globalization}}. Princeton University Press.
-\url{https://doi.org/10.2307/j.ctv131bw68}
 
@@
 Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
-(2021). Sujeto y cotidiano en la era neoliberal: El caso de la
-educaci{ó}n chilena. \emph{Revista Mexicana de Ciencias Pol{í}ticas y
+(2021). Sujeto y Cotidiano En La Era Neoliberal: El Caso de La
+Educaci{ó}n Chilena. \emph{Revista Mexicana de Ciencias Pol{í}ticas y
 Sociales}, \emph{67}(244).
@@
 
-\bibitem[\citeproctext]{ref-castillo_legitimacy_2011}
-Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
-Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
-Research}, \emph{24}(4), 314--340.
-\url{https://doi.org/10.1007/s11211-011-0144-5}
-
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
-(2023). A {Multidimensional Approach} for {Measuring Meritocratic
-Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
-Social Inequality Survey}. \emph{International Journal of Sociology},
-1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
-
-\bibitem[\citeproctext]{ref-castillo_socialization_2024}
-Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
-{Socialization} of {Meritocracy} and {Market Justice Preferences} at
-{School}. \emph{Societies}, \emph{14}(11), 214.
-\url{https://doi.org/10.3390/soc14110214}
-
-\bibitem[\citeproctext]{ref-castillo_meritocracia_2019}
-Castillo, J. C., Torres, A., Atria, J., \& Maldonado, L. (2019).
-Meritocracia y desigualdad econ{ó}mica: {Percepciones}, preferencias e
-implicancias. \emph{Revista Internacional de Sociolog{í}a},
-\emph{77}(1), 117. \url{https://doi.org/10.3989/ris.2019.77.1.17.114}
-
-\bibitem[\citeproctext]{ref-chancel_world_2022}
-Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
-inequality report 2022.
-https://bibliotecadigital.ccb.org.co/handle/11520/27510.
-
-\bibitem[\citeproctext]{ref-davis_principles_2001}
-Davis, K., \& Moore, W. E. (2001). Some {Principles} of
-{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
-{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
-Routledge.
+\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
+Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
+Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
+{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
+\emph{Under Review at Frontiers in Sociology}.
 
@@
 \emph{Social Psychological and Personality Science}, \emph{8}(3),
-267--274. \url{https://doi.org/10.1177/1948550616678454}
-
-\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
-Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
-movilidad social bajo un}.
-
-\bibitem[\citeproctext]{ref-eyles_social_2022}
-Eyles, A., Major, L. E., \& Machin, S. (2022). \emph{Social {Mobility} -
-{Past}, {Present} and {Future}: {The State} of {Play} in {Social
-Mobility}, on the 25th {Anniversary} of the {Sutton Trust}}. Sutton
-Trust.
+267-274. \url{https://doi.org/10.1177/1948550616678454}
 
 \bibitem[\citeproctext]{ref-ferre_welfare_2023}
-Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
+Ferre, J. C. (2023). Welfare Regimes in Twenty-First-Century {Latin
 America}. \emph{Journal of International and Comparative Social Policy},
-\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}
+\emph{39}(2), 101-127. \url{https://doi.org/10.1017/ics.2023.16}
 
@@
 in {Chile}: {A Historical Perspective} on {Income Inequality},
-1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
+1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850-874.
 \url{https://doi.org/10.1111/roiw.12441}
@@
 Gingrich, J. R. (2011). \emph{Making {Markets} in the {Welfare State}:
-{The Politics} of {Varying Market Reforms}} (1st ed.). Cambridge
+{The Politics} of {Varying Market Reforms}} (1.ª ed.). Cambridge
 University Press. \url{https://doi.org/10.1017/CBO9780511791529}
 
-\bibitem[\citeproctext]{ref-gugushvili_trends_2014}
-Gugushvili, A. (2014). Trends, {Covariates} and {Consequences} of
-{Intergenerational Social Mobility} in {Post- Socialist Societies}.
-
 \bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016c}
-Gugushvili, A. (2016a). Intergenerational objective and subjective
-mobility and attitudes towards income differences: Evidence from
-transition societies. \emph{Journal of International and Comparative
-Social Policy}, \emph{32}(3), 199--219.
+Gugushvili, A. (2016a). Intergenerational Objective and Subjective
+Mobility and Attitudes towards Income Differences: Evidence from
+Transition Societies. \emph{Journal of International and Comparative
+Social Policy}, \emph{32}(3), 199-219.
 \url{https://doi.org/10.1080/21699763.2016.1206482}
@@
 Explanations} of {Poverty}: {A Comparative Perspective}. \emph{Social
-Justice Research}, \emph{29}(4), 402--428.
+Justice Research}, \emph{29}(4), 402-428.
 \url{https://doi.org/10.1007/s11211-016-0275-9}
@@
 
-\bibitem[\citeproctext]{ref-gugushvili_heterogeneous_2024}
-Gugushvili, A. (2024). The heterogeneous well-being effects of
-intergenerational mobility perceptions. \emph{Journal of Health
-Psychology}, \emph{29}(2), 99--112.
-\url{https://doi.org/10.1177/13591053231187345}
-
-\bibitem[\citeproctext]{ref-hadjar_does_2015}
-Hadjar, A., \& Samuel, R. (2015). Does upward social mobility increase
-life satisfaction? {A} longitudinal analysis using {British} and {Swiss}
-panel data. \emph{Research in Social Stratification and Mobility},
-\emph{39}, 48--58. \url{https://doi.org/10.1016/j.rssm.2014.12.002}
-
-\bibitem[\citeproctext]{ref-harvey_breve_2015}
-Harvey, D. (2015). \emph{{Breve historia del neoliberalismo}}. Madrid
-(Espa{ñ}a): Ediciones Akal.
-
-\bibitem[\citeproctext]{ref-helgason_longterm_2023}
-Helgason, A. F., \& Rehm, P. (2023). Long-term income trajectories and
-the evolution of political attitudes. \emph{European Journal of
-Political Research}, \emph{62}(1), 264--284.
-\url{https://doi.org/10.1111/1475-6765.12506}
-
 \bibitem[\citeproctext]{ref-helgason_class_2025}
-Helgason, A. F., \& Rehm, P. (2025). Class experiences and the long-term
-evolution of economic values. \emph{Social Forces}, \emph{103}(3),
-1125--1143. \url{https://doi.org/10.1093/sf/soae135}
-
-\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
-Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
-D. (2023). Mindsets of poverty: {Implications} for redistributive policy
-support. \emph{Analyses of Social Issues and Public Policy},
-\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
-
-\bibitem[\citeproctext]{ref-immergut_theoretical_1998}
-Immergut, E. M. (1998). The {Theoretical Core} of the {New
-Institutionalism}. \emph{Politics \& Society}, \emph{26}(1), 5--34.
-\url{https://doi.org/10.1177/0032329298026001002}
+Helgason, A. F., \& Rehm, P. (2025). Class Experiences and the Long-Term
+Evolution of Economic Values. \emph{Social Forces}, \emph{103}(3),
+1125-1143. \url{https://doi.org/10.1093/sf/soae135}
+
+\bibitem[\citeproctext]{ref-imai_identification_2010}
+Imai, K., Keele, L., \& Yamamoto, T. (2010). Identification, {Inference}
+and {Sensitivity Analysis} for {Causal Mediation Effects}.
+\emph{Statistical Science}, \emph{25}(1).
+\url{https://doi.org/10.1214/10-STS321}
 
 \bibitem[\citeproctext]{ref-immergut_it_2020}
-Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
-affluent to be able to purchase {``better''} healthcare? {Existential}
-standards and institutional norms in healthcare attitudes across 28
-countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
+Immergut, E. M., \& Schneider, S. M. (2020). Is It Unfair for the
+Affluent to Be Able to Purchase {«Better»} Healthcare? {Existential}
+Standards and Institutional Norms in Healthcare Attitudes across 28
+Countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
 \url{https://doi.org/10.1016/j.socscimed.2020.113146}
@@
 \bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
-Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social mobility
-and demand for redistribution in {Europe}: A comparative analysis.
-\emph{The British Journal of Sociology}, \emph{70}(1), 138--165.
+Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social Mobility
+and Demand for Redistribution in {Europe}: A Comparative Analysis.
+\emph{The British Journal of Sociology}, \emph{70}(1), 138-165.
 \url{https://doi.org/10.1111/1468-4446.12363}
-
-\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
-Janmaat, J. G. (2013). Subjective {Inequality}: A {Review} of
-{International Comparative Studies} on {People}'s {Views} about
-{Inequality}. \emph{European Journal of Sociology}, \emph{54}(3),
-357--389. \url{https://doi.org/10.1017/S0003975613000209}
-
-\bibitem[\citeproctext]{ref-jasso_justice_1978}
-Jasso, Guillermina. (1978). On the {Justice} of {Earnings}: {A New
-Specification} of the {Justice Evaluation Function}. \emph{American
-Journal of Sociology}, \emph{83}(6), 1398--1419.
-\url{https://doi.org/10.1086/226706}
-
-\bibitem[\citeproctext]{ref-jasso_distributive_2016}
-Jasso, Guillermina, Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
-{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
-Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
-\url{https://doi.org/10.1007/978-1-4939-3216-0_11}
-
-\bibitem[\citeproctext]{ref-jasso_gender_1999}
-Jasso, G., \& Wegener, B. (1999). Gender and {Country Differences} in
-the {Sense} of {Justice}: {Justice Evaluation}, {Gender Earnings Gap},
-and {Earnings Functions} in {Thirteen Countries}. \emph{International
-Journal of Comparative Sociology}, \emph{40}(1), 94--115.
-\url{https://doi.org/10.1177/002071529904000106}
 
@@
 about {Market Justice}, 1991-1996. \emph{European Sociological Review},
-\emph{15}(3), 251--283. Retrieved from
+\emph{15}(3), 251-283. Recuperado de
 \url{https://www.jstor.org/stable/522731}
 
-\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
-Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
-\emph{Annual Review of Sociology}, 29--56.
-
 \bibitem[\citeproctext]{ref-koos_moral_2019}
-Koos, S., \& Sachweh, P. (2019). The moral economies of market
-societies: Popular attitudes towards market competition, redistribution
-and reciprocity in comparative perspective. \emph{Socio-Economic
-Review}, \emph{17}(4), 793--821.
-\url{https://doi.org/10.1093/ser/mwx045}
+Koos, S., \& Sachweh, P. (2019). The Moral Economies of Market
+Societies: Popular Attitudes towards Market Competition, Redistribution
+and Reciprocity in Comparative Perspective. \emph{Socio-Economic
+Review}, \emph{17}(4), 793-821. \url{https://doi.org/10.1093/ser/mwx045}
 
@@
 Lane, R. E. (1986). Market {Justice}, {Political Justice}.
-\emph{American Political Science Review}, \emph{80}(2), 383--402.
+\emph{American Political Science Review}, \emph{80}(2), 383-402.
 \url{https://doi.org/10.2307/1958264}
 
-\bibitem[\citeproctext]{ref-langsaether_explaining_2022}
-Langsæther, P. E., Evans, G., \& O'Grady, T. (2022). Explaining the
-{Relationship Between Class Position} and {Political Preferences}: {A
-Long-Term Panel Analysis} of {Intra-Generational Class Mobility}.
-\emph{British Journal of Political Science}, \emph{52}(2), 958--967.
-\url{https://doi.org/10.1017/S0007123420000599}
-
 \bibitem[\citeproctext]{ref-lee_fairness_2023}
-Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
-inequality: The effects of self-interest and neoliberal orientations.
+Lee, J.-S., \& Stacey, M. (2023). Fairness Perceptions of Educational
+Inequality: The Effects of Self-Interest and Neoliberal Orientations.
 \emph{The Australian Educational Researcher}.
 \url{https://doi.org/10.1007/s13384-023-00636-6}
-
-\bibitem[\citeproctext]{ref-lenski_status_1954}
-Lenski, G. E. (1954). Status {Crystallization}: {A Non-Vertical
-Dimension} of {Social Status}. \emph{American Sociological Review},
-\emph{19}(4), 405. \url{https://doi.org/10.2307/2087459}
 
@@
 17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
-887--910. \url{https://doi.org/10.1111/spol.12105}
-
-\bibitem[\citeproctext]{ref-lindh_bringing_2023}
-Lindh, A., \& McCall, L. (2023). Bringing the market in: An expanded
-framework for understanding popular responses to economic inequality.
-\emph{Socio-Economic Review}, \emph{21}(2), 1035--1055.
-\url{https://doi.org/10.1093/ser/mwac018}
+887-910. \url{https://doi.org/10.1111/spol.12105}
+
+\bibitem[\citeproctext]{ref-llorca-jana_historia_2021}
+Llorca-Jaña, M., \& Miller, R. M. D. (2021). \emph{{Historia econ{ó}mica
+de Chile desde la independencia}}. Santiago de Chile: RIL editores.
 
@@
 \bibitem[\citeproctext]{ref-madariaga_three_2020}
-Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
-economic policy trajectory in comparative perspective.
-\emph{Contemporary Politics}, \emph{26}(3), 308--329.
+Madariaga, A. (2020). The Three Pillars of Neoliberalism: {Chile}'s
+Economic Policy Trajectory in Comparative Perspective.
+\emph{Contemporary Politics}, \emph{26}(3), 308-329.
 \url{https://doi.org/10.1080/13569775.2020.1735021}
@@
 
-\bibitem[\citeproctext]{ref-mijs_paradox_2019}
-Mijs, J. (2019). The paradox of inequality: Income inequality and belief
-in meritocracy go hand in hand. \emph{Socio-Economic Review},
-\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
-
-\bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
-Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
-{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
-\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}
-
-\bibitem[\citeproctext]{ref-miller_selfserving_1975}
-Miller, D. T., \& Ross, M. (1975). Self-serving biases in the
-attribution of causality: {Fact} or fiction? \emph{Psychological
-Bulletin}, \emph{82}(2), 213--225.
-\url{https://doi.org/10.1037/h0076486}
-
 \bibitem[\citeproctext]{ref-molina_its_2019}
-Molina, M. D., Bucca, M., \& Macy, M. W. (2019). It's not just how the
-game is played, it's whether you win or lose. \emph{SCIENCE ADVANCES}.
+Molina, M. D., Bucca, M., \& Macy, M. W. (2019). It's Not Just How the
+Game Is Played, It's Whether You Win or Lose. \emph{SCIENCE ADVANCES}.
+
+\bibitem[\citeproctext]{ref-nucleodesociologiacontingente_informe_2020}
+Núcleo de Sociología Contingente, {[}NUDESOC{]}. (2020). \emph{Informe
+de Resultados Oficial {Encuesta Zona Cero}}. Santiago de Chile.
 
@@
 {Networks} and {Attitudes Towards Inequality}. \emph{Sociology},
-\emph{58}(4), 851--876. \url{https://doi.org/10.1177/00380385231217625}
-
-\bibitem[\citeproctext]{ref-polanyi_great_1975}
-Polanyi, K. (1975). \emph{The great transformation} (Repr). New York,
-NY: Octagon Books.
-
-\bibitem[\citeproctext]{ref-prag_subjective_2021}
-Präg, P., \& Gugushvili, A. (2021). Subjective social mobility and
-health in {Germany}. \emph{European Societies}, \emph{23}(4), 464--486.
-\url{https://doi.org/10.1080/14616696.2021.1887916}
-
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
-common good?} (First edition). New York: {Farrar, Straus and Giroux}.
-
-\bibitem[\citeproctext]{ref-satz_por_2019}
-Satz, D. (2019). \emph{{Por Qu{é} Algunas Cosas No Deber{í}an Estar en
-Venta: Los l{í}mites Morales Del Mercado}}. Ciudad Aut{ó}noma de Buenos
-Aires: Siglo XXI Editores.
+\emph{58}(4), 851-876. \url{https://doi.org/10.1177/00380385231217625}
+
+\bibitem[\citeproctext]{ref-panes_criticas_2020}
+Panes, D. (2020). \emph{Cr{í}ticas y Experiencias Obreras En Torno al
+Sistema de {AFP}'s En {Chile} (1981-2020)} (Tesis doctoral). Universidad
+de Chile, Santiago.
 
 \bibitem[\citeproctext]{ref-schmidt_experience_2011}
-Schmidt, A. W. (2011). The experience of social mobility and the
-formation of attitudes toward redistribution. In.
+Schmidt, A. W. (2011). The Experience of Social Mobility and the
+Formation of Attitudes toward Redistribution. En.
 
@@
 {Experimental Evidence}. \emph{Perspectives on Psychological Science},
-\emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}
+\emph{11}(3), 373-380. \url{https://doi.org/10.1177/1745691616635596}
 
@@
 Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
-water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
-Movement Studies}, \emph{20}(4), 495--502.
+Water in the Oasis: The {Chilean Spring} of 2019--2020. \emph{Social
+Movement Studies}, \emph{20}(4), 495-502.
 \url{https://doi.org/10.1080/14742837.2020.1727737}
 
-\bibitem[\citeproctext]{ref-sorokin_social_1927}
-Sorokin, P. A. (1927). \emph{Social {Mobility}}. Harper \& Brothers.
-
 \bibitem[\citeproctext]{ref-streeck_how_2016}
-Streeck, W. (2016). \emph{How will capitalism end? Essays on a failing
-system}. London: Verso.
+Streeck, W. (2016). \emph{How Will Capitalism End? Essays on a Failing
+System}. London: Verso.
 
@@
 {Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
-(1st ed.). Stanford University Press.
+(1.ª ed.). Stanford University Press.
 \url{https://doi.org/10.2307/j.ctvr0qv0q}
-
-\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
-Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
-Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
-{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
-\emph{International Review of Social Psychology}, \emph{38}(1), 4.
-\url{https://doi.org/10.5334/irsp.1013}
 
@@
 Latin American Case}. \emph{Annual Review of Sociology}, \emph{40}(1),
-619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
-
-\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
-Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
-care inequalities unfair? {A} study on public attitudes in 23 countries.
-\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
-\url{https://doi.org/10.1186/s12939-016-0350-8}
-
-\bibitem[\citeproctext]{ref-wen_does_2021}
-Wen, F., \& Witteveen, D. (2021). Does perceived social mobility shape
-attitudes toward government and family educational investment?
-\emph{Social Science Research}, \emph{98}, 102579.
-\url{https://doi.org/10.1016/j.ssresearch.2021.102579}
-
-\bibitem[\citeproctext]{ref-wiederkehr_belief_2015}
-Wiederkehr, V., Bonnot, V., Krauth-Gruber, S., \& Darnon, C. (2015).
-Belief in school meritocracy as a system-justifying tool for low status
-students. \emph{Frontiers in Psychology}, \emph{6}.
-
-\bibitem[\citeproctext]{ref-wilson_role_2003}
-Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
-Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
-\url{https://doi.org/10.1023/A:1024667228488}
+619-642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
 
 \bibitem[\citeproctext]{ref-young_rise_1958}
-Young, M. (1958). \emph{The rise of the meritocracy}. New Brunswick,
+Young, M. (1958). \emph{The Rise of the Meritocracy}. New Brunswick,
 N.J., U.S.A: Transaction Publishers.

### 2025-09-04 presentations/coloquio_2025/mjp-mobility.qmd
Up pp coloquio

@@
     mathjax: "default"
-  pdf:
-    format: beamer 
-    keep-tex: true  # Debug LaTeX output if needed
 editor_options: 
@@
 
----
-
-## ¿A dónde quiero llegar?
-
-::: {.fragment}
-
-<div style="text-align:center;">
-  <img src="images/idea.png" style="width:100%; max-width:none;">
-</div>
-
-:::
-
 # Contexto y motivación {.xlarge data-background-color="#523870"}
@@
 
-::: {.box-inv-4 .sp-after .fragment style="font-size: 100%;"}
+::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}
 
@@
 
-::: {.box-inv-5 .sp-after-half .fragment style="font-size: 100%;"}
+::: {.box-inv-5 .sp-after-half .fragment style="font-size: 110%;"}
 
-Preferencias por justicia de mercado [@busemeyer_skills_2014; @immergut_it_2020; @castillo_perceptions_2025; @koos_moral_2019; @lindh_public_2015]
+Preferencias por justicia de mercado [@busemeyer_skills_2014; @castillo_perceptions_2025; @koos_moral_2019; @lindh_public_2015]
 
@@
 
-## Preferencias por justicia de mercado
+## Preferencias por la comodificación de servicios
 
@@
 
-## Preferencias por justicia de mercado: Chile
+## Preferencias por la comodificación de servicios: Chile
 
@@
 
-Cómo la movilidad social afecta estas preferencias sigue sin respuesta
+El movimiento entre posiciones sociales expone a los individuos experiencias y mecanismos que afectan nociones de justicia [@alesina_intergenerational_2018; @ares_changing_2020; @gugushvili_subjective_2017; @day_movin_2017]
+
 :::
 
-</br></br>
+</br></br></br></br>
 
@@
 
-El movimiento entre posiciones expone experiencias y **mecanismos** que afectan nociones de justicia
+Meritocracia como mecanismo de sesgo de atribución[@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
 
@@
 ::::
-
-## Movilidad social
-
-::::: columns
-::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}
-
-### Afecta
-
-- Preferencias redistributivas [@alesina_intergenerational_2018; @ares_changing_2020; @schmidt_experience_2011; @breen_effects_2024]
-- Creencias sobre la desigualdad [@gugushvili_intergenerational_2016; @bucca_merit_2016; @shariff_income_2016; @day_movin_2017]
-- Apoyo al gasto social y políticas de bienestar [@gugushvili_subjective_2017; @gugushvili_intergenerational_2016c]
-
-:::
-
-::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}
-
-### Mecanismos
-
-- Intereses materiales [@helgason_class_2025]
-- Socialización [@jaime-castillo_social_2019]
-- **Self-serving bias** [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
-- Meritocracia como proxy al sesgo de atribuciones internas y externas [@young_rise_1958]
-:::
-:::::
 
@@
 ## Estrategia de identificación causal
+
+```{r}
+#| label: fig-dag
+#| fig-cap-location: top
+#| fig-cap: "DAG movilidad social, meritocracia y preferencias por comodificación"
+#| fig-align: center
+#| out-width: '200%'
+library(dagitty)
+library(ggdag)
+library(ggplot2)
+
+# Coordenadas para ubicar los nodos según el tiempo t1–t4
+coords <- list(
+  x = c(C=-1,   # Confusores pre
+        T= 0,   # Tratamiento
+        L= 1,   # Post-tratamiento
+        M= 1,   # Mediador
+        Y= 2),  # Outcome
+  y = c(C= 0.6,
+        T= 0.9,
+        L=-0.4,
+        M= 0.2,
+        Y= 0.9)
+)
+
+# Definición del DAG
+dag_mjp <- dagify(
+  T ~ C,          # Tratamiento depende de confusores pre
+  L ~ T + C,      # Confusor post-T inducido por T y C
+  M ~ T + C + L,  # Mediador depende de T, C y L
+  Y ~ T + M + C + L,  # Outcome depende de todo lo anterior
+  exposure = "T",
+  outcome  = "Y",
+  coords   = coords
+)
+
+# Pasar a formato tidy
+td <- tidy_dagitty(dag_mjp)
+
+# Asignar tiempos por nodo
+time_map <- c(
+  C = "t1",
+  T = "t2",
+  L = "t3",
+  M = "t3",
+  Y = "t4"
+)
+td$data$time <- unname(time_map[td$data$name])
+
+# Graficar DAG
+ggdag(td, node_size = 18, text_size = 6, edge_size = 1.2) +   # <--- aumenta nodos, texto de nodos y grosor de flechas
+  theme_classic() + 
+  remove_axes() + 
+  theme(axis.line = element_blank(),
+        text = element_text(size = 16)) +                     # <--- tamaño global del texto adicional
+  xlim(-1.5, 2.5) + ylim(-1, 1) +
+
+  # Etiquetas extra
+  geom_text(x=-1, y=-0.45, label="T = Movilidad social\n(upward / downward)", hjust=0, size=5) +
+  geom_text(x=-1, y=-0.70, label="Y = Preferencias por\ncomodificación", hjust=0, size=5) +
+  geom_text(x=-1, y=-0.95, label="M = Creencias\nmeritocráticas", hjust=0, size=5)  +
+  geom_text(x=0.55, y=-0.7, label="C = Confusores pre-tratamiento\n(propensity)", hjust=0, size=5)  +
+  geom_text(x=0.55, y=-0.9, label="L = Confusor post-T", hjust=0, size=5)
+
+```
 
@@
 2.
-**Propensity-score estimation**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles para cada estrato (categ.
-ref.).
-Estas puntuaciones de propensión se emplean posteriormente para ajustar la selección en la movilidad al evaluar su efecto.
+**Propensity-score estimation**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles (categ.
+ref.) para cada estrato.
+Creación de IPW o "clones"
 
@@
 
-## Estimand y estimando
+## Estimand
 
-- Métodos
+::: {.fragment style="font-size: 120%;"}
+
+- Efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran trasladado a un destino alternativo *k′* [@breen_effects_2024] (@eq-att).
+:::
+
+::: {.fragment}
+$$ 
+ATT_{jkk′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
+$$ {#eq-att}
+
+:::
+
+::: {.fragment style="font-size: 120%;"}
+
+- El estimador resultante es el efecto promedio del tratamiento sobre los tratados (ATT) 
+:::
+
+## Estimando
+
+::: {.incremental style="font-size: 120%;"}
+
+- Para estimar el efecto causal de la movilidad social, utilizo modelos de regresión lineal con ponderaciones de probabilidad inversa (IPW) para la movilidad condicional a la clase de origen [@breen_effects_2024]
+
+- Para mediación ocupo el estimador Average Causal Mediation Effect [@imai_identification_2010]
+
+:::
 
@@
   # etiquetas: coef en el punto
-  geom_text(aes(label = est_lab), vjust = -0.8, size = 3.3) +
+  geom_text(aes(label = est_lab), vjust = -0.8, size = 4) +
   scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
@@
 
+```
+
+## Mediación
+
+```{r}
+#| label: fig-acme
+#| fig-cap-location: top
+#| fig-cap: "ACME para meritocracia y preferencias por comodificación en pensiones"
+#| fig-align: center
+#| out-width: '180%'
+
+plot(med.fit)
 ```
@@
 2. **Contexto chileno**: profunda comodificación y privatización de servicios escenciales → trabajador-empresario 
-3. **Mediación**: rol de la meritocracia como mediador → ACME [@imai_identification_2010], moderación (?)
+3. **Mediación**: rol de la meritocracia como mediador → ACME supuestos muy fuertes =  alternativas (?)
 :::

### 2025-09-04 presentations/coloquio_2025/mjp-mobility.tex
Up pp coloquio

@@
 
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\subsection{¿A dónde quiero llegar?}\label{a-duxf3nde-quiero-llegar}
-
 \section{Contexto y motivación}\label{contexto-y-motivaciuxf3n}
@@
 Preferencias por justicia de mercado (Busemeyer, 2014; Castillo et~al.,
-2025; Immergut \& Schneider, 2020; Koos \& Sachweh, 2019; Lindh, 2015)
+2025; Koos \& Sachweh, 2019; Lindh, 2015)
 
@@
 
-\subsection{Preferencias por justicia de
-mercado}\label{preferencias-por-justicia-de-mercado-1}
+\subsection{Preferencias por la comodificación de
+servicios}\label{preferencias-por-la-comodificaciuxf3n-de-servicios}
 
@@
 
-\subsection{Preferencias por justicia de mercado:
-Chile}\label{preferencias-por-justicia-de-mercado-chile}
+\subsection{Preferencias por la comodificación de servicios:
+Chile}\label{preferencias-por-la-comodificaciuxf3n-de-servicios-chile}
 
@@
 
-Cómo la movilidad social afecta estas preferencias sigue sin respuesta
+El movimiento entre posiciones sociales expone a los individuos
+experiencias y mecanismos que afectan nociones de justicia (Alesina
+et~al., 2018; Ares, 2020; Day \& Fiske, 2017; Gugushvili, 2017)
 
-El movimiento entre posiciones expone experiencias y \textbf{mecanismos}
-que afectan nociones de justicia
+Meritocracia como mecanismo de sesgo de atribución(Gugushvili, 2016;
+Molina et~al., 2019; Schmidt, 2011).
 
@@
 relación en el contexto chileno?}}
-
-\subsection{Movilidad social}\label{movilidad-social}
-
-\subsubsection{Afecta}\label{afecta}
-
-\begin{itemize}
-\tightlist
-\item
-  Preferencias redistributivas (Alesina et~al., 2018; Ares, 2020; Breen
-  \& Ermisch, 2024; Schmidt, 2011)
-\item
-  Creencias sobre la desigualdad (Bucca, 2016; Day \& Fiske, 2017;
-  Gugushvili, 2016b; Shariff et~al., 2016)
-\item
-  Apoyo al gasto social y políticas de bienestar (Gugushvili, 2016a,
-  2017)
-\end{itemize}
-
-\subsubsection{Mecanismos}\label{mecanismos}
-
-\begin{itemize}
-\tightlist
-\item
-  Intereses materiales (Helgason \& Rehm, 2025)
-\item
-  Socialización (Jaime-Castillo \& Marqués-Perales, 2019)
-\item
-  \textbf{Self-serving bias} (Gugushvili, 2016a; Molina et~al., 2019;
-  Schmidt, 2011).
-\item
-  Meritocracia como proxy al sesgo de atribuciones internas y externas
-  (Young, 1958)
-\end{itemize}
 
@@
 causal}\label{estrategia-de-identificaciuxf3n-causal}
+
+\begin{figure}
+
+\caption{\label{fig-dag}DAG movilidad social, meritocracia y
+preferencias por comodificación}
+
+\centering{
+
+\includegraphics[width=2\linewidth,height=\textheight,keepaspectratio]{mjp-mobility_files/figure-pdf/fig-dag-1.pdf}
+
+}
+
+\end{figure}%
 
@@
   movilidad social mediante un modelo logístico multinomial en relación
-  con las trayectorias inmóviles para cada estrato (categ. ref.). Estas
-  puntuaciones de propensión se emplean posteriormente para ajustar la
-  selección en la movilidad al evaluar su efecto.
+  con las trayectorias inmóviles (categ. ref.) para cada estrato.
+  Creación de IPW o ``clones''
 \end{enumerate}
@@
 
-\subsection{Estimand y estimando}\label{estimand-y-estimando}
+\subsection{Estimand}\label{estimand}
 
@@
 \item
-  Métodos
+  Efecto causal promedio de pasar de una clase de origen \emph{j} a una
+  clase de destino \emph{k} comparando a las personas que realmente
+  alcanzan \emph{k} con su resultado contrafactual si se hubieran
+  trasladado a un destino alternativo \emph{k′} (Breen \& Ermisch, 2024)
+  (Ecuación~\ref{eq-att}).
+\end{itemize}
+
+\begin{equation}\phantomsection\label{eq-att}{ 
+ATT_{jkk′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
+}\end{equation}
+
+\begin{itemize}
+\tightlist
+\item
+  El estimador resultante es el efecto promedio del tratamiento sobre
+  los tratados (ATT)
+\end{itemize}
+
+\subsection{Estimando}\label{estimando}
+
+\begin{itemize}
+\item
+  Para estimar el efecto causal de la movilidad social, utilizo modelos
+  de regresión lineal con ponderaciones de probabilidad inversa (IPW)
+  para la movilidad condicional a la clase de origen (Breen \& Ermisch,
+  2024)
+\item
+  Para mediación ocupo el estimador Average Causal Mediation Effect
+  (Imai et~al., 2010)
 \end{itemize}
@@
 
+\subsection{Mediación}\label{mediaciuxf3n}
+
+\begin{figure}
+
+\caption{\label{fig-acme}ACME para meritocracia y preferencias por
+comodificación en pensiones}
+
+\centering{
+
+\includegraphics[width=1.8\linewidth,height=\textheight,keepaspectratio]{mjp-mobility_files/figure-pdf/fig-acme-1.pdf}
+
+}
+
+\end{figure}%
+
 \section{Discusión y proyecciones}\label{discusiuxf3n-y-proyecciones}
@@
 \item
-  \textbf{Mediación}: rol de la meritocracia como mediador → ACME (Imai
-  et~al., 2010), moderación (?)
+  \textbf{Mediación}: rol de la meritocracia como mediador → ACME
+  supuestos muy fuertes = alternativas (?)
 \end{enumerate}
@@
 \url{https://doi.org/10.15195/v11.a17}
-
-\bibitem[\citeproctext]{ref-bucca_merit_2016}
-Bucca, M. (2016). Merit and Blame in Unequal Societies: {Explaining
-Latin Americans}' Beliefs about Wealth and Poverty. \emph{Research in
-Social Stratification and Mobility}, \emph{44}, 98-112.
-\url{https://doi.org/10.1016/j.rssm.2016.02.005}
 
@@
 \bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016c}
-Gugushvili, A. (2016a). Intergenerational Objective and Subjective
+Gugushvili, A. (2016). Intergenerational Objective and Subjective
 Mobility and Attitudes towards Income Differences: Evidence from
@@
 
-\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016}
-Gugushvili, A. (2016b). Intergenerational {Social Mobility} and {Popular
-Explanations} of {Poverty}: {A Comparative Perspective}. \emph{Social
-Justice Research}, \emph{29}(4), 402-428.
-\url{https://doi.org/10.1007/s11211-016-0275-9}
-
 \bibitem[\citeproctext]{ref-gugushvili_subjective_2017}
@@
 {Support} for {Welfare State Programmes}.
-
-\bibitem[\citeproctext]{ref-helgason_class_2025}
-Helgason, A. F., \& Rehm, P. (2025). Class Experiences and the Long-Term
-Evolution of Economic Values. \emph{Social Forces}, \emph{103}(3),
-1125-1143. \url{https://doi.org/10.1093/sf/soae135}
 
@@
 \url{https://doi.org/10.1016/j.socscimed.2020.113146}
-
-\bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
-Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social Mobility
-and Demand for Redistribution in {Europe}: A Comparative Analysis.
-\emph{The British Journal of Sociology}, \emph{70}(1), 138-165.
-\url{https://doi.org/10.1111/1468-4446.12363}
 
@@
 
-\bibitem[\citeproctext]{ref-shariff_income_2016}
-Shariff, A. F., Wiwad, D., \& Aknin, L. B. (2016). Income {Mobility
-Breeds Tolerance} for {Income Inequality}: {Cross-National} and
-{Experimental Evidence}. \emph{Perspectives on Psychological Science},
-\emph{11}(3), 373-380. \url{https://doi.org/10.1177/1745691616635596}
-
 \bibitem[\citeproctext]{ref-somma_no_2021}

### 2025-09-04 presentations/coloquio_2025/mjp-mobility.qmd
Up VF pp coloquio

@@
 
-Meritocracia como mecanismo de sesgo de atribución[@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
+Meritocracia como mecanismo de sesgo de atribución[@newman_economic_2023; @gugushvili_intergenerational_2016c; @molina_its_2019].
 
@@
 # Método {.xlarge data-background-color="#523870"}
+
+## Enfoque: Mobility Effects
+
+::::: columns
+::: {.column width="50%"}
+::: {.fragment}
+
+![](images/breen.png)
+
+:::
+
+:::
+
+::: {.column width="50%" .incremental .highlight-last style="font-size: 100%;"}
+
+- @breen_effects_2024 proponen un método para estimar el efecto (causal) de la movilidad social intergernacional sobre un outcome
+- ¿En qué habría cambiado el resultado _Y_ entre las personas de origen _j_ que llegaron al destino _k_ si, hipotéticamente, hubieran llegado al destino _k'_?
+
+:::
+:::::
 
@@
 2.
-**Propensity-score estimation**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles (categ.
+**Propensity-score weighting**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles (categ.
 ref.) para cada estrato.
@@
 
+::: {.fragment style="font-size: 120%;"}
+
+- Mi interés es comparar con la situación de inmobilidad: cuando _k'_ = _j_
+:::
+
 ## Estimando
@@
 
-- Para estimar el efecto causal de la movilidad social, utilizo modelos de regresión lineal con ponderaciones de probabilidad inversa (IPW) para la movilidad condicional a la clase de origen [@breen_effects_2024]
+- Efecto movilidad: modelos de regresión lineal con ponderaciones de probabilidad inversa (IPW) para la movilidad condicional a la clase de origen [@breen_effects_2024] y SE clusterizados por id
 
-- Para mediación ocupo el estimador Average Causal Mediation Effect [@imai_identification_2010]
+- Mediación: Average Causal Mediation Effect [@imai_identification_2010]
 
@@
 #| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de mobilidad en preferencia por comodificación en salud"
+#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación en salud"
 #| fig-align: center
@@
 ggplot(df_health_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 0.8) +
   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
@@
 #| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de mobilidad en preferencia por comodificación en educación"
+#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación en educación"
 #| fig-align: center
@@
 ggplot(df_educ_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 0.8) +
   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
@@
 #| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de mobilidad en preferencia por comodificación en pensiones"
+#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación en pensiones"
 #| fig-align: center
@@
 ggplot(df_pension_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkorange", size = 0.5, alpha = 0.8) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 0.8) +
   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +

### 2025-09-24 presentations/coloquio_2025/educ_estrati.qmd
Up estrati

---
bibliography: "input/bib/tesis-magister.bib"
csl: "input/bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
  pdf:
    format: beamer 
    keep-tex: true  # Debug LaTeX output if needed
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(knitr)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

#pacman::p_load(tidyverse, 
#               sjmisc, 
#               sjPlot, 
##               sjlabelled, 
#               here,
#               kableExtra,
#               ggdist,
#               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data
#| include: false

```

</br>

::: columns

::: {.column width="20%"}

![](images/logo_isuc.png)

</br></br>

![](images/coes.png)

:::

::: {.column .column-right width="80%"}

# **El rol de la educación en la movilidad social**

------------------------------------------------------------------------

**Andreas Laffert^1,^^2^**

::: {.blue2 .medium}

**^1^Instituto de Sociología, Pontificia Universidad Católica de Chile**

**^2^Centro de Estudios de Conflicto y Cohesión Social - COES**

:::

Seminario Estratificación Social - SOL3016

24 Septiembre 2025, Santiago

:::
:::

---
## Textos

::: {.fragment}

<div style="text-align:center;">
  <img src="images/collage.png" style="width:100%; max-width:none;">
</div>

:::

# Antecedentes {.xlarge data-background-color="#523870"}

## Antecedentes: O-E-D

::: {.fragment}

<div style="text-align:center;">
  <img src="images/oed1.png" style="width:100%; max-width:none;">
</div>

:::

## Antecedentes: O-E-D

::: {.fragment}

<div style="text-align:center;">
  <img src="images/goldt.png" style="width:80%; max-width:none;">
</div>

:::

## Antecedentes:  O-E-D

::: {.incremental .highlight-last style="font-size: 120%;"}

- Es indiscutible que los hijos de familias aventajadas tienden a obtener destinos aventajados; transmición intergeneracional de ventajas/desventajas

- El efecto O-D es mucho más débil entre los más educados, i.e., la movilidad es mayor entre los más educados (Chetty et al. 2017; Hout, 1988; Torche, 2011)

- Hout (1988): Contar con un college degree de hecho "anula" el OD y, en consecuencia, la expansión educativa promueve la movilidad social total → poder meritocrático del degree

- Los hijos de familias de bajos ingresos tienen menos probabilidades de terminar la universidad, pero cuando lo logran suelen beneficiarse más que sus pares de familias acomodadas [@zhou_equalization_2019]

:::

## Antecedentes: la promesa meritocrática

::: {.incremental .highlight-last style="font-size: 120%;"}

- Meritocracia: _sistema_ donde los bienes y recompensas se distribuyen en base al mérito (esfuerzo + talento) [@young_rise_1958]

- Modernización e igualdad de oportunidades [@dubet_repensar_2011]

- Las personas pueden seguir creyendo normativamente en la meritocracia incluso ante la evidencia de que las oportunidades están distribuidas de manera desigual [@mijs_paradox_2019; @mijs_belief_2022]

- La meritocracia como una promesa incumplible en las sociedades modernas que permite la perpetuación de las desigualdades sociales [@mijs_unfulfillable_2016; @sandel_tyranny_2020]

:::

## Ejes de análisis

::: {style="font-size: 110%;"}
Las tres investigaciones buscan analizar empiricamente dos hipótesis rivales:
:::
::: {.incremental .highlight-last style="font-size: 110%;"}

1. **Equalization**: la expansión educacional promueve la movilidad social [@vandewerfhorst_meritocracy_2024]

    * Vía indirecta (O→E→D): al reducir la desigualdad de oportunidades educativas (el vínculo O-E) o al disminuir los retornos a la educación (el vínculo E-D)
    * Vía directa (O→D|E): si la movilidad es mayor entre los más educados, el hecho de que más gente alcance niveles educativos altos (expansión) reducirá la persistencia total en la sociedad (efecto de composición)

2. **Reproduction**: aunque expansión debilita la vía indirecta (O→E→D), las familias con recursos compensarán esta pérdida fortaleciendo la vía directa (O→D|E); reproducción _dentro_ de ciertos grupos usando nuevas formas [@zhou_equalization_2019; @witteveen_reconsidering_2020]

:::

# Preguntas {.xlarge data-background-color="#523870"}

:::: {.incremental style="font-size: 150%;"}

**_(1) ¿La educación permite mitigar la transmisión intergeneracional de ventajas y desventajas?_**

**_(2) ¿La educación promueve la movilidad social en general o para algunos?_**

::::

## @witteveen_reconsidering_2020

:::: {style="font-size: 110%;"}

::: {.incremental .highlight-last}

- **Objetivo**:  determinar si la asociación entre el origen familiar y el destino económico persiste entre los graduados universitarios en Estados Unidos

- **Datos**: National Survey of College Graduates (NSCG) de 1993, 2003 y 2015
- **Mediciones**:

    * Destino (VD): ingreso anual del hijo
    * Origen (VI): nivel educativo más alto alcanzado por cualquiera de los padres (desde menos de secundaria hasta doctorado)
    * Educación (M): nivel educativo del hijo
    * Controles: edad, género, raza/etnia y si el encuestado nació en el extranjero

- **Estrategia**: regresiones binomiales negativas (NB).  Analizan si el propio nivel educativo del hijo/a (licenciatura vs. posgrado) media la relación

:::
::::

## @witteveen_reconsidering_2020

:::: {style="font-size: 110%;"}
::: {.incremental .highlight-last}

- **Hallazgos**:

    - Contrario a la tesis de la igualación, encontraron una asociación "sustancial y persistente" entre la educación de los padres y los ingresos de sus hijos después de la universidad
    - La influencia de O persiste tanto para quienes solo obtienen una licenciatura como para quienes alcanzan títulos avanzados. Este efecto no disminuye en las cohortes más jóvenes
    - Factores como la selectividad de la universidad a la que se asiste solo median parcialmente la asociación OED

- **Contribuciones**:

    - Refuta empíricamente la "tesis de la igualación" en el contexto de Estados Unidos

:::
::::

## @vandewerfhorst_meritocracy_2024

:::: {style="font-size: 110%;"}
::: {.incremental .highlight-last}

- **Objetivo**: determinar si la expansión educativa conduce a una mayor movilidad ocupacional y si esta relación es contrarrestada por un fortalecimiento de la asociación O-D dentro de los mismos niveles educativos (críticas a la meritocracia)

- **Datos**: múltiples encuestas internacionales (ESS e ISSP) y cohortes
- **Mediciones**:

    * Movilidad (O-D): ridit scores por país-cohorte del ISEI del hijo/a y padre
    * Educación: tres categorías secundaria baja o menos, secundaria alta, y título universitario o más
    * Contextuales: expansión educativa y políticas que la promueven

- **Estrategia**: enfoque de descomposición en dos pasos inspirado en Bloome et al. (2018)

:::
::::

## @vandewerfhorst_meritocracy_2024

:::: {style="font-size: 110%;"}
::: {.incremental .highlight-last}

- **Hallazgos**:

    - El path OD es más fuerte entre los menos educados y más débil entre los más educados
    - Si lo anterior es cierto, la expansión educacional reducirá la persistencia intergeneracional total
    - La expansión educativa se asocia a una mayor movilidad ocupacional; expansión afecta negativamente la vía índirecta (O-E-D) y directa (OD|E)
    - No hay evidencia de que las familias de élite compensen la mayor igualdad de oportunidades educativas fortaleciendo la vía directa (OD|E)

- **Contribuciones**:

    - OD|E se reduce a medida que la educación superior se expande
    - Evidencia favorable al rol igualador de la educación como motor de la movilidad ocupacional a nivel macro

:::
::::

## @zhou_equalization_2019

:::: {style="font-size: 110%;"}

::: {.incremental .highlight-last}

- **Objetivo**: distinguir entre dos explicaciones para la alta movilidad intergeneracional observada entre los graduados universitarios en EE. UU.: la "ecualización" y la "selección"

- **Datos**: National Longitudinal Survey of Youth 1979
- **Mediciones**:

    * Movilidad (O-D): rank-rank income mobility padres e hijos
    * Educación: graduado universitario vs. no graduado. También se distingue entre graduados de universidades selectivas y no selectivas
    * Confusores Z: conjunto de características pre-universidad

- **Estrategia**: enfoque causal para distinguir entre movilidad condicional (unweighted) y movilidad controlada (reweighted) que corrige sesgo de selección

:::
::::

## @zhou_equalization_2019

:::: {style="font-size: 110%;"}
::: {.incremental .highlight-last}

- **Hallazgos**:

    - Condicionar por los college degree (grupo) induce un sesgo de selección (collider bias) en la estimación del efecto de la educación  en la movilidad social
    - La movilidad condicional es mayor entre los más educados en comparación a los menos educados
    - Una vez que la selección es ajustada, la movilidad intergeneracional por ingresos entre los graduados es muy similar a la de los no graduados

- **Contribuciones**:

    - La selección explica la movilidad: la movilidad entre universitarios se debe a selección, no a un efecto igualador real del título.
    - La mayor movilidad observada refleja que llegan al grupo jóvenes de origen bajo positivamente seleccionados (habilidad, motivación).
    - Evidencia consistente con la reproducción / “effectively maintained inequality”

:::
::::

## Diálogo entre las investigaciones

::: {.incremental .highlight-last style="font-size: 120%;"}

- ¿Ecualización o reproducción? → Apoyo a [@vandewerfhorst_meritocracy_2024] vs. críticas de [@zhou_equalization_2019; @witteveen_reconsidering_2020].

- @vandewerfhorst_meritocracy_2024 asume el patrón composicional sin modelar selección; @zhou_equalization_2019 cuestiona esa premisa.

- El efecto de la expansión educativa sobre OD puede ser en realidad selección [@zhou_equalization_2019].

- Desmantelando el O–E–D: descomposición en @vandewerfhorst_meritocracy_2024, vía directa en @witteveen_reconsidering_2020, y advertencia de @zhou_equalization_2019 sobre sesgos si no se ajusta por características previas.

:::

::: {.notes}

@vandewerfhorst_meritocracy_2024 no modela la selección a nivel individual. Su análisis macro asume que la menor persistencia observada entre los más educados es un patrón real y lo usa como base para su "hipótesis composicional". El trabajo de @zhou_equalization_2019 socava directamente esta premisa.

:::

## Discusión

::: {.incremental .highlight-last style="font-size: 120%;"}

1. Hallazgos dependen largamente del diseño métodologico: Macro vs. micro, decomposición vs. selección, movilidad ocupacional vs. ingresos

2. Implicaciones políticas: Si queremos más movilidad, ¿qué debemos hacer?

:::

# Gracias por su atención!

## Referencias

### 2025-09-24 modelos_mjp.R

@@ -1,0 +1,646 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data analysis for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               sjmisc,
+               PSweight,
+               texreg,
+               estimatr,
+               mediation,
+               conflicted)
+
+conflict_prefer("select", "dplyr")
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(here("input/data/proc/data_v1.RData"))
+
+# 3. Analysis -----------------------------------------------------------
+
+# 3.1 Regresiones con IPW -----------------------------------------------------------
+
+### Pensiones
+
+### Low to middle
+
+data <- datos1 %>% 
+  select(idencuesta, 
+         ola, 
+         y = just_pension, 
+         t = moblm, 
+         w = wlow_orig1, 
+         edad = m0_edad, 
+         sexo, 
+         nac = nacionalidad,
+         etnia,
+         hogar = hogar_bip_f, 
+         edu0 = educ_orig_f) %>% 
+  na.omit() %>% 
+  mutate(w = w / mean(w)) 
+
+data
+
+mp_lm <- data %>% 
+  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
+            weights = w,
+            se_type = "CR2",
+            clusters = idencuesta,
+            data = .) 
+
+mp_lm %>%  
+  screenreg()
+
+### Low to high
+
+data <- datos1 %>% 
+  select(idencuesta, 
+         ola, 
+         y = just_pension, 
+         t = moblh, 
+         w = wlow_orig2, 
+         edad = m0_edad, 
+         sexo, 
+         nac = nacionalidad,
+         etnia,
+         hogar = hogar_bip_f, 
[…]

### 2025-09-24 processing/prod_analysis.R

@@ -210,363 +210,6 @@ mp_hl <- data %>%
 mp_hl %>% 
   screenreg()
 
-
-### Salud
-
-### Low to middle
-
-data <- datos1 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_health, 
-         t = moblm, 
-         w = wlow_orig1, 
-         edad = m0_edad, 
-         sexo, 
-         nac = nacionalidad,
-         etnia,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  na.omit() %>% 
-  mutate(w = w / mean(w)) 
-
-data
-
-mh_lm <-data %>% 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = .) 
-    
-mh_lm %>% 
-  screenreg()
-
-### Low to high
-
-data <- datos1 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_health, 
-         t = moblh, 
-         w = wlow_orig2, 
-         edad = m0_edad, 
-         sexo, 
-         nac = nacionalidad,
-         etnia,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  na.omit() %>% 
-  mutate(w = w / mean(w)) 
-
-data
-
-mh_lh <- data %>% 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = .) 
-
-mh_lh %>% 
-  screenreg()
-
-### middle to high
-
-data <- datos2 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_health, 
-         t = mobmh, 
-         w = wm_mh, 
-         edad = m0_edad, 
-         sexo, 
-         nac = nacionalidad,
-         etnia,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  na.omit() %>% 
-  mutate(w = w / mean(w)) 
-
-data
-
-mh_mh <- data %>% 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = .) 
-        
-mh_mh %>% 
-  screenreg()
-
-
-### middle to low
-
[…]

### 2025-09-24 processing/prod_prep.R

@@ -211,9 +211,6 @@ db <- db %>%
     )
   )
 
-db$merit_i <- rowMeans(db[, c("merit_effort", "merit_talent")], na.rm = TRUE)
-
-
 # Crear estrato_ocupa basado en rangos de isei_ocupa
 
 #db <- db %>% 

### 2025-10-01 presentations/coloquio_2025/gender_estrati.qmd
Up ppt estrati gender

---
bibliography: "input/bib/tesis-magister.bib"
csl: "input/bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
  pdf:
    format: beamer 
    keep-tex: true  # Debug LaTeX output if needed
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(knitr)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

#pacman::p_load(tidyverse, 
#               sjmisc, 
#               sjPlot, 
##               sjlabelled, 
#               here,
#               kableExtra,
#               ggdist,
#               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data
#| include: false

```

</br>

::: columns

::: {.column width="20%"}

![](images/logo_isuc.png)

</br></br>

![](images/coes.png)

:::

::: {.column .column-right width="80%"}

# **Penalidad por maternidad en los ingresos: evidencia y métodos**

------------------------------------------------------------------------

**Andreas Laffert^1,^^2^**

::: {.blue2 .medium}

**^1^Instituto de Sociología, Pontificia Universidad Católica de Chile**

**^2^Centro de Estudios de Conflicto y Cohesión Social - COES**

:::

Seminario Estratificación Social - SOL3016

1 Octubre 2025, Santiago

:::
:::

---
## Textos

::: {.fragment}

<div style="text-align:center;">
  <img src="images/collage(1).png" style="width:100%; max-width:none;">
</div>

:::

# Antecedentes {.xlarge data-background-color="#523870"}

## Antecedentes

::: {.incremental .highlight-last style="font-size: 120%;"}

- La literatura muestra una penalidad por maternidad: las madres suelen ganar menos que las mujeres sin hijos [@budig_wage_2001; @waldfogel_understanding_1998].

- @budig_differences_2010: la penalidad es mayor para las mujeres de bajos salarios → evidencia con conditional quantile regression (CQR).

- Sin embargo, la penalidad no es uniforme: puede variar según posición en la distribución salarial, contexto institucional o régimen de bienestar.

- En Europa, la transición a la maternidad también genera una penalidad relativa dentro de la pareja, que puede compensarse con beneficios familiares [@dottisani_withincouple_2015].

:::

## Ejes de análisis

::: {style="font-size: 120%;"}
Los dos estudios ponen a prueba hipótesis rivales sobre cómo y dónde opera la penalidad maternal:
:::
::: {.incremental .highlight-last style="font-size: 120%;"}

1. Desigualdad en el mercado laboral [@killewald_motherhood_2014]

    - ¿La penalidad es mayor para mujeres con bajos salarios (como decía Budig & Hodges)?
    - O, por el contrario, ¿afecta más a las de la mitad de la distribución salarial?

2. Desigualdad dentro del hogar [@dottisani_withincouple_2015]

    - ¿El nacimiento de un hijo reduce el peso económico de la mujer frente a su pareja?
    - ¿Los beneficios familiares pueden compensar la pérdida de ingresos laborales?

:::

## @dottisani_withincouple_2015

:::: {style="font-size: 105%;"}
::: {.incremental .highlight-last}

- **Objetivo**: Analizar si el nacimiento de un hijo reduce los ingresos laborales de las mujeres en relación con los de su pareja, y hasta qué punto los beneficios familiares compensan esta pérdida.

- **Datos**: panel EU-SILC (2004–2008) de ocho países europeos (Austria, Bélgica, España, Irlanda, Italia, Luxemburgo, Portugal y Suecia).
- **Mediciones**:

    * Ingreso relativo de la mujer en la pareja: participación en ingresos laborales (y₁).
    * Ingreso relativo ampliado: ingresos laborales + beneficios familiares (y₂).
    * Controles: individuales (ej. edad, horas trabajadas) y contextuales (ej. régimen bienestar)

- **Estrategia**: modelos de efectos fijos (FE) → estiman el impacto de un nacimiento (niño 0–1 años en el hogar) sobre el ingreso relativo de la madre.

:::
::::

## @dottisani_withincouple_2015

:::: {style="font-size: 100%;"}

- **Hallazgos**:

::: {.fragment}

<div style="text-align:center;">
  <img src="images/resul1.png" style="width:70%; max-width:none;">
</div>

:::
::::

## @dottisani_withincouple_2015

:::: {style="font-size: 120%;"}

::: {.incremental .highlight-last}

- **Contribuciones**:

    - Evidencia de una penalidad relativa de maternidad en la mayoría de países, pero compensada por beneficios familiares en Estados con apoyos generosos (Suecia, Luxemburgo)
    - En países del sur de Europa (Italia, Portugal, España), la reducción es mínima, ya que las madres tienden a permanecer en la fuerza laboral.
    - Propone distinguir entre penalidad en ingresos laborales vs. penalidad en ingresos totales, mostrando que la primera no siempre implica la segunda.

:::
::::

## @killewald_motherhood_2014

:::: {style="font-size: 120%;"}
::: {.incremental .highlight-last}

- **Objetivo**: Evaluar si la penalidad por maternidad es mayor para mujeres con bajos salarios, como plantearon Budig & Hodges (2010) con CQR, pero ahora usando UQR.

- **Datos**: panel de mujeres blancas, NLSY79 (1979–2004).
- **Mediciones**:

    * Salarios: log de salarios transformados con RIF
    * Maternidad: dummy 0 = no madre, 1 = madre
    * Controles: edad, region, densidad poblacional y FE individuales

- **Estrategia**: Unconditional Quantile Regression (UQR) → permite estimar la penalidad en distintos puntos de la distribución salarial sin redefinir los cuantiles por covariables (a diferencia de CQR).

:::
::::

## @killewald_motherhood_2014

:::: {style="font-size: 110%;"}
- **Hallazgos**:

::: {.fragment}

<div style="text-align:center;">
  <img src="images/resul2.png" style="width:80%; max-width:none;">
</div>

:::
::::

## @killewald_motherhood_2014

:::: {style="font-size: 120%;"}

::: {.incremental .highlight-last}

- **Contribuciones**:

     - @budig_differences_2010 preguntan: ¿Qué madres enfrentan la mayor penalidad salarial? → concluyen que son las de bajos salarios.
     - @killewald_motherhood_2014 muestran que esta pregunta no puede responderse con CQR, ya que mide diferencias condicionales, no sobre la distribución salarial total.
     - Usando UQR, encuentran que la penalidad es más fuerte en la parte media de la distribución salarial, y menor en los extremos (bajos y altos salarios).

:::
::::

## Diálogo entre las investigaciones

:::: {style="font-size: 95%;"}

| Aspecto                | **Killewald & Bearak (2014)**                 | **Dotti Sani (2015)**                  |
|-------------------------|-----------------------------------------------|----------------------------------------|
| **Unidad de análisis**  | Mujeres y salarios individuales               | Parejas: mujer vs. varón               |
| **Pregunta**            | ¿Dónde es mayor la penalidad salarial?        | ¿Cambia el peso económico de la mujer? |
| **Datos / Método**      | EE.UU., NLSY79, UQR                           | 8 países UE, EU-SILC, FE               |
| **Hallazgos**           | Penalidad mayor en la **mitad** de la curva   | Penalidad relativa, compensada por BF  |
| **Similitudes**         | Ambas estudian penalidad maternal, no uniforme| Enfatizan heterogeneidad por contexto  |
| **Encuentros / Desenc.**| Corrigen a Budig: no son las de bajos salarios| Beneficios pueden neutralizar penalidad|
| **Contribución**        | Desigualdad *entre mujeres*                   | Desigualdad *dentro de la pareja*      |
| **Interés sociológico** | Escala **estructural** (mercado laboral)      | Escala **relacional** (hogar)          |

::::

## Discusión

::: {.incremental .highlight-last style="font-size: 120%;"}

1. ¿Qué aprendemos sobre la penalidad maternal cuando la miramos en el mercado laboral (entre mujeres) versus dentro del hogar (entre parejas)?

2. ¿Cómo interactúan la posición salarial y los regímenes de bienestar en la producción de desigualdades de género?

3. Si la penalidad no es “promedio” sino heterogénea, ¿qué implica esto para pensar la justicia social y la equidad de género?

:::

# Gracias por su atención!

## Referencias

### 2025-10-08 presentations/coloquio_2025/mjp-mobility.qmd
Up varios

@@
   geom_point(size = 3.5) +
-  # etiquetas: coef en el punto
-  geom_text(aes(label = est_lab), vjust = -0.8, size = 4) +
   scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
+  coord_flip() +
+  geom_text(aes(label = est_lab), hjust = -0.3, size = 4) +
   labs(x = "Estimate (β)", y = NULL) +
-  theme_ggdist() +
-  theme(axis.title = element_text(face = "bold"),
-        axis.text  = element_text(face = "bold"),
-        text = element_text(size = 18))
-
+  theme_ggdist() 
 ```

### 2025-10-08 presentations/seminario/mjp-mobility-seminario.qmd
Up varios

--- 2025-10-08 presentations/coloquio_2025/mjp-mobility.qmd
@@
 ---
-bibliography: "input/bib/tesis-magister.bib"
-csl: "input/bib/apa6.csl"
+bibliography: "bib/tesis-magister.bib"
+csl: "bib/apa6.csl"
 lang: es
@@
 ::: {.column width="20%"}
-</br>
 
@@
 
-</br></br>
+![](images/jusmer_trans.png)
+</br>
 
-![](images/jusmer_trans.png)
+![](images/coes.png)
 
@@
 
-# **Preferencias por la comodificación**
-
-## **El rol de la movilidad social intergeneracional y las creencias meritocráticas en Chile**
+# **Movilidad social y preferencias por la comodificación de pensiones en Chile**
 
@@
 
-::: {.medium}
+::: {.blue2 .medium}
 
@@
 
-Coloquio de Investigación en Justicia Distributiva y Desigualdad Socioeconómica
+Seminario de tesis
 
-4 Septiembre 2025, Santiago
+Supervisores: Prof. Mauricio Bucca y Prof. Andrea Canales
+
+8 Octubre 2025, Santiago
 
@@
 
+## Contexto chileno
+
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+- Crecimiento con elevada desigualdad [@llorca-jana_historia_2021; @flores_top_2020]
+- Rigida movilidad relativa y fuerte clausura cúspide [@lopez-roldan_comparative_2021; @torche_intergenerational_2014]
+- Profunda privatización y comodificación de áreas de reproducción social con fuerte dependencia Estatal [@madariaga_three_2020; @boccardo_30_2020]
+- Pensiones: sistema de **capitalización individual** administrado por **AFPs**; cubre ~11 millones de cotizantes, pero **27% de la fuerza laboral** queda fuera por **informalidad** [@superintendenciadepensiones_estadisticas_2024]
+- Conflictividad social previsional [@somma_no_2021; @nucleodesociologiacontingente_informe_2020] y legitimidad del sistema ("con mi plata no") [@castillo_perceptions_2025; @canalesceron_sujeto_2021; @panes_criticas_2020]
+:::
+
 ## Preferencias por la comodificación de servicios
@@
    * Las percepciones meritocráticas (esfuerzo) aumenta el respaldo a la justicia de mercado.
+:::
 
+::: {.fragment style="font-size: 120%;"}
+**Sin embargo...**
 :::
@@
 
-El movimiento entre posiciones sociales expone a los individuos experiencias y mecanismos que afectan nociones de justicia [@alesina_intergenerational_2018; @ares_changing_2020; @gugushvili_subjective_2017; @day_movin_2017]
+El movimiento entre posiciones sociales expone a los individuos a condiciones y experiencias que afectan sus nociones de justicia distributiva [@alesina_intergenerational_2018; @ares_changing_2020; @gugushvili_subjective_2017; @day_movin_2017]
 
@@
 
-Meritocracia como mecanismo de sesgo de atribución[@newman_economic_2023; @gugushvili_intergenerational_2016c; @molina_its_2019].
+¿Cómo?
+Movilidad ascendente refuerza ver al mercado como asignador justo, la descendente lo cuestiona.
+La meritocracia es un canal cognitivo para ese proceso [@mau_inequality_2015; @gugushvili_intergenerational_2016].
 
@@
 
-**_(1) ¿Cómo la movilidad social intergeneracional afecta las preferencias por comodificación en salud, pensiones y educación en Chile?_**
+**_(1) ¿Cómo la movilidad social intergeneracional afecta las preferencias por la comodificación de las pensiones en Chile?_**
 
-**_(2) ¿Cómo las creencias meritocráticas median esta relación en el contexto chileno?_**
+**_(2) ¿En qué medida las creencias meritocráticas condicionan esta relación?_**
 
@@
 
-## Contexto chileno
+## Movilidad social
 
@@
 
-- Crecimiento con elevada desigualdad [@llorca-jana_historia_2021; @flores_top_2020]
-- Profunda privatización y comodificación de áreas de reproducción social con fuerte dependencia Estatal [@madariaga_three_2020; @boccardo_30_2020]
-- Rigida movilidad relativa y fuerte clausura cúspide [@lopez-roldan_comparative_2021; @torche_intergenerational_2014]
-- Conflictividad social [@somma_no_2021; @nucleodesociologiacontingente_informe_2020] y legitimidad de la desigualdad [@castillo_perceptions_2025; @canalesceron_sujeto_2021; @panes_criticas_2020]
+- Preferencias redistributivas [@alesina_intergenerational_2018; @ares_changing_2020; @schmidt_experience_2011; @breen_effects_2024]
+- Legitimidad de la desigualdad económica [@gugushvili_intergenerational_2016], política social [@gugushvili_subjective_2017] y creencias sobre la desigualdad [@bucca_merit_2016; @day_movin_2017]
+- Diversos **mecanismos explicativos** [@helgason_class_2025]:
+
+    * self-interest (ej. intereses materiales, POUM)
+    * socialización (ej. aculturación, socialización valores, máximización estatus)
+    * **atribuciones** (ej. self-serving bias)
+
+:::
+
+## El mecanismo del *self-serving bias*
+
+::: {.incremental style="font-size:120%;"}
+
+- Las personas ajustan sus **creencias sobre justicia** según cómo explican su posición social [@gugushvili_intergenerational_2016; @miller_selfserving_1975].  
+- Las experiencias de movilidad activan **procesos de atribución causal**:  
+  - **Interna** → éxito por esfuerzo → el mercado es justo.  
+  - **Externa** → fracaso por factores estructurales → el mercado es injusto.  
+- **Ascenso:** refuerza la creencia de que el mercado premia el mérito.  
+- **Descenso:** puede generar disonancia y cuestionar esa justicia.
+
+:::
+
+## ¿Cómo?
+::: {.incremental .highlight-last style="font-size: 120%;"}
+- **No solo cambia intereses materiales**, también cambia **cómo las personas entienden la justicia** [@gugushvili_intergenerational_2016; @helgason_class_2025].
+- Al moverse, los individuos **recalibran sus creencias sobre la meritocracia** [@mijs_belief_2022a].
+- **Movilidad ascendente →** refuerza la creencia en un mercado justo (asignador legítimo de recompensas).
+- **Movilidad descendente →** desafía esa creencia (percepción de arbitrariedad o injusticia).  
+- La **meritocracia** filtra estas experiencias [@mau_inequality_2015]:  
+  - *Fuerte creencia en mérito →* mantiene la fe en la justicia del mercado (disonancia resuelta internamente).  
+  - *Débil creencia en mérito →* percibe el sistema como injusto (disonancia resuelta externamente).  
 :::
@@
 
-::: {.fragment}
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
-<div style="text-align:center;">
-  <img src="images/idea.png" style="width:100%; max-width:none;">
-</div>
+- **H1**: La movilidad ascendente refuerza las preferencias por comodificación de pensiones, mientras que la movilidad descendente las reduce.
 
+- **H2a.** El efecto positivo de la movilidad ascendente sobre el apoyo a la mercantilización es más fuerte entre las personas con creencias meritocráticas más sólidas.  
+→ *Self-serving bias*: el ascenso confirma que el sistema recompensa el esfuerzo.
+
+- **H2b.** El efecto negativo de la movilidad descendente sobre el apoyo a la mercantilización es más débil entre las personas con creencias meritocráticas más fuertes.  
+→ *Resolución de la disonancia*: los meritócratas internalizan la pérdida («mi culpa»), mientras que los no meritócratas la atribuyen al sistema («es injusto»).
 :::
@@
 #| fig-align: center
-#| out-width: '200%'
+#| out-width: '180%'
 library(dagitty)
@@
   geom_text(x=-1, y=-0.70, label="Y = Preferencias por\ncomodificación", hjust=0, size=5) +
-  geom_text(x=-1, y=-0.95, label="M = Creencias\nmeritocráticas", hjust=0, size=5)  +
+  geom_text(x=-1, y=-0.96, label="M = Meritocracia (post-T)\n— heterogeneidad descriptiva", hjust=0, size=5) +
   geom_text(x=0.55, y=-0.7, label="C = Confusores pre-tratamiento\n(propensity)", hjust=0, size=5)  +
   geom_text(x=0.55, y=-0.9, label="L = Confusor post-T", hjust=0, size=5)
+```
 
-```
+::: {.fragment style="font-size: 100%;"}
+- M es post-tratamiento; lo uso para caracterizar heterogeneidad del efecto, sin atribuir causalidad a 
+T×M. El efecto principal (ATT) no depende de M.
+::::
 
@@
 
-<div style="text-align:center;">
-  <img src="images/outcome.png" style="width:100%; max-width:none;">
-</div>
+```{r}
+#| label: fig-desc
+#| fig-cap-location: top
+#| fig-cap: "Distribución de respuestas para las preferencias por comodificación de pensiones Chile (2016-2023)"
+#| fig-align: center
+#| out-width: '130%'
+
+g1
+
+```
 
@@
 
-1.
-**Asignación de clase**: En base a la ocupación (ISCO-08) del padre/madre y del entrevistado se crean estratos de origen y destino (bajo, medio y alto) según terciles de la distribución del índice de estatus ocupacional (ISEI)
+1. **Asignación de clase (O y D):** Con ISCO-08/ISEI, defino origen (O ∈ {bajo, medio, alto}) y destino (D). Para cada origen \(O=j\), el tratamiento es llegar a \(D=k\) (móvil) vs inmóvil \(D=j\).
 
-2.
-**Propensity-score weighting**: Se estima la probabilidad de movilidad social mediante un modelo logístico multinomial en relación con las trayectorias inmóviles (categ.
-ref.) para cada estrato.
-Creación de IPW o "clones"
+2. **ATT por entropy balancing (no PS):**  
+   - En cada estrato \(O=j\), uso `ebal` para **reponderar solo a los controles** (inmóviles) e igualar sus valores de los confusores pre-tratamiento \(C\) a los de los tratados (móviles).  
+   - Obtengo **IPW-ATT estabilizados** (tratados=1; controles = pesos de `ebal`).
 
@@
 library(sjPlot)
-
-db <- db %>% 
-  mutate(
-    across(
-      .cols = c(estrato_orig, estrato_ocupa),
-      .fns = ~ car::recode(.,
-                           recodes = c("1 = 'Low';
-                                       2 = 'Middle';
-                                       3 = 'High'"),
-                           as.factor = T)
-    )
-  ) %>% 
-  mutate(
-    across(
-      .cols = c(estrato_orig, estrato_ocupa),
-      .fns = ~ factor(., levels = c("Low", "Middle", "High"))
-    )
-  )
-
-db$estrato_orig <- sjlabelled::set_label(db$estrato_orig, "Father stratum")
-
-db$estrato_ocupa <- sjlabelled::set_label(db$estrato_ocupa, "Offspring stratum")
 
@@
 
-## Mediador
+## Heterogenidad causal
 
@@
 
-- **Meritocracia**: se mide a partir de dos componentes, a saber, el esfuerzo y el talento [@young_rise_1958]
+- **Meritocracia**: se mide a partir de la percepción de recompensa del esfuerzo [@young_rise_1958]
 
@@
 
-    * Ítem esfuerzo: “En Chile, se recompensa a las personas por su esfuerzo”
-    * Ítem talento: “En Chile, se recompensa a las personas por su inteligencia y habilidades”
-    * Ambos ítems se responden en una escala Likert de 1 (totalmente en desacuerdo) a 5 (totalmente de acuerdo)
-
+    * Ítem: “En Chile, se recompensa a las personas por su esfuerzo”
+    * Escala Likert de 1 (totalmente en desacuerdo) a 5 (totalmente de acuerdo). 
+    * Fue dicotomizado de modo tal que baja meritocracia (≤ 3) y alta meritocracia (≥ 4) 
 :::
@@
 
-- Efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran trasladado a un destino alternativo *k′* [@breen_effects_2024] (@eq-att).
+- @breen_effects_2024: Efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran trasladado a un destino alternativo *k′* (@eq-breen).
 :::
@@
 $$ 
-ATT_{jkk′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
-$$ {#eq-att}
+ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
+$$ {#eq-breen}
 
@@
 
-- El estimador resultante es el efecto promedio del tratamiento sobre los tratados (ATT) 
+- Mi interés es comparar con la situación de inmobilidad: cuando _k'_ = _j_ (@eq-att)
 :::
 
-::: {.fragment style="font-size: 120%;"}
+::: {.fragment}
+$$ 
+ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
+$$ {#eq-att}
 
-- Mi interés es comparar con la situación de inmobilidad: cuando _k'_ = _j_
 :::
 
-## Estimando
+## Estimador
 
-::: {.incremental style="font-size: 120%;"}
+::: {style="font-size:110%;"}
 
-- Efecto movilidad: modelos de regresión lineal con ponderaciones de probabilidad inversa (IPW) para la movilidad condicional a la clase de origen [@breen_effects_2024] y SE clusterizados por id
+$$
+Y_i = \alpha + \beta T_i + X_i'\gamma + \lambda_{t(i)} + \varepsilon_i
+\quad \text{(WLS con IPW–ATT, CR2 por id)}
+$$
+:::
 
-- Mediación: Average Causal Mediation Effect [@imai_identification_2010]
+::: {.incremental style="font-size:110%;"}
+
+- **Muestra:** individuos con mismo origen de clase $O=j$  
+-  **Outcome:** $Y_i$ preferencia por comodificación de pensiones 
+- **Tratamiento:** $T_i=1$ si $D_i=k$ (*móvil*), $T_i=0$ si $D_i=j$ (*inmóvil*)  
+- **Pesos:** IPW–ATT *estabilizados* con `ebal`  
+- **Errores estándar:** CR2 *clusterizados por id*  
+- **Efectos fijos:** $\lambda_{t(i)}$ por ola  
+- **Controles:** $X_i'\gamma$ covariables propensity score movilidad
+- **Interpretación:** $\beta = \widehat{ATT}_{j,k,j}$  
+  → efecto causal de alcanzar destino $k$ respecto a permanecer en $j$
 
@@
 
-## Modelos
+## Mobility effects
 
-::: {.fragment}
+::: {.fragment style="font-size: 120%;"}
 
 ```{r}
-#| label: fig-mh
+#| label: fig-mp
 #| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación en salud"
+#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación de pensiones"
 #| fig-align: center
@@
 
-library(broom)
-library(forcats)
-conflicts_prefer(dplyr::filter)
+g2 +
+  theme(text = element_text(size = 16))
+```
 
-df_health <- bind_rows(
-tidy(mh_lm, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Low→Middle \n(N obs. = 532; N Clusters = 381)") 
-,
-tidy(mh_lh, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Low→High \n(N obs. = 452; N Clusters = 342)") 
-,
-tidy(mh_mh, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Middle→High \n(N obs. = 432; N Clusters = 320)") 
-,
-tidy(mh_ml, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Middle→Low \n(N obs. = 469; N Clusters = 345)") 
-,
-tidy(mh_hm, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "High→Middle \n(N obs. = 620; N Clusters = 444)") 
-,
-tidy(mh_hl, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "High→Low \n(N obs. = 576; N Clusters = 432)") 
+:::
+
+## Heterogenidad
+
+::: {.fragment style="font-size: 120%;"}
+
+```{r}
+#| label: fig-mp_i
+#| fig-cap-location: top
+#| fig-cap: "Interacción entre movilidad y meritocracia en preferencia por comodificación de pensiones"
+#| fig-align: center
+#| out-width: '180%'
+
+g3+
+  theme(text = element_text(size = 16))
+```
+
+:::
+
+## Heterogenidad
+
+::: {.fragment style="font-size: 120%;"}
+
+```{r}
+#| label: fig-marginal
+#| fig-cap-location: top
+#| fig-cap: "Efecto marginal de movilidad descendente High→Low por meritocracia"
+#| fig-align: center
+#| out-width: '180%'
+
+b <- coef(mp_hl_i)
+V <- vcov(mp_hl_i)   # ya es CR2 por cómo estimaste el modelo
+
+# 2) Efectos marginales de t por nivel de meritocracia
+est_m0 <- b["t"]                           # mAlta = 0
+se_m0  <- sqrt(V["t","t"])
+
+est_m1 <- b["t"] + b["t:mAlta"]            # mAlta = 1
+se_m1  <- sqrt(V["t","t"] + V["t:mAlta","t:mAlta"] + 2*V["t","t:mAlta"])
+
+# 3) Arma tabla y CIs 95%
+z <- qnorm(0.975)
+df_me <- tibble(
+  meritocracia = c("Baja meritocracia","Alta meritocracia"),
+  estimate = c(est_m0, est_m1),
+  se = c(se_m0, se_m1),
+  ci_lo = estimate - z*se,
+  ci_hi = estimate + z*se
 )
 
-df_health_lab <- df_health %>%
-  mutate(
-    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
-    est_lab = sprintf("%.2f%s", estimate, sig),
-    low_lab = sprintf("%.2f", conf.low),
-    high_lab= sprintf("%.2f", conf.high)
-  )
-
-ggplot(df_health_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 0.8) +
-  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
-  geom_point(size = 3.5) +
-  # etiquetas: coef en el punto
-  geom_text(aes(label = est_lab), vjust = -0.8, size = 4) +
-  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
-  labs(x = "Estimate (β)", y = NULL) +
-  theme_ggdist() +
-  theme(axis.title = element_text(face = "bold"),
-        axis.text  = element_text(face = "bold"),
-        text = element_text(size = 18))
+# 4) Grafica (efecto de High→Low por meritocracia)
+ggplot(df_me, aes(x = meritocracia, y = estimate)) +
+  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
+  geom_point(size = 4) +
+  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0, size = 1.2) +
+  labs(x = NULL, y = "Efecto de High→Low (β | meritocracia)") +
+  theme_ggdist(base_size = 16)
 
@@
 
-## Modelos
-
-```{r}
-#| label: fig-me
-#| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación en educación"
-#| fig-align: center
-#| out-width: '180%'
-
-df_educ <- bind_rows(
-tidy(me_lm, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Low→Middle \n(N obs. = 534; N Clusters = 380)") 
-,
-tidy(me_lh, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Low→High \n(N obs. = 450; N Clusters = 342)") 
-,
-tidy(me_mh, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Middle→High \n(N obs. = 424; N Clusters = 311)") 
-,
-tidy(me_ml, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Middle→Low \n(N obs. = 467; N Clusters = 342)") 
-,
-tidy(me_hm, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "High→Middle \n(N obs. = 611; N Clusters = 437)") 
-,
-tidy(me_hl, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "High→Low \n(N obs. = 569; N Clusters = 430)") 
-)
-
-df_educ_lab <- df_educ %>%
-  mutate(
-    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
-    est_lab = sprintf("%.2f%s", estimate, sig),
-    low_lab = sprintf("%.2f", conf.low),
-    high_lab= sprintf("%.2f", conf.high)
-  )
-
-ggplot(df_educ_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 0.8) +
-  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
-  geom_point(size = 3.5) +
-  # etiquetas: coef en el punto
-  geom_text(aes(label = est_lab), vjust = -0.8, size = 4) +
-  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
-  labs(x = "Estimate (β)", y = NULL) +
-  theme_ggdist() +
-   theme(axis.title = element_text(face = "bold"),
-        axis.text  = element_text(face = "bold"),
-        text = element_text(size = 18))
-
-```
-
-## Modelos
-
-```{r}
-#| label: fig-mp
-#| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación en pensiones"
-#| fig-align: center
-#| out-width: '180%'
-
-df_pension <- bind_rows(
-tidy(mp_lm, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Low→Middle \n(N obs. = 511; N Clusters = 365)") 
-,
-tidy(mp_lh, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Low→High \n(N obs. = 435; N Clusters = 333)") 
-,
-tidy(mp_mh, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Middle→High \n(N obs. = 396; N Clusters = 298)") 
-,
-tidy(mp_ml, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "Middle→Low \n(N obs. = 447; N Clusters = 335)") 
-,
-tidy(mp_hm, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "High→Middle \n(N obs. = 581; N Clusters = 421)") 
-,
-tidy(mp_hl, conf.int = TRUE) %>%
-  filter(term == "t") %>%
-  mutate(term = "High→Low \n(N obs. = 544; N Clusters = 415)") 
-)
-
-df_pension_lab <- df_pension %>%
-  mutate(
-    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
-    est_lab = sprintf("%.2f%s", estimate, sig),
-    low_lab = sprintf("%.2f", conf.low),
-    high_lab= sprintf("%.2f", conf.high)
-  )
-
-ggplot(df_pension_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 0.8) +
-  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 1) +
-  geom_point(size = 3.5) +
-  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
-  coord_flip() +
-  geom_text(aes(label = est_lab), hjust = -0.3, size = 4) +
-  labs(x = "Estimate (β)", y = NULL) +
-  theme_ggdist() 
-```
-
-## Mediación
-
-```{r}
-#| label: fig-acme
-#| fig-cap-location: top
-#| fig-cap: "ACME para meritocracia y preferencias por comodificación en pensiones"
-#| fig-align: center
-#| out-width: '180%'
-
-plot(med.fit)
-```
-
 # Discusión y proyecciones {data-background-color="#523870"}
 
-# Discusión y proyecciones {data-background-color="#523870"}
+## Discusión y proyecciones
 
-::: { .incremental .highlight-last2 style="font-size: 120%;"}
+::: {.incremental style="font-size: 120%;"}
 
-1. **Movilidad social**: ascendente (Middle→High) _incrementa_ la preferencia por comodificación en pensiones (dominio), no en otros. 
-2. **Contexto chileno**: profunda comodificación y privatización de servicios escenciales → trabajador-empresario 
-3. **Mediación**: rol de la meritocracia como mediador → ACME supuestos muy fuertes =  alternativas (?)
+1. **Movilidad social**: no hay efectos significativos. Sin embargo, existe cierta evidencia en la dirección de $H1$:
+
+    * Middle→High _incrementa_ la preferencia por comodificación en pensiones. 
+    * High→Low _disminuye_ la preferencia por comodificación en pensiones.
+
+2. **Meritocracia**: Entre quienes son escépticos con la meritocracia, descender High→Low implica preferir menos comodificación de pensiones ($H2b$).
+
+3. **Contexto chileno**: profunda comodificación y privatización de servicios escenciales → trabajador-empresario
+
 :::
@@
 ## Referencias
+
+Bien, y cómo entra la creencia meritocratica en este diagrama de hipotesis y DAG causal entonces?

### 2025-10-08 processing/prod_analysis.R

@@ -12,26 +12,122 @@ if (!require("pacman")) install.packages("pacman")
 
 pacman::p_load(tidyverse,
                here,
+               sjPlot,
                sjmisc,
-               PSweight,
                texreg,
                estimatr,
-               mediation,
-               conflicted)
+               conflicted,
+               ggdist,
+               summarytools)
 
-conflict_prefer("select", "dplyr")
+conflicts_prefer(dplyr::filter)
 options(scipen=999)
 rm(list = ls())
 
 # 2. Data -----------------------------------------------------------------
 
-load(here("input/data/proc/data_v1.RData"))
+#load(url("https://github.com/Andreas-Lafferte/mobility-market-justice/raw/refs/heads/main/input/data/proc/data_v1.RData"))
+
+load(here("input/data/proc/data_v2.RData"))
+
+glimpse(db)
 
 # 3. Analysis -----------------------------------------------------------
 
-# 3.1 Regresiones con IPW -----------------------------------------------------------
+bases <- c("datos1", "datos2", "datos3", "db")
 
-### Pensiones
+for (nm in bases) {
+  df <- get(nm)
+  
+  df <- df %>%
+    mutate(
+      merit_effort = if_else(merit_effort >= 4, 1, 0),
+      merit_effort = factor(merit_effort, levels = c(0,1), labels = c("Baja", "Alta"))
+      #just_pension = if_else(
+      #  just_pension %in% 4:5, 1L,
+      #  if_else(just_pension %in% 1:3, 0L, NA_integer_)),
+      #just_health = if_else(
+      #  just_health %in% 4:5, 1L,
+      #  if_else(just_health %in% 1:3, 0L, NA_integer_)),
+      #just_educ = if_else(
+      #  just_educ %in% 4:5, 1L,
+      #  if_else(just_educ %in% 1:3, 0L, NA_integer_))
+    )
+  
+  assign(nm, df, inherits = TRUE)
+}
+
+t1 <- db %>% 
+  select(idencuesta, wave, just_pension, estrato_orig, estrato_ocupa, merit_effort, nacionalidad, sexo,
+         m0_edad, etnia, hogar_bip_f, educ_orig_f)
+
+dfSummary(t1,
+          plain.ascii = FALSE,
[…]

### 2025-10-08 processing/prod_balance.R

@@ -1,0 +1,223 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data analysis for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               sjmisc,
+               cobalt,
+               conflicted)
+
+conflict_prefer("select", "dplyr")
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(here("input/data/proc/data_v1.RData"))
+
+# 3. Balance -----------------------------------------------------------
+
+# Covariables pre-tratamiento (las de tus modelos de PS)
+covars_ps <- c("m0_edad","sexo","nacionalidad","etnia","hogar_bip_f","educ_orig_f")
+
+# -----------------------------------------------------------------------
+# A) MIDDLE -> HIGH  (datos2: t = mobmh, w = wm_mh; PS: pm_mh (t=1), pm_mm (t=0))
+# -----------------------------------------------------------------------
+data_mh_att <- datos2 %>%
+  select(idencuesta, t = mobmh, pm_mm, pm_mh,
+         m0_edad, sexo, nacionalidad, etnia, hogar_bip_f, educ_orig_f) %>%
+  na.omit() %>%
+  mutate(ps_pair = pm_mh / (pm_mh + pm_mm),
+         w_att   = ifelse(t == 1, 1, ps_pair / (1 - ps_pair)))
+
+# Balance SMD + Razón de varianzas
+bt_mh_ATT <- cobalt::bal.tab(
+  t ~ m0_edad + sexo + nacionalidad + etnia + hogar_bip_f + educ_orig_f,
+  data     = data_mh_att,
+  weights  = list(IPW = data_mh_att$w_att),
+  estimand = "ATT",
+  un = TRUE,
+  s.d.denom = "pooled",
+  disp.v.ratio = TRUE,
+  m.threshold = 0.10,
+  v.threshold = 2
+)
+print(bt_mh_ATT)
+
[…]

### 2025-10-08 prod_prep_binom.R

@@ -19,8 +19,7 @@ pacman::p_load(tidyverse,
                summarytools,
                nnet,
                occupar,
-               cobalt,
-               WeightIt)
+               PSweight)
 
 options(scipen=999)
 rm(list = ls())
@@ -201,14 +200,13 @@ db <- db %>%
 
 db <- db %>% 
   select(idencuesta, ola, ponderador_long_total, segmento, estrato,
-         just_health, just_pension, just_educ, merit_effort,
+         just_pension, merit_effort,
          nacionalidad, sexo, m0_edad, etnia, hogar_bip_f,
          educ_orig_f, isei08_orig, 
          isei08_ocupa) %>% 
   mutate(just_pension = as.integer(just_pension),
-         just_health = as.integer(just_health),
-         just_educ = as.integer(just_educ),
-         merit_effort = as.integer(merit_effort))
+         merit_effort = as.integer(merit_effort),
+         merit_effort = if_else(merit_effort <= 3, 0, 1))
 
 # Crear estrato_ocupa basado en rangos de isei_ocupa
 
@@ -276,300 +274,164 @@ db$estrato_ocupa <- db$terc_ocupa
 
 table(db$estrato_orig, db$estrato_ocupa)
 
-# Combinations
+db <- db %>%
+  mutate(
+    inmobile = if_else(estrato_ocupa == estrato_orig, 1, 0),
+    upward = if_else(estrato_ocupa > estrato_orig, 1, 0),
+    downward = if_else(estrato_ocupa < estrato_orig, 1, 0))
+
+# Downward
 
 db <- db %>%
   mutate(
-    ll = if_else(estrato_orig == 1 & estrato_ocupa == 1, 1, 0),
-    lm = if_else(estrato_orig == 1 & estrato_ocupa == 2, 1, 0),
-    lh = if_else(estrato_orig == 1 & estrato_ocupa == 3, 1, 0),
-    
-    ml = if_else(estrato_orig == 2 & estrato_ocupa == 1, 1, 0),
-    mm = if_else(estrato_orig == 2 & estrato_ocupa == 2, 1, 0),
[…]

### 2025-10-08 presentations/seminario/mjp-mobility-seminario.qmd
up varios

@@
 ## ¿Cómo?
-::: {.incremental .highlight-last style="font-size: 120%;"}
+::: {.incremental .highlight-last style="font-size: 110%;"}
 - **No solo cambia intereses materiales**, también cambia **cómo las personas entienden la justicia** [@gugushvili_intergenerational_2016; @helgason_class_2025].
@@
 #| fig-align: center
-#| out-width: '180%'
+#| out-width: '160%'
 library(dagitty)
@@
 
-::: {.fragment style="font-size: 100%;"}
+::: {.fragment style="font-size: 60%;"}
 - M es post-tratamiento; lo uso para caracterizar heterogeneidad del efecto, sin atribuir causalidad a 

### 2025-11-02 paper/01-introduction.qmd
Update varios

--- 2025-06-27 advances/informe2.qmd
@@
----
-title: "Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
-format:
-  html: 
-    code-fold: true
-    html-math-method: katex
-    number-sections: true
-    code-link: true
-    title-block-banner: true
-    theme:
-      - pretty.scss
-  pdf: 
-    geometry: margin=2cm
-    template-partials: 
-      - title.tex
-    keep-tex: true
-    include-in-header:
-      text: |
-        \usepackage[noblocks]{authblk}
-        \renewcommand*{\Authsep}{, }
-        \renewcommand*{\Authand}{, }
-        \renewcommand*{\Authands}{, }
-        \renewcommand\Affilfont{\small}
-    number-sections: true
-editor: source
-bibliography: "../input/bib/tesis-magister.bib"
-csl: "../input/bib/apa6.csl"
-author:
-  - name: Andreas Laffert Tamayo
-    orcid: 0000-0002-9008-2454
-    email: alaffertt@estudiante.uc.cl 
-    affiliations:
-      - ref: 1
-affiliations: 
-  - id: 1
-    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
-link-citations: true
-linestretch: 1.5       
-mainfont: Times New Roman
-fontsize: 13pt          
-fig-height: 4
-fig-width: 7.5
----
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-```{r}
-# This document was last modified at {{< modified >}}
-# 
-# and it was last rendered at {{< now >}}
-
-```
-
 # Introduction
-
-:::: {.justify}
 
@@
 The final sections present the empirical findings, offer an interpretation of the results, and conclude with a discussion of their implications.
-
-# Theoretical and empirical background
-
-## Market justice preferences
-
-Beyond state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023].
-As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks.
-These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking [@koos_moral_2019; @immergut_theoretical_1998].
-In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007].
-While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@lindh_bringing_2023].
-In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023].
-Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020].
-This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@satz_por_2019; @mau_inequality_2015].
-
-The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles.
-Beyond stratification structures and socioeconomic conditions, the study of inequality also focuses on individuals’ beliefs about the origins of inequality, the normative frameworks that sustain those beliefs, the mechanisms that shape them, and their implications for attitudes and behavior [@kluegel_beliefs_1981].
-In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016].
-This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011].
-Most of the research on this field explores the legitimacy of wage inequality, specially salary gaps between jobs at opposite ends of the occupational hierarchy [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978].
-Recently, others examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, pensions or social security.
-Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015].
-In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic [@busemeyer_welfare_2020].
-
-Market justice preferences refer to normative beliefs that legitimize the idea that access to core social services—such as healthcare, education, or pensions—should be determined by market-based criteria.
-Following Janmaat’s [-@janmaat_subjective_2013, p.
-359] distinction, these preferences fall under the category of “beliefs,” understood as normative ideas about what inequality should look like, as opposed to “perceptions,” which refer to subjective evaluations of existing inequality.
-Market justice preferences reflect the view that access to these services should depend on individuals’ ability to pay, thus justifying inequalities generated by market mechanisms [@kluegel_legitimation_1999;  @lindh_public_2015].
-The concept draws on Lane’s [-@lane_market_1986] classic contrast between market justice and political justice: the former is grounded in the idea of earned deserts—where rewards reflect effort, productivity, and skill—while the latter prioritizes need and equality, typically expressed in welfare state policies.
-Lane argued that markets and states differ in purpose (efficiency vs.
-need), logic (individual vs.
-collective), and fairness criteria (merit vs.
-equality).
-Market justice assumes that markets are neutral, self-regulating systems in which fair procedures yield outcomes proportional to merit.
-Inequality, from this perspective, is not only expected but legitimate—as long as it arises from fair competition.
-In this way, market justice offers a moral lens through which individuals can view the commodification and stratified access to social services as fair and justified [@kluegel_legitimation_1999; @lindh_public_2015].
-
-The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation.
-A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income.
-This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems.
-Subsequent studies have extended this logic to a wider set of welfare goods.
-For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
-[-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare.
-Similarly, Lee and Stacey [-@lee_fairness_2023] apply this framework to education.
-Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation.
-Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al.
-[-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile.
-These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate.
-Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [-@lindh_public_2015].
-
-Comparative empirical research has identified several individual-level factors that influence support for market justice.
-Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles [@koos_moral_2019; @lindh_public_2015; @svallfors_political_2007].
-For example, Lindh [-@lindh_public_2015] finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries.
-In a comparative analysis, Svallfors [-@svallfors_political_2007] observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class.
-Busemeyer [-@busemeyer_skills_2014] similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
-[-@vondemknesebeck_are_2016] report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage.
-In Chile, Otero and Mendoza [-@otero_power_2024] show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions.
-Beliefs about inequality and political orientation also matter.
-In Chile, Castillo et al.
-[-@castillo_socialization_2024] show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism.
-In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.
-
-However, individual characteristics alone do not fully explain variation in support for market justice.
-A growing body of cross-national research highlights the importance of institutional arrangements in shaping these preferences.
-For instance, Immergut and Schneider [-@immergut_it_2020] find that in countries with higher public spending on healthcare, individuals are less likely to view income-based access as fair.
-Similarly, Busemeyer [-@busemeyer_skills_2014] shows that increased public investment in education is associated with lower support for privatized provision.
-Conversely, Lindh [-@lindh_public_2015] finds that in countries with more market-oriented welfare systems, support for market-based distribution tends to be higher, suggesting that individual attitudes often align with institutional outputs.
-These findings are consistent with neo-institutionalist and policy feedback theories [@campbell_institutional_2020], which argue that institutions do more than redistribute resources—they also shape the normative categories through which individuals assess who is deserving of support [@immergut_theoretical_1998].
-In this view, dominant values and preferences are both shaped by and embedded in institutional configurations [@busemeyer_skills_2014], reinforcing the notion that institutions are not neutral structures, but active producers of the moral frameworks that legitimize or challenge inequality.
-
-## Social mobility
-
-The study of social mobility, its drivers and consequences has long been central to sociology.
-Classic theorists explored not only movement across social hierarchies but also its broader implications for class conflict, norm stability, and institutional change [@breen_effects_2024].
-Sorokin [-@sorokin_social_1927] introduced the concept formally, defining mobility as the shift of individuals, values, or objects between positions within a stratification system, and distinguishing between horizontal and vertical forms.
-Later work differentiated intergenerational from intragenerational mobility, as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which origins constrain destinations [@eyles_social_2022].
-In Latin America, research shows high absolute but low relative mobility [@bucca_merit_2016]: although educational expansion and economic modernization have enabled some upward movement, status reproduction remains strong, especially among elites [@torche_intergenerational_2014; @lopez-roldan_comparative_2021].
-This reflects deep structural inequality, segmented education systems, stratified labor markets, and legacies of dependent development that restrict access to mobility channels and reinforce the intergenerational transmission of advantage.
-
-Beyond mapping mobility patterns, growing research has examined its subjective and attitudinal effects.
-Mobility effects—defined as outcomes resulting from movement between origin and destination classes [@breen_effects_2024]—have long attracted theoretical interest.
-Sorokin’s (1959) dissociative hypothesis posits that mobility, whether upward or downward, may produce psychological strain due to conflicting norms between class contexts, leading to lower life satisfaction.
-Similarly, Lenski [-@lenski_status_1954] argued that status inconsistency—mismatches among education, income, and occupation—can undermine well-being.
-These ideas underpin extensive empirical work linking intergenerational mobility to outcomes such as life satisfaction, mental health, and stress [@prag_subjective_2021; @gugushvili_heterogeneous_2024; @hadjar_does_2015].
-
-Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality.
-A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed.
-Alesina et al.
-[-@alesina_intergenerational_2018] show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies.
-Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility.
-Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [-@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support.
-However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.
-
-Beyond redistribution, other studies have explored the impact of mobility on broader beliefs about inequality.
-Gugushvili [-@gugushvili_intergenerational_2016] finds that upwardly mobile individuals are more likely to adopt individualistic attributions of poverty and to legitimize income inequality, particularly in post-socialist societies.
-Similarly, Bucca [-@bucca_merit_2016] shows that subjective upward mobility reinforces individualistic explanations of wealth in seven Latin American countries.
-At the macro level, Shariff et al.
-[-@shariff_income_2016] demonstrate that higher levels of national economic mobility correlate with greater tolerance of inequality.
-In contrast, Day and Fiske [-@day_movin_2017] find that low perceived mobility undermines belief in meritocracy and a just world, thereby weakening system justification.
-Taken together, this body of research suggests that mobility shapes attitudes toward inequality and justice through multiple, and sometimes contradictory, mechanisms.
-
-The literature on the effects of social mobility has proposed various mechanisms to explain how and why changes in social position may influence individual outcomes [@helgason_class_2025].
-One of the most prominent is the self-interest mechanism, which posits that individuals who experience upward or downward mobility undergo a shift in their material interests, thereby altering their perceptions and preferences [@helgason_longterm_2023; @ares_changing_2020; @langsaether_explaining_2022].
-Closely related to this logic is the Prospect of Upward Mobility (POUM) hypothesis, which suggests that individuals may oppose redistribution not because of their current position, but because they anticipate improving their status in the future [@benabou_social_2001].
-A second line of explanation draws on the framework of theories of socialization.
-Within this tradition, hypotheses such as acculturation, socialization, and status maximization propose that individuals adjust their attitudes based on the norms and values of either their class of origin or their destination, or a combination of both [@jaime-castillo_social_2019].
-However, most of these mechanisms focus primarily on the indirect effects of origin and destination positions, rather than on the direct effect of experience of movement itself.
-In response to this gap, recent research has highlighted the role of self-serving bias in causal attribution processes, suggesting that individuals tend to explain their mobility trajectories in ways that justify their current position, which in turn shapes their beliefs and preferences [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
-
-The self-serving bias mechanism builds on a value-oriented perspective, emphasizing that individuals’ experiences of social mobility shape their causal attributions, which in turn influence their beliefs about justice and distributive preferences [@gugushvili_trends_2014].
-Causal attribution refers to the process through which individuals generate explanations for their own behavior and outcomes, as well as those of others [@gugushvili_intergenerational_2016].
-In this view, people’s interpretations of economic inequality depend on whether they believe such disparities reflect unequal individual contributions.
-Individuals who adopt an internal attribution framework tend to see success or failure as rooted in personal characteristics such as effort, talent, or merit.
-In contrast, those who regard inequality as unjust are more likely to adopt an external attribution model, viewing outcomes as the result of structural barriers beyond individual control [@kluegel_beliefs_1981].
-This mechanism, often described as intrapersonal causal attribution, reflects how people explain their own socioeconomic positions—typically attributing their successes to internal qualities while blaming failures on external circumstances [@miller_selfserving_1975].
-Over time, individuals may revise their beliefs and attitudes: while early views are shaped by their social origin, these are later adjusted in light of personal experiences of mobility and the perceived role of ascribed versus achieved factors in determining socioeconomic outcomes [@gugushvili_intergenerational_2016].
-
-Empirical research has provided support for the self-serving bias mechanism in shaping redistributive preferences and attitudes toward the legitimacy of inequality.
-For instance, Schmidt [-@schmidt_experience_2011] finds that individuals who experience upward mobility are more likely to interpret their success as the result of personal effort or merit, and consequently perceive redistribution as less necessary.
-Conversely, individuals who experience downward mobility tend to attribute their decline to external circumstances—such as structural inequality or unemployment—and show stronger support for redistribution.
-In a comparative analysis across different welfare domains, Gugushvili [-@gugushvili_subjective_2017] demonstrates that upward mobility is associated with lower support for government spending on housing and pensions, while individuals who experience downward mobility express lower support for healthcare and education spending, but favor increased investment in housing and pensions, reflecting the material nature of these domains.
-Moreover, Gugushvili [-@gugushvili_intergenerational_2016c] finds that upward mobility is linked to greater justification of income inequality, suggesting that improved social standing reinforces an attributional view in which success is seen as the result of individual characteristics—thus legitimizing inequality as a fair outcome.
-
-Consequently, considering this theoretical and empirical background, the first hypothesis of this research is that:
-
-$H1$: Experiencing upward (downward) social mobility is positively (negatively) associated with greater support for market justice in healthcare, pensions, and education.
-
-## Meritocracy
-
-Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs.
-Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
-As initially conceptualized by Michael Young [-@young_rise_1958], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality.
-However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2019; @sandel_tyranny_2020].
-From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality.
-People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_meritocracia_2019; @batruch_belief_2023].
-Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.
-
-I adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through which individuals justify inequality [@young_rise_1958].
-Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility.
-A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed.
-Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023].
-The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions.
-For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023; @wiederkehr_belief_2015; @wilson_role_2003].
-This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage.
-Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.
-
-These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
-This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025].
-Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
-Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_meritocracia_2019; @castillo_socialization_2024].
-
-In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups.
-Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
-This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
-Moreover, Busemeyer et al.
-[-@busemeyer_positive_2021] argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
-This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
-
-Importantly, recent research has explored how meritocratic beliefs interact with experiences of intergenerational mobility to shape distributive attitudes.
-The belief that one’s success is earned can lead upwardly mobile individuals to internalize meritocratic narratives and justify existing inequalities, reinforcing support for market justice [@gugushvili_intergenerational_2016c; @molina_its_2019].
-Conversely, downwardly mobile individuals who maintain strong meritocratic beliefs may interpret their status as a personal failure, reducing their support for redistribution [@day_movin_2017].
-At the macro level, Shariff et al.
-[-@shariff_income_2016] show that higher perceived mobility increases tolerance for inequality, suggesting that meritocracy and mobility are mutually reinforcing.
-
-Taken together, this literature supports the idea that meritocratic beliefs can moderate the relationship between mobility and market justice preferences.
-Individuals who experience mobility—especially upward—may draw on meritocratic narratives to legitimize both their own status and broader inequalities, thereby strengthening their support for market-based distribution.
-Accordingly, the second hypothesis of this study is:
-
-$H2$: The positive (negative) relationship between upward (downward) social mobility and support for market justice in healthcare, pensions, and education is moderated by meritocratic beliefs; specifically, this association is stronger (weaker) among individuals with higher perceptions of meritocracy.
-
-## The Chilean context
-
-# Data, variables and methods
-
-## Data
-
-This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
-The survey is an annual panel collected from 2016 to 2023, comprising two independent samples (original and refreshment), featuring permanent and rotating questionnaire modules.
-It evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile.
-The sampling design is complex, probabilistic, clustered, multistage, and stratified by city size, targeting men and women aged 18--75 who habitually reside in private dwellings in urban areas spanning 40 cities (92 municipalities, 13 regions).
-This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample.
-Further details on sampling, attrition, and weighting can be found at https://coes.cl/encuesta-panel/, and the dataset is publicly available at https://dataverse.harvard.edu/dataverse/elsoc.
-
-## Variables
-
-*Outcome variables*
-
-**Market justice preferences**: The outcome variables in this study are market justice preferences.
-This construct is operationalized through three variables that address the degree of justification regarding whether access to social services in healthcare, pensions, and education should be income conditional.
-Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is asked for pensions and education.
-In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree").
-Although these items theoretically reflect the same underlying concept of market justice [@castillo_socialization_2024; @lindh_public_2015], they are analyzed separately because healthcare, pensions, and education may prompt distinct response patterns [@lee_fairness_2023; @busemeyer_skills_2014; @immergut_it_2020].
-
-*Independent variables*
-
-**Social mobility**: Intergenerational social mobility is treated as an exposure indicating whether respondents occupy a different class position from their fathers.
-Following Breen and Ermisch's [-@breen_effects_2024] framework for estimating causal effects of mobility, the variable is constructed in two stages.
-
-1.
-Class assignment.
-Using the International Socio-Economic Index of Occupational Status (ISEI) derived from three-digit ISCO-08 codes, both paternal (origin) and respondent (destination) occupations are grouped into three strata:
-
-    -   Upper (codes 100-299: managers, directors, professionals)
-    -   Middle (codes 300-499: skilled white-collar workers)
-    -   Lower (codes 500 and above: manual, semi-skilled, or unskilled workers)
-
-2.
-Propensity-score estimation.
-To model mobility as a treatment---i.e., a transition from origin *j* to destination *k*---I estimate multinomial logit propensity scores using covariates that influence an individual's likelihood of moving between strata: (a) father's educational level, (b) presence of both parents at age 15, (c) nationality, (d) age, (e) sex, and (f) ethnicity.
-
-These propensity scores are subsequently employed to adjust for selection into mobility when assessing its effect on market justice preferences.
-
-**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1958].
-The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").
-
-*Controls*
-
-Sociodemographic and attitudinal variables are included to control for potential composition effects in the population.
-In terms of sociodemographic characteristics, I incorporate per capita household income quantile, educational level (1=Primary or below, 2=Secondary, 3=Technical, 4=University or above), age (in years), and sex (1=Male, 2=Female), which have been shown to significantly influence market justice preferences [@castillo_socialization_2024; @lindh_public_2015].
-Regarding attitudinal variables, I include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and social mobility and meritocracy.
-
-## Methods
-
-Following Breen and Ermisch's [-@breen_effects_2024] strategy for estimating the causal effect of social mobility on market justice preferences, I employ linear regression models with inverse-probability weights (IPW) for mobility conditional on class of origin.
-This approach allows estimation of the average causal effect of moving from an origin class *j* to a destination class *k* by comparing individuals who actually reach *k* with their counterfactual outcome had they instead moved to an alternative destination *k′* [@breen_effects_2024].
-The resulting estimand is the average treatment effect on the treated (ATT).
-
-All the analyses will be conducted using R software.
-
-# References
-
-::: {#refs}
-:::
-
-::::

### 2025-11-02 paper/02-antecedents.qmd
Update varios

--- 2025-06-27 advances/informe3.qmd
@@
----
-title: "Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile"
-format:
-  html: 
-    code-fold: true
-    html-math-method: katex
-    number-sections: true
-    code-link: true
-    title-block-banner: true
-    theme:
-      - pretty.scss
-  pdf: 
-    geometry: margin=2cm
-    template-partials: 
-      - title.tex
-    keep-tex: true
-    include-in-header:
-      text: |
-        \usepackage[noblocks]{authblk}
-        \renewcommand*{\Authsep}{, }
-        \renewcommand*{\Authand}{, }
-        \renewcommand*{\Authands}{, }
-        \renewcommand\Affilfont{\small}
-    number-sections: true
-editor: source
-bibliography: "../input/bib/tesis-magister.bib"
-csl: "../input/bib/apa6.csl"
-author:
-  - name: Andreas Laffert Tamayo
-    orcid: 0000-0002-9008-2454
-    email: alaffertt@estudiante.uc.cl 
-    affiliations:
-      - ref: 1
-affiliations: 
-  - id: 1
-    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
-link-citations: true
-linestretch: 1.5       
-mainfont: Times New Roman
-fontsize: 13pt          
-fig-height: 4
-fig-width: 7.5
----
-
-```{r setup, include=FALSE}
- knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
-                       # , fig.align = 'center'
-                       )
- # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
- # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
- options(scipen=999)
- options(kableExtra.auto_format = FALSE)
- rm(list=ls())
- options(knitr.kable.NA = '')
- options(knitr.graphics.error = FALSE)
- Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
-```
-
-```{r}
-# This document was last modified at {{< modified >}}
-# 
-# and it was last rendered at {{< now >}}
-
-```
-
-# Introduction
-
-:::: {.justify}
-
-What is the legitimate extent of market inequality in the eyes of the public?
-Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016].
-In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023].
-This transformation reduced the role of public provision and increased the presence of private actors in core social services [@ferre_welfare_2023].
-Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a].
-Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @castillo_perceptions_2025; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
-Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].
-
-Existing literature shows that market justice preferences are shaped by both the economic and institutional context and individuals' positions within social stratification.
-Grounded in the notion that economic institutions influence people's normative attitudes [@immergut_theoretical_1998], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015].
-In such contexts, market justice preferences tend to rise as individuals "ascend" the social structure, with those in more privileged positions in terms of class, education, and income holding stronger preferences for market-based solutions compared to those in more disadvantaged or at-risk positions [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_socialization_2024; @otero_power_2024; @lindh_public_2015; @svallfors_political_2007].
-
-Market justice preferences are shaped not only by objective socioeconomic conditions but also by popular beliefs about inequality.
-Among these, meritocracy is a key normative principle underpinning market-based distributive preferences [@mau_inequality_2015].
-Studies show that individuals with stronger meritocratic beliefs tend to perceive less inequality and legitimize it by attributing economic differences to personal achievement [@mijs_paradox_2019; @wilson_role_2003; @batruch_belief_2023].
-In highly unequal societies where access to services is largely governed by market logic, such beliefs play a critical role in normalizing inequality.
-Recent evidence from Chile shows that students who believe effort and talent are rewarded in their country express stronger preferences for market-based access to healthcare, pensions, and education [@castillo_socialization_2024].
-
-Although it is clear that one's social position influences market justice preferences, the question of how upward or downward mobility within the social structure affects these preferences remains unanswered.
-This question is far from trivial, especially in Latin America, where many have experienced various forms of mobility amid high economic inequality and deep welfare privatization [@lopez-roldan_comparative_2021; @torche_intergenerational_2014; @ferre_welfare_2023].
-Social origins and destinations affect attitudes toward inequality in distinct ways [ @jaime-castillo_social_2019; @gugushvili_intergenerational_2016; @day_movin_2017; @gugushvili_subjective_2017; @wen_does_2021; @mijs_belief_2022], while movement between these positions exposes individuals to different experiences and mechanisms that shape their views on what is fair [@gugushvili_trends_2014; @mau_inequality_2015].
-Building on this research, examining the effects of social mobility on market justice preferences can help to illuminate how inequalities in access to social services are justified among individuals who have experienced, or not, changes in their social standing, and what are the normative mechanisms that guide this justification [@mau_inequality_2015].
-
-Beyond their isolated effects, social mobility and meritocratic beliefs interact in complex ways to shape market justice preferences.
-Among others, a key mechanism proposed in the literature to explain how mobility influences distributive justice preferences is the psychological process of self-serving bias in causal attribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c].
-This bias suggests that individuals attribute failures---such as downward mobility---to external factors, while crediting successes---such as upward mobility---to their own merit and effort [@miller_selfserving_1975].
-Those who experience upward mobility tend to view their social position as earned, making them more likely to believe that individuals are responsible for their own success or failure.
-Research shows that upward mobility is associated with weaker preferences for redistribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @jaime-castillo_social_2019; @alesina_intergenerational_2018] and stronger legitimacy of income inequality [@shariff_income_2016].
-In contrast, individuals who experience downward mobility tend to blame structural factors like inequality and are more supportive of redistribution while rejecting merit-based explanations [@gugushvili_trends_2014].
-Taken together, I argue that meritocratic beliefs may reinforce this self-serving attribution mechanism by legitimizing one's social status as the outcome of personal merit, closely tied to attribution bias.
-
-Against this background, this article pursues two main objectives: first, to analyze the extent to which intergenerational social mobility influences market justice preferences regarding healthcare, pensions, and education; and second, to examine how meritocratic beliefs may moderate this relationship.
-Building on a theoretical framework that emphasizes how neoliberal transformations---particularly through the privatization and commodification of key areas of social reproduction---have profoundly reshaped processes of subject formation [@mau_inequality_2015], the central argument is that upward mobility increases support for market justice preferences, while downward mobility decreases it.
-Moreover, meritocratic beliefs are expected to moderate this relationship by reflecting a self-serving attribution bias, whereby individuals justify their social position in terms of personal merit.
-
-This study focuses on Chile, a particularly intriguing case for examining market justice preferences.
-Despite significant economic growth and poverty reduction, Chile has some of the highest levels of inequality in Latin America and among OECD countries [@chancel_world_2022; @flores_top_2020].
-This inequality coexists with short-range upward mobility among lower-class segments moving into middle strata, though strong barriers remain to reaching higher positions [@torche_intergenerational_2014; @espinoza_estratificacion_2013].
-What makes Chile especially salient is that much of this inequality is rooted in deep neoliberal reforms that institutionalized the privatization and commodification of key social sectors [@madariaga_three_2020].
-Introduced during the dictatorship (1973--1989) and expanded in democracy, these reforms enabled the unprecedented emergence of markets in health, pensions, and education, with provision segmented by individuals' ability to pay and supported by public subsidies [@boccardo_30_2020].
-In parallel---and despite waves of protest against inequality and commodification from 2006 to 2019 [@somma_no_2021]---Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing their attitudes toward inequality and welfare distribution [@araujo_desafios_2012; @canalesceron_sujeto_2021].
-
-In this context, the questions that guide this research are as follows:
-
-(1) To what extent does intergenerational social mobility influence market justice preferences regarding healthcare, pensions, and education in Chile?
-(2) How do meritocratic beliefs condition or moderate this relationship in the Chilean context?
-
-To address these questions, this study draws on large-scale, representative survey data collected in 2018 from the urban Chilean population aged 18 to 75 (n = 2,726).
-The next section outlines the theoretical framework linking market justice preferences, social mobility, and meritocratic beliefs, and proposes a set of hypotheses.
-This is followed by a description of the data, variables, and analytical strategy.
-The final sections present the empirical findings, offer an interpretation of the results, and conclude with a discussion of their implications.
-
 # Theoretical and empirical background
@@
 [@castillo_perceptions_2025; @castillo_socialization_2024] find that individuals who endorse meritocratic beliefs—such as the idea that effort and talent are rewarded in Chile—are more likely to support market justice across welfare domains.
-
-# Data, variables and methods
-
-## Data
-
-This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
-The survey is an annual panel collected from 2016 to 2023, comprising two independent samples (original and refreshment), featuring permanent and rotating questionnaire modules.
-It evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile.
-The sampling design is complex, probabilistic, clustered, multistage, and stratified by city size, targeting men and women aged 18--75 who habitually reside in private dwellings in urban areas spanning 40 cities (92 municipalities, 13 regions).
-This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample.
-Further details on sampling, attrition, and weighting can be found at [https://coes.cl/encuesta-panel/](https://coes.cl/encuesta-panel/), and the dataset is publicly available at [https://dataverse.harvard.edu/dataverse/elsoc](https://dataverse.harvard.edu/dataverse/elsoc).
-
-## Variables
-
-*Outcome variables*
-
-**Market justice preferences**: The outcome variables in this study are market justice preferences.
-This construct is operationalized through three variables that address the degree of justification regarding whether access to social services in healthcare, pensions, and education should be income conditional.
-Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is asked for pensions and education.
-In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree").
-Although these items theoretically reflect the same underlying concept of market justice [@castillo_socialization_2024; @lindh_public_2015], they are analyzed separately because healthcare, pensions, and education may prompt distinct response patterns [@lee_fairness_2023; @busemeyer_skills_2014; @immergut_it_2020].
-
-*Independent variables*
-
-**Social mobility**: Intergenerational social mobility is treated as an exposure indicating whether respondents occupy a different class position from their fathers.
-Following Breen and Ermisch's [-@breen_effects_2024] framework for estimating causal effects of mobility, the variable is constructed in two stages.
-
-1.
-Class assignment.
-Using the International Socio-Economic Index of Occupational Status (ISEI) derived from three-digit ISCO-08 codes, both paternal (origin) and respondent (destination) occupations are grouped into three strata:
-
-    -   Upper (codes 100-299: managers, directors, professionals)
-    -   Middle (codes 300-499: skilled white-collar workers)
-    -   Lower (codes 500 and above: manual, semi-skilled, or unskilled workers)
-
-2.
-Propensity-score estimation.
-To model mobility as a treatment---i.e., a transition from origin *j* to destination *k*---I estimate multinomial logit propensity scores using covariates that influence an individual's likelihood of moving between strata: (a) father's educational level, (b) presence of both parents at age 15, (c) nationality, (d) age, (e) sex, and (f) ethnicity.
-
-These propensity scores are subsequently employed to adjust for selection into mobility when assessing its effect on market justice preferences.
-
-**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1958].
-The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").
-
-*Controls*
-
-Sociodemographic and attitudinal variables are included to control for potential composition effects in the population.
-In terms of sociodemographic characteristics, I incorporate per capita household income quantile, educational level (1=Primary or below, 2=Secondary, 3=Technical, 4=University or above), age (in years), and sex (1=Male, 2=Female), which have been shown to significantly influence market justice preferences [@castillo_socialization_2024; @lindh_public_2015].
-Regarding attitudinal variables, I include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and social mobility and meritocracy.
-
-## Methods
-
-Following Breen and Ermisch's [-@breen_effects_2024] strategy for estimating the causal effect of social mobility on market justice preferences, I employ linear regression models with inverse-probability weights (IPW) for mobility conditional on class of origin.
-This approach allows estimation of the average causal effect of moving from an origin class *j* to a destination class *k* by comparing individuals who actually reach *k* with their counterfactual outcome had they instead moved to an alternative destination *k′* [@breen_effects_2024].
-The resulting estimand is the average treatment effect on the treated (ATT).
-
-All the analyses will be conducted using R software.
-
-# References
-
-::: {#refs}
-:::
-
-::::

### 2025-11-02 paper/03-methods.qmd
Update varios

---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup2, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")

 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
```

```{r}
#| label: libraries1
#| message: false
#| echo: false

if (!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               here,
               sjPlot,
               sjmisc,
               texreg,
               estimatr,
               conflicted,
               ggdist,
               summarytools,
               dagitty,
               ggdag)

conflicts_prefer(dplyr::filter)
options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data1
#| message: false
#| echo: false

load(here("input/data/proc/data_v2.3.RData"))

```

# Method

Social mobility effects—understood as impacts on an outcome arising from movements between an origin state and a destination state (e.g., social class)—have long been a focus of sociological research [@eyles_social_2022; @langsaether_explaining_2022]. Yet, as Breen and Ermisch [-@breen_effects_2024, p.467] emphasize, most hypotheses are, at their core, individual-level counterfactual comparisons between the observed outcome under a mobility trajectory and the outcome the same person would have shown if they had remained in their origin class (or moved to an alternative destination). Standard specifications (e.g., SAM, DRM, and mobility-contrast models) struggle to retrieve these inherently counterfactual quantities due to identification constraints and their reliance on between-group contrasts constructed from additive terms and interactions, thereby yielding primarily associational evidence [@breen_effects_2024; @song_there_2025]. I therefore adopt Breen and Ermisch’s [-@breen_effects_2024] causal formulation, treating destination class as the treatment, conditioning on the class of origin, and estimating heterogeneous destination effects with observational data under explicit identification assumptions. To align design and target, I follow the MIDA template [@blair_research_2023]: set out the causal model and assumptions (M), define the inquiry and estimand (I), describe the data and variables (D), and detail the answer strategy (A) used to identify the causal effect of intergenerational social mobility on preferences for the commodification of pensions.

## A Causal Model for Mobility Effects

Breen and Ermisch [-@breen_effects_2024], building on a critical reassessment of the inferential limits of standard mobility models, propose a causal framework that reconceptualizes the “mobility effect” as the treatment effect of reaching a destination class, with impacts heterogeneous by class of origin. The core claim is that typical mobility hypotheses are fundamentally within-person counterfactual comparisons rather than between-person contrasts. Defining mobility as $(M = D - O)$ —a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination—places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the outcome under the realized destination but not under alternative destinations or under immobility.

Exploiting the temporal ordering of origin, destination and outcome, Breen and Ermisch [-@breen_effects_2024] frame mobility as a treatment process that motivates the causal question: "how would the outcome among people from origin *j* who entered destination *k* have been different if those people had, counterfactually, entered destination *k’* instead?" (p.472). The corresponding estimand is the conditional causal effect of destination given origin. It is of particular interest when *k' = j*, i.e., immobility. This estimand compares movers’ observed outcomes with the hypothetical outcomes those same individuals would have exhibited had they remained in their origin class. This formulation provides a coherent potential-outcomes basis for studying how class transitions shape individual preferences and attitudes.

Following Breen and Ermisch [-@breen_effects_2024], the identification of causal mobility effects from observational data requires the following assumptions:

**Positivity**: For all $(j,c)$ in the support of $(O,C)$ and for each relevant destination $(k)$, the probability of receiving $D=k$ is strictly between 0 and 1; substantively, every $(O,C)$-type has some chance of each comparison destination.

$$
 0<P(D=k\mid O=j, C=c)<1
$$ {#eq-posi}

**Stable Unit Treatment Value Assumption (SUTVA)**: each unit's potential outcomes $Y_i(D)$ does not depend on the mechanism used to assign treatments (destinations) and by the treatments assigned to other units (also called no interference), assuming a single, well-defined version of each treatment (consistency).

$$
Y_i(D_i) = Y_i\big(D_i, D_{-i}\big)\quad \forall, D_{-i}
$$ {#eq-sutva}

**Conditional Independence**: Also known as conditional unconfoundedness or exchangeability, this assumption emphasizes that, conditional on origin $O$ and pre-treatment covariates $C$ (e.g., parental education, ethnicity, cohort, early-life factors), assignment to $D$ is as good as random. This underlies IPW and regression adjustment, which seek exchangeability between mobile and immobile individuals.

$$
Y(D)\ \perp\kern-5pt \perp D\ \mid\ (O,C)
$$ {#eq-indep}

@fig-dag depicts the causal model guiding identification. In this DAG, $C$ represents pre-treatment confounders (e.g., family resources, household composition, and ascriptive status) affecting both origin and destination, and $U$ denotes unobserved factors influencing $C$. Class origin $O$ is determined by $C$ and, in turn, affects destination $D$. The outcome $Y$ (preferences for commodification) is caused by $D$ and can also be affected by $C$. There is no direct arrow from $O$ to $Y$, reflecting the assumption that the effect of origin on the outcome operates entirely through background factors $(C)$ and destination $(D)$.  Accordingly, conditioning on $(O,C)$ blocks the relevant back-door paths from $D$ to $Y$, specifically $D\leftarrow O\leftarrow C\rightarrow Y$ and, if present, $D \leftarrow O \rightarrow Y$, so $(O,C)$ is a minimal sufficient adjustment set.

```{r}
#| label: fig-dag
#| out-width: '85%'
#| fig-cap: "Causal graph of the effect of intergenerational social mobility on preferences for commodification. Y = preferences for pension commodification, my outcome of interest, D = an individual’s class of destination, O = an individual’s class of origin, C = different attributes determined in childhood or earlier that affects an individual’s class of origin and destination. Finally, U = unobserved factors influencing origin attributes."
#| fig-cap-location: bottom
#| echo: false
#| warning: false
#| message: false

dag_mjp <- dagify(
  Y ~ D + C,
  D ~ O + C,
  O ~ C,
  C ~ U,
  coords = time_ordered_coords(
    list(
      "U",
      c("C", "O"),
      "D",
      "Y"
    )
  ),
  exposure = "D",
  outcome = "Y",
  labels = c(
    Y = "Y",
    D = "D",
    O = "O",
    C = "C",
    U = "U"
  )
)

curvatures <- rep(0, 8)
curvatures[3] <- -.25

dag_mjp |>
  tidy_dagitty() |>
  ggplot(aes(x, y, xend = xend, yend = yend)) +
  geom_dag_point(shape = 21, fill = "black", size = 14, stroke = 0.8) +
  geom_dag_edges_arc(curvature = curvatures) +
  geom_dag_text(aes(label = label), color = "white", fontface = "bold", size = 4) +
  theme_dag()

```

Formally, the DAG encodes the following relations:

$$
U \rightarrow C,\quad C \rightarrow O,\quad C \rightarrow D,\quad C \rightarrow Y,\quad O \rightarrow D,\quad D \rightarrow Y
$$

This structure corresponds to a nonparametric structural causal model in which $(D)$ is the treatment variable, $(Y)$ the outcome, and $(O, C)$ the minimal sufficient adjustment set. Conditioning on these variables satisfies the back-door criterion for identifying the causal effect of mobility on preferences [@breen_effects_2024].

Why are the assumptions plausible here? Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$; entropy balancing/IPW enforces covariate balance within each origin stratum. Positivity is assessed via propensity scores and trimming of extreme weights, restricting to common support. SUTVA holds by treating the destination class as a single exposure measured before $(Y)$ and assuming no interference. Together, the DAG and assumptions define conditions for identifying the causal effect of intergenerational mobility on preferences for commodification.

## Inquiry and estimand

The causal inquiry guiding this study asks: How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile? Following the potential outcomes framework proposed by Breen and Ermisch [-@breen_effects_2024, p.473], the estimand of interest is the average treatment effect on the treated (ATT), defined as the mean causal effect of moving from an origin class $(O=j)$ to a destination class $(D=k)$, compared with the counterfactual outcome that those same individuals would have exhibited had they instead reached an alternative destination $(k')$:

$$ 
ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
$$ {#eq-breen}

In this study, I focus on the specific case where the counterfactual destination corresponds to class immobility $(k' = j)$, that is, how the outcome among people from origin *k* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility). Accordingly, the estimand becomes:

$$ 
ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
$$ {#eq-att}

which represents the average causal effect of attaining class *(k)*—for instance, moving upward or downward—relative to remaining in class *(j)*, among individuals who actually experienced such mobility. This estimand captures the within-origin counterfactual contrast at the heart of mobility effects: how a person’s preference for market-based pensions would differ if, counterfactually, they had stayed in their origin class rather than moved to their observed destination.

## Data and variables

### Data

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES). This survey is a nationally representative panel study of the urban adult population in Chile, conducted annually between 2016 and 2023. Designed to examine individuals’ attitudes, emotions, and behaviors regarding social conflict and cohesion, ELSOC employs a probabilistic, stratified, clustered, and multistage sampling design covering both major urban centers and smaller cities. The sampling frame was proportionally stratified into six categories of urban population size (e.g., large and small cities), followed by a random selection of households within 1,067 city blocks. The target population includes men and women aged 18 to 75 who are habitual residents of private dwellings.

Because respondent occupation is not measured in every wave, I restrict the analysis to the 2016, 2018, and 2023 waves. Parental occupation was collected only in 2023 as open-text labels. I coded these texts to the ISCO-08 two-digit scheme using the National Institute of Statistics of Chile’s automated coding API and retained cases with high-confidence matches (>95%). Treating parental occupation as a time-invariant origin attribute, I carried these codes back to 2016 and 2018. After listwise deletion and restricting to key variables (respondent occupation and the outcome measure), the final analytic sample comprises *N* = 3,435 observations nested within *N* = 1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with the study design, estimation proceeds within trajectory-specific subsamples (e.g., low→low; low→middle), retaining movers and matched non-movers from the same origin class. Then, I apply weights to construct the corresponding counterfactual control distributions for each trajectory. As Breen and Ermisch [-@breen_effects_2024] did, I use the three waves in the estimations that follow, allowing for correlation between the observations for each individual in calculating the standard errors (i.e. clustering on the personal identifier).

### Outcome variable

The outcome variable measures preferences regarding the commodification of pensions, operationalized with a single item addressing the degree of public support for market-based access to old-age pension benefits. Respondents were asked: “Is it fair in Chile that people with higher incomes have better pensions than people with lower incomes?” and were asked to indicate their level of agreement on a five-point Likert scale ranging from 1 (“strongly disagree”) to 5 (“strongly agree”). For estimation and interpretability, I dichotomized the item so that values 4-5 indicate agreement with market-based access to old-age pensions (coded as 1), while values 1-3 indicate disagreement (coded as 0). This binary coding treats neutrality as substantially closer to non-support than to support and produces a clear contrast between respondents who support market-based differentiation in pensions and those who do not. I use this item for two main reasons: on the one hand, to be able to make comparisons with existing literature [@lindh_public_2015; @otero_power_2024; @castillo_perceptions_2025] and, on the other hand, because it captures two fundamental dimensions of market-based welfare distribution: the role of economic resources as a key determinant of outcomes and the framing of services as as tradable commodities that can be bought and sold based on ability to pay [@lindh_public_2015].

@tbl-summary1 shows the descriptive statistics of the variables used in the study.

```{r}
#| label: tbl-summary1
#| tbl-cap: "Descriptive statistics for all variables."
#| tbl-cap-location: top
#| results: asis
#| echo: false

t1 <- db %>% 
  select(just_pension_f, 
         estrato_orig, 
         estrato_ocupa, 
         educ_orig_f,
         hogar_bip_f, 
         nacionalidad_f, 
         sexo_f,
         m0_edad_f, 
         etnia_f, 
         merit_i,
         wave

)

t1 <- t1 %>% 
  mutate(
    just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree")),
    hogar_bip_f  = factor(hogar_bip_f, levels = c(0,1), labels = c("No co-residence", "Co-residence")),
    #nacionalidad_f = if_else(nacionalidad_f == 0, 1, 0),
    nacionalidad_f  = factor(nacionalidad_f, levels = c(0,1), labels = c("Non-Chilean", "Chilean")),
    #sexo_f = if_else(sexo_f == 1, 0, 1),
    sexo_f  = factor(sexo_f, levels = c(0,1), labels = c("Male", "Female")),
    etnia_f = factor(etnia_f, levels = c(0,1), labels = c("Non-indigenous", "Indigenous"))
  )

t1$just_pension_f <- sjlabelled::set_label(t1$just_pension_f, "Preference for pension commodification")
t1$educ_orig_f <- sjlabelled::set_label(t1$educ_orig_f, "Parental education")
t1$hogar_bip_f <- sjlabelled::set_label(t1$hogar_bip_f, "Co-residence with both parents at age 15")
t1$nacionalidad_f <- sjlabelled::set_label(t1$nacionalidad_f, "Nacionality")
t1$sexo_f <- sjlabelled::set_label(t1$sexo_f, "Sex")
t1$m0_edad_f <- sjlabelled::set_label(t1$m0_edad_f, "Age (in years)")
t1$etnia_f <- sjlabelled::set_label(t1$etnia_f, "Indigenous ethnicity")
t1$merit_i <- sjlabelled::set_label(t1$merit_i, "Meritocracy perception")
t1$wave <- sjlabelled::set_label(t1$wave, "Wave")

df<-dfSummary(t1,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)

```

### Treatment

#### Intergenerational occupational mobility

I treat intergenerational occupational mobility as an exposure indicating whether respondents occupy a different socioeconomic status than their fathers, closely following Breen and Ermisch’s [-@breen_effects_2024] causal framework. Class assignment proceeds in two steps. First, I derive status for both origin (father) and destination (respondent) from two-digit ISCO-08 occupations using the International Socio-Economic Index of Occupational Status (ISEI). ISEI is a hierarchical, unidimensional scale that locates occupations by incumbents’ average education and earnings and was explicitly designed to harmonize occupational stratification across other class schemes and socioeconomic status measures [@ganzeboom_internationally_1996], thereby supporting cross-study comparability [@barozet_measurement_2021a]. Second, I group both origin and destination ISEI scores into terciles (low, middle, high), yielding a three-category schema for origin and destination stratums (see @tbl-matrix). Substantively, I use ISEI because occupation-based measures capture structural locations that organize life chances and attitudes more directly than income alone: they bundle access to the means of production, qualification credentials, and the authority/power inherent in labor relations; central dimensions of stratification and inequality [@barone_rise_2022; @wrightUnderstandingClass2015]. In short, using ISEI aligns the measurement of origin and destination on a common vertical status continuum that is theoretically meaningful for social mobility research and empirically feasible given the available information of ELSOC’s data.

```{r}
#| label: tbl-matrix
#| tbl-cap: "Occupational mobility by occupational groups."
#| tbl-cap-location: top
#| results: asis
#| echo: false

library(janitor)

db %>%
  tabyl(estrato_orig, estrato_ocupa) %>%
  adorn_totals(where = c("row", "col")) %>%     # agrega totales
  adorn_percentages("row") %>%                  # porcentajes por fila
  adorn_pct_formatting(digits = 1) %>%          # formato %
  adorn_ns() %>% 
  as_tibble() %>% 
  mutate(Offspring = NA) %>% 
  select(estrato_orig, Offspring, everything()) %>% 
  kableExtra::kable(format = table_format, 
                    booktabs= T,
                    align = 'c', 
                    col.names = c("Father↓", "Offspring→", "Low", "Middle", "High", "Total"),
                    caption = NULL) %>%
  kableExtra::kable_styling(latex_options = "hold_position", 
                            bootstrap_options = "striped",
                            full_width = F)

```

#### Pre-treatment control variables for selection into treatment

Within each origin stratum $O=j$, I estimate average treatment effects on the treated (ATT) by comparing movers to otherwise similar non-movers from the same origin. To adjust for selection into mobility, I reweight only the control group (immobile) using entropy balancing so that the distribution of pre-treatment origin covariates $C$ matches that of the treated on selected moments.

The selection of the adjustment set $C$ follows two principles. First, it mirrors the logic in Breen and Ermisch [-@breen_effects_2024]—who condition on attributes determined in childhood or earlier (e.g., parental resources, household structure, ascriptive traits)—and, second, it is constrained to origin-side characteristics available in ELSOC. Guided by theory and Chilean evidence on relative mobility determinants [mainly @brunori_inequality_2025; @celhay_intergenerational_2010; @espinoza_movilidad_2014;@torche_unequal_2005], I include six covariates covering family resources, household composition, and ascriptive status: (a) parental education (highest of father/mother), 10 categories coded ordinally from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes). These variables are fixed before the destination class, plausibly affect selection into mobility, and constitute the maximal origin set in ELSOC, thereby making the conditional independence assumption more credible within each $O=j$.

After balancing, I construct stabilized IPW-ATT, setting treated weights to 1 and assigning ebal-derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum. These weights are then used in the outcome stage to estimate the causal effect of mobility on preferences for the commodification of pensions.

### Effect heterogeneity

To test heterogeneity in mobility effects, I examine two moderators: meritocratic perceptions and time. Meritocracy is captured with two items [@young_rise_1958], one referring to effort (“In Chile, people are rewarded for their efforts”) and one to talent (“In Chile, people are rewarded for their intelligence and skills”), each answered on a five-point Likert scale (1 = “strongly disagree” to 5 = “strongly agree”). I average the two items into a single index and dichotomize it into low meritocracy (≤ 3) and high meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023) and entered as a categorical (dummy) variable. Both variables are used as moderators by interacting them with the mobility treatment to assess conditional average treatment effects—i.e., whether the effect of mobility varies across levels of meritocratic beliefs or across survey waves.

### Controls

All models include the same pre-treatment covariates $C$ used in the IPW construction, not to re-balance groups (entropy balancing already does so), but to (i) block backdoor paths from $C$ to the outcome $Y$ as implied by the DAG, and (ii) achieve double robustness: estimates remain consistent if either the weighting model or the outcome model is correctly specified. Concretely, I adjust for (a) father’s educational level, (b) co-residence with both parents at age 15, (c) nationality, (d) age, (e) sex, and (f) ethnicity, along with wave fixed effects to absorb secular trends.

## Analytical strategy

Because the dependent variable—the preference for pension commodification—is binary $(Y_{it}\in{0,1})$, I estimate weighted linear probability models (WLS) using stabilized inverse probability weights (IPW–ATT) obtained via entropy balancing. Although OLS estimator is often questioned with binary outcomes, it’s unbiased. It consistently estimates the conditional expectation $E(Y|X)$ under standard exogeneity $E[\varepsilon_i|X_i]=0$, with heteroskedasticity addressed by robust or clustered errors [@wooldridge_introductory_2009, ch. 7.5]. WLS further improves efficiency under heteroskedasticity or unequal sampling probabilities [@wooldridge_introductory_2009, ch. 8.4]. Following Gelman, Hill, and Vehtari [-@gelman_regression_2020, pp. 270–272], IPW-weighted regression transparently targets the ATT by recreating the counterfactual distribution. Linear models are preferred here because coefficients are average probability differences; nonlinear models (logit/probit) complicate weighting and interpretation [@gelman_regression_2020, ch. 13].

Formally, the model is:

$$
Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it},
$$ {#eq-final}

where $Y_{it}$ indicates whether individual $i$ in wave $t$ supports more market-based pension access; $\alpha$ is the baseline probability for the reference categories; $\beta T_i$ captures the intergenerational mobility contrast: $T_i=1$ when the observed destination is $D_i=k$ (mobile to $k$) and $T_i=0$ when $D_i=j$ (immobile), so that $\beta$ is the ATT within origin $O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})$, i.e., the percentage-point change in the probability of preferring pension commodification from reaching $k$ rather than remaining in $j$. The term $X_i'\gamma$ includes the pre-treatment covariates $(C)$ used to construct the entropy-balancing weights; $\lambda_t$ are wave fixed effects; and $\varepsilon_{it}$ is idiosyncratic error. The analytic sample is restricted to individuals sharing the same origin $(O=j)$. Estimation uses pooled WLS with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds if either the weighting model or the outcome model is correctly specified, while also improving efficiency [@wooldridge_introductory_2009]. I report both IPW-only and doubly robust specifications to document robustness. In addition, I assess causal heterogeneity by interacting the mobility treatment with (i) meritocratic perceptions and (ii) survey wave, reporting simple slopes (origin-specific ATTs) by moderator levels. To evaluate sensitivity to unmeasured confounding, I compute E-values for point estimates and confidence limits [@ding_sensitivity_2016]. To address multiplicity across the six mobility trajectories, I also report Holm-adjusted p-values alongside 95% confidence intervals.

All the analyses were conducted using R software and the _lm_robust_ package.

### 2025-11-02 paper/04-analysis.qmd
Update varios

---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Results

```{r setup3, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")

 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }
```

```{r}
#| label: libraries2
#| message: false
#| echo: false

if (!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               here,
               sjPlot,
               sjmisc,
               texreg,
               estimatr,
               conflicted,
               ggdist,
               summarytools,
               dagitty,
               ggdag)

conflicts_prefer(dplyr::filter)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data2
#| message: false
#| echo: false

load(here("input/data/proc/data_v2.3.RData"))

```

## Descriptive statistics

Lorem ipsoThe specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustnessThe specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness

```{r}
#| label: fig-alluvial
#| fig-cap: "Change in preferences for pension commodification over time (2016, 2018, and 2023)."
#| fig-cap-location: bottom
#| echo: false
#| results: asis
db %>% 
  mutate(just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree"))) %>%
  group_by(wave, just_pension_f) %>% 
  count() %>% 
  group_by(wave) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  ggplot(aes(x = wave, y = porcentaje, group = just_pension_f)) +
  geom_col(aes(fill = just_pension_f)) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(
    values = c("#5b59d4", "#c8237d"),
    breaks = c("Agree","Disagree")
  )  +
  coord_flip() +
  geom_text(
    aes(label = paste0(round(porcentaje * 100, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "white",
    size = 3,
    fontface = "bold"
  ) +
  labs(y = NULL,
       x = NULL,
       fill = NULL,
       caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)") +
  theme_ggdist() +
  theme(legend.position = "top") 
```

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness

```{r}
#| label: fig-mean
#| fig-cap: "Percentage agreement on preferences for pension commodification according to mobility trajectory and time (2016, 2018, and 2023)."
#| out-width: '90%'
#| fig-cap-location: bottom
#| echo: false
#| results: asis

pdat <- db %>% 
  select(wave, estrato_orig, estrato_ocupa, just_pension_f) %>% 
  drop_na() %>% 
  mutate(
    mobility = paste(estrato_orig, estrato_ocupa, sep = "-"),
    facet = case_when(
      mobility %in% c("High-Middle","High-Low","Middle-Low") ~ "Downward",
      estrato_orig == estrato_ocupa ~ "Inmobile",
      mobility %in% c("Middle-High","Low-Middle","Low-High") ~ "Upward"
    )
  ) %>% 
  group_by(wave, mobility, facet) %>% 
  summarise(y = mean(just_pension_f), .groups = "drop") %>% 
  group_by(mobility, facet) %>% 
  mutate(gm = mean(y)) %>% 
  ungroup()

# datos SOLO para el punto (uno por categoría)
pdat_gm <- pdat %>% distinct(mobility, facet, gm)

ggplot(pdat, aes(x = mobility, y = y, group = wave)) +
  geom_col(aes(fill = wave), position = "dodge", alpha = 0.9) +
  # Punto "X" (pch 4) con su propia leyenda (shape)
  geom_point(
    data = pdat_gm,
    aes(x = mobility, y = gm, shape = "All waves"),
    inherit.aes = FALSE,
    size = 3, stroke = 1, color = "black"
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
  scale_fill_manual(
    values = c("2016" = "#5b59d4", "2018" = "#b3b3b3ff", "2023" = "#c8237d"),
    name   = "Wave"
  ) +
  scale_shape_manual(                     # <-- leyenda separada para el shape
    name = NULL,                          # o "Marker" si quieres título
    values = c("All waves" = 4)          # 4 = "X"
  ) +
  guides(
    fill  = guide_legend(order = 1),
    shape = guide_legend(order = 2, override.aes = list(color = "black"))
  ) +
  facet_wrap(~ facet, scales = "free_x") +
  labs(
    y = "% of Agreement",
    x = NULL,
    caption = "Source: own elaboration with data from ELSOC 2016–2023 (N obs = 3,435)"
  ) +
  theme_ggdist() +
  theme(legend.position = "top")

```

## Models

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness

```{r}
#| label: models
#| echo: false
#| results: false

### Pensiones

### Low to middle

data_lm <- datos1 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = moblm, 
         w = w_low_lm_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_lm <-
  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0 + ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_lm)

### Low to high

data_lh <- datos1 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = moblh, 
         w = w_low_lh_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_lh <- 
  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_lh)

### middle to high

data_mh <- datos2 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobmh, 
         w = w_mid_mh_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_mh <-  
  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_mh)

### middle to low

data_ml <- datos2 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobml, 
         w = w_mid_ml_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_ml <- 
  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_ml)

### High to middle

data_hm <- datos3 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobhm, 
         w = w_hig_hm_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_hm <- 
  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_hm)

### High to low

data_hl <- datos3 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobhl, 
         w = w_hig_hl_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_hl <- 
  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0 + ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_hl)

```

```{r}
#| label: fig-reg
#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification. Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility. Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual; bars show 95% confidence intervals."
#| out-width: '90%'
#| fig-cap-location: bottom
#| echo: false
#| results: asis

conflicts_prefer(dplyr::filter)

df_pension <- bind_rows(
  tidy(mp_lm, conf.int = TRUE) %>%
    dplyr::filter(term == "t") %>%
    dplyr::mutate(term = paste0("Low-Middle\n", "(N obs. = ", mp_hl$nobs, ";", "\nN clusters = ", mp_hl$nclusters, ")")
) 
  ,
  tidy(mp_lh, conf.int = TRUE) %>%
    dplyr::filter(term == "t") %>%
  dplyr::mutate(term = paste0("Low-High\n", "(N obs. = ", mp_lh$nobs, ";", "\nN clusters = ", mp_lh$nclusters, ")"))
  ,
tidy(mp_mh, conf.int = TRUE) %>%
    dplyr::filter(term == "t") %>%
    dplyr::mutate(term = paste0("Middle-High\n", "(N obs. = ", mp_mh$nobs, ";", "\nN clusters = ", mp_mh$nclusters, ")"))
,
  tidy(mp_ml, conf.int = TRUE) %>%
    dplyr::filter(term == "t") %>%
  dplyr::mutate(term = paste0("Middle-Low\n", "(N obs. = ", mp_ml$nobs, ";", "\nN clusters = ", mp_ml$nclusters, ")"))
,
  tidy(mp_hm, conf.int = TRUE) %>%
    dplyr::filter(term == "t") %>%
  dplyr::mutate(term = paste0("High-Middle\n", "(N obs. = ", mp_hm$nobs, ";", "\nN clusters = ", mp_hm$nclusters, ")"))
,
  tidy(mp_hl, conf.int = TRUE) %>%
    dplyr::filter(term == "t") %>%
  dplyr::mutate(term = paste0("High-Low\n", "(N obs. = ", mp_hl$nobs, ";", "\nN clusters = ", mp_hl$nclusters, ")"))
)

df_pension_lab <- df_pension %>%
  dplyr::mutate(
    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
    est_lab = sprintf("%.2f%s", estimate, sig),
    low_lab = sprintf("%.2f", conf.low),
    high_lab= sprintf("%.2f", conf.high)
  )

df_pension_lab <- df_pension_lab %>%
  dplyr::mutate(
    term = factor(
      term,
      levels = term
      )
  )

ggplot(df_pension_lab, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 0.5, alpha = 1) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 0.7) +
  geom_point(size = 2.5) +
  scale_x_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +
  #geom_text(aes(label = est_lab), hjust = -0.05, size = 4, vjust = -0.5) +
  labs(x = "Estimate", y = NULL) +
  theme_ggdist() 
```

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustnessThe specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness

## Effect heterogeneity

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
```{r}
#| echo: false
#| results: false

# 3.4 Interacciones -------------------------------------------------------

# -----------
# Meritocracy 
# -----------

### Interacciones

### Low to middle

data_lm_i <- datos1 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = moblm, 
         w = w_low_lm_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f,
         m = merit_i) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

data_lm_i %>% 
  arrange(idencuesta)

mp_lm_i <-
  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0 + ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_lm_i)

mp_lm_i %>%  
  screenreg()

### Low to high

data_lh_i <- datos1 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = moblh, 
         w = w_low_lh_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f,
         m = merit_i) %>%  
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_lh_i <- 
  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_lh_i)

mp_lh_i %>% 
  screenreg()

### middle to high

data_mh_i <- datos2 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobmh, 
         w = w_mid_mh_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f,
         m = merit_i) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_mh_i <-  
  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_mh_i)

mp_mh_i %>% 
  screenreg()

### middle to low

data_ml_i <- datos2 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobml, 
         w = w_mid_ml_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f,
         m = merit_i) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_ml_i <- 
  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_ml_i)

mp_ml_i %>% 
  screenreg()

### High to middle

data_hm_i <- datos3 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobhm, 
         w = w_hig_hm_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f,
         m = merit_i) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_hm_i <- 
  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_hm_i)

mp_hm_i %>% 
  screenreg()

### High to low

data_hl_i <- datos3 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobhl, 
         w = w_hig_hl_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f,
         m = merit_i) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")))

mp_hl_i <- 
  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0 + ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_hl_i)

mp_hl_i %>% 
  screenreg()
```

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness

```{r}
#| label: fig-interact
#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by meritocratic perception. Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility, estimated from interaction models (T × Merit). Points show simple slopes of the mobility effect at low merit and high merit; bars are 95% confidence intervals. Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual."
#| out-width: '90%'
#| fig-cap-location: bottom
#| echo: false
#| results: asis

df_pension_i <-  bind_rows(
  tidy(mp_lm_i, conf.int = TRUE) %>%
    filter(term %in% c("t:mHigh", "t")) %>%
    mutate(name = paste0("Low-Middle\n", "(N obs. = ", mp_hl_i$nobs, ";", "\nN clusters = ", mp_hl_i$nclusters, ")")
    ) 
  ,
  tidy(mp_lh_i, conf.int = TRUE) %>%
    filter(term %in% c("t:mHigh", "t")) %>%
    mutate(name = paste0("Low-High\n", "(N obs. = ", mp_lh_i$nobs, ";", "\nN clusters = ", mp_lh_i$nclusters, ")"))
  ,
  tidy(mp_mh_i, conf.int = TRUE) %>%
    filter(term %in% c("t:mHigh", "t")) %>%
    mutate(name = paste0("Middle-High\n", "(N obs. = ", mp_mh_i$nobs, ";", "\nN clusters = ", mp_mh_i$nclusters, ")"))
  ,
  tidy(mp_ml_i, conf.int = TRUE) %>%
    filter(term %in% c("t:mHigh", "t")) %>%
    mutate(name = paste0("Middle-Low\n", "(N obs. = ", mp_ml_i$nobs, ";", "\nN clusters = ", mp_ml_i$nclusters, ")"))
  ,
  tidy(mp_hm_i, conf.int = TRUE) %>%
    filter(term %in% c("t:mHigh", "t")) %>%
    mutate(name = paste0("High-Middle\n", "(N obs. = ", mp_hm_i$nobs, ";", "\nN clusters = ", mp_hm_i$nclusters, ")"))
  ,
  tidy(mp_hl_i, conf.int = TRUE) %>%
    filter(term %in% c("t:mHigh", "t")) %>%
    mutate(name = paste0("High-Low\n", "(N obs. = ", mp_hl_i$nobs, ";", "\nN clusters = ", mp_hl_i$nclusters, ")"))
)

df_pension_lab_i <- df_pension_i %>%
  mutate(
    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
    est_lab = sprintf("%.2f%s", estimate, sig),
    low_lab = sprintf("%.2f", conf.low),
    high_lab= sprintf("%.2f", conf.high)
  )

df_pension_lab_i <- df_pension_lab_i %>%
  mutate(
    name = factor(
      name,
      levels = unique(name)
    ),
    term = if_else(term == "t", "Low Meritocracy", "High Meritocracy"),
    term = factor(term, levels = c("High Meritocracy", "Low Meritocracy"))
  )

ggplot(df_pension_lab_i, aes(x = estimate, y = name)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 0.5, alpha = 1) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 0.7) +
  geom_point(size = 2.5) +
  scale_x_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +
  facet_wrap(~term) + 
  #geom_text(aes(label = est_lab), hjust = -0.05, size = 4, vjust = -0.5) +
  labs(x = "Estimate", y = NULL) +
  theme_ggdist()

```

```{r}
#| label: tbl-inttime
#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by survey wave"
#| tbl-cap-location: top
#| results: asis
#| echo: false

# -----------
# Time 
# -----------

# Low to middle

mp_lm_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
          weights = w,
          se_type = "CR2",
          clusters = idencuesta,
          data = data_lm)

# Low to high

mp_lh_t <- lm_robust(y ~ t*ola+ edad + sexo + nac + etnia + hogar + edu0,
          weights = w,
          se_type = "CR2",
          clusters = idencuesta,
          data = data_lh)

# Middle to high

mp_mh_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
          weights = w,
          se_type = "CR2",
          clusters = idencuesta,
          data = data_mh)

# Middle to low

mp_ml_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
          weights = w,
          se_type = "CR2",
          clusters = idencuesta,
          data = data_ml)

# High to middle

# tiempo: descender se vuelve un efecto positivo en la ultima ola (con mi plata no)
mp_hm_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
          weights = w,
          se_type = "CR2",
          clusters = idencuesta,
          data = data_hm)

# High to low

mp_hl_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
          weights = w,
          se_type = "CR2",
          clusters = idencuesta,
          data = data_hl)

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment",
  "ola2018" = "Wave 2018",
  "ola2023" = "Wave 2023",
  "t:ola2018" = "Mobility treatment x Wave 2018",
  "t:ola2023" = "Mobility treatment x Wave 2023"
  )

texreg::texreg(list(mp_lm_t,
               mp_lh_t,
               mp_mh_t,               
               mp_ml_t,
               mp_hm_t,
               mp_hl_t),
          custom.model.names = c(
            "Low-Middle",
            "Low-High",
            "Middle-High",            
            "Middle-Low",
            "High-Middle",
            "High-Low"
          ),   caption.above = NULL,
               caption = NULL,
               custom.coef.map = ccoef,
               digits = 2,
               groups = list("Wave (Ref.= 2016)" = 3:4, "Mobility treatment x Wave (Ref.= 2016)" = 5:6),
               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.80,
               center = T,float.pos = "H!",
               custom.gof.rows = list("Controls"=rep("Yes",6)))

```

## Sensitivity analysis

The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness

### 2025-11-02 paper/paper.qmd
Update varios

---
title: "Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
    code-link: true
    title-block-banner: true
    theme:
      - pretty.scss
  pdf: 
    pdf-engine: xelatex
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
author:
  - name: Andreas Laffert Tamayo
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 1
affiliations: 
  - id: 1
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
citeproc: true
abstract: | 
 My abstract
  \newline
 **Keywords**: pension commodification, social mobility, causal inference, Chile, public preferences
link-citations: true
linestretch: 1.5
mainfont: Times New Roman
fontsize: 13pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning = FALSE,
                      message = FALSE, 
                      cache = TRUE,
                      out.width = '85%',
                      fig.pos= "H")
options(scipen=999)
options(kableExtra.auto_format = FALSE)
options(knitr.kable.NA = '')
options(knitr.graphics.error = FALSE)
Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }

library(conflicted)
conflicts_prefer(dplyr::filter)

```

This document was last modified at {{< modified >}}
and it was last rendered at {{< now >}}.

:::: {.justify}

{{< include 01-introduction.qmd >}}

{{< include 02-antecedents.qmd >}}

{{< include 03-methods.qmd >}}

{{< include 04-analysis.qmd >}}

{{< include 05-discussion.qmd >}}

{{< include 06-conclusion.qmd >}}

# References

::: {#refs}
:::
::::

{{< include 07-supplementary-material.qmd >}}

### 2025-11-02 paper/paper.tex
Update varios

--- 2025-06-27 advances/informe3.tex
@@
 ]{article}
-
+\usepackage{xcolor}
+\usepackage[margin=2cm]{geometry}
 \usepackage{amsmath,amssymb}
-\usepackage{setspace}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 }{}
+\usepackage{setspace}
 \makeatletter
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\makeatother
+
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
-  pdftitle={Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile},
+  pdftitle={Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility},
   pdfauthor={Andreas Laffert Tamayo},
@@
 
-\title{Preferences for mercantilization: the role of intergenerational
-social mobility and beliefs in meritocracy in Chile}
+\title{Preferences for the commodification of pensions in Chile: the rol
+of intergenerational social mobility}
 
@@
 \date{}
+
+% Colores extra por nombre
+\usepackage[dvipsnames]{xcolor}
+% Ajustes de hipervínculos
+% \hypersetup{
+%   colorlinks = true,
+%   citecolor  = DarkSlateBlue,     % << citas
+%   urlcolor   = DarkSlateBlue  % URLs/DOIs
+% }
+% 
 \begin{document}
 \maketitle
+\begin{abstract}
+My abstract \newline \textbf{Keywords}: pension commodification, social
+mobility, causal inference, Chile, public preferences
+\end{abstract}
 
 \setstretch{1.5}
+This document was last modified at 2025-11-02 05:55:15 and it was last
+rendered at 2025-11-02 05:55:15.
+
 \section{Introduction}\label{introduction}
@@
 COVID-19 pandemic---and structural inequalities have become more visible
-(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). Despite its
-economic achievements, Chile remains one of the most unequal countries
-in Latin America and the OCDE. The poorest 50\% of the population
-captures just 10\% of total income and holds negative net wealth, while
-the wealthiest 1\% receives nearly 27\% of all income and controls
-almost half (49.6\%) of the country's wealth
+(\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et al., 2021}).
+Despite its economic achievements, Chile remains one of the most unequal
+countries in Latin America and the OCDE. The poorest 50\% of the
+population captures just 10\% of total income and holds negative net
+wealth, while the wealthiest 1\% receives nearly 27\% of all income and
+controls almost half (49.6\%) of the country's wealth
 (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
@@
 increased household debt, labor market precarity, and the COVID-19
-crisis (\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}).
-However, the majority of the population identifies themselves as members
-of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo et al.,
-2013}). These dynamics have fueled a legitimacy crisis: meritocracy is
-increasingly questioned, and many perceive that advancement depends more
-on informal networks than on effort or earned credentials
-(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). The
-combination of constrained mobility, persistent inequality, and
-concentrated opportunity has eroded expectations of upward mobility and
-intensified social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD,
-2017}).
+crisis (\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et al.,
+2021}). However, the majority of the population identifies themselves as
+members of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo
+et al., 2013}). These dynamics have fueled a legitimacy crisis:
+meritocracy is increasingly questioned, and many perceive that
+advancement depends more on informal networks than on effort or earned
+credentials (\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et
+al., 2021}). The combination of constrained mobility, persistent
+inequality, and concentrated opportunity has eroded expectations of
+upward mobility and intensified social discontent
+(\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).
 
@@
 
-\section{Data, variables and methods}\label{data-variables-and-methods}
-
-\subsection{Data}\label{data}
+\section{Method}\label{method}
+
+Social mobility effects---understood as impacts on an outcome arising
+from movements between an origin state and a destination state (e.g.,
+social class)---have long been a focus of sociological research
+(\citeproc{ref-eyles_social_2022}{Eyles et al., 2022};
+\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
+Yet, as Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024, p.
+467}) emphasize, most hypotheses are, at their core, individual-level
+counterfactual comparisons between the observed outcome under a mobility
+trajectory and the outcome the same person would have shown if they had
+remained in their origin class (or moved to an alternative destination).
+Standard specifications (e.g., SAM, DRM, and mobility-contrast models)
+struggle to retrieve these inherently counterfactual quantities due to
+identification constraints and their reliance on between-group contrasts
+constructed from additive terms and interactions, thereby yielding
+primarily associational evidence
+(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
+\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). I therefore adopt
+Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024}) causal
+formulation, treating destination class as the treatment, conditioning
+on the class of origin, and estimating heterogeneous destination effects
+with observational data under explicit identification assumptions. To
+align design and target, I follow the MIDA template
+(\citeproc{ref-blair_research_2023}{Blair et al., 2023}): set out the
+causal model and assumptions (M), define the inquiry and estimand (I),
+describe the data and variables (D), and detail the answer strategy (A)
+used to identify the causal effect of intergenerational social mobility
+on preferences for the commodification of pensions.
+
+\subsection{A Causal Model for Mobility
+Effects}\label{a-causal-model-for-mobility-effects}
+
+Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}), building on
+a critical reassessment of the inferential limits of standard mobility
+models, propose a causal framework that reconceptualizes the ``mobility
+effect'' as the treatment effect of reaching a destination class, with
+impacts heterogeneous by class of origin. The core claim is that typical
+mobility hypotheses are fundamentally within-person counterfactual
+comparisons rather than between-person contrasts. Defining mobility as
+\((M = D - O)\) ---a change from origin \((O)\) to destination \((D)\),
+rather than an additive or interactive combination---places the focus
+squarely on the Neyman--Rubin problem: for any individual, we observe
+the outcome under the realized destination but not under alternative
+destinations or under immobility.
+
+Exploiting the temporal ordering of origin, destination and outcome,
+Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}) frame
+mobility as a treatment process that motivates the causal question:
+``how would the outcome among people from origin \emph{j} who entered
+destination \emph{k} have been different if those people had,
+counterfactually, entered destination \emph{k'} instead?'' (p.472). The
+corresponding estimand is the conditional causal effect of destination
+given origin. It is of particular interest when \emph{k' = j}, i.e.,
+immobility. This estimand compares movers' observed outcomes with the
+hypothetical outcomes those same individuals would have exhibited had
+they remained in their origin class. This formulation provides a
+coherent potential-outcomes basis for studying how class transitions
+shape individual preferences and attitudes.
+
+Following Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024}),
+the identification of causal mobility effects from observational data
+requires the following assumptions:
+
+\textbf{Positivity}: For all \((j,c)\) in the support of \((O,C)\) and
+for each relevant destination \((k)\), the probability of receiving
+\(D=k\) is strictly between 0 and 1; substantively, every \((O,C)\)-type
+has some chance of each comparison destination.
+
+\begin{equation}\phantomsection\label{eq-posi}{
+ 0<P(D=k\mid O=j, C=c)<1
+}\end{equation}
+
+\textbf{Stable Unit Treatment Value Assumption (SUTVA)}: each unit's
+potential outcomes \(Y_i(D)\) does not depend on the mechanism used to
+assign treatments (destinations) and by the treatments assigned to other
+units (also called no interference), assuming a single, well-defined
+version of each treatment (consistency).
+
+\begin{equation}\phantomsection\label{eq-sutva}{
+Y_i(D_i) = Y_i\big(D_i, D_{-i}\big)\quad \forall, D_{-i}
+}\end{equation}
+
+\textbf{Conditional Independence}: Also known as conditional
+unconfoundedness or exchangeability, this assumption emphasizes that,
+conditional on origin \(O\) and pre-treatment covariates \(C\) (e.g.,
+parental education, ethnicity, cohort, early-life factors), assignment
+to \(D\) is as good as random. This underlies IPW and regression
+adjustment, which seek exchangeability between mobile and immobile
+individuals.
+
+\begin{equation}\phantomsection\label{eq-indep}{
+Y(D)\ \perp\kern-5pt \perp D\ \mid\ (O,C)
+}\end{equation}
+
+Figure~\ref{fig-dag} depicts the causal model guiding identification. In
+this DAG, \(C\) represents pre-treatment confounders (e.g., family
+resources, household composition, and ascriptive status) affecting both
+origin and destination, and \(U\) denotes unobserved factors influencing
+\(C\). Class origin \(O\) is determined by \(C\) and, in turn, affects
+destination \(D\). The outcome \(Y\) (preferences for commodification)
+is caused by \(D\) and can also be affected by \(C\). There is no direct
+arrow from \(O\) to \(Y\), reflecting the assumption that the effect of
+origin on the outcome operates entirely through background factors
+\((C)\) and destination \((D)\). Accordingly, conditioning on \((O,C)\)
+blocks the relevant back-door paths from \(D\) to \(Y\), specifically
+\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
+\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
+adjustment set.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-dag-1.pdf}
+
+}
+
+\caption{\label{fig-dag}Causal graph of the effect of intergenerational
+social mobility on preferences for commodification. Y = preferences for
+pension commodification, my outcome of interest, D = an individual's
+class of destination, O = an individual's class of origin, C = different
+attributes determined in childhood or earlier that affects an
+individual's class of origin and destination. Finally, U = unobserved
+factors influencing origin attributes.}
+
+\end{figure}%
+
+Formally, the DAG encodes the following relations:
+
+\[
+U \rightarrow C,\quad C \rightarrow O,\quad C \rightarrow D,\quad C \rightarrow Y,\quad O \rightarrow D,\quad D \rightarrow Y
+\]
+
+This structure corresponds to a nonparametric structural causal model in
+which \((D)\) is the treatment variable, \((Y)\) the outcome, and
+\((O, C)\) the minimal sufficient adjustment set. Conditioning on these
+variables satisfies the back-door criterion for identifying the causal
+effect of mobility on preferences
+(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}).
+
+Why are the assumptions plausible here? Conditional independence is
+targeted by controlling for a plausible sufficient adjustment set
+\((O,C)\); entropy balancing/IPW enforces covariate balance within each
+origin stratum. Positivity is assessed via propensity scores and
+trimming of extreme weights, restricting to common support. SUTVA holds
+by treating the destination class as a single exposure measured before
+\((Y)\) and assuming no interference. Together, the DAG and assumptions
+define conditions for identifying the causal effect of intergenerational
+mobility on preferences for commodification.
+
+\subsection{Inquiry and estimand}\label{inquiry-and-estimand}
+
+The causal inquiry guiding this study asks: How does intergenerational
+social mobility affect individuals' preferences for the commodification
+of pensions in Chile? Following the potential outcomes framework
+proposed by Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024,
+p. 473}), the estimand of interest is the average treatment effect on
+the treated (ATT), defined as the mean causal effect of moving from an
+origin class \((O=j)\) to a destination class \((D=k)\), compared with
+the counterfactual outcome that those same individuals would have
+exhibited had they instead reached an alternative destination \((k')\):
+
+\begin{equation}\phantomsection\label{eq-breen}{ 
+ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
+}\end{equation}
+
+In this study, I focus on the specific case where the counterfactual
+destination corresponds to class immobility \((k' = j)\), that is, how
+the outcome among people from origin \emph{k} who entered destination
+\emph{k} would have bee different if those people had, counterfactually,
+remained in origin \emph{j} instead (inmobility). Accordingly, the
+estimand becomes:
+
+\begin{equation}\phantomsection\label{eq-att}{ 
+ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
+}\end{equation}
+
+which represents the average causal effect of attaining class
+\emph{(k)}---for instance, moving upward or downward---relative to
+remaining in class \emph{(j)}, among individuals who actually
+experienced such mobility. This estimand captures the within-origin
+counterfactual contrast at the heart of mobility effects: how a person's
+preference for market-based pensions would differ if, counterfactually,
+they had stayed in their origin class rather than moved to their
+observed destination.
+
+\subsection{Data and variables}\label{data-and-variables}
+
+\subsubsection{Data}\label{data}
 
@@
 (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES).
-The survey is an annual panel collected from 2016 to 2023, comprising
-two independent samples (original and refreshment), featuring permanent
-and rotating questionnaire modules. It evaluates how individuals think,
-feel, and behave regarding conflict and social cohesion in Chile. The
-sampling design is complex, probabilistic, clustered, multistage, and
-stratified by city size, targeting men and women aged 18--75 who
-habitually reside in private dwellings in urban areas spanning 40 cities
-(92 municipalities, 13 regions). This analysis focuses on 2023, the most
-recent wave, which includes 1,737 respondents in the original sample and
-989 in the refreshment sample. Further details on sampling, attrition,
-and weighting can be found at \url{https://coes.cl/encuesta-panel/}, and
-the dataset is publicly available at
-\url{https://dataverse.harvard.edu/dataverse/elsoc}.
-
-\subsection{Variables}\label{variables}
-
-\emph{Outcome variables}
-
-\textbf{Market justice preferences}: The outcome variables in this study
-are market justice preferences. This construct is operationalized
-through three variables that address the degree of justification
-regarding whether access to social services in healthcare, pensions, and
-education should be income conditional. Specifically, the justification
-of inequality in healthcare is assessed by the question: ``Is it fair in
-Chile that people with higher incomes can access better healthcare than
-people with lower incomes?'' The same question is asked for pensions and
-education. In all cases, respondents indicate their level of agreement
-on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
-(``strongly agree''). Although these items theoretically reflect the
-same underlying concept of market justice
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}), they are analyzed
-separately because healthcare, pensions, and education may prompt
-distinct response patterns
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}).
-
-\emph{Independent variables}
-
-\textbf{Social mobility}: Intergenerational social mobility is treated
-as an exposure indicating whether respondents occupy a different class
-position from their fathers. Following Breen and Ermisch's
-(\citeproc{ref-breen_effects_2024}{2024}) framework for estimating
-causal effects of mobility, the variable is constructed in two stages.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Class assignment. Using the International Socio-Economic Index of
-  Occupational Status (ISEI) derived from three-digit ISCO-08 codes,
-  both paternal (origin) and respondent (destination) occupations are
-  grouped into three strata:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Upper (codes 100-299: managers, directors, professionals)
-  \item
-    Middle (codes 300-499: skilled white-collar workers)
-  \item
-    Lower (codes 500 and above: manual, semi-skilled, or unskilled
-    workers)
-  \end{itemize}
-\item
-  Propensity-score estimation. To model mobility as a treatment---i.e.,
-  a transition from origin \emph{j} to destination \emph{k}---I estimate
-  multinomial logit propensity scores using covariates that influence an
-  individual's likelihood of moving between strata: (a) father's
-  educational level, (b) presence of both parents at age 15, (c)
-  nationality, (d) age, (e) sex, and (f) ethnicity.
-\end{enumerate}
-
-These propensity scores are subsequently employed to adjust for
-selection into mobility when assessing its effect on market justice
-preferences.
-
-\textbf{Meritocracy}: Meritocratic perception is operationalized through
-two components: one addressing effort and another focusing on talent
-(\citeproc{ref-young_rise_1958}{Young, 1958}). The item used to gauge
-effort is: ``In Chile, people are rewarded for their efforts,'' while
-the item for talent is: ``In Chile, people are rewarded for their
-intelligence and skills.'' In both cases, respondents indicate their
-level of agreement on a five-point Likert scale, ranging from 1
-(``strongly disagree'') to 5 (``strongly agree'').
-
-\emph{Controls}
-
-Sociodemographic and attitudinal variables are included to control for
-potential composition effects in the population. In terms of
-sociodemographic characteristics, I incorporate per capita household
-income quantile, educational level (1=Primary or below, 2=Secondary,
-3=Technical, 4=University or above), age (in years), and sex (1=Male,
-2=Female), which have been shown to significantly influence market
-justice preferences (\citeproc{ref-castillo_socialization_2024}{Castillo
-et al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
-attitudinal variables, I include political identification (1=Left,
-2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) because they may confound the relationship
-between market justice preferences and social mobility and meritocracy.
-
-\subsection{Methods}\label{methods}
-
-Following Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024})
-strategy for estimating the causal effect of social mobility on market
-justice preferences, I employ linear regression models with
-inverse-probability weights (IPW) for mobility conditional on class of
-origin. This approach allows estimation of the average causal effect of
-moving from an origin class \emph{j} to a destination class \emph{k} by
-comparing individuals who actually reach \emph{k} with their
-counterfactual outcome had they instead moved to an alternative
-destination \emph{k′} (\citeproc{ref-breen_effects_2024}{Breen \&
-Ermisch, 2024}). The resulting estimand is the average treatment effect
-on the treated (ATT).
-
-All the analyses will be conducted using R software.
+This survey is a nationally representative panel study of the urban
+adult population in Chile, conducted annually between 2016 and 2023.
+Designed to examine individuals' attitudes, emotions, and behaviors
+regarding social conflict and cohesion, ELSOC employs a probabilistic,
+stratified, clustered, and multistage sampling design covering both
+major urban centers and smaller cities. The sampling frame was
+proportionally stratified into six categories of urban population size
+(e.g., large and small cities), followed by a random selection of
+households within 1,067 city blocks. The target population includes men
+and women aged 18 to 75 who are habitual residents of private dwellings.
+
+Because respondent occupation is not measured in every wave, I restrict
+the analysis to the 2016, 2018, and 2023 waves. Parental occupation was
+collected only in 2023 as open-text labels. I coded these texts to the
+ISCO-08 two-digit scheme using the National Institute of Statistics of
+Chile's automated coding API and retained cases with high-confidence
+matches (\textgreater95\%). Treating parental occupation as a
+time-invariant origin attribute, I carried these codes back to 2016 and
+2018. After listwise deletion and restricting to key variables
+(respondent occupation and the outcome measure), the final analytic
+sample comprises \emph{N} = 3,435 observations nested within \emph{N} =
+1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144). Consistent with
+the study design, estimation proceeds within trajectory-specific
+subsamples (e.g., low→low; low→middle), retaining movers and matched
+non-movers from the same origin class. Then, I apply weights to
+construct the corresponding counterfactual control distributions for
+each trajectory. As Breen and Ermisch
+(\citeproc{ref-breen_effects_2024}{2024}) did, I use the three waves in
+the estimations that follow, allowing for correlation between the
+observations for each individual in calculating the standard errors
+(i.e.~clustering on the personal identifier).
+
+\subsubsection{Outcome variable}\label{outcome-variable}
+
+The outcome variable measures preferences regarding the commodification
+of pensions, operationalized with a single item addressing the degree of
+public support for market-based access to old-age pension benefits.
+Respondents were asked: ``Is it fair in Chile that people with higher
+incomes have better pensions than people with lower incomes?'' and were
+asked to indicate their level of agreement on a five-point Likert scale
+ranging from 1 (``strongly disagree'') to 5 (``strongly agree''). For
+estimation and interpretability, I dichotomized the item so that values
+4-5 indicate agreement with market-based access to old-age pensions
+(coded as 1), while values 1-3 indicate disagreement (coded as 0). This
+binary coding treats neutrality as substantially closer to non-support
+than to support and produces a clear contrast between respondents who
+support market-based differentiation in pensions and those who do not. I
+use this item for two main reasons: on the one hand, to be able to make
+comparisons with existing literature
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}) and, on the
+other hand, because it captures two fundamental dimensions of
+market-based welfare distribution: the role of economic resources as a
+key determinant of outcomes and the framing of services as as tradable
+commodities that can be bought and sold based on ability to pay
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
+
+Table~\ref{tbl-summary1} shows the descriptive statistics of the
+variables used in the study.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.4216}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2647}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2059}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.1078}}@{}}
+\caption{Descriptive statistics for all
+variables.}\label{tbl-summary1}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Label
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Stats / Values
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Freqs (\% of Valid)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Valid
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Label
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Stats / Values
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Freqs (\% of Valid)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Valid
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Preference for pension commodification &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. Disagree\\
+2. Agree\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+2702 (78.7\%)\\
+733 (21.3\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Father stratum & \begin{minipage}[t]{\linewidth}\raggedright
+1. Low\\
+2. Middle\\
+3. High\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1122 (32.7\%)\\
+1104 (32.1\%)\\
+1209 (35.2\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Offspring stratum & \begin{minipage}[t]{\linewidth}\raggedright
+1. Low\\
+2. Middle\\
+3. High\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1186 (34.5\%)\\
+1110 (32.3\%)\\
+1139 (33.2\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Parental education & \begin{minipage}[t]{\linewidth}\raggedright
+Mean (sd) : 4 (2.2)\\
+min \textless{} med \textless{} max:\\
+1 \textless{} 4 \textless{} 10\\
+IQR (CV) : 3 (0.6)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1 : 222 ( 6.5\%)\\
+2 : 905 (26.3\%)\\
+3 : 576 (16.8\%)\\
+4 : 321 ( 9.3\%)\\
+5 : 826 (24.0\%)\\
+6 : 30 ( 0.9\%)\\
+7 : 214 ( 6.2\%)\\
+8 : 75 ( 2.2\%)\\
+9 : 245 ( 7.1\%)\\
+10 : 21 ( 0.6\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Co-residence with both parents at age 15 &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. No co-residence\\
+2. Co-residence\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+996 (29.0\%)\\
+2439 (71.0\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Nacionality & \begin{minipage}[t]{\linewidth}\raggedright
+1. Non-Chilean\\
+2. Chilean\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+62 ( 1.8\%)\\
+3373 (98.2\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Sex & \begin{minipage}[t]{\linewidth}\raggedright
+1. Male\\
+2. Female\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1507 (43.9\%)\\
+1928 (56.1\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Age (in years) & \begin{minipage}[t]{\linewidth}\raggedright
+Mean (sd) : 42.6 (12.5)\\
+min \textless{} med \textless{} max:\\
+18 \textless{} 43 \textless{} 75\\
+IQR (CV) : 21 (0.3)\strut
+\end{minipage} & 58 distinct values &
+\begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Indigenous ethnicity & \begin{minipage}[t]{\linewidth}\raggedright
+1. Non-indigenous\\
+2. Indigenous\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3035 (88.4\%)\\
+400 (11.6\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Meritocracy perception & \begin{minipage}[t]{\linewidth}\raggedright
+1. Low\\
+2. High\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+2741 (79.8\%)\\
+694 (20.2\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+Wave & \begin{minipage}[t]{\linewidth}\raggedright
+1. 2016\\
+2. 2018\\
+3. 2023\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+914 (26.6\%)\\
+1377 (40.1\%)\\
+1144 (33.3\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+3435\\
+(100.0\%)\strut
+\end{minipage} \\
+\end{longtable}
+
+\subsubsection{Treatment}\label{treatment}
+
+\paragraph{Intergenerational occupational
+mobility}\label{intergenerational-occupational-mobility}
+
+I treat intergenerational occupational mobility as an exposure
+indicating whether respondents occupy a different socioeconomic status
+than their fathers, closely following Breen and Ermisch's
+(\citeproc{ref-breen_effects_2024}{2024}) causal framework. Class
+assignment proceeds in two steps. First, I derive status for both origin
+(father) and destination (respondent) from two-digit ISCO-08 occupations
+using the International Socio-Economic Index of Occupational Status
+(ISEI). ISEI is a hierarchical, unidimensional scale that locates
+occupations by incumbents' average education and earnings and was
+explicitly designed to harmonize occupational stratification across
+other class schemes and socioeconomic status measures
+(\citeproc{ref-ganzeboom_internationally_1996}{Ganzeboom \& Treiman,
+1996}), thereby supporting cross-study comparability
+(\citeproc{ref-barozet_measurement_2021a}{Barozet, Boado, et al.,
+2021}). Second, I group both origin and destination ISEI scores into
+terciles (low, middle, high), yielding a three-category schema for
+origin and destination stratums (see Table~\ref{tbl-matrix}).
+Substantively, I use ISEI because occupation-based measures capture
+structural locations that organize life chances and attitudes more
+directly than income alone: they bundle access to the means of
+production, qualification credentials, and the authority/power inherent
+in labor relations; central dimensions of stratification and inequality
+(\citeproc{ref-barone_rise_2022}{Barone et al., 2022};
+\citeproc{ref-wrightUnderstandingClass2015}{Wright, 2015}). In short,
+using ISEI aligns the measurement of origin and destination on a common
+vertical status continuum that is theoretically meaningful for social
+mobility research and empirically feasible given the available
+information of ELSOC's data.
+
+\begin{table}
+
+\caption{\label{tbl-matrix}Occupational mobility by occupational
+groups.}
+
+\centering{
+
+\centering
+\begin{tabular}{cccccc}
+\toprule
+Father↓ & Offspring→ & Low & Middle & High & Total\\
+\midrule
+Low &  & 44.7\%   (501) & 33.6\%   (377) & 21.7\%   (244) & 100.0\% (1,122)\\
+Middle &  & 35.1\%   (388) & 34.9\%   (385) & 30.0\%   (331) & 100.0\% (1,104)\\
+High &  & 24.6\%   (297) & 28.8\%   (348) & 46.7\%   (564) & 100.0\% (1,209)\\
+Total &  & 34.5\% (1,186) & 32.3\% (1,110) & 33.2\% (1,139) & 100.0\% (3,435)\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{Pre-treatment control variables for selection into
+treatment}\label{pre-treatment-control-variables-for-selection-into-treatment}
+
+Within each origin stratum \(O=j\), I estimate average treatment effects
+on the treated (ATT) by comparing movers to otherwise similar non-movers
+from the same origin. To adjust for selection into mobility, I reweight
+only the control group (immobile) using entropy balancing so that the
+distribution of pre-treatment origin covariates \(C\) matches that of
+the treated on selected moments.
+
+The selection of the adjustment set \(C\) follows two principles. First,
+it mirrors the logic in Breen and Ermisch
+(\citeproc{ref-breen_effects_2024}{2024})---who condition on attributes
+determined in childhood or earlier (e.g., parental resources, household
+structure, ascriptive traits)---and, second, it is constrained to
+origin-side characteristics available in ELSOC. Guided by theory and
+Chilean evidence on relative mobility determinants (mainly
+\citeproc{ref-brunori_inequality_2025}{Brunori et al., 2025};
+\citeproc{ref-celhay_intergenerational_2010}{Celhay et al., 2010};
+\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
+\citeproc{ref-torche_unequal_2005}{Torche, 2005}), I include six
+covariates covering family resources, household composition, and
+ascriptive status: (a) parental education (highest of father/mother), 10
+categories coded ordinally from no schooling to postgraduate studies;
+(b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c)
+nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0
+= male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
+These variables are fixed before the destination class, plausibly affect
+selection into mobility, and constitute the maximal origin set in ELSOC,
+thereby making the conditional independence assumption more credible
+within each \(O=j\).
+
+After balancing, I construct stabilized IPW-ATT, setting treated weights
+to 1 and assigning ebal-derived weights to controls, thus aligning the
+counterfactual distribution of non-movers with the covariate profile of
+movers in each origin stratum. These weights are then used in the
+outcome stage to estimate the causal effect of mobility on preferences
+for the commodification of pensions.
+
+\subsubsection{Effect heterogeneity}\label{effect-heterogeneity}
+
+To test heterogeneity in mobility effects, I examine two moderators:
+meritocratic perceptions and time. Meritocracy is captured with two
+items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
+effort (``In Chile, people are rewarded for their efforts'') and one to
+talent (``In Chile, people are rewarded for their intelligence and
+skills''), each answered on a five-point Likert scale (1 = ``strongly
+disagree'' to 5 = ``strongly agree''). I average the two items into a
+single index and dichotomize it into low meritocracy (≤ 3) and high
+meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
+and entered as a categorical (dummy) variable. Both variables are used
+as moderators by interacting them with the mobility treatment to assess
+conditional average treatment effects---i.e., whether the effect of
+mobility varies across levels of meritocratic beliefs or across survey
+waves.
+
+\subsubsection{Controls}\label{controls}
+
+All models include the same pre-treatment covariates \(C\) used in the
+IPW construction, not to re-balance groups (entropy balancing already
+does so), but to (i) block backdoor paths from \(C\) to the outcome
+\(Y\) as implied by the DAG, and (ii) achieve double robustness:
+estimates remain consistent if either the weighting model or the outcome
+model is correctly specified. Concretely, I adjust for (a) father's
+educational level, (b) co-residence with both parents at age 15, (c)
+nationality, (d) age, (e) sex, and (f) ethnicity, along with wave fixed
+effects to absorb secular trends.
+
+\subsection{Analytical strategy}\label{analytical-strategy}
+
+Because the dependent variable---the preference for pension
+commodification---is binary \((Y_{it}\in{0,1})\), I estimate weighted
+linear probability models (WLS) using stabilized inverse probability
+weights (IPW--ATT) obtained via entropy balancing. Although OLS
+estimator is often questioned with binary outcomes, it's unbiased. It
+consistently estimates the conditional expectation \(E(Y|X)\) under
+standard exogeneity \(E[\varepsilon_i|X_i]=0\), with heteroskedasticity
+addressed by robust or clustered errors
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
+ch.~7.5). WLS further improves efficiency under heteroskedasticity or
+unequal sampling probabilities
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
+ch.~8.4). Following Gelman, Hill, and Vehtari
+(\citeproc{ref-gelman_regression_2020}{2020, pp. 270--272}),
+IPW-weighted regression transparently targets the ATT by recreating the
+counterfactual distribution. Linear models are preferred here because
+coefficients are average probability differences; nonlinear models
+(logit/probit) complicate weighting and interpretation
+(\citeproc{ref-gelman_regression_2020}{Gelman et al., 2020}, ch.~13).
+
+Formally, the model is:
+
+\begin{equation}\phantomsection\label{eq-final}{
+Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it},
+}\end{equation}
+
+where \(Y_{it}\) indicates whether individual \(i\) in wave \(t\)
+supports more market-based pension access; \(\alpha\) is the baseline
+probability for the reference categories; \(\beta T_i\) captures the
+intergenerational mobility contrast: \(T_i=1\) when the observed
+destination is \(D_i=k\) (mobile to \(k\)) and \(T_i=0\) when \(D_i=j\)
+(immobile), so that \(\beta\) is the ATT within origin
+\(O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})\), i.e., the
+percentage-point change in the probability of preferring pension
+commodification from reaching \(k\) rather than remaining in \(j\). The
+term \(X_i'\gamma\) includes the pre-treatment covariates \((C)\) used
+to construct the entropy-balancing weights; \(\lambda_t\) are wave fixed
+effects; and \(\varepsilon_{it}\) is idiosyncratic error. The analytic
+sample is restricted to individuals sharing the same origin \((O=j)\).
+Estimation uses pooled WLS with stabilized IPW--ATT from entropy
+balancing and CR2 standard errors clustered by individual to address
+heteroskedasticity and within-person dependence.
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds if either the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). I
+report both IPW-only and doubly robust specifications to document
+robustness. In addition, I assess causal heterogeneity by interacting
+the mobility treatment with (i) meritocratic perceptions and (ii) survey
+wave, reporting simple slopes (origin-specific ATTs) by moderator
+levels. To evaluate sensitivity to unmeasured confounding, I compute
+E-values for point estimates and confidence limits
+(\citeproc{ref-ding_sensitivity_2016}{Ding \& VanderWeele, 2016}). To
+address multiplicity across the six mobility trajectories, I also report
+Holm-adjusted p-values alongside 95\% confidence intervals.
+
+All the analyses were conducted using R software and the
+\emph{lm\_robust} package.
+
+\section{Results}\label{results}
+
+\subsection{Descriptive statistics}\label{descriptive-statistics}
+
+Lorem ipsoThe specification is doubly robust: entropy-balancing weights
+are combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustnessThe specification is
+doubly robust: entropy-balancing weights are combined with covariate
+adjustment so consistency holds ifeither the weighting model or the
+outcome model is correctly specified, while also improving efficiency
+(Wooldridge, 2009).I report both IPW-only and doubly robust
+specifications to document robustness
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-alluvial-1.pdf}
+
+}
+
+\caption{\label{fig-alluvial}Change in preferences for pension
+commodification over time (2016, 2018, and 2023).}
+
+\end{figure}%
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustness
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-mean-1.pdf}
+
+}
+
+\caption{\label{fig-mean}Percentage agreement on preferences for pension
+commodification according to mobility trajectory and time (2016, 2018,
+and 2023).}
+
+\end{figure}%
+
+\subsection{Models}\label{models}
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustness
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-reg-1.pdf}
+
+}
+
+\caption{\label{fig-reg}Effects of intergenerational occupational
+mobility on preferences for pension commodification. Coefficient plot of
+origin-specific ATT estimates comparing mobility versus immobility.
+Estimates from pooled WLS with doubly robust adjustment, wave fixed
+effects, and standard errors clustered by individual; bars show 95\%
+confidence intervals.}
+
+\end{figure}%
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustnessThe specification is
+doubly robust: entropy-balancing weights are combined with covariate
+adjustment so consistency holds ifeither the weighting model or the
+outcome model is correctly specified, while also improving efficiency
+(Wooldridge, 2009).I report both IPW-only and doubly robust
+specifications to document robustness
+
+\subsection{Effect heterogeneity}\label{effect-heterogeneity-1}
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustness
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustness
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-1.pdf}
+
+}
+
+\caption{\label{fig-interact}Effects of intergenerational occupational
+mobility on preferences for pension commodification, by meritocratic
+perception. Coefficient plot of origin-specific ATT estimates comparing
+mobility versus immobility, estimated from interaction models (T ×
+Merit). Points show simple slopes of the mobility effect at low merit
+and high merit; bars are 95\% confidence intervals. Estimates from
+pooled WLS with doubly robust adjustment, wave fixed effects, and
+standard errors clustered by individual.}
+
+\end{figure}%
+
+\begin{table}
+
+\caption{\label{tbl-inttime}Effects of intergenerational occupational
+mobility on preferences for pension commodification, by survey wave}
+
+\centering{
+
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                              & $0.18$           & $-0.04$          & $0.28$           & $0.20$           & $0.18$            & $0.15$            \\
+                                       & $ [-0.36; 0.73]$ & $ [-0.36; 0.29]$ & $ [-0.01; 0.57]$ & $ [-0.11; 0.52]$ & $ [-0.15;  0.50]$ & $ [-0.27;  0.58]$ \\
+Mobility treatment                     & $-0.05$          & $-0.03$          & $0.13^{*}$       & $0.06$           & $-0.19^{*}$       & $-0.19^{*}$       \\
+                                       & $ [-0.16; 0.06]$ & $ [-0.15; 0.10]$ & $ [ 0.05; 0.22]$ & $ [-0.03; 0.15]$ & $ [-0.33; -0.05]$ & $ [-0.32; -0.05]$ \\
+Wave (Ref.= 2016)                      &                  &                  &                  &                  &                   &                   \\
+                                       &                  &                  &                  &                  &                   &                   \\
+\quad Wave 2018                        & $0.01$           & $0.01$           & $0.10^{*}$       & $0.09^{*}$       & $-0.04$           & $-0.02$           \\
+                                       & $ [-0.10; 0.12]$ & $ [-0.10; 0.12]$ & $ [ 0.03; 0.18]$ & $ [ 0.01; 0.17]$ & $ [-0.15;  0.07]$ & $ [-0.15;  0.10]$ \\
+\quad Wave 2023                        & $0.03$           & $0.01$           & $0.19^{*}$       & $0.18^{*}$       & $0.05$            & $0.04$            \\
+                                       & $ [-0.09; 0.14]$ & $ [-0.12; 0.13]$ & $ [ 0.10; 0.28]$ & $ [ 0.08; 0.27]$ & $ [-0.09;  0.18]$ & $ [-0.11;  0.19]$ \\
+Mobility treatment x Wave (Ref.= 2016) &                  &                  &                  &                  &                   &                   \\
+                                       &                  &                  &                  &                  &                   &                   \\
+\quad Mobility treatment x Wave 2018   & $0.03$           & $-0.06$          & $-0.04$          & $-0.09$          & $0.11$            & $0.06$            \\
+                                       & $ [-0.11; 0.17]$ & $ [-0.21; 0.09]$ & $ [-0.17; 0.09]$ & $ [-0.20; 0.03]$ & $ [-0.04;  0.26]$ & $ [-0.09;  0.22]$ \\
+\quad Mobility treatment x Wave 2023   & $0.07$           & $0.10$           & $-0.09$          & $-0.07$          & $0.19^{*}$        & $0.18$            \\
+                                       & $ [-0.09; 0.23]$ & $ [-0.08; 0.28]$ & $ [-0.23; 0.06]$ & $ [-0.22; 0.08]$ & $ [ 0.01;  0.38]$ & $ [-0.01;  0.37]$ \\
+\hline
+Controls                               & Yes              & Yes              & Yes              & Yes              & Yes               & Yes               \\
+R$^2$                                  & $0.04$           & $0.05$           & $0.07$           & $0.03$           & $0.06$            & $0.06$            \\
+Adj. R$^2$                             & $0.03$           & $0.04$           & $0.06$           & $0.02$           & $0.05$            & $0.05$            \\
+Num. obs.                              & $878$            & $745$            & $716$            & $773$            & $912$             & $861$             \\
+RMSE                                   & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$            & $0.42$            \\
+N Clusters                             & $497$            & $449$            & $443$            & $470$            & $536$             & $525$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+\subsection{Sensitivity analysis}\label{sensitivity-analysis}
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds ifeither the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency (Wooldridge, 2009).I report both IPW-only and
+doubly robust specifications to document robustness
+
+\section{Discussion}\label{discussion}
+
+\section{Conclusion}\label{conclusion}
 
@@
 
+\bibitem[\citeproctext]{ref-barone_rise_2022}
+Barone, C., Hertel, F. R., \& Smallenbroek, O. (2022). The rise of
+income and the demise of class and social status? {A} systematic review
+of measures of socio-economic position in stratification research.
+\emph{Research in Social Stratification and Mobility}, \emph{78},
+100678. \url{https://doi.org/10.1016/j.rssm.2022.100678}
+
+\bibitem[\citeproctext]{ref-barozet_measurement_2021a}
+Barozet, E., Boado, M., \& Marqués-Perales, I. (2021). The {Measurement}
+of {Social Stratification}: {Comparative Perspectives Between Europe}
+and {Latin America}. In P. López-Roldán \& S. Fachelli (Eds.),
+\emph{Towards a {Comparative Analysis} of {Social Inequalities} between
+{Europe} and {Latin America}} (pp. 171--202). Cham: Springer
+International Publishing.
+\url{https://doi.org/10.1007/978-3-030-48442-2_6}
+
 \bibitem[\citeproctext]{ref-barozet_clases_2021}
@@
 
+\bibitem[\citeproctext]{ref-blair_research_2023}
+Blair, G., Coppock, A., \& Humphreys, M. (2023). \emph{Research design
+in the social sciences: Declaration, diagnosis, and redesign}. Princeton
+(N.J.) Oxford: Princeton University press.
+
 \bibitem[\citeproctext]{ref-boccardo_30_2020}
@@
 \url{https://doi.org/10.15195/v11.a17}
+
+\bibitem[\citeproctext]{ref-brunori_inequality_2025}
+Brunori, P., Ferreira, F. H. G., \& Neidhöfer, G. (2025). Inequality of
+opportunity and intergenerational persistence in {Latin America}.
+\emph{Oxford Open Economics}, \emph{4}(Supplement\_1), i167--i199.
+\url{https://doi.org/10.1093/ooec/odae021}
 
@@
 
+\bibitem[\citeproctext]{ref-celhay_intergenerational_2010}
+Celhay, P., Sanhueza, C., \& Zubizarreta, J. R. (2010).
+{INTERGENERATIONAL MOBILITY OF INCOME AND SCHOOLING}: {CHILE} 1996-2006.
+\emph{Revista de an{á}lisis Econ{ó}mico}, \emph{25}(2), 43--63.
+\url{https://doi.org/10.4067/S0718-88702010000200003}
+
 \bibitem[\citeproctext]{ref-chancel_world_2022}
@@
 
+\bibitem[\citeproctext]{ref-ding_sensitivity_2016}
+Ding, P., \& VanderWeele, T. J. (2016). Sensitivity {Analysis Without
+Assumptions}: \emph{Epidemiology}, \emph{27}(3), 368--377.
+\url{https://doi.org/10.1097/EDE.0000000000000457}
+
 \bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
@@
 \url{https://doi.org/10.1111/roiw.12441}
+
+\bibitem[\citeproctext]{ref-ganzeboom_internationally_1996}
+Ganzeboom, H. B. G., \& Treiman, D. J. (1996). Internationally
+{Comparable Measures} of {Occupational Status} for the 1988
+{International Standard Classification} of {Occupations}. \emph{Social
+Science Research}, \emph{25}(3), 201--239.
+\url{https://doi.org/10.1006/ssre.1996.0010}
+
+\bibitem[\citeproctext]{ref-gelman_regression_2020}
+Gelman, A., Hill, J., \& Vehtari, A. (2020). \emph{Regression and {Other
+Stories}} (1st ed.). Cambridge University Press.
+\url{https://doi.org/10.1017/9781139161879}
 
@@
 
+\bibitem[\citeproctext]{ref-song_there_2025}
+Song, X., \& Zhou, X. (2025). Is {There} a {Mobility Effect}? {On
+Methodological Issues} in the {Mobility Contrast Model}.
+\emph{Sociological Methods \& Research}, \emph{54}(4), 1576--1593.
+\url{https://doi.org/10.1177/00491241251347983}
+
 \bibitem[\citeproctext]{ref-sorokin_social_1927}
@@
 
+\bibitem[\citeproctext]{ref-wooldridge_introductory_2009}
+Wooldridge, J. M. (2009). \emph{{Introductory econometrics: a modern
+approach}} (4th ed). Mason, OH: South Western, Cengage Learning.
+
+\bibitem[\citeproctext]{ref-wrightUnderstandingClass2015}
+Wright, E. O. (2015). \emph{Understanding {Class}} (Verso).
+
 \bibitem[\citeproctext]{ref-young_rise_1958}
@@
 
+\section{Supplementary material}\label{supplementary-material}
+
 \end{document}

### 2025-11-02 presentations/seminario/mjp-mobility-seminario.qmd
Update varios

@@
         T= 0,   # Tratamiento
-        L= 1,   # Post-tratamiento
-        M= 1,   # Mediador
         Y= 2),  # Outcome
@@
         T= 0.9,
-        L=-0.4,
-        M= 0.2,
         Y= 0.9)
@@
   T ~ C,          # Tratamiento depende de confusores pre
-  L ~ T + C,      # Confusor post-T inducido por T y C
-  M ~ T + C + L,  # Mediador depende de T, C y L
-  Y ~ T + M + C + L,  # Outcome depende de todo lo anterior
+  Y ~ T  + C ,  # Outcome depende de todo lo anterior
   exposure = "T",
@@
   C = "t1",
-  T = "t2",
-  L = "t3",
-  M = "t3",
-  Y = "t4"
+  T = "t2",,
+  Y = "t3"
 )

### 2025-11-02 processing/otros/borrador_matriz.R

@@ -1,0 +1,92 @@
+
+# tabla cruzada con totales, % fila y % columna
+
+
+tab_mob <- db %>%
+  count(estrato_orig, estrato_ocupa) %>%
+  group_by(estrato_orig) %>%
+  mutate(row_total = sum(n),
+         row_pct = 100 * n / row_total) %>%
+  ungroup() %>%
+  group_by(estrato_ocupa) %>%
+  mutate(col_total = sum(n),
+         col_pct = 100 * n / col_total) %>%
+  ungroup()
+
+# totales por fila y por columna
+row_totals <- tab_mob %>%
+  group_by(estrato_orig) %>%
+  summarise(n = sum(n), .groups = "drop") %>%
+  mutate(estrato_ocupa = "Total")
+
+col_totals <- tab_mob %>%
+  group_by(estrato_ocupa) %>%
+  summarise(n = sum(n), .groups = "drop") %>%
+  mutate(estrato_orig = "Total")
+
+grand_total <- tibble(estrato_orig = "Total", estrato_ocupa = "Total",
+                      n = sum(tab_mob$n))
+
+tab_mob <- tab_mob %>% 
+  select(-c(row_total, col_total, col_pct)) %>% 
+  pivot_longer(cols = c(n, row_pct),
+               names_to = "name",
+               values_to = "value") %>% 
+  pivot_wider(
+    id_cols   = c(estrato_orig, name),   
+    names_from  = estrato_ocupa,         
+    values_from = value,
+    values_fill = 0
+  ) 
+
+tab_mob <- dplyr::add_row(tab_mob, 
+                          estrato_orig = "Total",
+                          name = "n",
+                          Low = as.numeric(col_totals[1,2]),
+                          Middle = as.numeric(col_totals[2,2]),
+                          High = as.numeric(col_totals[3,2]),
+)
+
+
+tab_mob <- tab_mob %>% 
+  rowwise() %>% 
+  mutate(total = sum(Low, Middle, High)) %>% 
+  ungroup()
+
+tab_mob <- add_row(tab_mob, 
+                   estrato_orig = "Total",
+                   name = "row_pct",
+                   Low = round((tab_mob$Low[tab_mob$name=="n"&tab_mob$estrato_orig=="Total"]/tab_mob$total[tab_mob$name=="n"&tab_mob$estrato_orig=="Total"])*100,1),
[…]

### 2025-11-02 processing/prod_analysis.R

@@ -1,630 +1,0 @@
-# 0. Identification ---------------------------------------------------
-
-# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
-# Responsible: Andreas Laffert
-
-# Executive Summary: This script contains the code to data analysis for thesis project
-# Date: May 1, 2025
-
-# 1. Packages  -----------------------------------------------------
-
-if (!require("pacman")) install.packages("pacman")
-
-pacman::p_load(tidyverse,
-               here,
-               sjPlot,
-               sjmisc,
-               texreg,
-               estimatr,
-               conflicted,
-               ggdist,
-               summarytools)
-
-conflicts_prefer(dplyr::filter)
-options(scipen=999)
-rm(list = ls())
-
-# 2. Data -----------------------------------------------------------------
-
-#load(url("https://github.com/Andreas-Lafferte/mobility-market-justice/raw/refs/heads/main/input/data/proc/data_v1.RData"))
-
-load(here("input/data/proc/data_v2.RData"))
-
-glimpse(db)
-
-# 3. Analysis -----------------------------------------------------------
-
-bases <- c("datos1", "datos2", "datos3", "db")
-
-for (nm in bases) {
-  df <- get(nm)
-  
-  df <- df %>%
-    mutate(
-      merit_effort = if_else(merit_effort >= 4, 1, 0),
-      merit_effort = factor(merit_effort, levels = c(0,1), labels = c("Baja", "Alta"))
-      #just_pension = if_else(
-      #  just_pension %in% 4:5, 1L,
-      #  if_else(just_pension %in% 1:3, 0L, NA_integer_)),
-      #just_health = if_else(
-      #  just_health %in% 4:5, 1L,
-      #  if_else(just_health %in% 1:3, 0L, NA_integer_)),
-      #just_educ = if_else(
-      #  just_educ %in% 4:5, 1L,
-      #  if_else(just_educ %in% 1:3, 0L, NA_integer_))
-    )
-  
-  assign(nm, df, inherits = TRUE)
-}
-
-t1 <- db %>% 
-  select(idencuesta, wave, just_pension, estrato_orig, estrato_ocupa, merit_effort, nacionalidad, sexo,
[…]

### 2025-11-02 processing/prod_analysis_binom.R

@@ -1,0 +1,433 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data analysis for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               sjPlot,
+               sjmisc,
+               texreg,
+               estimatr,
+               conflicted,
+               ggdist,
+               summarytools)
+
+conflicts_prefer(dplyr::filter)
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+#load(url("https://github.com/Andreas-Lafferte/mobility-market-justice/raw/refs/heads/main/input/data/proc/data_v1.RData"))
+
+load(here("input/data/proc/data_v3.RData"))
+
+glimpse(db)
+
+# 3. Analysis -----------------------------------------------------------
+
+bases <- c("datos1", "datos2", "datos3", "db")
+
+for (nm in bases) {
+  df <- get(nm)
+  
+  df <- df %>%
+    mutate(
+      just_pension_f = if_else(just_pension >= 4, 1, 0),
+      merit_i = if_else(merit_i >= 4, 1, 0),
+      merit_i = factor(merit_i, levels = c(0,1), labels = c("Baja", "Alta"))
+      #just_pension = if_else(
+      #  just_pension %in% 4:5, 1L,
+      #  if_else(just_pension %in% 1:3, 0L, NA_integer_)),
+      #just_health = if_else(
+      #  just_health %in% 4:5, 1L,
+      #  if_else(just_health %in% 1:3, 0L, NA_integer_)),
+      #just_educ = if_else(
+      #  just_educ %in% 4:5, 1L,
+      #  if_else(just_educ %in% 1:3, 0L, NA_integer_))
+    )
+  
+  assign(nm, df, inherits = TRUE)
+}
+
+bases <- c("datos1", "datos2", "datos3")
+
+for (nm in bases) {
+  df <- get(nm)
+  
[…]

### 2025-11-02 processing/prod_analysis_multinom.R

@@ -1,0 +1,952 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data analysis for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               sjPlot,
+               sjmisc,
+               texreg,
+               estimatr,
+               conflicted,
+               ggdist,
+               summarytools)
+
+conflicts_prefer(dplyr::filter)
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+#load(url("https://github.com/Andreas-Lafferte/mobility-market-justice/raw/refs/heads/main/input/data/proc/data_v1.RData"))
+
+load(here("input/data/proc/data_v2.3.RData"))
+
+glimpse(db)
+
+# 3. Analysis -----------------------------------------------------------
+
+# Descriptive statistics
+
+# Table summary
+t1 <- db %>% 
+  select(idencuesta, 
+         wave, 
+         just_pension_f, 
+         estrato_orig, 
+         estrato_ocupa, 
+         merit_i, 
+         nacionalidad_f, 
+         sexo_f,
+         m0_edad_f, 
+         etnia_f, 
+         hogar_bip_f, 
+         educ_orig_f)
+
+dfSummary(t1,
+          plain.ascii = FALSE,
+          style = "multiline",
+          tmp.img.dir = "/tmp",
+          graph.magnif = 0.75,
+          headings = F,  # encabezado
+          varnumbers = F, # num variable
+          labels.col = T, # etiquetas
+          na.col = F,    # missing
+          graph.col = F, # plot
+          valid.col = T, # n valido
+          col.widths = c(30,10,10,10))
+
+# Mobility matrix
+sjPlot::sjt.xtab(db$estrato_orig, 
[…]

### 2025-11-02 processing/prod_prep.R

@@ -1,575 +1,0 @@
-# 0. Identification ---------------------------------------------------
-
-# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
-# Responsible: Andreas Laffert
-
-# Executive Summary: This script contains the code to data preparation for thesis project
-# Date: May 1, 2025
-
-# 1. Packages  -----------------------------------------------------
-
-if (!require("pacman")) install.packages("pacman")
-
-pacman::p_load(tidyverse,
-               here,
-               car,
-               sjmisc,
-               sjlabelled,
-               data.table,
-               summarytools,
-               nnet,
-               occupar,
-               cobalt,
-               WeightIt)
-
-options(scipen=999)
-rm(list = ls())
-
-# 2. Data -----------------------------------------------------------------
-
-load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
-load(here("input/data/elsoc_ciuo08.RData"))
-
-glimpse(elsoc_long_2016_2023)
-
-# 3. Processing -----------------------------------------------------------
-
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
-elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
-
-elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
-  mutate(just_educ = d02_02, 
-         just_pension = d02_01, 
-         just_health = d02_03, 
-         merit_effort = c18_09,
-         merit_talent = c18_10) %>% 
-  as_tibble() %>% 
-  sjlabelled::drop_labels(., drop.na = FALSE)
-
-# Market Justice Preferences
-
-frq(elsoc_long_2016_2023$just_health)
-frq(elsoc_long_2016_2023$just_pension)
-frq(elsoc_long_2016_2023$just_educ)
-
-elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
-  mutate(
-    across(
[…]

### 2025-11-02 processing/prod_prep_binom.R

@@ -1,0 +1,527 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               summarytools,
+               occupar,
+               cobalt,
+               WeightIt)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
+      .cols = c(just_health, just_pension, just_educ),
[…]

### 2025-11-02 processing/prod_prep_multinom.R

@@ -1,0 +1,628 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               summarytools,
+               nnet,
+               occupar,
+               cobalt,
+               WeightIt)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# Market Justice Preferences
+
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(
+    across(
[…]

### 2025-11-02 paper/03-methods.qmd
Up results faltan anexos

@@
 
-After balancing, I construct stabilized IPW-ATT, setting treated weights to 1 and assigning ebal-derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum.
+After balancing, I construct stabilized IPW-ATT, setting treated weights to 1 and assigning ebal-derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum (see [Supplementary Material](#anexo) for the evaluation of balance).
 These weights are then used in the outcome stage to estimate the causal effect of mobility on preferences for the commodification of pensions.
@@
 $$
-Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it},
+Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
 $$ {#eq-final}

### 2025-11-02 paper/04-analysis.qmd
Up results faltan anexos

@@
 
-Lorem ipsoThe specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustnessThe specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
+@fig-alluvial shows the annual frequencies of preferences for pension commodification in 2016, 2018, and 2023.
+Each year presents stacked percentages frequencies for the level of agreement and disagreement.
+Overall, a large majority rejects market-based access to old-age benefits, but this opposition has eased over time: 83.5% in 2016, 81.2% in 2018, and 71.8% in 2023.
+The 2016–2018 change is modest, whereas 2023 registers a 9.4 point drop in disagreement relative to 2018, mirrored by rising agreement: 16.5% (2016), 18.8% (2018), 28.2% (2023).
+Substantively, while most respondents continue to oppose the idea that higher-income individuals should obtain better pensions via the market, a non-trivial and growing portion endorses commodification, with the sharpest expansion concentrated in the latest wave (+9.4 points from 2018 to 2023).
 
@@
 
-The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
+Regarding the relationship between preferences for the commodification of pensions and intergenerational occupational mobility, @fig-mean shows the percentage of agreement with the market provision of pension benefits according to mobility trajectories and survey waves.
+Averaging across waves, agreement is highest among the High-High immobile (27.8%), followed by the Middle-High upwardly mobile (24.6%) and the High-Middle downwardly mobile (24.0%); the lowest levels are found among Low-High (17.3%), Low-Low (17.4%), and descending High-Low (17.6%) groups.
+The specific profiles of each wave accentuate these gradients, especially in 2023: agreement reaches 31.9% for high-high, 29.4% for upward middle-high, and peaks at 36.8% for downward high-middle (the highest of all trajectories for that year).
+Descriptively, immobility at the top is associated with greater support for commodification; among those who move, support is greatest for the upward Middle-High trajectory and the downward High-Middle trajectory, patterns that intensify in 2023.
+These shifts anticipate the heterogeneity documented below.
 
@@
 
-The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
+@fig-reg reports the doubly robust estimates of the effect of intergenerational mobility on preferences for pension commodification, obtained from weighted models combining stabilized IPW–ATT with covariate adjustment and wave fixed effects.
+This specification ensures consistency if either the weighting or the outcome model is correctly specified, while improving efficiency [@wooldridge_introductory_2009].
+Results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive change in the size or significance of the mobility coefficients.
 
@@
 
-The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustnessThe specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
+Results reveal a clear directional asymmetry between upward and downward mobility trajectories.
+Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]).
+Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high class—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle class.
+Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level.
+In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-class origins express weaker preferences for market provision.
+The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts.
+These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them.
 
@@
 
-The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
+To probe potential mechanisms underlying the mobility effects, I estimate origin-specific ATT models that interact the mobility treatment with high meritocratic perceptions (ref.
+= low), using stabilized IPW–ATT weights, covariate adjustment, and wave fixed effects.
+I also estimate an analogous specification interacting mobility with survey waves (ref.
+= 2016) to assess temporal heterogeneity in the treatment effect (see [Supplementary Material](#anexo) for complete models).
+
 ```{r}
@@
 
-The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
-
 ```{r}
@@
 ```
+
+@fig-interact shows that neither the interaction terms nor the mobility coefficients reach conventional significance across trajectories, indicating no detectable moderation of the causal mobility effect by meritocratic perceptions at the 95% level.
+Directionally, the interaction is positive for Middle→High ($\beta$ = 0.15, 95% CI [−0.03, 0.33]) and Low→High ($\beta$ = 0.05, [−0.14, 0.24]), suggesting, if anything, somewhat larger pro-commodification effects among high-merit respondents when upwardly mobile, whereas it is negative for High→Low ($\beta$ = −0.14, [−0.38, 0.10]) and High→Middle ($\beta$ = −0.03, [−0.23, 0.17]), showing weaker pro-commodification effects under downward mobility at high merit (left panel).
+Focusing on the treatment effect at low meritocracy (the reference category), the mobility coefficient captures the origin-specific ATT comparing movers to their immobility counterfactual among low-merit respondents (right panel).
+These effects are small and imprecise across trajectories: Low→Middle ($\beta$ = −0.02, [−0.09, 0.04]) and Low→High ($\beta$ = −0.03, [−0.10, 0.05]) are near zero.
+Middle→High is positive but not significant ($\beta$ = 0.07, [−0.01, 0.14]), Middle→Low is null ($\beta$ = -0.00, [−0.07, 0.06]), and downward from high shows negative, borderline estimates: High→Middle ($\beta$ = −0.07, [−0.15, 0.00]) and High→Low ($\beta$ = −0.08, [−0.15, 0.00]).
+Substantively, I do not find evidence that meritocratic perceptions moderate the causal effect of mobility, although the signs align with the idea that upward mobility at high merit tilts attitudes toward commodification, and downward mobility at high merit tilts them away.
 
@@
 
+@tbl-inttime shows the causal heterogeneity of the mobility effect across the different waves of the survey.
+The treatment effect in 2016 (reference category) closely replicates the reference ATT patterns: positive for Middle→High ($\beta$ = 0.13, 95% CI [0.05, 0.22]) and negative for High→Middle ($\beta$ = −0.19, [−0.33, −0.05]) and High→Low ($\beta$ = −0.19, [−0.32, −0.05]).
+This indicates that, holding all other predictors constant, individuals who moved from Middle to high status in 2016 were, on average, 13 percentage points more likely to support access to the pension market compared to their counterfactual of immobility, while those who moved down from higher classes were approximately 19 percentage points less likely.
+Beyond these reference effects, there is a notable temporal change in the High→Middle trajectory: the interaction between mobility and the 2023 wave is positive and statistically significant ($\beta$ = 0.19, [0.01, 0.38]).
+This suggests that, although people with downward mobility from high backgrounds were less supportive of commodification in 2016, by 2023 the same trajectory shows a change in sign, implying that the negative causal effect of downward mobility attenuated and eventually turned positive.
+No comparable changes are observed in other trajectories.
+In essence, this pattern reveals time-conditional heterogeneity in the mobility effect: among those who experienced downward mobility, preferences for market-based access to old-age pensions appear to have realigned in the most recent wave, reflecting a notable shift toward greater support for commodification within this trajectory by 2023.
+
 ## Sensitivity analysis
 
-The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds ifeither the weighting model or the outcome model is correctly specified, while also improving efficiency (Wooldridge, 2009).I report both IPW-only and doubly robust specifications to document robustness
+The first sensitivity check corresponds to @ding_sensitivity_2016 approach to assess unmeasured confounding (E values).
+This indicates the minimum strength of association that an unobserved confounder would need to have with both the treatment and the outcome—on the relative risk scale—to explain an effect, the Middle-High ATT would require an $RR ≈ 1.74$ to offset the point estimate ($E_{point}$ = 1.736) and an $RR ≈ 1.26$ to move its CI to include the null value ($E_{CI}$ = 1.264).
+High→Low shows comparable robustness ($E_{point}$ = 1.759; $E_{CI}$ = 1.239), while High→Middle is comparatively fragile near zero ($E_{point}$ = 1.591; $E_{CI}$ = 1.037).
+For non-significant trajectories (Low→Middle, Low→High, Middle→Low), $E_{CI}$ = 1, as expected.
+In conclusion, the directional pattern—stronger in favor of commodification in Middle→High and weaker in the descent from High—seems reasonably robust to moderate unmeasured confounding in Middle→High and High→Low, but borderline in High→Middle.
+These results justify treating signs and relative magnitudes as informative, while recognizing limited protection against omitted variables for High→Middle (see [Supplementary Material](#anexo) for complete tables).
+
+The second sensitivity check applies Holm’s step-down correction—a multiple-testing procedure that controls the familywise error rate (FWER) while retaining more power than Bonferroni—across the six mobility trajectories.
+After FWER control, none of the effects remain significant at $\alpha$ = .05: Middle→High ($p_{raw}$ = .0116 → $p_{Holm}$ = .0698), High→Low (.0167 → .0834), and High→Middle (.0481 → .1924).
+Accordingly, claims of statistical significance should be tempered under strict multiplicity control.
+Substantively, however, this adjustment affects the strength of inference, not its direction: the pattern remains consistent with the theorized asymmetry (positive for Middle→High; negative for downward from high).
+I therefore treat signs and effect sizes as the primary evidence and regard p-values as exploratory once multiplicity is accounted for (see [Supplementary Material](#anexo) for complete tables).

### 2025-11-02 paper/paper.tex
Up results faltan anexos

@@
 \setstretch{1.5}
-This document was last modified at 2025-11-02 05:55:15 and it was last
-rendered at 2025-11-02 05:55:15.
+This document was last modified at 2025-11-02 18:14:07 and it was last
+rendered at 2025-11-02 18:14:07.
 
@@
 counterfactual distribution of non-movers with the covariate profile of
-movers in each origin stratum. These weights are then used in the
-outcome stage to estimate the causal effect of mobility on preferences
-for the commodification of pensions.
+movers in each origin stratum (see \hyperref[anexo]{Supplementary
+Material} for the evaluation of balance). These weights are then used in
+the outcome stage to estimate the causal effect of mobility on
+preferences for the commodification of pensions.
 
@@
 \begin{equation}\phantomsection\label{eq-final}{
-Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it},
+Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
 }\end{equation}
@@
 
-Lorem ipsoThe specification is doubly robust: entropy-balancing weights
-are combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustnessThe specification is
-doubly robust: entropy-balancing weights are combined with covariate
-adjustment so consistency holds ifeither the weighting model or the
-outcome model is correctly specified, while also improving efficiency
-(Wooldridge, 2009).I report both IPW-only and doubly robust
-specifications to document robustness
+Figure~\ref{fig-alluvial} shows the annual frequencies of preferences
+for pension commodification in 2016, 2018, and 2023. Each year presents
+stacked percentages frequencies for the level of agreement and
+disagreement. Overall, a large majority rejects market-based access to
+old-age benefits, but this opposition has eased over time: 83.5\% in
+2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018 change is
+modest, whereas 2023 registers a 9.4 point drop in disagreement relative
+to 2018, mirrored by rising agreement: 16.5\% (2016), 18.8\% (2018),
+28.2\% (2023). Substantively, while most respondents continue to oppose
+the idea that higher-income individuals should obtain better pensions
+via the market, a non-trivial and growing portion endorses
+commodification, with the sharpest expansion concentrated in the latest
+wave (+9.4 points from 2018 to 2023).
 
@@
 
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustness
+Regarding the relationship between preferences for the commodification
+of pensions and intergenerational occupational mobility,
+Figure~\ref{fig-mean} shows the percentage of agreement with the market
+provision of pension benefits according to mobility trajectories and
+survey waves. Averaging across waves, agreement is highest among the
+High-High immobile (27.8\%), followed by the Middle-High upwardly mobile
+(24.6\%) and the High-Middle downwardly mobile (24.0\%); the lowest
+levels are found among Low-High (17.3\%), Low-Low (17.4\%), and
+descending High-Low (17.6\%) groups. The specific profiles of each wave
+accentuate these gradients, especially in 2023: agreement reaches 31.9\%
+for high-high, 29.4\% for upward middle-high, and peaks at 36.8\% for
+downward high-middle (the highest of all trajectories for that year).
+Descriptively, immobility at the top is associated with greater support
+for commodification; among those who move, support is greatest for the
+upward Middle-High trajectory and the downward High-Middle trajectory,
+patterns that intensify in 2023. These shifts anticipate the
+heterogeneity documented below.
 
@@
 
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustness
+Figure~\ref{fig-reg} reports the doubly robust estimates of the effect
+of intergenerational mobility on preferences for pension
+commodification, obtained from weighted models combining stabilized
+IPW--ATT with covariate adjustment and wave fixed effects. This
+specification ensures consistency if either the weighting or the outcome
+model is correctly specified, while improving efficiency
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). Results
+are virtually identical to baseline models without covariates (see
+\hyperref[anexo]{Supplementary Material} for complete models), with no
+substantive change in the size or significance of the mobility
+coefficients.
 
@@
 
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustnessThe specification is
-doubly robust: entropy-balancing weights are combined with covariate
-adjustment so consistency holds ifeither the weighting model or the
-outcome model is correctly specified, while also improving efficiency
-(Wooldridge, 2009).I report both IPW-only and doubly robust
-specifications to document robustness
+Results reveal a clear directional asymmetry between upward and downward
+mobility trajectories. Among upwardly mobile respondents, only the
+Middle→High trajectory exhibits a significant positive effect on support
+for pension commodification (\(\beta\) = 0.09, 95\% CI {[}0.02,
+0.16{]}). Interpreted as an origin-specific ATT, this estimate
+compares---for individuals who actually moved from the middle to the
+high class---their observed endorsement with the counterfactual
+endorsement they would have expressed had they remained immobile in the
+middle class. Holding all other predictors constant, the point estimate
+implies a 9 percentage point higher probability of endorsing
+market-based access to pension benefits for Middle→High movers relative
+to their own immobility counterfactual, statistically significant at the
+95\% confidence level. In contrast, two downward trajectories:
+High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
+(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
+effects, suggesting that individuals who experience downward mobility
+from higher-class origins express weaker preferences for market
+provision. The remaining pathways (Low→Middle, Low→High, Middle→Low)
+display no significant differences relative to their non-mobile
+counterparts. These patterns indicate that upward movement within the
+upper segment (Middle→High) reinforces pro-commodification attitudes,
+whereas downward movement from privileged origins reduces them.
 
@@
 
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustness
-
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustness
+To probe potential mechanisms underlying the mobility effects, I
+estimate origin-specific ATT models that interact the mobility treatment
+with high meritocratic perceptions (ref. = low), using stabilized
+IPW--ATT weights, covariate adjustment, and wave fixed effects. I also
+estimate an analogous specification interacting mobility with survey
+waves (ref. = 2016) to assess temporal heterogeneity in the treatment
+effect (see \hyperref[anexo]{Supplementary Material} for complete
+models).
 
@@
 \end{figure}%
+
+Figure~\ref{fig-interact} shows that neither the interaction terms nor
+the mobility coefficients reach conventional significance across
+trajectories, indicating no detectable moderation of the causal mobility
+effect by meritocratic perceptions at the 95\% level. Directionally, the
+interaction is positive for Middle→High (\(\beta\) = 0.15, 95\% CI
+{[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05, {[}−0.14, 0.24{]}),
+suggesting, if anything, somewhat larger pro-commodification effects
+among high-merit respondents when upwardly mobile, whereas it is
+negative for High→Low (\(\beta\) = −0.14, {[}−0.38, 0.10{]}) and
+High→Middle (\(\beta\) = −0.03, {[}−0.23, 0.17{]}), showing weaker
+pro-commodification effects under downward mobility at high merit (left
+panel). Focusing on the treatment effect at low meritocracy (the
+reference category), the mobility coefficient captures the
+origin-specific ATT comparing movers to their immobility counterfactual
+among low-merit respondents (right panel). These effects are small and
+imprecise across trajectories: Low→Middle (\(\beta\) = −0.02, {[}−0.09,
+0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}) are near
+zero. Middle→High is positive but not significant (\(\beta\) = 0.07,
+{[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00, {[}−0.07,
+0.06{]}), and downward from high shows negative, borderline estimates:
+High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and High→Low
+(\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not find
+evidence that meritocratic perceptions moderate the causal effect of
+mobility, although the signs align with the idea that upward mobility at
+high merit tilts attitudes toward commodification, and downward mobility
+at high merit tilts them away.
 
@@
 
+Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
+effect across the different waves of the survey. The treatment effect in
+2016 (reference category) closely replicates the reference ATT patterns:
+positive for Middle→High (\(\beta\) = 0.13, 95\% CI {[}0.05, 0.22{]})
+and negative for High→Middle (\(\beta\) = −0.19, {[}−0.33, −0.05{]}) and
+High→Low (\(\beta\) = −0.19, {[}−0.32, −0.05{]}). This indicates that,
+holding all other predictors constant, individuals who moved from Middle
+to high status in 2016 were, on average, 13 percentage points more
+likely to support access to the pension market compared to their
+counterfactual of immobility, while those who moved down from higher
+classes were approximately 19 percentage points less likely. Beyond
+these reference effects, there is a notable temporal change in the
+High→Middle trajectory: the interaction between mobility and the 2023
+wave is positive and statistically significant (\(\beta\) = 0.19,
+{[}0.01, 0.38{]}). This suggests that, although people with downward
+mobility from high backgrounds were less supportive of commodification
+in 2016, by 2023 the same trajectory shows a change in sign, implying
+that the negative causal effect of downward mobility attenuated and
+eventually turned positive. No comparable changes are observed in other
+trajectories. In essence, this pattern reveals time-conditional
+heterogeneity in the mobility effect: among those who experienced
+downward mobility, preferences for market-based access to old-age
+pensions appear to have realigned in the most recent wave, reflecting a
+notable shift toward greater support for commodification within this
+trajectory by 2023.
+
 \subsection{Sensitivity analysis}\label{sensitivity-analysis}
 
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds ifeither the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency (Wooldridge, 2009).I report both IPW-only and
-doubly robust specifications to document robustness
+The first sensitivity check corresponds to Ding \& VanderWeele
+(\citeproc{ref-ding_sensitivity_2016}{2016}) approach to assess
+unmeasured confounding (E values). This indicates the minimum strength
+of association that an unobserved confounder would need to have with
+both the treatment and the outcome---on the relative risk scale---to
+explain an effect, the Middle-High ATT would require an \(RR ≈ 1.74\) to
+offset the point estimate (\(E_{point}\) = 1.736) and an \(RR ≈ 1.26\)
+to move its CI to include the null value (\(E_{CI}\) = 1.264). High→Low
+shows comparable robustness (\(E_{point}\) = 1.759; \(E_{CI}\) = 1.239),
+while High→Middle is comparatively fragile near zero (\(E_{point}\) =
+1.591; \(E_{CI}\) = 1.037). For non-significant trajectories
+(Low→Middle, Low→High, Middle→Low), \(E_{CI}\) = 1, as expected. In
+conclusion, the directional pattern---stronger in favor of
+commodification in Middle→High and weaker in the descent from
+High---seems reasonably robust to moderate unmeasured confounding in
+Middle→High and High→Low, but borderline in High→Middle. These results
+justify treating signs and relative magnitudes as informative, while
+recognizing limited protection against omitted variables for High→Middle
+(see \hyperref[anexo]{Supplementary Material} for complete tables).
+
+The second sensitivity check applies Holm's step-down correction---a
+multiple-testing procedure that controls the familywise error rate
+(FWER) while retaining more power than Bonferroni---across the six
+mobility trajectories. After FWER control, none of the effects remain
+significant at \(\alpha\) = .05: Middle→High (\(p_{raw}\) = .0116 →
+\(p_{Holm}\) = .0698), High→Low (.0167 → .0834), and High→Middle (.0481
+→ .1924). Accordingly, claims of statistical significance should be
+tempered under strict multiplicity control. Substantively, however, this
+adjustment affects the strength of inference, not its direction: the
+pattern remains consistent with the theorized asymmetry (positive for
+Middle→High; negative for downward from high). I therefore treat signs
+and effect sizes as the primary evidence and regard p-values as
+exploratory once multiplicity is accounted for (see
+\hyperref[anexo]{Supplementary Material} for complete tables).
 
@@
 
-\section{Supplementary material}\label{supplementary-material}
+\section{Supplementary material}\label{anexo}
 

### 2025-11-02 processing/prod_analysis_multinom.R

@@ -576,7 +576,7 @@ evalues_tbl <- imap_dfr(mods, function(mod, name) {
 })
 
 # Tabla bonita (kable); puedes usar 'gt(evalues_tbl)' si prefieres
-kable(evalues_tbl,
+kableExtra::kable(evalues_tbl,
       caption = "E-values for ATT estimates by mobility trajectory (Ding & VanderWeele, 2016 approximation)",
       align = "lrrrrrrrrr")
 
@@ -884,38 +884,68 @@ mp_hl_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
 
 
 
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "t" = "Mobility treatment"
+)
 
-screenreg(list(mp_lm,
-               mp_lh,
-               mp_ml,
-               mp_mh,
-               mp_hm,
-               mp_hl),
-          custom.model.names = c(
-            "Low-Middle",
-            "Low-High",
-            "Middle-Low",
-            "Middle-High",
-            "High-Middle",
-            "High-Low"
-          ))
+texreg::screenreg(list(mp_lm,
+                       mp_lh,
+                       mp_mh,
+                       mp_ml,
+                       mp_hm,
+                       mp_hl),
+                  custom.model.names = c(
+                    "Low-Middle",
+                    "Low-High",
+                    "Middle-High",
+                    "Middle-Low",
+                    "High-Middle",
+                    "High-Low"
+                  ),   caption.above = NULL,
+               caption = NULL,
+               custom.coef.map = ccoef,
+               digits = 2,
+               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.80,
+               center = T,float.pos = "H!",
+               custom.gof.rows = list("Controls"=rep("Yes",6)))
 
-
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "t" = "Mobility treatment",
+  "mHigh" = "High meritocracy perception (Ref.= Low)",
+  "t:mHigh" = "Mobility treatment x High meritocracy perception (Ref.= Low)"
+)
 
 screenreg(list(mp_lm_i,
                mp_lh_i,
-               mp_ml_i,
                mp_mh_i,
+               mp_ml_i,
                mp_hm_i,
                mp_hl_i),
           custom.model.names = c(
             "Low-Middle",
[…]

### 2025-11-03 paper/03-methods.qmd
Update entrega 1

@@
 ```{r}
-#| label: libraries1
+#| label: data1
+#| echo: false
+#| results: false
 #| message: false
-#| echo: false
-
-if (!require("pacman")) install.packages("pacman")
-
-pacman::p_load(tidyverse,
-               here,
-               sjPlot,
-               sjmisc,
-               texreg,
-               estimatr,
-               conflicted,
-               ggdist,
-               summarytools,
-               dagitty,
-               ggdag)
-
-conflicts_prefer(dplyr::filter)
-options(scipen=999)
-rm(list = ls())
-```
-
-```{r}
-#| label: data1
-#| message: false
-#| echo: false
+#| warning: false
+library(here)
+library(janitor)
+library(dagitty)
+library(ggdag)
+library(tidyverse)
 
 load(here("input/data/proc/data_v2.3.RData"))
-
 ```
@@
 The core claim is that typical mobility hypotheses are fundamentally within-person counterfactual comparisons rather than between-person contrasts.
-Defining mobility as $(M = D - O)$ —a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination—places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the outcome under the realized destination but not under alternative destinations or under immobility.
+Defining mobility as $M = D - O$ —a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination—places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the result in the social position they currently occupy, but not in the situation of other alternative destinations or immobility.
 
@@
 This estimand compares movers’ observed outcomes with the hypothetical outcomes those same individuals would have exhibited had they remained in their origin class.
-This formulation provides a coherent potential-outcomes basis for studying how class transitions shape individual preferences and attitudes.
+This formulation provides a coherent potential-outcomes basis for studying how social mobility shape individual preferences and attitudes.
 
@@
 
-**Positivity**: For all $(j,c)$ in the support of $(O,C)$ and for each relevant destination $(k)$, the probability of receiving $D=k$ is strictly between 0 and 1; substantively, every $(O,C)$-type has some chance of each comparison destination.
+**Positivity**: For all $(j,c)$ in the support of $(O,C)$ and for each relevant destination $(k)$, the probability of receiving $D=k$ is strictly between 0 and 1; substantively, every $(O,C)$-type has non-zero chance of each comparison destination.
 
@@
 **Conditional Independence**: Also known as conditional unconfoundedness or exchangeability, this assumption emphasizes that, conditional on origin $O$ and pre-treatment covariates $C$ (e.g., parental education, ethnicity, cohort, early-life factors), assignment to $D$ is as good as random.
-This underlies IPW and regression adjustment, which seek exchangeability between mobile and immobile individuals.
+This underlies IPW and regression adjustment, which seek exchangeability between mobile (treated) and immobile individuals (controls).
 
@@
 
+In fact, establishing the causal relationship between social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference.
+Below, I use directed acyclic graphs (DAGs) to illustrate some of these challenges and evaluate possible strategies for identifying the effects of social mobility.
 @fig-dag depicts the causal model guiding identification.
-In this DAG, $C$ represents pre-treatment confounders (e.g., family resources, household composition, and ascriptive status) affecting both origin and destination, and $U$ denotes unobserved factors influencing $C$.
-Class origin $O$ is determined by $C$ and, in turn, affects destination $D$.
-The outcome $Y$ (preferences for commodification) is caused by $D$ and can also be affected by $C$.
+In this model, a person's preference ($Y$) is directly caused by her class destination ($D$) and can also be affected by different attributes determined in childhood or earlier ($C$), such as family resources, household composition, and ascriptive status (pre-treatment confounders).
+Class origin ($O$) is determined by those origin-attributes ($C$) and, in turn, affects destination ($D$).
+In addition, denotes unobserved factors ($U$) can also influence origin-attributes ($C$).
 There is no direct arrow from $O$ to $Y$, reflecting the assumption that the effect of origin on the outcome operates entirely through background factors $(C)$ and destination $(D)$.
-Accordingly, conditioning on $(O,C)$ blocks the relevant back-door paths from $D$ to $Y$, specifically $D\leftarrow O\leftarrow C\rightarrow Y$ and, if present, $D \leftarrow O \rightarrow Y$, so $(O,C)$ is a minimal sufficient adjustment set.
+Accordingly, conditioning on $(O,C)$ could blocks the relevant back-door paths from $D$ to $Y$, specifically $D\leftarrow O\leftarrow C\rightarrow Y$ and, if present, $D \leftarrow O \rightarrow Y$, so $(O,C)$ is a minimal sufficient adjustment set.
 
@@
 
-In this study, I focus on the specific case where the counterfactual destination corresponds to class immobility $(k' = j)$, that is, how the outcome among people from origin *k* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility).
+In this study, I focus on the specific case where the counterfactual destination corresponds to class immobility $(k' = j)$, that is, how the outcome among people from origin *j* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility).
 Accordingly, the estimand becomes:
@@
 Consistent with the study design, estimation proceeds within trajectory-specific subsamples (e.g., low→low; low→middle), retaining movers and matched non-movers from the same origin class.
-Then, I apply weights to construct the corresponding counterfactual control distributions for each trajectory.
-As Breen and Ermisch [-@breen_effects_2024] did, I use the three waves in the estimations that follow, allowing for correlation between the observations for each individual in calculating the standard errors (i.e.
+Then, I apply weights to construct the corresponding counterfactual control sample for each trajectory.
+Following Breen and Ermisch [-@breen_effects_2024], I use the three waves in the estimations that follow, allowing for correlation between the observations for each individual in calculating the standard errors (i.e.
 clustering on the personal identifier).
@@
 
-The outcome variable measures preferences regarding the commodification of pensions, operationalized with a single item addressing the degree of public support for market-based access to old-age pension benefits.
+The outcome variable measures preferences regarding the commodification of pensions, operationalized with a single item addressing how strongly individuals justify conditioning access to old-age pension benefits based on individual income.
 Respondents were asked: “Is it fair in Chile that people with higher incomes have better pensions than people with lower incomes?” and were asked to indicate their level of agreement on a five-point Likert scale ranging from 1 (“strongly disagree”) to 5 (“strongly agree”).
@@
 
-library(janitor)
-
 db %>%
@@
 
-After balancing, I construct stabilized IPW-ATT, setting treated weights to 1 and assigning ebal-derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum (see [Supplementary Material](#anexo) for the evaluation of balance).
+After balancing, I construct stabilized IPW, setting treated weights to 1 and assigning derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum (see [Supplementary Material](#anexo) for the evaluation of balance).
 These weights are then used in the outcome stage to estimate the causal effect of mobility on preferences for the commodification of pensions.
@@
 
-Because the dependent variable—the preference for pension commodification—is binary $(Y_{it}\in{0,1})$, I estimate weighted linear probability models (WLS) using stabilized inverse probability weights (IPW–ATT) obtained via entropy balancing.
+Because the dependent variable—the preference for pension commodification—is binary $(Y_{it}\in{0,1})$, I estimate weighted linear probability models (WLS) using stabilized inverse probability weights (IPW) obtained via entropy balancing.
 Although OLS estimator is often questioned with binary outcomes, it’s unbiased.
@@
 The specification is doubly robust: entropy-balancing weights are combined with covariate adjustment so consistency holds if either the weighting model or the outcome model is correctly specified, while also improving efficiency [@wooldridge_introductory_2009].
-I report both IPW-only and doubly robust specifications to document robustness.
+These results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive changes in the mobility coefficients.
 In addition, I assess causal heterogeneity by interacting the mobility treatment with (i) meritocratic perceptions and (ii) survey wave, reporting simple slopes (origin-specific ATTs) by moderator levels.

### 2025-11-03 paper/04-analysis.qmd
Update entrega 1

@@
----
-output: html_document
-editor_options: 
-  chunk_output_type: console
----
-
 # Results
@@
 ```{r}
-#| label: libraries2
+#| label: source1
+#| echo: false
+#| results: false
 #| message: false
-#| echo: false
+#| warning: false
+#| include: false
 
-if (!require("pacman")) install.packages("pacman")
-
-pacman::p_load(tidyverse,
-               here,
-               sjPlot,
-               sjmisc,
-               texreg,
-               estimatr,
-               conflicted,
-               ggdist,
-               summarytools,
-               dagitty,
-               ggdag)
-
-conflicts_prefer(dplyr::filter)
-
-options(scipen=999)
-rm(list = ls())
-```
-
-```{r}
-#| label: data2
-#| message: false
-#| echo: false
-
-load(here("input/data/proc/data_v2.3.RData"))
-
+library(here)
+source(here("processing/prod_analysis_multinom.R"))
 ```
@@
 #| results: asis
-db %>% 
-  mutate(just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree"))) %>%
-  group_by(wave, just_pension_f) %>% 
-  count() %>% 
-  group_by(wave) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  ggplot(aes(x = wave, y = porcentaje, group = just_pension_f)) +
-  geom_col(aes(fill = just_pension_f)) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(
-    values = c("#5b59d4", "#c8237d"),
-    breaks = c("Agree","Disagree")
-  )  +
-  coord_flip() +
-  geom_text(
-    aes(label = paste0(round(porcentaje * 100, 1), "%")),
-    position = position_stack(vjust = 0.5),
-    color = "white",
-    size = 3,
-    fontface = "bold"
-  ) +
-  labs(y = NULL,
-       x = NULL,
-       fill = NULL,
-       caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)") +
-  theme_ggdist() +
-  theme(legend.position = "top") 
+g1
 ```
@@
 
-pdat <- db %>% 
-  select(wave, estrato_orig, estrato_ocupa, just_pension_f) %>% 
-  drop_na() %>% 
-  mutate(
-    mobility = paste(estrato_orig, estrato_ocupa, sep = "-"),
-    facet = case_when(
-      mobility %in% c("High-Middle","High-Low","Middle-Low") ~ "Downward",
-      estrato_orig == estrato_ocupa ~ "Inmobile",
-      mobility %in% c("Middle-High","Low-Middle","Low-High") ~ "Upward"
-    )
-  ) %>% 
-  group_by(wave, mobility, facet) %>% 
-  summarise(y = mean(just_pension_f), .groups = "drop") %>% 
-  group_by(mobility, facet) %>% 
-  mutate(gm = mean(y)) %>% 
-  ungroup()
-
-# datos SOLO para el punto (uno por categoría)
-pdat_gm <- pdat %>% distinct(mobility, facet, gm)
-
-ggplot(pdat, aes(x = mobility, y = y, group = wave)) +
-  geom_col(aes(fill = wave), position = "dodge", alpha = 0.9) +
-  # Punto "X" (pch 4) con su propia leyenda (shape)
-  geom_point(
-    data = pdat_gm,
-    aes(x = mobility, y = gm, shape = "All waves"),
-    inherit.aes = FALSE,
-    size = 3, stroke = 1, color = "black"
-  ) +
-  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) + 
-  scale_fill_manual(
-    values = c("2016" = "#5b59d4", "2018" = "#b3b3b3ff", "2023" = "#c8237d"),
-    name   = "Wave"
-  ) +
-  scale_shape_manual(                     # <-- leyenda separada para el shape
-    name = NULL,                          # o "Marker" si quieres título
-    values = c("All waves" = 4)          # 4 = "X"
-  ) +
-  guides(
-    fill  = guide_legend(order = 1),
-    shape = guide_legend(order = 2, override.aes = list(color = "black"))
-  ) +
-  facet_wrap(~ facet, scales = "free_x") +
-  labs(
-    y = "% of Agreement",
-    x = NULL,
-    caption = "Source: own elaboration with data from ELSOC 2016–2023 (N obs = 3,435)"
-  ) +
-  theme_ggdist() +
-  theme(legend.position = "top")
+g2
 
@@
 
-@fig-reg reports the doubly robust estimates of the effect of intergenerational mobility on preferences for pension commodification, obtained from weighted models combining stabilized IPW–ATT with covariate adjustment and wave fixed effects.
-This specification ensures consistency if either the weighting or the outcome model is correctly specified, while improving efficiency [@wooldridge_introductory_2009].
-Results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive change in the size or significance of the mobility coefficients.
-
-```{r}
-#| label: models
-#| echo: false
-#| results: false
-
-### Pensiones
-
-### Low to middle
-
-data_lm <- datos1 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = moblm, 
-         w = w_low_lm_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_lm <-
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0 + ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_lm)
-
-### Low to high
-
-data_lh <- datos1 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = moblh, 
-         w = w_low_lh_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_lh <- 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_lh)
-
-### middle to high
-
-data_mh <- datos2 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobmh, 
-         w = w_mid_mh_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_mh <-  
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_mh)
-
-### middle to low
-
-data_ml <- datos2 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobml, 
-         w = w_mid_ml_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_ml <- 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_ml)
-
-### High to middle
-
-data_hm <- datos3 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobhm, 
-         w = w_hig_hm_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_hm <- 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_hm)
-
-### High to low
-
-data_hl <- datos3 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobhl, 
-         w = w_hig_hl_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_hl <- 
-  lm_robust(y ~ t + edad + sexo + nac + etnia + hogar + edu0 + ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_hl)
-
-```
+@fig-reg reports the doubly robust estimates of the effect of intergenerational mobility on preferences for pension commodification, obtained from weighted models combining stabilized IPW with covariate adjustment and wave fixed effects.
+Results reveal a clear directional asymmetry between upward and downward mobility trajectories.
+Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]).
+Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high class—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle class.
+Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level.
+In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-class origins express weaker preferences for market provision.
+The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts.
+These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them.
 
@@
 
-conflicts_prefer(dplyr::filter)
-
-df_pension <- bind_rows(
-  tidy(mp_lm, conf.int = TRUE) %>%
-    dplyr::filter(term == "t") %>%
-    dplyr::mutate(term = paste0("Low-Middle\n", "(N obs. = ", mp_hl$nobs, ";", "\nN clusters = ", mp_hl$nclusters, ")")
-) 
-  ,
-  tidy(mp_lh, conf.int = TRUE) %>%
-    dplyr::filter(term == "t") %>%
-  dplyr::mutate(term = paste0("Low-High\n", "(N obs. = ", mp_lh$nobs, ";", "\nN clusters = ", mp_lh$nclusters, ")"))
-  ,
-tidy(mp_mh, conf.int = TRUE) %>%
-    dplyr::filter(term == "t") %>%
-    dplyr::mutate(term = paste0("Middle-High\n", "(N obs. = ", mp_mh$nobs, ";", "\nN clusters = ", mp_mh$nclusters, ")"))
-,
-  tidy(mp_ml, conf.int = TRUE) %>%
-    dplyr::filter(term == "t") %>%
-  dplyr::mutate(term = paste0("Middle-Low\n", "(N obs. = ", mp_ml$nobs, ";", "\nN clusters = ", mp_ml$nclusters, ")"))
-,
-  tidy(mp_hm, conf.int = TRUE) %>%
-    dplyr::filter(term == "t") %>%
-  dplyr::mutate(term = paste0("High-Middle\n", "(N obs. = ", mp_hm$nobs, ";", "\nN clusters = ", mp_hm$nclusters, ")"))
-,
-  tidy(mp_hl, conf.int = TRUE) %>%
-    dplyr::filter(term == "t") %>%
-  dplyr::mutate(term = paste0("High-Low\n", "(N obs. = ", mp_hl$nobs, ";", "\nN clusters = ", mp_hl$nclusters, ")"))
-)
-
-df_pension_lab <- df_pension %>%
-  dplyr::mutate(
-    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
-    est_lab = sprintf("%.2f%s", estimate, sig),
-    low_lab = sprintf("%.2f", conf.low),
-    high_lab= sprintf("%.2f", conf.high)
-  )
-
-df_pension_lab <- df_pension_lab %>%
-  dplyr::mutate(
-    term = factor(
-      term,
-      levels = term
-      )
-  )
-
-ggplot(df_pension_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 0.5, alpha = 1) +
-  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 0.7) +
-  geom_point(size = 2.5) +
-  scale_x_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +
-  #geom_text(aes(label = est_lab), hjust = -0.05, size = 4, vjust = -0.5) +
-  labs(x = "Estimate", y = NULL) +
-  theme_ggdist() 
+g3
 ```
-
-Results reveal a clear directional asymmetry between upward and downward mobility trajectories.
-Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]).
-Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high class—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle class.
-Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level.
-In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-class origins express weaker preferences for market provision.
-The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts.
-These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them.
 
@@
 To probe potential mechanisms underlying the mobility effects, I estimate origin-specific ATT models that interact the mobility treatment with high meritocratic perceptions (ref.
-= low), using stabilized IPW–ATT weights, covariate adjustment, and wave fixed effects.
+= low), using stabilized IPW weights, covariate adjustment, and wave fixed effects.
 I also estimate an analogous specification interacting mobility with survey waves (ref.
 = 2016) to assess temporal heterogeneity in the treatment effect (see [Supplementary Material](#anexo) for complete models).
-
-```{r}
-#| echo: false
-#| results: false
-
-# 3.4 Interacciones -------------------------------------------------------
-
-# -----------
-# Meritocracy 
-# -----------
-
-### Interacciones
-
-### Low to middle
-
-data_lm_i <- datos1 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = moblm, 
-         w = w_low_lm_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f,
-         m = merit_i) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-data_lm_i %>% 
-  arrange(idencuesta)
-
-mp_lm_i <-
-  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0 + ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_lm_i)
-
-mp_lm_i %>%  
-  screenreg()
-
-### Low to high
-
-data_lh_i <- datos1 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = moblh, 
-         w = w_low_lh_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f,
-         m = merit_i) %>%  
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_lh_i <- 
-  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_lh_i)
-
-mp_lh_i %>% 
-  screenreg()
-
-### middle to high
-
-data_mh_i <- datos2 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobmh, 
-         w = w_mid_mh_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f,
-         m = merit_i) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_mh_i <-  
-  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_mh_i)
-
-mp_mh_i %>% 
-  screenreg()
-
-### middle to low
-
-data_ml_i <- datos2 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobml, 
-         w = w_mid_ml_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f,
-         m = merit_i) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_ml_i <- 
-  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_ml_i)
-
-mp_ml_i %>% 
-  screenreg()
-
-### High to middle
-
-data_hm_i <- datos3 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobhm, 
-         w = w_hig_hm_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f,
-         m = merit_i) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_hm_i <- 
-  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0+ ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_hm_i)
-
-mp_hm_i %>% 
-  screenreg()
-
-### High to low
-
-data_hl_i <- datos3 %>% 
-  select(idencuesta, 
-         ola, 
-         y = just_pension_f, 
-         t = mobhl, 
-         w = w_hig_hl_t, 
-         edad = m0_edad_f, 
-         sexo = sexo_f, 
-         nac = nacionalidad_f,
-         etnia = etnia_f,
-         hogar = hogar_bip_f, 
-         edu0 = educ_orig_f,
-         m = merit_i) %>% 
-  filter(!is.na(t), !is.na(w)) %>% 
-  mutate(ola = factor(ola, levels = c("1", "2", "3"),
-                      labels = c("2016", "2018", "2023")))
-
-mp_hl_i <- 
-  lm_robust(y ~ t*m + edad + sexo + nac + etnia + hogar + edu0 + ola,
-            weights = w,
-            se_type = "CR2",
-            clusters = idencuesta,
-            data = data_hl_i)
-
-mp_hl_i %>% 
-  screenreg()
-```
 
@@
 
-df_pension_i <-  bind_rows(
-  tidy(mp_lm_i, conf.int = TRUE) %>%
-    filter(term %in% c("t:mHigh", "t")) %>%
-    mutate(name = paste0("Low-Middle\n", "(N obs. = ", mp_hl_i$nobs, ";", "\nN clusters = ", mp_hl_i$nclusters, ")")
-    ) 
-  ,
-  tidy(mp_lh_i, conf.int = TRUE) %>%
-    filter(term %in% c("t:mHigh", "t")) %>%
-    mutate(name = paste0("Low-High\n", "(N obs. = ", mp_lh_i$nobs, ";", "\nN clusters = ", mp_lh_i$nclusters, ")"))
-  ,
-  tidy(mp_mh_i, conf.int = TRUE) %>%
-    filter(term %in% c("t:mHigh", "t")) %>%
-    mutate(name = paste0("Middle-High\n", "(N obs. = ", mp_mh_i$nobs, ";", "\nN clusters = ", mp_mh_i$nclusters, ")"))
-  ,
-  tidy(mp_ml_i, conf.int = TRUE) %>%
-    filter(term %in% c("t:mHigh", "t")) %>%
-    mutate(name = paste0("Middle-Low\n", "(N obs. = ", mp_ml_i$nobs, ";", "\nN clusters = ", mp_ml_i$nclusters, ")"))
-  ,
-  tidy(mp_hm_i, conf.int = TRUE) %>%
-    filter(term %in% c("t:mHigh", "t")) %>%
-    mutate(name = paste0("High-Middle\n", "(N obs. = ", mp_hm_i$nobs, ";", "\nN clusters = ", mp_hm_i$nclusters, ")"))
-  ,
-  tidy(mp_hl_i, conf.int = TRUE) %>%
-    filter(term %in% c("t:mHigh", "t")) %>%
-    mutate(name = paste0("High-Low\n", "(N obs. = ", mp_hl_i$nobs, ";", "\nN clusters = ", mp_hl_i$nclusters, ")"))
-)
-
-df_pension_lab_i <- df_pension_i %>%
-  mutate(
-    sig     = ifelse(conf.low * conf.high > 0, "*", ""),  # IC no cruza 0
-    est_lab = sprintf("%.2f%s", estimate, sig),
-    low_lab = sprintf("%.2f", conf.low),
-    high_lab= sprintf("%.2f", conf.high)
-  )
-
-df_pension_lab_i <- df_pension_lab_i %>%
-  mutate(
-    name = factor(
-      name,
-      levels = unique(name)
-    ),
-    term = if_else(term == "t", "Low Meritocracy", "High Meritocracy"),
-    term = factor(term, levels = c("High Meritocracy", "Low Meritocracy"))
-  )
-
-ggplot(df_pension_lab_i, aes(x = estimate, y = name)) +
-  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 0.5, alpha = 1) +
-  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 0.7) +
-  geom_point(size = 2.5) +
-  scale_x_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +
-  facet_wrap(~term) + 
-  #geom_text(aes(label = est_lab), hjust = -0.05, size = 4, vjust = -0.5) +
-  labs(x = "Estimate", y = NULL) +
-  theme_ggdist()
-
+g4
 ```
@@
 #| echo: false
-
-# -----------
-# Time 
-# -----------
-
-# Low to middle
-
-mp_lm_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
-          weights = w,
-          se_type = "CR2",
-          clusters = idencuesta,
-          data = data_lm)
-
-# Low to high
-
-mp_lh_t <- lm_robust(y ~ t*ola+ edad + sexo + nac + etnia + hogar + edu0,
-          weights = w,
-          se_type = "CR2",
-          clusters = idencuesta,
-          data = data_lh)
-
-# Middle to high
-
-mp_mh_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
-          weights = w,
-          se_type = "CR2",
-          clusters = idencuesta,
-          data = data_mh)
-
-# Middle to low
-
-mp_ml_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
-          weights = w,
-          se_type = "CR2",
-          clusters = idencuesta,
-          data = data_ml)
-
-# High to middle
-
-# tiempo: descender se vuelve un efecto positivo en la ultima ola (con mi plata no)
-mp_hm_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
-          weights = w,
-          se_type = "CR2",
-          clusters = idencuesta,
-          data = data_hm)
-
-# High to low
-
-mp_hl_t <- lm_robust(y ~ t*ola + edad + sexo + nac + etnia + hogar + edu0,
-          weights = w,
-          se_type = "CR2",
-          clusters = idencuesta,
-          data = data_hl)
 

### 2025-11-03 paper/07-supplementary-material.qmd
Update entrega 1

---
title: "Supplementary material"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=0.5cm
    template-partials: 
      - title.tex
    keep-tex: true
    number-sections: true
editor: source
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
editor_options: 
  chunk_output_type: console
---

This section presents the supplementary material for this study. First, graphs are shown for the evaluation of the balance of IPW-ATT via entropy balance. Second, the complete estimated regression models are shown. Finally, the sensitivity analyses performed are shown.

```{r}
#| label: source2
#| echo: false
#| results: false
#| message: false
#| warning: false
#| include: false

library(here)
library(cobalt)
library(ggplot2)
library(cowplot)

source(here("processing/prod_analysis_multinom.R"))

table_format <- if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }

```

```{r}
#| label: fig-interact
#| fig-cap: "Balance SMD — Mobility treatment (ATT)"
#| fig-cap-location: bottom
#| echo: false
#| results: asis
#| warning: false
#| out-width: '90%'
#| fig-asp: 0.9

p1 <- cobalt::love.plot(W_lm_ll, stats = "m", abs = TRUE, threshold = .10,
                  var.order = "unadjusted", line = TRUE,
                  colors = c("grey70","black")) +
  ggtitle("Low-Middle")

p2 <- cobalt::love.plot(W_lh_ll, stats = "m", abs = TRUE, threshold = .10,
                  var.order = "unadjusted", line = TRUE,
                  colors = c("grey70","black")) +
  ggtitle("Low-High")

p3 <- cobalt::love.plot(W_mh_mm, stats = "m", abs = TRUE, threshold = .10,
                  var.order = "unadjusted", line = TRUE,
                  colors = c("grey70","black")) +
  ggtitle("Middle-High")

p4 <- cobalt::love.plot(W_ml_mm, stats = "m", abs = TRUE, threshold = .10,
                  var.order = "unadjusted", line = TRUE,
                  colors = c("grey70","black")) +
  ggtitle("Middle-Low")

p5 <- cobalt::love.plot(W_hm_hh, stats = "m", abs = TRUE, threshold = .10,
                  var.order = "unadjusted", line = TRUE,
                  colors = c("grey70","black")) +
  ggtitle("High-Middle")

p6 <- cobalt::love.plot(W_hl_hh, stats = "m", abs = TRUE, threshold = .10,
                  var.order = "unadjusted", line = TRUE,
                  colors = c("grey70","black")) +
  ggtitle("High-Low")

cowplot::plot_grid(p1, p2, p3, p4, p5, p6, label_size = 11, ncol = 2, nrow = 3)

```

```{r}
#| label: tbl-complete1
#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, with covariates and wave fixed effects."
#| tbl-cap-location: top
#| results: asis
#| echo: false

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment",
  edad = "Age (in years)",
  sexo = "Female (Ref. = Male)",
  nac = "Chilean nacionality (Ref. = Non-Chilean)",
  etnia = "Indigenous ethnicity (Ref. = Non-indigenous)",
  hogar = "Co-residence with both parents (Ref. = No co-residence)",
  edu0 = "Parental education",
  ola2018 = "Wave 2018",
  ola2023 = "Wave 2023"

)

texreg::texreg(list(mp_lm,
                       mp_lh,
                       mp_mh,
                       mp_ml,
                       mp_hm,
                       mp_hl),
                  custom.model.names = c(
                    "Low-Middle",
                    "Low-High",
                    "Middle-High",
                    "Middle-Low",
                    "High-Middle",
                    "High-Low"
                  ),   caption.above = NULL,
               caption = NULL,
               custom.coef.map = ccoef,
               digits = 2,
               groups = list("Wave (Ref.= 2016)" = 9:10),
               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.7,
               center = T,
               float.pos = "H!")
```

```{r}
#| label: tbl-complete2
#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification."
#| tbl-cap-location: top
#| results: asis
#| echo: false

# Low to middle

mp_lm_2 <- lm_robust(y ~ t,
                     weights = w,
                     se_type = "CR2",
                     clusters = idencuesta,
                     data = data_lm)

# Low to high

mp_lh_2 <- lm_robust(y ~ t,
                     weights = w,
                     se_type = "CR2",
                     clusters = idencuesta,
                     data = data_lh)

# Middle to high

mp_mh_2 <- lm_robust(y ~ t,
                     weights = w,
                     se_type = "CR2",
                     clusters = idencuesta,
                     data = data_mh)

# Middle to low

mp_ml_2 <- lm_robust(y ~ t,
                     weights = w,
                     se_type = "CR2",
                     clusters = idencuesta,
                     data = data_ml)

# High to middle

mp_hm_2 <- lm_robust(y ~ t,
                     weights = w,
                     se_type = "CR2",
                     clusters = idencuesta,
                     data = data_hm)

# High to low

mp_hl_2 <- lm_robust(y ~ t,
                     weights = w,
                     se_type = "CR2",
                     clusters = idencuesta,
                     data = data_hl)

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment"
)

texreg::texreg(list(mp_lm_2,
                       mp_lh_2,
                       mp_mh_2,
                       mp_ml_2,
                       mp_hm_2,
                       mp_hl_2),
                  custom.model.names = c(
                    "Low-Middle",
                    "Low-High",
                    "Middle-High",
                    "Middle-Low",
                    "High-Middle",
                    "High-Low"
                  ),   caption.above = NULL,
               caption = NULL,
               custom.coef.map = ccoef,
               digits = 2,
               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               center = T,
               float.pos = "H!")
```

```{r}
#| label: tbl-complete3
#| tbl-cap: "Interactions effects between intergenerational occupational mobility and perceived meritocracy on preferences for pension commodification, with covariates and wave fixed effects."
#| tbl-cap-location: top
#| results: asis
#| echo: false

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment",
  "mHigh" = "High meritocracy perception (Ref.= Low)",
  "t:mHigh" = "Mobility treatment x High meritocracy perception (Ref.= Low)"
)

texreg(list(mp_lm_i,
               mp_lh_i,
               mp_mh_i,
               mp_ml_i,
               mp_hm_i,
               mp_hl_i),
          custom.model.names = c(
            "Low-Middle",
            "Low-High",
            "Middle-High",
            "Middle-Low",
            "High-Middle",
            "High-Low"
          ),
          caption.above = NULL,
          caption = NULL,
          custom.coef.map = ccoef,
          digits = 2,
          custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
          leading.zero = T,
          use.packages = F,
          booktabs = F,
          scalebox = 0.7,
          center = T,float.pos = "H!",
          custom.gof.rows = list("Controls"=rep("Yes",6)))

```

```{r}
#| label: tbl-holm
#| tbl-cap: "Holm correction"
#| tbl-cap-location: top
#| results: asis
#| echo: false

data.frame(traj = c("Low→Middle","Low→High","Middle→Low","Middle→High","High→Middle","High→Low"),
           p_raw = p, p_holm = p_holm) %>% 
  mutate_at(.vars = c(2,3), .funs = ~ round(.,3)) %>% 
  mutate_at(.vars = c(2,3), .funs = ~ format(., nsmall=3)) %>% 
  mutate_all(~ as.character(.)) %>% 
  kableExtra::kable(format = table_format, 
                    col.names = c("Mobility treatment", "p value", "Holm"),
                    row.names = F,
                    booktabs = T, align = 'l',
                    caption = NULL) %>% 
  kableExtra::kable_styling(latex_options = "hold_position", 
                            position = "center") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kableExtra::row_spec(0, bold = T) %>% 
  kableExtra::add_footnote(label = c("Note: p values derived from Table 1"), notation = "none")

```

```{r}
#| label: tbl-evalues
#| tbl-cap: "E-values for ATT estimates by mobility trajectory (Ding & VanderWeele, 2016 approximation)"
#| tbl-cap-location: top
#| results: asis
#| echo: false

evalues_tbl %>% 
  mutate_at(.vars = c(2:12), .funs = ~ round(.,3)) %>% 
  mutate_at(.vars = c(2:11), .funs = ~ format(., nsmall=3)) %>% 
  mutate_all(~ as.character(.)) %>% 
  tidyr::unite("ci", ci_low:ci_high, sep = ";") %>% 
  kableExtra::kable(format = table_format, 
                    col.names = c("Mobility treatment", "Beta", "CI 95% [low-upper]", "sdY", "RR point", "RR point upper", "E point", "RR CI", "RR CI Upper", "E CI", "CI Crosses 0"),
                    row.names = F,
                    booktabs = T, 
                    align = 'l',
                    caption = NULL) %>% 
  kableExtra::kable_styling(latex_options = "hold_position", 
                            position = "center") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, font_size = 8) %>% 
  kableExtra::row_spec(0, bold = T)
```

### 2025-11-03 paper/paper.tex
Update entrega 1

@@
 \setstretch{1.5}
-This document was last modified at 2025-11-02 18:14:07 and it was last
-rendered at 2025-11-02 18:14:07.
+This document was last modified at 2025-11-03 01:00:05 and it was last
+rendered at 2025-11-03 01:00:05.
 
@@
 comparisons rather than between-person contrasts. Defining mobility as
-\((M = D - O)\) ---a change from origin \((O)\) to destination \((D)\),
+\(M = D - O\) ---a change from origin \((O)\) to destination \((D)\),
 rather than an additive or interactive combination---places the focus
 squarely on the Neyman--Rubin problem: for any individual, we observe
-the outcome under the realized destination but not under alternative
-destinations or under immobility.
+the result in the social position they currently occupy, but not in the
+situation of other alternative destinations or immobility.
 
@@
 they remained in their origin class. This formulation provides a
-coherent potential-outcomes basis for studying how class transitions
-shape individual preferences and attitudes.
+coherent potential-outcomes basis for studying how social mobility shape
+individual preferences and attitudes.
 
@@
 \(D=k\) is strictly between 0 and 1; substantively, every \((O,C)\)-type
-has some chance of each comparison destination.
+has non-zero chance of each comparison destination.
 
@@
 to \(D\) is as good as random. This underlies IPW and regression
-adjustment, which seek exchangeability between mobile and immobile
-individuals.
+adjustment, which seek exchangeability between mobile (treated) and
+immobile individuals (controls).
 
@@
 
-Figure~\ref{fig-dag} depicts the causal model guiding identification. In
-this DAG, \(C\) represents pre-treatment confounders (e.g., family
-resources, household composition, and ascriptive status) affecting both
-origin and destination, and \(U\) denotes unobserved factors influencing
-\(C\). Class origin \(O\) is determined by \(C\) and, in turn, affects
-destination \(D\). The outcome \(Y\) (preferences for commodification)
-is caused by \(D\) and can also be affected by \(C\). There is no direct
-arrow from \(O\) to \(Y\), reflecting the assumption that the effect of
-origin on the outcome operates entirely through background factors
-\((C)\) and destination \((D)\). Accordingly, conditioning on \((O,C)\)
-blocks the relevant back-door paths from \(D\) to \(Y\), specifically
+In fact, establishing the causal relationship between social mobility
+and subjective outcomes, such as preferences, poses several challenges
+for causal inference. Below, I use directed acyclic graphs (DAGs) to
+illustrate some of these challenges and evaluate possible strategies for
+identifying the effects of social mobility. Figure~\ref{fig-dag} depicts
+the causal model guiding identification. In this model, a person's
+preference (\(Y\)) is directly caused by her class destination (\(D\))
+and can also be affected by different attributes determined in childhood
+or earlier (\(C\)), such as family resources, household composition, and
+ascriptive status (pre-treatment confounders). Class origin (\(O\)) is
+determined by those origin-attributes (\(C\)) and, in turn, affects
+destination (\(D\)). In addition, denotes unobserved factors (\(U\)) can
+also influence origin-attributes (\(C\)). There is no direct arrow from
+\(O\) to \(Y\), reflecting the assumption that the effect of origin on
+the outcome operates entirely through background factors \((C)\) and
+destination \((D)\). Accordingly, conditioning on \((O,C)\) could blocks
+the relevant back-door paths from \(D\) to \(Y\), specifically
 \(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
@@
 destination corresponds to class immobility \((k' = j)\), that is, how
-the outcome among people from origin \emph{k} who entered destination
+the outcome among people from origin \emph{j} who entered destination
 \emph{k} would have bee different if those people had, counterfactually,
@@
 non-movers from the same origin class. Then, I apply weights to
-construct the corresponding counterfactual control distributions for
-each trajectory. As Breen and Ermisch
-(\citeproc{ref-breen_effects_2024}{2024}) did, I use the three waves in
-the estimations that follow, allowing for correlation between the
+construct the corresponding counterfactual control sample for each
+trajectory. Following Breen and Ermisch
+(\citeproc{ref-breen_effects_2024}{2024}), I use the three waves in the
+estimations that follow, allowing for correlation between the
 observations for each individual in calculating the standard errors
@@
 The outcome variable measures preferences regarding the commodification
-of pensions, operationalized with a single item addressing the degree of
-public support for market-based access to old-age pension benefits.
-Respondents were asked: ``Is it fair in Chile that people with higher
-incomes have better pensions than people with lower incomes?'' and were
-asked to indicate their level of agreement on a five-point Likert scale
-ranging from 1 (``strongly disagree'') to 5 (``strongly agree''). For
-estimation and interpretability, I dichotomized the item so that values
-4-5 indicate agreement with market-based access to old-age pensions
-(coded as 1), while values 1-3 indicate disagreement (coded as 0). This
-binary coding treats neutrality as substantially closer to non-support
-than to support and produces a clear contrast between respondents who
-support market-based differentiation in pensions and those who do not. I
-use this item for two main reasons: on the one hand, to be able to make
+of pensions, operationalized with a single item addressing how strongly
+individuals justify conditioning access to old-age pension benefits
+based on individual income. Respondents were asked: ``Is it fair in
+Chile that people with higher incomes have better pensions than people
+with lower incomes?'' and were asked to indicate their level of
+agreement on a five-point Likert scale ranging from 1 (``strongly
+disagree'') to 5 (``strongly agree''). For estimation and
+interpretability, I dichotomized the item so that values 4-5 indicate
+agreement with market-based access to old-age pensions (coded as 1),
+while values 1-3 indicate disagreement (coded as 0). This binary coding
+treats neutrality as substantially closer to non-support than to support
+and produces a clear contrast between respondents who support
+market-based differentiation in pensions and those who do not. I use
+this item for two main reasons: on the one hand, to be able to make
 comparisons with existing literature
@@
 
-After balancing, I construct stabilized IPW-ATT, setting treated weights
-to 1 and assigning ebal-derived weights to controls, thus aligning the
+After balancing, I construct stabilized IPW, setting treated weights to
+1 and assigning derived weights to controls, thus aligning the
 counterfactual distribution of non-movers with the covariate profile of
@@
 linear probability models (WLS) using stabilized inverse probability
-weights (IPW--ATT) obtained via entropy balancing. Although OLS
-estimator is often questioned with binary outcomes, it's unbiased. It
-consistently estimates the conditional expectation \(E(Y|X)\) under
-standard exogeneity \(E[\varepsilon_i|X_i]=0\), with heteroskedasticity
-addressed by robust or clustered errors
+weights (IPW) obtained via entropy balancing. Although OLS estimator is
+often questioned with binary outcomes, it's unbiased. It consistently
+estimates the conditional expectation \(E(Y|X)\) under standard
+exogeneity \(E[\varepsilon_i|X_i]=0\), with heteroskedasticity addressed
+by robust or clustered errors
 (\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
@@
 improving efficiency
-(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). I
-report both IPW-only and doubly robust specifications to document
-robustness. In addition, I assess causal heterogeneity by interacting
-the mobility treatment with (i) meritocratic perceptions and (ii) survey
-wave, reporting simple slopes (origin-specific ATTs) by moderator
-levels. To evaluate sensitivity to unmeasured confounding, I compute
-E-values for point estimates and confidence limits
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
+results are virtually identical to baseline models without covariates
+(see \hyperref[anexo]{Supplementary Material} for complete models), with
+no substantive changes in the mobility coefficients. In addition, I
+assess causal heterogeneity by interacting the mobility treatment with
+(i) meritocratic perceptions and (ii) survey wave, reporting simple
+slopes (origin-specific ATTs) by moderator levels. To evaluate
+sensitivity to unmeasured confounding, I compute E-values for point
+estimates and confidence limits
 (\citeproc{ref-ding_sensitivity_2016}{Ding \& VanderWeele, 2016}). To
@@
 of intergenerational mobility on preferences for pension
-commodification, obtained from weighted models combining stabilized
-IPW--ATT with covariate adjustment and wave fixed effects. This
-specification ensures consistency if either the weighting or the outcome
-model is correctly specified, while improving efficiency
-(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). Results
-are virtually identical to baseline models without covariates (see
-\hyperref[anexo]{Supplementary Material} for complete models), with no
-substantive change in the size or significance of the mobility
-coefficients.
+commodification, obtained from weighted models combining stabilized IPW
+with covariate adjustment and wave fixed effects. Results reveal a clear
+directional asymmetry between upward and downward mobility trajectories.
+Among upwardly mobile respondents, only the Middle→High trajectory
+exhibits a significant positive effect on support for pension
+commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
+Interpreted as an origin-specific ATT, this estimate compares---for
+individuals who actually moved from the middle to the high class---their
+observed endorsement with the counterfactual endorsement they would have
+expressed had they remained immobile in the middle class. Holding all
+other predictors constant, the point estimate implies a 9 percentage
+point higher probability of endorsing market-based access to pension
+benefits for Middle→High movers relative to their own immobility
+counterfactual, statistically significant at the 95\% confidence level.
+In contrast, two downward trajectories: High→Middle (\(\beta\) = −0.07,
+CI {[}−0.15, −0.00{]}) and High→Low (\(\beta\) = −0.10, CI {[}−0.18,
+−0.02{]}), show significant negative effects, suggesting that
+individuals who experience downward mobility from higher-class origins
+express weaker preferences for market provision. The remaining pathways
+(Low→Middle, Low→High, Middle→Low) display no significant differences
+relative to their non-mobile counterparts. These patterns indicate that
+upward movement within the upper segment (Middle→High) reinforces
+pro-commodification attitudes, whereas downward movement from privileged
+origins reduces them.
 
@@
 
-Results reveal a clear directional asymmetry between upward and downward
-mobility trajectories. Among upwardly mobile respondents, only the
-Middle→High trajectory exhibits a significant positive effect on support
-for pension commodification (\(\beta\) = 0.09, 95\% CI {[}0.02,
-0.16{]}). Interpreted as an origin-specific ATT, this estimate
-compares---for individuals who actually moved from the middle to the
-high class---their observed endorsement with the counterfactual
-endorsement they would have expressed had they remained immobile in the
-middle class. Holding all other predictors constant, the point estimate
-implies a 9 percentage point higher probability of endorsing
-market-based access to pension benefits for Middle→High movers relative
-to their own immobility counterfactual, statistically significant at the
-95\% confidence level. In contrast, two downward trajectories:
-High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
-(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
-effects, suggesting that individuals who experience downward mobility
-from higher-class origins express weaker preferences for market
-provision. The remaining pathways (Low→Middle, Low→High, Middle→Low)
-display no significant differences relative to their non-mobile
-counterparts. These patterns indicate that upward movement within the
-upper segment (Middle→High) reinforces pro-commodification attitudes,
-whereas downward movement from privileged origins reduces them.
-
 \subsection{Effect heterogeneity}\label{effect-heterogeneity-1}
@@
 estimate origin-specific ATT models that interact the mobility treatment
-with high meritocratic perceptions (ref. = low), using stabilized
-IPW--ATT weights, covariate adjustment, and wave fixed effects. I also
-estimate an analogous specification interacting mobility with survey
-waves (ref. = 2016) to assess temporal heterogeneity in the treatment
-effect (see \hyperref[anexo]{Supplementary Material} for complete
-models).
+with high meritocratic perceptions (ref. = low), using stabilized IPW
+weights, covariate adjustment, and wave fixed effects. I also estimate
+an analogous specification interacting mobility with survey waves (ref.
+= 2016) to assess temporal heterogeneity in the treatment effect (see
+\hyperref[anexo]{Supplementary Material} for complete models).
 
@@
 
-\section{Supplementary material}\label{anexo}
-
 \end{document}

### 2025-11-03 processing/prod_analysis_multinom.R

@@ -1,10 +1,10 @@
 # 0. Identification ---------------------------------------------------
 
-# Title: Data analysis for thesis project Social Mobilty and Market Justice Preferencees
+# Title: Data analysis for thesis project Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility
 # Responsible: Andreas Laffert
 
 # Executive Summary: This script contains the code to data analysis for thesis project
-# Date: May 1, 2025
+# Date: Nov 2, 2025
 
 # 1. Packages  -----------------------------------------------------
 
@@ -18,15 +18,15 @@ pacman::p_load(tidyverse,
                estimatr,
                conflicted,
                ggdist,
-               summarytools)
+               summarytools,
+               broom)
 
 conflicts_prefer(dplyr::filter)
+
 options(scipen=999)
 rm(list = ls())
 
 # 2. Data -----------------------------------------------------------------
-
-#load(url("https://github.com/Andreas-Lafferte/mobility-market-justice/raw/refs/heads/main/input/data/proc/data_v1.RData"))
 
 load(here("input/data/proc/data_v2.3.RData"))
 
@@ -34,9 +34,10 @@ glimpse(db)
 
 # 3. Analysis -----------------------------------------------------------
 
-# Descriptive statistics
+# 3.1 Descriptive statistics
 
 # Table summary
+
 t1 <- db %>% 
   select(idencuesta, 
          wave, 
@@ -65,6 +66,7 @@ dfSummary(t1,
           col.widths = c(30,10,10,10))
 
 # Mobility matrix
+
 sjPlot::sjt.xtab(db$estrato_orig, 
                  db$estrato_ocupa, 
                  show.row.prc = T,
@@ -72,7 +74,8 @@ sjPlot::sjt.xtab(db$estrato_orig,
                  statistics = NULL)
 
 # MJP per year
-db %>% 
+
+g1 <- db %>% 
   mutate(just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree"))) %>%
   group_by(wave, just_pension_f) %>% 
   count() %>% 
@@ -101,7 +104,9 @@ db %>%
   theme_ggdist() +
[…]

### 2025-11-03 paper/03-methods.qmd
Update entrega

@@
 Social mobility effects—understood as impacts on an outcome arising from movements between an origin state and a destination state (e.g., social class)—have long been a focus of sociological research [@eyles_social_2022; @langsaether_explaining_2022].
-Yet, as Breen and Ermisch [-@breen_effects_2024, p.467] emphasize, most hypotheses are, at their core, individual-level counterfactual comparisons between the observed outcome under a mobility trajectory and the outcome the same person would have shown if they had remained in their origin class (or moved to an alternative destination).
+Yet, as Breen and Ermisch [-@breen_effects_2024, p.467] emphasize, most of the mobility hypotheses are, at their core, individual-level counterfactual comparisons between the observed outcome under a mobility trajectory and the outcome the same person would have shown if they had remained in their origin class (or moved to an alternative destination).
 Standard specifications (e.g., SAM, DRM, and mobility-contrast models) struggle to retrieve these inherently counterfactual quantities due to identification constraints and their reliance on between-group contrasts constructed from additive terms and interactions, thereby yielding primarily associational evidence [@breen_effects_2024; @song_there_2025].
@@
 
-**Positivity**: For all $(j,c)$ in the support of $(O,C)$ and for each relevant destination $(k)$, the probability of receiving $D=k$ is strictly between 0 and 1; substantively, every $(O,C)$-type has non-zero chance of each comparison destination.
+**Positivity**: For all $(j,c)$ in the support of $(O,C)$ and for each relevant destination $(k)$, the probability of receiving $D=k$ is strictly between 0 and 1; substantively, each type $(O,C)$ has a nonzero probability of entering each comparison destination.
 
@@
 
-In fact, establishing the causal relationship between social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference.
+In fact, establishing the causal relationship between objective social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference.
 Below, I use directed acyclic graphs (DAGs) to illustrate some of these challenges and evaluate possible strategies for identifying the effects of social mobility.
@@
 Class origin ($O$) is determined by those origin-attributes ($C$) and, in turn, affects destination ($D$).
-In addition, denotes unobserved factors ($U$) can also influence origin-attributes ($C$).
+In addition, unobserved factors ($U$) can also influence origin-attributes ($C$).
 There is no direct arrow from $O$ to $Y$, reflecting the assumption that the effect of origin on the outcome operates entirely through background factors $(C)$ and destination $(D)$.
@@
 #| fig-cap: "Causal graph of the effect of intergenerational social mobility on preferences for commodification.
-Y = preferences for pension commodification, my outcome of interest, D = an individual’s class of destination, O = an individual’s class of origin, C = different attributes determined in childhood or earlier that affects an individual’s class of origin and destination.
+Y = preferences for pension commodification, D = an individual’s class of destination, O = an individual’s class of origin, C = different attributes determined in childhood or earlier that affects an individual’s class of origin and destination.
 Finally, U = unobserved factors influencing origin attributes."
@@
 
-This structure corresponds to a nonparametric structural causal model in which $(D)$ is the treatment variable, $(Y)$ the outcome, and $(O, C)$ the minimal sufficient adjustment set.
+This structure corresponds to a nonparametric structural causal model in which $D$ is the treatment variable, $Y$ the outcome, and $(O, C)$ the minimal sufficient adjustment set.
 Conditioning on these variables satisfies the back-door criterion for identifying the causal effect of mobility on preferences [@breen_effects_2024].
@@
 Why are the assumptions plausible here?
-Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$; entropy balancing/IPW enforces covariate balance within each origin stratum.
+Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$; inverse probability weights (IPW) obtained via entropy balancing enforces covariate balance within each origin stratum.
 Positivity is assessed via propensity scores and trimming of extreme weights, restricting to common support.
 SUTVA holds by treating the destination class as a single exposure measured before $(Y)$ and assuming no interference.
-Together, the DAG and assumptions define conditions for identifying the causal effect of intergenerational mobility on preferences for commodification.
+Together, the DAG and assumptions define conditions for identifying the causal effect of intergenerational social mobility on preferences for commodification.
 
@@
 Time is measured by survey wave (2016, 2018, 2023) and entered as a categorical (dummy) variable.
-Both variables are used as moderators by interacting them with the mobility treatment to assess conditional average treatment effects—i.e., whether the effect of mobility varies across levels of meritocratic beliefs or across survey waves.
+Both variables are used as moderators by interacting them with the mobility treatment to assess conditional average treatment effects—i.e., whether the effect of mobility varies across levels of meritocratic perceptions or across survey waves.
 
@@
 The analytic sample is restricted to individuals sharing the same origin $(O=j)$.
-Estimation uses pooled WLS with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.
+Therefore, estimation uses pooled WLS with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.
 

### 2025-11-03 paper/04-analysis.qmd
Update entrega

@@
 Each year presents stacked percentages frequencies for the level of agreement and disagreement.
-Overall, a large majority rejects market-based access to old-age benefits, but this opposition has eased over time: 83.5% in 2016, 81.2% in 2018, and 71.8% in 2023.
+Overall, a large majority rejects market-based access to old-age pension benefits, but this opposition has eased over time: 83.5% in 2016, 81.2% in 2018, and 71.8% in 2023.
 The 2016–2018 change is modest, whereas 2023 registers a 9.4 point drop in disagreement relative to 2018, mirrored by rising agreement: 16.5% (2016), 18.8% (2018), 28.2% (2023).
-Substantively, while most respondents continue to oppose the idea that higher-income individuals should obtain better pensions via the market, a non-trivial and growing portion endorses commodification, with the sharpest expansion concentrated in the latest wave (+9.4 points from 2018 to 2023).
+Substantively, while most respondents continue to oppose the idea that higher-income individuals should obtain better pensions via the market, a non-trivial and growing portion endorses this statement, with the sharpest expansion concentrated in the latest wave (+9.4 points from 2018 to 2023).
 
@@
 
-Regarding the relationship between preferences for the commodification of pensions and intergenerational occupational mobility, @fig-mean shows the percentage of agreement with the market provision of pension benefits according to mobility trajectories and survey waves.
-Averaging across waves, agreement is highest among the High-High immobile (27.8%), followed by the Middle-High upwardly mobile (24.6%) and the High-Middle downwardly mobile (24.0%); the lowest levels are found among Low-High (17.3%), Low-Low (17.4%), and descending High-Low (17.6%) groups.
-The specific profiles of each wave accentuate these gradients, especially in 2023: agreement reaches 31.9% for high-high, 29.4% for upward middle-high, and peaks at 36.8% for downward high-middle (the highest of all trajectories for that year).
+Regarding the relationship between preferences for the commodification of pensions and intergenerational occupational mobility, @fig-mean shows the percentage of agreement with market-based access to pension benefits according to mobility trajectories and survey waves.
+Averaging across waves, agreement is highest among the High-High immobile (27.8%), followed by the Middle-High upwardly mobile (24.6%) and the High-Middle downwardly mobile (24.0%).
+The lowest levels are found among ascending Low-High (17.3%), and immobile Low-Low (17.4%) groups.
+The specific profiles of each wave accentuate these gradients in most trajectories, especially in 2023: agreement reaches 31.9% for High-High, 29.4% for upward Middle-High, and peaks at 36.8% for downward High-Middle (the highest of all trajectories for that year).
 Descriptively, immobility at the top is associated with greater support for commodification; among those who move, support is greatest for the upward Middle-High trajectory and the downward High-Middle trajectory, patterns that intensify in 2023.
@@
 
-@fig-reg reports the doubly robust estimates of the effect of intergenerational mobility on preferences for pension commodification, obtained from weighted models combining stabilized IPW with covariate adjustment and wave fixed effects.
+@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects.
 Results reveal a clear directional asymmetry between upward and downward mobility trajectories.
 Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]).
-Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high class—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle class.
+Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high socioeconomic status—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle socioeconomic status.
 Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level.
-In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-class origins express weaker preferences for market provision.
+In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-status origins express weaker preferences for market-based pension benefits.
 The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts.
@@
 @fig-interact shows that neither the interaction terms nor the mobility coefficients reach conventional significance across trajectories, indicating no detectable moderation of the causal mobility effect by meritocratic perceptions at the 95% level.
-Directionally, the interaction is positive for Middle→High ($\beta$ = 0.15, 95% CI [−0.03, 0.33]) and Low→High ($\beta$ = 0.05, [−0.14, 0.24]), suggesting, if anything, somewhat larger pro-commodification effects among high-merit respondents when upwardly mobile, whereas it is negative for High→Low ($\beta$ = −0.14, [−0.38, 0.10]) and High→Middle ($\beta$ = −0.03, [−0.23, 0.17]), showing weaker pro-commodification effects under downward mobility at high merit (left panel).
-Focusing on the treatment effect at low meritocracy (the reference category), the mobility coefficient captures the origin-specific ATT comparing movers to their immobility counterfactual among low-merit respondents (right panel).
+Directionally, in the left panel, the interaction is positive for Middle→High ($\beta$ = 0.15, 95% CI [−0.03, 0.33]) and Low→High ($\beta$ = 0.05, [−0.14, 0.24]), suggesting, if anything, somewhat larger pro-commodification effects among high-merit respondents when upwardly mobile, whereas it is negative for High→Low ($\beta$ = −0.14, [−0.38, 0.10]) and High→Middle ($\beta$ = −0.03, [−0.23, 0.17]), showing weaker pro-commodification effects under downward mobility at high merit.
+Focusing on the treatment effect at low meritocracy (the reference category) at the right panel, the mobility coefficient captures the origin-specific ATT comparing movers to their immobility counterfactual among low-merit respondents.
 These effects are small and imprecise across trajectories: Low→Middle ($\beta$ = −0.02, [−0.09, 0.04]) and Low→High ($\beta$ = −0.03, [−0.10, 0.05]) are near zero.
@@
 The treatment effect in 2016 (reference category) closely replicates the reference ATT patterns: positive for Middle→High ($\beta$ = 0.13, 95% CI [0.05, 0.22]) and negative for High→Middle ($\beta$ = −0.19, [−0.33, −0.05]) and High→Low ($\beta$ = −0.19, [−0.32, −0.05]).
-This indicates that, holding all other predictors constant, individuals who moved from Middle to high status in 2016 were, on average, 13 percentage points more likely to support access to the pension market compared to their counterfactual of immobility, while those who moved down from higher classes were approximately 19 percentage points less likely.
+This indicates that, holding all other predictors constant, individuals who moved from middle to high socioeconomic status in 2016 were, on average, 13 percentage points more likely to support market-based access to the pension benefits compared to their counterfactual of immobility, while those who moved down from higher status were approximately 19 percentage points less likely.
 Beyond these reference effects, there is a notable temporal change in the High→Middle trajectory: the interaction between mobility and the 2023 wave is positive and statistically significant ($\beta$ = 0.19, [0.01, 0.38]).
@@
 
-The first sensitivity check corresponds to @ding_sensitivity_2016 approach to assess unmeasured confounding (E values).
-This indicates the minimum strength of association that an unobserved confounder would need to have with both the treatment and the outcome—on the relative risk scale—to explain an effect, the Middle-High ATT would require an $RR ≈ 1.74$ to offset the point estimate ($E_{point}$ = 1.736) and an $RR ≈ 1.26$ to move its CI to include the null value ($E_{CI}$ = 1.264).
+The first sensitivity check corresponds to the @ding_sensitivity_2016 approach to assess unmeasured confounding (E values).
+This values indicates the minimum strength of association that an unobserved confounder would need to have with both the treatment and the outcome—on the relative risk scale—to explain an effect.
+The Middle-High ATT would require an $RR ≈ 1.74$ to offset the point estimate ($E_{point}$ = 1.736) and an $RR ≈ 1.26$ to move its CI to include the null value ($E_{CI}$ = 1.264).
 High→Low shows comparable robustness ($E_{point}$ = 1.759; $E_{CI}$ = 1.239), while High→Middle is comparatively fragile near zero ($E_{point}$ = 1.591; $E_{CI}$ = 1.037).
@@
 
-The second sensitivity check applies Holm’s step-down correction—a multiple-testing procedure that controls the familywise error rate (FWER) while retaining more power than Bonferroni—across the six mobility trajectories.
+The second sensitivity check applies Holm’s step-down correction, which is a multiple-testing procedure that controls the familywise error rate (FWER) while retaining more power than Bonferroni, across the six mobility trajectories.
 After FWER control, none of the effects remain significant at $\alpha$ = .05: Middle→High ($p_{raw}$ = .0116 → $p_{Holm}$ = .0698), High→Low (.0167 → .0834), and High→Middle (.0481 → .1924).

### 2025-11-03 paper/paper.tex
Update entrega

@@
 \setstretch{1.5}
-This document was last modified at 2025-11-03 01:00:05 and it was last
-rendered at 2025-11-03 01:00:05.
+This document was last modified at 2025-11-03 17:04:08 and it was last
+rendered at 2025-11-03 17:04:08.
 
@@
 Yet, as Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024, p.
-467}) emphasize, most hypotheses are, at their core, individual-level
-counterfactual comparisons between the observed outcome under a mobility
-trajectory and the outcome the same person would have shown if they had
-remained in their origin class (or moved to an alternative destination).
-Standard specifications (e.g., SAM, DRM, and mobility-contrast models)
-struggle to retrieve these inherently counterfactual quantities due to
-identification constraints and their reliance on between-group contrasts
-constructed from additive terms and interactions, thereby yielding
-primarily associational evidence
+467}) emphasize, most of the mobility hypotheses are, at their core,
+individual-level counterfactual comparisons between the observed outcome
+under a mobility trajectory and the outcome the same person would have
+shown if they had remained in their origin class (or moved to an
+alternative destination). Standard specifications (e.g., SAM, DRM, and
+mobility-contrast models) struggle to retrieve these inherently
+counterfactual quantities due to identification constraints and their
+reliance on between-group contrasts constructed from additive terms and
+interactions, thereby yielding primarily associational evidence
 (\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
@@
 for each relevant destination \((k)\), the probability of receiving
-\(D=k\) is strictly between 0 and 1; substantively, every \((O,C)\)-type
-has non-zero chance of each comparison destination.
+\(D=k\) is strictly between 0 and 1; substantively, each type \((O,C)\)
+has a nonzero probability of entering each comparison destination.
 
@@
 
-In fact, establishing the causal relationship between social mobility
-and subjective outcomes, such as preferences, poses several challenges
-for causal inference. Below, I use directed acyclic graphs (DAGs) to
-illustrate some of these challenges and evaluate possible strategies for
-identifying the effects of social mobility. Figure~\ref{fig-dag} depicts
-the causal model guiding identification. In this model, a person's
-preference (\(Y\)) is directly caused by her class destination (\(D\))
-and can also be affected by different attributes determined in childhood
-or earlier (\(C\)), such as family resources, household composition, and
-ascriptive status (pre-treatment confounders). Class origin (\(O\)) is
-determined by those origin-attributes (\(C\)) and, in turn, affects
-destination (\(D\)). In addition, denotes unobserved factors (\(U\)) can
-also influence origin-attributes (\(C\)). There is no direct arrow from
-\(O\) to \(Y\), reflecting the assumption that the effect of origin on
-the outcome operates entirely through background factors \((C)\) and
-destination \((D)\). Accordingly, conditioning on \((O,C)\) could blocks
-the relevant back-door paths from \(D\) to \(Y\), specifically
+In fact, establishing the causal relationship between objective social
+mobility and subjective outcomes, such as preferences, poses several
+challenges for causal inference. Below, I use directed acyclic graphs
+(DAGs) to illustrate some of these challenges and evaluate possible
+strategies for identifying the effects of social mobility.
+Figure~\ref{fig-dag} depicts the causal model guiding identification. In
+this model, a person's preference (\(Y\)) is directly caused by her
+class destination (\(D\)) and can also be affected by different
+attributes determined in childhood or earlier (\(C\)), such as family
+resources, household composition, and ascriptive status (pre-treatment
+confounders). Class origin (\(O\)) is determined by those
+origin-attributes (\(C\)) and, in turn, affects destination (\(D\)). In
+addition, unobserved factors (\(U\)) can also influence
+origin-attributes (\(C\)). There is no direct arrow from \(O\) to \(Y\),
+reflecting the assumption that the effect of origin on the outcome
+operates entirely through background factors \((C)\) and destination
+\((D)\). Accordingly, conditioning on \((O,C)\) could blocks the
+relevant back-door paths from \(D\) to \(Y\), specifically
 \(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
@@
 social mobility on preferences for commodification. Y = preferences for
-pension commodification, my outcome of interest, D = an individual's
-class of destination, O = an individual's class of origin, C = different
-attributes determined in childhood or earlier that affects an
-individual's class of origin and destination. Finally, U = unobserved
-factors influencing origin attributes.}
+pension commodification, D = an individual's class of destination, O =
+an individual's class of origin, C = different attributes determined in
+childhood or earlier that affects an individual's class of origin and
+destination. Finally, U = unobserved factors influencing origin
+attributes.}
 
@@
 This structure corresponds to a nonparametric structural causal model in
-which \((D)\) is the treatment variable, \((Y)\) the outcome, and
-\((O, C)\) the minimal sufficient adjustment set. Conditioning on these
-variables satisfies the back-door criterion for identifying the causal
-effect of mobility on preferences
-(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}).
+which \(D\) is the treatment variable, \(Y\) the outcome, and \((O, C)\)
+the minimal sufficient adjustment set. Conditioning on these variables
+satisfies the back-door criterion for identifying the causal effect of
+mobility on preferences (\citeproc{ref-breen_effects_2024}{Breen \&
+Ermisch, 2024}).
 
@@
 targeted by controlling for a plausible sufficient adjustment set
-\((O,C)\); entropy balancing/IPW enforces covariate balance within each
-origin stratum. Positivity is assessed via propensity scores and
-trimming of extreme weights, restricting to common support. SUTVA holds
-by treating the destination class as a single exposure measured before
-\((Y)\) and assuming no interference. Together, the DAG and assumptions
-define conditions for identifying the causal effect of intergenerational
+\((O,C)\); inverse probability weights (IPW) obtained via entropy
+balancing enforces covariate balance within each origin stratum.
+Positivity is assessed via propensity scores and trimming of extreme
+weights, restricting to common support. SUTVA holds by treating the
+destination class as a single exposure measured before \((Y)\) and
+assuming no interference. Together, the DAG and assumptions define
+conditions for identifying the causal effect of intergenerational social
 mobility on preferences for commodification.
@@
 conditional average treatment effects---i.e., whether the effect of
-mobility varies across levels of meritocratic beliefs or across survey
-waves.
+mobility varies across levels of meritocratic perceptions or across
+survey waves.
 
@@
 sample is restricted to individuals sharing the same origin \((O=j)\).
-Estimation uses pooled WLS with stabilized IPW--ATT from entropy
-balancing and CR2 standard errors clustered by individual to address
-heteroskedasticity and within-person dependence.
+Therefore, estimation uses pooled WLS with stabilized IPW--ATT from
+entropy balancing and CR2 standard errors clustered by individual to
+address heteroskedasticity and within-person dependence.
 
@@
 disagreement. Overall, a large majority rejects market-based access to
-old-age benefits, but this opposition has eased over time: 83.5\% in
-2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018 change is
-modest, whereas 2023 registers a 9.4 point drop in disagreement relative
-to 2018, mirrored by rising agreement: 16.5\% (2016), 18.8\% (2018),
-28.2\% (2023). Substantively, while most respondents continue to oppose
-the idea that higher-income individuals should obtain better pensions
-via the market, a non-trivial and growing portion endorses
-commodification, with the sharpest expansion concentrated in the latest
-wave (+9.4 points from 2018 to 2023).
+old-age pension benefits, but this opposition has eased over time:
+83.5\% in 2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018
+change is modest, whereas 2023 registers a 9.4 point drop in
+disagreement relative to 2018, mirrored by rising agreement: 16.5\%
+(2016), 18.8\% (2018), 28.2\% (2023). Substantively, while most
+respondents continue to oppose the idea that higher-income individuals
+should obtain better pensions via the market, a non-trivial and growing
+portion endorses this statement, with the sharpest expansion
+concentrated in the latest wave (+9.4 points from 2018 to 2023).
 
@@
 of pensions and intergenerational occupational mobility,
-Figure~\ref{fig-mean} shows the percentage of agreement with the market
-provision of pension benefits according to mobility trajectories and
-survey waves. Averaging across waves, agreement is highest among the
-High-High immobile (27.8\%), followed by the Middle-High upwardly mobile
-(24.6\%) and the High-Middle downwardly mobile (24.0\%); the lowest
-levels are found among Low-High (17.3\%), Low-Low (17.4\%), and
-descending High-Low (17.6\%) groups. The specific profiles of each wave
-accentuate these gradients, especially in 2023: agreement reaches 31.9\%
-for high-high, 29.4\% for upward middle-high, and peaks at 36.8\% for
-downward high-middle (the highest of all trajectories for that year).
-Descriptively, immobility at the top is associated with greater support
-for commodification; among those who move, support is greatest for the
-upward Middle-High trajectory and the downward High-Middle trajectory,
-patterns that intensify in 2023. These shifts anticipate the
-heterogeneity documented below.
+Figure~\ref{fig-mean} shows the percentage of agreement with
+market-based access to pension benefits according to mobility
+trajectories and survey waves. Averaging across waves, agreement is
+highest among the High-High immobile (27.8\%), followed by the
+Middle-High upwardly mobile (24.6\%) and the High-Middle downwardly
+mobile (24.0\%). The lowest levels are found among ascending Low-High
+(17.3\%), and immobile Low-Low (17.4\%) groups. The specific profiles of
+each wave accentuate these gradients in most trajectories, especially in
+2023: agreement reaches 31.9\% for High-High, 29.4\% for upward
+Middle-High, and peaks at 36.8\% for downward High-Middle (the highest
+of all trajectories for that year). Descriptively, immobility at the top
+is associated with greater support for commodification; among those who
+move, support is greatest for the upward Middle-High trajectory and the
+downward High-Middle trajectory, patterns that intensify in 2023. These
+shifts anticipate the heterogeneity documented below.
 
@@
 
-Figure~\ref{fig-reg} reports the doubly robust estimates of the effect
-of intergenerational mobility on preferences for pension
-commodification, obtained from weighted models combining stabilized IPW
-with covariate adjustment and wave fixed effects. Results reveal a clear
-directional asymmetry between upward and downward mobility trajectories.
-Among upwardly mobile respondents, only the Middle→High trajectory
-exhibits a significant positive effect on support for pension
+Figure~\ref{fig-reg} reports the double-robust estimates of the effect
+of intergenerational occupational mobility on preferences for pension
+commodification, obtained from weighted models that combine stabilized
+IPW with covariate adjustment and fixed wave effects. Results reveal a
+clear directional asymmetry between upward and downward mobility
+trajectories. Among upwardly mobile respondents, only the Middle→High
+trajectory exhibits a significant positive effect on support for pension
 commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
 Interpreted as an origin-specific ATT, this estimate compares---for
-individuals who actually moved from the middle to the high class---their
-observed endorsement with the counterfactual endorsement they would have
-expressed had they remained immobile in the middle class. Holding all
-other predictors constant, the point estimate implies a 9 percentage
-point higher probability of endorsing market-based access to pension
-benefits for Middle→High movers relative to their own immobility
-counterfactual, statistically significant at the 95\% confidence level.
-In contrast, two downward trajectories: High→Middle (\(\beta\) = −0.07,
-CI {[}−0.15, −0.00{]}) and High→Low (\(\beta\) = −0.10, CI {[}−0.18,
-−0.02{]}), show significant negative effects, suggesting that
-individuals who experience downward mobility from higher-class origins
-express weaker preferences for market provision. The remaining pathways
-(Low→Middle, Low→High, Middle→Low) display no significant differences
-relative to their non-mobile counterparts. These patterns indicate that
-upward movement within the upper segment (Middle→High) reinforces
-pro-commodification attitudes, whereas downward movement from privileged
-origins reduces them.
+individuals who actually moved from the middle to the high socioeconomic
+status---their observed endorsement with the counterfactual endorsement
+they would have expressed had they remained immobile in the middle
+socioeconomic status. Holding all other predictors constant, the point
+estimate implies a 9 percentage point higher probability of endorsing
+market-based access to pension benefits for Middle→High movers relative
+to their own immobility counterfactual, statistically significant at the
+95\% confidence level. In contrast, two downward trajectories:
+High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
+(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
+effects, suggesting that individuals who experience downward mobility
+from higher-status origins express weaker preferences for market-based
+pension benefits. The remaining pathways (Low→Middle, Low→High,
+Middle→Low) display no significant differences relative to their
+non-mobile counterparts. These patterns indicate that upward movement
+within the upper segment (Middle→High) reinforces pro-commodification
+attitudes, whereas downward movement from privileged origins reduces
+them.
 
@@
 trajectories, indicating no detectable moderation of the causal mobility
-effect by meritocratic perceptions at the 95\% level. Directionally, the
-interaction is positive for Middle→High (\(\beta\) = 0.15, 95\% CI
-{[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05, {[}−0.14, 0.24{]}),
-suggesting, if anything, somewhat larger pro-commodification effects
-among high-merit respondents when upwardly mobile, whereas it is
-negative for High→Low (\(\beta\) = −0.14, {[}−0.38, 0.10{]}) and
-High→Middle (\(\beta\) = −0.03, {[}−0.23, 0.17{]}), showing weaker
-pro-commodification effects under downward mobility at high merit (left
-panel). Focusing on the treatment effect at low meritocracy (the
-reference category), the mobility coefficient captures the
-origin-specific ATT comparing movers to their immobility counterfactual
-among low-merit respondents (right panel). These effects are small and
-imprecise across trajectories: Low→Middle (\(\beta\) = −0.02, {[}−0.09,
-0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}) are near
-zero. Middle→High is positive but not significant (\(\beta\) = 0.07,
-{[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00, {[}−0.07,
-0.06{]}), and downward from high shows negative, borderline estimates:
-High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and High→Low
-(\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not find
-evidence that meritocratic perceptions moderate the causal effect of
-mobility, although the signs align with the idea that upward mobility at
-high merit tilts attitudes toward commodification, and downward mobility
-at high merit tilts them away.
+effect by meritocratic perceptions at the 95\% level. Directionally, in
+the left panel, the interaction is positive for Middle→High (\(\beta\) =
+0.15, 95\% CI {[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05,
+{[}−0.14, 0.24{]}), suggesting, if anything, somewhat larger
+pro-commodification effects among high-merit respondents when upwardly
+mobile, whereas it is negative for High→Low (\(\beta\) = −0.14,
+{[}−0.38, 0.10{]}) and High→Middle (\(\beta\) = −0.03, {[}−0.23,
+0.17{]}), showing weaker pro-commodification effects under downward
+mobility at high merit. Focusing on the treatment effect at low
+meritocracy (the reference category) at the right panel, the mobility
+coefficient captures the origin-specific ATT comparing movers to their
+immobility counterfactual among low-merit respondents. These effects are
+small and imprecise across trajectories: Low→Middle (\(\beta\) = −0.02,
+{[}−0.09, 0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]})
+are near zero. Middle→High is positive but not significant (\(\beta\) =
+0.07, {[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00,
+{[}−0.07, 0.06{]}), and downward from high shows negative, borderline
+estimates: High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and
+High→Low (\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not
+find evidence that meritocratic perceptions moderate the causal effect
+of mobility, although the signs align with the idea that upward mobility
+at high merit tilts attitudes toward commodification, and downward
+mobility at high merit tilts them away.
 
@@
 High→Low (\(\beta\) = −0.19, {[}−0.32, −0.05{]}). This indicates that,
-holding all other predictors constant, individuals who moved from Middle
-to high status in 2016 were, on average, 13 percentage points more
-likely to support access to the pension market compared to their
-counterfactual of immobility, while those who moved down from higher
-classes were approximately 19 percentage points less likely. Beyond
-these reference effects, there is a notable temporal change in the
-High→Middle trajectory: the interaction between mobility and the 2023
-wave is positive and statistically significant (\(\beta\) = 0.19,
-{[}0.01, 0.38{]}). This suggests that, although people with downward
-mobility from high backgrounds were less supportive of commodification
-in 2016, by 2023 the same trajectory shows a change in sign, implying
-that the negative causal effect of downward mobility attenuated and
-eventually turned positive. No comparable changes are observed in other
-trajectories. In essence, this pattern reveals time-conditional
-heterogeneity in the mobility effect: among those who experienced
-downward mobility, preferences for market-based access to old-age
-pensions appear to have realigned in the most recent wave, reflecting a
-notable shift toward greater support for commodification within this
-trajectory by 2023.
+holding all other predictors constant, individuals who moved from middle
+to high socioeconomic status in 2016 were, on average, 13 percentage
+points more likely to support market-based access to the pension
+benefits compared to their counterfactual of immobility, while those who
+moved down from higher status were approximately 19 percentage points
+less likely. Beyond these reference effects, there is a notable temporal
+change in the High→Middle trajectory: the interaction between mobility
+and the 2023 wave is positive and statistically significant (\(\beta\) =
+0.19, {[}0.01, 0.38{]}). This suggests that, although people with
+downward mobility from high backgrounds were less supportive of
+commodification in 2016, by 2023 the same trajectory shows a change in
+sign, implying that the negative causal effect of downward mobility
+attenuated and eventually turned positive. No comparable changes are
+observed in other trajectories. In essence, this pattern reveals
+time-conditional heterogeneity in the mobility effect: among those who
+experienced downward mobility, preferences for market-based access to
+old-age pensions appear to have realigned in the most recent wave,
+reflecting a notable shift toward greater support for commodification
+within this trajectory by 2023.
 
@@
 
-The first sensitivity check corresponds to Ding \& VanderWeele
+The first sensitivity check corresponds to the Ding \& VanderWeele
 (\citeproc{ref-ding_sensitivity_2016}{2016}) approach to assess
-unmeasured confounding (E values). This indicates the minimum strength
-of association that an unobserved confounder would need to have with
-both the treatment and the outcome---on the relative risk scale---to
-explain an effect, the Middle-High ATT would require an \(RR ≈ 1.74\) to
-offset the point estimate (\(E_{point}\) = 1.736) and an \(RR ≈ 1.26\)
-to move its CI to include the null value (\(E_{CI}\) = 1.264). High→Low
-shows comparable robustness (\(E_{point}\) = 1.759; \(E_{CI}\) = 1.239),
-while High→Middle is comparatively fragile near zero (\(E_{point}\) =
-1.591; \(E_{CI}\) = 1.037). For non-significant trajectories
-(Low→Middle, Low→High, Middle→Low), \(E_{CI}\) = 1, as expected. In
-conclusion, the directional pattern---stronger in favor of
+unmeasured confounding (E values). This values indicates the minimum
+strength of association that an unobserved confounder would need to have
+with both the treatment and the outcome---on the relative risk
+scale---to explain an effect. The Middle-High ATT would require an
+\(RR ≈ 1.74\) to offset the point estimate (\(E_{point}\) = 1.736) and
+an \(RR ≈ 1.26\) to move its CI to include the null value (\(E_{CI}\) =
+1.264). High→Low shows comparable robustness (\(E_{point}\) = 1.759;
+\(E_{CI}\) = 1.239), while High→Middle is comparatively fragile near
+zero (\(E_{point}\) = 1.591; \(E_{CI}\) = 1.037). For non-significant
+trajectories (Low→Middle, Low→High, Middle→Low), \(E_{CI}\) = 1, as
+expected. In conclusion, the directional pattern---stronger in favor of
 commodification in Middle→High and weaker in the descent from
@@
 
-The second sensitivity check applies Holm's step-down correction---a
-multiple-testing procedure that controls the familywise error rate
-(FWER) while retaining more power than Bonferroni---across the six
+The second sensitivity check applies Holm's step-down correction, which
+is a multiple-testing procedure that controls the familywise error rate
+(FWER) while retaining more power than Bonferroni, across the six
 mobility trajectories. After FWER control, none of the effects remain

### 2025-11-03 processing/prod_analysis_multinom.R

@@ -153,7 +153,7 @@ g2 <- ggplot(pdat, aes(x = mobility, y = y, group = wave)) +
   labs(
     y = "% of Agreement",
     x = NULL,
-    caption = "Source: own elaboration with data from ELSOC 2016–2023 (N obs = 3,435)"
+    caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)"
   ) +
   theme_ggdist() +
   theme(legend.position = "top")

### 2025-11-03 paper/07-supplementary-material.qmd
Up SP

@@
----
-title: "Supplementary material"
-format:
-  html: 
-    code-fold: true
-    html-math-method: katex
-    number-sections: true
-  docx:
-    number-sections: true
-  pdf: 
-    geometry: margin=0.5cm
-    template-partials: 
-      - title.tex
-    keep-tex: true
-    number-sections: true
-editor: source
-link-citations: true
-linestretch: 1.15       
-mainfont: Times New Roman
-fontsize: 12pt          
-colorlinks: true
-fig-height: 4
-fig-width: 7.5
-editor_options: 
-  chunk_output_type: console
----
+# Supplementary material {#anexo}
 
@@
 ```{r}
-#| label: fig-interact
+#| label: fig-balance
 #| fig-cap: "Balance SMD — Mobility treatment (ATT)"

### 2025-11-03 paper/paper.tex
Up SP

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 \setstretch{1.5}
-This document was last modified at 2025-11-03 17:04:08 and it was last
-rendered at 2025-11-03 17:04:08.
+This document was last modified at 2025-11-03 17:13:25 and it was last
+rendered at 2025-11-03 17:13:25.
 
@@
 
+\section{Supplementary material}\label{anexo}
+
+This section presents the supplementary material for this study. First,
+graphs are shown for the evaluation of the balance of IPW-ATT via
+entropy balance. Second, the complete estimated regression models are
+shown. Finally, the sensitivity analyses performed are shown.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-balance-1.pdf}
+
+}
+
+\caption{\label{fig-balance}Balance SMD --- Mobility treatment (ATT)}
+
+\end{figure}%
+
+\begin{table}
+
+\caption{\label{tbl-complete1}Effects of intergenerational occupational
+mobility on preferences for pension commodification, with covariates and
+wave fixed effects.}
+
+\centering{
+
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                                               & $0.18$            & $-0.02$           & $0.30^{*}$        & $0.23$            & $0.13$            & $0.14$            \\
+                                                        & $ [-0.36;  0.71]$ & $ [-0.35;  0.31]$ & $ [ 0.01;  0.58]$ & $ [-0.08;  0.54]$ & $ [-0.19;  0.46]$ & $ [-0.28;  0.56]$ \\
+Mobility treatment                                      & $-0.02$           & $-0.02$           & $0.09^{*}$        & $0.00$            & $-0.07^{*}$       & $-0.10^{*}$       \\
+                                                        & $ [-0.08;  0.05]$ & $ [-0.09;  0.05]$ & $ [ 0.02;  0.16]$ & $ [-0.06;  0.06]$ & $ [-0.15; -0.00]$ & $ [-0.18; -0.02]$ \\
+Age (in years)                                          & $0.00$            & $0.00$            & $-0.00$           & $-0.00$           & $0.00$            & $0.00$            \\
+                                                        & $ [-0.00;  0.00]$ & $ [-0.00;  0.00]$ & $ [-0.00;  0.00]$ & $ [-0.00;  0.00]$ & $ [-0.00;  0.01]$ & $ [-0.00;  0.01]$ \\
+Female (Ref. = Male)                                    & $-0.14^{*}$       & $-0.14^{*}$       & $-0.13^{*}$       & $-0.07^{*}$       & $-0.11^{*}$       & $-0.12^{*}$       \\
+                                                        & $ [-0.20; -0.08]$ & $ [-0.22; -0.06]$ & $ [-0.20; -0.06]$ & $ [-0.14; -0.01]$ & $ [-0.19; -0.03]$ & $ [-0.21; -0.03]$ \\
+Chilean nacionality (Ref. = Non-Chilean)                & $-0.03$           & $0.20$            & $0.08$            & $0.04$            & $-0.00$           & $0.06$            \\
+                                                        & $ [-0.63;  0.57]$ & $ [-0.05;  0.46]$ & $ [-0.19;  0.36]$ & $ [-0.26;  0.34]$ & $ [-0.27;  0.27]$ & $ [-0.30;  0.42]$ \\
+Indigenous ethnicity (Ref. = Non-indigenous)            & $0.03$            & $-0.05$           & $0.05$            & $-0.01$           & $0.07$            & $0.01$            \\
+                                                        & $ [-0.06;  0.12]$ & $ [-0.15;  0.05]$ & $ [-0.08;  0.18]$ & $ [-0.10;  0.08]$ & $ [-0.08;  0.22]$ & $ [-0.13;  0.14]$ \\
+Co-residence with both parents (Ref. = No co-residence) & $-0.00$           & $0.04$            & $-0.06$           & $-0.01$           & $0.10^{*}$        & $0.10^{*}$        \\
+                                                        & $ [-0.08;  0.07]$ & $ [-0.04;  0.12]$ & $ [-0.14;  0.01]$ & $ [-0.09;  0.06]$ & $ [ 0.01;  0.19]$ & $ [ 0.01;  0.19]$ \\
+Parental education                                      & $0.02$            & $0.01$            & $-0.02^{*}$       & $-0.01$           & $-0.00$           & $-0.01$           \\
+                                                        & $ [-0.00;  0.05]$ & $ [-0.01;  0.04]$ & $ [-0.03; -0.01]$ & $ [-0.02;  0.01]$ & $ [-0.02;  0.02]$ & $ [-0.03;  0.02]$ \\
+Wave (Ref.= 2016)                                       &                   &                   &                   &                   &                   &                   \\
+                                                        &                   &                   &                   &                   &                   &                   \\
+\quad Wave 2018                                         & $0.02$            & $-0.01$           & $0.08^{*}$        & $0.04$            & $-0.00$           & $-0.00$           \\
+                                                        & $ [-0.05;  0.09]$ & $ [-0.09;  0.07]$ & $ [ 0.02;  0.14]$ & $ [-0.02;  0.10]$ & $ [-0.08;  0.08]$ & $ [-0.09;  0.09]$ \\
+\quad Wave 2023                                         & $0.06$            & $0.04$            & $0.15^{*}$        & $0.14^{*}$        & $0.12^{*}$        & $0.10$            \\
+                                                        & $ [-0.02;  0.13]$ & $ [-0.06;  0.13]$ & $ [ 0.08;  0.22]$ & $ [ 0.07;  0.21]$ & $ [ 0.02;  0.21]$ & $ [-0.00;  0.21]$ \\
+\hline
+R$^2$                                                   & $0.04$            & $0.04$            & $0.07$            & $0.03$            & $0.05$            & $0.06$            \\
+Adj. R$^2$                                              & $0.03$            & $0.03$            & $0.06$            & $0.02$            & $0.04$            & $0.05$            \\
+Num. obs.                                               & $878$             & $745$             & $716$             & $773$             & $912$             & $861$             \\
+RMSE                                                    & $0.40$            & $0.37$            & $0.39$            & $0.39$            & $0.44$            & $0.42$            \\
+N Clusters                                              & $497$             & $449$             & $443$             & $470$             & $536$             & $525$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-complete2}Effects of intergenerational occupational
+mobility on preferences for pension commodification.}
+
+\centering{
+
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept          & $0.21^{*}$       & $0.18^{*}$       & $0.17^{*}$      & $0.19^{*}$       & $0.32^{*}$       & $0.28^{*}$        \\
+                   & $ [ 0.16; 0.26]$ & $ [ 0.13; 0.23]$ & $ [0.12; 0.21]$ & $ [ 0.15; 0.24]$ & $ [ 0.26; 0.38]$ & $ [ 0.21;  0.35]$ \\
+Mobility treatment & $-0.01$          & $-0.02$          & $0.08^{*}$      & $-0.00$          & $-0.07$          & $-0.09^{*}$       \\
+                   & $ [-0.08; 0.05]$ & $ [-0.08; 0.05]$ & $ [0.01; 0.15]$ & $ [-0.06; 0.06]$ & $ [-0.15; 0.00]$ & $ [-0.17; -0.02]$ \\
+\hline
+R$^2$              & $0.00$           & $0.00$           & $0.01$          & $0.00$           & $0.01$           & $0.01$            \\
+Adj. R$^2$         & $-0.00$          & $-0.00$          & $0.01$          & $-0.00$          & $0.01$           & $0.01$            \\
+Num. obs.          & $878$            & $745$            & $716$           & $773$            & $912$            & $861$             \\
+RMSE               & $0.40$           & $0.38$           & $0.40$          & $0.39$           & $0.45$           & $0.43$            \\
+N Clusters         & $497$            & $449$            & $443$           & $470$            & $536$            & $525$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-complete3}Interactions effects between
+intergenerational occupational mobility and perceived meritocracy on
+preferences for pension commodification, with covariates and wave fixed
+effects.}
+
+\centering{
+
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                                                    & $0.15$           & $-0.04$          & $0.30^{*}$       & $0.21$           & $0.08$           & $0.05$           \\
+                                                             & $ [-0.36; 0.67]$ & $ [-0.40; 0.31]$ & $ [ 0.04; 0.56]$ & $ [-0.09; 0.51]$ & $ [-0.25; 0.40]$ & $ [-0.30; 0.41]$ \\
+Mobility treatment                                           & $-0.02$          & $-0.03$          & $0.07$           & $-0.00$          & $-0.07$          & $-0.08$          \\
+                                                             & $ [-0.09; 0.04]$ & $ [-0.10; 0.05]$ & $ [-0.01; 0.14]$ & $ [-0.07; 0.06]$ & $ [-0.15; 0.00]$ & $ [-0.15; 0.00]$ \\
+High meritocracy perception (Ref.= Low)                      & $0.05$           & $0.06$           & $-0.01$          & $0.03$           & $0.17^{*}$       & $0.26^{*}$       \\
+                                                             & $ [-0.06; 0.16]$ & $ [-0.07; 0.19]$ & $ [-0.09; 0.08]$ & $ [-0.07; 0.13]$ & $ [ 0.00; 0.33]$ & $ [ 0.04; 0.48]$ \\
+Mobility treatment x High meritocracy perception (Ref.= Low) & $0.04$           & $0.05$           & $0.15$           & $0.04$           & $-0.03$          & $-0.14$          \\
+                                                             & $ [-0.12; 0.20]$ & $ [-0.14; 0.24]$ & $ [-0.03; 0.33]$ & $ [-0.11; 0.19]$ & $ [-0.23; 0.17]$ & $ [-0.38; 0.10]$ \\
+\hline
+Controls                                                     & Yes              & Yes              & Yes              & Yes              & Yes              & Yes              \\
+R$^2$                                                        & $0.04$           & $0.05$           & $0.08$           & $0.04$           & $0.07$           & $0.10$           \\
+Adj. R$^2$                                                   & $0.03$           & $0.04$           & $0.07$           & $0.02$           & $0.06$           & $0.08$           \\
+Num. obs.                                                    & $878$            & $745$            & $716$            & $773$            & $912$            & $861$            \\
+RMSE                                                         & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$           & $0.41$           \\
+N Clusters                                                   & $497$            & $449$            & $443$            & $470$            & $536$            & $525$            \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-holm}Holm correction}
+
+\centering{
+
+\centering\centering
+\begin{tabular}{lll}
+\toprule
+\textbf{Mobility treatment} & \textbf{p value} & \textbf{Holm}\\
+\midrule
+Low→Middle & 0.616 & 1.000\\
+Low→High & 0.594 & 1.000\\
+Middle→Low & 0.904 & 1.000\\
+Middle→High & 0.012 & 0.070\\
+High→Middle & 0.048 & 0.192\\
+\addlinespace
+High→Low & 0.017 & 0.083\\
+\bottomrule
+\multicolumn{3}{l}{\textsuperscript{} Note: p values derived from Table 1}\\
+\end{tabular}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-evalues}E-values for ATT estimates by mobility
+trajectory (Ding \& VanderWeele, 2016 approximation)}
+
+\centering{
+
+\centering\centering
+\fontsize{8}{10}\selectfont
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\toprule
+\textbf{Mobility treatment} & \textbf{Beta} & \textbf{CI 95\% [low-upper]} & \textbf{sdY} & \textbf{RR point} & \textbf{RR point upper} & \textbf{E point} & \textbf{RR CI} & \textbf{RR CI Upper} & \textbf{E CI} & \textbf{CI Crosses 0}\\
+\midrule
+Low→Middle & -0.016 & -0.078; 0.046 & 0.403 & 0.965 & 1.036 & 1.230 & 1.110 & 1.110 & 1.000 & 1\\
+Low→High & -0.019 & -0.088; 0.051 & 0.383 & 0.956 & 1.046 & 1.264 & 1.128 & 1.128 & 1.000 & 1\\
+Middle→Low & 0.004 & -0.056; 0.063 & 0.392 & 1.008 & 1.008 & 1.101 & 0.879 & 1.138 & 1.000 & 1\\
+Middle→High & 0.088 & 0.020; 0.156 & 0.403 & 1.219 & 1.219 & 1.736 & 1.046 & 1.046 & 1.264 & 0\\
+High→Middle & -0.074 & -0.148;-0.001 & 0.455 & 0.862 & 1.160 & 1.590 & 0.999 & 1.001 & 1.037 & 0\\
+\addlinespace
+High→Low & -0.098 & -0.177;-0.018 & 0.431 & 0.814 & 1.229 & 1.759 & 0.963 & 1.039 & 1.239 & 0\\
+\bottomrule
+\end{tabu}
+
+}
+
+\end{table}%
+
 \end{document}

### 2025-11-11 paper/03-methods.qmd
up v2

@@
 
+library(summarytools)
 t1 <- db %>% 

### 2025-11-11 paper/04-analysis.qmd
up v2

@@
 #| results: asis
-g1
+g1 +
+  theme(legend.text = element_text(size = 10))
 ```
@@
 #| out-width: '90%'
+#| fig-asp: 0.7
 #| fig-cap-location: bottom
@@
 
-g2
+g2 +
+  theme(legend.text = element_text(size = 10))
 
@@
 #| out-width: '90%'
+#| fig-asp: 0.8
 #| fig-cap-location: bottom
@@
 
-g4
+g4+
+  theme(legend.text = element_text(size = 10))
 ```

### 2025-11-11 processing/estrati2.R

@@ -1,0 +1,787 @@
+# 0. Identification ---------------------------------------------------
+
+# Title: Data preparation for thesis project Social Mobilty and Market Justice Preferencees
+# Responsible: Andreas Laffert
+
+# Executive Summary: This script contains the code to data preparation for thesis project
+# Date: May 1, 2025
+
+# 1. Packages  -----------------------------------------------------
+
+if (!require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse,
+               here,
+               car,
+               sjmisc,
+               sjlabelled,
+               summarytools,
+               nnet,
+               occupar,
+               cobalt,
+               WeightIt)
+
+options(scipen=999)
+rm(list = ls())
+
+# 2. Data -----------------------------------------------------------------
+
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
+load(here("input/data/elsoc_ciuo08.RData"))
+
+glimpse(elsoc_long_2016_2023)
+
+# 3. Processing -----------------------------------------------------------
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(merit_effort = c18_09,
+         merit_talent = c18_10,
+         perc_inequality2 = c18_11,
+         perc_sal_gerente = d03_01,
+         perc_sal_obrero = d03_02) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+
+# perc inequality gap 
+
+frq(elsoc_long_2016_2023$perc_sal_gerente)
+frq(elsoc_long_2016_2023$perc_sal_obrero)
+
[…]

### 2025-11-11 processing/prod_analysis_multinom.R

@@ -34,13 +34,26 @@ glimpse(db)
 
 # 3. Analysis -----------------------------------------------------------
 
+my_pretty_theme <- function(border_width = 0.4) {
+  theme_ggdist() %+replace%
+    theme(
+      panel.border = element_blank(),
+      axis.line.x  = element_line(colour = "black", linewidth = border_width),
+      axis.line.y  = element_line(colour = "black", linewidth = border_width),
+      axis.ticks = element_line(colour = "black", linewidth = border_width),
+      axis.text = element_text(colour = "black"),
+      #legend.text = element_text(size = 12),
+      #plot.caption = element_text(size = 11, hjust = 1)
+      )
+}
+
+
 # 3.1 Descriptive statistics
 
 # Table summary
 
 t1 <- db %>% 
-  select(idencuesta, 
-         wave, 
+  select(wave, 
          just_pension_f, 
          estrato_orig, 
          estrato_ocupa, 
@@ -86,7 +99,7 @@ g1 <- db %>%
   geom_col(aes(fill = just_pension_f)) +
   scale_y_continuous(labels = scales::percent) + 
   scale_fill_manual(
-    values = c("#5b59d4", "#c8237d"),
+    values = c("grey50", "grey20"),
     breaks = c("Agree","Disagree")
   )  +
   coord_flip() +
@@ -101,8 +114,8 @@ g1 <- db %>%
        x = NULL,
        fill = NULL,
        caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)") +
-  theme_ggdist() +
-  theme(legend.position = "top") 
+  my_pretty_theme() +
+  theme(legend.position = "bottom") 
 
 
 # MJP per mobility
@@ -112,6 +125,16 @@ pdat <- db %>%
   drop_na() %>% 
   mutate(
     mobility = paste(estrato_orig, estrato_ocupa, sep = "-"),
+    mobility = factor(mobility, 
+                      levels = c("High-Middle",
+                                 "High-Low",
+                                 "Middle-Low",
+                                 "Low-Low",
+                                 "Middle-Middle",
+                                 "High-High",
+                                 "Low-Middle",
+                                 "Low-High",
[…]

### 2025-11-12 presentations/seminario/mjp-mobility-seminario.qmd
Up presentacion seminario

@@
 
-```{r}
-#| label: setup
-#| include: false
-library(knitr)
-knitr::opts_chunk$set(echo = F,
-                      warning = F,
-                      error = F, 
-                      message = F) 
+```{r setup2, include=FALSE}
+ knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
+                       # , fig.align = 'center'
+                       )
+ # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
+ # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
+ options(scipen=999)
+ options(kableExtra.auto_format = FALSE)
+ options(knitr.kable.NA = '')
+ options(knitr.graphics.error = FALSE)
+ Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
+
+ table_format = if (knitr::is_html_output()) {
+   #conditional instructions for kable
+   "html" #if html set "html" in format
+ } else if (knitr::is_latex_output()) {
+   "latex"#if latex set "latex" in format
+ }
+
+ library(conflicted)
+
+conflicts_prefer(dplyr::filter)
 ```
@@
 
-pacman::p_load(tidyverse, 
-               sjmisc, 
-               sjPlot, 
-               sjlabelled, 
-               here,
-               kableExtra,
-               ggdist,
-               summarytools)
+pacman::p_load(here,
+               kableExtra)
 
@@
 rm(list = ls())
+
 ```
@@
 
-source(file = here("processing/prod_analysis.R"))
+source(file = here("processing/prod_analysis_multinom.R"))
 
@@
 
-8 Octubre 2025, Santiago
+19 Noviembre 2025, Santiago
 
@@
 # Contexto y motivación {.xlarge data-background-color="#523870"}
-
-## Contexto y motivación
-
-::: {.fragment}
-<div style="text-align:center;">
-  <img src="images/corte.png" style="width:100%; max-width:none;">
-</div>
-:::
 
@@
 
-Preferencias por justicia de mercado [@busemeyer_skills_2014; @castillo_perceptions_2025; @koos_moral_2019; @lindh_public_2015]
+Preferencias por la comodificación de servicios [@busemeyer_skills_2014; @castillo_perceptions_2025; @koos_moral_2019; @lindh_public_2015]
 
@@
 
-## Preferencias por justicia de mercado
+## El caso de las pensiones
 
-::: {.incremental .highlight-last style="font-size: 120%;"}
+:: {.incremental .highlight-last style="font-size: 120%;"}
 
-- Lane [-@lane_market_1986]: justicia de mercado vs. justicia política
+- Ciclos de privatización de sistemas de pensiones en AL: ampliación de cuentas individuales de contribución definida, obligatorias o voluntarias, administradas por empresas privadas [@ferre_welfare_2023; @huber_political_2000; @oecd_pensions_2023].
+Chile destaca [@solimano_rise_2021]
 
-- Creencias normativas que legitiman la idea de que el acceso a los servicios sociales esenciales —como la salud, la educación o las pensiones— debe determinarse según criterios basados en el mercado [@lindh_public_2015, p.895]
+- Investigaciones sobre pensiones justas [@sabbagh_perceived_2014; @castillo_deserving_2019] y preferencias por alternativas de reforma público vs.
+privado [@jaime-castillo_public_2013; @quadagno_has_2012]
 
-- Medición: evaluar si las personas consideran justo que el acceso a dichos servicios dependa de los ingresos [@lindh_public_2015; @kluegel_legitimation_1999; @castillo_perceptions_2025]
+- No hay estudio sistemático sobre las preferencias por comodificación de pensiones y de los motivos que llevan a abrazar estos principios, menos en AL [@mau_inequality_2015; @kerner_pension_2020]
 
-:::
+- Refleja los principios de justicia que las personas consideran legítimos para organizar un ámbito central del bienestar que distribuye riesgos y define la pobreza en la vejez [@borzutzky_pension_2012; @madero-cabib_private_2019]
 
-::: {.box-inv-5 .sp-after .fragment style="font-size: 120%;"}
-
-Preferencias por la comodificación de servicios
-
-:::
-
-## Contexto chileno
-
-::: {.incremental .highlight-last style="font-size: 120%;"}
-
-- Crecimiento con elevada desigualdad [@llorca-jana_historia_2021; @flores_top_2020]
-- Rigida movilidad relativa y fuerte clausura cúspide [@lopez-roldan_comparative_2021; @torche_intergenerational_2014]
-- Profunda privatización y comodificación de áreas de reproducción social con fuerte dependencia Estatal [@madariaga_three_2020; @boccardo_30_2020]
-- Pensiones: sistema de **capitalización individual** administrado por **AFPs**; cubre ~11 millones de cotizantes, pero **27% de la fuerza laboral** queda fuera por **informalidad** [@superintendenciadepensiones_estadisticas_2024]
-- Conflictividad social previsional [@somma_no_2021; @nucleodesociologiacontingente_informe_2020] y legitimidad del sistema ("con mi plata no") [@castillo_perceptions_2025; @canalesceron_sujeto_2021; @panes_criticas_2020]
 :::
@@
 
-## Preferencias por la comodificación de servicios: Chile
+## Contexto chileno
 
-::: {.incremental .highlight-last style="font-size: 110%;"}
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
-1. @otero_power_2024 proveen evidencia sobre:
-
-   * Personas en posiciones sociales bajas/subordinadas muestran menor apoyo a la justicia de mercado que aquellas en posiciones altas/privilegiadas.
-
-2. @castillo_perceptions_2025 encuentran que:
-
-   * La percepción de la desigualdad económica reduce el apoyo a la justicia de mercado.
-   * Las percepciones meritocráticas (esfuerzo) aumenta el respaldo a la justicia de mercado.
-:::
-
-::: {.fragment style="font-size: 120%;"}
-**Sin embargo...**
+- Crecimiento con elevada desigualdad [@llorca-jana_historia_2021; @flores_top_2020]
+- Rigida movilidad relativa y fuerte clausura cúspide [@lopez-roldan_comparative_2021; @torche_intergenerational_2014]
+- Profunda privatización y comodificación de áreas de reproducción social con fuerte dependencia Estatal [@madariaga_three_2020; @boccardo_30_2020]
+- Pensiones: sistema de **capitalización individual** administrado por **AFPs**; cubre ~11 millones de cotizantes, pero **27% de la fuerza laboral** queda fuera por **informalidad** [@superintendenciadepensiones_estadisticas_2024]
+- Conflictividad social previsional [@somma_no_2021] y legitimidad del sistema ("con mi plata no") [@castillo_perceptions_2025; @canalesceron_sujeto_2021]
 :::
@@
 
-El movimiento entre posiciones sociales expone a los individuos a condiciones y experiencias que afectan sus nociones de justicia distributiva [@alesina_intergenerational_2018; @ares_changing_2020; @gugushvili_subjective_2017; @day_movin_2017]
+El movimiento entre posiciones sociales expone a los individuos a condiciones y experiencias que afectan sus nociones de justicia distributiva [@alesina_intergenerational_2018; @ares_changing_2020; @gugushvili_subjective_2017; @langsaether_explaining_2022]
 
@@
 ¿Cómo?
-Movilidad ascendente refuerza ver al mercado como asignador justo, la descendente lo cuestiona.
-La meritocracia es un canal cognitivo para ese proceso [@mau_inequality_2015; @gugushvili_intergenerational_2016].
+Vía exposición diferenciada a condiciones materiales, marcos normativos y señales/información sobre cómo operan mercado y Estado [@helgason_class_2025; @mau_inequality_2015; @schmidt_experience_2011].
 
@@
 
-**_(2) ¿En qué medida las creencias meritocráticas condicionan esta relación?_**
+**_(2) ¿En qué medida las creencias meritocráticas y el tiempo condicionan esta relación?_**
 
@@
 
-    * self-interest (ej. intereses materiales, POUM)
-    * socialización (ej. aculturación, socialización valores, máximización estatus)
-    * **atribuciones** (ej. self-serving bias)
+    * Exposición y aprendizaje (ej. intereses materiales, POUM)
+    * Socialización y redes (ej. aculturación, socialización valores, máximización estatus)
+    * Atribuciones causales (ej. self-serving bias)
 
@@
 
-## El mecanismo del *self-serving bias*
+## ¿Por qué la movilidad tendría un efecto?
 
-::: {.incremental style="font-size:120%;"}
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
-- Las personas ajustan sus **creencias sobre justicia** según cómo explican su posición social [@gugushvili_intergenerational_2016; @miller_selfserving_1975].  
-- Las experiencias de movilidad activan **procesos de atribución causal**:  
-  - **Interna** → éxito por esfuerzo → el mercado es justo.  
-  - **Externa** → fracaso por factores estructurales → el mercado es injusto.  
-- **Ascenso:** refuerza la creencia de que el mercado premia el mérito.  
-- **Descenso:** puede generar disonancia y cuestionar esa justicia.
+- **No solo cambia intereses materiales**, también cambia **las experiencias, redes, y marcos de referencia normativos**  [@gugushvili_intergenerational_2016; @helgason_class_2025]
 
-:::
+- @ares_changing_2020 y @helgason_class_2025 muestran que, al moverse, los individuos **recalibran sus interpretaciones sobre la desigualdad y lo que es justo en cuanto a riesgos** en base a experiencias de clase acumuladas
 
-## ¿Cómo?
-::: {.incremental .highlight-last style="font-size: 110%;"}
-- **No solo cambia intereses materiales**, también cambia **cómo las personas entienden la justicia** [@gugushvili_intergenerational_2016; @helgason_class_2025].
-- Al moverse, los individuos **recalibran sus creencias sobre la meritocracia** [@mijs_belief_2022a].
 - **Movilidad ascendente →** refuerza la creencia en un mercado justo (asignador legítimo de recompensas).
+
 - **Movilidad descendente →** desafía esa creencia (percepción de arbitrariedad o injusticia).
-- La **meritocracia** filtra estas experiencias [@mau_inequality_2015]:  
-  - *Fuerte creencia en mérito →* mantiene la fe en la justicia del mercado (disonancia resuelta internamente).  
-  - *Débil creencia en mérito →* percibe el sistema como injusto (disonancia resuelta externamente).  
+
 :::
@@
 
-- **H2a.** El efecto positivo de la movilidad ascendente sobre el apoyo a la mercantilización es más fuerte entre las personas con creencias meritocráticas más sólidas.  
-→ *Self-serving bias*: el ascenso confirma que el sistema recompensa el esfuerzo.
+- **H2** : El efecto de la movilidad es moderado por las creencias meritocráticas.
 
-- **H2b.** El efecto negativo de la movilidad descendente sobre el apoyo a la mercantilización es más débil entre las personas con creencias meritocráticas más fuertes.  
-→ *Resolución de la disonancia*: los meritócratas internalizan la pérdida («mi culpa»), mientras que los no meritócratas la atribuyen al sistema («es injusto»).
+    * Entre quienes sostienen fuertes creencias meritocráticas, los efectos de la movilidad ascendente serán más pro-mercado y los de la descendente más débiles (porque atribuyen la posición a esfuerzo o mérito).
+
+- **H3** La fuerza del efecto de la movilidad varía con el tiempo.
+
+    * A lo largo del período 2016–2023, los efectos pueden intensificarse o atenuarse según los cambios contextuales, lo que indicaría procesos de aprendizaje o realineamiento normativo.
 :::
@@
 
-## Enfoque: Mobility Effects
+(M)odel, (I)nquiry, (D)ata, (A)nswer estrategy [@blair_research_2023]
+
+## Un modelo causal para los efectos de la movilidad (M)
 
@@
 #| fig-cap-location: top
-#| fig-cap: "DAG movilidad social, meritocracia y preferencias por comodificación"
+#| fig-cap: "DAG movilidad social y preferencias por comodificación"
 #| fig-align: center
@@
 
-# Coordenadas para ubicar los nodos según el tiempo t1–t4
-coords <- list(
-  x = c(C=-1,   # Confusores pre
-        T= 0,   # Tratamiento
-        Y= 2),  # Outcome
-  y = c(C= 0.6,
-        T= 0.9,
-        Y= 0.9)
+dag_mjp <- dagify(
+  Y ~ D + C,
+  D ~ O + C,
+  O ~ C,
+  C ~ U,
+  coords = time_ordered_coords(
+    list(
+      "U",
+      c("C", "O"),
+      "D",
+      "Y"
+    )
+  ),
+  exposure = "D",
+  outcome = "Y",
+  labels = c(
+    Y = "Y",
+    D = "D",
+    O = "O",
+    C = "C",
+    U = "U"
+  )
 )
 
-# Definición del DAG
-dag_mjp <- dagify(
-  T ~ C,          # Tratamiento depende de confusores pre
-  Y ~ T  + C ,  # Outcome depende de todo lo anterior
-  exposure = "T",
-  outcome  = "Y",
-  coords   = coords
-)
+curvatures <- rep(0, 8)
+curvatures[3] <- -.25
 
-# Pasar a formato tidy
-td <- tidy_dagitty(dag_mjp)
-
-# Asignar tiempos por nodo
-time_map <- c(
-  C = "t1",
-  T = "t2",,
-  Y = "t3"
-)
-td$data$time <- unname(time_map[td$data$name])
-
-# Graficar DAG
-ggdag(td, node_size = 18, text_size = 6, edge_size = 1.2) +   # <--- aumenta nodos, texto de nodos y grosor de flechas
-  theme_classic() + 
-  remove_axes() + 
-  theme(axis.line = element_blank(),
-        text = element_text(size = 16)) +                     # <--- tamaño global del texto adicional
-  xlim(-1.5, 2.5) + ylim(-1, 1) +
-
-  # Etiquetas extra
-  geom_text(x=-1, y=-0.45, label="T = Movilidad social\n(upward / downward)", hjust=0, size=5) +
-  geom_text(x=-1, y=-0.70, label="Y = Preferencias por\ncomodificación", hjust=0, size=5) +
-  geom_text(x=-1, y=-0.96, label="M = Meritocracia (post-T)\n— heterogeneidad descriptiva", hjust=0, size=5) +
-  geom_text(x=0.55, y=-0.7, label="C = Confusores pre-tratamiento\n(propensity)", hjust=0, size=5)  +
-  geom_text(x=0.55, y=-0.9, label="L = Confusor post-T", hjust=0, size=5)
+dag_mjp |>
+  tidy_dagitty() |>
+  ggplot(aes(x, y, xend = xend, yend = yend)) +
+  geom_dag_point(shape = 21, fill = "black", size = 14, stroke = 0.8) +
+  geom_dag_edges_arc(curvature = curvatures) +
+  geom_dag_text(aes(label = label), color = "white", fontface = "bold", size = 6) +
+  theme_dag()
 ```
 
-::: {.fragment style="font-size: 60%;"}
-- M es post-tratamiento; lo uso para caracterizar heterogeneidad del efecto, sin atribuir causalidad a 
-T×M. El efecto principal (ATT) no depende de M.
+::: {style="font-size: 60%;"}
+- Y = preferences for pension commodification, D = an individual’s class of destination, O = an individual’s class of origin, C = different attributes determined in childhood or earlier that affects an individual’s class of origin and destination.
+Finally, U = unobserved factors influencing origin attributes.
 ::::
 
-## Datos
+## Pregunta y estimand (I)
+
+::: {.incremental style="font-size: 120%;"}
+
+- *¿Cómo la movilidad social intergeneracional afecta las preferencias por mercantilización de las pensiones en Chile?*
+
+- @breen_effects_2024: Efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran mantenido en su origen *j* (@eq-att).
+:::
+
+::: {.fragment}
+$$ 
+ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
+$$ {#eq-att}
+
+:::
+
+## Datos y variables (D)
 
@@
 
-- Encuesta Longitudinal Social de Chile [(ELSOC)](https://coes.cl/elsoc/) de COES.
+- [(ELSOC)](https://coes.cl/elsoc/) (COES): panel urbano nacional 2016–2023; muestreo probabilístico estratificado, por conglomerados y multietápico; población objetivo 18–75 años residentes habituales
 
-- Muestra analítica: olas 2016 (N = 2.927), 2018 (N = 3.748) y 2023 (N = 2.726)
+- Analizo 2016, 2018 y 2023 por disponibilidad ocupacional; ocupación parental (solo 2023) codificada a vía API del INE y tratada como atributo de origen invariante en el tiempo
 
-- Encuesta panel representativa de zonas urbanas, localizadas en 40 ciudades del país (92 comunas y 13 regiones)
+- Muestra analítica: *N*=3.435 observaciones anidadas en *N*=1.787 individuos (2016: 914; 2018: 1.377; 2023: 1.144)
 
-- Diseño muestral complejo: probabilístico, por conglomerados, multietápico y estratificado según el tamaño de las ciudades.
-
-- Población objetivo incluye a mujeres y hombres de entre 18 y 75 años que residen habitualmente en viviendas privadas
+- Submuestras por trayectorias de movilidad (p. ej., baja→baja, baja→media), reteniendo movers y no-movers apareados desde la misma clase de origen
 
@@
 
-g1
+g1 +
+  labs(title = "It is fair that people with higher incomes have better pensions than people\nwith lower incomes?") +
+  theme(legend.position = "top",
+        plot.title = element_text(face = "bold"),
+        text = element_text(size = 16))
 
@@
 ## Tratamiento
+
 ::: {style="font-size: 120%;"}
@@
 
-2. **ATT por entropy balancing (no PS):**  
+2. **ATT por entropy balancing:**  
    - En cada estrato \(O=j\), uso `ebal` para **reponderar solo a los controles** (inmóviles) e igualar sus valores de los confusores pre-tratamiento \(C\) a los de los tratados (móviles).  
@@
 ```{r}
-#| label: tbl-ocup
+#| label: tbl-matrix
+#| tbl-cap: "Movilidad ocupacional por grupos ocupacionales"
 #| tbl-cap-location: top
-#| tbl-cap: "Movilidad ocupacional por grupos ocupacionales"
+#| results: asis
+#| echo: false
+library(janitor)
+db %>%
+  tabyl(estrato_orig, estrato_ocupa) %>%
+  adorn_totals(where = c("row", "col")) %>%     # agrega totales
+  adorn_percentages("row") %>%                  # porcentajes por fila
+  adorn_pct_formatting(digits = 1) %>%          # formato %
+  adorn_ns() %>% 
+  as_tibble() %>% 
+  mutate(Offspring = NA) %>% 
+  select(estrato_orig, Offspring, everything()) %>% 
+  kableExtra::kable(format = "html", 
+                    booktabs= T,
+                    align = 'c', 
+                    col.names = c("Father↓", "Offspring→", "Low", "Middle", "High", "Total"),
+                    caption = NULL) %>%
+  kableExtra::kable_styling(latex_options = "hold_position", 
+                            bootstrap_options = "striped",
+                            full_width = F)
 
-library(sjPlot)
-
-sjPlot::sjt.xtab(db$estrato_orig, 
-                 db$estrato_ocupa, 
-                 show.row.prc = T,
-                 show.col.prc = T,
-                 statistics = NULL)
 ```
-
-- _N_ (id único) = 4.447. Porcentajes en azul corresponde a las filas y en verde a las columnas.
 
@@
 
-- **Meritocracia**: se mide a partir de la percepción de recompensa del esfuerzo [@young_rise_1958]
+- Meritocracia [@young_rise_1958]: dos ítems (“recompensa al esfuerzo” y “a la inteligencia/habilidades”), Likert 1–5; se promedian en un índice y se dicotomiza: baja ≤3 vs. alta ≥4.
 
-- Medición:
-
-    * Ítem: “En Chile, se recompensa a las personas por su esfuerzo”
-    * Escala Likert de 1 (totalmente en desacuerdo) a 5 (totalmente de acuerdo). 
-    * Fue dicotomizado de modo tal que baja meritocracia (≤ 3) y alta meritocracia (≥ 4) 
-:::
-
-## Estimand
-
-::: {.fragment style="font-size: 120%;"}
-
-- @breen_effects_2024: Efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran trasladado a un destino alternativo *k′* (@eq-breen).
-:::
-
-::: {.fragment}
-$$ 
-ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
-$$ {#eq-breen}
+- Tiempo: ola de encuesta 2016, 2018, 2023, ingresada como variable categórica (dummies).
 
@@
 
-::: {.fragment style="font-size: 120%;"}
+## Controles
+::: {.incremental .highlight-last style="font-size: 120%;"}
 
-- Mi interés es comparar con la situación de inmobilidad: cuando _k'_ = _j_ (@eq-att)
+- Incluyo los mismos pretratamientos $C$ que en IPW no para re-balancear, sino para bloquear backdoors (según el DAG) y 
+- Lograr doble robustez: consistencia si el modelo de pesos o el de resultados está bien especificado [@wooldridge_introductory_2009]
 :::
 
-::: {.fragment}
-$$ 
-ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
-$$ {#eq-att}
-
-:::
-
-## Estimador
-
-::: {style="font-size:110%;"}
+## Estrategia analítica (A)
 
 $$
-Y_i = \alpha + \beta T_i + X_i'\gamma + \lambda_{t(i)} + \varepsilon_i
-\quad \text{(WLS con IPW–ATT, CR2 por id)}
-$$
-:::
+Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
+$$ {#eq-final}
 
@@
 - **Muestra:** individuos con mismo origen de clase $O=j$  
--  **Outcome:** $Y_i$ preferencia por comodificación de pensiones 
+-  **Outcome:** $Y_{it}$ preferencia por comodificación de pensiones 
 - **Tratamiento:** $T_i=1$ si $D_i=k$ (*móvil*), $T_i=0$ si $D_i=j$ (*inmóvil*)  
@@
 
+## Descriptivos
+
+```{r}
+#| label: fig-mean
+#| fig-cap: "Porcentaje de acuerdo sobre las preferencias respecto a la mercantilización de las pensiones según la trayectoria de movilidad y el tiempo (2016, 2018 y 2023)"
+#| out-width: '90%'
+#| fig-asp: 0.7
+#| fig-cap-location: top
+#| echo: false
+#| results: asis
+
+g2+
+  theme(text = element_text(size = 14))
+
+```
+
 ## Mobility effects
 
-::: {.fragment style="font-size: 120%;"}
+::: {.fragment style="font-size: 110%;"}
 
@@
 #| fig-cap-location: top
-#| fig-cap: "Estimaciones de efectos de movilidad en preferencia por comodificación de pensiones"
+#| fig-cap: "Efectos de movilidad en preferencias por la mercantilización de las pensiones"
 #| fig-align: center
 #| out-width: '180%'
+#| fig-asp: 0.55
 
-g2 +
-  theme(text = element_text(size = 16))
+g3 +
+  theme(text = element_text(size = 14))
 ```
@@
 
-::: {.fragment style="font-size: 120%;"}
+::: {.fragment style="font-size: 110%;"}
 
@@
 #| fig-cap-location: top
-#| fig-cap: "Interacción entre movilidad y meritocracia en preferencia por comodificación de pensiones"
+#| fig-cap: "Efectos de movilidad en preferencias por la mercantilización de las pensiones, según  percepción meritocrática"
 #| fig-align: center
-#| out-width: '180%'
+#| out-width: '160%'
+#| fig-asp: 0.55
 
-g3+
-  theme(text = element_text(size = 16))
+g4+
+  theme(text = element_text(size = 14),
+        legend.position = "right")
 ```
@@
 
-::: {.fragment style="font-size: 120%;"}
+::: {.fragment style="font-size: 20%;"}
 
 ```{r}
-#| label: fig-marginal
-#| fig-cap-location: top
-#| fig-cap: "Efecto marginal de movilidad descendente High→Low por meritocracia"
-#| fig-align: center
-#| out-width: '180%'
+#| label: tbl-inttime
+#| tbl-cap: "Efectos de movilidad en preferencias por la mercantilización de las pensiones, por ola"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
 
-b <- coef(mp_hl_i)
-V <- vcov(mp_hl_i)   # ya es CR2 por cómo estimaste el modelo
-
-# 2) Efectos marginales de t por nivel de meritocracia
-est_m0 <- b["t"]                           # mAlta = 0
-se_m0  <- sqrt(V["t","t"])
-
-est_m1 <- b["t"] + b["t:mAlta"]            # mAlta = 1
-se_m1  <- sqrt(V["t","t"] + V["t:mAlta","t:mAlta"] + 2*V["t","t:mAlta"])
-
-# 3) Arma tabla y CIs 95%
-z <- qnorm(0.975)
-df_me <- tibble(
-  meritocracia = c("Baja meritocracia","Alta meritocracia"),
-  estimate = c(est_m0, est_m1),
-  se = c(se_m0, se_m1),
-  ci_lo = estimate - z*se,
-  ci_hi = estimate + z*se
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "t" = "Mobility treatment",
+  "ola2018" = "Wave 2018",
+  "ola2023" = "Wave 2023",
+  "t:ola2018" = "Mobility treatment x Wave 2018",
+  "t:ola2023" = "Mobility treatment x Wave 2023"
   )
 
-# 4) Grafica (efecto de High→Low por meritocracia)
-ggplot(df_me, aes(x = meritocracia, y = estimate)) +
-  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
-  geom_point(size = 4) +
-  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0, size = 1.2) +
-  labs(x = NULL, y = "Efecto de High→Low (β | meritocracia)") +
-  theme_ggdist(base_size = 16)
+texreg::htmlreg(list(mp_lm_t,
+               mp_lh_t,
+               mp_mh_t,               
+               mp_ml_t,
+               mp_hm_t,
+               mp_hl_t),
+          custom.model.names = c(
+            "Low-Middle",
+            "Low-High",
+            "Middle-High",            
+            "Middle-Low",
+            "High-Middle",
+            "High-Low"
+          ),   caption.above = NULL,
+               caption = NULL,
+               custom.coef.map = ccoef,
+               digits = 2,
+               groups = list("Wave (Ref.= 2016)" = 3:4, "Mobility treatment x Wave (Ref.= 2016)" = 5:6),
+               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.5,
+               center = T,float.pos = "H!",
+               custom.gof.rows = list("Controls"=rep("Yes",6)))
 
@@
 
-# Discusión y proyecciones {data-background-color="#523870"}
+## Análisis de sensibilidad
 
-## Discusión y proyecciones
+::: {.incremental .highlight-last style="font-size: 120%;"}
+
+- **E-values** [@ding_sensitivity_2016]: Para anular el ATT puntual se requeriría un confusor con $(RR \approx 1.74)$ en Middle→High ($E_{CI}=1.264$) y ($RR \approx 1.76$) en High→Low ($E_{CI}=1.239$); High→Middle es frágil (($E_{point}=1.591$); ($E_{CI}=1.037$)).
+Conclusión: **robustez moderada** en Middle→High y High→Low; borderline en High→Middle.
+
+- **Corrección de Holm**: Tras ajustar por multiplicidad (seis trayectorias) ningún efecto mantiene a un $p<.05$. Interpretación: se atenúa la significancia, no el signo/patrón.
+:::
+
+# Discusión y conclusiones {data-background-color="#523870"}
+
+## Discusión y conclusiones
 
@@
 
-1. **Movilidad social**: no hay efectos significativos. Sin embargo, existe cierta evidencia en la dirección de $H1$:
+1. **Movilidad social**: Existe evidencia en la dirección de $H1$:
 
-    * Middle→High _incrementa_ la preferencia por comodificación en pensiones. 
-    * High→Low _disminuye_ la preferencia por comodificación en pensiones.
+    * Middle→High _incrementa_ la preferencia por comodificación en pensiones en 9 pp.
+    * High→Middle y High→Low _disminuye_ la preferencia por comodificación en pensiones en 7 y 10 pp.
+    * Hay diferencias entre quienes ascienden a las posiciones más altas y quienes descienden de ellas [@atria_economic_2020; @carranza_what_2024]
+    * No puedo testear mecanismos, pero la movilidad social se asocia con cambios en los intereses, información y sesgos que pueden alterar las nociones de justicia distributiva [@ares_changing_2020; @helgason_class_2025; @gugushvili_subjective_2017]
+:::
 
-2. **Meritocracia**: Entre quienes son escépticos con la meritocracia, descender High→Low implica preferir menos comodificación de pensiones ($H2b$).
+## Discusión y conclusiones
 
-3. **Contexto chileno**: profunda comodificación y privatización de servicios escenciales → trabajador-empresario
+::: {.incremental style="font-size: 120%;"}
+
+2. **Heterogeneidad**: No hay evidencia que apoye heterogeneidad causal, pero:
+
+    * Los efectos de movilidad x meritocracia van en la dirección esperada; ascender y mayor meritocracia se asocia con mayores preferencias [@castillo_perceptions_2025]
+    * Aunque descriptivamente hay cambios en las preferencias por comodificación de pensiones en el tiempo, la movilidad no juega un rol en ello
+
+3.
+**Policy feedback y contexto chileno**: En un escenario de profunda privatización de las pensiones, la experiencia de movilidad genera preferencias distintas respecto a su comodificación (negative feedback effects) [@mau_inequality_2015; @busemeyer_positive_2021]
 
@@
 ## Referencias
-
-Bien, y cómo entra la creencia meritocratica en este diagrama de hipotesis y DAG causal entonces?

### 2025-11-12 processing/prod_analysis_multinom.R

@@ -89,7 +89,8 @@ sjPlot::sjt.xtab(db$estrato_orig,
 # MJP per year
 
 g1 <- db %>% 
-  mutate(just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree"))) %>%
+  mutate(just_pension_f = factor(just_pension_f, levels = c(0,1), 
+                                 labels = c("Disagreement", "Agreement"))) %>%
   group_by(wave, just_pension_f) %>% 
   count() %>% 
   group_by(wave) %>% 
@@ -100,20 +101,20 @@ g1 <- db %>%
   scale_y_continuous(labels = scales::percent) + 
   scale_fill_manual(
     values = c("grey50", "grey20"),
-    breaks = c("Agree","Disagree")
+    breaks = c("Agreement","Disagreement")
   )  +
   coord_flip() +
   geom_text(
     aes(label = paste0(round(porcentaje * 100, 1), "%")),
     position = position_stack(vjust = 0.5),
     color = "white",
-    size = 3,
+    size = 4,
     fontface = "bold"
   ) +
   labs(y = NULL,
        x = NULL,
        fill = NULL,
-       caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)") +
+       caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)\nNote: Agreement ≥4, Disagreement ≤3") +
   my_pretty_theme() +
   theme(legend.position = "bottom") 
 
@@ -143,9 +144,35 @@ pdat <- db %>%
   ) %>% 
   group_by(wave, mobility, facet) %>% 
   summarise(y = mean(just_pension_f), .groups = "drop") %>% 
-  group_by(mobility, facet) %>% 
-  mutate(gm = mean(y)) %>% 
   ungroup()
+  
+pdat2 <- db %>% 
+  select(wave, estrato_orig, estrato_ocupa, just_pension_f) %>% 
+  drop_na() %>% 
+  mutate(
+    mobility = paste(estrato_orig, estrato_ocupa, sep = "-"),
+    mobility = factor(mobility, 
+                      levels = c("High-Middle",
+                                 "High-Low",
+                                 "Middle-Low",
+                                 "Low-Low",
+                                 "Middle-Middle",
+                                 "High-High",
+                                 "Low-Middle",
[…]

### 2025-11-12 presentations/seminario/mjp-mobility-seminario.qmd
Up presentacion

@@
 #| out-width: '90%'
-#| fig-asp: 0.7
+#| fig-asp: 0.6
 #| fig-cap-location: top
@@
 g2+
-  theme(text = element_text(size = 14))
+  theme(text = element_text(size = 14),
+        legend.position = "top")
 

### 2025-11-12 processing/prod_analysis_multinom.R

@@ -144,35 +144,9 @@ pdat <- db %>%
   ) %>% 
   group_by(wave, mobility, facet) %>% 
   summarise(y = mean(just_pension_f), .groups = "drop") %>% 
+  group_by(mobility, facet) %>% 
+  mutate(gm = mean(y)) %>% 
   ungroup()
-  
-pdat2 <- db %>% 
-  select(wave, estrato_orig, estrato_ocupa, just_pension_f) %>% 
-  drop_na() %>% 
-  mutate(
-    mobility = paste(estrato_orig, estrato_ocupa, sep = "-"),
-    mobility = factor(mobility, 
-                      levels = c("High-Middle",
-                                 "High-Low",
-                                 "Middle-Low",
-                                 "Low-Low",
-                                 "Middle-Middle",
-                                 "High-High",
-                                 "Low-Middle",
-                                 "Low-High",
-                                 "Middle-High")),
-    facet = case_when(
-      mobility %in% c("High-Middle","High-Low","Middle-Low") ~ "Downward",
-      estrato_orig == estrato_ocupa ~ "Inmobile",
-      mobility %in% c("Middle-High","Low-Middle","Low-High") ~ "Upward"
-    )
-  ) %>% 
-  group_by(mobility) %>% 
-  mutate(gm = mean(just_pension_f)) %>% 
-  ungroup() %>% 
-  select(wave, mobility, facet, gm)
-
-pdat <- left_join(pdat, pdat2)
 
 pdat <- pdat %>% 
   pivot_longer(cols = c(y, gm),
@@ -181,9 +155,6 @@ pdat <- pdat %>%
   mutate(wave = if_else(name == "gm", "All waves", wave),
          wave = factor(wave, levels = c("2016", "2018", "2023", "All waves")))
 
-
-pdat <- pdat %>% 
-  distinct(.keep_all = T)
 
 # datos SOLO para el punto (uno por categoría)
 #pdat_gm <- pdat %>% distinct(mobility, facet, gm)
@@ -200,7 +171,7 @@ g2 <- ggplot(pdat, aes(x = mobility, y = value, group = wave)) +
     values = c("2016" = "#999999",
                "2018" = "#666666",
                "2023" = "#000000",
-               "All waves" = "darkred")) +
+               "All waves" = "black")) +
   scale_shape_manual(
     name   = "Wave",
     breaks = breaks_w,

### 2025-11-17 paper/03-methods.qmd
Big up methods and results

--- 2025-11-03 paper/03-methods.qmd
@@
 Standard specifications (e.g., SAM, DRM, and mobility-contrast models) struggle to retrieve these inherently counterfactual quantities due to identification constraints and their reliance on between-group contrasts constructed from additive terms and interactions, thereby yielding primarily associational evidence [@breen_effects_2024; @song_there_2025].
-I therefore adopt Breen and Ermisch’s [-@breen_effects_2024] causal formulation, treating destination class as the treatment, conditioning on the class of origin, and estimating heterogeneous destination effects with observational data under explicit identification assumptions.
-To align design and target, I follow the MIDA template [@blair_research_2023]: set out the causal model and assumptions (M), define the inquiry and estimand (I), describe the data and variables (D), and detail the answer strategy (A) used to identify the causal effect of intergenerational social mobility on preferences for the commodification of pensions.
+Following the causal framework of Breen and Ermisch [-@breen_effects_2024], I conceive of the destination class as a treatment, conditioning on the origin class and estimating the heterogeneous effects of the destination with observational data under explicit identification assumptions.
+To align design and target, I follow the MIDA template [@blair_research_2023]: set out the causal model and assumptions (M), define the inquiry and estimand (I), describe the data and variables (D), and detail the answer strategy (A) used to identify the causal effect of intergenerational occupational mobility on preferences for the commodification of pensions.
 
@@
 The core claim is that typical mobility hypotheses are fundamentally within-person counterfactual comparisons rather than between-person contrasts.
-Defining mobility as $M = D - O$ —a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination—places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the result in the social position they currently occupy, but not in the situation of other alternative destinations or immobility.
+Defining mobility as $M = D - O$; a change from origin $(O)$ to destination $(D)$, rather than an additive or interactive combination, places the focus squarely on the Neyman–Rubin problem: for any individual, we observe the result in the social position they currently occupy, but not in the situation of other alternative destinations or immobility.
 
@@
 
-In fact, establishing the causal relationship between objective social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference.
+Establishing the causal relationship between objective social mobility and subjective outcomes, such as preferences, poses several challenges for causal inference.
 Below, I use directed acyclic graphs (DAGs) to illustrate some of these challenges and evaluate possible strategies for identifying the effects of social mobility.
 @fig-dag depicts the causal model guiding identification.
-In this model, a person's preference ($Y$) is directly caused by her class destination ($D$) and can also be affected by different attributes determined in childhood or earlier ($C$), such as family resources, household composition, and ascriptive status (pre-treatment confounders).
-Class origin ($O$) is determined by those origin-attributes ($C$) and, in turn, affects destination ($D$).
-In addition, unobserved factors ($U$) can also influence origin-attributes ($C$).
-There is no direct arrow from $O$ to $Y$, reflecting the assumption that the effect of origin on the outcome operates entirely through background factors $(C)$ and destination $(D)$.
-Accordingly, conditioning on $(O,C)$ could blocks the relevant back-door paths from $D$ to $Y$, specifically $D\leftarrow O\leftarrow C\rightarrow Y$ and, if present, $D \leftarrow O \rightarrow Y$, so $(O,C)$ is a minimal sufficient adjustment set.
+In this model, a respondent’s preference $Y$ is directly affected by her class destination $D$, and indirectly influenced by a set of pre-treatment attributes $C$ rooted in childhood—family resources, household composition, parental education, and other ascriptive characteristics.
+Class origin $O$ is itself shaped by these background factors $C$ and, in turn, affects destination $D$.
+I allow for unobserved determinants $U$ that influence background factors $C$, but I assume that any such unobserved variation operates only through $C$.
+Thus, there are no direct paths $U \rightarrow D$ or $U \rightarrow Y$ beyond those mediated by $C$.
+
+Under this structure, conditioning on $(O, C)$ blocks all relevant backdoor paths from $D$ to $Y$—notably $D \leftarrow O \leftarrow C \rightarrow Y$.
+This renders $(O, C)$ a minimal sufficient adjustment set for identifying the causal effect of mobility on preferences, in line with the framework proposed by @breen_effects_2024.
+Identification therefore relies on the assumption that background factors $C$ adequately summarize pre-treatment characteristics that jointly shape both mobility and attitudes.
 
@@
 
-Formally, the DAG encodes the following relations:
-
-$$
-U \rightarrow C,\quad C \rightarrow O,\quad C \rightarrow D,\quad C \rightarrow Y,\quad O \rightarrow D,\quad D \rightarrow Y
-$$
-
-This structure corresponds to a nonparametric structural causal model in which $D$ is the treatment variable, $Y$ the outcome, and $(O, C)$ the minimal sufficient adjustment set.
-Conditioning on these variables satisfies the back-door criterion for identifying the causal effect of mobility on preferences [@breen_effects_2024].
-
 Why are the assumptions plausible here?
-Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$; inverse probability weights (IPW) obtained via entropy balancing enforces covariate balance within each origin stratum.
-Positivity is assessed via propensity scores and trimming of extreme weights, restricting to common support.
+Conditional independence is targeted by controlling for a plausible sufficient adjustment set $(O,C)$ and by using inverse probability weights obtained via entropy balancing [@hainmueller_entropy_2012], which enforce covariate balance within each origin stratum and mitigate selection on observables.
+Positivity is assessed by inspecting the distribution of the entropy-balancing weights and trimming observations with extreme weights, thus restricting inference to regions of common support.
 SUTVA holds by treating the destination class as a single exposure measured before $(Y)$ and assuming no interference.
@@
 The causal inquiry guiding this study asks: How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile?
-Following the potential outcomes framework proposed by Breen and Ermisch [-@breen_effects_2024, p.473], the estimand of interest is the average treatment effect on the treated (ATT), defined as the mean causal effect of moving from an origin class $(O=j)$ to a destination class $(D=k)$, compared with the counterfactual outcome that those same individuals would have exhibited had they instead reached an alternative destination $(k')$:
+Substantively, I am interested in the effect of mobility relative to immobility within a given class of origin.
+
+Following the potential outcomes framework proposed by Breen and Ermisch [-@breen_effects_2024, p.473], the estimand of interest is an average treatment effect on the treated (ATT) defined within each origin class.
+I focus on the specific case in which the counterfactual destination corresponds to class immobility $(k' = j)$.
+In other words, I ask how the outcome among people from origin *j* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility).
+
+Formally,  for individuals with origin $(O=j)$ who attain destination $(D=k)$, the estimand is:
 
 $$
-ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
-$$ {#eq-breen}
 
-In this study, I focus on the specific case where the counterfactual destination corresponds to class immobility $(k' = j)$, that is, how the outcome among people from origin *j* who entered destination *k* would have bee different if those people had, counterfactually, remained in origin *j* instead (inmobility).
-Accordingly, the estimand becomes:
+ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
 
-$$ 
-ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
 $$ {#eq-att}
 
-which represents the average causal effect of attaining class *(k)*—for instance, moving upward or downward—relative to remaining in class *(j)*, among individuals who actually experienced such mobility.
-This estimand captures the within-origin counterfactual contrast at the heart of mobility effects: how a person’s preference for market-based pensions would differ if, counterfactually, they had stayed in their origin class rather than moved to their observed destination.
+which represents the mean causal effect of moving from origin class $(j)$ to destination class $(k)$, compared to the counterfactual outcome those same individuals would have exhibited had they instead remained in $(j)$.
+Thus, this ATT captures the counterfactual contrast within origin that lies at the heart of mobility effects: how preferences for market-based pensions would differ if, for the same upwardly or downwardly mobile individuals, their observed destination class were replaced by immobility in their origin class.
 
@@
 
-Because respondent occupation is not measured in every wave, I restrict the analysis to the 2016, 2018, and 2023 waves.
-Parental occupation was collected only in 2023 as open-text labels.
-I coded these texts to the ISCO-08 two-digit scheme using the National Institute of Statistics of Chile’s automated coding API and retained cases with high-confidence matches (>95%).
-Treating parental occupation as a time-invariant origin attribute, I carried these codes back to 2016 and 2018.
+Because respondent occupation[^1] is not measured in every wave, I restrict the analysis to the 2016, 2018, and 2023 waves.
 After listwise deletion and restricting to key variables (respondent occupation and the outcome measure), the final analytic sample comprises *N* = 3,435 observations nested within *N* = 1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144).
@@
 
+[^1]: Parental occupation was collected only in 2023 as open-text labels.
+I coded these texts to the ISCO-08 two-digit scheme using the National Institute of Statistics of Chile’s automated coding API and retained cases with high-confidence matches (>95%).
+Treating parental occupation as a time-invariant origin attribute, I carried these codes back to 2016 and 2018.
+
 ### Outcome variable
@@
 The outcome variable measures preferences regarding the commodification of pensions, operationalized with a single item addressing how strongly individuals justify conditioning access to old-age pension benefits based on individual income.
-Respondents were asked: “Is it fair in Chile that people with higher incomes have better pensions than people with lower incomes?” and were asked to indicate their level of agreement on a five-point Likert scale ranging from 1 (“strongly disagree”) to 5 (“strongly agree”).
-For estimation and interpretability, I dichotomized the item so that values 4-5 indicate agreement with market-based access to old-age pensions (coded as 1), while values 1-3 indicate disagreement (coded as 0).
-This binary coding treats neutrality as substantially closer to non-support than to support and produces a clear contrast between respondents who support market-based differentiation in pensions and those who do not.
-I use this item for two main reasons: on the one hand, to be able to make comparisons with existing literature [@lindh_public_2015; @otero_power_2024; @castillo_perceptions_2025] and, on the other hand, because it captures two fundamental dimensions of market-based welfare distribution: the role of economic resources as a key determinant of outcomes and the framing of services as as tradable commodities that can be bought and sold based on ability to pay [@lindh_public_2015].
+Respondents were asked: “Is it fair in Chile that people with higher incomes have better pensions than people with lower incomes?” and answered on a five-point Likert scale ranging from 1 (“strongly disagree”) to 5 (“strongly agree”).
+For estimation and interpretability, I construct a binary indicator in which values 4–5 indicate agreement with market-based access to old-age pensions (coded 1), while values 1–3 indicate non-agreement (coded 0).
+Substantively, the dependent variable is meant to capture *endorsement* of pension commodification: only responses in the upper tail of the scale represent an active, explicit justification of market-based differentiation, whereas neutrality corresponds to an absence of such endorsement and is therefore treated as closer to non-support than to support.[^2] This binary coding thus yields a clear contrast between respondents who *support* market-based differentiation in pensions and those who do not.
+I use this item for two main reasons: first, to enable comparisons with existing work on market-based justice in social policy [@lindh_public_2015; @otero_power_2024; @castillo_perceptions_2025]; and second, because it taps two core dimensions of market-based welfare distribution—(i) the centrality of economic resources as a criterion for allocating outcomes and (ii) the framing of pensions as tradable commodities that can be bought and sold according to ability to pay [@lindh_public_2015].
 
-@tbl-summary1 shows the descriptive statistics of the variables used in the study.
-
-```{r}
-#| label: tbl-summary1
-#| tbl-cap: "Descriptive statistics for all variables."
-#| tbl-cap-location: top
-#| results: asis
-#| echo: false
-
-t1 <- db %>% 
-  select(just_pension_f, 
-         estrato_orig, 
-         estrato_ocupa, 
-         educ_orig_f,
-         hogar_bip_f, 
-         nacionalidad_f, 
-         sexo_f,
-         m0_edad_f, 
-         etnia_f, 
-         merit_i,
-         wave
-
-)
-
-t1 <- t1 %>% 
-  mutate(
-    just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree")),
-    hogar_bip_f  = factor(hogar_bip_f, levels = c(0,1), labels = c("No co-residence", "Co-residence")),
-    #nacionalidad_f = if_else(nacionalidad_f == 0, 1, 0),
-    nacionalidad_f  = factor(nacionalidad_f, levels = c(0,1), labels = c("Non-Chilean", "Chilean")),
-    #sexo_f = if_else(sexo_f == 1, 0, 1),
-    sexo_f  = factor(sexo_f, levels = c(0,1), labels = c("Male", "Female")),
-    etnia_f = factor(etnia_f, levels = c(0,1), labels = c("Non-indigenous", "Indigenous"))
-  )
-
-t1$just_pension_f <- sjlabelled::set_label(t1$just_pension_f, "Preference for pension commodification")
-t1$educ_orig_f <- sjlabelled::set_label(t1$educ_orig_f, "Parental education")
-t1$hogar_bip_f <- sjlabelled::set_label(t1$hogar_bip_f, "Co-residence with both parents at age 15")
-t1$nacionalidad_f <- sjlabelled::set_label(t1$nacionalidad_f, "Nacionality")
-t1$sexo_f <- sjlabelled::set_label(t1$sexo_f, "Sex")
-t1$m0_edad_f <- sjlabelled::set_label(t1$m0_edad_f, "Age (in years)")
-t1$etnia_f <- sjlabelled::set_label(t1$etnia_f, "Indigenous ethnicity")
-t1$merit_i <- sjlabelled::set_label(t1$merit_i, "Meritocracy perception")
-t1$wave <- sjlabelled::set_label(t1$wave, "Wave")
-
-df<-dfSummary(t1,
-               plain.ascii = FALSE,
-               style = "multiline",
-               tmp.img.dir = "/tmp",
-               graph.magnif = 0.75,
-               headings = F,  # encabezado
-               varnumbers = F, # num variable
-               labels.col = T, # etiquetas
-               na.col = F,    # missing
-               graph.col = F, # plot
-               valid.col = T, # n valido
-               col.widths = c(30,10,10,10))
-
-df$Variable <- NULL # delete variable column
-
-print(df)
-
-```
+[^2]: A potential concern is that the neutral category may reflect genuine ambivalence rather than disagreement.
+To address this, I conduct robustness checks using alternative codings: (i) excluding neutral respondents from the analysis; (ii) combining the neutral and agreement categories (3–5 vs.
+1–2); and (iii) estimating models with the original 5-point scale using linear specifications.
+The main mobility effects reported below are substantively unchanged across these specifications (see [Supplementary Material](#anexo)).
 
@@
 
-I treat intergenerational occupational mobility as an exposure indicating whether respondents occupy a different socioeconomic status than their fathers, closely following Breen and Ermisch’s [-@breen_effects_2024] causal framework.
-Class assignment proceeds in two steps.
-First, I derive status for both origin (father) and destination (respondent) from two-digit ISCO-08 occupations using the International Socio-Economic Index of Occupational Status (ISEI).
-ISEI is a hierarchical, unidimensional scale that locates occupations by incumbents’ average education and earnings and was explicitly designed to harmonize occupational stratification across other class schemes and socioeconomic status measures [@ganzeboom_internationally_1996], thereby supporting cross-study comparability [@barozet_measurement_2021a].
-Second, I group both origin and destination ISEI scores into terciles (low, middle, high), yielding a three-category schema for origin and destination stratums (see @tbl-matrix).
-Substantively, I use ISEI because occupation-based measures capture structural locations that organize life chances and attitudes more directly than income alone: they bundle access to the means of production, qualification credentials, and the authority/power inherent in labor relations; central dimensions of stratification and inequality [@barone_rise_2022; @wrightUnderstandingClass2015].
+I treat intergenerational occupational mobility as an exposure indicating whether respondents occupy a different occupational status than their fathers, closely following Breen and Ermisch’s [-@breen_effects_2024] causal framework.
+Occupational assignment proceeds in two steps.
+First, I derive occupational status for both origin (father) and destination (respondent) from two-digit ISCO-08 codes using the International Socio-Economic Index of Occupational Status (ISEI).
+Second, I group these ISEI scores into terciles (low, middle, high), yielding a three-category schema for both origin and destination strata (see @tbl-matrix).
+Substantively, I use ISEI because occupational status is a core indicator of stratification and a reliable mobility measure: it locates jobs on a hierarchical continuum defined by incumbents’ typical education level and earnings [@hauser_intergenerational_2010; @salgado_uplifting_2025], and thus approximates long-run socio-economic position more closely than volatile, single-year income, which is also prone to recall and reporting error for both parental and own resources [@barone_rise_2022].
+This occupational focus is consistent with a long tradition that interprets the occupational structure as the backbone of the stratification system and a key determinant of life chances [@wrightUnderstandingClass2015].
+Moreover, ISEI was explicitly designed to harmonize occupational stratification across class schemes and SES measures [@ganzeboom_internationally_1996], supporting cross-study comparability.
 In short, using ISEI aligns the measurement of origin and destination on a common vertical status continuum that is theoretically meaningful for social mobility research and empirically feasible given the available information of ELSOC’s data.
@@
 
-Within each origin stratum $O=j$, I estimate average treatment effects on the treated (ATT) by comparing movers to otherwise similar non-movers from the same origin.
-To adjust for selection into mobility, I reweight only the control group (immobile) using entropy balancing so that the distribution of pre-treatment origin covariates $C$ matches that of the treated on selected moments.
+Within each origin stratum ($O = j$), I estimate average treatment effects on the treated (ATT) by comparing movers to otherwise similar non-movers from the same origin.
+To reduce bias from non-random selection into mobility based on observed characteristics, I preprocess the data using entropy balancing, reweighting only the control group (the immobile) so that the distribution of pre-treatment origin covariates $C$ among non-movers matches that of movers on selected moments.
+Entropy balancing implements a maximum-entropy reweighting scheme that chooses unit weights for non-movers to satisfy a set of balance constraints (e.g., equality of means and, where relevant, higher moments), while keeping the new weights as close as possible to the original weights [@hainmueller_entropy_2012].
+Substantively, this procedure addresses selection on observables: it reweights immobile respondents so that, within each origin class, they resemble mobile respondents in their observed background characteristics, making any remaining differences in the outcome more plausibly attributable to mobility rather than to pre-existing measured advantages.
 
 The selection of the adjustment set $C$ follows two principles.
-First, it mirrors the logic in Breen and Ermisch [-@breen_effects_2024]—who condition on attributes determined in childhood or earlier (e.g., parental resources, household structure, ascriptive traits)—and, second, it is constrained to origin-side characteristics available in ELSOC.
-Guided by theory and Chilean evidence on relative mobility determinants [mainly @brunori_inequality_2025; @celhay_intergenerational_2010; @espinoza_movilidad_2014;@torche_unequal_2005], I include six covariates covering family resources, household composition, and ascriptive status: (a) parental education (highest of father/mother), 10 categories coded ordinally from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
-These variables are fixed before the destination class, plausibly affect selection into mobility, and constitute the maximal origin set in ELSOC, thereby making the conditional independence assumption more credible within each $O=j$.
+First, it mirrors the logic in Breen and Ermisch [-@breen_effects_2024], who condition on attributes determined in childhood or earlier (e.g., parental resources, household structure, ascriptive traits).
+Second, it is constrained to origin-side characteristics available in ELSOC.
+Guided by theory and Chilean evidence on the determinants of relative mobility [mainly @brunori_inequality_2025; @salgado_uplifting_2025; @espinoza_movilidad_2014; @torche_unequal_2005], I include six covariates covering family resources, household composition, and ascriptive status: (a) parental education (highest of father/mother), in 10 ordinal categories from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
+This set of covariates is fixed prior to the destination class and constitutes the maximum set on the origin side in ELSOC, thus allowing for the plausible capture of both family and contextual influences that may affect career trajectories, making the conditional independence hypothesis more credible within each $O = j$.
+Descriptive statistics for all variables are presented in Supplementary Material Table A1.
 
-After balancing, I construct stabilized IPW, setting treated weights to 1 and assigning derived weights to controls, thus aligning the counterfactual distribution of non-movers with the covariate profile of movers in each origin stratum (see [Supplementary Material](#anexo) for the evaluation of balance).
-These weights are then used in the outcome stage to estimate the causal effect of mobility on preferences for the commodification of pensions.
+After achieving balance, I construct stabilized IPW-ATT weights by setting treated (mobile) cases to weight 1 and assigning the entropy-balancing weights to controls, rescaled within each origin stratum so that the mean weight remains close to one.
+This aligns the counterfactual distribution of non-movers with the covariate profile of movers in each origin class (see [Supplementary Material](#anexo) for balance diagnostics).
+The resulting weights are then used in the outcome stage to estimate the causal effect of intergenerational mobility on preferences for the commodification of pensions.
 
@@
 Because the dependent variable—the preference for pension commodification—is binary $(Y_{it}\in{0,1})$, I estimate weighted linear probability models (WLS) using stabilized inverse probability weights (IPW) obtained via entropy balancing.
-Although OLS estimator is often questioned with binary outcomes, it’s unbiased.
-It consistently estimates the conditional expectation $E(Y|X)$ under standard exogeneity $E[\varepsilon_i|X_i]=0$, with heteroskedasticity addressed by robust or clustered errors [@wooldridge_introductory_2009, ch.
+Although OLS is often questioned with binary outcomes, it consistently estimates the conditional mean $E(Y|X)$ under standard exogeneity $E[\varepsilon_i|X_i]=0$, and heteroskedasticity can be handled with robust or clustered standard errors [@wooldridge_introductory_2009, ch.
 7.5].
-WLS further improves efficiency under heteroskedasticity or unequal sampling probabilities [@wooldridge_introductory_2009, ch.
-8.4].
-Following Gelman, Hill, and Vehtari [-@gelman_regression_2020, pp.
-270–272], IPW-weighted regression transparently targets the ATT by recreating the counterfactual distribution.
-Linear models are preferred here because coefficients are average probability differences; nonlinear models (logit/probit) complicate weighting and interpretation [@gelman_regression_2020, ch.
+WLS with IPW further improves efficiency and implements the ATT estimand by recreating the counterfactual distribution of non-movers [@gelman_regression_2020, pp.
+270–272].
+Linear models are preferred here because coefficients are directly interpretable as average differences in predicted probabilities, whereas non-linear models complicate both weighting and interpretation [@gelman_regression_2020, ch.
 13].
@@
 These results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive changes in the mobility coefficients.
-In addition, I assess causal heterogeneity by interacting the mobility treatment with (i) meritocratic perceptions and (ii) survey wave, reporting simple slopes (origin-specific ATTs) by moderator levels.
-To evaluate sensitivity to unmeasured confounding, I compute E-values for point estimates and confidence limits [@ding_sensitivity_2016].
-To address multiplicity across the six mobility trajectories, I also report Holm-adjusted p-values alongside 95% confidence intervals.
+Heterogeneous effects by meritocratic perceptions and survey wave are examined through interaction models described in the previous section.
 
-All the analyses were conducted using R software and the _lm_robust_ package.
+To evaluate the robustness of the findings, I implement two sets of sensitivity analyses.
+First, I conduct standard robustness checks by re-estimating the models under alternative codings of the outcome: (i) excluding neutral responses from the analysis, (ii) combining neutral responses with agreement (3–5 vs.
+1–2), and (iii) using the original 5-point item in a linear specification.
+Second, I assess sensitivity to unmeasured confounding using the omitted-variable–bias framework of Cinelli and Hazlett [-@cinelli_making_2020], which quantifies how strong an unobserved confounder would need to be to explain away the estimated ATT.

### 2025-11-17 paper/04-analysis.qmd
Big up methods and results

@@
 
-## Models
+## Mobility effects models
 
@@
 @tbl-inttime shows the causal heterogeneity of the mobility effect across the different waves of the survey.
-The treatment effect in 2016 (reference category) closely replicates the reference ATT patterns: positive for Middle→High ($\beta$ = 0.13, 95% CI [0.05, 0.22]) and negative for High→Middle ($\beta$ = −0.19, [−0.33, −0.05]) and High→Low ($\beta$ = −0.19, [−0.32, −0.05]).
-This indicates that, holding all other predictors constant, individuals who moved from middle to high socioeconomic status in 2016 were, on average, 13 percentage points more likely to support market-based access to the pension benefits compared to their counterfactual of immobility, while those who moved down from higher status were approximately 19 percentage points less likely.
-Beyond these reference effects, there is a notable temporal change in the High→Middle trajectory: the interaction between mobility and the 2023 wave is positive and statistically significant ($\beta$ = 0.19, [0.01, 0.38]).
-This suggests that, although people with downward mobility from high backgrounds were less supportive of commodification in 2016, by 2023 the same trajectory shows a change in sign, implying that the negative causal effect of downward mobility attenuated and eventually turned positive.
-No comparable changes are observed in other trajectories.
-In essence, this pattern reveals time-conditional heterogeneity in the mobility effect: among those who experienced downward mobility, preferences for market-based access to old-age pensions appear to have realigned in the most recent wave, reflecting a notable shift toward greater support for commodification within this trajectory by 2023.
+Regarding temporal heterogeneity, the only statistically significant interaction term is observed for the High→Middle trajectory in 2023 ($\beta_{\text{int}} = 0.19$), 95% CI [0.01, 0.38]).
+However, when this interaction is added to the baseline effect in 2016 (-0.19), the marginal effect of High→Middle mobility in 2023 is approximately zero, and its confidence interval includes the null.
+Thus, rather than indicating a reversal toward a positive effect, the evidence suggests that the negative impact of downward mobility from high backgrounds is confined to 2016 and attenuates over time.
+No comparable temporal changes are observed in the remaining trajectories.
+Overall, the results provide little support for strong time-conditional heterogeneity in mobility effects, beyond the fading of the initially negative High→Middle effect.
 
-## Sensitivity analysis
+## Robustness check and sensitivity analysis
 
-The first sensitivity check corresponds to the @ding_sensitivity_2016 approach to assess unmeasured confounding (E values).
-This values indicates the minimum strength of association that an unobserved confounder would need to have with both the treatment and the outcome—on the relative risk scale—to explain an effect.
-The Middle-High ATT would require an $RR ≈ 1.74$ to offset the point estimate ($E_{point}$ = 1.736) and an $RR ≈ 1.26$ to move its CI to include the null value ($E_{CI}$ = 1.264).
-High→Low shows comparable robustness ($E_{point}$ = 1.759; $E_{CI}$ = 1.239), while High→Middle is comparatively fragile near zero ($E_{point}$ = 1.591; $E_{CI}$ = 1.037).
-For non-significant trajectories (Low→Middle, Low→High, Middle→Low), $E_{CI}$ = 1, as expected.
-In conclusion, the directional pattern—stronger in favor of commodification in Middle→High and weaker in the descent from High—seems reasonably robust to moderate unmeasured confounding in Middle→High and High→Low, but borderline in High→Middle.
-These results justify treating signs and relative magnitudes as informative, while recognizing limited protection against omitted variables for High→Middle (see [Supplementary Material](#anexo) for complete tables).
+As a robustness check, I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs.
+1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification.
+Across these alternative models (see [Supplementary Material](#anexo) for complete tables), the substantive pattern of the main effects is preserved: the Middle→High trajectory remains positively associated with support for pension commodification, and the High→Low trajectory remains negatively associated with it, with both coefficients generally increasing in magnitude.
+By contrast, the negative High→Middle effect becomes statistically indistinguishable from zero in the robustness specifications, indicating that this particular estimate is more sensitive to how the outcome is coded than the other two trajectories.
 
-The second sensitivity check applies Holm’s step-down correction, which is a multiple-testing procedure that controls the familywise error rate (FWER) while retaining more power than Bonferroni, across the six mobility trajectories.
-After FWER control, none of the effects remain significant at $\alpha$ = .05: Middle→High ($p_{raw}$ = .0116 → $p_{Holm}$ = .0698), High→Low (.0167 → .0834), and High→Middle (.0481 → .1924).
-Accordingly, claims of statistical significance should be tempered under strict multiplicity control.
-Substantively, however, this adjustment affects the strength of inference, not its direction: the pattern remains consistent with the theorized asymmetry (positive for Middle→High; negative for downward from high).
-I therefore treat signs and effect sizes as the primary evidence and regard p-values as exploratory once multiplicity is accounted for (see [Supplementary Material](#anexo) for complete tables).
+To assess the extent to which the estimated mobility effects may be driven by unmeasured confounding, I implement the sensitivity analysis proposed by Cinelli and Hazlett [-@cinelli_making_2020], which quantifies how strong an unobserved confounder would need to be—in terms of its joint explanatory power for both treatment and outcome—to attenuate or explain away the ATT, using gender as a benchmark covariate (the predictor with the largest partial association with the outcome aside from treatment).
+For the Middle→High trajectory ($\beta \approx 0.09$), partial $R^2 \approx 0.013$, an unobserved confounder would need to explain about 10–11% of the residual variance of both treatment and outcome to reduce the effect to zero and roughly 4% to render it statistically insignificant, substantially more than gender does.
+For the High→Low effect ($\beta \approx -0.10$), partial $R^2 \approx 0.012$, the required strength is of similar magnitude (around 10–11% to explain away the effect and about 4% to remove significance), again exceeding the explanatory power of the main observed covariates.
+By contrast, the High→Middle effect ($\beta \approx -0.07$), partial $R^2 \approx 0.007$ is somewhat less robust: a confounder explaining around 8% of residual variance in both treatment and outcome could nullify the estimate, and one explaining about 2% could make it non-significant.
+Overall, the sensitivity analysis suggests that the positive Middle→High and negative High→Low effects are moderately robust to unmeasured confounding, whereas the High→Middle effect is more vulnerable to relatively modest omitted variables (see [Supplementary Material](#anexo) for complete tables).

### 2025-11-17 paper/07-supplementary-material.qmd
Big up methods and results

@@
 This section presents the supplementary material for this study.
-First, graphs are shown for the evaluation of the balance of IPW-ATT via entropy balance.
-Second, the complete estimated regression models are shown.
-Finally, the sensitivity analyses performed are shown.
 
@@
 ```
+
+## Descriptive statistics
+
+```{r}
+#| label: tbl-summary1
+#| tbl-cap: "Descriptive statistics for all variables."
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+library(summarytools)
+t1 <- db %>% 
+  select(just_pension_f, 
+         estrato_orig, 
+         estrato_ocupa, 
+         educ_orig_f,
+         hogar_bip_f, 
+         nacionalidad_f, 
+         sexo_f,
+         m0_edad_f, 
+         etnia_f, 
+         merit_i,
+         wave
+
+)
+
+t1 <- t1 %>% 
+  mutate(
+    just_pension_f = factor(just_pension_f, levels = c(0,1), labels = c("Disagree", "Agree")),
+    hogar_bip_f  = factor(hogar_bip_f, levels = c(0,1), labels = c("No co-residence", "Co-residence")),
+    #nacionalidad_f = if_else(nacionalidad_f == 0, 1, 0),
+    nacionalidad_f  = factor(nacionalidad_f, levels = c(0,1), labels = c("Non-Chilean", "Chilean")),
+    #sexo_f = if_else(sexo_f == 1, 0, 1),
+    sexo_f  = factor(sexo_f, levels = c(0,1), labels = c("Male", "Female")),
+    etnia_f = factor(etnia_f, levels = c(0,1), labels = c("Non-indigenous", "Indigenous"))
+  )
+
+t1$just_pension_f <- sjlabelled::set_label(t1$just_pension_f, "Preference for pension commodification")
+t1$educ_orig_f <- sjlabelled::set_label(t1$educ_orig_f, "Parental education")
+t1$hogar_bip_f <- sjlabelled::set_label(t1$hogar_bip_f, "Co-residence with both parents at age 15")
+t1$nacionalidad_f <- sjlabelled::set_label(t1$nacionalidad_f, "Nacionality")
+t1$sexo_f <- sjlabelled::set_label(t1$sexo_f, "Sex")
+t1$m0_edad_f <- sjlabelled::set_label(t1$m0_edad_f, "Age (in years)")
+t1$etnia_f <- sjlabelled::set_label(t1$etnia_f, "Indigenous ethnicity")
+t1$merit_i <- sjlabelled::set_label(t1$merit_i, "Meritocracy perception")
+t1$wave <- sjlabelled::set_label(t1$wave, "Wave")
+
+df<-dfSummary(t1,
+               plain.ascii = FALSE,
+               style = "multiline",
+               tmp.img.dir = "/tmp",
+               graph.magnif = 0.75,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = F, # plot
+               valid.col = T, # n valido
+               col.widths = c(30,10,10,10))
+
+df$Variable <- NULL # delete variable column
+
+print(df)
+
+```
+
+## Balance evaluation
 
@@
 ```
+
+## Mobility effects models
 
@@
 
+{{< pagebreak >}}
+
+## Robustness check and sensitivity analysis
+
 ```{r}
-#| label: tbl-holm
-#| tbl-cap: "Holm correction"
+#| label: tbl-robus1
+#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification coded as 3-5 vs. 1-2, with covariates and wave fixed effects."
 #| tbl-cap-location: top
@@
 
-data.frame(traj = c("Low→Middle","Low→High","Middle→Low","Middle→High","High→Middle","High→Low"),
-           p_raw = p, p_holm = p_holm) %>% 
-  mutate_at(.vars = c(2,3), .funs = ~ round(.,3)) %>% 
-  mutate_at(.vars = c(2,3), .funs = ~ format(., nsmall=3)) %>% 
-  mutate_all(~ as.character(.)) %>% 
-  kableExtra::kable(format = table_format, 
-                    col.names = c("Mobility treatment", "p value", "Holm"),
-                    row.names = F,
-                    booktabs = T, align = 'l',
-                    caption = NULL) %>% 
-  kableExtra::kable_styling(latex_options = "hold_position", 
-                            position = "center") %>%
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
-  kableExtra::row_spec(0, bold = T) %>% 
-  kableExtra::add_footnote(label = c("Note: p values derived from Table 1"), notation = "none")
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "t" = "Mobility treatment",
+  edad = "Age (in years)",
+  sexo = "Female (Ref. = Male)",
+  nac = "Chilean nacionality (Ref. = Non-Chilean)",
+  etnia = "Indigenous ethnicity (Ref. = Non-indigenous)",
+  hogar = "Co-residence with both parents (Ref. = No co-residence)",
+  edu0 = "Parental education",
+  ola2018 = "Wave 2018",
+  ola2023 = "Wave 2023"
 
+)
+
+texreg::texreg(list(mp_lm_r_i,
+                       mp_lh_r_i,
+                       mp_mh_r_i,
+                       mp_ml_r_i,
+                       mp_hm_r_i,
+                       mp_hl_r_i),
+                  custom.model.names = c(
+                    "Low-Middle",
+                    "Low-High",
+                    "Middle-High",
+                    "Middle-Low",
+                    "High-Middle",
+                    "High-Low"
+                  ),   caption.above = NULL,
+               caption = NULL,
+               custom.coef.map = ccoef,
+               digits = 2,
+               groups = list("Wave (Ref.= 2016)" = 9:10),
+               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.7,
+               center = T,
+               float.pos = "H!")
 ```
@@
 ```{r}
-#| label: tbl-evalues
-#| tbl-cap: "E-values for ATT estimates by mobility trajectory (Ding & VanderWeele, 2016 approximation)"
+#| label: tbl-robus2
+#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification excluding intermediate category, with covariates and wave fixed effects."
 #| tbl-cap-location: top
@@
 
-evalues_tbl %>% 
-  mutate_at(.vars = c(2:12), .funs = ~ round(.,3)) %>% 
-  mutate_at(.vars = c(2:11), .funs = ~ format(., nsmall=3)) %>% 
-  mutate_all(~ as.character(.)) %>% 
-  tidyr::unite("ci", ci_low:ci_high, sep = ";") %>% 
-  kableExtra::kable(format = table_format, 
-                    col.names = c("Mobility treatment", "Beta", "CI 95% [low-upper]", "sdY", "RR point", "RR point upper", "E point", "RR CI", "RR CI Upper", "E CI", "CI Crosses 0"),
-                    row.names = F,
-                    booktabs = T, 
-                    align = 'l',
-                    caption = NULL) %>% 
-  kableExtra::kable_styling(latex_options = "hold_position", 
-                            position = "center") %>%
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, font_size = 8) %>% 
-  kableExtra::row_spec(0, bold = T)
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "t" = "Mobility treatment",
+  edad = "Age (in years)",
+  sexo = "Female (Ref. = Male)",
+  nac = "Chilean nacionality (Ref. = Non-Chilean)",
+  etnia = "Indigenous ethnicity (Ref. = Non-indigenous)",
+  hogar = "Co-residence with both parents (Ref. = No co-residence)",
+  edu0 = "Parental education",
+  ola2018 = "Wave 2018",
+  ola2023 = "Wave 2023"
+
+)
+
+texreg::texreg(list(mp_lm_r_ii,
+                       mp_lh_r_ii,
+                       mp_mh_r_ii,
+                       mp_ml_r_ii,
+                       mp_hm_r_ii,
+                       mp_hl_r_ii),
+                  custom.model.names = c(
+                    "Low-Middle",
+                    "Low-High",
+                    "Middle-High",
+                    "Middle-Low",
+                    "High-Middle",
+                    "High-Low"
+                  ),   caption.above = NULL,
+               caption = NULL,
+               custom.coef.map = ccoef,
+               digits = 2,
+               groups = list("Wave (Ref.= 2016)" = 9:10),
+               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.7,
+               center = T,
+               float.pos = "H!")
 ```
+
+```{r}
+#| label: tbl-robus3
+#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification using the original 5-point likert scale, with covariates and wave fixed effects."
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "t" = "Mobility treatment",
+  edad = "Age (in years)",
+  sexo = "Female (Ref. = Male)",
+  nac = "Chilean nacionality (Ref. = Non-Chilean)",
+  etnia = "Indigenous ethnicity (Ref. = Non-indigenous)",
+  hogar = "Co-residence with both parents (Ref. = No co-residence)",
+  edu0 = "Parental education",
+  ola2018 = "Wave 2018",
+  ola2023 = "Wave 2023"
+
+)
+
+texreg::texreg(list(mp_lm_r_iii,
+                       mp_lh_r_iii,
+                       mp_mh_r_iii,
+                       mp_ml_r_iii,
+                       mp_hm_r_iii,
+                       mp_hl_r_iii),
+                  custom.model.names = c(
+                    "Low-Middle",
+                    "Low-High",
+                    "Middle-High",
+                    "Middle-Low",
+                    "High-Middle",
+                    "High-Low"
+                  ),   caption.above = NULL,
+               caption = NULL,
+               custom.coef.map = ccoef,
+               digits = 2,
+               groups = list("Wave (Ref.= 2016)" = 9:10),
+               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.7,
+               center = T,
+               float.pos = "H!")
+```
+
+```{r}
+#| label: tbl-evalues1
+#| tbl-cap: "E-values for ATT estimates for Middle-High trajectory (Cinelli & Hazlett, 2020 approximation)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+ovb_minimal_reporting(mh_evalue, format = "latex")
+```
+
+```{r}
+#| label: tbl-evalues2
+#| tbl-cap: "E-values for ATT estimates for High-Middle trajectory (Cinelli & Hazlett, 2020 approximation)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+ovb_minimal_reporting(hm_evalue, format = "latex")
+```
+
+```{r}
+#| label: tbl-evalues3
+#| tbl-cap: "E-values for ATT estimates for High-Low trajectory (Cinelli & Hazlett, 2020 approximation)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+ovb_minimal_reporting(hl_evalue, format = "latex")
+```

### 2025-11-17 paper/paper.qmd
Update entrega 1
Up SP
Big up methods and results

@@
   \newline
- **Keywords**: pension commodification, social mobility, causal inference, Chile, public preferences
+ **Keywords**: Pension commodification · Intergenerational mobility · Causal inference · Meritocracy · Chile
 link-citations: true
@@
 
+::: {pagebreak}
+:::
+
 {{< include 07-supplementary-material.qmd >}}

### 2025-11-17 paper/paper.tex
up v2
Big up methods and results

@@
 \begin{abstract}
-My abstract \newline \textbf{Keywords}: pension commodification, social
-mobility, causal inference, Chile, public preferences
+My abstract \newline \textbf{Keywords}: Pension commodification ·
+Intergenerational mobility · Causal inference · Meritocracy · Chile
 \end{abstract}
@@
 \setstretch{1.5}
-This document was last modified at 2025-11-03 17:13:25 and it was last
-rendered at 2025-11-03 17:13:25.
+This document was last modified at 2025-11-17 19:08:04 and it was last
+rendered at 2025-11-17 19:08:04.
 
@@
 COVID-19 pandemic---and structural inequalities have become more visible
-(\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et al., 2021}).
-Despite its economic achievements, Chile remains one of the most unequal
-countries in Latin America and the OCDE. The poorest 50\% of the
-population captures just 10\% of total income and holds negative net
-wealth, while the wealthiest 1\% receives nearly 27\% of all income and
-controls almost half (49.6\%) of the country's wealth
+(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). Despite its
+economic achievements, Chile remains one of the most unequal countries
+in Latin America and the OCDE. The poorest 50\% of the population
+captures just 10\% of total income and holds negative net wealth, while
+the wealthiest 1\% receives nearly 27\% of all income and controls
+almost half (49.6\%) of the country's wealth
 (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
@@
 increased household debt, labor market precarity, and the COVID-19
-crisis (\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et al.,
-2021}). However, the majority of the population identifies themselves as
-members of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo
-et al., 2013}). These dynamics have fueled a legitimacy crisis:
-meritocracy is increasingly questioned, and many perceive that
-advancement depends more on informal networks than on effort or earned
-credentials (\citeproc{ref-barozet_clases_2021}{Barozet, Contreras, et
-al., 2021}). The combination of constrained mobility, persistent
-inequality, and concentrated opportunity has eroded expectations of
-upward mobility and intensified social discontent
-(\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).
+crisis (\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}).
+However, the majority of the population identifies themselves as members
+of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo et al.,
+2013}). These dynamics have fueled a legitimacy crisis: meritocracy is
+increasingly questioned, and many perceive that advancement depends more
+on informal networks than on effort or earned credentials
+(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). The
+combination of constrained mobility, persistent inequality, and
+concentrated opportunity has eroded expectations of upward mobility and
+intensified social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD,
+2017}).
 
@@
 (\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
-\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). I therefore adopt
-Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024}) causal
-formulation, treating destination class as the treatment, conditioning
-on the class of origin, and estimating heterogeneous destination effects
-with observational data under explicit identification assumptions. To
-align design and target, I follow the MIDA template
-(\citeproc{ref-blair_research_2023}{Blair et al., 2023}): set out the
-causal model and assumptions (M), define the inquiry and estimand (I),
-describe the data and variables (D), and detail the answer strategy (A)
-used to identify the causal effect of intergenerational social mobility
-on preferences for the commodification of pensions.
+\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). Following the
+causal framework of Breen and Ermisch
+(\citeproc{ref-breen_effects_2024}{2024}), I conceive of the destination
+class as a treatment, conditioning on the origin class and estimating
+the heterogeneous effects of the destination with observational data
+under explicit identification assumptions. To align design and target, I
+follow the MIDA template (\citeproc{ref-blair_research_2023}{Blair et
+al., 2023}): set out the causal model and assumptions (M), define the
+inquiry and estimand (I), describe the data and variables (D), and
+detail the answer strategy (A) used to identify the causal effect of
+intergenerational occupational mobility on preferences for the
+commodification of pensions.
 
@@
 comparisons rather than between-person contrasts. Defining mobility as
-\(M = D - O\) ---a change from origin \((O)\) to destination \((D)\),
-rather than an additive or interactive combination---places the focus
+\(M = D - O\); a change from origin \((O)\) to destination \((D)\),
+rather than an additive or interactive combination, places the focus
 squarely on the Neyman--Rubin problem: for any individual, we observe
@@
 
-In fact, establishing the causal relationship between objective social
-mobility and subjective outcomes, such as preferences, poses several
-challenges for causal inference. Below, I use directed acyclic graphs
-(DAGs) to illustrate some of these challenges and evaluate possible
-strategies for identifying the effects of social mobility.
-Figure~\ref{fig-dag} depicts the causal model guiding identification. In
-this model, a person's preference (\(Y\)) is directly caused by her
-class destination (\(D\)) and can also be affected by different
-attributes determined in childhood or earlier (\(C\)), such as family
-resources, household composition, and ascriptive status (pre-treatment
-confounders). Class origin (\(O\)) is determined by those
-origin-attributes (\(C\)) and, in turn, affects destination (\(D\)). In
-addition, unobserved factors (\(U\)) can also influence
-origin-attributes (\(C\)). There is no direct arrow from \(O\) to \(Y\),
-reflecting the assumption that the effect of origin on the outcome
-operates entirely through background factors \((C)\) and destination
-\((D)\). Accordingly, conditioning on \((O,C)\) could blocks the
-relevant back-door paths from \(D\) to \(Y\), specifically
-\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
-\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
-adjustment set.
+Establishing the causal relationship between objective social mobility
+and subjective outcomes, such as preferences, poses several challenges
+for causal inference. Below, I use directed acyclic graphs (DAGs) to
+illustrate some of these challenges and evaluate possible strategies for
+identifying the effects of social mobility. Figure~\ref{fig-dag} depicts
+the causal model guiding identification. In this model, a respondent's
+preference \(Y\) is directly affected by her class destination \(D\),
+and indirectly influenced by a set of pre-treatment attributes \(C\)
+rooted in childhood---family resources, household composition, parental
+education, and other ascriptive characteristics. Class origin \(O\) is
+itself shaped by these background factors \(C\) and, in turn, affects
+destination \(D\). I allow for unobserved determinants \(U\) that
+influence background factors \(C\), but I assume that any such
+unobserved variation operates only through \(C\). Thus, there are no
+direct paths \(U \rightarrow D\) or \(U \rightarrow Y\) beyond those
+mediated by \(C\).
+
+Under this structure, conditioning on \((O, C)\) blocks all relevant
+backdoor paths from \(D\) to \(Y\)---notably
+\(D \leftarrow O \leftarrow C \rightarrow Y\). This renders \((O, C)\) a
+minimal sufficient adjustment set for identifying the causal effect of
+mobility on preferences, in line with the framework proposed by Breen \&
+Ermisch (\citeproc{ref-breen_effects_2024}{2024}). Identification
+therefore relies on the assumption that background factors \(C\)
+adequately summarize pre-treatment characteristics that jointly shape
+both mobility and attitudes.
 
@@
 
-Formally, the DAG encodes the following relations:
-
-\[
-U \rightarrow C,\quad C \rightarrow O,\quad C \rightarrow D,\quad C \rightarrow Y,\quad O \rightarrow D,\quad D \rightarrow Y
-\]
-
-This structure corresponds to a nonparametric structural causal model in
-which \(D\) is the treatment variable, \(Y\) the outcome, and \((O, C)\)
-the minimal sufficient adjustment set. Conditioning on these variables
-satisfies the back-door criterion for identifying the causal effect of
-mobility on preferences (\citeproc{ref-breen_effects_2024}{Breen \&
-Ermisch, 2024}).
-
 Why are the assumptions plausible here? Conditional independence is
 targeted by controlling for a plausible sufficient adjustment set
-\((O,C)\); inverse probability weights (IPW) obtained via entropy
-balancing enforces covariate balance within each origin stratum.
-Positivity is assessed via propensity scores and trimming of extreme
-weights, restricting to common support. SUTVA holds by treating the
-destination class as a single exposure measured before \((Y)\) and
-assuming no interference. Together, the DAG and assumptions define
-conditions for identifying the causal effect of intergenerational social
-mobility on preferences for commodification.
+\((O,C)\) and by using inverse probability weights obtained via entropy
+balancing (\citeproc{ref-hainmueller_entropy_2012}{Hainmueller, 2012}),
+which enforce covariate balance within each origin stratum and mitigate
+selection on observables. Positivity is assessed by inspecting the
+distribution of the entropy-balancing weights and trimming observations
+with extreme weights, thus restricting inference to regions of common
+support. SUTVA holds by treating the destination class as a single
+exposure measured before \((Y)\) and assuming no interference. Together,
+the DAG and assumptions define conditions for identifying the causal
+effect of intergenerational social mobility on preferences for
+commodification.
 
@@
 social mobility affect individuals' preferences for the commodification
-of pensions in Chile? Following the potential outcomes framework
-proposed by Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024,
-p. 473}), the estimand of interest is the average treatment effect on
-the treated (ATT), defined as the mean causal effect of moving from an
-origin class \((O=j)\) to a destination class \((D=k)\), compared with
-the counterfactual outcome that those same individuals would have
-exhibited had they instead reached an alternative destination \((k')\):
-
-\begin{equation}\phantomsection\label{eq-breen}{ 
-ATT_{j,k,k′} =E[Y(D=k|O=j, D=k)]−E[Y(D=k′|O=j, D=k)]
-}\end{equation}
-
-In this study, I focus on the specific case where the counterfactual
-destination corresponds to class immobility \((k' = j)\), that is, how
-the outcome among people from origin \emph{j} who entered destination
-\emph{k} would have bee different if those people had, counterfactually,
-remained in origin \emph{j} instead (inmobility). Accordingly, the
-estimand becomes:
-
-\begin{equation}\phantomsection\label{eq-att}{ 
-ATT_{j,k,j} =E[Y(D=k|O=j, D=k)]−E[Y(D=j|O=j, D=k)]
-}\end{equation}
-
-which represents the average causal effect of attaining class
-\emph{(k)}---for instance, moving upward or downward---relative to
-remaining in class \emph{(j)}, among individuals who actually
-experienced such mobility. This estimand captures the within-origin
-counterfactual contrast at the heart of mobility effects: how a person's
-preference for market-based pensions would differ if, counterfactually,
-they had stayed in their origin class rather than moved to their
-observed destination.
+of pensions in Chile? Substantively, I am interested in the effect of
+mobility relative to immobility within a given class of origin.
+
+Following the potential outcomes framework proposed by Breen and Ermisch
+(\citeproc{ref-breen_effects_2024}{2024, p. 473}), the estimand of
+interest is an average treatment effect on the treated (ATT) defined
+within each origin class. I focus on the specific case in which the
+counterfactual destination corresponds to class immobility \((k' = j)\).
+In other words, I ask how the outcome among people from origin \emph{j}
+who entered destination \emph{k} would have bee different if those
+people had, counterfactually, remained in origin \emph{j} instead
+(inmobility).
+
+Formally, for individuals with origin \((O=j)\) who attain destination
+\((D=k)\), the estimand is:
+
+\$\$
+
+ATT\_\{j,k,j\} = E\big[Y(D = k) \mid O = j, D = k \big] -
+E\big[Y(D = j) \mid O = j, D = k \big]
+
+\$\$ \{\#eq-att\}
+
+which represents the mean causal effect of moving from origin class
+\((j)\) to destination class \((k)\), compared to the counterfactual
+outcome those same individuals would have exhibited had they instead
+remained in \((j)\). Thus, this ATT captures the counterfactual contrast
+within origin that lies at the heart of mobility effects: how
+preferences for market-based pensions would differ if, for the same
+upwardly or downwardly mobile individuals, their observed destination
+class were replaced by immobility in their origin class.
 
@@
 
-Because respondent occupation is not measured in every wave, I restrict
-the analysis to the 2016, 2018, and 2023 waves. Parental occupation was
-collected only in 2023 as open-text labels. I coded these texts to the
-ISCO-08 two-digit scheme using the National Institute of Statistics of
-Chile's automated coding API and retained cases with high-confidence
-matches (\textgreater95\%). Treating parental occupation as a
-time-invariant origin attribute, I carried these codes back to 2016 and
-2018. After listwise deletion and restricting to key variables
+Because respondent occupation\footnote{Parental occupation was collected
+  only in 2023 as open-text labels. I coded these texts to the ISCO-08
+  two-digit scheme using the National Institute of Statistics of Chile's
+  automated coding API and retained cases with high-confidence matches
+  (\textgreater95\%). Treating parental occupation as a time-invariant
+  origin attribute, I carried these codes back to 2016 and 2018.} is not
+measured in every wave, I restrict the analysis to the 2016, 2018, and
+2023 waves. After listwise deletion and restricting to key variables
 (respondent occupation and the outcome measure), the final analytic
@@
 Chile that people with higher incomes have better pensions than people
-with lower incomes?'' and were asked to indicate their level of
-agreement on a five-point Likert scale ranging from 1 (``strongly
-disagree'') to 5 (``strongly agree''). For estimation and
-interpretability, I dichotomized the item so that values 4-5 indicate
-agreement with market-based access to old-age pensions (coded as 1),
-while values 1-3 indicate disagreement (coded as 0). This binary coding
-treats neutrality as substantially closer to non-support than to support
-and produces a clear contrast between respondents who support
+with lower incomes?'' and answered on a five-point Likert scale ranging
+from 1 (``strongly disagree'') to 5 (``strongly agree''). For estimation
+and interpretability, I construct a binary indicator in which values
+4--5 indicate agreement with market-based access to old-age pensions
+(coded 1), while values 1--3 indicate non-agreement (coded 0).
+Substantively, the dependent variable is meant to capture
+\emph{endorsement} of pension commodification: only responses in the
+upper tail of the scale represent an active, explicit justification of
+market-based differentiation, whereas neutrality corresponds to an
+absence of such endorsement and is therefore treated as closer to
+non-support than to support.\footnote{A potential concern is that the
+  neutral category may reflect genuine ambivalence rather than
+  disagreement. To address this, I conduct robustness checks using
+  alternative codings: (i) excluding neutral respondents from the
+  analysis; (ii) combining the neutral and agreement categories (3--5
+  vs.~1--2); and (iii) estimating models with the original 5-point scale
+  using linear specifications. The main mobility effects reported below
+  are substantively unchanged across these specifications (see
+  \hyperref[anexo]{Supplementary Material}).} This binary coding thus
+yields a clear contrast between respondents who \emph{support}
 market-based differentiation in pensions and those who do not. I use
-this item for two main reasons: on the one hand, to be able to make
-comparisons with existing literature
+this item for two main reasons: first, to enable comparisons with
+existing work on market-based justice in social policy
 (\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
 2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}) and, on the
-other hand, because it captures two fundamental dimensions of
-market-based welfare distribution: the role of economic resources as a
-key determinant of outcomes and the framing of services as as tradable
-commodities that can be bought and sold based on ability to pay
+\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}); and second,
+because it taps two core dimensions of market-based welfare
+distribution---(i) the centrality of economic resources as a criterion
+for allocating outcomes and (ii) the framing of pensions as tradable
+commodities that can be bought and sold according to ability to pay
 (\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
 
-Table~\ref{tbl-summary1} shows the descriptive statistics of the
-variables used in the study.
+\subsubsection{Treatment}\label{treatment}
+
+\paragraph{Intergenerational occupational
+mobility}\label{intergenerational-occupational-mobility}
+
+I treat intergenerational occupational mobility as an exposure
+indicating whether respondents occupy a different occupational status
+than their fathers, closely following Breen and Ermisch's
+(\citeproc{ref-breen_effects_2024}{2024}) causal framework. Occupational
+assignment proceeds in two steps. First, I derive occupational status
+for both origin (father) and destination (respondent) from two-digit
+ISCO-08 codes using the International Socio-Economic Index of
+Occupational Status (ISEI). Second, I group these ISEI scores into
+terciles (low, middle, high), yielding a three-category schema for both
+origin and destination strata (see Table~\ref{tbl-matrix}).
+Substantively, I use ISEI because occupational status is a core
+indicator of stratification and a reliable mobility measure: it locates
+jobs on a hierarchical continuum defined by incumbents' typical
+education level and earnings
+(\citeproc{ref-hauser_intergenerational_2010}{Hauser, 2010};
+\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025}), and thus
+approximates long-run socio-economic position more closely than
+volatile, single-year income, which is also prone to recall and
+reporting error for both parental and own resources
+(\citeproc{ref-barone_rise_2022}{Barone et al., 2022}). This
+occupational focus is consistent with a long tradition that interprets
+the occupational structure as the backbone of the stratification system
+and a key determinant of life chances
+(\citeproc{ref-wrightUnderstandingClass2015}{Wright, 2015}). Moreover,
+ISEI was explicitly designed to harmonize occupational stratification
+across class schemes and SES measures
+(\citeproc{ref-ganzeboom_internationally_1996}{Ganzeboom \& Treiman,
+1996}), supporting cross-study comparability. In short, using ISEI
+aligns the measurement of origin and destination on a common vertical
+status continuum that is theoretically meaningful for social mobility
+research and empirically feasible given the available information of
+ELSOC's data.
+
+\begin{table}
+
+\caption{\label{tbl-matrix}Occupational mobility by occupational
+groups.}
+
+\centering{
+
+\centering
+\begin{tabular}{cccccc}
+\toprule
+Father↓ & Offspring→ & Low & Middle & High & Total\\
+\midrule
+Low &  & 44.7\%   (501) & 33.6\%   (377) & 21.7\%   (244) & 100.0\% (1,122)\\
+Middle &  & 35.1\%   (388) & 34.9\%   (385) & 30.0\%   (331) & 100.0\% (1,104)\\
+High &  & 24.6\%   (297) & 28.8\%   (348) & 46.7\%   (564) & 100.0\% (1,209)\\
+Total &  & 34.5\% (1,186) & 32.3\% (1,110) & 33.2\% (1,139) & 100.0\% (3,435)\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{Pre-treatment control variables for selection into
+treatment}\label{pre-treatment-control-variables-for-selection-into-treatment}
+
+Within each origin stratum (\(O = j\)), I estimate average treatment
+effects on the treated (ATT) by comparing movers to otherwise similar
+non-movers from the same origin. To reduce bias from non-random
+selection into mobility based on observed characteristics, I preprocess
+the data using entropy balancing, reweighting only the control group
+(the immobile) so that the distribution of pre-treatment origin
+covariates \(C\) among non-movers matches that of movers on selected
+moments. Entropy balancing implements a maximum-entropy reweighting
+scheme that chooses unit weights for non-movers to satisfy a set of
+balance constraints (e.g., equality of means and, where relevant, higher
+moments), while keeping the new weights as close as possible to the
+original weights (\citeproc{ref-hainmueller_entropy_2012}{Hainmueller,
+2012}). Substantively, this procedure addresses selection on
+observables: it reweights immobile respondents so that, within each
+origin class, they resemble mobile respondents in their observed
+background characteristics, making any remaining differences in the
+outcome more plausibly attributable to mobility rather than to
+pre-existing measured advantages.
+
+The selection of the adjustment set \(C\) follows two principles. First,
+it mirrors the logic in Breen and Ermisch
+(\citeproc{ref-breen_effects_2024}{2024}), who condition on attributes
+determined in childhood or earlier (e.g., parental resources, household
+structure, ascriptive traits). Second, it is constrained to origin-side
+characteristics available in ELSOC. Guided by theory and Chilean
+evidence on the determinants of relative mobility (mainly
+\citeproc{ref-brunori_inequality_2025}{Brunori et al., 2025};
+\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
+\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
+\citeproc{ref-torche_unequal_2005}{Torche, 2005}), I include six
+covariates covering family resources, household composition, and
+ascriptive status: (a) parental education (highest of father/mother), in
+10 ordinal categories from no schooling to postgraduate studies; (b)
+co-residence with both parents at age 15 (0 = no, 1 = yes); (c)
+nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0
+= male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
+This set of covariates is fixed prior to the destination class and
+constitutes the maximum set on the origin side in ELSOC, thus allowing
+for the plausible capture of both family and contextual influences that
+may affect career trajectories, making the conditional independence
+hypothesis more credible within each \(O = j\). Descriptive statistics
+for all variables are presented in Supplementary Material Table A1.
+
+After achieving balance, I construct stabilized IPW-ATT weights by
+setting treated (mobile) cases to weight 1 and assigning the
+entropy-balancing weights to controls, rescaled within each origin
+stratum so that the mean weight remains close to one. This aligns the
+counterfactual distribution of non-movers with the covariate profile of
+movers in each origin class (see \hyperref[anexo]{Supplementary
+Material} for balance diagnostics). The resulting weights are then used
+in the outcome stage to estimate the causal effect of intergenerational
+mobility on preferences for the commodification of pensions.
+
+\subsubsection{Effect heterogeneity}\label{effect-heterogeneity}
+
+To test heterogeneity in mobility effects, I examine two moderators:
+meritocratic perceptions and time. Meritocracy is captured with two
+items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
+effort (``In Chile, people are rewarded for their efforts'') and one to
+talent (``In Chile, people are rewarded for their intelligence and
+skills''), each answered on a five-point Likert scale (1 = ``strongly
+disagree'' to 5 = ``strongly agree''). I average the two items into a
+single index and dichotomize it into low meritocracy (≤ 3) and high
+meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
+and entered as a categorical (dummy) variable. Both variables are used
+as moderators by interacting them with the mobility treatment to assess
+conditional average treatment effects---i.e., whether the effect of
+mobility varies across levels of meritocratic perceptions or across
+survey waves.
+
+\subsubsection{Controls}\label{controls}
+
+All models include the same pre-treatment covariates \(C\) used in the
+IPW construction, not to re-balance groups (entropy balancing already
+does so), but to (i) block backdoor paths from \(C\) to the outcome
+\(Y\) as implied by the DAG, and (ii) achieve double robustness:
+estimates remain consistent if either the weighting model or the outcome
+model is correctly specified. Concretely, I adjust for (a) father's
+educational level, (b) co-residence with both parents at age 15, (c)
+nationality, (d) age, (e) sex, and (f) ethnicity, along with wave fixed
+effects to absorb secular trends.
+
+\subsection{Analytical strategy}\label{analytical-strategy}
+
+Because the dependent variable---the preference for pension
+commodification---is binary \((Y_{it}\in{0,1})\), I estimate weighted
+linear probability models (WLS) using stabilized inverse probability
+weights (IPW) obtained via entropy balancing. Although OLS is often
+questioned with binary outcomes, it consistently estimates the
+conditional mean \(E(Y|X)\) under standard exogeneity
+\(E[\varepsilon_i|X_i]=0\), and heteroskedasticity can be handled with
+robust or clustered standard errors
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
+ch.~7.5). WLS with IPW further improves efficiency and implements the
+ATT estimand by recreating the counterfactual distribution of non-movers
+(\citeproc{ref-gelman_regression_2020}{Gelman et al., 2020, pp.
+270--272}). Linear models are preferred here because coefficients are
+directly interpretable as average differences in predicted
+probabilities, whereas non-linear models complicate both weighting and
+interpretation (\citeproc{ref-gelman_regression_2020}{Gelman et al.,
+2020}, ch.~13).
+
+Formally, the model is:
+
+\begin{equation}\phantomsection\label{eq-final}{
+Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
+}\end{equation}
+
+where \(Y_{it}\) indicates whether individual \(i\) in wave \(t\)
+supports more market-based pension access; \(\alpha\) is the baseline
+probability for the reference categories; \(\beta T_i\) captures the
+intergenerational mobility contrast: \(T_i=1\) when the observed
+destination is \(D_i=k\) (mobile to \(k\)) and \(T_i=0\) when \(D_i=j\)
+(immobile), so that \(\beta\) is the ATT within origin
+\(O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})\), i.e., the
+percentage-point change in the probability of preferring pension
+commodification from reaching \(k\) rather than remaining in \(j\). The
+term \(X_i'\gamma\) includes the pre-treatment covariates \((C)\) used
+to construct the entropy-balancing weights; \(\lambda_t\) are wave fixed
+effects; and \(\varepsilon_{it}\) is idiosyncratic error. The analytic
+sample is restricted to individuals sharing the same origin \((O=j)\).
+Therefore, estimation uses pooled WLS with stabilized IPW--ATT from
+entropy balancing and CR2 standard errors clustered by individual to
+address heteroskedasticity and within-person dependence.
+
+The specification is doubly robust: entropy-balancing weights are
+combined with covariate adjustment so consistency holds if either the
+weighting model or the outcome model is correctly specified, while also
+improving efficiency
+(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
+results are virtually identical to baseline models without covariates
+(see \hyperref[anexo]{Supplementary Material} for complete models), with
+no substantive changes in the mobility coefficients. Heterogeneous
+effects by meritocratic perceptions and survey wave are examined through
+interaction models described in the previous section.
+
+To evaluate the robustness of the findings, I implement two sets of
+sensitivity analyses. First, I conduct standard robustness checks by
+re-estimating the models under alternative codings of the outcome: (i)
+excluding neutral responses from the analysis, (ii) combining neutral
+responses with agreement (3--5 vs.~1--2), and (iii) using the original
+5-point item in a linear specification. Second, I assess sensitivity to
+unmeasured confounding using the omitted-variable--bias framework of
+Cinelli and Hazlett (\citeproc{ref-cinelli_making_2020}{2020}), which
+quantifies how strong an unobserved confounder would need to be to
+explain away the estimated ATT.
+
+\section{Results}\label{results}
+
+\subsection{Descriptive statistics}\label{descriptive-statistics}
+
+Figure~\ref{fig-alluvial} shows the annual frequencies of preferences
+for pension commodification in 2016, 2018, and 2023. Each year presents
+stacked percentages frequencies for the level of agreement and
+disagreement. Overall, a large majority rejects market-based access to
+old-age pension benefits, but this opposition has eased over time:
+83.5\% in 2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018
+change is modest, whereas 2023 registers a 9.4 point drop in
+disagreement relative to 2018, mirrored by rising agreement: 16.5\%
+(2016), 18.8\% (2018), 28.2\% (2023). Substantively, while most
+respondents continue to oppose the idea that higher-income individuals
+should obtain better pensions via the market, a non-trivial and growing
+portion endorses this statement, with the sharpest expansion
+concentrated in the latest wave (+9.4 points from 2018 to 2023).
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-alluvial-1.pdf}
+
+}
+
+\caption{\label{fig-alluvial}Change in preferences for pension
+commodification over time (2016, 2018, and 2023).}
+
+\end{figure}%
+
+Regarding the relationship between preferences for the commodification
+of pensions and intergenerational occupational mobility,
+Figure~\ref{fig-mean} shows the percentage of agreement with
+market-based access to pension benefits according to mobility
+trajectories and survey waves. Averaging across waves, agreement is
+highest among the High-High immobile (27.8\%), followed by the
+Middle-High upwardly mobile (24.6\%) and the High-Middle downwardly
+mobile (24.0\%). The lowest levels are found among ascending Low-High
+(17.3\%), and immobile Low-Low (17.4\%) groups. The specific profiles of
+each wave accentuate these gradients in most trajectories, especially in
+2023: agreement reaches 31.9\% for High-High, 29.4\% for upward
+Middle-High, and peaks at 36.8\% for downward High-Middle (the highest
+of all trajectories for that year). Descriptively, immobility at the top
+is associated with greater support for commodification; among those who
+move, support is greatest for the upward Middle-High trajectory and the
+downward High-Middle trajectory, patterns that intensify in 2023. These
+shifts anticipate the heterogeneity documented below.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-mean-1.pdf}
+
+}
+
+\caption{\label{fig-mean}Percentage agreement on preferences for pension
+commodification according to mobility trajectory and time (2016, 2018,
+and 2023).}
+
+\end{figure}%
+
+\subsection{Mobility effects models}\label{mobility-effects-models}
+
+Figure~\ref{fig-reg} reports the double-robust estimates of the effect
+of intergenerational occupational mobility on preferences for pension
+commodification, obtained from weighted models that combine stabilized
+IPW with covariate adjustment and fixed wave effects. Results reveal a
+clear directional asymmetry between upward and downward mobility
+trajectories. Among upwardly mobile respondents, only the Middle→High
+trajectory exhibits a significant positive effect on support for pension
+commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
+Interpreted as an origin-specific ATT, this estimate compares---for
+individuals who actually moved from the middle to the high socioeconomic
+status---their observed endorsement with the counterfactual endorsement
+they would have expressed had they remained immobile in the middle
+socioeconomic status. Holding all other predictors constant, the point
+estimate implies a 9 percentage point higher probability of endorsing
+market-based access to pension benefits for Middle→High movers relative
+to their own immobility counterfactual, statistically significant at the
+95\% confidence level. In contrast, two downward trajectories:
+High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
+(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
+effects, suggesting that individuals who experience downward mobility
+from higher-status origins express weaker preferences for market-based
+pension benefits. The remaining pathways (Low→Middle, Low→High,
+Middle→Low) display no significant differences relative to their
+non-mobile counterparts. These patterns indicate that upward movement
+within the upper segment (Middle→High) reinforces pro-commodification
+attitudes, whereas downward movement from privileged origins reduces
+them.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-reg-1.pdf}
+
+}
+
+\caption{\label{fig-reg}Effects of intergenerational occupational
+mobility on preferences for pension commodification. Coefficient plot of
+origin-specific ATT estimates comparing mobility versus immobility.
+Estimates from pooled WLS with doubly robust adjustment, wave fixed
+effects, and standard errors clustered by individual; bars show 95\%
+confidence intervals.}
+
+\end{figure}%
+
+\subsection{Effect heterogeneity}\label{effect-heterogeneity-1}
+
+To probe potential mechanisms underlying the mobility effects, I
+estimate origin-specific ATT models that interact the mobility treatment
+with high meritocratic perceptions (ref. = low), using stabilized IPW
+weights, covariate adjustment, and wave fixed effects. I also estimate
+an analogous specification interacting mobility with survey waves (ref.
+= 2016) to assess temporal heterogeneity in the treatment effect (see
+\hyperref[anexo]{Supplementary Material} for complete models).
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-1.pdf}
+
+}
+
+\caption{\label{fig-interact}Effects of intergenerational occupational
+mobility on preferences for pension commodification, by meritocratic
+perception. Coefficient plot of origin-specific ATT estimates comparing
+mobility versus immobility, estimated from interaction models (T ×
+Merit). Points show simple slopes of the mobility effect at low merit
+and high merit; bars are 95\% confidence intervals. Estimates from
+pooled WLS with doubly robust adjustment, wave fixed effects, and
+standard errors clustered by individual.}
+
+\end{figure}%
+
+Figure~\ref{fig-interact} shows that neither the interaction terms nor
+the mobility coefficients reach conventional significance across
+trajectories, indicating no detectable moderation of the causal mobility
+effect by meritocratic perceptions at the 95\% level. Directionally, in
+the left panel, the interaction is positive for Middle→High (\(\beta\) =
+0.15, 95\% CI {[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05,
+{[}−0.14, 0.24{]}), suggesting, if anything, somewhat larger
+pro-commodification effects among high-merit respondents when upwardly
+mobile, whereas it is negative for High→Low (\(\beta\) = −0.14,
+{[}−0.38, 0.10{]}) and High→Middle (\(\beta\) = −0.03, {[}−0.23,
+0.17{]}), showing weaker pro-commodification effects under downward
+mobility at high merit. Focusing on the treatment effect at low
+meritocracy (the reference category) at the right panel, the mobility
+coefficient captures the origin-specific ATT comparing movers to their
+immobility counterfactual among low-merit respondents. These effects are
+small and imprecise across trajectories: Low→Middle (\(\beta\) = −0.02,
+{[}−0.09, 0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]})
+are near zero. Middle→High is positive but not significant (\(\beta\) =
+0.07, {[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00,
+{[}−0.07, 0.06{]}), and downward from high shows negative, borderline
+estimates: High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and
+High→Low (\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not
+find evidence that meritocratic perceptions moderate the causal effect
+of mobility, although the signs align with the idea that upward mobility
+at high merit tilts attitudes toward commodification, and downward
+mobility at high merit tilts them away.
+
+\begin{table}
+
+\caption{\label{tbl-inttime}Effects of intergenerational occupational
+mobility on preferences for pension commodification, by survey wave}
+
+\centering{
+
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                              & $0.18$           & $-0.04$          & $0.28$           & $0.20$           & $0.18$            & $0.15$            \\
+                                       & $ [-0.36; 0.73]$ & $ [-0.36; 0.29]$ & $ [-0.01; 0.57]$ & $ [-0.11; 0.52]$ & $ [-0.15;  0.50]$ & $ [-0.27;  0.58]$ \\
+Mobility treatment                     & $-0.05$          & $-0.03$          & $0.13^{*}$       & $0.06$           & $-0.19^{*}$       & $-0.19^{*}$       \\
+                                       & $ [-0.16; 0.06]$ & $ [-0.15; 0.10]$ & $ [ 0.05; 0.22]$ & $ [-0.03; 0.15]$ & $ [-0.33; -0.05]$ & $ [-0.32; -0.05]$ \\
+Wave (Ref.= 2016)                      &                  &                  &                  &                  &                   &                   \\
+                                       &                  &                  &                  &                  &                   &                   \\
+\quad Wave 2018                        & $0.01$           & $0.01$           & $0.10^{*}$       & $0.09^{*}$       & $-0.04$           & $-0.02$           \\
+                                       & $ [-0.10; 0.12]$ & $ [-0.10; 0.12]$ & $ [ 0.03; 0.18]$ & $ [ 0.01; 0.17]$ & $ [-0.15;  0.07]$ & $ [-0.15;  0.10]$ \\
+\quad Wave 2023                        & $0.03$           & $0.01$           & $0.19^{*}$       & $0.18^{*}$       & $0.05$            & $0.04$            \\
+                                       & $ [-0.09; 0.14]$ & $ [-0.12; 0.13]$ & $ [ 0.10; 0.28]$ & $ [ 0.08; 0.27]$ & $ [-0.09;  0.18]$ & $ [-0.11;  0.19]$ \\
+Mobility treatment x Wave (Ref.= 2016) &                  &                  &                  &                  &                   &                   \\
+                                       &                  &                  &                  &                  &                   &                   \\
+\quad Mobility treatment x Wave 2018   & $0.03$           & $-0.06$          & $-0.04$          & $-0.09$          & $0.11$            & $0.06$            \\
+                                       & $ [-0.11; 0.17]$ & $ [-0.21; 0.09]$ & $ [-0.17; 0.09]$ & $ [-0.20; 0.03]$ & $ [-0.04;  0.26]$ & $ [-0.09;  0.22]$ \\
+\quad Mobility treatment x Wave 2023   & $0.07$           & $0.10$           & $-0.09$          & $-0.07$          & $0.19^{*}$        & $0.18$            \\
+                                       & $ [-0.09; 0.23]$ & $ [-0.08; 0.28]$ & $ [-0.23; 0.06]$ & $ [-0.22; 0.08]$ & $ [ 0.01;  0.38]$ & $ [-0.01;  0.37]$ \\
+\hline
+Controls                               & Yes              & Yes              & Yes              & Yes              & Yes               & Yes               \\
+R$^2$                                  & $0.04$           & $0.05$           & $0.07$           & $0.03$           & $0.06$            & $0.06$            \\
+Adj. R$^2$                             & $0.03$           & $0.04$           & $0.06$           & $0.02$           & $0.05$            & $0.05$            \\
+Num. obs.                              & $878$            & $745$            & $716$            & $773$            & $912$             & $861$             \\
+RMSE                                   & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$            & $0.42$            \\
+N Clusters                             & $497$            & $449$            & $443$            & $470$            & $536$             & $525$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
+effect across the different waves of the survey. Regarding temporal
+heterogeneity, the only statistically significant interaction term is
+observed for the High→Middle trajectory in 2023
+(\(\beta_{\text{int}} = 0.19\)), 95\% CI {[}0.01, 0.38{]}). However,
+when this interaction is added to the baseline effect in 2016 (-0.19),
+the marginal effect of High→Middle mobility in 2023 is approximately
+zero, and its confidence interval includes the null. Thus, rather than
+indicating a reversal toward a positive effect, the evidence suggests
+that the negative impact of downward mobility from high backgrounds is
+confined to 2016 and attenuates over time. No comparable temporal
+changes are observed in the remaining trajectories. Overall, the results
+provide little support for strong time-conditional heterogeneity in
+mobility effects, beyond the fading of the initially negative
+High→Middle effect.
+
+\subsection{Robustness check and sensitivity
+analysis}\label{robustness-check-and-sensitivity-analysis}
+
+As a robustness check, I re-estimate the models under three alternative
+codings of the outcome: (i) recoding the neutral category as agreement
+(3--5 vs.~1--2), (ii) excluding neutral responses from the analysis, and
+(iii) using the original 5-point item in a linear specification. Across
+these alternative models (see \hyperref[anexo]{Supplementary Material}
+for complete tables), the substantive pattern of the main effects is
+preserved: the Middle→High trajectory remains positively associated with
+support for pension commodification, and the High→Low trajectory remains
+negatively associated with it, with both coefficients generally
+increasing in magnitude. By contrast, the negative High→Middle effect
+becomes statistically indistinguishable from zero in the robustness
+specifications, indicating that this particular estimate is more
+sensitive to how the outcome is coded than the other two trajectories.
+
+To assess the extent to which the estimated mobility effects may be
+driven by unmeasured confounding, I implement the sensitivity analysis
+proposed by Cinelli and Hazlett
+(\citeproc{ref-cinelli_making_2020}{2020}), which quantifies how strong
+an unobserved confounder would need to be---in terms of its joint
+explanatory power for both treatment and outcome---to attenuate or
+explain away the ATT, using gender as a benchmark covariate (the
+predictor with the largest partial association with the outcome aside
+from treatment). For the Middle→High trajectory
+(\(\beta \approx 0.09\)), partial \(R^2 \approx 0.013\), an unobserved
+confounder would need to explain about 10--11\% of the residual variance
+of both treatment and outcome to reduce the effect to zero and roughly
+4\% to render it statistically insignificant, substantially more than
+gender does. For the High→Low effect (\(\beta \approx -0.10\)), partial
+\(R^2 \approx 0.012\), the required strength is of similar magnitude
+(around 10--11\% to explain away the effect and about 4\% to remove
+significance), again exceeding the explanatory power of the main
+observed covariates. By contrast, the High→Middle effect
+(\(\beta \approx -0.07\)), partial \(R^2 \approx 0.007\) is somewhat
+less robust: a confounder explaining around 8\% of residual variance in
+both treatment and outcome could nullify the estimate, and one
+explaining about 2\% could make it non-significant. Overall, the
+sensitivity analysis suggests that the positive Middle→High and negative
+High→Low effects are moderately robust to unmeasured confounding,
+whereas the High→Middle effect is more vulnerable to relatively modest
+omitted variables (see \hyperref[anexo]{Supplementary Material} for
+complete tables).
+
+\section{Discussion}\label{discussion}
+
+\section{Conclusion}\label{conclusion}
+
+\section{References}\label{references}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-alesina_intergenerational_2018}
+Alesina, A., Stantcheva, S., \& Teso, E. (2018). Intergenerational
+{Mobility} and {Preferences} for {Redistribution}. \emph{American
+Economic Review}, \emph{108}(2), 521--554.
+\url{https://doi.org/10.1257/aer.20162015}
+
+\bibitem[\citeproctext]{ref-araujo_desafios_2012}
+Araujo, K., \& Martuccelli, D. (2012). \emph{Desafíos comunes: Retrato
+de la sociedad chilena y sus individuos} (1a. ed). Santiago: LOM
+Ediciones.
+
+\bibitem[\citeproctext]{ref-ares_changing_2020}
+Ares, M. (2020). Changing classes, changing preferences: How social
+class mobility affects economic preferences. \emph{West European
+Politics}, \emph{43}(6), 1211--1237.
+\url{https://doi.org/10.1080/01402382.2019.1644575}
+
+\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
+Arrizabalo, X. (1995). \emph{{Milagro o quimera: la economía chilena
+durante la dictadura}}. Libros de la Catarata.
+
+\bibitem[\citeproctext]{ref-barone_rise_2022}
+Barone, C., Hertel, F. R., \& Smallenbroek, O. (2022). The rise of
+income and the demise of class and social status? {A} systematic review
+of measures of socio-economic position in stratification research.
+\emph{Research in Social Stratification and Mobility}, \emph{78},
+100678. \url{https://doi.org/10.1016/j.rssm.2022.100678}
+
+\bibitem[\citeproctext]{ref-barozet_clases_2021}
+Barozet, E., Contreras, D., Espinoza, V., Gayo, M., \& Méndez, M. L.
+(2021). \emph{{Clases medias en tiempos de crisis: vulnerabilidad
+persistente, desafíos para la cohesión y un nuevo pacto social en
+Chile}}. Santiago: Comisión Económica para América Latina y el Caribe
+(CEPAL).
+
+\bibitem[\citeproctext]{ref-batruch_belief_2023}
+Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
+(2023). Belief in {School Meritocracy} and the {Legitimization} of
+{Social} and {Income Inequality}. \emph{Social Psychological and
+Personality Science}, \emph{14}(5), 621--635.
+\url{https://doi.org/10.1177/19485506221111017}
+
+\bibitem[\citeproctext]{ref-benabou_social_2001}
+Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
+{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
+Economics}, \emph{116}(2), 447--487.
+\url{https://doi.org/10.1162/00335530151144078}
+
+\bibitem[\citeproctext]{ref-blair_research_2023}
+Blair, G., Coppock, A., \& Humphreys, M. (2023). \emph{Research design
+in the social sciences: Declaration, diagnosis, and redesign}. Princeton
+(N.J.) Oxford: Princeton University press.
+
+\bibitem[\citeproctext]{ref-boccardo_30_2020}
+Boccardo, G. (2020). \emph{30 años de privatizaciones en {Chile}: Lo que
+la pandemia reveló} (Nodo XXI). Santiago.
+
+\bibitem[\citeproctext]{ref-breen_effects_2024}
+Breen, R., \& Ermisch, J. (2024). The {Effects} of {Social Mobility}.
+\emph{Sociological Science}, \emph{11}, 467--488.
+\url{https://doi.org/10.15195/v11.a17}
+
+\bibitem[\citeproctext]{ref-brunori_inequality_2025}
+Brunori, P., Ferreira, F. H. G., \& Neidhöfer, G. (2025). Inequality of
+opportunity and intergenerational persistence in {Latin America}.
+\emph{Oxford Open Economics}, \emph{4}(Supplement\_1), i167--i199.
+\url{https://doi.org/10.1093/ooec/odae021}
+
+\bibitem[\citeproctext]{ref-bucca_merit_2016}
+Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
+Latin Americans}' beliefs about wealth and poverty. \emph{Research in
+Social Stratification and Mobility}, \emph{44}, 98--112.
+\url{https://doi.org/10.1016/j.rssm.2016.02.005}
+
+\bibitem[\citeproctext]{ref-busemeyer_skills_2014}
+Busemeyer, M. (2014). \emph{Skills and {Inequality}: {Partisan Politics}
+and the {Political Economy} of {Education Reforms} in {Western Welfare
+States}}. Cambridge University Press.
+
+\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
+Busemeyer, M., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive} and
+{Negative}: {New Perspectives} on {Feedback Effects} in {Public Opinion}
+on the {Welfare State}. \emph{British Journal of Political Science},
+\emph{51}(1), 137--162. \url{https://doi.org/10.1017/S0007123418000534}
+
+\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
+Busemeyer, M., \& Iversen, T. (2020). The {Welfare State} with {Private
+Alternatives}: {The Transformation} of {Popular Support} for {Social
+Insurance}. \emph{The Journal of Politics}, \emph{82}(2), 671--686.
+\url{https://doi.org/10.1086/706980}
+
+\bibitem[\citeproctext]{ref-campbell_institutional_2020}
+Campbell, J. L. (2020). \emph{Institutional {Change} and
+{Globalization}}. Princeton University Press.
+\url{https://doi.org/10.2307/j.ctv131bw68}
+
+\bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
+Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
+(2021). Sujeto y cotidiano en la era neoliberal: El caso de la educación
+chilena. \emph{Revista Mexicana de Ciencias Políticas y Sociales},
+\emph{67}(244).
+\url{https://doi.org/10.22201/fcpys.2448492xe.2022.244.70386}
+
+\bibitem[\citeproctext]{ref-castillo_legitimacy_2011}
+Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
+Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
+Research}, \emph{24}(4), 314--340.
+\url{https://doi.org/10.1007/s11211-011-0144-5}
+
+\bibitem[\citeproctext]{ref-castillo_changes_2025}
+Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
+{Justification} of {Educational Inequalities}: {The Role} of
+{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
+Pandemic}. \emph{Social Justice Research}.
+\url{https://doi.org/10.1007/s11211-025-00458-0}
+
+\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
+Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
+(2023). A {Multidimensional Approach} for {Measuring Meritocratic
+Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
+Social Inequality Survey}. \emph{International Journal of Sociology},
+1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
+
+\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
+Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
+Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
+{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
+\emph{Under Review at Frontiers in Sociology}.
+
+\bibitem[\citeproctext]{ref-castillo_todos_2013}
+Castillo, J. C., Miranda, D., \& Cabib, I. M. (2013). Todos somos de
+clase media: {Sobre} el estatus social subjetivo en {Chile}. \emph{Latin
+American Research Review}, \emph{48}(1), 155--173.
+\url{https://doi.org/10.1353/lar.2013.0006}
+
+\bibitem[\citeproctext]{ref-castillo_socialization_2024}
+Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
+{Socialization} of {Meritocracy} and {Market Justice Preferences} at
+{School}. \emph{Societies}, \emph{14}(11), 214.
+\url{https://doi.org/10.3390/soc14110214}
+
+\bibitem[\citeproctext]{ref-castillo_meritocracia_2019}
+Castillo, J. C., Torres, A., Atria, J., \& Maldonado, L. (2019).
+Meritocracia y desigualdad económica: {Percepciones}, preferencias e
+implicancias. \emph{Revista Internacional de Sociología}, \emph{77}(1),
+117. \url{https://doi.org/10.3989/ris.2019.77.1.17.114}
+
+\bibitem[\citeproctext]{ref-chancel_world_2022}
+Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
+inequality report 2022.
+https://bibliotecadigital.ccb.org.co/handle/11520/27510.
+
+\bibitem[\citeproctext]{ref-cinelli_making_2020}
+Cinelli, C., \& Hazlett, C. (2020). Making {Sense} of {Sensitivity}:
+{Extending Omitted Variable Bias}. \emph{Journal of the Royal
+Statistical Society Series B: Statistical Methodology}, \emph{82}(1),
+39--67. \url{https://doi.org/10.1111/rssb.12348}
+
+\bibitem[\citeproctext]{ref-day_movin_2017}
+Day, M. V., \& Fiske, S. T. (2017). Movin' on {Up}? {How Perceptions} of
+{Social Mobility Affect Our Willingness} to {Defend} the {System}.
+\emph{Social Psychological and Personality Science}, \emph{8}(3),
+267--274. \url{https://doi.org/10.1177/1948550616678454}
+
+\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
+Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificación y
+movilidad social bajo un modelo neoliberal: El caso de Chile}.
+\emph{Lavboratorio}, (25).
+
+\bibitem[\citeproctext]{ref-espinoza_movilidad_2014}
+Espinoza, V., \& Núñez, J. (2014). Movilidad ocupacional en {Chile}
+2001-2009. \textquestiondown{{Desigualdad}} de ingresos con igualdad de
+oportunidades? \emph{Revista Internacional de Sociología}, \emph{72}(1),
+57--82. \url{https://doi.org/10.3989/ris.2011.11.08}
+
+\bibitem[\citeproctext]{ref-eyles_social_2022}
+Eyles, A., Major, L. E., \& Machin, S. (2022). \emph{Social {Mobility} -
+{Past}, {Present} and {Future}: {The State} of {Play} in {Social
+Mobility}, on the 25th {Anniversary} of the {Sutton Trust}}. Sutton
+Trust.
+
+\bibitem[\citeproctext]{ref-ferre_welfare_2023}
+Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
+America}. \emph{Journal of International and Comparative Social Policy},
+\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}
+
+\bibitem[\citeproctext]{ref-flores_top_2020}
+Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
+in {Chile}: {A Historical Perspective} on {Income Inequality},
+1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
+\url{https://doi.org/10.1111/roiw.12441}
+
+\bibitem[\citeproctext]{ref-ganzeboom_internationally_1996}
+Ganzeboom, H. B. G., \& Treiman, D. J. (1996). Internationally
+{Comparable Measures} of {Occupational Status} for the 1988
+{International Standard Classification} of {Occupations}. \emph{Social
+Science Research}, \emph{25}(3), 201--239.
+\url{https://doi.org/10.1006/ssre.1996.0010}
+
+\bibitem[\citeproctext]{ref-gelman_regression_2020}
+Gelman, A., Hill, J., \& Vehtari, A. (2020). \emph{Regression and {Other
+Stories}} (1st ed.). Cambridge University Press.
+\url{https://doi.org/10.1017/9781139161879}
+
+\bibitem[\citeproctext]{ref-gingrich_making_2011}
+Gingrich, J. R. (2011). \emph{Making {Markets} in the {Welfare State}:
+{The Politics} of {Varying Market Reforms}} (1st ed.). Cambridge
+University Press. \url{https://doi.org/10.1017/CBO9780511791529}
+
+\bibitem[\citeproctext]{ref-gugushvili_trends_2014}
+Gugushvili, A. (2014). Trends, {Covariates} and {Consequences} of
+{Intergenerational Social Mobility} in {Post- Socialist Societies}.
+
+\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016c}
+Gugushvili, A. (2016a). Intergenerational objective and subjective
+mobility and attitudes towards income differences: Evidence from
+transition societies. \emph{Journal of International and Comparative
+Social Policy}, \emph{32}(3), 199--219.
+\url{https://doi.org/10.1080/21699763.2016.1206482}
+
+\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016}
+Gugushvili, A. (2016b). Intergenerational {Social Mobility} and {Popular
+Explanations} of {Poverty}: {A Comparative Perspective}. \emph{Social
+Justice Research}, \emph{29}(4), 402--428.
+\url{https://doi.org/10.1007/s11211-016-0275-9}
+
+\bibitem[\citeproctext]{ref-gugushvili_subjective_2017}
+Gugushvili, A. (2017). Subjective {Intergenerational Mobility} and
+{Support} for {Welfare State Programmes}.
+
+\bibitem[\citeproctext]{ref-gugushvili_heterogeneous_2024}
+Gugushvili, A. (2024). The heterogeneous well-being effects of
+intergenerational mobility perceptions. \emph{Journal of Health
+Psychology}, \emph{29}(2), 99--112.
+\url{https://doi.org/10.1177/13591053231187345}
+
+\bibitem[\citeproctext]{ref-hadjar_does_2015}
+Hadjar, A., \& Samuel, R. (2015). Does upward social mobility increase
+life satisfaction? {A} longitudinal analysis using {British} and {Swiss}
+panel data. \emph{Research in Social Stratification and Mobility},
+\emph{39}, 48--58. \url{https://doi.org/10.1016/j.rssm.2014.12.002}
+
+\bibitem[\citeproctext]{ref-hainmueller_entropy_2012}
+Hainmueller, J. (2012). Entropy {Balancing} for {Causal Effects}: {A
+Multivariate Reweighting Method} to {Produce Balanced Samples} in
+{Observational Studies}. \emph{Political Analysis}, \emph{20}(1),
+25--46. \url{https://doi.org/10.1093/pan/mpr025}
+
+\bibitem[\citeproctext]{ref-hauser_intergenerational_2010}
+Hauser, R. M. (2010). Intergenerational {Economic Mobility} in the
+{United States} ---{Measures}, {Differentials}, and {Trends}.
+
+\bibitem[\citeproctext]{ref-helgason_longterm_2023}
+Helgason, A. F., \& Rehm, P. (2023). Long-term income trajectories and
+the evolution of political attitudes. \emph{European Journal of
+Political Research}, \emph{62}(1), 264--284.
+\url{https://doi.org/10.1111/1475-6765.12506}
+
+\bibitem[\citeproctext]{ref-helgason_class_2025}
+Helgason, A. F., \& Rehm, P. (2025). Class experiences and the long-term
+evolution of economic values. \emph{Social Forces}, \emph{103}(3),
+1125--1143. \url{https://doi.org/10.1093/sf/soae135}
+
+\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
+Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
+D. (2023). Mindsets of poverty: {Implications} for redistributive policy
+support. \emph{Analyses of Social Issues and Public Policy},
+\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
+
+\bibitem[\citeproctext]{ref-immergut_theoretical_1998}
+Immergut, E. M. (1998). The {Theoretical Core} of the {New
+Institutionalism}. \emph{Politics \& Society}, \emph{26}(1), 5--34.
+\url{https://doi.org/10.1177/0032329298026001002}
+
+\bibitem[\citeproctext]{ref-immergut_it_2020}
+Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
+affluent to be able to purchase {``better''} healthcare? {Existential}
+standards and institutional norms in healthcare attitudes across 28
+countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
+\url{https://doi.org/10.1016/j.socscimed.2020.113146}
+
+\bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
+Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social mobility
+and demand for redistribution in {Europe}: A comparative analysis.
+\emph{The British Journal of Sociology}, \emph{70}(1), 138--165.
+\url{https://doi.org/10.1111/1468-4446.12363}
+
+\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
+Janmaat, J. G. (2013). Subjective {Inequality}: A {Review} of
+{International Comparative Studies} on {People}'s {Views} about
+{Inequality}. \emph{European Journal of Sociology}, \emph{54}(3),
+357--389. \url{https://doi.org/10.1017/S0003975613000209}
+
+\bibitem[\citeproctext]{ref-jasso_justice_1978}
+Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification}
+of the {Justice Evaluation Function}. \emph{American Journal of
+Sociology}, \emph{83}(6), 1398--1419.
+\url{https://doi.org/10.1086/226706}
+
+\bibitem[\citeproctext]{ref-jasso_distributive_2016}
+Jasso, G., Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
+{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
+Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
+\url{https://doi.org/10.1007/978-1-4939-3216-0_11}
+
+\bibitem[\citeproctext]{ref-jasso_gender_1999}
+Jasso, G., \& Wegener, B. (1999). Gender and {Country Differences} in
+the {Sense} of {Justice}: {Justice Evaluation}, {Gender Earnings Gap},
+and {Earnings Functions} in {Thirteen Countries}. \emph{International
+Journal of Comparative Sociology}, \emph{40}(1), 94--115.
+\url{https://doi.org/10.1177/002071529904000106}
+
+\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
+of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
+about {Market Justice}, 1991-1996. \emph{European Sociological Review},
+\emph{15}(3), 251--283. Retrieved from
+\url{https://www.jstor.org/stable/522731}
+
+\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
+Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
+\emph{Annual Review of Sociology}, 29--56.
+
+\bibitem[\citeproctext]{ref-koos_moral_2019}
+Koos, S., \& Sachweh, P. (2019). The moral economies of market
+societies: Popular attitudes towards market competition, redistribution
+and reciprocity in comparative perspective. \emph{Socio-Economic
+Review}, \emph{17}(4), 793--821.
+\url{https://doi.org/10.1093/ser/mwx045}
+
+\bibitem[\citeproctext]{ref-lane_market_1986}
+Lane, R. E. (1986). Market {Justice}, {Political Justice}.
+\emph{American Political Science Review}, \emph{80}(2), 383--402.
+\url{https://doi.org/10.2307/1958264}
+
+\bibitem[\citeproctext]{ref-langsaether_explaining_2022}
+Langsæther, P. E., Evans, G., \& O'Grady, T. (2022). Explaining the
+{Relationship Between Class Position} and {Political Preferences}: {A
+Long-Term Panel Analysis} of {Intra-Generational Class Mobility}.
+\emph{British Journal of Political Science}, \emph{52}(2), 958--967.
+\url{https://doi.org/10.1017/S0007123420000599}
+
+\bibitem[\citeproctext]{ref-lee_fairness_2023}
+Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
+inequality: The effects of self-interest and neoliberal orientations.
+\emph{The Australian Educational Researcher}.
+\url{https://doi.org/10.1007/s13384-023-00636-6}
+
+\bibitem[\citeproctext]{ref-lenski_status_1954}
+Lenski, G. E. (1954). Status {Crystallization}: {A Non-Vertical
+Dimension} of {Social Status}. \emph{American Sociological Review},
+\emph{19}(4), 405. \url{https://doi.org/10.2307/2087459}
+
+\bibitem[\citeproctext]{ref-lindh_public_2015}
+Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
+towards {Market Distribution} of {Social Services} -- {A Comparison} of
+17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
+887--910. \url{https://doi.org/10.1111/spol.12105}
+
+\bibitem[\citeproctext]{ref-lindh_bringing_2023}
+Lindh, A., \& McCall, L. (2023). Bringing the market in: An expanded
+framework for understanding popular responses to economic inequality.
+\emph{Socio-Economic Review}, \emph{21}(2), 1035--1055.
+\url{https://doi.org/10.1093/ser/mwac018}
+
+\bibitem[\citeproctext]{ref-llorca-jana_historia_2021}
+Llorca-Jaña, M., \& Miller, R. M. D. (2021). \emph{{Historia económica
+de Chile desde la independencia}}. Santiago de Chile: RIL editores.
+
+\bibitem[\citeproctext]{ref-lopez-roldan_comparative_2021}
+López-Roldán, P., \& Fachelli, S. (Eds.). (2021). \emph{Towards a
+{Comparative Analysis} of {Social Inequalities} between {Europe} and
+{Latin America}}. Cham: Springer International Publishing.
+\url{https://doi.org/10.1007/978-3-030-48442-2}
+
+\bibitem[\citeproctext]{ref-mac-clure_justicia_2024}
+Mac-Clure, O., Barozet, E., \& Franetovic, G. (2024). {Justicia
+distributiva y posición social subjetiva: \textquestiondown la
+meritocracia justifica la desigualdad de ingresos?} \emph{Convergencia
+Revista de Ciencias Sociales}, \emph{31}, 1.
+\url{https://doi.org/10.29101/crcs.v31i0.22258}
+
+\bibitem[\citeproctext]{ref-madariaga_three_2020}
+Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
+economic policy trajectory in comparative perspective.
+\emph{Contemporary Politics}, \emph{26}(3), 308--329.
+\url{https://doi.org/10.1080/13569775.2020.1735021}
+
+\bibitem[\citeproctext]{ref-mau_inequality_2015}
+Mau, S. (2015). \emph{Inequality, {Marketization} and the {Majority
+Class}: {Why Did} the {European Middle Classes Accept Neo-Liberalism}?}
+Houndmills: Palgrave Macmillan.
+
+\bibitem[\citeproctext]{ref-mideplan_evolucion_2017}
+MIDEPLAN. (2017). \emph{Evolución de la pobreza 1990-2017}. Ministerio
+de Desarrollo Social y Familia - PNUD.
+
+\bibitem[\citeproctext]{ref-mideplan_estadisticas_2024}
+MIDEPLAN. (2024). \emph{Estadísticas {Encuesta Casen} 2022, sección
+{Salud}}. Observatorio social.
+
+\bibitem[\citeproctext]{ref-mijs_paradox_2019}
+Mijs, J. (2019). The paradox of inequality: Income inequality and belief
+in meritocracy go hand in hand. \emph{Socio-Economic Review},
+\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
+
+\bibitem[\citeproctext]{ref-mijs_belief_2022}
+Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
+Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
+{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
+\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}
+
+\bibitem[\citeproctext]{ref-miller_selfserving_1975}
+Miller, D. T., \& Ross, M. (1975). Self-serving biases in the
+attribution of causality: {Fact} or fiction? \emph{Psychological
+Bulletin}, \emph{82}(2), 213--225.
+\url{https://doi.org/10.1037/h0076486}
+
+\bibitem[\citeproctext]{ref-ministeriodeeducacion_resumen_2023}
+Ministerio de Educación. (2023). \emph{Resumen de matrícula por
+establecimiento educacional}. Datos abiertos Ministerio de Educación.
+
+\bibitem[\citeproctext]{ref-molina_its_2019}
+Molina, M. D., Bucca, M., \& Macy, M. W. (2019). It's not just how the
+game is played, it's whether you win or lose. \emph{SCIENCE ADVANCES}.
+
+\bibitem[\citeproctext]{ref-nucleodesociologiacontingente_informe_2020}
+Núcleo de Sociología Contingente, {[}NUDESOC{]}. (2020). \emph{Informe
+de resultados oficial {Encuesta Zona Cero}}. Santiago de Chile.
+
+\bibitem[\citeproctext]{ref-otero_power_2024}
+Otero, G., \& Mendoza, M. (2024). The {Power} of {Diversity}: {Class},
+{Networks} and {Attitudes Towards Inequality}. \emph{Sociology},
+\emph{58}(4), 851--876. \url{https://doi.org/10.1177/00380385231217625}
+
+\bibitem[\citeproctext]{ref-otero_lives_2022}
+Otero, G., Volker, B., Rözer, J., \& Mollenhorst, G. (2022). The lives
+of others: {Class} divisions, network segregation, and attachment to
+society in {Chile}. \emph{The British Journal of Sociology},
+\emph{73}(4), 754--785. \url{https://doi.org/10.1111/1468-4446.12966}
+
+\bibitem[\citeproctext]{ref-panes_criticas_2020}
+Panes, D. (2020). \emph{Críticas y experiencias obreras en torno al
+sistema de {AFP}'s en {Chile} (1981-2020)} (PhD thesis). Universidad de
+Chile, Santiago.
+
+\bibitem[\citeproctext]{ref-perez-ahumada_clases_2019}
+Pérez-Ahumada, P. (2019). {Clases sociales, sectores económicos y
+cambios en la estructura social chilena entre 1992 y 2013}.
+\emph{Revista de la CEPAL}, \emph{2018}(126), 171--192.
+\url{https://doi.org/10.18356/0ce0faed-es}
+
+\bibitem[\citeproctext]{ref-pnud_desiguales_2017}
+PNUD (Ed.). (2017). \emph{{Desiguales: orígenes, cambios y desafíos de
+la brecha social en Chile}}. Santiago, Chile: PNUD : Uqbar Editores.
+
+\bibitem[\citeproctext]{ref-polanyi_great_1975}
+Polanyi, K. (1975). \emph{The great transformation} (Repr). New York,
+NY: Octagon Books.
+
+\bibitem[\citeproctext]{ref-prag_subjective_2021}
+Präg, P., \& Gugushvili, A. (2021). Subjective social mobility and
+health in {Germany}. \emph{European Societies}, \emph{23}(4), 464--486.
+\url{https://doi.org/10.1080/14616696.2021.1887916}
+
+\bibitem[\citeproctext]{ref-ruiz_chilenos_2014}
+Ruiz, C., \& Boccardo, G. (2014). \emph{Los chilenos bajo el
+neoliberalismo: Clases y conflicto social} (Primera edición). Santiago
+de Chile: Ediciones y Publicaciones El Buen Aire.
+
+\bibitem[\citeproctext]{ref-salgado_uplifting_2025}
+Salgado, M., Díaz, M., Gamarra, C., \& Núñez, J. (2025). Uplifting
+mechanism or equalizing force? {The} expansion of higher education and
+its role in intergenerational mobility in {Chile}. \emph{Higher
+Education}. \url{https://doi.org/10.1007/s10734-025-01560-7}
+
+\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
+Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
+common good?} (First edition). New York: {Farrar, Straus and Giroux}.
+
+\bibitem[\citeproctext]{ref-satz_por_2019}
+Satz, D. (2019). \emph{{Por Qué Algunas Cosas No Deberían Estar en
+Venta: Los límites Morales Del Mercado}}. Ciudad Autónoma de Buenos
+Aires: Siglo XXI Editores.
+
+\bibitem[\citeproctext]{ref-schmidt_experience_2011}
+Schmidt, A. W. (2011). The experience of social mobility and the
+formation of attitudes toward redistribution. In.
+
+\bibitem[\citeproctext]{ref-shariff_income_2016}
+Shariff, A. F., Wiwad, D., \& Aknin, L. B. (2016). Income {Mobility
+Breeds Tolerance} for {Income Inequality}: {Cross-National} and
+{Experimental Evidence}. \emph{Perspectives on Psychological Science},
+\emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}
+
+\bibitem[\citeproctext]{ref-somma_no_2021}
+Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
+water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
+Movement Studies}, \emph{20}(4), 495--502.
+\url{https://doi.org/10.1080/14742837.2020.1727737}
+
+\bibitem[\citeproctext]{ref-song_there_2025}
+Song, X., \& Zhou, X. (2025). Is {There} a {Mobility Effect}? {On
+Methodological Issues} in the {Mobility Contrast Model}.
+\emph{Sociological Methods \& Research}, \emph{54}(4), 1576--1593.
+\url{https://doi.org/10.1177/00491241251347983}
+
+\bibitem[\citeproctext]{ref-sorokin_social_1927}
+Sorokin, P. A. (1927). \emph{Social {Mobility}}. Harper \& Brothers.
+
+\bibitem[\citeproctext]{ref-streeck_how_2016}
+Streeck, W. (2016). \emph{How will capitalism end? Essays on a failing
+system}. London: Verso.
+
+\bibitem[\citeproctext]{ref-superintendenciadepensiones_estadisticas_2024}
+Superintendencia de Pensiones. (2024). \emph{Estadísticas de {Afiliados}
+en {AFP} en {Sistema} de {Pensiones}. {Abril} 2024}. Superintendencia de
+Pensiones.
+
+\bibitem[\citeproctext]{ref-svallforsMoralEconomyClass2006a}
+Svallfors, S. (2006). \emph{The {Moral Economy} of {Class}: {Class} and
+{Attitudes} in {Comparative Perspective}}. Stanford University Press.
+
+\bibitem[\citeproctext]{ref-svallfors_political_2007}
+Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
+{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
+(1st ed.). Stanford University Press.
+\url{https://doi.org/10.2307/j.ctvr0qv0q}
+
+\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
+Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
+Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
+{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
+\emph{International Review of Social Psychology}, \emph{38}(1), 4.
+\url{https://doi.org/10.5334/irsp.1013}
+
+\bibitem[\citeproctext]{ref-torche_unequal_2005}
+Torche, F. (2005). Unequal {But Fluid}: {Social Mobility} in {Chile} in
+{Comparative Perspective}. \emph{American Sociological Review},
+\emph{70}(3), 422--450. \url{https://doi.org/10.1177/000312240507000304}
+
+\bibitem[\citeproctext]{ref-torche_intergenerational_2014}
+Torche, F. (2014). Intergenerational {Mobility} and {Inequality}: {The
+Latin American Case}. \emph{Annual Review of Sociology}, \emph{40}(1),
+619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
+
+\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
+Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
+care inequalities unfair? {A} study on public attitudes in 23 countries.
+\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
+\url{https://doi.org/10.1186/s12939-016-0350-8}
+
+\bibitem[\citeproctext]{ref-wen_does_2021}
+Wen, F., \& Witteveen, D. (2021). Does perceived social mobility shape
+attitudes toward government and family educational investment?
+\emph{Social Science Research}, \emph{98}, 102579.
+\url{https://doi.org/10.1016/j.ssresearch.2021.102579}
+
+\bibitem[\citeproctext]{ref-wiederkehr_belief_2015}
+Wiederkehr, V., Bonnot, V., Krauth-Gruber, S., \& Darnon, C. (2015).
+Belief in school meritocracy as a system-justifying tool for low status
+students. \emph{Frontiers in Psychology}, \emph{6}.
+
+\bibitem[\citeproctext]{ref-wilson_role_2003}
+Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
+Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
+\url{https://doi.org/10.1023/A:1024667228488}
+
+\bibitem[\citeproctext]{ref-wooldridge_introductory_2009}
+Wooldridge, J. M. (2009). \emph{{Introductory econometrics: a modern
+approach}} (4th ed). Mason, OH: South Western, Cengage Learning.
+
+\bibitem[\citeproctext]{ref-wrightUnderstandingClass2015}
+Wright, E. O. (2015). \emph{Understanding {Class}} (Verso).
+
+\bibitem[\citeproctext]{ref-young_rise_1958}
+Young, M. (1958). \emph{The rise of the meritocracy}. New Brunswick,
+N.J., U.S.A: Transaction Publishers.
+
+\end{CSLReferences}
+
+\section{Supplementary material}\label{anexo}
+
+This section presents the supplementary material for this study.
+
+\subsection{Descriptive statistics}\label{descriptive-statistics-1}
 
@@
 
-\subsubsection{Treatment}\label{treatment}
-
-\paragraph{Intergenerational occupational
-mobility}\label{intergenerational-occupational-mobility}
-
-I treat intergenerational occupational mobility as an exposure
-indicating whether respondents occupy a different socioeconomic status
-than their fathers, closely following Breen and Ermisch's
-(\citeproc{ref-breen_effects_2024}{2024}) causal framework. Class
-assignment proceeds in two steps. First, I derive status for both origin
-(father) and destination (respondent) from two-digit ISCO-08 occupations
-using the International Socio-Economic Index of Occupational Status
-(ISEI). ISEI is a hierarchical, unidimensional scale that locates
-occupations by incumbents' average education and earnings and was
-explicitly designed to harmonize occupational stratification across
-other class schemes and socioeconomic status measures
-(\citeproc{ref-ganzeboom_internationally_1996}{Ganzeboom \& Treiman,
-1996}), thereby supporting cross-study comparability
-(\citeproc{ref-barozet_measurement_2021a}{Barozet, Boado, et al.,
-2021}). Second, I group both origin and destination ISEI scores into
-terciles (low, middle, high), yielding a three-category schema for
-origin and destination stratums (see Table~\ref{tbl-matrix}).
-Substantively, I use ISEI because occupation-based measures capture
-structural locations that organize life chances and attitudes more
-directly than income alone: they bundle access to the means of
-production, qualification credentials, and the authority/power inherent
-in labor relations; central dimensions of stratification and inequality
-(\citeproc{ref-barone_rise_2022}{Barone et al., 2022};
-\citeproc{ref-wrightUnderstandingClass2015}{Wright, 2015}). In short,
-using ISEI aligns the measurement of origin and destination on a common
-vertical status continuum that is theoretically meaningful for social
-mobility research and empirically feasible given the available
-information of ELSOC's data.
-
-\begin{table}
-
-\caption{\label{tbl-matrix}Occupational mobility by occupational
-groups.}
+\subsection{Balance evaluation}\label{balance-evaluation}
+
+\begin{figure}[H]
 
@@
 
-\centering
-\begin{tabular}{cccccc}
-\toprule
-Father↓ & Offspring→ & Low & Middle & High & Total\\
-\midrule
-Low &  & 44.7\%   (501) & 33.6\%   (377) & 21.7\%   (244) & 100.0\% (1,122)\\
-Middle &  & 35.1\%   (388) & 34.9\%   (385) & 30.0\%   (331) & 100.0\% (1,104)\\
-High &  & 24.6\%   (297) & 28.8\%   (348) & 46.7\%   (564) & 100.0\% (1,209)\\
-Total &  & 34.5\% (1,186) & 32.3\% (1,110) & 33.2\% (1,139) & 100.0\% (3,435)\\
-\bottomrule
-\end{tabular}
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-balance-1.pdf}
 
@@
 
-\end{table}%
-
-\paragraph{Pre-treatment control variables for selection into
-treatment}\label{pre-treatment-control-variables-for-selection-into-treatment}
-
-Within each origin stratum \(O=j\), I estimate average treatment effects
-on the treated (ATT) by comparing movers to otherwise similar non-movers
-from the same origin. To adjust for selection into mobility, I reweight
-only the control group (immobile) using entropy balancing so that the
-distribution of pre-treatment origin covariates \(C\) matches that of
-the treated on selected moments.
-
-The selection of the adjustment set \(C\) follows two principles. First,
-it mirrors the logic in Breen and Ermisch
-(\citeproc{ref-breen_effects_2024}{2024})---who condition on attributes
-determined in childhood or earlier (e.g., parental resources, household
-structure, ascriptive traits)---and, second, it is constrained to
-origin-side characteristics available in ELSOC. Guided by theory and
-Chilean evidence on relative mobility determinants (mainly
-\citeproc{ref-brunori_inequality_2025}{Brunori et al., 2025};
-\citeproc{ref-celhay_intergenerational_2010}{Celhay et al., 2010};
-\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
-\citeproc{ref-torche_unequal_2005}{Torche, 2005}), I include six
-covariates covering family resources, household composition, and
-ascriptive status: (a) parental education (highest of father/mother), 10
-categories coded ordinally from no schooling to postgraduate studies;
-(b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c)
-nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0
-= male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
-These variables are fixed before the destination class, plausibly affect
-selection into mobility, and constitute the maximal origin set in ELSOC,
-thereby making the conditional independence assumption more credible
-within each \(O=j\).
-
-After balancing, I construct stabilized IPW, setting treated weights to
-1 and assigning derived weights to controls, thus aligning the
-counterfactual distribution of non-movers with the covariate profile of
-movers in each origin stratum (see \hyperref[anexo]{Supplementary
-Material} for the evaluation of balance). These weights are then used in
-the outcome stage to estimate the causal effect of mobility on
-preferences for the commodification of pensions.
-
-\subsubsection{Effect heterogeneity}\label{effect-heterogeneity}
-
-To test heterogeneity in mobility effects, I examine two moderators:
-meritocratic perceptions and time. Meritocracy is captured with two
-items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
-effort (``In Chile, people are rewarded for their efforts'') and one to
-talent (``In Chile, people are rewarded for their intelligence and
-skills''), each answered on a five-point Likert scale (1 = ``strongly
-disagree'' to 5 = ``strongly agree''). I average the two items into a
-single index and dichotomize it into low meritocracy (≤ 3) and high
-meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
-and entered as a categorical (dummy) variable. Both variables are used
-as moderators by interacting them with the mobility treatment to assess
-conditional average treatment effects---i.e., whether the effect of
-mobility varies across levels of meritocratic perceptions or across
-survey waves.
-
-\subsubsection{Controls}\label{controls}
-
-All models include the same pre-treatment covariates \(C\) used in the
-IPW construction, not to re-balance groups (entropy balancing already
-does so), but to (i) block backdoor paths from \(C\) to the outcome
-\(Y\) as implied by the DAG, and (ii) achieve double robustness:
-estimates remain consistent if either the weighting model or the outcome
-model is correctly specified. Concretely, I adjust for (a) father's
-educational level, (b) co-residence with both parents at age 15, (c)
-nationality, (d) age, (e) sex, and (f) ethnicity, along with wave fixed
-effects to absorb secular trends.
-
-\subsection{Analytical strategy}\label{analytical-strategy}
-
-Because the dependent variable---the preference for pension
-commodification---is binary \((Y_{it}\in{0,1})\), I estimate weighted
-linear probability models (WLS) using stabilized inverse probability
-weights (IPW) obtained via entropy balancing. Although OLS estimator is
-often questioned with binary outcomes, it's unbiased. It consistently
-estimates the conditional expectation \(E(Y|X)\) under standard
-exogeneity \(E[\varepsilon_i|X_i]=0\), with heteroskedasticity addressed
-by robust or clustered errors
-(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
-ch.~7.5). WLS further improves efficiency under heteroskedasticity or
-unequal sampling probabilities
-(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009},
-ch.~8.4). Following Gelman, Hill, and Vehtari
-(\citeproc{ref-gelman_regression_2020}{2020, pp. 270--272}),
-IPW-weighted regression transparently targets the ATT by recreating the
-counterfactual distribution. Linear models are preferred here because
-coefficients are average probability differences; nonlinear models
-(logit/probit) complicate weighting and interpretation
-(\citeproc{ref-gelman_regression_2020}{Gelman et al., 2020}, ch.~13).
-
-Formally, the model is:
-
-\begin{equation}\phantomsection\label{eq-final}{
-Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
-}\end{equation}
-
-where \(Y_{it}\) indicates whether individual \(i\) in wave \(t\)
-supports more market-based pension access; \(\alpha\) is the baseline
-probability for the reference categories; \(\beta T_i\) captures the
-intergenerational mobility contrast: \(T_i=1\) when the observed
-destination is \(D_i=k\) (mobile to \(k\)) and \(T_i=0\) when \(D_i=j\)
-(immobile), so that \(\beta\) is the ATT within origin
-\(O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})\), i.e., the
-percentage-point change in the probability of preferring pension
-commodification from reaching \(k\) rather than remaining in \(j\). The
-term \(X_i'\gamma\) includes the pre-treatment covariates \((C)\) used
-to construct the entropy-balancing weights; \(\lambda_t\) are wave fixed
-effects; and \(\varepsilon_{it}\) is idiosyncratic error. The analytic
-sample is restricted to individuals sharing the same origin \((O=j)\).
-Therefore, estimation uses pooled WLS with stabilized IPW--ATT from
-entropy balancing and CR2 standard errors clustered by individual to
-address heteroskedasticity and within-person dependence.
-
-The specification is doubly robust: entropy-balancing weights are
-combined with covariate adjustment so consistency holds if either the
-weighting model or the outcome model is correctly specified, while also
-improving efficiency
-(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
-results are virtually identical to baseline models without covariates
-(see \hyperref[anexo]{Supplementary Material} for complete models), with
-no substantive changes in the mobility coefficients. In addition, I
-assess causal heterogeneity by interacting the mobility treatment with
-(i) meritocratic perceptions and (ii) survey wave, reporting simple
-slopes (origin-specific ATTs) by moderator levels. To evaluate
-sensitivity to unmeasured confounding, I compute E-values for point
-estimates and confidence limits
-(\citeproc{ref-ding_sensitivity_2016}{Ding \& VanderWeele, 2016}). To
-address multiplicity across the six mobility trajectories, I also report
-Holm-adjusted p-values alongside 95\% confidence intervals.
-
-All the analyses were conducted using R software and the
-\emph{lm\_robust} package.
-
-\section{Results}\label{results}
-
-\subsection{Descriptive statistics}\label{descriptive-statistics}
-
-Figure~\ref{fig-alluvial} shows the annual frequencies of preferences
-for pension commodification in 2016, 2018, and 2023. Each year presents
-stacked percentages frequencies for the level of agreement and
-disagreement. Overall, a large majority rejects market-based access to
-old-age pension benefits, but this opposition has eased over time:
-83.5\% in 2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018
-change is modest, whereas 2023 registers a 9.4 point drop in
-disagreement relative to 2018, mirrored by rising agreement: 16.5\%
-(2016), 18.8\% (2018), 28.2\% (2023). Substantively, while most
-respondents continue to oppose the idea that higher-income individuals
-should obtain better pensions via the market, a non-trivial and growing
-portion endorses this statement, with the sharpest expansion
-concentrated in the latest wave (+9.4 points from 2018 to 2023).
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-alluvial-1.pdf}
-
-}
-
-\caption{\label{fig-alluvial}Change in preferences for pension
-commodification over time (2016, 2018, and 2023).}
+\caption{\label{fig-balance}Balance SMD --- Mobility treatment (ATT)}
 
@@
 
-Regarding the relationship between preferences for the commodification
-of pensions and intergenerational occupational mobility,
-Figure~\ref{fig-mean} shows the percentage of agreement with
-market-based access to pension benefits according to mobility
-trajectories and survey waves. Averaging across waves, agreement is
-highest among the High-High immobile (27.8\%), followed by the
-Middle-High upwardly mobile (24.6\%) and the High-Middle downwardly
-mobile (24.0\%). The lowest levels are found among ascending Low-High
-(17.3\%), and immobile Low-Low (17.4\%) groups. The specific profiles of
-each wave accentuate these gradients in most trajectories, especially in
-2023: agreement reaches 31.9\% for High-High, 29.4\% for upward
-Middle-High, and peaks at 36.8\% for downward High-Middle (the highest
-of all trajectories for that year). Descriptively, immobility at the top
-is associated with greater support for commodification; among those who
-move, support is greatest for the upward Middle-High trajectory and the
-downward High-Middle trajectory, patterns that intensify in 2023. These
-shifts anticipate the heterogeneity documented below.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-mean-1.pdf}
-
-}
-
-\caption{\label{fig-mean}Percentage agreement on preferences for pension
-commodification according to mobility trajectory and time (2016, 2018,
-and 2023).}
-
-\end{figure}%
-
-\subsection{Models}\label{models}
-
-Figure~\ref{fig-reg} reports the double-robust estimates of the effect
-of intergenerational occupational mobility on preferences for pension
-commodification, obtained from weighted models that combine stabilized
-IPW with covariate adjustment and fixed wave effects. Results reveal a
-clear directional asymmetry between upward and downward mobility
-trajectories. Among upwardly mobile respondents, only the Middle→High
-trajectory exhibits a significant positive effect on support for pension
-commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
-Interpreted as an origin-specific ATT, this estimate compares---for
-individuals who actually moved from the middle to the high socioeconomic
-status---their observed endorsement with the counterfactual endorsement
-they would have expressed had they remained immobile in the middle
-socioeconomic status. Holding all other predictors constant, the point
-estimate implies a 9 percentage point higher probability of endorsing
-market-based access to pension benefits for Middle→High movers relative
-to their own immobility counterfactual, statistically significant at the
-95\% confidence level. In contrast, two downward trajectories:
-High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
-(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
-effects, suggesting that individuals who experience downward mobility
-from higher-status origins express weaker preferences for market-based
-pension benefits. The remaining pathways (Low→Middle, Low→High,
-Middle→Low) display no significant differences relative to their
-non-mobile counterparts. These patterns indicate that upward movement
-within the upper segment (Middle→High) reinforces pro-commodification
-attitudes, whereas downward movement from privileged origins reduces
-them.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-reg-1.pdf}
-
-}
-
-\caption{\label{fig-reg}Effects of intergenerational occupational
-mobility on preferences for pension commodification. Coefficient plot of
-origin-specific ATT estimates comparing mobility versus immobility.
-Estimates from pooled WLS with doubly robust adjustment, wave fixed
-effects, and standard errors clustered by individual; bars show 95\%
-confidence intervals.}
-
-\end{figure}%
-
-\subsection{Effect heterogeneity}\label{effect-heterogeneity-1}
-
-To probe potential mechanisms underlying the mobility effects, I
-estimate origin-specific ATT models that interact the mobility treatment
-with high meritocratic perceptions (ref. = low), using stabilized IPW
-weights, covariate adjustment, and wave fixed effects. I also estimate
-an analogous specification interacting mobility with survey waves (ref.
-= 2016) to assess temporal heterogeneity in the treatment effect (see
-\hyperref[anexo]{Supplementary Material} for complete models).
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-1.pdf}
-
-}
-
-\caption{\label{fig-interact}Effects of intergenerational occupational
-mobility on preferences for pension commodification, by meritocratic
-perception. Coefficient plot of origin-specific ATT estimates comparing
-mobility versus immobility, estimated from interaction models (T ×
-Merit). Points show simple slopes of the mobility effect at low merit
-and high merit; bars are 95\% confidence intervals. Estimates from
-pooled WLS with doubly robust adjustment, wave fixed effects, and
-standard errors clustered by individual.}
-
-\end{figure}%
-
-Figure~\ref{fig-interact} shows that neither the interaction terms nor
-the mobility coefficients reach conventional significance across
-trajectories, indicating no detectable moderation of the causal mobility
-effect by meritocratic perceptions at the 95\% level. Directionally, in
-the left panel, the interaction is positive for Middle→High (\(\beta\) =
-0.15, 95\% CI {[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05,
-{[}−0.14, 0.24{]}), suggesting, if anything, somewhat larger
-pro-commodification effects among high-merit respondents when upwardly
-mobile, whereas it is negative for High→Low (\(\beta\) = −0.14,
-{[}−0.38, 0.10{]}) and High→Middle (\(\beta\) = −0.03, {[}−0.23,
-0.17{]}), showing weaker pro-commodification effects under downward
-mobility at high merit. Focusing on the treatment effect at low
-meritocracy (the reference category) at the right panel, the mobility
-coefficient captures the origin-specific ATT comparing movers to their
-immobility counterfactual among low-merit respondents. These effects are
-small and imprecise across trajectories: Low→Middle (\(\beta\) = −0.02,
-{[}−0.09, 0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]})
-are near zero. Middle→High is positive but not significant (\(\beta\) =
-0.07, {[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00,
-{[}−0.07, 0.06{]}), and downward from high shows negative, borderline
-estimates: High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and
-High→Low (\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not
-find evidence that meritocratic perceptions moderate the causal effect
-of mobility, although the signs align with the idea that upward mobility
-at high merit tilts attitudes toward commodification, and downward
-mobility at high merit tilts them away.
-
-\begin{table}
-
-\caption{\label{tbl-inttime}Effects of intergenerational occupational
-mobility on preferences for pension commodification, by survey wave}
-
-\centering{
-
-\begin{center}
-\scalebox{0.8}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
-\hline
-Intercept                              & $0.18$           & $-0.04$          & $0.28$           & $0.20$           & $0.18$            & $0.15$            \\
-                                       & $ [-0.36; 0.73]$ & $ [-0.36; 0.29]$ & $ [-0.01; 0.57]$ & $ [-0.11; 0.52]$ & $ [-0.15;  0.50]$ & $ [-0.27;  0.58]$ \\
-Mobility treatment                     & $-0.05$          & $-0.03$          & $0.13^{*}$       & $0.06$           & $-0.19^{*}$       & $-0.19^{*}$       \\
-                                       & $ [-0.16; 0.06]$ & $ [-0.15; 0.10]$ & $ [ 0.05; 0.22]$ & $ [-0.03; 0.15]$ & $ [-0.33; -0.05]$ & $ [-0.32; -0.05]$ \\
-Wave (Ref.= 2016)                      &                  &                  &                  &                  &                   &                   \\
-                                       &                  &                  &                  &                  &                   &                   \\
-\quad Wave 2018                        & $0.01$           & $0.01$           & $0.10^{*}$       & $0.09^{*}$       & $-0.04$           & $-0.02$           \\
-                                       & $ [-0.10; 0.12]$ & $ [-0.10; 0.12]$ & $ [ 0.03; 0.18]$ & $ [ 0.01; 0.17]$ & $ [-0.15;  0.07]$ & $ [-0.15;  0.10]$ \\
-\quad Wave 2023                        & $0.03$           & $0.01$           & $0.19^{*}$       & $0.18^{*}$       & $0.05$            & $0.04$            \\
-                                       & $ [-0.09; 0.14]$ & $ [-0.12; 0.13]$ & $ [ 0.10; 0.28]$ & $ [ 0.08; 0.27]$ & $ [-0.09;  0.18]$ & $ [-0.11;  0.19]$ \\
-Mobility treatment x Wave (Ref.= 2016) &                  &                  &                  &                  &                   &                   \\
-                                       &                  &                  &                  &                  &                   &                   \\
-\quad Mobility treatment x Wave 2018   & $0.03$           & $-0.06$          & $-0.04$          & $-0.09$          & $0.11$            & $0.06$            \\
-                                       & $ [-0.11; 0.17]$ & $ [-0.21; 0.09]$ & $ [-0.17; 0.09]$ & $ [-0.20; 0.03]$ & $ [-0.04;  0.26]$ & $ [-0.09;  0.22]$ \\
-\quad Mobility treatment x Wave 2023   & $0.07$           & $0.10$           & $-0.09$          & $-0.07$          & $0.19^{*}$        & $0.18$            \\
-                                       & $ [-0.09; 0.23]$ & $ [-0.08; 0.28]$ & $ [-0.23; 0.06]$ & $ [-0.22; 0.08]$ & $ [ 0.01;  0.38]$ & $ [-0.01;  0.37]$ \\
-\hline
-Controls                               & Yes              & Yes              & Yes              & Yes              & Yes               & Yes               \\
-R$^2$                                  & $0.04$           & $0.05$           & $0.07$           & $0.03$           & $0.06$            & $0.06$            \\
-Adj. R$^2$                             & $0.03$           & $0.04$           & $0.06$           & $0.02$           & $0.05$            & $0.05$            \\
-Num. obs.                              & $878$            & $745$            & $716$            & $773$            & $912$             & $861$             \\
-RMSE                                   & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$            & $0.42$            \\
-N Clusters                             & $497$            & $449$            & $443$            & $470$            & $536$             & $525$             \\
-\hline
-\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
-\end{tabular}
-}
-\label{table:coefficients}
-\end{center}
-
-}
-
-\end{table}%
-
-Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
-effect across the different waves of the survey. The treatment effect in
-2016 (reference category) closely replicates the reference ATT patterns:
-positive for Middle→High (\(\beta\) = 0.13, 95\% CI {[}0.05, 0.22{]})
-and negative for High→Middle (\(\beta\) = −0.19, {[}−0.33, −0.05{]}) and
-High→Low (\(\beta\) = −0.19, {[}−0.32, −0.05{]}). This indicates that,
-holding all other predictors constant, individuals who moved from middle
-to high socioeconomic status in 2016 were, on average, 13 percentage
-points more likely to support market-based access to the pension
-benefits compared to their counterfactual of immobility, while those who
-moved down from higher status were approximately 19 percentage points
-less likely. Beyond these reference effects, there is a notable temporal
-change in the High→Middle trajectory: the interaction between mobility
-and the 2023 wave is positive and statistically significant (\(\beta\) =
-0.19, {[}0.01, 0.38{]}). This suggests that, although people with
-downward mobility from high backgrounds were less supportive of
-commodification in 2016, by 2023 the same trajectory shows a change in
-sign, implying that the negative causal effect of downward mobility
-attenuated and eventually turned positive. No comparable changes are
-observed in other trajectories. In essence, this pattern reveals
-time-conditional heterogeneity in the mobility effect: among those who
-experienced downward mobility, preferences for market-based access to
-old-age pensions appear to have realigned in the most recent wave,
-reflecting a notable shift toward greater support for commodification
-within this trajectory by 2023.
-
-\subsection{Sensitivity analysis}\label{sensitivity-analysis}
-
-The first sensitivity check corresponds to the Ding \& VanderWeele
-(\citeproc{ref-ding_sensitivity_2016}{2016}) approach to assess
-unmeasured confounding (E values). This values indicates the minimum
-strength of association that an unobserved confounder would need to have
-with both the treatment and the outcome---on the relative risk
-scale---to explain an effect. The Middle-High ATT would require an
-\(RR ≈ 1.74\) to offset the point estimate (\(E_{point}\) = 1.736) and
-an \(RR ≈ 1.26\) to move its CI to include the null value (\(E_{CI}\) =
-1.264). High→Low shows comparable robustness (\(E_{point}\) = 1.759;
-\(E_{CI}\) = 1.239), while High→Middle is comparatively fragile near
-zero (\(E_{point}\) = 1.591; \(E_{CI}\) = 1.037). For non-significant
-trajectories (Low→Middle, Low→High, Middle→Low), \(E_{CI}\) = 1, as
-expected. In conclusion, the directional pattern---stronger in favor of
-commodification in Middle→High and weaker in the descent from
-High---seems reasonably robust to moderate unmeasured confounding in
-Middle→High and High→Low, but borderline in High→Middle. These results
-justify treating signs and relative magnitudes as informative, while
-recognizing limited protection against omitted variables for High→Middle
-(see \hyperref[anexo]{Supplementary Material} for complete tables).
-
-The second sensitivity check applies Holm's step-down correction, which
-is a multiple-testing procedure that controls the familywise error rate
-(FWER) while retaining more power than Bonferroni, across the six
-mobility trajectories. After FWER control, none of the effects remain
-significant at \(\alpha\) = .05: Middle→High (\(p_{raw}\) = .0116 →
-\(p_{Holm}\) = .0698), High→Low (.0167 → .0834), and High→Middle (.0481
-→ .1924). Accordingly, claims of statistical significance should be
-tempered under strict multiplicity control. Substantively, however, this
-adjustment affects the strength of inference, not its direction: the
-pattern remains consistent with the theorized asymmetry (positive for
-Middle→High; negative for downward from high). I therefore treat signs
-and effect sizes as the primary evidence and regard p-values as
-exploratory once multiplicity is accounted for (see
-\hyperref[anexo]{Supplementary Material} for complete tables).
-
-\section{Discussion}\label{discussion}
-
-\section{Conclusion}\label{conclusion}
-
-\section{References}\label{references}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-alesina_intergenerational_2018}
-Alesina, A., Stantcheva, S., \& Teso, E. (2018). Intergenerational
-{Mobility} and {Preferences} for {Redistribution}. \emph{American
-Economic Review}, \emph{108}(2), 521--554.
-\url{https://doi.org/10.1257/aer.20162015}
-
-\bibitem[\citeproctext]{ref-araujo_desafios_2012}
-Araujo, K., \& Martuccelli, D. (2012). \emph{Desaf{í}os comunes: Retrato
-de la sociedad chilena y sus individuos} (1a. ed). Santiago: LOM
-Ediciones.
-
-\bibitem[\citeproctext]{ref-ares_changing_2020}
-Ares, M. (2020). Changing classes, changing preferences: How social
-class mobility affects economic preferences. \emph{West European
-Politics}, \emph{43}(6), 1211--1237.
-\url{https://doi.org/10.1080/01402382.2019.1644575}
-
-\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
-Arrizabalo, X. (1995). \emph{{Milagro o quimera: la econom{í}a chilena
-durante la dictadura}}. Libros de la Catarata.
-
-\bibitem[\citeproctext]{ref-barone_rise_2022}
-Barone, C., Hertel, F. R., \& Smallenbroek, O. (2022). The rise of
-income and the demise of class and social status? {A} systematic review
-of measures of socio-economic position in stratification research.
-\emph{Research in Social Stratification and Mobility}, \emph{78},
-100678. \url{https://doi.org/10.1016/j.rssm.2022.100678}
-
-\bibitem[\citeproctext]{ref-barozet_measurement_2021a}
-Barozet, E., Boado, M., \& Marqués-Perales, I. (2021). The {Measurement}
-of {Social Stratification}: {Comparative Perspectives Between Europe}
-and {Latin America}. In P. López-Roldán \& S. Fachelli (Eds.),
-\emph{Towards a {Comparative Analysis} of {Social Inequalities} between
-{Europe} and {Latin America}} (pp. 171--202). Cham: Springer
-International Publishing.
-\url{https://doi.org/10.1007/978-3-030-48442-2_6}
-
-\bibitem[\citeproctext]{ref-barozet_clases_2021}
-Barozet, E., Contreras, D., Espinoza, V., Gayo, M., \& Méndez, M. L.
-(2021). \emph{{Clases medias en tiempos de crisis: vulnerabilidad
-persistente, desaf{í}os para la cohesi{ó}n y un nuevo pacto social en
-Chile}}. Santiago: Comisi{ó}n Econ{ó}mica para Am{é}rica Latina y el
-Caribe (CEPAL).
-
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023). Belief in {School Meritocracy} and the {Legitimization} of
-{Social} and {Income Inequality}. \emph{Social Psychological and
-Personality Science}, \emph{14}(5), 621--635.
-\url{https://doi.org/10.1177/19485506221111017}
-
-\bibitem[\citeproctext]{ref-benabou_social_2001}
-Benabou, R., \& Ok, E. A. (2001). Social {Mobility} and the {Demand} for
-{Redistribution}: {The Poum Hypothesis}. \emph{The Quarterly Journal of
-Economics}, \emph{116}(2), 447--487.
-\url{https://doi.org/10.1162/00335530151144078}
-
-\bibitem[\citeproctext]{ref-blair_research_2023}
-Blair, G., Coppock, A., \& Humphreys, M. (2023). \emph{Research design
-in the social sciences: Declaration, diagnosis, and redesign}. Princeton
-(N.J.) Oxford: Princeton University press.
-
-\bibitem[\citeproctext]{ref-boccardo_30_2020}
-Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
-que la pandemia revel{ó}} (Nodo XXI). Santiago.
-
-\bibitem[\citeproctext]{ref-breen_effects_2024}
-Breen, R., \& Ermisch, J. (2024). The {Effects} of {Social Mobility}.
-\emph{Sociological Science}, \emph{11}, 467--488.
-\url{https://doi.org/10.15195/v11.a17}
-
-\bibitem[\citeproctext]{ref-brunori_inequality_2025}
-Brunori, P., Ferreira, F. H. G., \& Neidhöfer, G. (2025). Inequality of
-opportunity and intergenerational persistence in {Latin America}.
-\emph{Oxford Open Economics}, \emph{4}(Supplement\_1), i167--i199.
-\url{https://doi.org/10.1093/ooec/odae021}
-
-\bibitem[\citeproctext]{ref-bucca_merit_2016}
-Bucca, M. (2016). Merit and blame in unequal societies: {Explaining
-Latin Americans}' beliefs about wealth and poverty. \emph{Research in
-Social Stratification and Mobility}, \emph{44}, 98--112.
-\url{https://doi.org/10.1016/j.rssm.2016.02.005}
-
-\bibitem[\citeproctext]{ref-busemeyer_skills_2014}
-Busemeyer, M. (2014). \emph{Skills and {Inequality}: {Partisan Politics}
-and the {Political Economy} of {Education Reforms} in {Western Welfare
-States}}. Cambridge University Press.
-
-\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
-Busemeyer, M., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive} and
-{Negative}: {New Perspectives} on {Feedback Effects} in {Public Opinion}
-on the {Welfare State}. \emph{British Journal of Political Science},
-\emph{51}(1), 137--162. \url{https://doi.org/10.1017/S0007123418000534}
-
-\bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
-Busemeyer, M., \& Iversen, T. (2020). The {Welfare State} with {Private
-Alternatives}: {The Transformation} of {Popular Support} for {Social
-Insurance}. \emph{The Journal of Politics}, \emph{82}(2), 671--686.
-\url{https://doi.org/10.1086/706980}
-
-\bibitem[\citeproctext]{ref-campbell_institutional_2020}
-Campbell, J. L. (2020). \emph{Institutional {Change} and
-{Globalization}}. Princeton University Press.
-\url{https://doi.org/10.2307/j.ctv131bw68}
-
-\bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
-Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
-(2021). Sujeto y cotidiano en la era neoliberal: El caso de la
-educaci{ó}n chilena. \emph{Revista Mexicana de Ciencias Pol{í}ticas y
-Sociales}, \emph{67}(244).
-\url{https://doi.org/10.22201/fcpys.2448492xe.2022.244.70386}
-
-\bibitem[\citeproctext]{ref-castillo_legitimacy_2011}
-Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
-Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
-Research}, \emph{24}(4), 314--340.
-\url{https://doi.org/10.1007/s11211-011-0144-5}
-
-\bibitem[\citeproctext]{ref-castillo_changes_2025}
-Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
-{Justification} of {Educational Inequalities}: {The Role} of
-{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
-Pandemic}. \emph{Social Justice Research}.
-\url{https://doi.org/10.1007/s11211-025-00458-0}
-
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
-(2023). A {Multidimensional Approach} for {Measuring Meritocratic
-Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
-Social Inequality Survey}. \emph{International Journal of Sociology},
-1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
-
-\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
-Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
-Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
-{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
-\emph{Under Review at Frontiers in Sociology}.
-
-\bibitem[\citeproctext]{ref-castillo_todos_2013}
-Castillo, J. C., Miranda, D., \& Cabib, I. M. (2013). Todos somos de
-clase media: {Sobre} el estatus social subjetivo en {Chile}. \emph{Latin
-American Research Review}, \emph{48}(1), 155--173.
-\url{https://doi.org/10.1353/lar.2013.0006}
-
-\bibitem[\citeproctext]{ref-castillo_socialization_2024}
-Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
-{Socialization} of {Meritocracy} and {Market Justice Preferences} at
-{School}. \emph{Societies}, \emph{14}(11), 214.
-\url{https://doi.org/10.3390/soc14110214}
-
-\bibitem[\citeproctext]{ref-castillo_meritocracia_2019}
-Castillo, J. C., Torres, A., Atria, J., \& Maldonado, L. (2019).
-Meritocracia y desigualdad econ{ó}mica: {Percepciones}, preferencias e
-implicancias. \emph{Revista Internacional de Sociolog{í}a},
-\emph{77}(1), 117. \url{https://doi.org/10.3989/ris.2019.77.1.17.114}
-
-\bibitem[\citeproctext]{ref-celhay_intergenerational_2010}
-Celhay, P., Sanhueza, C., \& Zubizarreta, J. R. (2010).
-{INTERGENERATIONAL MOBILITY OF INCOME AND SCHOOLING}: {CHILE} 1996-2006.
-\emph{Revista de an{á}lisis Econ{ó}mico}, \emph{25}(2), 43--63.
-\url{https://doi.org/10.4067/S0718-88702010000200003}
-
-\bibitem[\citeproctext]{ref-chancel_world_2022}
-Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
-inequality report 2022.
-https://bibliotecadigital.ccb.org.co/handle/11520/27510.
-
-\bibitem[\citeproctext]{ref-day_movin_2017}
-Day, M. V., \& Fiske, S. T. (2017). Movin' on {Up}? {How Perceptions} of
-{Social Mobility Affect Our Willingness} to {Defend} the {System}.
-\emph{Social Psychological and Personality Science}, \emph{8}(3),
-267--274. \url{https://doi.org/10.1177/1948550616678454}
-
-\bibitem[\citeproctext]{ref-ding_sensitivity_2016}
-Ding, P., \& VanderWeele, T. J. (2016). Sensitivity {Analysis Without
-Assumptions}: \emph{Epidemiology}, \emph{27}(3), 368--377.
-\url{https://doi.org/10.1097/EDE.0000000000000457}
-
-\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
-Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificaci{ó}n y
-movilidad social bajo un modelo neoliberal: El caso de Chile}.
-\emph{Lavboratorio}, (25).
-
-\bibitem[\citeproctext]{ref-espinoza_movilidad_2014}
-Espinoza, V., \& Núñez, J. (2014). Movilidad ocupacional en {Chile}
-2001-2009. {\textquestiondown}{Desigualdad} de ingresos con igualdad de
-oportunidades? \emph{Revista Internacional de Sociolog{í}a},
-\emph{72}(1), 57--82. \url{https://doi.org/10.3989/ris.2011.11.08}
-
-\bibitem[\citeproctext]{ref-eyles_social_2022}
-Eyles, A., Major, L. E., \& Machin, S. (2022). \emph{Social {Mobility} -
-{Past}, {Present} and {Future}: {The State} of {Play} in {Social
-Mobility}, on the 25th {Anniversary} of the {Sutton Trust}}. Sutton
-Trust.
-
-\bibitem[\citeproctext]{ref-ferre_welfare_2023}
-Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
-America}. \emph{Journal of International and Comparative Social Policy},
-\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}
-
-\bibitem[\citeproctext]{ref-flores_top_2020}
-Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
-in {Chile}: {A Historical Perspective} on {Income Inequality},
-1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
-\url{https://doi.org/10.1111/roiw.12441}
-
-\bibitem[\citeproctext]{ref-ganzeboom_internationally_1996}
-Ganzeboom, H. B. G., \& Treiman, D. J. (1996). Internationally
-{Comparable Measures} of {Occupational Status} for the 1988
-{International Standard Classification} of {Occupations}. \emph{Social
-Science Research}, \emph{25}(3), 201--239.
-\url{https://doi.org/10.1006/ssre.1996.0010}
-
-\bibitem[\citeproctext]{ref-gelman_regression_2020}
-Gelman, A., Hill, J., \& Vehtari, A. (2020). \emph{Regression and {Other
-Stories}} (1st ed.). Cambridge University Press.
-\url{https://doi.org/10.1017/9781139161879}
-
-\bibitem[\citeproctext]{ref-gingrich_making_2011}
-Gingrich, J. R. (2011). \emph{Making {Markets} in the {Welfare State}:
-{The Politics} of {Varying Market Reforms}} (1st ed.). Cambridge
-University Press. \url{https://doi.org/10.1017/CBO9780511791529}
-
-\bibitem[\citeproctext]{ref-gugushvili_trends_2014}
-Gugushvili, A. (2014). Trends, {Covariates} and {Consequences} of
-{Intergenerational Social Mobility} in {Post- Socialist Societies}.
-
-\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016c}
-Gugushvili, A. (2016a). Intergenerational objective and subjective
-mobility and attitudes towards income differences: Evidence from
-transition societies. \emph{Journal of International and Comparative
-Social Policy}, \emph{32}(3), 199--219.
-\url{https://doi.org/10.1080/21699763.2016.1206482}
-
-\bibitem[\citeproctext]{ref-gugushvili_intergenerational_2016}
-Gugushvili, A. (2016b). Intergenerational {Social Mobility} and {Popular
-Explanations} of {Poverty}: {A Comparative Perspective}. \emph{Social
-Justice Research}, \emph{29}(4), 402--428.
-\url{https://doi.org/10.1007/s11211-016-0275-9}
-
-\bibitem[\citeproctext]{ref-gugushvili_subjective_2017}
-Gugushvili, A. (2017). Subjective {Intergenerational Mobility} and
-{Support} for {Welfare State Programmes}.
-
-\bibitem[\citeproctext]{ref-gugushvili_heterogeneous_2024}
-Gugushvili, A. (2024). The heterogeneous well-being effects of
-intergenerational mobility perceptions. \emph{Journal of Health
-Psychology}, \emph{29}(2), 99--112.
-\url{https://doi.org/10.1177/13591053231187345}
-
-\bibitem[\citeproctext]{ref-hadjar_does_2015}
-Hadjar, A., \& Samuel, R. (2015). Does upward social mobility increase
-life satisfaction? {A} longitudinal analysis using {British} and {Swiss}
-panel data. \emph{Research in Social Stratification and Mobility},
-\emph{39}, 48--58. \url{https://doi.org/10.1016/j.rssm.2014.12.002}
-
-\bibitem[\citeproctext]{ref-helgason_longterm_2023}
-Helgason, A. F., \& Rehm, P. (2023). Long-term income trajectories and
-the evolution of political attitudes. \emph{European Journal of
-Political Research}, \emph{62}(1), 264--284.
-\url{https://doi.org/10.1111/1475-6765.12506}
-
-\bibitem[\citeproctext]{ref-helgason_class_2025}
-Helgason, A. F., \& Rehm, P. (2025). Class experiences and the long-term
-evolution of economic values. \emph{Social Forces}, \emph{103}(3),
-1125--1143. \url{https://doi.org/10.1093/sf/soae135}
-
-\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
-Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
-D. (2023). Mindsets of poverty: {Implications} for redistributive policy
-support. \emph{Analyses of Social Issues and Public Policy},
-\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
-
-\bibitem[\citeproctext]{ref-immergut_theoretical_1998}
-Immergut, E. M. (1998). The {Theoretical Core} of the {New
-Institutionalism}. \emph{Politics \& Society}, \emph{26}(1), 5--34.
-\url{https://doi.org/10.1177/0032329298026001002}
-
-\bibitem[\citeproctext]{ref-immergut_it_2020}
-Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
-affluent to be able to purchase {``better''} healthcare? {Existential}
-standards and institutional norms in healthcare attitudes across 28
-countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
-\url{https://doi.org/10.1016/j.socscimed.2020.113146}
-
-\bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
-Jaime-Castillo, A. M., \& Marqués-Perales, I. (2019). Social mobility
-and demand for redistribution in {Europe}: A comparative analysis.
-\emph{The British Journal of Sociology}, \emph{70}(1), 138--165.
-\url{https://doi.org/10.1111/1468-4446.12363}
-
-\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
-Janmaat, J. G. (2013). Subjective {Inequality}: A {Review} of
-{International Comparative Studies} on {People}'s {Views} about
-{Inequality}. \emph{European Journal of Sociology}, \emph{54}(3),
-357--389. \url{https://doi.org/10.1017/S0003975613000209}
-
-\bibitem[\citeproctext]{ref-jasso_justice_1978}
-Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification}
-of the {Justice Evaluation Function}. \emph{American Journal of
-Sociology}, \emph{83}(6), 1398--1419.
-\url{https://doi.org/10.1086/226706}
-
-\bibitem[\citeproctext]{ref-jasso_distributive_2016}
-Jasso, G., Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
-{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
-Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
-\url{https://doi.org/10.1007/978-1-4939-3216-0_11}
-
-\bibitem[\citeproctext]{ref-jasso_gender_1999}
-Jasso, G., \& Wegener, B. (1999). Gender and {Country Differences} in
-the {Sense} of {Justice}: {Justice Evaluation}, {Gender Earnings Gap},
-and {Earnings Functions} in {Thirteen Countries}. \emph{International
-Journal of Comparative Sociology}, \emph{40}(1), 94--115.
-\url{https://doi.org/10.1177/002071529904000106}
-
-\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
-of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
-about {Market Justice}, 1991-1996. \emph{European Sociological Review},
-\emph{15}(3), 251--283. Retrieved from
-\url{https://www.jstor.org/stable/522731}
-
-\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
-Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
-\emph{Annual Review of Sociology}, 29--56.
-
-\bibitem[\citeproctext]{ref-koos_moral_2019}
-Koos, S., \& Sachweh, P. (2019). The moral economies of market
-societies: Popular attitudes towards market competition, redistribution
-and reciprocity in comparative perspective. \emph{Socio-Economic
-Review}, \emph{17}(4), 793--821.
-\url{https://doi.org/10.1093/ser/mwx045}
-
-\bibitem[\citeproctext]{ref-lane_market_1986}
-Lane, R. E. (1986). Market {Justice}, {Political Justice}.
-\emph{American Political Science Review}, \emph{80}(2), 383--402.
-\url{https://doi.org/10.2307/1958264}
-
-\bibitem[\citeproctext]{ref-langsaether_explaining_2022}
-Langsæther, P. E., Evans, G., \& O'Grady, T. (2022). Explaining the
-{Relationship Between Class Position} and {Political Preferences}: {A
-Long-Term Panel Analysis} of {Intra-Generational Class Mobility}.
-\emph{British Journal of Political Science}, \emph{52}(2), 958--967.
-\url{https://doi.org/10.1017/S0007123420000599}
-
-\bibitem[\citeproctext]{ref-lee_fairness_2023}
-Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
-inequality: The effects of self-interest and neoliberal orientations.
-\emph{The Australian Educational Researcher}.
-\url{https://doi.org/10.1007/s13384-023-00636-6}
-
-\bibitem[\citeproctext]{ref-lenski_status_1954}
-Lenski, G. E. (1954). Status {Crystallization}: {A Non-Vertical
-Dimension} of {Social Status}. \emph{American Sociological Review},
-\emph{19}(4), 405. \url{https://doi.org/10.2307/2087459}
-
-\bibitem[\citeproctext]{ref-lindh_public_2015}
-Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
-towards {Market Distribution} of {Social Services} -- {A Comparison} of
-17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
-887--910. \url{https://doi.org/10.1111/spol.12105}
-
-\bibitem[\citeproctext]{ref-lindh_bringing_2023}
-Lindh, A., \& McCall, L. (2023). Bringing the market in: An expanded
-framework for understanding popular responses to economic inequality.
-\emph{Socio-Economic Review}, \emph{21}(2), 1035--1055.
-\url{https://doi.org/10.1093/ser/mwac018}
-
-\bibitem[\citeproctext]{ref-llorca-jana_historia_2021}
-Llorca-Jaña, M., \& Miller, R. M. D. (2021). \emph{{Historia econ{ó}mica
-de Chile desde la independencia}}. Santiago de Chile: RIL editores.
-
-\bibitem[\citeproctext]{ref-lopez-roldan_comparative_2021}
-López-Roldán, P., \& Fachelli, S. (Eds.). (2021). \emph{Towards a
-{Comparative Analysis} of {Social Inequalities} between {Europe} and
-{Latin America}}. Cham: Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-48442-2}
-
-\bibitem[\citeproctext]{ref-mac-clure_justicia_2024}
-Mac-Clure, O., Barozet, E., \& Franetovic, G. (2024). {Justicia
-distributiva y posici{ó}n social subjetiva: {\textquestiondown}la
-meritocracia justifica la desigualdad de ingresos?} \emph{Convergencia
-Revista de Ciencias Sociales}, \emph{31}, 1.
-\url{https://doi.org/10.29101/crcs.v31i0.22258}
-
-\bibitem[\citeproctext]{ref-madariaga_three_2020}
-Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
-economic policy trajectory in comparative perspective.
-\emph{Contemporary Politics}, \emph{26}(3), 308--329.
-\url{https://doi.org/10.1080/13569775.2020.1735021}
-
-\bibitem[\citeproctext]{ref-mau_inequality_2015}
-Mau, S. (2015). \emph{Inequality, {Marketization} and the {Majority
-Class}: {Why Did} the {European Middle Classes Accept Neo-Liberalism}?}
-Houndmills: Palgrave Macmillan.
-
-\bibitem[\citeproctext]{ref-mideplan_evolucion_2017}
-MIDEPLAN. (2017). \emph{Evoluci{ó}n de la pobreza 1990-2017}. Ministerio
-de Desarrollo Social y Familia - PNUD.
-
-\bibitem[\citeproctext]{ref-mideplan_estadisticas_2024}
-MIDEPLAN. (2024). \emph{Estad{í}sticas {Encuesta Casen} 2022, secci{ó}n
-{Salud}}. Observatorio social.
-
-\bibitem[\citeproctext]{ref-mijs_paradox_2019}
-Mijs, J. (2019). The paradox of inequality: Income inequality and belief
-in meritocracy go hand in hand. \emph{Socio-Economic Review},
-\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
-
-\bibitem[\citeproctext]{ref-mijs_belief_2022}
-Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
-Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
-{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
-\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}
-
-\bibitem[\citeproctext]{ref-miller_selfserving_1975}
-Miller, D. T., \& Ross, M. (1975). Self-serving biases in the
-attribution of causality: {Fact} or fiction? \emph{Psychological
-Bulletin}, \emph{82}(2), 213--225.
-\url{https://doi.org/10.1037/h0076486}
-
-\bibitem[\citeproctext]{ref-ministeriodeeducacion_resumen_2023}
-Ministerio de Educación. (2023). \emph{Resumen de matr{í}cula por
-establecimiento educacional}. Datos abiertos Ministerio de Educaci{ó}n.
-
-\bibitem[\citeproctext]{ref-molina_its_2019}
-Molina, M. D., Bucca, M., \& Macy, M. W. (2019). It's not just how the
-game is played, it's whether you win or lose. \emph{SCIENCE ADVANCES}.
-
-\bibitem[\citeproctext]{ref-nucleodesociologiacontingente_informe_2020}
-Núcleo de Sociología Contingente, {[}NUDESOC{]}. (2020). \emph{Informe
-de resultados oficial {Encuesta Zona Cero}}. Santiago de Chile.
-
-\bibitem[\citeproctext]{ref-otero_power_2024}
-Otero, G., \& Mendoza, M. (2024). The {Power} of {Diversity}: {Class},
-{Networks} and {Attitudes Towards Inequality}. \emph{Sociology},
-\emph{58}(4), 851--876. \url{https://doi.org/10.1177/00380385231217625}
-
-\bibitem[\citeproctext]{ref-otero_lives_2022}
-Otero, G., Volker, B., Rözer, J., \& Mollenhorst, G. (2022). The lives
-of others: {Class} divisions, network segregation, and attachment to
-society in {Chile}. \emph{The British Journal of Sociology},
-\emph{73}(4), 754--785. \url{https://doi.org/10.1111/1468-4446.12966}
-
-\bibitem[\citeproctext]{ref-panes_criticas_2020}
-Panes, D. (2020). \emph{Cr{í}ticas y experiencias obreras en torno al
-sistema de {AFP}'s en {Chile} (1981-2020)} (PhD thesis). Universidad de
-Chile, Santiago.
-
-\bibitem[\citeproctext]{ref-perez-ahumada_clases_2019}
-Pérez-Ahumada, P. (2019). {Clases sociales, sectores econ{ó}micos y
-cambios en la estructura social chilena entre 1992 y 2013}.
-\emph{Revista de la CEPAL}, \emph{2018}(126), 171--192.
-\url{https://doi.org/10.18356/0ce0faed-es}
-
-\bibitem[\citeproctext]{ref-pnud_desiguales_2017}
-PNUD (Ed.). (2017). \emph{{Desiguales: or{í}genes, cambios y desaf{í}os
-de la brecha social en Chile}}. Santiago, Chile: PNUD : Uqbar Editores.
-
-\bibitem[\citeproctext]{ref-polanyi_great_1975}
-Polanyi, K. (1975). \emph{The great transformation} (Repr). New York,
-NY: Octagon Books.
-
-\bibitem[\citeproctext]{ref-prag_subjective_2021}
-Präg, P., \& Gugushvili, A. (2021). Subjective social mobility and
-health in {Germany}. \emph{European Societies}, \emph{23}(4), 464--486.
-\url{https://doi.org/10.1080/14616696.2021.1887916}
-
-\bibitem[\citeproctext]{ref-ruiz_chilenos_2014}
-Ruiz, C., \& Boccardo, G. (2014). \emph{Los chilenos bajo el
-neoliberalismo: Clases y conflicto social} (Primera edici{ó}n). Santiago
-de Chile: Ediciones y Publicaciones El Buen Aire.
-
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
-common good?} (First edition). New York: {Farrar, Straus and Giroux}.
-
-\bibitem[\citeproctext]{ref-satz_por_2019}
-Satz, D. (2019). \emph{{Por Qu{é} Algunas Cosas No Deber{í}an Estar en
-Venta: Los l{í}mites Morales Del Mercado}}. Ciudad Aut{ó}noma de Buenos
-Aires: Siglo XXI Editores.
-
-\bibitem[\citeproctext]{ref-schmidt_experience_2011}
-Schmidt, A. W. (2011). The experience of social mobility and the
-formation of attitudes toward redistribution. In.
-
-\bibitem[\citeproctext]{ref-shariff_income_2016}
-Shariff, A. F., Wiwad, D., \& Aknin, L. B. (2016). Income {Mobility
-Breeds Tolerance} for {Income Inequality}: {Cross-National} and
-{Experimental Evidence}. \emph{Perspectives on Psychological Science},
-\emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}
-
-\bibitem[\citeproctext]{ref-somma_no_2021}
-Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
-water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
-Movement Studies}, \emph{20}(4), 495--502.
-\url{https://doi.org/10.1080/14742837.2020.1727737}
-
-\bibitem[\citeproctext]{ref-song_there_2025}
-Song, X., \& Zhou, X. (2025). Is {There} a {Mobility Effect}? {On
-Methodological Issues} in the {Mobility Contrast Model}.
-\emph{Sociological Methods \& Research}, \emph{54}(4), 1576--1593.
-\url{https://doi.org/10.1177/00491241251347983}
-
-\bibitem[\citeproctext]{ref-sorokin_social_1927}
-Sorokin, P. A. (1927). \emph{Social {Mobility}}. Harper \& Brothers.
-
-\bibitem[\citeproctext]{ref-streeck_how_2016}
-Streeck, W. (2016). \emph{How will capitalism end? Essays on a failing
-system}. London: Verso.
-
-\bibitem[\citeproctext]{ref-superintendenciadepensiones_estadisticas_2024}
-Superintendencia de Pensiones. (2024). \emph{Estad{í}sticas de
-{Afiliados} en {AFP} en {Sistema} de {Pensiones}. {Abril} 2024}.
-Superintendencia de Pensiones.
-
-\bibitem[\citeproctext]{ref-svallforsMoralEconomyClass2006a}
-Svallfors, S. (2006). \emph{The {Moral Economy} of {Class}: {Class} and
-{Attitudes} in {Comparative Perspective}}. Stanford University Press.
-
-\bibitem[\citeproctext]{ref-svallfors_political_2007}
-Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
-{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
-(1st ed.). Stanford University Press.
-\url{https://doi.org/10.2307/j.ctvr0qv0q}
-
-\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
-Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
-Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
-{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
-\emph{International Review of Social Psychology}, \emph{38}(1), 4.
-\url{https://doi.org/10.5334/irsp.1013}
-
-\bibitem[\citeproctext]{ref-torche_unequal_2005}
-Torche, F. (2005). Unequal {But Fluid}: {Social Mobility} in {Chile} in
-{Comparative Perspective}. \emph{American Sociological Review},
-\emph{70}(3), 422--450. \url{https://doi.org/10.1177/000312240507000304}
-
-\bibitem[\citeproctext]{ref-torche_intergenerational_2014}
-Torche, F. (2014). Intergenerational {Mobility} and {Inequality}: {The
-Latin American Case}. \emph{Annual Review of Sociology}, \emph{40}(1),
-619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
-
-\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
-Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
-care inequalities unfair? {A} study on public attitudes in 23 countries.
-\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
-\url{https://doi.org/10.1186/s12939-016-0350-8}
-
-\bibitem[\citeproctext]{ref-wen_does_2021}
-Wen, F., \& Witteveen, D. (2021). Does perceived social mobility shape
-attitudes toward government and family educational investment?
-\emph{Social Science Research}, \emph{98}, 102579.
-\url{https://doi.org/10.1016/j.ssresearch.2021.102579}
-
-\bibitem[\citeproctext]{ref-wiederkehr_belief_2015}
-Wiederkehr, V., Bonnot, V., Krauth-Gruber, S., \& Darnon, C. (2015).
-Belief in school meritocracy as a system-justifying tool for low status
-students. \emph{Frontiers in Psychology}, \emph{6}.
-
-\bibitem[\citeproctext]{ref-wilson_role_2003}
-Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
-Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
-\url{https://doi.org/10.1023/A:1024667228488}
-
-\bibitem[\citeproctext]{ref-wooldridge_introductory_2009}
-Wooldridge, J. M. (2009). \emph{{Introductory econometrics: a modern
-approach}} (4th ed). Mason, OH: South Western, Cengage Learning.
-
-\bibitem[\citeproctext]{ref-wrightUnderstandingClass2015}
-Wright, E. O. (2015). \emph{Understanding {Class}} (Verso).
-
-\bibitem[\citeproctext]{ref-young_rise_1958}
-Young, M. (1958). \emph{The rise of the meritocracy}. New Brunswick,
-N.J., U.S.A: Transaction Publishers.
-
-\end{CSLReferences}
-
-\section{Supplementary material}\label{anexo}
-
-This section presents the supplementary material for this study. First,
-graphs are shown for the evaluation of the balance of IPW-ATT via
-entropy balance. Second, the complete estimated regression models are
-shown. Finally, the sensitivity analyses performed are shown.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-balance-1.pdf}
-
-}
-
-\caption{\label{fig-balance}Balance SMD --- Mobility treatment (ATT)}
-
-\end{figure}%
+\subsection{Mobility effects models}\label{mobility-effects-models-1}
 
@@
 
+\newpage{}
+
+\subsection{Robustness check and sensitivity
+analysis}\label{robustness-check-and-sensitivity-analysis-1}
+
 \begin{table}
 
-\caption{\label{tbl-holm}Holm correction}
+\caption{\label{tbl-robus1}Effects of intergenerational occupational
+mobility on preferences for pension commodification coded as 3-5
+vs.~1-2, with covariates and wave fixed effects.}
 
@@
 
-\centering\centering
-\begin{tabular}{lll}
-\toprule
-\textbf{Mobility treatment} & \textbf{p value} & \textbf{Holm}\\
-\midrule
-Low→Middle & 0.616 & 1.000\\
-Low→High & 0.594 & 1.000\\
-Middle→Low & 0.904 & 1.000\\
-Middle→High & 0.012 & 0.070\\
-High→Middle & 0.048 & 0.192\\
-\addlinespace
-High→Low & 0.017 & 0.083\\
-\bottomrule
-\multicolumn{3}{l}{\textsuperscript{} Note: p values derived from Table 1}\\
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                                               & $0.16$            & $0.01$            & $0.59^{*}$        & $0.57^{*}$        & $0.48^{*}$        & $0.29$            \\
+                                                        & $ [-0.38;  0.70]$ & $ [-0.37;  0.40]$ & $ [ 0.17;  1.01]$ & $ [ 0.23;  0.92]$ & $ [ 0.04;  0.92]$ & $ [-0.32;  0.90]$ \\
+Mobility treatment                                      & $0.01$            & $0.05$            & $0.09^{*}$        & $-0.02$           & $-0.08$           & $-0.13^{*}$       \\
+                                                        & $ [-0.07;  0.08]$ & $ [-0.02;  0.13]$ & $ [ 0.01;  0.16]$ & $ [-0.09;  0.05]$ & $ [-0.15;  0.00]$ & $ [-0.21; -0.04]$ \\
+Age (in years)                                          & $0.00$            & $-0.00$           & $-0.00^{*}$       & $-0.00$           & $0.00$            & $0.00$            \\
+                                                        & $ [-0.00;  0.00]$ & $ [-0.01;  0.00]$ & $ [-0.01; -0.00]$ & $ [-0.00;  0.00]$ & $ [-0.00;  0.00]$ & $ [-0.00;  0.01]$ \\
+Female (Ref. = Male)                                    & $-0.15^{*}$       & $-0.16^{*}$       & $-0.17^{*}$       & $-0.09^{*}$       & $-0.11^{*}$       & $-0.10^{*}$       \\
+                                                        & $ [-0.22; -0.08]$ & $ [-0.24; -0.07]$ & $ [-0.25; -0.09]$ & $ [-0.16; -0.02]$ & $ [-0.19; -0.03]$ & $ [-0.20; -0.00]$ \\
+Chilean nacionality (Ref. = Non-Chilean)                & $0.03$            & $0.28$            & $-0.10$           & $-0.24$           & $-0.21$           & $-0.02$           \\
+                                                        & $ [-0.57;  0.63]$ & $ [-0.01;  0.57]$ & $ [-0.54;  0.33]$ & $ [-0.56;  0.08]$ & $ [-0.71;  0.29]$ & $ [-0.79;  0.76]$ \\
+Indigenous ethnicity (Ref. = Non-indigenous)            & $-0.03$           & $-0.10$           & $0.02$            & $-0.02$           & $0.02$            & $-0.07$           \\
+                                                        & $ [-0.12;  0.07]$ & $ [-0.22;  0.01]$ & $ [-0.11;  0.15]$ & $ [-0.12;  0.08]$ & $ [-0.13;  0.16]$ & $ [-0.21;  0.07]$ \\
+Co-residence with both parents (Ref. = No co-residence) & $0.01$            & $0.06$            & $0.01$            & $0.00$            & $0.15^{*}$        & $0.12^{*}$        \\
+                                                        & $ [-0.08;  0.09]$ & $ [-0.03;  0.14]$ & $ [-0.07;  0.10]$ & $ [-0.08;  0.08]$ & $ [ 0.06;  0.24]$ & $ [ 0.02;  0.22]$ \\
+Parental education                                      & $0.02$            & $0.02$            & $-0.02$           & $0.00$            & $0.00$            & $0.00$            \\
+                                                        & $ [-0.01;  0.05]$ & $ [-0.01;  0.05]$ & $ [-0.03;  0.00]$ & $ [-0.02;  0.02]$ & $ [-0.01;  0.02]$ & $ [-0.02;  0.02]$ \\
+Wave (Ref.= 2016)                                       &                   &                   &                   &                   &                   &                   \\
+                                                        &                   &                   &                   &                   &                   &                   \\
+\quad Wave 2018                                         & $0.05$            & $0.00$            & $0.07$            & $0.03$            & $0.02$            & $-0.00$           \\
+                                                        & $ [-0.02;  0.12]$ & $ [-0.08;  0.08]$ & $ [-0.01;  0.15]$ & $ [-0.04;  0.11]$ & $ [-0.06;  0.10]$ & $ [-0.09;  0.09]$ \\
+\quad Wave 2023                                         & $0.13^{*}$        & $0.11^{*}$        & $0.23^{*}$        & $0.20^{*}$        & $0.15^{*}$        & $0.15^{*}$        \\
+                                                        & $ [ 0.04;  0.22]$ & $ [ 0.00;  0.21]$ & $ [ 0.14;  0.32]$ & $ [ 0.12;  0.28]$ & $ [ 0.05;  0.25]$ & $ [ 0.05;  0.26]$ \\
+\hline
+R$^2$                                                   & $0.05$            & $0.07$            & $0.09$            & $0.05$            & $0.06$            & $0.07$            \\
+Adj. R$^2$                                              & $0.04$            & $0.05$            & $0.08$            & $0.04$            & $0.05$            & $0.06$            \\
+Num. obs.                                               & $878$             & $745$             & $716$             & $773$             & $912$             & $861$             \\
+RMSE                                                    & $0.44$            & $0.43$            & $0.45$            & $0.44$            & $0.48$            & $0.45$            \\
+N Clusters                                              & $497$             & $449$             & $443$             & $470$             & $536$             & $525$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
 \end{tabular}
-
 }
+\label{table:coefficients}
+\end{center}
+
+}
 
@@
 
-\caption{\label{tbl-evalues}E-values for ATT estimates by mobility
-trajectory (Ding \& VanderWeele, 2016 approximation)}
+\caption{\label{tbl-robus2}Effects of intergenerational occupational
+mobility on preferences for pension commodification excluding
+intermediate category, with covariates and wave fixed effects.}
 
@@
 
-\centering\centering
-\fontsize{8}{10}\selectfont
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\toprule
-\textbf{Mobility treatment} & \textbf{Beta} & \textbf{CI 95\% [low-upper]} & \textbf{sdY} & \textbf{RR point} & \textbf{RR point upper} & \textbf{E point} & \textbf{RR CI} & \textbf{RR CI Upper} & \textbf{E CI} & \textbf{CI Crosses 0}\\
-\midrule
-Low→Middle & -0.016 & -0.078; 0.046 & 0.403 & 0.965 & 1.036 & 1.230 & 1.110 & 1.110 & 1.000 & 1\\
-Low→High & -0.019 & -0.088; 0.051 & 0.383 & 0.956 & 1.046 & 1.264 & 1.128 & 1.128 & 1.000 & 1\\
-Middle→Low & 0.004 & -0.056; 0.063 & 0.392 & 1.008 & 1.008 & 1.101 & 0.879 & 1.138 & 1.000 & 1\\
-Middle→High & 0.088 & 0.020; 0.156 & 0.403 & 1.219 & 1.219 & 1.736 & 1.046 & 1.046 & 1.264 & 0\\
-High→Middle & -0.074 & -0.148;-0.001 & 0.455 & 0.862 & 1.160 & 1.590 & 0.999 & 1.001 & 1.037 & 0\\
-\addlinespace
-High→Low & -0.098 & -0.177;-0.018 & 0.431 & 0.814 & 1.229 & 1.759 & 0.963 & 1.039 & 1.239 & 0\\
-\bottomrule
-\end{tabu}
-
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                                               & $1.84^{*}$        & $1.76^{*}$        & $2.39^{*}$        & $2.40^{*}$        & $2.14^{*}$        & $2.35^{*}$        \\
+                                                        & $ [ 1.08;  2.60]$ & $ [ 1.02;  2.49]$ & $ [ 1.80;  2.97]$ & $ [ 1.88;  2.92]$ & $ [ 1.45;  2.82]$ & $ [ 1.53;  3.18]$ \\
+Mobility treatment                                      & $-0.03$           & $-0.03$           & $0.24^{*}$        & $-0.01$           & $-0.12$           & $-0.22^{*}$       \\
+                                                        & $ [-0.16;  0.09]$ & $ [-0.16;  0.11]$ & $ [ 0.10;  0.38]$ & $ [-0.13;  0.11]$ & $ [-0.26;  0.01]$ & $ [-0.37; -0.07]$ \\
+Age (in years)                                          & $0.00$            & $-0.00$           & $-0.00$           & $-0.00$           & $0.00$            & $-0.00$           \\
+                                                        & $ [-0.01;  0.01]$ & $ [-0.01;  0.01]$ & $ [-0.01;  0.00]$ & $ [-0.01;  0.00]$ & $ [-0.00;  0.01]$ & $ [-0.01;  0.01]$ \\
+Female (Ref. = Male)                                    & $-0.24^{*}$       & $-0.23^{*}$       & $-0.27^{*}$       & $-0.16^{*}$       & $-0.16^{*}$       & $-0.22^{*}$       \\
+                                                        & $ [-0.35; -0.12]$ & $ [-0.39; -0.07]$ & $ [-0.42; -0.12]$ & $ [-0.29; -0.03]$ & $ [-0.30; -0.02]$ & $ [-0.39; -0.06]$ \\
+Chilean nacionality (Ref. = Non-Chilean)                & $0.07$            & $0.28$            & $-0.16$           & $-0.34$           & $-0.20$           & $-0.07$           \\
+                                                        & $ [-0.73;  0.87]$ & $ [-0.33;  0.89]$ & $ [-0.69;  0.37]$ & $ [-0.76;  0.07]$ & $ [-0.82;  0.43]$ & $ [-0.75;  0.61]$ \\
+Indigenous ethnicity (Ref. = Non-indigenous)            & $0.03$            & $-0.08$           & $-0.02$           & $0.04$            & $0.05$            & $-0.21$           \\
+                                                        & $ [-0.14;  0.20]$ & $ [-0.28;  0.12]$ & $ [-0.28;  0.25]$ & $ [-0.15;  0.23]$ & $ [-0.20;  0.31]$ & $ [-0.49;  0.07]$ \\
+Co-residence with both parents (Ref. = No co-residence) & $0.00$            & $0.05$            & $-0.09$           & $-0.01$           & $0.15$            & $0.15$            \\
+                                                        & $ [-0.14;  0.15]$ & $ [-0.10;  0.21]$ & $ [-0.23;  0.06]$ & $ [-0.16;  0.13]$ & $ [-0.01;  0.32]$ & $ [-0.03;  0.33]$ \\
+Parental education                                      & $0.01$            & $0.02$            & $-0.04^{*}$       & $-0.01$           & $0.00$            & $-0.01$           \\
+                                                        & $ [-0.04;  0.07]$ & $ [-0.04;  0.07]$ & $ [-0.08; -0.01]$ & $ [-0.05;  0.02]$ & $ [-0.03;  0.04]$ & $ [-0.05;  0.03]$ \\
+Wave (Ref.= 2016)                                       &                   &                   &                   &                   &                   &                   \\
+                                                        &                   &                   &                   &                   &                   &                   \\
+\quad Wave 2018                                         & $0.02$            & $-0.13$           & $0.10$            & $-0.01$           & $-0.08$           & $-0.18$           \\
+                                                        & $ [-0.11;  0.16]$ & $ [-0.28;  0.03]$ & $ [-0.05;  0.26]$ & $ [-0.14;  0.13]$ & $ [-0.24;  0.07]$ & $ [-0.37;  0.01]$ \\
+\quad Wave 2023                                         & $0.21^{*}$        & $0.03$            & $0.39^{*}$        & $0.37^{*}$        & $0.30^{*}$        & $0.16$            \\
+                                                        & $ [ 0.06;  0.36]$ & $ [-0.15;  0.21]$ & $ [ 0.23;  0.54]$ & $ [ 0.23;  0.51]$ & $ [ 0.13;  0.46]$ & $ [-0.04;  0.36]$ \\
+\hline
+R$^2$                                                   & $0.04$            & $0.04$            & $0.10$            & $0.06$            & $0.06$            & $0.08$            \\
+Adj. R$^2$                                              & $0.03$            & $0.02$            & $0.09$            & $0.05$            & $0.05$            & $0.07$            \\
+Num. obs.                                               & $812$             & $677$             & $629$             & $693$             & $808$             & $770$             \\
+RMSE                                                    & $0.77$            & $0.73$            & $0.75$            & $0.75$            & $0.82$            & $0.79$            \\
+N Clusters                                              & $480$             & $431$             & $415$             & $445$             & $499$             & $494$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
 }
+\label{table:coefficients}
+\end{center}
+
+}
 
@@
 
+\begin{table}
+
+\caption{\label{tbl-robus3}Effects of intergenerational occupational
+mobility on preferences for pension commodification using the original
+5-point likert scale, with covariates and wave fixed effects.}
+
+\centering{
+
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                                               & $1.99^{*}$        & $1.76^{*}$        & $2.91^{*}$        & $2.86^{*}$        & $2.53^{*}$        & $2.59^{*}$        \\
+                                                        & $ [ 0.81;  3.17]$ & $ [ 0.82;  2.70]$ & $ [ 2.19;  3.64]$ & $ [ 2.24;  3.47]$ & $ [ 1.59;  3.47]$ & $ [ 1.34;  3.83]$ \\
+Mobility treatment                                      & $-0.03$           & $0.03$            & $0.30^{*}$        & $-0.02$           & $-0.19$           & $-0.33^{*}$       \\
+                                                        & $ [-0.21;  0.15]$ & $ [-0.16;  0.22]$ & $ [ 0.11;  0.48]$ & $ [-0.18;  0.14]$ & $ [-0.38;  0.00]$ & $ [-0.54; -0.12]$ \\
+Age (in years)                                          & $0.00$            & $-0.00$           & $-0.00$           & $-0.00$           & $0.00$            & $0.00$            \\
+                                                        & $ [-0.01;  0.01]$ & $ [-0.01;  0.01]$ & $ [-0.01;  0.00]$ & $ [-0.01;  0.00]$ & $ [-0.01;  0.01]$ & $ [-0.01;  0.01]$ \\
+Female (Ref. = Male)                                    & $-0.37^{*}$       & $-0.37^{*}$       & $-0.40^{*}$       & $-0.23^{*}$       & $-0.26^{*}$       & $-0.32^{*}$       \\
+                                                        & $ [-0.54; -0.20]$ & $ [-0.58; -0.15]$ & $ [-0.60; -0.21]$ & $ [-0.41; -0.06]$ & $ [-0.46; -0.06]$ & $ [-0.55; -0.08]$ \\
+Chilean nacionality (Ref. = Non-Chilean)                & $0.10$            & $0.56$            & $-0.25$           & $-0.49^{*}$       & $-0.32$           & $-0.05$           \\
+                                                        & $ [-1.17;  1.38]$ & $ [-0.08;  1.20]$ & $ [-0.89;  0.39]$ & $ [-0.93; -0.05]$ & $ [-1.23;  0.59]$ & $ [-1.24;  1.15]$ \\
+Indigenous ethnicity (Ref. = Non-indigenous)            & $-0.00$           & $-0.18$           & $-0.00$           & $0.02$            & $0.08$            & $-0.26$           \\
+                                                        & $ [-0.24;  0.24]$ & $ [-0.46;  0.10]$ & $ [-0.35;  0.35]$ & $ [-0.24;  0.28]$ & $ [-0.29;  0.44]$ & $ [-0.66;  0.13]$ \\
+Co-residence with both parents (Ref. = No co-residence) & $0.01$            & $0.11$            & $-0.06$           & $-0.01$           & $0.28^{*}$        & $0.26^{*}$        \\
+                                                        & $ [-0.20;  0.22]$ & $ [-0.11;  0.33]$ & $ [-0.27;  0.15]$ & $ [-0.21;  0.19]$ & $ [ 0.05;  0.50]$ & $ [ 0.01;  0.50]$ \\
+Parental education                                      & $0.04$            & $0.03$            & $-0.06^{*}$       & $-0.01$           & $0.01$            & $-0.01$           \\
+                                                        & $ [-0.04;  0.11]$ & $ [-0.04;  0.11]$ & $ [-0.10; -0.01]$ & $ [-0.05;  0.04]$ & $ [-0.04;  0.05]$ & $ [-0.07;  0.05]$ \\
+Wave (Ref.= 2016)                                       &                   &                   &                   &                   &                   &                   \\
+                                                        &                   &                   &                   &                   &                   &                   \\
+\quad Wave 2018                                         & $0.08$            & $-0.12$           & $0.16$            & $0.03$            & $-0.05$           & $-0.17$           \\
+                                                        & $ [-0.11;  0.26]$ & $ [-0.32;  0.09]$ & $ [-0.04;  0.36]$ & $ [-0.15;  0.21]$ & $ [-0.26;  0.16]$ & $ [-0.42;  0.09]$ \\
+\quad Wave 2023                                         & $0.33^{*}$        & $0.14$            & $0.55^{*}$        & $0.53^{*}$        & $0.42^{*}$        & $0.30^{*}$        \\
+                                                        & $ [ 0.11;  0.54]$ & $ [-0.11;  0.39]$ & $ [ 0.35;  0.76]$ & $ [ 0.35;  0.72]$ & $ [ 0.18;  0.65]$ & $ [ 0.02;  0.57]$ \\
+\hline
+R$^2$                                                   & $0.04$            & $0.05$            & $0.10$            & $0.06$            & $0.06$            & $0.08$            \\
+Adj. R$^2$                                              & $0.03$            & $0.04$            & $0.09$            & $0.05$            & $0.05$            & $0.07$            \\
+Num. obs.                                               & $878$             & $745$             & $716$             & $773$             & $912$             & $861$             \\
+RMSE                                                    & $1.10$            & $1.04$            & $1.06$            & $1.07$            & $1.17$            & $1.12$            \\
+N Clusters                                              & $497$             & $449$             & $443$             & $470$             & $536$             & $525$             \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-evalues1}E-values for ATT estimates for Middle-High
+trajectory (Cinelli \& Hazlett, 2020 approximation)}
+
+\centering{
+
+\centering
+\begin{tabular}{lrrrrrr}
+\multicolumn{7}{c}{Outcome: \textit{y}} \\
+\hline \hline 
+Treatment: & Est. & S.E. & t-value & $R^2_{Y \sim D |{\bf X}}$ & $RV_{q = 1}$ & $RV_{q = 1, \alpha = 0.05}$  \\ 
+\hline 
+\textit{t} & 0.088 & 0.029 & 2.995 & 1.3\% & 10.7\% & 3.8\% \\ 
+\hline 
+df = 706 & & \multicolumn{5}{r}{ \small \textit{Bound (1x sexo)}: $R^2_{Y\sim Z| {\bf X}, D}$ = 2.7\%, $R^2_{D\sim Z| {\bf X} }$ = 3.6\%} \\
+\end{tabular}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-evalues2}E-values for ATT estimates for High-Middle
+trajectory (Cinelli \& Hazlett, 2020 approximation)}
+
+\centering{
+
+\centering
+\begin{tabular}{lrrrrrr}
+\multicolumn{7}{c}{Outcome: \textit{y}} \\
+\hline \hline 
+Treatment: & Est. & S.E. & t-value & $R^2_{Y \sim D |{\bf X}}$ & $RV_{q = 1}$ & $RV_{q = 1, \alpha = 0.05}$  \\ 
+\hline 
+\textit{t} & -0.074 & 0.03 & -2.447 & 0.7\% & 7.8\% & 1.6\% \\ 
+\hline 
+df = 902 & & \multicolumn{5}{r}{ \small \textit{Bound (1x sexo)}: $R^2_{Y\sim Z| {\bf X}, D}$ = 1.5\%, $R^2_{D\sim Z| {\bf X} }$ = 3.9\%} \\
+\end{tabular}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-evalues3}E-values for ATT estimates for High-Low
+trajectory (Cinelli \& Hazlett, 2020 approximation)}
+
+\centering{
+
+\centering
+\begin{tabular}{lrrrrrr}
+\multicolumn{7}{c}{Outcome: \textit{y}} \\
+\hline \hline 
+Treatment: & Est. & S.E. & t-value & $R^2_{Y \sim D |{\bf X}}$ & $RV_{q = 1}$ & $RV_{q = 1, \alpha = 0.05}$  \\ 
+\hline 
+\textit{t} & -0.098 & 0.03 & -3.245 & 1.2\% & 10.5\% & 4.3\% \\ 
+\hline 
+df = 851 & & \multicolumn{5}{r}{ \small \textit{Bound (1x sexo)}: $R^2_{Y\sim Z| {\bf X}, D}$ = 1.7\%, $R^2_{D\sim Z| {\bf X} }$ = 0.8\%} \\
+\end{tabular}
+
+}
+
+\end{table}%
+
 \end{document}

### 2025-11-17 processing/prod_analysis_multinom.R

@@ -411,7 +411,7 @@ df_pension_lab <- df_pension_lab %>%
   )
 
 g3 <- ggplot(df_pension_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 0.5, alpha = 1) +
+  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 1) +
   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, size = 0.7) +
   geom_point(size = 2.5) +
   scale_x_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +
@@ -653,18 +653,29 @@ g4 <- ggplot(
   df_pension_lab_i,
   aes(x = estimate, y = name, group = term, colour = term, shape = term)
 ) +
-  geom_vline(xintercept = 0, linetype = "dashed", colour = "darkred", linewidth = 0.6) +
+  geom_vline(xintercept = 0, linetype = "solid", colour = "darkred", linewidth = 0.6) +
   geom_errorbarh(
-    aes(xmin = conf.low, xmax = conf.high),
+    aes(xmin = conf.low, xmax = conf.high, linetype = term),  # <- acá
     height = 0, size = 0.7, position = pos
   ) +
   geom_point(size = 2.8, position = pos) +
-  scale_shape_manual(values = c(16, 15)) +     # 16 = círculo, 15 = cuadrado
-  scale_colour_manual(values = c("black", "grey50")) +
+  scale_shape_manual(values = c(
+    "High Meritocracy" = 16,  # círculo
+    "Low Meritocracy"  = 15   # cuadrado
+  )) +
+  scale_colour_manual(values = c(
+    "High Meritocracy" = "black",
+    "Low Meritocracy"  = "grey50"
+  )) +
+  scale_linetype_manual(values = c(
+    "High Meritocracy" = "solid",
+    "Low Meritocracy"  = "dashed"   # <- low en dashed
+  )) +
   scale_x_continuous(limits = c(-0.5, 0.5), breaks = seq(-0.5, 0.5, 0.25)) +
-  labs(x = "Treatment Effect", y = NULL, colour = NULL, shape = NULL) +
+  labs(x = "Treatment Effect", y = NULL, colour = NULL, shape = NULL, linetype = NULL) +
   my_pretty_theme() +
   theme(legend.position = "bottom")
+
 
 # -----------
 # Time 
[…]

### 2025-11-21 paper/01-introduction.qmd
up antes de borrar tiempo

@@
 
-What is the legitimate extent of market inequality in the eyes of the public?
-Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016].
-In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023].
-This transformation reduced the role of public provision and increased the presence of private actors in core social services [@ferre_welfare_2023].
-Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a].
-Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @castillo_perceptions_2025; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
-Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].
+What is the legitimate scope of market inequality in the eyes of the public?
+Since the early 1980s, many countries have experienced a rollback of universal welfare programs and a shift toward the privatisation and commodification of public goods and social services [@gingrich_making_2011; @streeck_how_2016].
+In Latin America, these reforms extended market logic into domains of social reproduction long governed by the state, shrinking public provision and expanding private actors in core welfare sectors [@ferre_welfare_2023].
+From a moral-economy perspective, this diffusion of market rules has reconfigured the normative order: through policy feedback, welfare institutions increasingly embed market-based criteria of “fair” allocation and shape how citizens understand deservingness and the balance between state and market [@svallforsMoralEconomyClass2006a; @koos_moral_2019].
+Against this backdrop, growing scholarship examines *market justice preferences*—the extent to which citizens regard market-based criteria as a fair basis for access to essential services such as healthcare, education and pensions [@castillo_perceptions_2025; @svallfors_political_2007; @lindh_public_2015].
+These orientations matter because they help legitimise economic inequality by portraying stratified outcomes as a fair result of individual responsibility and limited state intervention.
 
-Existing literature shows that market justice preferences are shaped by both the economic and institutional context and individuals' positions within social stratification.
-Grounded in the notion that economic institutions influence people's normative attitudes [@immergut_theoretical_1998], studies find that countries with stronger public provision or more expansive welfare states exhibit lower levels of market justice preferences [@busemeyer_skills_2014; @immergut_it_2020], while more privatized contexts show stronger support for market-based criteria [@castillo_socialization_2024; @lindh_public_2015].
-In such contexts, market justice preferences tend to rise as individuals "ascend" the social structure, with those in more privileged positions in terms of class, education, and income holding stronger preferences for market-based solutions compared to those in more disadvantaged or at-risk positions [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_socialization_2024; @otero_power_2024; @lindh_public_2015; @svallfors_political_2007].
+This article examines public support for the commodification of pension welfare—specifically, whether people consider it fair that access to better pensions depends on income or contributions.
+Pension privatisation lies at the core of Latin America’s trajectory of commodification [@huber_political_2000], making pension systems a strategic domain for analysing how market principles become normatively embedded [@mau_inequality_2015].
+As many countries expanded mandatory or voluntary individual-capitalisation schemes—most radically in Chile—pensions became organised around contribution-based entitlements and market performance [@arenas_sistemas_2019; @verbic_political_2019].
+This shift reshapes not only how pension institutions allocate old-age risk, heightening exposure to poverty in retirement and tying benefit levels to lifetime labour-market trajectories [@oecd_pensions_2023; @madero-cabib_private_2019], but also reshape the justice principles citizens associate with old-age security: people have to judge whether higher benefits should follow market-derived criteria or solidaristic principles [@arrizabalomontoro_debate_2019; @borzutzky_pension_2012].
+Although we know something about what people consider a “just” monetary pension [@castillo_deserving_2019] and about support for public versus private provision [@busemeyer_welfare_2020; @jaime-castillo_public_2013], we still lack systematic evidence on how citizens assess the fairness of market-dependent access to pensions.
 
-Market justice preferences are shaped not only by objective socioeconomic conditions but also by popular beliefs about inequality.
-Among these, meritocracy is a key normative principle underpinning market-based distributive preferences [@mau_inequality_2015].
-Studies show that individuals with stronger meritocratic beliefs tend to perceive less inequality and legitimize it by attributing economic differences to personal achievement [@mijs_paradox_2019; @wilson_role_2003; @batruch_belief_2023].
-In highly unequal societies where access to services is largely governed by market logic, such beliefs play a critical role in normalizing inequality.
-Recent evidence from Chile shows that students who believe effort and talent are rewarded in their country express stronger preferences for market-based access to healthcare, pensions, and education [@castillo_socialization_2024].
+Research shows that support for market-based access to welfare services is strongly stratified: individuals with higher income, education or class position are more likely to endorse market-based criteria than disadvantaged groups [@busemeyer_skills_2014; @immergut_it_2020;  @lindh_public_2015;  @svallfors_political_2007].
+Evidence for pensions points in the same direction: higher income and greater financial gains in funded schemes correlate with stronger support for market-oriented pension policies [@kerner_pension_2020].
+Yet we know very little about how movement within the class structure shapes preferences for pension commodification.
+This gap is especially salient in Latin America, where mobility has unfolded under high inequality and deeply privatised welfare regimes [@lopez-roldan_comparative_2021].
+Social origins and destinations anchor material interests and justice principles [@alesina_intergenerational_2018; @langsaether_explaining_2022; @gugushvili_subjective_2017], while movement between these positions exposes individuals to new risks, resources and normative environments that reshape fairness judgements [@gugushvili_trends_2014; @helgason_longterm_2023].
+As @mau_inequality_2015 argues, social mobility can foster market-justice orientations, helping explain how inequalities in access to welfare services become normatively sustained.
 
-Although it is clear that one's social position influences market justice preferences, the question of how upward or downward mobility within the social structure affects these preferences remains unanswered.
-This question is far from trivial, especially in Latin America, where many have experienced various forms of mobility amid high economic inequality and deep welfare privatization [@lopez-roldan_comparative_2021; @torche_intergenerational_2014; @ferre_welfare_2023].
-Social origins and destinations affect attitudes toward inequality in distinct ways [ @jaime-castillo_social_2019; @gugushvili_intergenerational_2016; @day_movin_2017; @gugushvili_subjective_2017; @wen_does_2021; @mijs_belief_2022], while movement between these positions exposes individuals to different experiences and mechanisms that shape their views on what is fair [@gugushvili_trends_2014; @mau_inequality_2015].
-Building on this research, examining the effects of social mobility on market justice preferences can help to illuminate how inequalities in access to social services are justified among individuals who have experienced, or not, changes in their social standing, and what are the normative mechanisms that guide this justification [@mau_inequality_2015].
+The literature identifies several mechanisms through which mobility may shape distributive-justice preferences—material self-interest [@benabou_social_2001], socialisation and acculturation [@jaime-castillo_social_2019], learning through long-term exposure [@helgason_longterm_2023], and cognitive attributions such as self-serving bias or privilege awareness [@gugushvili_intergenerational_2016c; @molina_its_2019].
+While distinct, these accounts converge in highlighting two broad channels relevant for market-based welfare: shifts in material stakes and shifts in normative frames [@mau_inequality_2015].
+In this paper, I do not treat these mechanisms as mediators but use them to guide hypotheses about heterogeneity: if mobility alters preferences through material interests and normative updating, its effects should be most potent among those who interpret social outcomes through merit-based evaluations of effort [@castillo_perceptions_2025; @mijs_belief_2022].
 
-Beyond their isolated effects, social mobility and meritocratic beliefs interact in complex ways to shape market justice preferences.
-Among others, a key mechanism proposed in the literature to explain how mobility influences distributive justice preferences is the psychological process of self-serving bias in causal attribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c].
-This bias suggests that individuals attribute failures---such as downward mobility---to external factors, while crediting successes---such as upward mobility---to their own merit and effort [@miller_selfserving_1975].
-Those who experience upward mobility tend to view their social position as earned, making them more likely to believe that individuals are responsible for their own success or failure.
-Research shows that upward mobility is associated with weaker preferences for redistribution [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @jaime-castillo_social_2019; @alesina_intergenerational_2018] and stronger legitimacy of income inequality [@shariff_income_2016].
-In contrast, individuals who experience downward mobility tend to blame structural factors like inequality and are more supportive of redistribution while rejecting merit-based explanations [@gugushvili_trends_2014].
-Taken together, I argue that meritocratic beliefs may reinforce this self-serving attribution mechanism by legitimizing one's social status as the outcome of personal merit, closely tied to attribution bias.
+Against this background, this article pursues two aims: first, to assess whether intergenerational mobility causally affects support for the commodification of pensions; and second, to examine whether this effect varies by meritocratic beliefs.
+I argue that mobility works through two main channels.
+Upward mobility increases earnings, expected returns in a funded system and perceived economic security, making contribution-based access more appealing.
+At the same time, integration into higher-class networks exposes people to more individualistic norms and effort-based understandings of fairness.
+Downward mobility, in turn, heightens economic insecurity, dependence on collective protection and attachment to solidaristic principles, lowering support for market allocation.
+Meritocratic beliefs are conceptualised not as mediators but as moderators.
+Among those who see mobility as a fair reward for effort, both material and normative effects should be amplified, producing higher support for pension commodification.
+The paper contributes to research on market justice by focusing on a specific welfare domain and by providing causal evidence on how mobility shapes distributive preferences.
 
-Against this background, this article pursues two main objectives: first, to analyze the extent to which intergenerational social mobility influences market justice preferences regarding healthcare, pensions, and education; and second, to examine how meritocratic beliefs may moderate this relationship.
-Building on a theoretical framework that emphasizes how neoliberal transformations---particularly through the privatization and commodification of key areas of social reproduction---have profoundly reshaped processes of subject formation [@mau_inequality_2015], the central argument is that upward mobility increases support for market justice preferences, while downward mobility decreases it.
-Moreover, meritocratic beliefs are expected to moderate this relationship by reflecting a self-serving attribution bias, whereby individuals justify their social position in terms of personal merit.
+This study focuses on Chile, a particularly instructive case for examining preferences for pension commodification.
+Despite sustained economic growth, Chile combines high and persistent inequality with short-range upward mobility and substantial barriers to higher-class positions [@flores_top_2020; @salgado_uplifting_2025; @torche_intergenerational_2014].
+It was the first country worldwide to fully replace a public pay-as-you-go system with a mandatory, privately administered (through pension fund administrators, AFPs) defined-contribution scheme, while the state has progressively expanded a segmented solidarity pillar to compensate for insufficient benefits [@madariaga_three_2020; @boccardo_30_2020; @solimano_rise_2021].
+In parallel—and despite waves of protest from 2006 to 2019 targeting the AFP system [@somma_no_2021]—Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing attitudes toward pension distribution and inequality [@araujo_desafios_2012; @galvez_con_2023].
+The combination of profound inequality, market-driven old-age security and contested legitimacy makes Chile a critical case for analysing how mobility and meritocratic beliefs structure support for pension commodification.
 
-This study focuses on Chile, a particularly intriguing case for examining market justice preferences.
-Despite significant economic growth and poverty reduction, Chile has some of the highest levels of inequality in Latin America and among OECD countries [@chancel_world_2022; @flores_top_2020].
-This inequality coexists with short-range upward mobility among lower-class segments moving into middle strata, though strong barriers remain to reaching higher positions [@torche_intergenerational_2014; @espinoza_estratificacion_2013].
-What makes Chile especially salient is that much of this inequality is rooted in deep neoliberal reforms that institutionalized the privatization and commodification of key social sectors [@madariaga_three_2020].
-Introduced during the dictatorship (1973--1989) and expanded in democracy, these reforms enabled the unprecedented emergence of markets in health, pensions, and education, with provision segmented by individuals' ability to pay and supported by public subsidies [@boccardo_30_2020].
-In parallel---and despite waves of protest against inequality and commodification from 2006 to 2019 [@somma_no_2021]---Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing their attitudes toward inequality and welfare distribution [@araujo_desafios_2012; @canalesceron_sujeto_2021].
-
-In this context, the questions that guide this research are as follows:
-
-(1) To what extent does intergenerational social mobility influence market justice preferences regarding healthcare, pensions, and education in Chile?
-(2) How do meritocratic beliefs condition or moderate this relationship in the Chilean context?
-
-To address these questions, this study draws on large-scale, representative survey data collected in 2018 from the urban Chilean population aged 18 to 75 (n = 2,726).
-The next section outlines the theoretical framework linking market justice preferences, social mobility, and meritocratic beliefs, and proposes a set of hypotheses.
-This is followed by a description of the data, variables, and analytical strategy.
-The final sections present the empirical findings, offer an interpretation of the results, and conclude with a discussion of their implications.
+To address these questions, the study draws on large-scale, representative survey data for the urban Chilean population collected in 2016, 2018 and 2023.
+It adopts a causal framework of mobility effects inspired by @breen_effects_2024.
+The following section outlines the theoretical framework linking preferences for market-based welfare, social mobility and expected heterogeneity by meritocratic beliefs and time, and derives hypotheses.
+A subsequent section presents the causal identification strategy, the data, the variables, and the analytical approach.
+The empirical sections report the findings and interpret their implications.
+The article concludes by discussing what these results reveal about the politics of market-based pension welfare and social mobility.

### 2025-11-21 paper/04-analysis.qmd
up antes de borrar tiempo

@@
 Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility, estimated from interaction models (T × Merit).
-Points show simple slopes of the mobility effect at low merit and high merit; bars are 95% confidence intervals.
+Points show the marginal effects (simple slopes) of the mobility treatment at low and high merit; bars are 95% confidence intervals.
 Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual."
@@
 
-@fig-interact shows that neither the interaction terms nor the mobility coefficients reach conventional significance across trajectories, indicating no detectable moderation of the causal mobility effect by meritocratic perceptions at the 95% level.
-Directionally, in the left panel, the interaction is positive for Middle→High ($\beta$ = 0.15, 95% CI [−0.03, 0.33]) and Low→High ($\beta$ = 0.05, [−0.14, 0.24]), suggesting, if anything, somewhat larger pro-commodification effects among high-merit respondents when upwardly mobile, whereas it is negative for High→Low ($\beta$ = −0.14, [−0.38, 0.10]) and High→Middle ($\beta$ = −0.03, [−0.23, 0.17]), showing weaker pro-commodification effects under downward mobility at high merit.
-Focusing on the treatment effect at low meritocracy (the reference category) at the right panel, the mobility coefficient captures the origin-specific ATT comparing movers to their immobility counterfactual among low-merit respondents.
-These effects are small and imprecise across trajectories: Low→Middle ($\beta$ = −0.02, [−0.09, 0.04]) and Low→High ($\beta$ = −0.03, [−0.10, 0.05]) are near zero.
-Middle→High is positive but not significant ($\beta$ = 0.07, [−0.01, 0.14]), Middle→Low is null ($\beta$ = -0.00, [−0.07, 0.06]), and downward from high shows negative, borderline estimates: High→Middle ($\beta$ = −0.07, [−0.15, 0.00]) and High→Low ($\beta$ = −0.08, [−0.15, 0.00]).
-Substantively, I do not find evidence that meritocratic perceptions moderate the causal effect of mobility, although the signs align with the idea that upward mobility at high merit tilts attitudes toward commodification, and downward mobility at high merit tilts them away.
+@fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic beliefs.
+For low-merit respondents, mobility effects are uniformly small and imprecise across trajectories: Low→Middle ($\beta$ = −0.02, 95% CI [−0.09, 0.04]), Low→High ($\beta$ = −0.03, [−0.10, 0.05]), Middle→Low ($\beta$ = −0.00, [−0.07, 0.06]), and a modest, non-significant positive estimate for Middle→High ($\beta$ = 0.07, [−0.01, 0.14]), while downward moves from high origin show borderline negative effects (High→Middle: $\beta$ = −0.07, [−0.15, 0.00]; High→Low: $\beta$ = −0.08, [−0.15, 0.00]).
+Among high-merit respondents, the pattern is directionally consistent with meritocratic moderation but statistically robust only for one trajectory: the ATT for Middle→High at high merit is positive and significant ($\beta$ = 0.22, [0.05, 0.38]), whereas Low→Middle ($\beta$ = 0.02, [−0.13, 0.16]), Low→High ($\beta$ = 0.02, [−0.15, 0.20]), High→Middle ($\beta$ = −0.10, [−0.28, 0.09]), and High→Low ($\beta$ = −0.22, [−0.44, 0.00]) remain statistically indistinguishable from zero.
+Overall, these results provide only limited and trajectory-specific evidence that meritocratic beliefs moderate the causal effect of intergenerational occupational mobility on preferences for pension commodification, with a clearer pro-commodification sign only for upward mobility from middle to high status among high-merit respondents.
 
 ```{r}
-#| label: tbl-inttime
-#| tbl-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by survey wave"
-#| tbl-cap-location: top
+#| label: fig-interact-wave
+#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by survey wave.
+Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility, estimated from interaction models (T × Wave).
+Points show the marginal effects (simple slopes) of the mobility treatment in 2016, 2018, and 2023; bars are 95% confidence intervals.
+Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual."
+#| out-width: '90%'
+#| fig-asp: 0.8
+#| fig-cap-location: bottom
+#| echo: false
 #| results: asis
-#| echo: false
 
-ccoef <- list(
-  "(Intercept)" = "Intercept",
-  "t" = "Mobility treatment",
-  "ola2018" = "Wave 2018",
-  "ola2023" = "Wave 2023",
-  "t:ola2018" = "Mobility treatment x Wave 2018",
-  "t:ola2023" = "Mobility treatment x Wave 2023"
-  )
-
-texreg::texreg(list(mp_lm_t,
-               mp_lh_t,
-               mp_mh_t,               
-               mp_ml_t,
-               mp_hm_t,
-               mp_hl_t),
-          custom.model.names = c(
-            "Low-Middle",
-            "Low-High",
-            "Middle-High",            
-            "Middle-Low",
-            "High-Middle",
-            "High-Low"
-          ),   caption.above = NULL,
-               caption = NULL,
-               custom.coef.map = ccoef,
-               digits = 2,
-               groups = list("Wave (Ref.= 2016)" = 3:4, "Mobility treatment x Wave (Ref.= 2016)" = 5:6),
-               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
-               leading.zero = T,
-               use.packages = F,
-               booktabs = F,
-               scalebox = 0.80,
-               center = T,float.pos = "H!",
-               custom.gof.rows = list("Controls"=rep("Yes",6)))
-
+g5+
+  theme(legend.text = element_text(size = 10))
 ```
 
-@tbl-inttime shows the causal heterogeneity of the mobility effect across the different waves of the survey.
-Regarding temporal heterogeneity, the only statistically significant interaction term is observed for the High→Middle trajectory in 2023 ($\beta_{\text{int}} = 0.19$), 95% CI [0.01, 0.38]).
-However, when this interaction is added to the baseline effect in 2016 (-0.19), the marginal effect of High→Middle mobility in 2023 is approximately zero, and its confidence interval includes the null.
-Thus, rather than indicating a reversal toward a positive effect, the evidence suggests that the negative impact of downward mobility from high backgrounds is confined to 2016 and attenuates over time.
-No comparable temporal changes are observed in the remaining trajectories.
-Overall, the results provide little support for strong time-conditional heterogeneity in mobility effects, beyond the fading of the initially negative High→Middle effect.
+I find no consistent evidence of heterogeneous causal effects across survey waves.
+@fig-interact-wave displays the marginal effects of mobility trajectories on market-justice preferences for pensions in 2016, 2018, and 2023.
+Across trajectories, the estimated ATTs remain small and their confidence intervals largely overlap over time.
+Upward mobility from middle origin shows a positive and statistically significant effect in 2016 (Middle→High, $\beta$ = 0.13, 95% CI [0.05, 0.22]) and downward mobility from high origin displays negative effects at baseline (High→Middle and High→Low, both $\beta$ ≈ −0.19, CIs excluding zero), but these patterns attenuate and become statistically indistinguishable from zero by 2023.
+Taken together, the results suggest that any initial pro-commodification response to upward mobility and de-commodifying response to downward mobility are not stable over time, and the causal effect of mobility on preferences for pension commodification is effectively time-invariant over the 2016–2023 period.
 

### 2025-11-21 paper/07-supplementary-material.qmd
up antes de borrar tiempo

@@
 #| label: tbl-complete3
-#| tbl-cap: "Interactions effects between intergenerational occupational mobility and perceived meritocracy on preferences for pension commodification, with covariates and wave fixed effects."
+#| tbl-cap: "Marginal effects of mobility by meritocratic perception (Low vs High Meritocracy)."
 #| tbl-cap-location: top
@@
 
-ccoef <- list(
-  "(Intercept)" = "Intercept",
-  "t" = "Mobility treatment",
-  "mHigh" = "High meritocracy perception (Ref.= Low)",
-  "t:mHigh" = "Mobility treatment x High meritocracy perception (Ref.= Low)"
-)
+df_tab <- df_pension_lab_i %>%
+  mutate(
+    Trajectory = sub("\\n.*", "", name),
+    Merit = term,
+    Estimate = round(estimate, 3),
+    `Std. Error` = round(std.error, 3),
+    `CI Low` = round(conf.low, 3),
+    `CI High` = round(conf.high, 3),
+  ) %>%
+  select(Trajectory, Merit, Estimate, `Std. Error`, `CI Low`, `CI High`)
 
-texreg(list(mp_lm_i,
-               mp_lh_i,
-               mp_mh_i,
-               mp_ml_i,
-               mp_hm_i,
-               mp_hl_i),
-          custom.model.names = c(
-            "Low-Middle",
-            "Low-High",
-            "Middle-High",
-            "Middle-Low",
-            "High-Middle",
-            "High-Low"
-          ),
-          caption.above = NULL,
+kableExtra::kable(df_tab,
+      format = table_format,
       caption = NULL,
-          custom.coef.map = ccoef,
-          digits = 2,
-          custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
-          leading.zero = T,
-          use.packages = F,
-          booktabs = F,
-          scalebox = 0.7,
-          center = T,float.pos = "H!",
-          custom.gof.rows = list("Controls"=rep("Yes",6)))
+      align = "lcccccc") %>%
+kableExtra::kable_classic(full_width = F, html_font = "Cambria")
+```
+
+```{r}
+#| label: tbl-complete4
+#| tbl-cap: "Marginal effects of mobility by survey wave (2016, 2018, and 2023)."
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+df_tab <- df_pension_lab_t %>%
+  mutate(
+    Trajectory = sub("\\n.*", "", name),
+    Wave = term,
+    Estimate = round(estimate, 3),
+    `Std. Error` = round(std.error, 3),
+    `CI Low` = round(conf.low, 3),
+    `CI High` = round(conf.high, 3),
+  ) %>%
+  select(Trajectory, Wave, Estimate, `Std. Error`, `CI Low`, `CI High`)
+
+kableExtra::kable(df_tab,
+      format = table_format,
+      caption = NULL,
+      align = "lcccccc") %>%
+  kableExtra::kable_classic(full_width = F, html_font = "Cambria")
 

### 2025-11-21 paper/paper 2.tex
up antes de borrar tiempo

--- 2025-11-17 paper/paper.tex
@@
 \setstretch{1.5}
-This document was last modified at 2025-11-17 19:08:04 and it was last
-rendered at 2025-11-17 19:08:04.
+This document was last modified at 2025-11-19 13:57:11 and it was last
+rendered at 2025-11-19 13:57:11.
 
@@
 
-\$\$
-
-ATT\_\{j,k,j\} = E\big[Y(D = k) \mid O = j, D = k \big] -
-E\big[Y(D = j) \mid O = j, D = k \big]
-
-\$\$ \{\#eq-att\}
+\begin{equation}\phantomsection\label{eq-att}{ 
+ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
+}\end{equation}
 
@@
 mobility versus immobility, estimated from interaction models (T ×
-Merit). Points show simple slopes of the mobility effect at low merit
-and high merit; bars are 95\% confidence intervals. Estimates from
-pooled WLS with doubly robust adjustment, wave fixed effects, and
-standard errors clustered by individual.}
+Merit). Points show the marginal effects (simple slopes) of the mobility
+treatment at low and high merit; bars are 95\% confidence intervals.
+Estimates from pooled WLS with doubly robust adjustment, wave fixed
+effects, and standard errors clustered by individual.}
 
@@
 
-Figure~\ref{fig-interact} shows that neither the interaction terms nor
-the mobility coefficients reach conventional significance across
-trajectories, indicating no detectable moderation of the causal mobility
-effect by meritocratic perceptions at the 95\% level. Directionally, in
-the left panel, the interaction is positive for Middle→High (\(\beta\) =
-0.15, 95\% CI {[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05,
-{[}−0.14, 0.24{]}), suggesting, if anything, somewhat larger
-pro-commodification effects among high-merit respondents when upwardly
-mobile, whereas it is negative for High→Low (\(\beta\) = −0.14,
-{[}−0.38, 0.10{]}) and High→Middle (\(\beta\) = −0.03, {[}−0.23,
-0.17{]}), showing weaker pro-commodification effects under downward
-mobility at high merit. Focusing on the treatment effect at low
-meritocracy (the reference category) at the right panel, the mobility
-coefficient captures the origin-specific ATT comparing movers to their
-immobility counterfactual among low-merit respondents. These effects are
-small and imprecise across trajectories: Low→Middle (\(\beta\) = −0.02,
-{[}−0.09, 0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]})
-are near zero. Middle→High is positive but not significant (\(\beta\) =
-0.07, {[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00,
-{[}−0.07, 0.06{]}), and downward from high shows negative, borderline
-estimates: High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and
-High→Low (\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not
-find evidence that meritocratic perceptions moderate the causal effect
-of mobility, although the signs align with the idea that upward mobility
-at high merit tilts attitudes toward commodification, and downward
-mobility at high merit tilts them away.
-
-\begin{table}
-
-\caption{\label{tbl-inttime}Effects of intergenerational occupational
-mobility on preferences for pension commodification, by survey wave}
+Figure~\ref{fig-interact} displays the origin-specific ATT estimates
+from the meritocracy interaction models as marginal effects of the
+mobility treatment at low and high meritocratic beliefs. For low-merit
+respondents, mobility effects are uniformly small and imprecise across
+trajectories: Low→Middle (\(\beta\) = −0.02, 95\% CI {[}−0.09, 0.04{]}),
+Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}), Middle→Low (\(\beta\) =
+−0.00, {[}−0.07, 0.06{]}), and a modest, non-significant positive
+estimate for Middle→High (\(\beta\) = 0.07, {[}−0.01, 0.14{]}), while
+downward moves from high origin show borderline negative effects
+(High→Middle: \(\beta\) = −0.07, {[}−0.15, 0.00{]}; High→Low: \(\beta\)
+= −0.08, {[}−0.15, 0.00{]}). Among high-merit respondents, the pattern
+is directionally consistent with meritocratic moderation but
+statistically robust only for one trajectory: the ATT for Middle→High at
+high merit is positive and significant (\(\beta\) = 0.22, {[}0.05,
+0.38{]}), whereas Low→Middle (\(\beta\) = 0.02, {[}−0.13, 0.16{]}),
+Low→High (\(\beta\) = 0.02, {[}−0.15, 0.20{]}), High→Middle (\(\beta\) =
+−0.10, {[}−0.28, 0.09{]}), and High→Low (\(\beta\) = −0.22, {[}−0.44,
+0.00{]}) remain statistically indistinguishable from zero. Overall,
+these results provide only limited and trajectory-specific evidence that
+meritocratic beliefs moderate the causal effect of intergenerational
+occupational mobility on preferences for pension commodification, with a
+clearer pro-commodification sign only for upward mobility from middle to
+high status among high-merit respondents.
+
+\begin{figure}[H]
 
@@
 
-\begin{center}
-\scalebox{0.8}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
-\hline
-Intercept                              & $0.18$           & $-0.04$          & $0.28$           & $0.20$           & $0.18$            & $0.15$            \\
-                                       & $ [-0.36; 0.73]$ & $ [-0.36; 0.29]$ & $ [-0.01; 0.57]$ & $ [-0.11; 0.52]$ & $ [-0.15;  0.50]$ & $ [-0.27;  0.58]$ \\
-Mobility treatment                     & $-0.05$          & $-0.03$          & $0.13^{*}$       & $0.06$           & $-0.19^{*}$       & $-0.19^{*}$       \\
-                                       & $ [-0.16; 0.06]$ & $ [-0.15; 0.10]$ & $ [ 0.05; 0.22]$ & $ [-0.03; 0.15]$ & $ [-0.33; -0.05]$ & $ [-0.32; -0.05]$ \\
-Wave (Ref.= 2016)                      &                  &                  &                  &                  &                   &                   \\
-                                       &                  &                  &                  &                  &                   &                   \\
-\quad Wave 2018                        & $0.01$           & $0.01$           & $0.10^{*}$       & $0.09^{*}$       & $-0.04$           & $-0.02$           \\
-                                       & $ [-0.10; 0.12]$ & $ [-0.10; 0.12]$ & $ [ 0.03; 0.18]$ & $ [ 0.01; 0.17]$ & $ [-0.15;  0.07]$ & $ [-0.15;  0.10]$ \\
-\quad Wave 2023                        & $0.03$           & $0.01$           & $0.19^{*}$       & $0.18^{*}$       & $0.05$            & $0.04$            \\
-                                       & $ [-0.09; 0.14]$ & $ [-0.12; 0.13]$ & $ [ 0.10; 0.28]$ & $ [ 0.08; 0.27]$ & $ [-0.09;  0.18]$ & $ [-0.11;  0.19]$ \\
-Mobility treatment x Wave (Ref.= 2016) &                  &                  &                  &                  &                   &                   \\
-                                       &                  &                  &                  &                  &                   &                   \\
-\quad Mobility treatment x Wave 2018   & $0.03$           & $-0.06$          & $-0.04$          & $-0.09$          & $0.11$            & $0.06$            \\
-                                       & $ [-0.11; 0.17]$ & $ [-0.21; 0.09]$ & $ [-0.17; 0.09]$ & $ [-0.20; 0.03]$ & $ [-0.04;  0.26]$ & $ [-0.09;  0.22]$ \\
-\quad Mobility treatment x Wave 2023   & $0.07$           & $0.10$           & $-0.09$          & $-0.07$          & $0.19^{*}$        & $0.18$            \\
-                                       & $ [-0.09; 0.23]$ & $ [-0.08; 0.28]$ & $ [-0.23; 0.06]$ & $ [-0.22; 0.08]$ & $ [ 0.01;  0.38]$ & $ [-0.01;  0.37]$ \\
-\hline
-Controls                               & Yes              & Yes              & Yes              & Yes              & Yes               & Yes               \\
-R$^2$                                  & $0.04$           & $0.05$           & $0.07$           & $0.03$           & $0.06$            & $0.06$            \\
-Adj. R$^2$                             & $0.03$           & $0.04$           & $0.06$           & $0.02$           & $0.05$            & $0.05$            \\
-Num. obs.                              & $878$            & $745$            & $716$            & $773$            & $912$             & $861$             \\
-RMSE                                   & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$            & $0.42$            \\
-N Clusters                             & $497$            & $449$            & $443$            & $470$            & $536$             & $525$             \\
-\hline
-\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
-\end{tabular}
+\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-wave-1.pdf}
+
 }
-\label{table:coefficients}
-\end{center}
-
-}
-
-\end{table}%
-
-Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
-effect across the different waves of the survey. Regarding temporal
-heterogeneity, the only statistically significant interaction term is
-observed for the High→Middle trajectory in 2023
-(\(\beta_{\text{int}} = 0.19\)), 95\% CI {[}0.01, 0.38{]}). However,
-when this interaction is added to the baseline effect in 2016 (-0.19),
-the marginal effect of High→Middle mobility in 2023 is approximately
-zero, and its confidence interval includes the null. Thus, rather than
-indicating a reversal toward a positive effect, the evidence suggests
-that the negative impact of downward mobility from high backgrounds is
-confined to 2016 and attenuates over time. No comparable temporal
-changes are observed in the remaining trajectories. Overall, the results
-provide little support for strong time-conditional heterogeneity in
-mobility effects, beyond the fading of the initially negative
-High→Middle effect.
+
+\caption{\label{fig-interact-wave}Effects of intergenerational
+occupational mobility on preferences for pension commodification, by
+survey wave. Coefficient plot of origin-specific ATT estimates comparing
+mobility versus immobility, estimated from interaction models (T ×
+Wave). Points show the marginal effects (simple slopes) of the mobility
+treatment in 2016, 2018, and 2023; bars are 95\% confidence intervals.
+Estimates from pooled WLS with doubly robust adjustment, wave fixed
+effects, and standard errors clustered by individual.}
+
+\end{figure}%
+
+I find no consistent evidence of heterogeneous causal effects across
+survey waves. Figure~\ref{fig-interact-wave} displays the marginal
+effects of mobility trajectories on market-justice preferences for
+pensions in 2016, 2018, and 2023. Across trajectories, the estimated
+ATTs remain small and their confidence intervals largely overlap over
+time. Upward mobility from middle origin shows a positive and
+statistically significant effect in 2016 (Middle→High, \(\beta\) = 0.13,
+95\% CI {[}0.05, 0.22{]}) and downward mobility from high origin
+displays negative effects at baseline (High→Middle and High→Low, both
+\(\beta\) ≈ −0.19, CIs excluding zero), but these patterns attenuate and
+become statistically indistinguishable from zero by 2023. Taken
+together, the results suggest that any initial pro-commodification
+response to upward mobility and de-commodifying response to downward
+mobility are not stable over time, and the causal effect of mobility on
+preferences for pension commodification is effectively time-invariant
+over the 2016--2023 period.
 
@@
 
-\caption{\label{tbl-complete3}Interactions effects between
-intergenerational occupational mobility and perceived meritocracy on
-preferences for pension commodification, with covariates and wave fixed
-effects.}
+\caption{\label{tbl-complete3}Marginal effects of mobility by
+meritocratic perception (Low vs High Meritocracy).}
 
@@
 
-\begin{center}
-\scalebox{0.7}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
-\hline
-Intercept                                                    & $0.15$           & $-0.04$          & $0.30^{*}$       & $0.21$           & $0.08$           & $0.05$           \\
-                                                             & $ [-0.36; 0.67]$ & $ [-0.40; 0.31]$ & $ [ 0.04; 0.56]$ & $ [-0.09; 0.51]$ & $ [-0.25; 0.40]$ & $ [-0.30; 0.41]$ \\
-Mobility treatment                                           & $-0.02$          & $-0.03$          & $0.07$           & $-0.00$          & $-0.07$          & $-0.08$          \\
-                                                             & $ [-0.09; 0.04]$ & $ [-0.10; 0.05]$ & $ [-0.01; 0.14]$ & $ [-0.07; 0.06]$ & $ [-0.15; 0.00]$ & $ [-0.15; 0.00]$ \\
-High meritocracy perception (Ref.= Low)                      & $0.05$           & $0.06$           & $-0.01$          & $0.03$           & $0.17^{*}$       & $0.26^{*}$       \\
-                                                             & $ [-0.06; 0.16]$ & $ [-0.07; 0.19]$ & $ [-0.09; 0.08]$ & $ [-0.07; 0.13]$ & $ [ 0.00; 0.33]$ & $ [ 0.04; 0.48]$ \\
-Mobility treatment x High meritocracy perception (Ref.= Low) & $0.04$           & $0.05$           & $0.15$           & $0.04$           & $-0.03$          & $-0.14$          \\
-                                                             & $ [-0.12; 0.20]$ & $ [-0.14; 0.24]$ & $ [-0.03; 0.33]$ & $ [-0.11; 0.19]$ & $ [-0.23; 0.17]$ & $ [-0.38; 0.10]$ \\
-\hline
-Controls                                                     & Yes              & Yes              & Yes              & Yes              & Yes              & Yes              \\
-R$^2$                                                        & $0.04$           & $0.05$           & $0.08$           & $0.04$           & $0.07$           & $0.10$           \\
-Adj. R$^2$                                                   & $0.03$           & $0.04$           & $0.07$           & $0.02$           & $0.06$           & $0.08$           \\
-Num. obs.                                                    & $878$            & $745$            & $716$            & $773$            & $912$            & $861$            \\
-RMSE                                                         & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$           & $0.41$           \\
-N Clusters                                                   & $497$            & $449$            & $443$            & $470$            & $536$            & $525$            \\
-\hline
-\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\centering
+\begin{tabular}{l|c|c|c|c|c}
+\hline
+Trajectory & Merit & Estimate & Std. Error & CI Low & CI High\\
+\hline
+Low-Middle & Low Meritocracy & -0.024 & 0.034 & -0.091 & 0.042\\
+\hline
+Low-Middle & High Meritocracy & 0.015 & 0.072 & -0.127 & 0.157\\
+\hline
+Low-High & Low Meritocracy & -0.029 & 0.038 & -0.104 & 0.046\\
+\hline
+Low-High & High Meritocracy & 0.024 & 0.088 & -0.148 & 0.195\\
+\hline
+Middle-High & Low Meritocracy & 0.066 & 0.038 & -0.009 & 0.140\\
+\hline
+Middle-High & High Meritocracy & 0.215 & 0.084 & 0.051 & 0.379\\
+\hline
+Middle-Low & Low Meritocracy & -0.004 & 0.034 & -0.071 & 0.063\\
+\hline
+Middle-Low & High Meritocracy & 0.038 & 0.068 & -0.095 & 0.171\\
+\hline
+High-Middle & Low Meritocracy & -0.074 & 0.039 & -0.150 & 0.003\\
+\hline
+High-Middle & High Meritocracy & -0.100 & 0.094 & -0.285 & 0.085\\
+\hline
+High-Low & Low Meritocracy & -0.077 & 0.039 & -0.153 & 0.000\\
+\hline
+High-Low & High Meritocracy & -0.216 & 0.112 & -0.436 & 0.005\\
+\hline
 \end{tabular}
+
 }
-\label{table:coefficients}
-\end{center}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-complete4}Marginal effects of mobility by survey
+wave (2016, 2018, and 2023).}
+
+\centering{
+
+\centering
+\begin{tabular}{l|c|c|c|c|c}
+\hline
+Trajectory & Wave & Estimate & Std. Error & CI Low & CI High\\
+\hline
+Low-Middle & 2016 & -0.048 & 0.055 & -0.156 & 0.060\\
+\hline
+Low-Middle & 2018 & -0.022 & 0.048 & -0.116 & 0.072\\
+\hline
+Low-Middle & 2023 & 0.022 & 0.058 & -0.092 & 0.135\\
+\hline
+Low-High & 2016 & -0.025 & 0.062 & -0.146 & 0.096\\
+\hline
+Low-High & 2018 & -0.088 & 0.050 & -0.186 & 0.011\\
+\hline
+Low-High & 2023 & 0.072 & 0.066 & -0.058 & 0.202\\
+\hline
+Middle-High & 2016 & 0.132 & 0.044 & 0.047 & 0.218\\
+\hline
+Middle-High & 2018 & 0.091 & 0.051 & -0.008 & 0.190\\
+\hline
+Middle-High & 2023 & 0.045 & 0.061 & -0.074 & 0.164\\
+\hline
+Middle-Low & 2016 & 0.061 & 0.044 & -0.025 & 0.147\\
+\hline
+Middle-Low & 2018 & -0.027 & 0.046 & -0.118 & 0.064\\
+\hline
+Middle-Low & 2023 & -0.010 & 0.057 & -0.122 & 0.103\\
+\hline
+High-Middle & 2016 & -0.187 & 0.070 & -0.324 & -0.050\\
+\hline
+High-Middle & 2018 & -0.078 & 0.050 & -0.175 & 0.019\\
+\hline
+High-Middle & 2023 & 0.006 & 0.063 & -0.118 & 0.130\\
+\hline
+High-Low & 2016 & -0.186 & 0.068 & -0.319 & -0.053\\
+\hline
+High-Low & 2018 & -0.125 & 0.053 & -0.228 & -0.022\\
+\hline
+High-Low & 2023 & -0.009 & 0.068 & -0.141 & 0.124\\
+\hline
+\end{tabular}
 

### 2025-11-21 paper/paper.tex
up antes de borrar tiempo

@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 \setstretch{1.5}
-This document was last modified at 2025-11-17 19:08:04 and it was last
-rendered at 2025-11-17 19:08:04.
+This document was last modified at 2025-11-21 16:32:59 and it was last
+rendered at 2025-11-21 16:32:59.
 
@@
 
-What is the legitimate extent of market inequality in the eyes of the
+What is the legitimate scope of market inequality in the eyes of the
 public? Since the early 1980s, many countries have experienced a
-widespread retreat from universal welfare programs and a shift toward
-the privatization and commodification of public goods, welfare policies,
-and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
-2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
-America, as elsewhere, neoliberal reforms reshaped welfare-state
-institutions by extending market logic into domains of social
-reproduction that were traditionally governed by the state
-(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
-reduced the role of public provision and increased the presence of
-private actors in core social services
-(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Echoing Polanyi's
-(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
-constitute a distinct moral order, the institutional diffusion of market
-rules has fostered a corresponding moral economy: a constellation of
-norms and values concerning fair allocation, embedded in institutions
-and shaping individual subjectivities
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
-this framework, a growing body of research examines the extent to which,
-and the mechanisms by which, citizens consider it fair that the
-allocation of services like health care, pensions, and education be
-governed by market-based criteria---a phenomenon known as \emph{market
-justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
-2014}; \citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et
-al., 2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
-2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+rollback of universal welfare programs and a shift toward the
+privatisation and commodification of public goods and social services
+(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
+\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
+reforms extended market logic into domains of social reproduction long
+governed by the state, shrinking public provision and expanding private
+actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
+2023}). From a moral-economy perspective, this diffusion of market rules
+has reconfigured the normative order: through policy feedback, welfare
+institutions increasingly embed market-based criteria of ``fair''
+allocation and shape how citizens understand deservingness and the
+balance between state and market (\citeproc{ref-koos_moral_2019}{Koos \&
+Sachweh, 2019};
+\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
+Against this backdrop, growing scholarship examines \emph{market justice
+preferences}---the extent to which citizens regard market-based criteria
+as a fair basis for access to essential services such as healthcare,
+education and pensions
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
+orientations matter because they help legitimise economic inequality by
+portraying stratified outcomes as a fair result of individual
+responsibility and limited state intervention.
+
+This article examines public support for the commodification of pension
+welfare---specifically, whether people consider it fair that access to
+better pensions depends on income or contributions. Pension
+privatisation lies at the core of Latin America's trajectory of
+commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
+2000}), making pension systems a strategic domain for analysing how
+market principles become normatively embedded
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As many countries
+expanded mandatory or voluntary individual-capitalisation schemes---most
+radically in Chile---pensions became organised around contribution-based
+entitlements and market performance
+(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
+\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
+reshapes not only how pension institutions allocate old-age risk,
+heightening exposure to poverty in retirement and tying benefit levels
+to lifetime labour-market trajectories
+(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
+\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also reshape the
+justice principles citizens associate with old-age security: people have
+to judge whether higher benefits should follow market-derived criteria
+or solidaristic principles
+(\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
+2019}; \citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Although
+we know something about what people consider a ``just'' monetary pension
+(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
+and about support for public versus private provision
+(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
+\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
+still lack systematic evidence on how citizens assess the fairness of
+market-dependent access to pensions.
+
+Research shows that support for market-based access to welfare services
+is strongly stratified: individuals with higher income, education or
+class position are more likely to endorse market-based criteria than
+disadvantaged groups (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
+2014}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
 \citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
-Understanding these preferences is crucial, as they contribute to
-legitimizing economic inequality by framing it as the fair result of
-individual responsibility and limited state intervention
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
-
-Existing literature shows that market justice preferences are shaped by
-both the economic and institutional context and individuals' positions
-within social stratification. Grounded in the notion that economic
-institutions influence people's normative attitudes
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}), studies find
-that countries with stronger public provision or more expansive welfare
-states exhibit lower levels of market justice preferences
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
-more privatized contexts show stronger support for market-based criteria
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
-justice preferences tend to rise as individuals ``ascend'' the social
-structure, with those in more privileged positions in terms of class,
-education, and income holding stronger preferences for market-based
-solutions compared to those in more disadvantaged or at-risk positions
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
-\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
-2016}).
-
-Market justice preferences are shaped not only by objective
-socioeconomic conditions but also by popular beliefs about inequality.
-Among these, meritocracy is a key normative principle underpinning
-market-based distributive preferences
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Studies show that
-individuals with stronger meritocratic beliefs tend to perceive less
-inequality and legitimize it by attributing economic differences to
-personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
-societies where access to services is largely governed by market logic,
-such beliefs play a critical role in normalizing inequality. Recent
-evidence from Chile shows that students who believe effort and talent
-are rewarded in their country express stronger preferences for
-market-based access to healthcare, pensions, and education
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
-
-Although it is clear that one's social position influences market
-justice preferences, the question of how upward or downward mobility
-within the social structure affects these preferences remains
-unanswered. This question is far from trivial, especially in Latin
-America, where many have experienced various forms of mobility amid high
-economic inequality and deep welfare privatization
-(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
-\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}).
-Social origins and destinations affect attitudes toward inequality in
-distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
-\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b},
-\citeproc{ref-gugushvili_subjective_2017}{2017};
-\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
-2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
-movement between these positions exposes individuals to different
-experiences and mechanisms that shape their views on what is fair
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
+pensions points in the same direction: higher income and greater
+financial gains in funded schemes correlate with stronger support for
+market-oriented pension policies
+(\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
+little about how movement within the class structure shapes preferences
+for pension commodification. This gap is especially salient in Latin
+America, where mobility has unfolded under high inequality and deeply
+privatised welfare regimes
+(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
+2021}). Social origins and destinations anchor material interests and
+justice principles
+(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
+\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
+\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
+while movement between these positions exposes individuals to new risks,
+resources and normative environments that reshape fairness judgements
 (\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
-\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
-research, examining the effects of social mobility on market justice
-preferences can help to illuminate how inequalities in access to social
-services are justified among individuals who have experienced, or not,
-changes in their social standing, and what are the normative mechanisms
-that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
-2015}).
-
-Beyond their isolated effects, social mobility and meritocratic beliefs
-interact in complex ways to shape market justice preferences. Among
-others, a key mechanism proposed in the literature to explain how
-mobility influences distributive justice preferences is the
-psychological process of self-serving bias in causal attribution
+\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). As Mau
+(\citeproc{ref-mau_inequality_2015}{2015}) argues, social mobility can
+foster market-justice orientations, helping explain how inequalities in
+access to welfare services become normatively sustained.
+
+The literature identifies several mechanisms through which mobility may
+shape distributive-justice preferences---material self-interest
+(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}), socialisation
+and acculturation
+(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}), learning through long-term exposure
+(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}), and
+cognitive attributions such as self-serving bias or privilege awareness
 (\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
-suggests that individuals attribute failures---such as downward
-mobility---to external factors, while crediting successes---such as
-upward mobility---to their own merit and effort
-(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
-who experience upward mobility tend to view their social position as
-earned, making them more likely to believe that individuals are
-responsible for their own success or failure. Research shows that upward
-mobility is associated with weaker preferences for redistribution
-(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
-\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
-2011}) and stronger legitimacy of income inequality
-(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
-individuals who experience downward mobility tend to blame structural
-factors like inequality and are more supportive of redistribution while
-rejecting merit-based explanations
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
-together, I argue that meritocratic beliefs may reinforce this
-self-serving attribution mechanism by legitimizing one's social status
-as the outcome of personal merit, closely tied to attribution bias.
-
-Against this background, this article pursues two main objectives:
-first, to analyze the extent to which intergenerational social mobility
-influences market justice preferences regarding healthcare, pensions,
-and education; and second, to examine how meritocratic beliefs may
-moderate this relationship. Building on a theoretical framework that
-emphasizes how neoliberal transformations---particularly through the
-privatization and commodification of key areas of social
-reproduction---have profoundly reshaped processes of subject formation
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}), the central argument is
-that upward mobility increases support for market justice preferences,
-while downward mobility decreases it. Moreover, meritocratic beliefs are
-expected to moderate this relationship by reflecting a self-serving
-attribution bias, whereby individuals justify their social position in
-terms of personal merit.
-
-This study focuses on Chile, a particularly intriguing case for
-examining market justice preferences. Despite significant economic
-growth and poverty reduction, Chile has some of the highest levels of
-inequality in Latin America and among OECD countries
-(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
-\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
-coexists with short-range upward mobility among lower-class segments
-moving into middle strata, though strong barriers remain to reaching
-higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
-et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
-2014}). What makes Chile especially salient is that much of this
-inequality is rooted in deep neoliberal reforms that institutionalized
-the privatization and commodification of key social sectors
-(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
-during the dictatorship (1973--1989) and expanded in democracy, these
-reforms enabled the unprecedented emergence of markets in health,
-pensions, and education, with provision segmented by individuals'
-ability to pay and supported by public subsidies
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
-despite waves of protest against inequality and commodification from
-2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
-2021})---Chilean subjectivities have been increasingly shaped by
-neoliberal discourses and market logics, influencing their attitudes
-toward inequality and welfare distribution
-(\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli, 2012};
-\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
-
-In this context, the questions that guide this research are as follows:
-
-\begin{enumerate}
-\def\labelenumi{(\arabic{enumi})}
-\tightlist
-\item
-  To what extent does intergenerational social mobility influence market
-  justice preferences regarding healthcare, pensions, and education in
-  Chile?
-\item
-  How do meritocratic beliefs condition or moderate this relationship in
-  the Chilean context?
-\end{enumerate}
-
-To address these questions, this study draws on large-scale,
-representative survey data collected in 2018 from the urban Chilean
-population aged 18 to 75 (n = 2,726). The next section outlines the
-theoretical framework linking market justice preferences, social
-mobility, and meritocratic beliefs, and proposes a set of hypotheses.
-This is followed by a description of the data, variables, and analytical
-strategy. The final sections present the empirical findings, offer an
-interpretation of the results, and conclude with a discussion of their
-implications.
+\citeproc{ref-molina_its_2019}{Molina et al., 2019}). While distinct,
+these accounts converge in highlighting two broad channels relevant for
+market-based welfare: shifts in material stakes and shifts in normative
+frames (\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In this paper, I
+do not treat these mechanisms as mediators but use them to guide
+hypotheses about heterogeneity: if mobility alters preferences through
+material interests and normative updating, its effects should be most
+potent among those who interpret social outcomes through merit-based
+evaluations of effort
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).
+
+Against this background, this article pursues two aims: first, to assess
+whether intergenerational mobility causally affects support for the
+commodification of pensions; and second, to examine whether this effect
+varies by meritocratic beliefs and over time. I argue that mobility
+works through two main channels. Upward mobility increases earnings,
+expected returns in a funded system and perceived economic security,
+making contribution-based access more appealing. At the same time,
+integration into higher-class networks exposes people to more
+individualistic norms and effort-based understandings of fairness.
+Downward mobility, in turn, heightens economic insecurity, dependence on
+collective protection and attachment to solidaristic principles,
+lowering support for market allocation. Meritocratic beliefs are
+conceptualised not as mediators but as moderators. Among those who see
+mobility as a fair reward for effort, both material and normative
+effects should be amplified, producing higher support for pension
+commodification. The paper contributes to research on market justice by
+focusing on a specific welfare domain and by providing causal evidence
+on how mobility shapes distributive preferences.
+
+This study focuses on Chile, a particularly instructive case for
+examining preferences for pension commodification. Despite sustained
+economic growth, Chile combines high and persistent inequality with
+short-range upward mobility and substantial barriers to higher-class
+positions (\citeproc{ref-flores_top_2020}{Flores et al., 2020};
+\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
+\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
+first country worldwide to fully replace a public pay-as-you-go system
+with a mandatory, privately administered (through pension fund
+administrators, AFPs) defined-contribution scheme, while the state has
+progressively expanded a segmented solidarity pillar to compensate for
+insufficient benefits (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
+\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
+despite waves of protest from 2006 to 2019 targeting the AFP system
+(\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
+subjectivities have been increasingly shaped by neoliberal discourses
+and market logics, influencing attitudes toward pension distribution and
+inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
+2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
+profound inequality, market-driven old-age security and contested
+legitimacy makes Chile a critical case for analysing how mobility and
+meritocratic beliefs structure support for pension commodification.
+
+To address these questions, the study draws on large-scale,
+representative survey data for the urban Chilean population collected in
+2016, 2018 and 2023. It adopts a causal framework of mobility effects
+inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
+The following section outlines the theoretical framework linking
+preferences for market-based welfare, social mobility and expected
+heterogeneity by meritocratic beliefs and time, and derives hypotheses.
+A subsequent section presents the causal identification strategy, the
+data, the variables, and the analytical approach. The empirical sections
+report the findings and interpret their implications. The article
+concludes by discussing what these results reveal about the politics of
+market-based pension welfare and social mobility.
 
@@
 legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
-Conversely, if they see outcomes as driven by luck, social origin, or
-systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural
-inequality, where public narratives often emphasize personal
+\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
+2019}). Conversely, if they see outcomes as driven by luck, social
+origin, or systemic barriers, inequality is more likely to be perceived
+as unjust. This distinction becomes crucial in societies with persistent
+structural inequality, where public narratives often emphasize personal
 responsibility and merit while overlooking entrenched disadvantages.
@@
 which are associated with lower perceived injustice
-(\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
-Mac-Clure et al. (\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et
-al., 2024}) find that low and lower-middle status individuals are less
-likely than upper-status groups to view educational income differences
-as unfair. Qualitative research further reveals the internalization of
-the enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
+(\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
+2019}). Mac-Clure et al.
+(\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}) find
+that low and lower-middle status individuals are less likely than
+upper-status groups to view educational income differences as unfair.
+Qualitative research further reveals the internalization of the
+enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
 understood as a way in which the subjectivization process arises in line
@@
 
+\bibitem[\citeproctext]{ref-arenas_sistemas_2019}
+Arenas, A. (2019). \emph{{Los sistemas de pensiones en la encrucijada:
+desafíos para la sostenibilidad en América Latina}}. Santiago: Naciones
+Unidas, CEPAL.
+
 \bibitem[\citeproctext]{ref-ares_changing_2020}
@@
 \url{https://doi.org/10.1080/01402382.2019.1644575}
+
+\bibitem[\citeproctext]{ref-arrizabalomontoro_debate_2019}
+Arrizabalo Montoro, X., Del Rosal, M., \& Javier Murillo Arroyo, F.
+(2019). The {Debate} on {Pension Systems}: {The Paradigmatic Cases} of
+{Chile} and {Spain}. \emph{The American Journal of Economics and
+Sociology}, \emph{78}(1), 195--223.
+\url{https://doi.org/10.1111/ajes.12262}
 
@@
 
+\bibitem[\citeproctext]{ref-borzutzky_pension_2012}
+Borzutzky, S. (2012). Pension market failure in {Chile}: Foundations,
+analysis and policy reforms. \emph{Journal of Comparative Social
+Welfare}, \emph{28}(2), 103--112.
+\url{https://doi.org/10.1080/17486831.2012.655980}
+
 \bibitem[\citeproctext]{ref-breen_effects_2024}
@@
 \bibitem[\citeproctext]{ref-castillo_perceptions_2025}
-Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
-Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
-{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
-\emph{Under Review at Frontiers in Sociology}.
+Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra-Sanhueza, J.
+(2025). Perceptions of inequality and meritocracy: Their interplay in
+shaping preferences for market justice in {Chile} (2016--2023).
+\emph{Frontiers in Sociology}, \emph{10}, 1634219.
+\url{https://doi.org/10.3389/fsoc.2025.1634219}
 
@@
 
+\bibitem[\citeproctext]{ref-castillo_deserving_2019}
+Castillo, J. C., Olivos, F., \& Azar, A. (2019). Deserving a {Just
+Pension}: {A Factorial Survey Approach}*. \emph{Social Science
+Quarterly}, \emph{100}(1), 359--378.
+\url{https://doi.org/10.1111/ssqu.12539}
+
 \bibitem[\citeproctext]{ref-castillo_socialization_2024}
@@
 \url{https://doi.org/10.1111/roiw.12441}
+
+\bibitem[\citeproctext]{ref-galvez_con_2023}
+Gálvez, R. (2023). Con mi plata no\ldots{} alcanza. \emph{CNN Chile}.
 
@@
 
+\bibitem[\citeproctext]{ref-huber_political_2000}
+Huber, E., \& Stephens, J. D. (2000). \emph{The political economy of
+pension reform: {Latin America} in comparative perspective}. Geneva:
+UNRISD.
+
 \bibitem[\citeproctext]{ref-immergut_theoretical_1998}
@@
 
+\bibitem[\citeproctext]{ref-jaime-castillo_public_2013}
+Jaime-Castillo, A. M. (2013). Public opinion and the reform of the
+pension systems in {Europe}: The influence of solidarity principles.
+\emph{Journal of European Social Policy}, \emph{23}(4), 390--405.
+\url{https://doi.org/10.1177/0958928713507468}
+
 \bibitem[\citeproctext]{ref-jaime-castillo_social_2019}
@@
 
+\bibitem[\citeproctext]{ref-kerner_pension_2020}
+Kerner, A. (2020). Pension {Returns} and {Popular Support} for
+{Neoliberalism} in {Post-Pension Reform Latin America}. \emph{British
+Journal of Political Science}, \emph{50}(2), 585--620.
+\url{https://doi.org/10.1017/S0007123417000710}
+
 \bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
@@
 
+\bibitem[\citeproctext]{ref-madero-cabib_private_2019}
+Madero-Cabib, I., Biehl, A., Sehnbruch, K., Calvo, E., \& Bertranou, F.
+(2019). Private {Pension Systems Built} on {Precarious Foundations}: {A
+Cohort Study} of {Labor-Force Trajectories} in {Chile}. \emph{Research
+on Aging}, \emph{41}(10), 961--987.
+\url{https://doi.org/10.1177/0164027519874687}
+
 \bibitem[\citeproctext]{ref-mau_inequality_2015}
@@
 
+\bibitem[\citeproctext]{ref-oecd_pensions_2023}
+OECD. (2023). \emph{Pensions at a {Glance} 2023: {OECD} and {G20
+Indicators}}. OECD Publishing. \url{https://doi.org/10.1787/678055dd-en}
+
 \bibitem[\citeproctext]{ref-otero_power_2024}
@@
 \emph{11}(3), 373--380. \url{https://doi.org/10.1177/1745691616635596}
+
+\bibitem[\citeproctext]{ref-solimano_rise_2021}
+Solimano, A. (2021). \emph{The {Rise} and {Fall} of the {Privatized
+Pension System} in {Chile}: {An International Perspective}} (1st ed).
+London: Anthem Press.
 
@@
 
+\bibitem[\citeproctext]{ref-verbic_political_2019}
+Verbič, M., \& Spruk, R. (2019). Political economy of pension reforms:
+An empirical investigation. \emph{European Journal of Law and
+Economics}, \emph{47}(2), 171--232.
+\url{https://doi.org/10.1007/s10657-018-9606-7}
+
 \bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
@@
 \url{https://doi.org/10.1186/s12939-016-0350-8}
-
-\bibitem[\citeproctext]{ref-wen_does_2021}
-Wen, F., \& Witteveen, D. (2021). Does perceived social mobility shape
-attitudes toward government and family educational investment?
-\emph{Social Science Research}, \emph{98}, 102579.
-\url{https://doi.org/10.1016/j.ssresearch.2021.102579}
 

### 2025-11-21 processing/prod_analysis_multinom.R

@@ -100,7 +100,7 @@ g1 <- db %>%
   geom_col(aes(fill = just_pension_f)) +
   scale_y_continuous(labels = scales::percent) + 
   scale_fill_manual(
-    values = c("grey50", "grey20"),
+    values = c("#DE4968FF", "grey20"),
     breaks = c("Agreement","Disagreement")
   )  +
   coord_flip() +
@@ -114,7 +114,7 @@ g1 <- db %>%
   labs(y = NULL,
        x = NULL,
        fill = NULL,
-       caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)\nNote: Agreement ≥4, Disagreement ≤3") +
+       caption = "Source: own elaboration with data from ELSOC 2016-2023 (N obs = 3,435)\nNote: Agreement 5-4, Disagreement 1-3") +
   my_pretty_theme() +
   theme(legend.position = "bottom") 
 
@@ -168,10 +168,10 @@ g2 <- ggplot(pdat, aes(x = mobility, y = value, group = wave)) +
   scale_color_manual(
     name   = "Wave",
     breaks = breaks_w,
-    values = c("2016" = "#999999",
-               "2018" = "#666666",
-               "2023" = "#000000",
-               "All waves" = "black")) +
+    values = c("2016" = "#DE4968FF",
+               "2018" = "#8C2981FF",
+               "2023" = "#3B0F70FF",
+               "All waves" = "#000004FF")) +
   scale_shape_manual(
     name   = "Wave",
     breaks = breaks_w,
@@ -189,6 +189,7 @@ g2 <- ggplot(pdat, aes(x = mobility, y = value, group = wave)) +
   my_pretty_theme() +
   theme(legend.position = "bottom",
         axis.text.x=element_text(angle=45,hjust = 1))
+
 
 # 3.2 Pooled WLS with IPW-ATT and CR2 -----------------------------------------------------------
 
@@ -411,9 +412,11 @@ df_pension_lab <- df_pension_lab %>%
   )
 
 g3 <- ggplot(df_pension_lab, aes(x = estimate, y = term)) +
-  geom_vline(xintercept = 0, linetype = "solid", color = "darkred", size = 0.5, alpha = 1) +
[…]

### 2025-11-22 paper/01-introduction.qmd
Up V2

@@
 In Latin America, these reforms extended market logic into domains of social reproduction long governed by the state, shrinking public provision and expanding private actors in core welfare sectors [@ferre_welfare_2023].
-From a moral-economy perspective, this diffusion of market rules has reconfigured the normative order: through policy feedback, welfare institutions increasingly embed market-based criteria of “fair” allocation and shape how citizens understand deservingness and the balance between state and market [@svallforsMoralEconomyClass2006a; @koos_moral_2019].
+From a moral-economy perspective, this diffusion of market rules has reconfigured the normative order: through policy feedback, welfare institutions embed market-based criteria of “fair” allocation and shape how citizens understand deservingness and the balance between state and market [@svallforsMoralEconomyClass2006a; @koos_moral_2019].
 Against this backdrop, growing scholarship examines *market justice preferences*—the extent to which citizens regard market-based criteria as a fair basis for access to essential services such as healthcare, education and pensions [@castillo_perceptions_2025; @svallfors_political_2007; @lindh_public_2015].
@@
 This article examines public support for the commodification of pension welfare—specifically, whether people consider it fair that access to better pensions depends on income or contributions.
-Pension privatisation lies at the core of Latin America’s trajectory of commodification [@huber_political_2000], making pension systems a strategic domain for analysing how market principles become normatively embedded [@mau_inequality_2015].
+Pension privatisation lies at the core of Latin America’s trajectory of commodification [@huber_political_2000].
 As many countries expanded mandatory or voluntary individual-capitalisation schemes—most radically in Chile—pensions became organised around contribution-based entitlements and market performance [@arenas_sistemas_2019; @verbic_political_2019].
-This shift reshapes not only how pension institutions allocate old-age risk, heightening exposure to poverty in retirement and tying benefit levels to lifetime labour-market trajectories [@oecd_pensions_2023; @madero-cabib_private_2019], but also reshape the justice principles citizens associate with old-age security: people have to judge whether higher benefits should follow market-derived criteria or solidaristic principles [@arrizabalomontoro_debate_2019; @borzutzky_pension_2012].
+This shift reshapes how pension institutions allocate old-age risk, tying benefit levels to lifetime labour-market trajectories and increasing exposure to poverty in retirement [@oecd_pensions_2023; @madero-cabib_private_2019].
+It also transforms the justice principles citizens associate with old-age security: they must judge whether higher benefits should follow market-derived criteria or solidaristic principles [@arrizabalomontoro_debate_2019; @borzutzky_pension_2012].
 Although we know something about what people consider a “just” monetary pension [@castillo_deserving_2019] and about support for public versus private provision [@busemeyer_welfare_2020; @jaime-castillo_public_2013], we still lack systematic evidence on how citizens assess the fairness of market-dependent access to pensions.
@@
 Social origins and destinations anchor material interests and justice principles [@alesina_intergenerational_2018; @langsaether_explaining_2022; @gugushvili_subjective_2017], while movement between these positions exposes individuals to new risks, resources and normative environments that reshape fairness judgements [@gugushvili_trends_2014; @helgason_longterm_2023].
-As @mau_inequality_2015 argues, social mobility can foster market-justice orientations, helping explain how inequalities in access to welfare services become normatively sustained.
+Social mobility can thus foster market-justice orientations, helping explain how inequalities in access to welfare services become normatively sustained [@mau_inequality_2015].
 
-The literature identifies several mechanisms through which mobility may shape distributive-justice preferences—material self-interest [@benabou_social_2001], socialisation and acculturation [@jaime-castillo_social_2019], learning through long-term exposure [@helgason_longterm_2023], and cognitive attributions such as self-serving bias or privilege awareness [@gugushvili_intergenerational_2016c; @molina_its_2019].
-While distinct, these accounts converge in highlighting two broad channels relevant for market-based welfare: shifts in material stakes and shifts in normative frames [@mau_inequality_2015].
-In this paper, I do not treat these mechanisms as mediators but use them to guide hypotheses about heterogeneity: if mobility alters preferences through material interests and normative updating, its effects should be most potent among those who interpret social outcomes through merit-based evaluations of effort [@castillo_perceptions_2025; @mijs_belief_2022].
+Existing accounts identify several mechanisms through which intergenerational mobility may shape distributive-justice preferences but converge on two channels particularly relevant for market-based welfare: changes in material stakes and shifts in normative frames [@ares_changing_2020; @jaime-castillo_social_2019; @helgason_longterm_2023; @gugushvili_intergenerational_2016c].
+Recent research shows that meritocratic beliefs—seeing that effort and ability are rewarded—are strongly associated with support for market allocation [@castillo_perceptions_2025].
+Perceived upward mobility reinforces meritocratic interpretations of success, whereas downward mobility is more closely linked to structural explanations of inequality without necessarily eroding meritocratic convictions [@mijs_belief_2022].
+Thus, meritocracy and mobility operate on partly separate tracks yet intersect in systematic ways.
+I treat meritocratic beliefs as a post-treatment moderator—a diagnostic lens on potential mechanisms.
+If mobility partly shapes preferences by strengthening merit-based understandings of fairness and altering material stakes, its effects should be most visible—and most pro-market—among individuals who already interpret social outcomes through meritocratic lenses.
 
-Against this background, this article pursues two aims: first, to assess whether intergenerational mobility causally affects support for the commodification of pensions; and second, to examine whether this effect varies by meritocratic beliefs.
-I argue that mobility works through two main channels.
+This article pursues two aims: first, to assess whether intergenerational occupational mobility causally affects support for the commodification of pensions; and second, to examine whether this effect varies by meritocratic beliefs as a diagnostic of possible transmission channels.
+I argue that social mobility operates through two main pathways.
 Upward mobility increases earnings, expected returns in a funded system and perceived economic security, making contribution-based access more appealing.
-At the same time, integration into higher-class networks exposes people to more individualistic norms and effort-based understandings of fairness.
-Downward mobility, in turn, heightens economic insecurity, dependence on collective protection and attachment to solidaristic principles, lowering support for market allocation.
+At the same time, exposure to higher-class networks reinforces individualistic and effort-based understandings of fairness.
+Downward mobility tends to heighten economic insecurity, dependence on public protection and attachment to external attributions of poverty, reducing support for market allocation.
 Meritocratic beliefs are conceptualised not as mediators but as moderators.
-Among those who see mobility as a fair reward for effort, both material and normative effects should be amplified, producing higher support for pension commodification.
-The paper contributes to research on market justice by focusing on a specific welfare domain and by providing causal evidence on how mobility shapes distributive preferences.
+Among individuals with stronger meritocratic beliefs, upward mobility should heighten support for market-based access.
+In contrast, downward mobility should weaken—or even reverse—its usual de-commodifying pattern, as meritocratic interpretations of effort lead individuals to view market outcomes as fair regardless of direction.
+The paper contributes to research on market justice by identifying the causal effects of mobility and by using meritocracy to probe whether these effects align with a meritocratic channel.
 
-This study focuses on Chile, a particularly instructive case for examining preferences for pension commodification.
+The study focuses on Chile, an instructive case for examining preferences for pension commodification.
 Despite sustained economic growth, Chile combines high and persistent inequality with short-range upward mobility and substantial barriers to higher-class positions [@flores_top_2020; @salgado_uplifting_2025; @torche_intergenerational_2014].
-It was the first country worldwide to fully replace a public pay-as-you-go system with a mandatory, privately administered (through pension fund administrators, AFPs) defined-contribution scheme, while the state has progressively expanded a segmented solidarity pillar to compensate for insufficient benefits [@madariaga_three_2020; @boccardo_30_2020; @solimano_rise_2021].
-In parallel—and despite waves of protest from 2006 to 2019 targeting the AFP system [@somma_no_2021]—Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing attitudes toward pension distribution and inequality [@araujo_desafios_2012; @galvez_con_2023].
+It was the first country worldwide to fully replace a public pay-as-you-go system with a mandatory, privately administered defined-contribution scheme, while the state has progressively expanded a segmented solidarity pillar to compensate for insufficient benefits [@madariaga_three_2020; @boccardo_30_2020; @solimano_rise_2021].
+In parallel—and despite waves of protest from 2006 to 2019 against the private pension system [@somma_no_2021]—Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing attitudes toward pension distribution and inequality [@araujo_desafios_2012; @galvez_con_2023].
 The combination of profound inequality, market-driven old-age security and contested legitimacy makes Chile a critical case for analysing how mobility and meritocratic beliefs structure support for pension commodification.
@@
 It adopts a causal framework of mobility effects inspired by @breen_effects_2024.
-The following section outlines the theoretical framework linking preferences for market-based welfare, social mobility and expected heterogeneity by meritocratic beliefs and time, and derives hypotheses.
-A subsequent section presents the causal identification strategy, the data, the variables, and the analytical approach.
-The empirical sections report the findings and interpret their implications.
-The article concludes by discussing what these results reveal about the politics of market-based pension welfare and social mobility.
+The next section outlines the theoretical framework linking preferences for market-based welfare, social mobility and expected heterogeneity by meritocratic beliefs and time, and derives hypotheses.
+A subsequent section presents the causal identification strategy, the data and the analytical approach.
+The empirical sections report the findings, and the article concludes by discussing what these results reveal about the politics of market-based pension welfare and social mobility.

### 2025-11-22 paper/02-antecedents.qmd
Up V2

@@
 
-## Market justice preferences
+## Preferences for market-based welfare
 
@@
 [@castillo_perceptions_2025; @castillo_socialization_2024] find that individuals who endorse meritocratic beliefs—such as the idea that effort and talent are rewarded in Chile—are more likely to support market justice across welfare domains.
+
+\begin{table}[htbp]
+\centering
+\caption{Overview of hypotheses with expected directional patterns}
+\begin{tabular}{p{3.2cm} p{3.2cm} p{4cm} p{5cm}}
+\toprule
+\textbf{Hypothesis} & \textbf{Focus} & \textbf{Expected pattern} & \textbf{Interpretation} \\
+\midrule
+\textbf{H1: Average mobility effect} &
+Effect of intergenerational mobility on preferences for pension commodification &
+Upward mobility $\uparrow$ higher support; downward mobility $\downarrow$ lower support relative to immobility &
+Mobility shapes preferences through material stakes and normative frames \\
+\midrule
+\textbf{H2a: Upward $\times$ Meritocracy} &
+Conditional effect of upward mobility by meritocracy &
+Under high meritocracy: amplified pro-market effect $\uparrow\uparrow$; under low meritocracy: modest increase $\uparrow$ &
+Meritocratic lenses strengthen pro-market reactions to upward mobility \\
+\midrule
+\textbf{H2b: Downward $\times$ Meritocracy} &
+Conditional effect of downward mobility by meritocracy &
+Under low meritocracy: moderate decrease $\downarrow$; under high meritocracy: negative effect attenuated $\searrow$ toward neutral or slightly positive attitudes &
+Meritocratic lenses weaken de-commodifying reactions to downward mobility \\
+\bottomrule
+\end{tabular}
+\end{table}

### 2025-11-22 paper/03-methods.qmd
up antes de borrar tiempo
Up V2

@@
 $$ 
-
 ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
-
-$$ {#eq-att}
+$${#eq-att}
 
@@
 
-To test heterogeneity in mobility effects, I examine two moderators: meritocratic perceptions and time.
-Meritocracy is captured with two items [@young_rise_1958], one referring to effort (“In Chile, people are rewarded for their efforts”) and one to talent (“In Chile, people are rewarded for their intelligence and skills”), each answered on a five-point Likert scale (1 = “strongly disagree” to 5 = “strongly agree”).
-I average the two items into a single index and dichotomize it into low meritocracy (≤ 3) and high meritocracy (≥ 4).
-Time is measured by survey wave (2016, 2018, 2023) and entered as a categorical (dummy) variable.
-Both variables are used as moderators by interacting them with the mobility treatment to assess conditional average treatment effects—i.e., whether the effect of mobility varies across levels of meritocratic perceptions or across survey waves.
+To explore whether mobility effects vary across key contextual dimensions, I examine meritocratic perceptions as a moderator.
+Meritocracy is captured with two items [@young_rise_1958], one referring to effort (“In Chile, people are rewarded for their efforts”) and one to talent (“In Chile, people are rewarded for their intelligence and skills”), combined into a single index and dichotomized into low (≤ 3) and high (≥ 4) meritocracy.
+Because meritocracy is measured after mobility, it is used not as a causal effect modifier but as a post-treatment moderator to probe potential mechanisms.
+I estimate interaction models between mobility and meritocracy and report conditional treatment effects (simple slopes) at low and high meritocratic levels.
 
@@
 These results are virtually identical to baseline models without covariates (see [Supplementary Material](#anexo) for complete models), with no substantive changes in the mobility coefficients.
-Heterogeneous effects by meritocratic perceptions and survey wave are examined through interaction models described in the previous section.
+Heterogeneous effects by meritocratic perceptions are examined through interaction models described in the previous section.
 

### 2025-11-22 paper/04-analysis.qmd
Up V2

@@
 
-## Effect heterogeneity
+## The role of meritocracy
 
-To probe potential mechanisms underlying the mobility effects, I estimate origin-specific ATT models that interact the mobility treatment with high meritocratic perceptions (ref.
-= low), using stabilized IPW weights, covariate adjustment, and wave fixed effects.
-I also estimate an analogous specification interacting mobility with survey waves (ref.
-= 2016) to assess temporal heterogeneity in the treatment effect (see [Supplementary Material](#anexo) for complete models).
+To probe whether meritocratic beliefs channel the impact of mobility on preferences for pension commodification, I estimate interaction models between mobility and meritocratic beliefs and examine the conditional treatment effects of mobility at low and high levels of meritocracy (see [Supplementary Material](#anexo) for complete models).
 
@@
 @fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic beliefs.
-For low-merit respondents, mobility effects are uniformly small and imprecise across trajectories: Low→Middle ($\beta$ = −0.02, 95% CI [−0.09, 0.04]), Low→High ($\beta$ = −0.03, [−0.10, 0.05]), Middle→Low ($\beta$ = −0.00, [−0.07, 0.06]), and a modest, non-significant positive estimate for Middle→High ($\beta$ = 0.07, [−0.01, 0.14]), while downward moves from high origin show borderline negative effects (High→Middle: $\beta$ = −0.07, [−0.15, 0.00]; High→Low: $\beta$ = −0.08, [−0.15, 0.00]).
-Among high-merit respondents, the pattern is directionally consistent with meritocratic moderation but statistically robust only for one trajectory: the ATT for Middle→High at high merit is positive and significant ($\beta$ = 0.22, [0.05, 0.38]), whereas Low→Middle ($\beta$ = 0.02, [−0.13, 0.16]), Low→High ($\beta$ = 0.02, [−0.15, 0.20]), High→Middle ($\beta$ = −0.10, [−0.28, 0.09]), and High→Low ($\beta$ = −0.22, [−0.44, 0.00]) remain statistically indistinguishable from zero.
-Overall, these results provide only limited and trajectory-specific evidence that meritocratic beliefs moderate the causal effect of intergenerational occupational mobility on preferences for pension commodification, with a clearer pro-commodification sign only for upward mobility from middle to high status among high-merit respondents.
-
-```{r}
-#| label: fig-interact-wave
-#| fig-cap: "Effects of intergenerational occupational mobility on preferences for pension commodification, by survey wave.
-Coefficient plot of origin-specific ATT estimates comparing mobility versus immobility, estimated from interaction models (T × Wave).
-Points show the marginal effects (simple slopes) of the mobility treatment in 2016, 2018, and 2023; bars are 95% confidence intervals.
-Estimates from pooled WLS with doubly robust adjustment, wave fixed effects, and standard errors clustered by individual."
-#| out-width: '90%'
-#| fig-asp: 0.8
-#| fig-cap-location: bottom
-#| echo: false
-#| results: asis
-
-g5+
-  theme(legend.text = element_text(size = 10))
-```
-
-I find no consistent evidence of heterogeneous causal effects across survey waves.
-@fig-interact-wave displays the marginal effects of mobility trajectories on market-justice preferences for pensions in 2016, 2018, and 2023.
-Across trajectories, the estimated ATTs remain small and their confidence intervals largely overlap over time.
-Upward mobility from middle origin shows a positive and statistically significant effect in 2016 (Middle→High, $\beta$ = 0.13, 95% CI [0.05, 0.22]) and downward mobility from high origin displays negative effects at baseline (High→Middle and High→Low, both $\beta$ ≈ −0.19, CIs excluding zero), but these patterns attenuate and become statistically indistinguishable from zero by 2023.
-Taken together, the results suggest that any initial pro-commodification response to upward mobility and de-commodifying response to downward mobility are not stable over time, and the causal effect of mobility on preferences for pension commodification is effectively time-invariant over the 2016–2023 period.
+Across trajectories, mobility effects for low-merit respondents are uniformly small and imprecise, hovering around zero for both upward and downward moves.
+Among high-merit respondents, the pattern aligns only partially with the proposed mechanism: upward mobility from middle to high status shows a clear and statistically significant pro-commodification effect ($\beta$ = 0.22, 95% CI [0.05, 0.38]), whereas all other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero or display negative estimates that do not systematically weaken under high merit.
+Formal interaction tests also fail to detect strong heterogeneity by meritocratic beliefs (see [Supplementary Material](#anexo) for interaction models).
+Taken together, the results provide only limited support for the mechanism hypothesis: meritocracy appears to amplify the pro-market consequences of upward mobility in one specific trajectory, but there is no general or consistent evidence that meritocratic beliefs constitute a stable channel through which intergenerational mobility shapes preferences for pension commodification.
 

### 2025-11-22 paper/paper.qmd
Up V2

@@
   html: 
+    latex-environment: true
     code-fold: true

### 2025-11-22 paper/paper.tex
Up V2

@@
 \setstretch{1.5}
-This document was last modified at 2025-11-21 16:32:59 and it was last
-rendered at 2025-11-21 16:32:59.
+This document was last modified at 2025-11-21 21:08:43 and it was last
+rendered at 2025-11-21 21:08:43.
 
@@
 has reconfigured the normative order: through policy feedback, welfare
-institutions increasingly embed market-based criteria of ``fair''
-allocation and shape how citizens understand deservingness and the
-balance between state and market (\citeproc{ref-koos_moral_2019}{Koos \&
-Sachweh, 2019};
+institutions embed market-based criteria of ``fair'' allocation and
+shape how citizens understand deservingness and the balance between
+state and market (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
 \citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
@@
 commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
-2000}), making pension systems a strategic domain for analysing how
-market principles become normatively embedded
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As many countries
-expanded mandatory or voluntary individual-capitalisation schemes---most
-radically in Chile---pensions became organised around contribution-based
-entitlements and market performance
-(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
+2000}). As many countries expanded mandatory or voluntary
+individual-capitalisation schemes---most radically in Chile---pensions
+became organised around contribution-based entitlements and market
+performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
 \citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
-reshapes not only how pension institutions allocate old-age risk,
-heightening exposure to poverty in retirement and tying benefit levels
-to lifetime labour-market trajectories
+reshapes how pension institutions allocate old-age risk, tying benefit
+levels to lifetime labour-market trajectories and increasing exposure to
+poverty in retirement
 (\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
-\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also reshape the
-justice principles citizens associate with old-age security: people have
-to judge whether higher benefits should follow market-derived criteria
-or solidaristic principles
+\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). It also transforms the
+justice principles citizens associate with old-age security: they must
+judge whether higher benefits should follow market-derived criteria or
+solidaristic principles
 (\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
@@
 (\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
-\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). As Mau
-(\citeproc{ref-mau_inequality_2015}{2015}) argues, social mobility can
-foster market-justice orientations, helping explain how inequalities in
-access to welfare services become normatively sustained.
-
-The literature identifies several mechanisms through which mobility may
-shape distributive-justice preferences---material self-interest
-(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}), socialisation
-and acculturation
-(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}), learning through long-term exposure
-(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}), and
-cognitive attributions such as self-serving bias or privilege awareness
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-molina_its_2019}{Molina et al., 2019}). While distinct,
-these accounts converge in highlighting two broad channels relevant for
-market-based welfare: shifts in material stakes and shifts in normative
-frames (\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In this paper, I
-do not treat these mechanisms as mediators but use them to guide
-hypotheses about heterogeneity: if mobility alters preferences through
-material interests and normative updating, its effects should be most
-potent among those who interpret social outcomes through merit-based
-evaluations of effort
+\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). Social
+mobility can thus foster market-justice orientations, helping explain
+how inequalities in access to welfare services become normatively
+sustained (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
+
+Existing accounts identify several mechanisms through which
+intergenerational mobility may shape distributive-justice preferences
+but converge on two channels particularly relevant for market-based
+welfare: changes in material stakes and shifts in normative frames
+(\citeproc{ref-ares_changing_2020}{Ares, 2020};
+\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
+\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
+\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}). Recent research shows that meritocratic
+beliefs---seeing that effort and ability are rewarded---are strongly
+associated with support for market allocation
 (\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).
-
-Against this background, this article pursues two aims: first, to assess
-whether intergenerational mobility causally affects support for the
+2025}). Perceived upward mobility reinforces meritocratic
+interpretations of success, whereas downward mobility is more closely
+linked to structural explanations of inequality without necessarily
+eroding meritocratic convictions (\citeproc{ref-mijs_belief_2022}{Mijs
+et al., 2022}). Thus, meritocracy and mobility operate on partly
+separate tracks yet intersect in systematic ways. I treat meritocratic
+beliefs as a post-treatment moderator---a diagnostic lens on potential
+mechanisms. If mobility partly shapes preferences by strengthening
+merit-based understandings of fairness and altering material stakes, its
+effects should be most visible---and most pro-market---among individuals
+who already interpret social outcomes through meritocratic lenses.
+
+This article pursues two aims: first, to assess whether
+intergenerational occupational mobility causally affects support for the
 commodification of pensions; and second, to examine whether this effect
-varies by meritocratic beliefs and over time. I argue that mobility
-works through two main channels. Upward mobility increases earnings,
-expected returns in a funded system and perceived economic security,
-making contribution-based access more appealing. At the same time,
-integration into higher-class networks exposes people to more
-individualistic norms and effort-based understandings of fairness.
-Downward mobility, in turn, heightens economic insecurity, dependence on
-collective protection and attachment to solidaristic principles,
-lowering support for market allocation. Meritocratic beliefs are
-conceptualised not as mediators but as moderators. Among those who see
-mobility as a fair reward for effort, both material and normative
-effects should be amplified, producing higher support for pension
-commodification. The paper contributes to research on market justice by
-focusing on a specific welfare domain and by providing causal evidence
-on how mobility shapes distributive preferences.
-
-This study focuses on Chile, a particularly instructive case for
-examining preferences for pension commodification. Despite sustained
-economic growth, Chile combines high and persistent inequality with
-short-range upward mobility and substantial barriers to higher-class
-positions (\citeproc{ref-flores_top_2020}{Flores et al., 2020};
+varies by meritocratic beliefs as a diagnostic of possible transmission
+channels. I argue that social mobility operates through two main
+pathways. Upward mobility increases earnings, expected returns in a
+funded system and perceived economic security, making contribution-based
+access more appealing. At the same time, exposure to higher-class
+networks reinforces individualistic and effort-based understandings of
+fairness. Downward mobility tends to heighten economic insecurity,
+dependence on public protection and attachment to external attributions
+of poverty, reducing support for market allocation. Meritocratic beliefs
+are conceptualised not as mediators but as moderators. Among individuals
+with stronger meritocratic beliefs, upward mobility should heighten
+support for market-based access. In contrast, downward mobility should
+weaken---or even reverse---its usual de-commodifying pattern, as
+meritocratic interpretations of effort lead individuals to view market
+outcomes as fair regardless of direction. The paper contributes to
+research on market justice by identifying the causal effects of mobility
+and by using meritocracy to probe whether these effects align with a
+meritocratic channel.
+
+The study focuses on Chile, an instructive case for examining
+preferences for pension commodification. Despite sustained economic
+growth, Chile combines high and persistent inequality with short-range
+upward mobility and substantial barriers to higher-class positions
+(\citeproc{ref-flores_top_2020}{Flores et al., 2020};
 \citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
@@
 first country worldwide to fully replace a public pay-as-you-go system
-with a mandatory, privately administered (through pension fund
-administrators, AFPs) defined-contribution scheme, while the state has
-progressively expanded a segmented solidarity pillar to compensate for
-insufficient benefits (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+with a mandatory, privately administered defined-contribution scheme,
+while the state has progressively expanded a segmented solidarity pillar
+to compensate for insufficient benefits
+(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
 \citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
 \citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
-despite waves of protest from 2006 to 2019 targeting the AFP system
-(\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
+despite waves of protest from 2006 to 2019 against the private pension
+system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
 subjectivities have been increasingly shaped by neoliberal discourses
@@
 inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
-The following section outlines the theoretical framework linking
-preferences for market-based welfare, social mobility and expected
-heterogeneity by meritocratic beliefs and time, and derives hypotheses.
-A subsequent section presents the causal identification strategy, the
-data, the variables, and the analytical approach. The empirical sections
-report the findings and interpret their implications. The article
-concludes by discussing what these results reveal about the politics of
-market-based pension welfare and social mobility.
+The next section outlines the theoretical framework linking preferences
+for market-based welfare, social mobility and expected heterogeneity by
+meritocratic beliefs and time, and derives hypotheses. A subsequent
+section presents the causal identification strategy, the data and the
+analytical approach. The empirical sections report the findings, and the
+article concludes by discussing what these results reveal about the
+politics of market-based pension welfare and social mobility.
 
@@
 
-\subsection{Market justice
-preferences}\label{market-justice-preferences}
+\subsection{Preferences for market-based
+welfare}\label{preferences-for-market-based-welfare}
 
@@
 
+\begin{table}[htbp]
+\centering
+\caption{Overview of hypotheses with expected directional patterns}
+\begin{tabular}{p{3.2cm} p{3.2cm} p{4cm} p{5cm}}
+\toprule
+\textbf{Hypothesis} & \textbf{Focus} & \textbf{Expected pattern} & \textbf{Interpretation} \\
+\midrule
+\textbf{H1: Average mobility effect} &
+Effect of intergenerational mobility on preferences for pension commodification &
+Upward mobility $\uparrow$ higher support; downward mobility $\downarrow$ lower support relative to immobility &
+Mobility shapes preferences through material stakes and normative frames \\
+\midrule
+\textbf{H2a: Upward $\times$ Meritocracy} &
+Conditional effect of upward mobility by meritocracy &
+Under high meritocracy: amplified pro-market effect $\uparrow\uparrow$; under low meritocracy: modest increase $\uparrow$ &
+Meritocratic lenses strengthen pro-market reactions to upward mobility \\
+\midrule
+\textbf{H2b: Downward $\times$ Meritocracy} &
+Conditional effect of downward mobility by meritocracy &
+Under low meritocracy: moderate decrease $\downarrow$; under high meritocracy: negative effect attenuated $\searrow$ toward neutral or slightly positive attitudes &
+Meritocratic lenses weaken de-commodifying reactions to downward mobility \\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \section{Method}\label{method}
@@
 
-To test heterogeneity in mobility effects, I examine two moderators:
-meritocratic perceptions and time. Meritocracy is captured with two
-items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
-effort (``In Chile, people are rewarded for their efforts'') and one to
-talent (``In Chile, people are rewarded for their intelligence and
-skills''), each answered on a five-point Likert scale (1 = ``strongly
-disagree'' to 5 = ``strongly agree''). I average the two items into a
-single index and dichotomize it into low meritocracy (≤ 3) and high
-meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
-and entered as a categorical (dummy) variable. Both variables are used
-as moderators by interacting them with the mobility treatment to assess
-conditional average treatment effects---i.e., whether the effect of
-mobility varies across levels of meritocratic perceptions or across
-survey waves.
+To explore whether mobility effects vary across key contextual
+dimensions, I examine meritocratic perceptions as a moderator.
+Meritocracy is captured with two items
+(\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to effort
+(``In Chile, people are rewarded for their efforts'') and one to talent
+(``In Chile, people are rewarded for their intelligence and skills''),
+combined into a single index and dichotomized into low (≤ 3) and high (≥
+4) meritocracy. Because meritocracy is measured after mobility, it is
+used not as a causal effect modifier but as a post-treatment moderator
+to probe potential mechanisms. I estimate interaction models between
+mobility and meritocracy and report conditional treatment effects
+(simple slopes) at low and high meritocratic levels.
 
@@
 no substantive changes in the mobility coefficients. Heterogeneous
-effects by meritocratic perceptions and survey wave are examined through
-interaction models described in the previous section.
+effects by meritocratic perceptions are examined through interaction
+models described in the previous section.
 
@@
 
-\subsection{Effect heterogeneity}\label{effect-heterogeneity-1}
-
-To probe potential mechanisms underlying the mobility effects, I
-estimate origin-specific ATT models that interact the mobility treatment
-with high meritocratic perceptions (ref. = low), using stabilized IPW
-weights, covariate adjustment, and wave fixed effects. I also estimate
-an analogous specification interacting mobility with survey waves (ref.
-= 2016) to assess temporal heterogeneity in the treatment effect (see
+\subsection{The role of meritocracy}\label{the-role-of-meritocracy}
+
+To probe whether meritocratic beliefs channel the impact of mobility on
+preferences for pension commodification, I estimate interaction models
+between mobility and meritocratic beliefs and examine the conditional
+treatment effects of mobility at low and high levels of meritocracy (see
 \hyperref[anexo]{Supplementary Material} for complete models).
@@
 from the meritocracy interaction models as marginal effects of the
-mobility treatment at low and high meritocratic beliefs. For low-merit
-respondents, mobility effects are uniformly small and imprecise across
-trajectories: Low→Middle (\(\beta\) = −0.02, 95\% CI {[}−0.09, 0.04{]}),
-Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}), Middle→Low (\(\beta\) =
-−0.00, {[}−0.07, 0.06{]}), and a modest, non-significant positive
-estimate for Middle→High (\(\beta\) = 0.07, {[}−0.01, 0.14{]}), while
-downward moves from high origin show borderline negative effects
-(High→Middle: \(\beta\) = −0.07, {[}−0.15, 0.00{]}; High→Low: \(\beta\)
-= −0.08, {[}−0.15, 0.00{]}). Among high-merit respondents, the pattern
-is directionally consistent with meritocratic moderation but
-statistically robust only for one trajectory: the ATT for Middle→High at
-high merit is positive and significant (\(\beta\) = 0.22, {[}0.05,
-0.38{]}), whereas Low→Middle (\(\beta\) = 0.02, {[}−0.13, 0.16{]}),
-Low→High (\(\beta\) = 0.02, {[}−0.15, 0.20{]}), High→Middle (\(\beta\) =
-−0.10, {[}−0.28, 0.09{]}), and High→Low (\(\beta\) = −0.22, {[}−0.44,
-0.00{]}) remain statistically indistinguishable from zero. Overall,
-these results provide only limited and trajectory-specific evidence that
-meritocratic beliefs moderate the causal effect of intergenerational
-occupational mobility on preferences for pension commodification, with a
-clearer pro-commodification sign only for upward mobility from middle to
-high status among high-merit respondents.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-wave-1.pdf}
-
-}
-
-\caption{\label{fig-interact-wave}Effects of intergenerational
-occupational mobility on preferences for pension commodification, by
-survey wave. Coefficient plot of origin-specific ATT estimates comparing
-mobility versus immobility, estimated from interaction models (T ×
-Wave). Points show the marginal effects (simple slopes) of the mobility
-treatment in 2016, 2018, and 2023; bars are 95\% confidence intervals.
-Estimates from pooled WLS with doubly robust adjustment, wave fixed
-effects, and standard errors clustered by individual.}
-
-\end{figure}%
-
-I find no consistent evidence of heterogeneous causal effects across
-survey waves. Figure~\ref{fig-interact-wave} displays the marginal
-effects of mobility trajectories on market-justice preferences for
-pensions in 2016, 2018, and 2023. Across trajectories, the estimated
-ATTs remain small and their confidence intervals largely overlap over
-time. Upward mobility from middle origin shows a positive and
-statistically significant effect in 2016 (Middle→High, \(\beta\) = 0.13,
-95\% CI {[}0.05, 0.22{]}) and downward mobility from high origin
-displays negative effects at baseline (High→Middle and High→Low, both
-\(\beta\) ≈ −0.19, CIs excluding zero), but these patterns attenuate and
-become statistically indistinguishable from zero by 2023. Taken
-together, the results suggest that any initial pro-commodification
-response to upward mobility and de-commodifying response to downward
-mobility are not stable over time, and the causal effect of mobility on
-preferences for pension commodification is effectively time-invariant
-over the 2016--2023 period.
+mobility treatment at low and high meritocratic beliefs. Across
+trajectories, mobility effects for low-merit respondents are uniformly
+small and imprecise, hovering around zero for both upward and downward
+moves. Among high-merit respondents, the pattern aligns only partially
+with the proposed mechanism: upward mobility from middle to high status
+shows a clear and statistically significant pro-commodification effect
+(\(\beta\) = 0.22, 95\% CI {[}0.05, 0.38{]}), whereas all other
+trajectories---Low→Middle, Low→High, Middle→Low, High→Middle, and
+High→Low---remain statistically indistinguishable from zero or display
+negative estimates that do not systematically weaken under high merit.
+Formal interaction tests also fail to detect strong heterogeneity by
+meritocratic beliefs (see \hyperref[anexo]{Supplementary Material} for
+interaction models). Taken together, the results provide only limited
+support for the mechanism hypothesis: meritocracy appears to amplify the
+pro-market consequences of upward mobility in one specific trajectory,
+but there is no general or consistent evidence that meritocratic beliefs
+constitute a stable channel through which intergenerational mobility
+shapes preferences for pension commodification.
 
@@
 
-\caption{\label{tbl-complete3}Marginal effects of mobility by
+\caption{\label{tbl-complete3}Interactions effects between
+intergenerational occupational mobility and perceived meritocracy on
+preferences for pension commodification, with covariates and wave fixed
+effects.}
+
+\centering{
+
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
+\hline
+Intercept                                                    & $0.15$           & $-0.04$          & $0.30^{*}$       & $0.21$           & $0.08$           & $0.05$           \\
+                                                             & $ [-0.36; 0.67]$ & $ [-0.40; 0.31]$ & $ [ 0.04; 0.56]$ & $ [-0.09; 0.51]$ & $ [-0.25; 0.40]$ & $ [-0.30; 0.41]$ \\
+Mobility treatment                                           & $-0.02$          & $-0.03$          & $0.07$           & $-0.00$          & $-0.07$          & $-0.08$          \\
+                                                             & $ [-0.09; 0.04]$ & $ [-0.10; 0.05]$ & $ [-0.01; 0.14]$ & $ [-0.07; 0.06]$ & $ [-0.15; 0.00]$ & $ [-0.15; 0.00]$ \\
+High meritocracy perception (Ref.= Low)                      & $0.05$           & $0.06$           & $-0.01$          & $0.03$           & $0.17^{*}$       & $0.26^{*}$       \\
+                                                             & $ [-0.06; 0.16]$ & $ [-0.07; 0.19]$ & $ [-0.09; 0.08]$ & $ [-0.07; 0.13]$ & $ [ 0.00; 0.33]$ & $ [ 0.04; 0.48]$ \\
+Mobility treatment x High meritocracy perception (Ref.= Low) & $0.04$           & $0.05$           & $0.15$           & $0.04$           & $-0.03$          & $-0.14$          \\
+                                                             & $ [-0.12; 0.20]$ & $ [-0.14; 0.24]$ & $ [-0.03; 0.33]$ & $ [-0.11; 0.19]$ & $ [-0.23; 0.17]$ & $ [-0.38; 0.10]$ \\
+\hline
+Controls                                                     & Yes              & Yes              & Yes              & Yes              & Yes              & Yes              \\
+R$^2$                                                        & $0.04$           & $0.05$           & $0.08$           & $0.04$           & $0.07$           & $0.10$           \\
+Adj. R$^2$                                                   & $0.03$           & $0.04$           & $0.07$           & $0.02$           & $0.06$           & $0.08$           \\
+Num. obs.                                                    & $878$            & $745$            & $716$            & $773$            & $912$            & $861$            \\
+RMSE                                                         & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$           & $0.41$           \\
+N Clusters                                                   & $497$            & $449$            & $443$            & $470$            & $536$            & $525$            \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-complete4}Marginal effects of mobility by
 meritocratic perception (Low vs High Meritocracy).}
@@
 High-Low & High Meritocracy & -0.216 & 0.112 & -0.436 & 0.005\\
-\hline
-\end{tabular}
-
-}
-
-\end{table}%
-
-\begin{table}
-
-\caption{\label{tbl-complete4}Marginal effects of mobility by survey
-wave (2016, 2018, and 2023).}
-
-\centering{
-
-\centering
-\begin{tabular}{l|c|c|c|c|c}
-\hline
-Trajectory & Wave & Estimate & Std. Error & CI Low & CI High\\
-\hline
-Low-Middle & 2016 & -0.048 & 0.055 & -0.156 & 0.060\\
-\hline
-Low-Middle & 2018 & -0.022 & 0.048 & -0.116 & 0.072\\
-\hline
-Low-Middle & 2023 & 0.022 & 0.058 & -0.092 & 0.135\\
-\hline
-Low-High & 2016 & -0.025 & 0.062 & -0.146 & 0.096\\
-\hline
-Low-High & 2018 & -0.088 & 0.050 & -0.186 & 0.011\\
-\hline
-Low-High & 2023 & 0.072 & 0.066 & -0.058 & 0.202\\
-\hline
-Middle-High & 2016 & 0.132 & 0.044 & 0.047 & 0.218\\
-\hline
-Middle-High & 2018 & 0.091 & 0.051 & -0.008 & 0.190\\
-\hline
-Middle-High & 2023 & 0.045 & 0.061 & -0.074 & 0.164\\
-\hline
-Middle-Low & 2016 & 0.061 & 0.044 & -0.025 & 0.147\\
-\hline
-Middle-Low & 2018 & -0.027 & 0.046 & -0.118 & 0.064\\
-\hline
-Middle-Low & 2023 & -0.010 & 0.057 & -0.122 & 0.103\\
-\hline
-High-Middle & 2016 & -0.187 & 0.070 & -0.324 & -0.050\\
-\hline
-High-Middle & 2018 & -0.078 & 0.050 & -0.175 & 0.019\\
-\hline
-High-Middle & 2023 & 0.006 & 0.063 & -0.118 & 0.130\\
-\hline
-High-Low & 2016 & -0.186 & 0.068 & -0.319 & -0.053\\
-\hline
-High-Low & 2018 & -0.125 & 0.053 & -0.228 & -0.022\\
-\hline
-High-Low & 2023 & -0.009 & 0.068 & -0.141 & 0.124\\
 \hline

### 2025-11-30 paper/01-introduction.qmd
Up antecedentes

@@
 From a moral-economy perspective, this diffusion of market rules has reconfigured the normative order: through policy feedback, welfare institutions embed market-based criteria of “fair” allocation and shape how citizens understand deservingness and the balance between state and market [@svallforsMoralEconomyClass2006a; @koos_moral_2019].
-Against this backdrop, growing scholarship examines *market justice preferences*—the extent to which citizens regard market-based criteria as a fair basis for access to essential services such as healthcare, education and pensions [@castillo_perceptions_2025; @svallfors_political_2007; @lindh_public_2015].
-These orientations matter because they help legitimise economic inequality by portraying stratified outcomes as a fair result of individual responsibility and limited state intervention.
+Against this backdrop, growing scholarship examines *market justice preferences*—the extent to which citizens regard market-based criteria as a fair basis for the allocation of essential services such as healthcare, education and pensions [@castillo_perceptions_2025; @svallfors_political_2007; @lindh_public_2015].
+These orientations matter because they legitimize unequal outcomes as the product of individual responsibility and promote the conception of basic social services as commodities.
 
@@
 As many countries expanded mandatory or voluntary individual-capitalisation schemes—most radically in Chile—pensions became organised around contribution-based entitlements and market performance [@arenas_sistemas_2019; @verbic_political_2019].
-This shift reshapes how pension institutions allocate old-age risk, tying benefit levels to lifetime labour-market trajectories and increasing exposure to poverty in retirement [@oecd_pensions_2023; @madero-cabib_private_2019].
-It also transforms the justice principles citizens associate with old-age security: they must judge whether higher benefits should follow market-derived criteria or solidaristic principles [@arrizabalomontoro_debate_2019; @borzutzky_pension_2012].
+This shift reshapes not only how pension institutions allocate old-age risk and tie benefit levels to lifetime labour-market trajectories [@oecd_pensions_2023; @madero-cabib_private_2019], but also transforms the justice principles citizens associate with old-age security: they must judge whether higher benefits should follow market-derived criteria or solidaristic principles [@mau_inequality_2015; @borzutzky_pension_2012].
 Although we know something about what people consider a “just” monetary pension [@castillo_deserving_2019] and about support for public versus private provision [@busemeyer_welfare_2020; @jaime-castillo_public_2013], we still lack systematic evidence on how citizens assess the fairness of market-dependent access to pensions.
 
-Research shows that support for market-based access to welfare services is strongly stratified: individuals with higher income, education or class position are more likely to endorse market-based criteria than disadvantaged groups [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015; @svallfors_political_2007].
-Evidence for pensions points in the same direction: higher income and greater financial gains in funded schemes correlate with stronger support for market-oriented pension policies [@kerner_pension_2020].
+Research shows that support for market-based access to welfare services is strongly stratified: people with higher socioeconomic status are more likely to support market-based criteria than disadvantaged groups [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015; @svallfors_political_2007].
+Evidence for pensions points in the same direction: higher income and greater financial gains in funded schemes correlate with stronger support for market-oriented pension policies [@busemeyer_welfare_2020; @kerner_pension_2020].
 Yet we know very little about how movement within the class structure shapes preferences for pension commodification.
@@
 
-Existing accounts identify several mechanisms through which intergenerational mobility may shape distributive-justice preferences but converge on two channels particularly relevant for market-based welfare: changes in material stakes and shifts in normative frames [@ares_changing_2020; @jaime-castillo_social_2019; @helgason_longterm_2023; @gugushvili_intergenerational_2016c].
-Recent research shows that meritocratic beliefs—seeing that effort and ability are rewarded—are strongly associated with support for market allocation [@castillo_perceptions_2025].
-Perceived upward mobility reinforces meritocratic interpretations of success, whereas downward mobility is more closely linked to structural explanations of inequality without necessarily eroding meritocratic convictions [@mijs_belief_2022].
-Thus, meritocracy and mobility operate on partly separate tracks yet intersect in systematic ways.
-I treat meritocratic beliefs as a post-treatment moderator—a diagnostic lens on potential mechanisms.
-If mobility partly shapes preferences by strengthening merit-based understandings of fairness and altering material stakes, its effects should be most visible—and most pro-market—among individuals who already interpret social outcomes through meritocratic lenses.
+Existing accounts identify several mechanisms through which intergenerational mobility may shape distributive justice preferences but converge on two channels especially relevant for market-based welfare: changes in material interests and shifts in normative frames [@ares_changing_2020; @jaime-castillo_social_2019; @helgason_longterm_2023; @gugushvili_intergenerational_2016c].
+Mobility alters individuals’ material stakes by changing their occupational status, income, and risk exposure.
+At the same time, it reshapes their fairness judgments.
+Empirically, upward mobility is often associated with stronger meritocratic interpretations of success and fairness, whereas downward mobility is more closely linked to structural explanations of inequality without necessarily eroding meritocratic convictions [@mijs_belief_2022; @bucca_merit_2016; @deng_its_2025].
+Regarding market-based preferences, recent research shows that meritocratic beliefs – the conviction that effort and ability are rewarded – are strongly associated with support for market allocation [@castillo_perceptions_2025].
+In such contexts, the upwardly mobile are especially likely to view markets as just distributors, since their own trajectories appear to confirm that markets appropriately reward effort; those who experience downward mobility may still interpret outcomes through a meritocratic lens, often reading losses as personal failure rather than as evidence against market fairness.
+Thus, mobility and meritocracy evolve on partly separate tracks yet intersect systematically [@mau_inequality_2015].
 
-This article pursues two aims: first, to assess whether intergenerational occupational mobility causally affects support for the commodification of pensions; and second, to examine whether this effect varies by meritocratic beliefs as a diagnostic of possible transmission channels.
-I argue that social mobility operates through two main pathways.
-Upward mobility increases earnings, expected returns in a funded system and perceived economic security, making contribution-based access more appealing.
-At the same time, exposure to higher-class networks reinforces individualistic and effort-based understandings of fairness.
-Downward mobility tends to heighten economic insecurity, dependence on public protection and attachment to external attributions of poverty, reducing support for market allocation.
-Meritocratic beliefs are conceptualised not as mediators but as moderators.
-Among individuals with stronger meritocratic beliefs, upward mobility should heighten support for market-based access.
-In contrast, downward mobility should weaken—or even reverse—its usual de-commodifying pattern, as meritocratic interpretations of effort lead individuals to view market outcomes as fair regardless of direction.
-The paper contributes to research on market justice by identifying the causal effects of mobility and by using meritocracy to probe whether these effects align with a meritocratic channel.
+Against this backdrop, the central aim of this article is to assess whether intergenerational occupational mobility causally affects support for the commodification of pensions.
+A second, more exploratory aim is to examine whether this effect varies by meritocratic beliefs, using this heterogeneity as an indication of possible underlying mechanisms.
+I argue that social mobility operates through two main pathways [@mau_inequality_2015].
+Upward mobility increases earnings, expected returns in a funded system, and perceived economic security, making contribution-based access more appealing.
+Simultaneously, upward moves reinforce individualistic, effort-based understandings of success and, thus, the view that markets are just assignors.
+Downward mobility tends to heighten economic insecurity, dependence on public protection, and attachment to external attributions of poverty, reducing support for market allocation.
+Meritocratic beliefs are treated not as mediators but as moderators of these effects: among strong meritocrats, upward mobility should amplify pro-commodification attitudes, whereas downward mobility should attenuate – or even reverse – its expected de-commodifying impact, as meritocratic interpretations of effort lead individuals to view market outcomes as fair regardless of direction.
+This inquiry contributes to research on market justice by identifying the causal effects of mobility and using variation in meritocratic beliefs to assess whether these effects follow a meritocratic pathway.
 
-The study focuses on Chile, an instructive case for examining preferences for pension commodification.
+This study focuses on Chile, an instructive case for examining preferences for pension commodification.
 Despite sustained economic growth, Chile combines high and persistent inequality with short-range upward mobility and substantial barriers to higher-class positions [@flores_top_2020; @salgado_uplifting_2025; @torche_intergenerational_2014].
-It was the first country worldwide to fully replace a public pay-as-you-go system with a mandatory, privately administered defined-contribution scheme, while the state has progressively expanded a segmented solidarity pillar to compensate for insufficient benefits [@madariaga_three_2020; @boccardo_30_2020; @solimano_rise_2021].
+It was the first country worldwide in 1981 to fully replace a public pay-as-you-go system with a mandatory, privately administered defined-contribution scheme, while the state has progressively expanded a segmented solidarity pillar to compensate for insufficient benefits [@madariaga_three_2020; @boccardo_30_2020; @solimano_rise_2021].
 In parallel—and despite waves of protest from 2006 to 2019 against the private pension system [@somma_no_2021]—Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing attitudes toward pension distribution and inequality [@araujo_desafios_2012; @galvez_con_2023].
 The combination of profound inequality, market-driven old-age security and contested legitimacy makes Chile a critical case for analysing how mobility and meritocratic beliefs structure support for pension commodification.
+
+In this context, the questions that guide this research are as follows:
+
+(1) How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile?
+(2) How do meritocratic beliefs condition or moderate this relationship?
 

### 2025-11-30 paper/02-antecedents.qmd
Up antecedentes

@@
+---
+bibliography: "../input/bib/tesis-magister.bib"
+csl: "../input/bib/apa6.csl"
+---
+
 # Theoretical and empirical background
 
-## Preferences for market-based welfare
+## Preferences for pensions commodification
 
-Beyond state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023].
-As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks.
-These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking [@koos_moral_2019; @immergut_theoretical_1998].
-In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007].
-While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@lindh_bringing_2023].
-In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023].
-Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020].
-This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@satz_por_2019; @mau_inequality_2015].
+Beyond the state’s capacity to redistribute resources, market institutions are central arenas for allocating socially valuable goods and risks [@lindh_bringing_2023].
+Markets are not mere aggregates of individual choices but social institutions endowed with rules and normative meanings, built through “moral projects, saturated with normativity” that permeate everyday thinking around fairness, effort and deservingness [@fourcade_moral_2007; @koos_moral_2019].
+In this sense, the economic order is mirrored in a moral economy: shared norms and beliefs about what is considered a just distribution, embedded and reinforced through institutions and policy feedback [@svallforsMoralEconomyClass2006a].
+In an era of privatisation and commodification, market logic has expanded into core areas of social reproduction—childcare, healthcare, education and pensions—stratifying access and quality [@ferre_welfare_2023; @gingrich_making_2011] and generating feedback effects on public beliefs about welfare and market inequality [@lindh_public_2015].
+In this context, support for market-based welfare provision has grown worldwide, particularly among higher-income groups who view private alternatives as more efficient or higher-quality [@busemeyer_welfare_2020; @lindh_public_2015].
+These institutional transformations provide the backdrop for analysing how far citizens accept the commodification of social protection [@satz_por_2019].
 
-The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles.
-Beyond stratification structures and socioeconomic conditions, the study of inequality also focuses on individuals’ beliefs about the origins of inequality, the normative frameworks that sustain those beliefs, the mechanisms that shape them, and their implications for attitudes and behavior [@kluegel_beliefs_1981].
-In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016].
-This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011].
-Most of the research on this field explores the legitimacy of wage inequality, specially salary gaps between jobs at opposite ends of the occupational hierarchy [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978].
-Recently, others examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, pensions or social security.
-Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015].
-In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic [@busemeyer_welfare_2020].
+The legitimacy of market-based welfare is closely tied to beliefs about distributive justice grounded in market principles.
+Beyond stratification structures, citizens hold structured beliefs about how resources ought to be allocated, reflecting causal attributions for inequality, normative principles and expectations about deservingness [@kluegel_beliefs_1981].
+Research on distributive justice examines conceptions of how goods and rewards should be distributed and when inequality is considered just [@castillo_legitimacy_2011; @jasso_gender_1999], while work on redistributive preferences focuses on support for state-led mechanisms that reduce inequality [@cavaille_fair_2025].
+Situated at this intersection, the emerging literature on market justice preferences conceptualises the market itself as a site of redistribution and asks whether individuals regard market criteria as fair bases for allocating wages and access to core welfare services [@lindh_public_2015; @lindh_bringing_2023; @castillo_perceptions_2025].
+In this perspective, the expansion of market logic into welfare domains implies that social services are treated as legitimate commodities that can be priced and stratified [@busemeyer_welfare_2020], and empirical work suggests that such institutional changes feed back into attitudes, reinforcing market-conforming understandings of fairness and responsibility [@zhu_policy_2015].
 
-Market justice preferences refer to normative beliefs that legitimize the idea that access to core social services—such as healthcare, education, or pensions—should be determined by market-based criteria.
-Following Janmaat’s [-@janmaat_subjective_2013, p.
-359] distinction, these preferences fall under the category of “beliefs,” understood as normative ideas about what inequality should look like, as opposed to “perceptions,” which refer to subjective evaluations of existing inequality.
-Market justice preferences reflect the view that access to these services should depend on individuals’ ability to pay, thus justifying inequalities generated by market mechanisms [@kluegel_legitimation_1999;  @lindh_public_2015].
-The concept draws on Lane’s [-@lane_market_1986] classic contrast between market justice and political justice: the former is grounded in the idea of earned deserts—where rewards reflect effort, productivity, and skill—while the latter prioritizes need and equality, typically expressed in welfare state policies.
-Lane argued that markets and states differ in purpose (efficiency vs.
-need), logic (individual vs.
-collective), and fairness criteria (merit vs.
-equality).
-Market justice assumes that markets are neutral, self-regulating systems in which fair procedures yield outcomes proportional to merit.
-Inequality, from this perspective, is not only expected but legitimate—as long as it arises from fair competition.
-In this way, market justice offers a moral lens through which individuals can view the commodification and stratified access to social services as fair and justified [@kluegel_legitimation_1999; @lindh_public_2015].
+Market justice preferences refer to normative beliefs that legitimise allocating goods and services according to market criteria such as income and purchasing power [@kluegel_legitimation_1999; @lindh_public_2015].
+Building on Lane’s [-@lane_market_1986] contrast between market and political justice, market justice is a distributive principle that endorses rewarding effort, productivity and skill rather than need and equality as emphasised in welfare-state policies.
+In this study, I treat market justice less as an abstract doctrine and more concretely as preferences for commodification: support for treating welfare services as commodities, so that better services follow from higher income or ability to pay, and the resulting stratified access is seen as fair [@lindh_public_2015].
+Empirically, such orientations are typically measured with items asking whether it is fair that access to services depends on income, a strategy rooted in research on justifications of capitalist inequality [@kluegel_legitimation_1999] and applied to healthcare and education [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_changes_2025].
+Moving beyond single domains, @lindh_public_2015 develops a comparative index of income-based access to healthcare and education, while @castillo_perceptions_2025 use the same single-item measure but add the pensions domain in Chile.
+These constructs gauge how individuals view market-generated inequalities as legitimate and capture two core dimensions of market distribution: the role of income in determining attainment and the framing of services as commodities that can be bought and sold according to ability to pay [@lindh_public_2015].
 
-The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation.
-A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income.
-This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems.
-Subsequent studies have extended this logic to a wider set of welfare goods.
-For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
-[-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare.
-Similarly, Lee and Stacey [-@lee_fairness_2023] and @castillo_changes_2025 apply this framework to education.
-Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation.
-Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al.
-[-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile.
-These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate.
-Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [@lindh_public_2015].
-
-Comparative empirical research has identified several individual-level factors that influence support for market justice.
-Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles [@koos_moral_2019; @lindh_public_2015; @svallfors_political_2007].
-For example, Lindh [-@lindh_public_2015] finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries.
-In a comparative analysis, Svallfors [-@svallfors_political_2007] observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class.
-Busemeyer [-@busemeyer_skills_2014] similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al.
-[-@vondemknesebeck_are_2016] report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage.
-In Chile, Otero and Mendoza [-@otero_power_2024] show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions.
-Beliefs about inequality and political orientation also matter.
-Castillo et al.
-[-@castillo_socialization_2024] and @castillo_perceptions_2025 show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution in Chile, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism.
-In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.
-
-However, individual characteristics alone do not fully explain variation in support for market justice.
-A growing body of cross-national research highlights the importance of institutional arrangements in shaping these preferences.
-For instance, Immergut and Schneider [-@immergut_it_2020] find that in countries with higher public spending on healthcare, individuals are less likely to view income-based access as fair.
-Similarly, Busemeyer [-@busemeyer_skills_2014] shows that increased public investment in education is associated with lower support for privatized provision.
-Conversely, Lindh [-@lindh_public_2015] finds that in countries with more market-oriented welfare systems, support for market-based distribution tends to be higher, suggesting that individual attitudes often align with institutional outputs.
-These findings are consistent with neo-institutionalist and policy feedback theories [@campbell_institutional_2020], which argue that institutions do more than redistribute resources—they also shape the normative categories through which individuals assess who is deserving of support [@immergut_theoretical_1998].
-In this view, dominant values and preferences are both shaped by and embedded in institutional configurations [@busemeyer_skills_2014], reinforcing the notion that institutions are not neutral structures, but active producers of the moral frameworks that legitimize or challenge inequality.
+This study examines support for the commodification of pensions, defined as normative judgements on whether better pensions should depend on income or individual contributions.
+I treat this as a domain-specific form of market justice: pensions are conceived as commodities whose allocation should follow contribution- and return-based criteria rather than principles of need [@lane_market_1986; @mau_inequality_2015].
+The spread of funded, privately managed schemes built on individual accounts has transformed both the distributive and normative structure of old-age security [@huber_political_2000; @arenas_sistemas_2019].
+Distributively, these schemes reassign risk from the collective to the individual by tying benefits to formal employment, contribution density, earnings and financial returns, making retirement income heavily dependent on labour-market trajectories and capital-market success [@madero-cabib_private_2019; @oecd_pensions_2023].
+Normatively, they recast pensions as the outcome of personal investment decisions rather than collective obligations, promoting ideals of individual responsibility, self-economisation and contribution-based desert [@arza_pension_2008; @borzutzky_pension_2012].
+Contributions are increasingly framed as private assets rather than risk pooling, and workers are expected to act as employee–entrepreneurs who choose providers, monitor portfolios and “make their money work” over the life course [@kerner_pension_2020; @mau_inequality_2015; @borzutzky_chiles_2016].
+As more people tie their security expectations to private investments and financial markets, their material interests become more deeply rooted in the market, and their imperatives gain stronger normative support [@mau_inequality_2015].
+When citizens are asked whether higher pensions should be allocated to those who earn and contribute more, they reflect support for the market allocation of a critical social service and the legitimacy of the resulting risks and inequalities in old age [@busemeyer_welfare_2020; @lindh_public_2015].
 
@@
 
-The study of social mobility, its drivers and consequences has long been central to sociology.
-Classic theorists explored not only movement across social hierarchies but also its broader implications for class conflict, norm stability, and institutional change [@breen_effects_2024].
-Sorokin [-@sorokin_social_1927] introduced the concept formally, defining mobility as the shift of individuals, values, or objects between positions within a stratification system, and distinguishing between horizontal and vertical forms.
-Later work differentiated intergenerational from intragenerational mobility, as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which origins constrain destinations [@eyles_social_2022].
-In Latin America, research shows high absolute but low relative mobility [@bucca_merit_2016]: although educational expansion and economic modernization have enabled some upward movement, status reproduction remains strong, especially among elites [@torche_intergenerational_2014; @lopez-roldan_comparative_2021].
-This reflects deep structural inequality, segmented education systems, stratified labor markets, and legacies of dependent development that restrict access to mobility channels and reinforce the intergenerational transmission of advantage.
+The drivers and consequences of social mobility have long been central to sociology.
+Social mobility denotes movements between positions in a stratification system.
+Classical and later work distinguish intragenerational (within the life course) from intergenerational mobility (between parents and children), as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which social origins constrain destinations [@eyles_social_2022].
+Beyond mapping mobility patterns, a large body of research examines how mobility shapes attitudes and behaviours.
+In this literature, “mobility effects” refer to outcomes that arise from movement between origin and destination classes [@breen_effects_2024].
+A growing number of studies analyse the consequences of intergenerational mobility for attitudes towards economic inequality [e.g.
+@alesina_intergenerational_2018; @bucca_merit_2016; @day_movin_2017; @gugushvili_intergenerational_2016; @helgason_class_2025].
+These inquiries rest on the idea that, if attitudes to inequality are strongly stratified by class—because classes embody distinct material interests and moral economies [@edlundDemocraticClassStruggle2015a; @kulinClassValuesAttitudes2013]—, then people who move between classes should adjust their attitudes towards those typical of their new class location.
+This study contributes to this line of work by examining how intergenerational mobility is associated with preferences for the commodification of pensions in a highly stratified and commodified welfare regime such as Chile’s.
 
-Beyond mapping mobility patterns, growing research has examined its subjective and attitudinal effects.
-Mobility effects—defined as outcomes resulting from movement between origin and destination classes [@breen_effects_2024]—have long attracted theoretical interest.
-Sorokin’s (1959) dissociative hypothesis posits that mobility, whether upward or downward, may produce psychological strain due to conflicting norms between class contexts, leading to lower life satisfaction.
-Similarly, Lenski [-@lenski_status_1954] argued that status inconsistency—mismatches among education, income, and occupation—can undermine well-being.
-These ideas underpin extensive empirical work linking intergenerational mobility to outcomes such as life satisfaction, mental health, and stress [@prag_subjective_2021; @gugushvili_heterogeneous_2024; @hadjar_does_2015].
+The literature on the effects of social mobility identifies several mechanisms through which changes in class position may affect individual attitudes [@helgason_longterm_2023].
+A first family of accounts emphasises material self-interest and can be divided into myopic and anticipatory theories.
+Myopic theories assume that individuals rapidly align their attitudes with their current class position, so mobility triggers shifts in material interests, information and perceived risks that update preferences [@helgason_longterm_2023; @ares_changing_2020; @langsaether_explaining_2022].
+Anticipatory theories stress that forward-looking agents align their current attitudes with expected future income, as in the Prospect of Upward Mobility (POUM) hypothesis, in which individuals may oppose redistribution because they anticipate moving up [@benabou_social_2001].
 
-Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality.
-A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed.
-Alesina et al.
-[-@alesina_intergenerational_2018] show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies.
-Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility.
-Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [-@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support.
-However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.
+A second family foregrounds culture and socialisation.
+Acculturation accounts posit that people gradually adapt their views to those prevailing in their destination class through exposure to new information, class-homophilous networks, group pressure and the assimilation of principles and normative beliefs [@jaime-castillo_social_2019; @helgason_class_2025].
+Socialisation perspectives, instead, argue that core attitudes are formed early in life in the class of origin through families, schools, and critical events, and remain largely stable despite later mobility [@jaime-castillo_social_2019].
+Status-maximisation theories propose that individuals align their attitudes with the highest class they have occupied, so upwardly mobile respondents converge towards their destination class, while downwardly mobile respondents retain attitudes from their higher-status origins [@jaime-castillo_social_2019].
+Helgason and Rehm [-@helgason_longterm_2023;-@helgason_class_2025] refine these ideas in a learning framework, in which attitudes evolve slowly through cumulative class experiences that combine economic resources, socialisation into class-specific values, information, and networks.
+Finally, attributional mechanisms highlight self-serving biases, suggesting that individuals explain their mobility trajectories in ways that legitimise their current position, thereby shaping beliefs about effort, deservingness and inequality [@gugushvili_intergenerational_2016; @molina_its_2019; @schmidt_experience_2011].
+In what follows, I focus on two mechanisms that are particularly relevant for market-based pensions: material self-interest and acculturation into meritocratic distributive norms.
 
-Beyond redistribution, other studies have explored the impact of mobility on broader beliefs about inequality.
-Gugushvili [-@gugushvili_intergenerational_2016] finds that upwardly mobile individuals are more likely to adopt individualistic attributions of poverty and to legitimize income inequality, particularly in post-socialist societies.
-Similarly, Bucca [-@bucca_merit_2016] shows that subjective upward mobility reinforces individualistic explanations of wealth in seven Latin American countries.
-At the macro level, Shariff et al.
-[-@shariff_income_2016] demonstrate that higher levels of national economic mobility correlate with greater tolerance of inequality.
-In contrast, Day and Fiske [-@day_movin_2017] find that low perceived mobility undermines belief in meritocracy and a just world, thereby weakening system justification.
-Taken together, this body of research suggests that mobility shapes attitudes toward inequality and justice through multiple, and sometimes contradictory, mechanisms.
+How intergenerational social mobility shapes preferences for the commodification of pensions?
+A plausible mechanism runs through changes in material self-interest.
+Social mobility alters individuals’ incomes, job security and access to economic resources [@helgason_longterm_2023; @mau_inequality_2015], thereby modifying the costs and benefits they face in market-based pension systems.
+Material interests can be understood as the pursuit of economic well-being under the constraints and trade-offs implied by one’s class position [@wrightClassCountsComparative1997a].
+Upward mobility typically entails higher and more stable earnings, greater capacity to contribute to funded schemes, greater tolerance for investment risk, and the ability to purchase supplementary coverage, whereas downward mobility reduces income, undermines contribution histories, and increases dependence on public or solidarity-based pillars [@madero-cabib_private_2019; @galvez_pensiones_2024].
+These diverging pension prospects imply that upwardly and downwardly mobile individuals face systematically different stakes in market-based pension systems, making material self-interest a first channel through which mobility may shape support for pension commodification.
 
-The literature on the effects of social mobility has proposed various mechanisms to explain how and why changes in social position may influence individual outcomes [@helgason_class_2025].
-One of the most prominent is the self-interest mechanism, which posits that individuals who experience upward or downward mobility undergo a shift in their material interests, thereby altering their perceptions and preferences [@helgason_longterm_2023; @ares_changing_2020; @langsaether_explaining_2022].
-Closely related to this logic is the Prospect of Upward Mobility (POUM) hypothesis, which suggests that individuals may oppose redistribution not because of their current position, but because they anticipate improving their status in the future [@benabou_social_2001].
-A second line of explanation draws on the framework of theories of socialization.
-Within this tradition, hypotheses such as acculturation, socialization, and status maximization propose that individuals adjust their attitudes based on the norms and values of either their class of origin or their destination, or a combination of both [@jaime-castillo_social_2019].
-However, most of these mechanisms focus primarily on the indirect effects of origin and destination positions, rather than on the direct effect of experience of movement itself.
-In response to this gap, recent research has highlighted the role of self-serving bias in causal attribution processes, suggesting that individuals tend to explain their mobility trajectories in ways that justify their current position, which in turn shapes their beliefs and preferences [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].
+Consistent with this reasoning, existing research indicates that support for welfare commodification and pension marketisation is strongly structured by material advantage.
+Individuals with higher income, education and occupational class are more likely to endorse market-based distributive principles and to view private provision as a way to secure or enhance relative advantage  [@koos_moral_2019; @vondemknesebeck_are_2016; @immergut_it_2020; @busemeyer_skills_2014; @lindh_public_2015; @svallfors_political_2007].
+In the pension field, institutional design sharpens these incentives.
+@busemeyer_welfare_2020 show that in countries where private alternatives to public pensions exist, support for public pension spending among upper-income groups erodes as they exit public schemes into private plans whose contributions and benefits closely track their income and risk.
+In Latin America, @kerner_pension_2020 finds that higher income and higher perceived financial returns from pension funds are associated with preferences for a stronger market role in pensions, suggesting that favourable returns are interpreted as evidence that pension neoliberalism “works”.
+Evidence from Chile points in the same direction: higher-income and university-educated respondents express stronger support for market allocation in healthcare, education and pensions [@castillo_socialization_2024; @otero_power_2024], and economic elites—especially those socialised in highly prestigious schools—hold more favourable attitudes towards inequality that can be traced to their material interests against redistribution [@carranza_what_2024].
+Systems that link benefits to individual contributions thus allow better-off groups to capitalise on their financial independence and minimise reliance on public support, implying that mobility into or out of these advantaged positions should alter individuals’ stakes in pension commodification.
 
-The self-serving bias mechanism builds on a value-oriented perspective, emphasizing that individuals’ experiences of social mobility shape their causal attributions, which in turn influence their beliefs about justice and distributive preferences [@gugushvili_trends_2014].
-Causal attribution refers to the process through which individuals generate explanations for their own behavior and outcomes, as well as those of others [@gugushvili_intergenerational_2016].
-In this view, people’s interpretations of economic inequality depend on whether they believe such disparities reflect unequal individual contributions.
-Individuals who adopt an internal attribution framework tend to see success or failure as rooted in personal characteristics such as effort, talent, or merit.
-In contrast, those who regard inequality as unjust are more likely to adopt an external attribution model, viewing outcomes as the result of structural barriers beyond individual control [@kluegel_beliefs_1981].
-This mechanism, often described as intrapersonal causal attribution, reflects how people explain their own socioeconomic positions—typically attributing their successes to internal qualities while blaming failures on external circumstances [@miller_selfserving_1975].
-Over time, individuals may revise their beliefs and attitudes: while early views are shaped by their social origin, these are later adjusted in light of personal experiences of mobility and the perceived role of ascribed versus achieved factors in determining socioeconomic outcomes [@gugushvili_intergenerational_2016].
+Research on redistribution and inequality offers additional, indirect support for this self-interest mechanism.
+Individuals with pessimistic mobility expectations—especially those anticipating downward movement—display stronger support for redistributive policies [@alesina_intergenerational_2018].
+Studies tracking realised mobility show that, compared to immobility, upward moves are associated with more economically conservative, less redistributive attitudes, whereas downward moves strengthen demand for state intervention and redistribution [@ares_changing_2020; @langsaether_explaining_2022; @schmidt_experience_2011; @jaime-castillo_social_2019; @rodriguez_intergenerational_2024].
+Over the life course, cumulative exposure to higher-class positions—whether by staying or moving up—is linked to more market-liberal economic values, while sustained exposure to lower-class positions—through stability or downward mobility—reinforces pro-redistributive orientations [@helgason_longterm_2023;@helgason_class_2025].
+Domain-specific evidence also aligns with changing material stakes: @gugushvili_subjective_2017 finds that upward mobility is associated with lower support for government spending on housing and pensions, whereas downward mobility reduces support for healthcare and education but increases support for housing and pensions, reflecting the material nature of these domains.
+Beyond redistribution, upward mobility fosters more individualistic attributions of poverty and wealth and greater legitimation of income inequality [@bucca_merit_2016; @gugushvili_intergenerational_2016; @gugushvili_intergenerational_2016c], and societies with higher economic mobility display more tolerance of inequality [@shariff_income_2016].
+Taken together, this evidence suggests that social mobility realigns attitudes with changing material stakes and distributive interests.
+Although the present study cannot directly test the underlying mechanism, it treats myopic self-interest as the central interpretive framework for the expected effects of intergenerational mobility on support for pension commodification.
 
-Empirical research has provided support for the self-serving bias mechanism in shaping redistributive preferences and attitudes toward the legitimacy of inequality.
-For instance, Schmidt [-@schmidt_experience_2011] finds that individuals who experience upward mobility are more likely to interpret their success as the result of personal effort or merit, and consequently perceive redistribution as less necessary.
-Conversely, individuals who experience downward mobility tend to attribute their decline to external circumstances—such as structural inequality or unemployment—and show stronger support for redistribution.
-In a comparative analysis across different welfare domains, Gugushvili [-@gugushvili_subjective_2017] demonstrates that upward mobility is associated with lower support for government spending on housing and pensions, while individuals who experience downward mobility express lower support for healthcare and education spending, but favor increased investment in housing and pensions, reflecting the material nature of these domains.
-Moreover, Gugushvili [-@gugushvili_intergenerational_2016c] finds that upward mobility is linked to greater justification of income inequality, suggesting that improved social standing reinforces an attributional view in which success is seen as the result of individual characteristics—thus legitimizing inequality as a fair outcome.
+Consequently, the main hypothesis of this research is:
 
-Consequently, considering this theoretical and empirical background, the first hypothesis of this research is that:
+$H_1$: Compared to immobility, intergenerational upward mobility is associated with stronger, and downward mobility with weaker, support for the commodification of pensions.
 
-$H1$: Experiencing upward (downward) social mobility is positively (negatively) associated with greater support for market justice in healthcare, pensions, and education.
+While material self-interest is the main lens, mobility also reshapes norms and interpretive frames.
+Mobile individuals move across social milieus and networks and are exposed to new standards and beliefs about fairness and deservingness [@helgason_longterm_2023].
+Acculturation accounts hold that people gradually adopt the attitudes of their destination class [@jaime-castillo_social_2019].
+Mobility can thus reshape how individuals interpret success, failure and inequality, with upward moves fostering more individualistic and meritocratic understandings and downward moves encouraging structural accounts of disadvantage [@gugushvili_subjective_2017; @mau_inequality_2015; @mijs_belief_2022] Applied to pensions, those who rise may come to see market-based schemes as fair systems that reward effort and foresight.
+In contrast, those who fall may view commodified pensions as unjust and insufficiently protective.
+Therefore, shifts in meritocratic norms can be a second channel through which the mobility effect operates, serving as a complementary acculturation mechanism that may influence support for pension commodification, as examined in the next section.
 
@@
 
-Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs.
-Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
-As initially conceptualized by Michael Young [-@young_rise_1958], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality.
-However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2019; @sandel_tyranny_2020].
-From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality.
-People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_meritocracia_2019; @batruch_belief_2023].
-Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.
+Meritocracy refers to a distributive system in which individual merit—typically defined as effort and talent—is the primary criterion for allocating resources and rewards, rather than social origins or inherited privilege [@young_rise_1958; @sen_merit_2000].
+Originally coined by Young in a dystopian critique of a society where power and status are justified by achievement and mobility only driven by “merit”, the term has been re-appropriated as a positive ideal of fairness, especially in liberal and market-oriented societies [@dubet_repensar_2011; @mijs_paradox_2019; @vandewerfhorst_meritocracy_2024].
+From a sociological perspective, belief in meritocracy is not just a cognitive judgement; it constitutes a moral lens through which individuals interpret social and economic disparities [@castillo_multidimensional_2023].
+The conviction that economic inequalities are justified because they reflect differential merit has been identified as a key mechanism behind the persistence of inequality, since it recasts structural advantages and disadvantages as outcomes of individual performance and encourages “winners” to see their position as earned and “losers” to internalise blame rather than question the underlying structure of opportunity [@mijs_unfulfillable_2016; @sandel_tyranny_2020; @wilson_role_2003].
+Understanding how these beliefs in meritocracy are structured is therefore crucial for analysing attitudes toward inequality.
 
-I adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through which individuals justify inequality [@young_rise_1958].
-Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility.
-A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed.
-Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023].
-The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions.
-For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023; @wiederkehr_belief_2015; @wilson_role_2003].
-This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage.
-Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.
+Recent research has emphasised the need to decompose the term “meritocratic beliefs” and to distinguish between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023; @zhu_meritocratic_2025].
+Preferences refer to normative ideas about how rewards should be allocated—whether people think effort and ability ought to determine life chances—whereas perceptions capture evaluations of how meritocracy actually operates in society, that is, whether people believe that existing inequalities reflect merit-based processes [@janmaat_subjective_2013].
+This distinction is crucial for understanding how people interpret inequality in stratified societies because it clarifies the (mis)alignment between normative support for meritocracy and evaluations about its actual implementation [@castillo_multidimensional_2023; @mijs_paradox_2019; @mijs_how_2022; @newman_economic_2023; @lindner_does_2024].
+In this study, I focus on meritocratic perceptions of effort and talent as key dimensions of broader meritocratic beliefs, given their central role in shaping attitudes toward inequality and welfare.
 
-These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
-This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025].
-Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
-Prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_perceptions_2025; @castillo_socialization_2024].
+Several empirical studies have examined the social foundations and consequences of meritocratic perceptions.
+Individuals with higher levels of education, income and occupational prestige are more likely to endorse merit-based explanations for social outcomes [@mijs_paradox_2019; @duru-bellat_who_2012; @garcia-sanchez_perceptions_2018].
+These perceptions function as a normative framework that legitimises several unequal outcomes, especially when access to social goods is stratified by income or social background.
+Individuals who see their society as meritocratic tend to show lower support for redistribution [@hoyt_mindsets_2023; @tejero-peregrina_perceived_2025], greater legitimisation of class inequality [@darnon_where_2018], increased tolerance of income gaps [@batruch_belief_2023], more support for system-justifying ideologies [@wiederkehr_belief_2015], and even lower perceived levels of economic inequality itself [@castillo_meritocracia_2019].
+This evidence provides a basis for extending the study of the attitudinal consequences of meritocracy to preferences for market-based welfare.
 
-In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups.
-Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
-This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
-Moreover, Busemeyer et al.
-[-@busemeyer_positive_2021] argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
-This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
+Preferences for the commodification of social welfare services are anchored in normative frames.
+Studies show that support for market-based welfare is higher among individuals who endorse meritocratic and liberal principles—that benefits should depend on individual effort and contributions—and among those with economically conservative orientations, whereas egalitarian views reduce support for market allocation [@lee_fairness_2023; @quadagno_has_2012; @jaime-castillo_public_2013].
+In the pension field, factorial evidence from Chile indicates that merit-based criteria such as educational attainment and years in the labour force weigh more heavily than need-based attributes in judgments of just pension levels, with respondents willing to accept very low pensions for those perceived as low achievers [@castillo_deserving_2019].
+Recently, @castillo_perceptions_2025 show that agreement with the notion that it is fair for higher-income individuals to access better welfare services was relatively low in 2016 but rose significantly by 2023, especially in the pension domain.
+They also find that individuals who endorse more meritocratic perceptions–specifically the perception that effort is rewarded– are more likely to support market-based access for healthcare, education and pensions.
+All in all, meritocratic perceptions are a powerful source of inequality legitimation, anchoring who is seen as deserving to be better or worse off, and are strongly linked to how individuals understand unequal life chances and movements up and down the social hierarchy.
 
-Importantly, recent research has explored how meritocratic beliefs interact with experiences of intergenerational mobility to shape distributive attitudes.
-The belief that one’s success is earned can lead upwardly mobile individuals to internalize meritocratic narratives and justify existing inequalities, reinforcing support for market justice [@gugushvili_intergenerational_2016c; @molina_its_2019].
-Conversely, downwardly mobile individuals who maintain strong meritocratic beliefs may interpret their status as a personal failure, reducing their support for redistribution [@day_movin_2017].
-At the macro level, Shariff et al.
-[-@shariff_income_2016] show that higher perceived mobility increases tolerance for inequality, suggesting that meritocracy and mobility are mutually reinforcing.
+Social mobility and meritocratic perceptions are closely intertwined in the acculturation mechanism, although the direction of this link can vary across trajectories and contexts.
+A large body of work consistently shows that upwardly mobile individuals tend to interpret their trajectory through a meritocratic lens, reading their success as evidence that effort and ability are rewarded and reinforcing individualistic attributions of poverty and wealth [@bucca_merit_2016; @gugushvili_intergenerational_2016; @shariff_income_2016; @mijs_belief_2022], even in a highly unequal and low mobility setting were the narratives of individual success reinforces [@deng_its_2025].
+Studies of economic elites in Chile similarly report strong endorsement of meritocracy, with respondents attributing their ascent mainly to talent and business or leadership skills [@atria_economic_2020], in line with self-serving attribution patterns [@kluegel_beliefs_1981; @miller_selfserving_1975].Evidence on downward or blocked mobility is more ambivalent.
+Exposure to low-mobility contexts and experiences of downward mobility are associated with weaker beliefs that effort is rewarded, lower just-world and opportunity beliefs, and stronger structural attributions for success and failure [@gugushvili_intergenerational_2016; @davidai_why_2018; @day_movin_2017; @mijs_how_2022].
+Yet, under strong meritocratic convictions, status loss can also be internalised as personal failure, sustaining system-justifying attitudes and the legitimation of existing hierarchies as a psychological defence strategy [@day_movin_2017; @mau_inequality_2015; @deng_its_2025].
+In sum, upward mobility is generally associated with stronger meritocratic perceptions, whereas downward mobility is more closely linked to structural explanations of inequality, without necessarily eroding meritocratic convictions.
 
-Taken together, this literature supports the idea that meritocratic beliefs can moderate the relationship between mobility and market justice preferences.
-Individuals who experience mobility—especially upward—may draw on meritocratic narratives to legitimize both their own status and broader inequalities, thereby strengthening their support for market-based distribution.
-Accordingly, the second hypothesis of this study is:
+The ways in which social mobility trajectories shape interpretations of meritocracy can determine whether mobility translates into stronger or weaker support for market-based pension welfare.
+Following Mau’s [-@mau_inequality_2015] account of the “majority class”, upward mobility is expected to strengthen market-based welfare not only because material interests change, but because those who rise come to read their story as one of individual effort and talent rewarded.
+In this view, markets are seen as fair arenas that properly allocate pensions and other welfare benefits according to merit, so upwardly mobile individuals with strong meritocratic convictions should be especially prone to endorse commodified pensions.
+In contrast, downward mobility typically fosters de-commodifying demands, yet under strong meritocratic beliefs, status loss may be read as individually deserved, weakening—or even reversing—its usual de-commodifying pattern, as market outcomes are still viewed as broadly fair regardless of direction.
+Empirical research on redistribution and attitudes toward inequality supports this mechanism.
+Survey studies suggest that subjective social mobility shapes support for redistribution and pension spending through self-serving bias [@schmidt_experience_2011; @gugushvili_subjective_2017], although recent work indicates that this mechanism is evident only under an explicit experimental design [@molina_its_2019].
+In such designs, exposure to upward-mobility information or self-made narratives reduces support for redistribution, increases tolerance of inequality and system justification, and even legitimises exploitative work conditions, whereas exposure to low or downward mobility produces the opposite pattern; crucially, these effects are partly mediated by stronger meritocratic beliefs [@matamoros-lima_social_2025; @shariff_income_2016; @deng_its_2025].
+In this framework, meritocratic beliefs condition the attitudinal consequences of mobility by amplifying the pro-market effects of upward mobility and muting the critical potential of downward mobility; thus, I treat meritocratic perceptions as a moderating mechanism linking intergenerational mobility to preferences for pension commodification.
 
-$H2$: The positive (negative) relationship between upward (downward) social mobility and support for market justice in healthcare, pensions, and education is moderated by meritocratic beliefs; specifically, this association is stronger (weaker) among individuals with higher perceptions of meritocracy.
+Accordingly, the second set of hypotheses of this study is:
 
-## The Chilean context
+$H_{2a}$: Under high meritocracy, upward mobility should generate a stronger increase in support for market-based pension access than under low meritocracy.
 
-Chile offers a compelling case for examining individual preferences for market justice in the access to core social services.
-Since the early 1990s, the country experienced a long period of sustained economic growth, with per capita GDP increasing by nearly 4% annually until the mid-2010s [@llorca-jana_historia_2021].
-This growth brought notable improvements in living standards, including a strong decline in income poverty—from 38.6% in 1990 to 6.5% in 2022 [@mideplan_evolucion_2017].
-However, this success story has been tempered by growing concerns over the unequal distribution of its benefits.
-In recent years, economic growth has slowed—exacerbated by the global impact of the COVID-19 pandemic—and structural inequalities have become more visible [@barozet_clases_2021].
-Despite its economic achievements, Chile remains one of the most unequal countries in Latin America and the OCDE.
-The poorest 50% of the population captures just 10% of total income and holds negative net wealth, while the wealthiest 1% receives nearly 27% of all income and controls almost half (49.6%) of the country’s wealth [@chancel_world_2022], a figure that has seen little change since the 1990s [@flores_top_2020].
-These stark disparities extend beyond income, reflecting deep inequalities in access to essential social services [@pnud_desiguales_2017].
+$H_{2b}$: Under low meritocracy, downward mobility should moderately reduce support for market-based pension access, whereas under high meritocracy, this negative effect should be attenuated toward neutral or slightly positive attitudes.
 
-A significant share of Chile’s inequality stems from neoliberal reforms that privatized and commodified key domains of social reproduction [@arrizabalo_milagro_1995; @ferre_welfare_2023].
-Introduced during the military dictatorship (1973–1989) and expanded under democratic governments, these reforms embedded market logic into public services through concessions, subsidies, demand-side vouchers, and pro-private regulatory frameworks [@madariaga_three_2020; @boccardo_30_2020].
-This "crowded-out" welfare model benefits higher-income groups, leaving lower-income individuals to rely on limited public options.
-Scholars argue that this neoliberal shift repurposed the state to create niches of capitalist accumulation -from water to education-, giving rise to a model of “public-service capitalism” heavily reliant on state funding [@boccardo_30_2020].
-In healthcare, around 79% of the population is covered by the public insurer FONASA, while 15% are enrolled in private ISAPREs, which offer faster and higher-quality services to wealthier groups [@mideplan_estadisticas_2024].
-The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors through private fund administrators (AFPs), yet 27% of the labor force remains outside the system due to informality [@superintendenciadepensiones_estadisticas_2024].
-Education is similarly stratified: only 30.6% of students attend fully public schools, while 54% are in voucher-funded private institutions and 9.3% in fully private schools that largely serve affluent families [@ministeriodeeducacion_resumen_2023].
-This arrangement has created a structurally segmented system, where access and quality of services depend on ability to pay, reinforced by state subsidies to private providers [@pnud_desiguales_2017].
-
-Over recent decades, Chile has undergone a significant transformation in its class structure, shaped by sustained economic growth, the expansion of the service sector, the massification of education, and major changes in the role of the state and social policy [@ruiz_chilenos_2014].
-These changes contributed to the growth of the middle sectors, a phenomena labelled as “mesocratrization” of society [@espinoza_estratificacion_2013].
-However, this expansion did not reduce structural inequalities nor facilitate upward mobility into elite positions [@espinoza_estratificacion_2013; @lopez-roldan_comparative_2021].
-While Chile shows high absolute mobility, most flows occur between adjacent classes—particularly from working classes into lower service-class positions—while access to upper classes remains limited, indicating strong elite closure and reproduction [@espinoza_estratificacion_2013; @lopez-roldan_comparative_2021; @torche_unequal_2005; @espinoza_movilidad_2014; @perez-ahumada_clases_2019].
-Though often described as “unequal but fluid”, recent evidence points to growing rigidity in relative mobility, especially after the economic slowdown, increased household debt, labor market precarity, and the COVID-19 crisis [@barozet_clases_2021].
-However, the majority of the population identifies themselves as members of the middle class [@castillo_todos_2013].
-These dynamics have fueled a legitimacy crisis: meritocracy is increasingly questioned, and many perceive that advancement depends more on informal networks than on effort or earned credentials [@barozet_clases_2021].
-The combination of constrained mobility, persistent inequality, and concentrated opportunity has eroded expectations of upward mobility and intensified social discontent [@pnud_desiguales_2017].
-
-Despite recurrent social unrest, Chile presents a paradoxical coexistence between strong conflict over inequality and widespread public legitimation of it.
-The October 2019 “social outburst”, a period of mass protests and severe political repression throughout the country,  crystallized discontent over the privatization and commodification of social services, alongside a crisis of political legitimacy [@somma_no_2021].
-A survey during the protests identified pensions, healthcare, and education as the top demands [@nucleodesociologiacontingente_informe_2020], reflecting dissatisfaction with Chile’s stratified welfare regime.
-Yet, this unrest coexists with strong meritocratic beliefs and high tolerance for income inequality.
-Castillo [-@castillo_legitimacy_2011] shows that wage gaps between occupations are consistently justified, and the larger the perceived gap, the greater its legitimation, an effect amplified by Chile’s structural inequality.
-Moreover, perceptions of unfairness due to non-meritocratic factors often reinforce belief in effort and talent as legitimate bases for success [@castillo_multidimensional_2023], which are associated with lower perceived injustice [@castillo_meritocracia_2019].
-Mac-Clure et al.
-[@mac-clure_justicia_2024] find that low and lower-middle status individuals are less likely than upper-status groups to view educational income differences as unfair.
-Qualitative research further reveals the internalization of the enterprising self [@mau_inequality_2015], understood as a way in which the subjectivization process arises in line with positioning the self in the market, generating its own market value and subjecting it to competition.
-Canales et al.
-[-@canalesceron_sujeto_2021] describes how families navigate school choice as investment decisions, while Panes [-@panes_criticas_2020] finds that many workers frame pension contributions as individual investments aimed at maximizing future returns.
-These representations suggest that market-based moral orientations are deeply embedded in Chilean subjectivities, offering fertile ground to study how support for market justice varies across welfare domains.
-
-Although still limited, recent research on market justice preferences in Chile indicates that support for these beliefs has increased in recent years across domains such as healthcare, pensions, and education, and that both objective and subjective dimensions of inequality shape them.
-Castillo et al.
-[-@castillo_perceptions_2025] show that agreement with the notion that it is fair for higher-income individuals to access better services was relatively low in 2016, but rose significantly by 2023—especially in the pension domain.
-Otero and Mendoza [-@otero_power_2024] provide evidence that social class, measured using the EGP scheme, influences these preferences: individuals in lower or subordinate class positions are less supportive of market-based access to welfare services compared to those in higher or privileged positions.
-Additionally, greater diversity in social class networks is associated with lower support for market justice.
-This is particularly relevant in the Chilean context, where one of the main determinants of class-diverse networks is intergenerational social mobility, which exposes individuals to relationships with people from different class backgrounds [@otero_lives_2022].
-Regarding subjective factors, Castillo et al.
-[@castillo_perceptions_2025; @castillo_socialization_2024] find that individuals who endorse meritocratic beliefs—such as the idea that effort and talent are rewarded in Chile—are more likely to support market justice across welfare domains.
+A summary of the hypotheses and their expected directional patterns is presented in @tbl-hipo.
 
@@
 \centering
-\caption{Overview of hypotheses with expected directional patterns}
+\caption{Overview of hypotheses with expected directional patterns {#tbl-hipo}}
 \begin{tabular}{p{3.2cm} p{3.2cm} p{4cm} p{5cm}}
@@
 \end{table}
+
+## The Chilean context
+
+Chile offers a compelling case for examining preferences for market-based pension welfare.
+Despite sustained economic growth, it remains one of the most unequal countries in Latin America and the OECD.
+The poorest 50% of the population captures just 10% of total income and holds negative net wealth, while the wealthiest 1% receives nearly 27% of all income and controls almost half of the country’s wealth [@chancel_world_2022], a pattern that has changed little since the 1990s [@flores_top_2020].
+Research on intergenerational mobility shows that Chile combines high absolute mobility with mostly short-range moves—mainly from working classes into lower service-class positions—while access to upper and elite classes remains tightly restricted, and that despite the expansion of higher education, relative mobility remains rigid and intergenerational inequalities persist [@espinoza_movilidad_2014; @lopez-roldan_comparative_2021; @salgado_uplifting_2025; @torche_intergenerational_2014].
+Crucially, a significant share of Chile’s inequality is rooted in neoliberal reforms that privatised and commodified key social services, stratifying and segmenting access to high-quality education, healthcare and housing between classes [@ferre_welfare_2023; @pnud_desiguales_2017].
+
+Introduced under the military dictatorship (1973–1989) and expanded by democratic governments, neoliberal reforms in Chile embedded market logic into public services through concessions, subsidies, vouchers and pro-private regulation [@madariaga_three_2020; @boccardo_30_2020].
+This “crowded-out” welfare model benefits higher-income groups, leaving lower-income individuals to rely on limited public options.
+Scholars argue that this neoliberal shift repurposed the state to create niches of capitalist accumulation—from water to education—giving rise to a model of “public-service capitalism” heavily reliant on state funding [@boccardo_30_2020].
+In this setting, class trajectories unfold under conditions of high inequality and market-based provision in core welfare domains, such as pensions, making Chile a beneficial context for studying support for pension commodification.
+
+In pensions, Chile was the first country to fully replace a public pay-as-you-go social insurance system with mandatory individual capitalisation: in 1981, the tripartite cajas de previsión were replaced by a fully funded defined-contribution scheme managed by private Pension Fund Administrators (AFP) [@arenas_sistemas_2019; @huber_political_2000].
+Framed by its principal architect, José Piñera, as “the mother of all battles”, the reform was a way to roll back the redistributive role of the state, shift “freedom of choice” to individuals and create a “society of worker-capitalists” [@borzutzky_pension_2012; @solimano_rise_2021].
+Dependent workers contribute 10% of their wages to individual accounts, plus an administrative fee, while employers contribute nothing; AFPs invest these mandatory savings in domestic and international financial markets, and since the 2000s affiliates choose (or are defaulted into) “multifunds” (A–E) with different risk profiles, directly exposing pensions to market volatility [@hyde_chiles_2015; @barriga_pensiones_2024].
+After more than four decades, this architecture has produced an open pension crisis: for most retirees, AFP-financed pensions fall below the minimum wage and are clearly insufficient to avoid old-age poverty, with replacement rates around 30% even among cohorts with long contribution histories [@galvez_pensiones_2024; @madero-cabib_private_2019].
+To prevent destitution, the state has been forced to re-enter through tax-funded solidarity pillars —the Nuevo Pilar Solidario (2008) and the Pensión Garantizada Universal (PGU, 2022)— that partially repair these failures [@boccardo_30_2020; @solimano_rise_2021] by topping up low or non-existent contributory pensions for the poorest and lower-middle older adults, so that for many low earners these tax-funded contributions are the main component of their total pension income [@galvez_pensiones_2024].
+Meanwhile, the AFP industry has remained highly concentrated [@hyde_chiles_2015], posting very high returns on equity (around US$ 613 million in 2024) and fostering powerful national economic actors [@ruiz_formacion_2020], leaving Chile with a structurally fragile, segmented and regressive pension system that combines poor pensions, heavy reliance on public subsidies and high, sustained profits for private financial intermediaries.
+Chile’s combination of constrained mobility, persistent inequality and deep privatisation has eroded expectations of upward mobility and fuelled social discontent [@pnud_desiguales_2017], with pensions at the centre of this conflict.
+
+Despite recurrent social unrest, Chile exhibits a paradoxical coexistence of intense conflict over inequality and its widespread legitimation, vividly illustrated by pensions.
+The October 2019 “social outburst”, a period of mass protests and severe political repression throughout the country, crystallised discontent over the privatisation and commodification of social services, alongside a crisis of political legitimacy [@somma_no_2021].
+One of the main conflicts in this period centred on grievances over low pensions and the profits of private pension funds, building on earlier mobilisation by the No+AFP movement since 2016; a survey during the 2019 protests identified pensions as the top public demand [@nudesoc_informe_2020].
+This conflict helped place comprehensive pension reform on the political agenda after 2022.
+However, contestation coexists with enduring market-conforming orientations.
+Research documents strong meritocratic beliefs and high tolerance for inequality [@castillo_meritocracia_2019; @mac-clure_justicia_2024], and the COVID-19 emergency withdrawals of pension savings gave rise to the slogan “with my money, no”, expressing opposition both to greater pay-as-you-go elements and to continued AFP control [@galvez_con_2023].
+This stance reflects the internalisation of the “enterprising self” [@mau_inequality_2015]: families navigate school choice as investment strategies [@canalesceron_sujeto_2021], and workers denounce the pension system as illegitimate in its design and outcomes while simultaneously treating their contributions as personal investments aimed at maximising returns and asserting that they themselves —rather than AFPs, the state or “society”— are the legitimate owners of their pension funds [@panes_criticas_2020].
+These ambivalent representations indicate that market-based moral orientations are deeply embedded, making Chile a fertile setting to study how support for pension commodification varies across social groups and meritocratic beliefs.

### 2025-11-30 paper/04-analysis.qmd
Up antecedentes

@@
 
-## Descriptive statistics
+## Descriptive analysis
 
@@
 @fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic beliefs.
-Across trajectories, mobility effects for low-merit respondents are uniformly small and imprecise, hovering around zero for both upward and downward moves.
-Among high-merit respondents, the pattern aligns only partially with the proposed mechanism: upward mobility from middle to high status shows a clear and statistically significant pro-commodification effect ($\beta$ = 0.22, 95% CI [0.05, 0.38]), whereas all other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero or display negative estimates that do not systematically weaken under high merit.
-Formal interaction tests also fail to detect strong heterogeneity by meritocratic beliefs (see [Supplementary Material](#anexo) for interaction models).
-Taken together, the results provide only limited support for the mechanism hypothesis: meritocracy appears to amplify the pro-market consequences of upward mobility in one specific trajectory, but there is no general or consistent evidence that meritocratic beliefs constitute a stable channel through which intergenerational mobility shapes preferences for pension commodification.
+Across trajectories, mobility effects among low-merit respondents are uniformly small, imprecise, and centered around zero for both upward and downward moves.
+Among high-merit respondents, only upward mobility from middle to high status shows a clear pro-commodification pattern, with a positive and statistically significant effect on MJP for pensions ($\beta = 0.22$, 95% CI [0.05, 0.38]).
+All other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero.
+Nonetheless, the point estimates broadly align with the theoretical expectations for upward moves: upwardly mobile individuals with stronger meritocratic beliefs tend to display higher support for pension commodification than their low-merit counterparts.
+By contrast, downward trajectories exhibit a pattern opposite to the mechanism hypothesis, with high-merit respondents generally showing more de-commodifying attitudes than those with weaker meritocratic beliefs.
+However, formal interaction tests do not detect robust heterogeneity in mobility effects by meritocratic beliefs (see [Supplementary Material](#anexo) for full interaction models): even for the Middle→High trajectory, the simple effect under high meritocracy does not differ significantly from its low-merit counterpart.
+Taken together, these results do not provide convincing support for the mechanism hypothesis: meritocratic beliefs do not operate as a consistent or general channel through which intergenerational mobility shapes preferences for the commodification of pensions.
 

### 2025-11-30 paper/paper.tex
Up antecedentes

@@
 \setstretch{1.5}
-This document was last modified at 2025-11-21 21:08:43 and it was last
-rendered at 2025-11-21 21:08:43.
+This document was last modified at 2025-11-29 23:44:43 and it was last
+rendered at 2025-11-29 23:44:43.
 
@@
 preferences}---the extent to which citizens regard market-based criteria
-as a fair basis for access to essential services such as healthcare,
-education and pensions
+as a fair basis for the allocation of essential services such as
+healthcare, education and pensions
 (\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
@@
 \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
-orientations matter because they help legitimise economic inequality by
-portraying stratified outcomes as a fair result of individual
-responsibility and limited state intervention.
+orientations matter because they legitimize unequal outcomes as the
+product of individual responsibility and promote the conception of basic
+social services as commodities.
 
@@
 \citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
-reshapes how pension institutions allocate old-age risk, tying benefit
-levels to lifetime labour-market trajectories and increasing exposure to
-poverty in retirement
+reshapes not only how pension institutions allocate old-age risk and tie
+benefit levels to lifetime labour-market trajectories
 (\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
-\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). It also transforms the
+\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also transforms the
 justice principles citizens associate with old-age security: they must
@@
 solidaristic principles
-(\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
-2019}; \citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Although
-we know something about what people consider a ``just'' monetary pension
+(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Although we know
+something about what people consider a ``just'' monetary pension
 (\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
@@
 Research shows that support for market-based access to welfare services
-is strongly stratified: individuals with higher income, education or
-class position are more likely to endorse market-based criteria than
-disadvantaged groups (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
-2014}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+is strongly stratified: people with higher socioeconomic status are more
+likely to support market-based criteria than disadvantaged groups
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
 \citeproc{ref-lindh_public_2015}{Lindh, 2015};
@@
 market-oriented pension policies
-(\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
+(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
+\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
 little about how movement within the class structure shapes preferences
@@
 Existing accounts identify several mechanisms through which
-intergenerational mobility may shape distributive-justice preferences
-but converge on two channels particularly relevant for market-based
-welfare: changes in material stakes and shifts in normative frames
+intergenerational mobility may shape distributive justice preferences
+but converge on two channels especially relevant for market-based
+welfare: changes in material interests and shifts in normative frames
 (\citeproc{ref-ares_changing_2020}{Ares, 2020};
@@
 \citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}). Recent research shows that meritocratic
-beliefs---seeing that effort and ability are rewarded---are strongly
-associated with support for market allocation
+Marqués-Perales, 2019}). Mobility alters individuals' material stakes by
+changing their occupational status, income, and risk exposure. At the
+same time, it reshapes their fairness judgments. Empirically, upward
+mobility is often associated with stronger meritocratic interpretations
+of success and fairness, whereas downward mobility is more closely
+linked to structural explanations of inequality without necessarily
+eroding meritocratic convictions (\citeproc{ref-bucca_merit_2016}{Bucca,
+2016}; \citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
+\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). Regarding
+market-based preferences, recent research shows that meritocratic
+beliefs -- the conviction that effort and ability are rewarded -- are
+strongly associated with support for market allocation
 (\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}). Perceived upward mobility reinforces meritocratic
-interpretations of success, whereas downward mobility is more closely
-linked to structural explanations of inequality without necessarily
-eroding meritocratic convictions (\citeproc{ref-mijs_belief_2022}{Mijs
-et al., 2022}). Thus, meritocracy and mobility operate on partly
-separate tracks yet intersect in systematic ways. I treat meritocratic
-beliefs as a post-treatment moderator---a diagnostic lens on potential
-mechanisms. If mobility partly shapes preferences by strengthening
-merit-based understandings of fairness and altering material stakes, its
-effects should be most visible---and most pro-market---among individuals
-who already interpret social outcomes through meritocratic lenses.
-
-This article pursues two aims: first, to assess whether
-intergenerational occupational mobility causally affects support for the
-commodification of pensions; and second, to examine whether this effect
-varies by meritocratic beliefs as a diagnostic of possible transmission
-channels. I argue that social mobility operates through two main
-pathways. Upward mobility increases earnings, expected returns in a
-funded system and perceived economic security, making contribution-based
-access more appealing. At the same time, exposure to higher-class
-networks reinforces individualistic and effort-based understandings of
-fairness. Downward mobility tends to heighten economic insecurity,
-dependence on public protection and attachment to external attributions
+2025}). In such contexts, the upwardly mobile are especially likely to
+view markets as just distributors, since their own trajectories appear
+to confirm that markets appropriately reward effort; those who
+experience downward mobility may still interpret outcomes through a
+meritocratic lens, often reading losses as personal failure rather than
+as evidence against market fairness. Thus, mobility and meritocracy
+evolve on partly separate tracks yet intersect systematically
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
+
+Against this backdrop, the central aim of this article is to assess
+whether intergenerational occupational mobility causally affects support
+for the commodification of pensions. A second, more exploratory aim is
+to examine whether this effect varies by meritocratic beliefs, using
+this heterogeneity as an indication of possible underlying mechanisms. I
+argue that social mobility operates through two main pathways
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Upward mobility
+increases earnings, expected returns in a funded system, and perceived
+economic security, making contribution-based access more appealing.
+Simultaneously, upward moves reinforce individualistic, effort-based
+understandings of success and, thus, the view that markets are just
+assignors. Downward mobility tends to heighten economic insecurity,
+dependence on public protection, and attachment to external attributions
 of poverty, reducing support for market allocation. Meritocratic beliefs
-are conceptualised not as mediators but as moderators. Among individuals
-with stronger meritocratic beliefs, upward mobility should heighten
-support for market-based access. In contrast, downward mobility should
-weaken---or even reverse---its usual de-commodifying pattern, as
-meritocratic interpretations of effort lead individuals to view market
-outcomes as fair regardless of direction. The paper contributes to
-research on market justice by identifying the causal effects of mobility
-and by using meritocracy to probe whether these effects align with a
-meritocratic channel.
-
-The study focuses on Chile, an instructive case for examining
+are treated not as mediators but as moderators of these effects: among
+strong meritocrats, upward mobility should amplify pro-commodification
+attitudes, whereas downward mobility should attenuate -- or even reverse
+-- its expected de-commodifying impact, as meritocratic interpretations
+of effort lead individuals to view market outcomes as fair regardless of
+direction. This inquiry contributes to research on market justice by
+identifying the causal effects of mobility and using variation in
+meritocratic beliefs to assess whether these effects follow a
+meritocratic pathway.
+
+This study focuses on Chile, an instructive case for examining
 preferences for pension commodification. Despite sustained economic
@@
 \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
-first country worldwide to fully replace a public pay-as-you-go system
-with a mandatory, privately administered defined-contribution scheme,
-while the state has progressively expanded a segmented solidarity pillar
-to compensate for insufficient benefits
+first country worldwide in 1981 to fully replace a public pay-as-you-go
+system with a mandatory, privately administered defined-contribution
+scheme, while the state has progressively expanded a segmented
+solidarity pillar to compensate for insufficient benefits
 (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
@@
 
+In this context, the questions that guide this research are as follows:
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\tightlist
+\item
+  How does intergenerational social mobility affect individuals'
+  preferences for the commodification of pensions in Chile?
+\item
+  How do meritocratic beliefs condition or moderate this relationship?
+\end{enumerate}
+
 To address these questions, the study draws on large-scale,
@@
 
-\subsection{Preferences for market-based
-welfare}\label{preferences-for-market-based-welfare}
-
-Beyond state's capacity to reallocate resources from the advantaged to
-the vulnerable, market institutions also play a central role in
-distributing socially valuable goods and rewards
-(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As Polanyi
-(\citeproc{ref-polanyi_great_1975}{1975}) observed, economic integration
-in capitalist societies is primarily organized through market exchange,
-governed by a self-regulating price system embedded in institutional
-frameworks. These institutions are not mere aggregates of individual
-behavior, but social realities endowed with rules, mechanisms, and
-normative meanings that shape everyday thinking
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
+\subsection{Preferences for pensions
+commodification}\label{preferences-for-pensions-commodification}
+
+Beyond the state's capacity to redistribute resources, market
+institutions are central arenas for allocating socially valuable goods
+and risks (\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
+Markets are not mere aggregates of individual choices but social
+institutions endowed with rules and normative meanings, built through
+``moral projects, saturated with normativity'' that permeate everyday
+thinking around fairness, effort and deservingness
+(\citeproc{ref-fourcade_moral_2007}{Fourcade \& Healy, 2007};
 \citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
-the economic order is mirrored in a moral economy: collectively shared
-norms and beliefs about justice in distribution, embedded and reinforced
-through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
-research from this perspective has focused on welfare institutions,
-recent scholarship has brought the market back into focus as a site of
-distributive justice beliefs and institutional responses to inequality
-(\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In many
-countries, privatization and commodification have expanded market logic
-into core areas of social reproduction, such as healthcare, education,
-pensions, and social security, deepening inequality in access to these
-services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
-\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
-for market-based welfare provision has grown, even in traditional
-welfare states, particularly among higher-income groups who view private
-alternatives as more efficient or higher in quality
-(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
-This shift calls for broader inquiry into how institutions like markets
-structure access to resources and legitimize inequalities
-(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
-\citeproc{ref-satz_por_2019}{Satz, 2019}).
-
-The legitimacy of market-based inequalities is closely tied to beliefs
-about distributive justice grounded in market principles. Beyond
-stratification structures and socioeconomic conditions, the study of
-inequality also focuses on individuals' beliefs about the origins of
-inequality, the normative frameworks that sustain those beliefs, the
-mechanisms that shape them, and their implications for attitudes and
-behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
-In this regard, empirical research on distributive justice focuses on
-individuals' conceptions of how goods and rewards \emph{should} be
-distributed in society (\citeproc{ref-jasso_distributive_2016}{Jasso et
-al., 2016}). This line of inquiry allows for examining the extent to
-which economic inequality is perceived as just, or in a certain sense,
-legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
-Most of the research on this field explores the legitimacy of wage
-inequality, specially salary gaps between jobs at opposite ends of the
-occupational hierarchy
+the economic order is mirrored in a moral economy: shared norms and
+beliefs about what is considered a just distribution, embedded and
+reinforced through institutions and policy feedback
+(\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). In an
+era of privatisation and commodification, market logic has expanded into
+core areas of social reproduction---childcare, healthcare, education and
+pensions---stratifying access and quality
+(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
+\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}) and generating
+feedback effects on public beliefs about welfare and market inequality
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In this context,
+support for market-based welfare provision has grown worldwide,
+particularly among higher-income groups who view private alternatives as
+more efficient or higher-quality
+(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These institutional
+transformations provide the backdrop for analysing how far citizens
+accept the commodification of social protection
+(\citeproc{ref-satz_por_2019}{Satz, 2019}).
+
+The legitimacy of market-based welfare is closely tied to beliefs about
+distributive justice grounded in market principles. Beyond
+stratification structures, citizens hold structured beliefs about how
+resources ought to be allocated, reflecting causal attributions for
+inequality, normative principles and expectations about deservingness
+(\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}). Research
+on distributive justice examines conceptions of how goods and rewards
+should be distributed and when inequality is considered just
 (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
-\citeproc{ref-jasso_justice_1978}{Jasso, 1978};
-\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}). Recently,
-others examines how individuals justify market-generated inequalities in
-access to core social services such as healthcare, education, pensions
-or social security. Here, legitimacy stems from the belief that access
-to these goods should follow market-based criteria
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
-services are treated as legitimate commodities; goods that can be
-traded, priced, and evaluated through market logic
-(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
-
-Market justice preferences refer to normative beliefs that legitimize
-the idea that access to core social services---such as healthcare,
-education, or pensions---should be determined by market-based criteria.
-Following Janmaat's (\citeproc{ref-janmaat_subjective_2013}{2013, p.
-359}) distinction, these preferences fall under the category of
-``beliefs,'' understood as normative ideas about what inequality should
-look like, as opposed to ``perceptions,'' which refer to subjective
-evaluations of existing inequality. Market justice preferences reflect
-the view that access to these services should depend on individuals'
-ability to pay, thus justifying inequalities generated by market
-mechanisms (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al.,
-1999}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). The concept draws
-on Lane's (\citeproc{ref-lane_market_1986}{1986}) classic contrast
-between market justice and political justice: the former is grounded in
-the idea of earned deserts---where rewards reflect effort, productivity,
-and skill---while the latter prioritizes need and equality, typically
-expressed in welfare state policies. Lane argued that markets and states
-differ in purpose (efficiency vs.~need), logic (individual
-vs.~collective), and fairness criteria (merit vs.~equality). Market
-justice assumes that markets are neutral, self-regulating systems in
-which fair procedures yield outcomes proportional to merit. Inequality,
-from this perspective, is not only expected but legitimate---as long as
-it arises from fair competition. In this way, market justice offers a
-moral lens through which individuals can view the commodification and
-stratified access to social services as fair and justified
+\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}), while work on
+redistributive preferences focuses on support for state-led mechanisms
+that reduce inequality (\citeproc{ref-cavaille_fair_2025}{Cavaillé,
+2025}). Situated at this intersection, the emerging literature on market
+justice preferences conceptualises the market itself as a site of
+redistribution and asks whether individuals regard market criteria as
+fair bases for allocating wages and access to core welfare services
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In this
+perspective, the expansion of market logic into welfare domains implies
+that social services are treated as legitimate commodities that can be
+priced and stratified (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
+\& Iversen, 2020}), and empirical work suggests that such institutional
+changes feed back into attitudes, reinforcing market-conforming
+understandings of fairness and responsibility
+(\citeproc{ref-zhu_policy_2015}{Ling Zhu \& Lipsmeyer, 2015}).
+
+Market justice preferences refer to normative beliefs that legitimise
+allocating goods and services according to market criteria such as
+income and purchasing power
 (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
-
-The empirical study of market justice preferences has employed various
-strategies to capture how individuals assess inequalities arising from
-market allocation. A common approach evaluates whether people find it
-fair that access to essential social services---such as healthcare,
-education, or pensions---depends on income. This builds on foundational
-work by Kluegel and Smith
-(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
-justifications for capitalist systems. Subsequent studies have extended
-this logic to a wider set of welfare goods. For example, Immergut and
-Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
-et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
-respondents believe it is just that higher-income individuals receive
-better healthcare. Similarly, Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) and Castillo, Iturra, et al.
-(\citeproc{ref-castillo_changes_2025}{2025}) apply this framework to
-education. Complementing these efforts, more recent research has
-introduced composite indicators to capture broader orientations toward
-market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
-for instance, constructs an index averaging support for income-based
-access to healthcare and education in comparative perspective, while
-Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
-propose a single-item measure covering health, education, and pensions
-in Chile. These instruments seek to gauge to what extent individuals
-view market-generated inequalities as legitimate. Taken together, they
-capture two core dimensions of market distribution: the role of economic
-resources as a key determinant of outcomes, and the framing of services
-as tradable commodities that can be bought and sold according to ability
-to pay (\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
-
-Comparative empirical research has identified several individual-level
-factors that influence support for market justice. Individuals in more
-advantaged socioeconomic positions---those with higher income,
-education, and occupational status---are consistently more likely to
-endorse market-based distributive principles
-(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
-Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
-from the service class are more likely to support market-based access to
-healthcare and education than skilled and unskilled workers across 17
-relatively affluent countries. In a comparative analysis, Svallfors
-(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
-expected class pattern appears clearly only in Sweden, where support for
-private education and healthcare varies systematically by class.
-Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
-that support for private education is stronger among high-income groups,
-while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
-von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
-report comparable findings for healthcare, suggesting that wealthier
-individuals perceive private provision as a means of maintaining
-relative advantage. In Chile, Otero and Mendoza
-(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
-higher income and university education express stronger support for
-market allocation in healthcare, education, and pensions. Beliefs about
-inequality and political orientation also matter. Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) and Castillo,
-Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
-that individuals who have strong meritocratic beliefs are more likely to
-support market-based distribution in Chile, while Lee and Stacy
-(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
-preferences are also greater as people lean toward economic
-conservatism. In this sense, market justice preferences are shaped by
-the interplay between structural position and normative reasoning.
-
-However, individual characteristics alone do not fully explain variation
-in support for market justice. A growing body of cross-national research
-highlights the importance of institutional arrangements in shaping these
-preferences. For instance, Immergut and Schneider
-(\citeproc{ref-immergut_it_2020}{2020}) find that in countries with
-higher public spending on healthcare, individuals are less likely to
-view income-based access as fair. Similarly, Busemeyer
-(\citeproc{ref-busemeyer_skills_2014}{2014}) shows that increased public
-investment in education is associated with lower support for privatized
-provision. Conversely, Lindh (\citeproc{ref-lindh_public_2015}{2015})
-finds that in countries with more market-oriented welfare systems,
-support for market-based distribution tends to be higher, suggesting
-that individual attitudes often align with institutional outputs. These
-findings are consistent with neo-institutionalist and policy feedback
-theories (\citeproc{ref-campbell_institutional_2020}{Campbell, 2020}),
-which argue that institutions do more than redistribute resources---they
-also shape the normative categories through which individuals assess who
-is deserving of support
-(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}). In this
-view, dominant values and preferences are both shaped by and embedded in
-institutional configurations
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014}), reinforcing the
-notion that institutions are not neutral structures, but active
-producers of the moral frameworks that legitimize or challenge
-inequality.
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Building on Lane's
+(\citeproc{ref-lane_market_1986}{1986}) contrast between market and
+political justice, market justice is a distributive principle that
+endorses rewarding effort, productivity and skill rather than need and
+equality as emphasised in welfare-state policies. In this study, I treat
+market justice less as an abstract doctrine and more concretely as
+preferences for commodification: support for treating welfare services
+as commodities, so that better services follow from higher income or
+ability to pay, and the resulting stratified access is seen as fair
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Empirically, such
+orientations are typically measured with items asking whether it is fair
+that access to services depends on income, a strategy rooted in research
+on justifications of capitalist inequality
+(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999}) and
+applied to healthcare and education
+(\citeproc{ref-castillo_changes_2025}{Castillo, Iturra, et al., 2025};
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
+\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
+2016}). Moving beyond single domains, Lindh
+(\citeproc{ref-lindh_public_2015}{2015}) develops a comparative index of
+income-based access to healthcare and education, while Castillo,
+Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) use the
+same single-item measure but add the pensions domain in Chile. These
+constructs gauge how individuals view market-generated inequalities as
+legitimate and capture two core dimensions of market distribution: the
+role of income in determining attainment and the framing of services as
+commodities that can be bought and sold according to ability to pay
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
+
+This study examines support for the commodification of pensions, defined
+as normative judgements on whether better pensions should depend on
+income or individual contributions. I treat this as a domain-specific
+form of market justice: pensions are conceived as commodities whose
+allocation should follow contribution- and return-based criteria rather
+than principles of need (\citeproc{ref-lane_market_1986}{Lane, 1986};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}). The spread of funded,
+privately managed schemes built on individual accounts has transformed
+both the distributive and normative structure of old-age security
+(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
+\citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
+Distributively, these schemes reassign risk from the collective to the
+individual by tying benefits to formal employment, contribution density,
+earnings and financial returns, making retirement income heavily
+dependent on labour-market trajectories and capital-market success
+(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
+\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). Normatively, they recast
+pensions as the outcome of personal investment decisions rather than
+collective obligations, promoting ideals of individual responsibility,
+self-economisation and contribution-based desert
+(\citeproc{ref-arza_pension_2008}{Arza, 2008};
+\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Contributions
+are increasingly framed as private assets rather than risk pooling, and
+workers are expected to act as employee--entrepreneurs who choose
+providers, monitor portfolios and ``make their money work'' over the
+life course (\citeproc{ref-borzutzky_chiles_2016}{Borzutzky \& Hyde,
+2016}; \citeproc{ref-kerner_pension_2020}{Kerner, 2020};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As more people tie their
+security expectations to private investments and financial markets,
+their material interests become more deeply rooted in the market, and
+their imperatives gain stronger normative support
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). When citizens are asked
+whether higher pensions should be allocated to those who earn and
+contribute more, they reflect support for the market allocation of a
+critical social service and the legitimacy of the resulting risks and
+inequalities in old age (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
+\& Iversen, 2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).
 
@@
 
-The study of social mobility, its drivers and consequences has long been
-central to sociology. Classic theorists explored not only movement
-across social hierarchies but also its broader implications for class
-conflict, norm stability, and institutional change
-(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}). Sorokin
-(\citeproc{ref-sorokin_social_1927}{1927}) introduced the concept
-formally, defining mobility as the shift of individuals, values, or
-objects between positions within a stratification system, and
-distinguishing between horizontal and vertical forms. Later work
-differentiated intergenerational from intragenerational mobility, as
-well as absolute mobility---driven by structural change---from relative
-mobility, which captures the extent to which origins constrain
-destinations (\citeproc{ref-eyles_social_2022}{Eyles et al., 2022}). In
-Latin America, research shows high absolute but low relative mobility
-(\citeproc{ref-bucca_merit_2016}{Bucca, 2016}): although educational
-expansion and economic modernization have enabled some upward movement,
-status reproduction remains strong, especially among elites
-(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). This
-reflects deep structural inequality, segmented education systems,
-stratified labor markets, and legacies of dependent development that
-restrict access to mobility channels and reinforce the intergenerational
-transmission of advantage.
-
-Beyond mapping mobility patterns, growing research has examined its
-subjective and attitudinal effects. Mobility effects---defined as
-outcomes resulting from movement between origin and destination classes
-(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024})---have long
-attracted theoretical interest. Sorokin's (1959) dissociative hypothesis
-posits that mobility, whether upward or downward, may produce
-psychological strain due to conflicting norms between class contexts,
-leading to lower life satisfaction. Similarly, Lenski
-(\citeproc{ref-lenski_status_1954}{1954}) argued that status
-inconsistency---mismatches among education, income, and occupation---can
-undermine well-being. These ideas underpin extensive empirical work
-linking intergenerational mobility to outcomes such as life
-satisfaction, mental health, and stress
-(\citeproc{ref-gugushvili_heterogeneous_2024}{Gugushvili, 2024};
-\citeproc{ref-hadjar_does_2015}{Hadjar \& Samuel, 2015};
-\citeproc{ref-prag_subjective_2021}{Präg \& Gugushvili, 2021}).
-
-Research on intergenerational social mobility has increasingly examined
-its effects on attitudes toward economic inequality. A key area of
-inquiry focuses on how upward and downward mobility influence support
-for redistribution, though findings remain mixed. Alesina et al.
-(\citeproc{ref-alesina_intergenerational_2018}{2018}) show that
-individuals with pessimistic expectations about their
-mobility---particularly those anticipating downward movement---are more
-likely to support generous redistributive policies. Similarly, Ares
-(\citeproc{ref-ares_changing_2020}{2020}) finds that upwardly mobile
-individuals tend to be less supportive of state-led redistribution
-compared to those who have experienced downward mobility. Comparative
-studies by Schmidt (\citeproc{ref-schmidt_experience_2011}{2011}) and
-Gugushvili (\citeproc{ref-gugushvili_subjective_2017}{2017}) likewise
-report that subjective upward mobility is associated with weaker
-preferences for redistribution, while downward mobility strengthens
-redistributive support. However, recent causal evidence from Breen and
-Ermisch (\citeproc{ref-breen_effects_2024}{2024}) suggests the opposite:
-upward mobility may increase redistributive preferences, while downward
-mobility may reduce them.
-
-Beyond redistribution, other studies have explored the impact of
-mobility on broader beliefs about inequality. Gugushvili
-(\citeproc{ref-gugushvili_intergenerational_2016}{2016b}) finds that
-upwardly mobile individuals are more likely to adopt individualistic
-attributions of poverty and to legitimize income inequality,
-particularly in post-socialist societies. Similarly, Bucca
-(\citeproc{ref-bucca_merit_2016}{2016}) shows that subjective upward
-mobility reinforces individualistic explanations of wealth in seven
-Latin American countries. At the macro level, Shariff et al.
-(\citeproc{ref-shariff_income_2016}{2016}) demonstrate that higher
-levels of national economic mobility correlate with greater tolerance of
-inequality. In contrast, Day and Fiske
-(\citeproc{ref-day_movin_2017}{2017}) find that low perceived mobility
-undermines belief in meritocracy and a just world, thereby weakening
-system justification. Taken together, this body of research suggests
-that mobility shapes attitudes toward inequality and justice through
-multiple, and sometimes contradictory, mechanisms.
-
-The literature on the effects of social mobility has proposed various
-mechanisms to explain how and why changes in social position may
-influence individual outcomes
-(\citeproc{ref-helgason_class_2025}{Helgason \& Rehm, 2025}). One of the
-most prominent is the self-interest mechanism, which posits that
-individuals who experience upward or downward mobility undergo a shift
-in their material interests, thereby altering their perceptions and
-preferences (\citeproc{ref-ares_changing_2020}{Ares, 2020};
+The drivers and consequences of social mobility have long been central
+to sociology. Social mobility denotes movements between positions in a
+stratification system. Classical and later work distinguish
+intragenerational (within the life course) from intergenerational
+mobility (between parents and children), as well as absolute
+mobility---driven by structural change---from relative mobility, which
+captures the extent to which social origins constrain destinations
+(\citeproc{ref-eyles_social_2022}{Eyles et al., 2022}). Beyond mapping
+mobility patterns, a large body of research examines how mobility shapes
+attitudes and behaviours. In this literature, ``mobility effects'' refer
+to outcomes that arise from movement between origin and destination
+classes (\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}). A
+growing number of studies analyse the consequences of intergenerational
+mobility for attitudes towards economic inequality (e.g.
+\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
+\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
+\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
+\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
+\citeproc{ref-helgason_class_2025}{Helgason \& Rehm, 2025}). These
+inquiries rest on the idea that, if attitudes to inequality are strongly
+stratified by class---because classes embody distinct material interests
+and moral economies
+(\citeproc{ref-edlundDemocraticClassStruggle2015a}{Edlund \& Lindh,
+2015}; \citeproc{ref-kulinClassValuesAttitudes2013}{Kulin \& Svallfors,
+2013})---, then people who move between classes should adjust their
+attitudes towards those typical of their new class location. This study
+contributes to this line of work by examining how intergenerational
+mobility is associated with preferences for the commodification of
+pensions in a highly stratified and commodified welfare regime such as
+Chile's.
+
+The literature on the effects of social mobility identifies several
+mechanisms through which changes in class position may affect individual
+attitudes (\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm,
+2023}). A first family of accounts emphasises material self-interest and
+can be divided into myopic and anticipatory theories. Myopic theories
+assume that individuals rapidly align their attitudes with their current
+class position, so mobility triggers shifts in material interests,
+information and perceived risks that update preferences
+(\citeproc{ref-ares_changing_2020}{Ares, 2020};
 \citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
 \citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
-Closely related to this logic is the Prospect of Upward Mobility (POUM)
-hypothesis, which suggests that individuals may oppose redistribution
-not because of their current position, but because they anticipate
-improving their status in the future
-(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}). A second line
-of explanation draws on the framework of theories of socialization.
-Within this tradition, hypotheses such as acculturation, socialization,
-and status maximization propose that individuals adjust their attitudes
-based on the norms and values of either their class of origin or their
-destination, or a combination of both
+Anticipatory theories stress that forward-looking agents align their
+current attitudes with expected future income, as in the Prospect of
+Upward Mobility (POUM) hypothesis, in which individuals may oppose
+redistribution because they anticipate moving up
+(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}).
+
+A second family foregrounds culture and socialisation. Acculturation
+accounts posit that people gradually adapt their views to those
+prevailing in their destination class through exposure to new
+information, class-homophilous networks, group pressure and the
+assimilation of principles and normative beliefs
+(\citeproc{ref-helgason_class_2025}{Helgason \& Rehm, 2025};
+\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}). Socialisation perspectives, instead, argue that
+core attitudes are formed early in life in the class of origin through
+families, schools, and critical events, and remain largely stable
+despite later mobility
 (\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
-Marqués-Perales, 2019}). However, most of these mechanisms focus
-primarily on the indirect effects of origin and destination positions,
-rather than on the direct effect of experience of movement itself. In
-response to this gap, recent research has highlighted the role of
-self-serving bias in causal attribution processes, suggesting that
-individuals tend to explain their mobility trajectories in ways that
-justify their current position, which in turn shapes their beliefs and
-preferences
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
+Marqués-Perales, 2019}). Status-maximisation theories propose that
+individuals align their attitudes with the highest class they have
+occupied, so upwardly mobile respondents converge towards their
+destination class, while downwardly mobile respondents retain attitudes
+from their higher-status origins
+(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}). Helgason and Rehm
+(\citeproc{ref-helgason_longterm_2023}{2023},
+\citeproc{ref-helgason_class_2025}{2025}) refine these ideas in a
+learning framework, in which attitudes evolve slowly through cumulative
+class experiences that combine economic resources, socialisation into
+class-specific values, information, and networks. Finally, attributional
+mechanisms highlight self-serving biases, suggesting that individuals
+explain their mobility trajectories in ways that legitimise their
+current position, thereby shaping beliefs about effort, deservingness
+and inequality
+(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
 \citeproc{ref-molina_its_2019}{Molina et al., 2019};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}).
-
-The self-serving bias mechanism builds on a value-oriented perspective,
-emphasizing that individuals' experiences of social mobility shape their
-causal attributions, which in turn influence their beliefs about justice
-and distributive preferences
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Causal
-attribution refers to the process through which individuals generate
-explanations for their own behavior and outcomes, as well as those of
-others (\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
-2016b}). In this view, people's interpretations of economic inequality
-depend on whether they believe such disparities reflect unequal
-individual contributions. Individuals who adopt an internal attribution
-framework tend to see success or failure as rooted in personal
-characteristics such as effort, talent, or merit. In contrast, those who
-regard inequality as unjust are more likely to adopt an external
-attribution model, viewing outcomes as the result of structural barriers
-beyond individual control (\citeproc{ref-kluegel_beliefs_1981}{Kluegel
-\& Smith, 1981}). This mechanism, often described as intrapersonal
-causal attribution, reflects how people explain their own socioeconomic
-positions---typically attributing their successes to internal qualities
-while blaming failures on external circumstances
-(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Over
-time, individuals may revise their beliefs and attitudes: while early
-views are shaped by their social origin, these are later adjusted in
-light of personal experiences of mobility and the perceived role of
-ascribed versus achieved factors in determining socioeconomic outcomes
-(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b}).
-
-Empirical research has provided support for the self-serving bias
-mechanism in shaping redistributive preferences and attitudes toward the
-legitimacy of inequality. For instance, Schmidt
-(\citeproc{ref-schmidt_experience_2011}{2011}) finds that individuals
-who experience upward mobility are more likely to interpret their
-success as the result of personal effort or merit, and consequently
-perceive redistribution as less necessary. Conversely, individuals who
-experience downward mobility tend to attribute their decline to external
-circumstances---such as structural inequality or unemployment---and show
-stronger support for redistribution. In a comparative analysis across
-different welfare domains, Gugushvili
-(\citeproc{ref-gugushvili_subjective_2017}{2017}) demonstrates that
-upward mobility is associated with lower support for government spending
-on housing and pensions, while individuals who experience downward
-mobility express lower support for healthcare and education spending,
-but favor increased investment in housing and pensions, reflecting the
-material nature of these domains. Moreover, Gugushvili
-(\citeproc{ref-gugushvili_intergenerational_2016c}{2016a}) finds that
-upward mobility is linked to greater justification of income inequality,
-suggesting that improved social standing reinforces an attributional
-view in which success is seen as the result of individual
-characteristics---thus legitimizing inequality as a fair outcome.
-
-Consequently, considering this theoretical and empirical background, the
-first hypothesis of this research is that:
-
-\(H1\): Experiencing upward (downward) social mobility is positively
-(negatively) associated with greater support for market justice in
-healthcare, pensions, and education.
+\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). In what follows,
+I focus on two mechanisms that are particularly relevant for
+market-based pensions: material self-interest and acculturation into
+meritocratic distributive norms.
+
+How intergenerational social mobility shapes preferences for the
+commodification of pensions? A plausible mechanism runs through changes
+in material self-interest. Social mobility alters individuals' incomes,
+job security and access to economic resources
+(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}), thereby modifying the
+costs and benefits they face in market-based pension systems. Material
+interests can be understood as the pursuit of economic well-being under
+the constraints and trade-offs implied by one's class position
+(\citeproc{ref-wrightClassCountsComparative1997a}{Wright, 1997}). Upward
+mobility typically entails higher and more stable earnings, greater
+capacity to contribute to funded schemes, greater tolerance for
+investment risk, and the ability to purchase supplementary coverage,
+whereas downward mobility reduces income, undermines contribution
+histories, and increases dependence on public or solidarity-based
+pillars (\citeproc{ref-galvez_pensiones_2024}{Gálvez et al., 2024};
+\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019}).
+These diverging pension prospects imply that upwardly and downwardly
+mobile individuals face systematically different stakes in market-based
+pension systems, making material self-interest a first channel through
+which mobility may shape support for pension commodification.
+
+Consistent with this reasoning, existing research indicates that support
+for welfare commodification and pension marketisation is strongly
+structured by material advantage. Individuals with higher income,
+education and occupational class are more likely to endorse market-based
+distributive principles and to view private provision as a way to secure
+or enhance relative advantage
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
+\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
+2016}). In the pension field, institutional design sharpens these
+incentives. Busemeyer \& Iversen
+(\citeproc{ref-busemeyer_welfare_2020}{2020}) show that in countries
+where private alternatives to public pensions exist, support for public
+pension spending among upper-income groups erodes as they exit public
+schemes into private plans whose contributions and benefits closely
+track their income and risk. In Latin America, Kerner
+(\citeproc{ref-kerner_pension_2020}{2020}) finds that higher income and
+higher perceived financial returns from pension funds are associated
+with preferences for a stronger market role in pensions, suggesting that
+favourable returns are interpreted as evidence that pension
+neoliberalism ``works''. Evidence from Chile points in the same
+direction: higher-income and university-educated respondents express
+stronger support for market allocation in healthcare, education and
+pensions (\citeproc{ref-castillo_socialization_2024}{Castillo et al.,
+2024}; \citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}), and
+economic elites---especially those socialised in highly prestigious
+schools---hold more favourable attitudes towards inequality that can be
+traced to their material interests against redistribution
+(\citeproc{ref-carranza_what_2024}{Carranza et al., 2024}). Systems that
+link benefits to individual contributions thus allow better-off groups
+to capitalise on their financial independence and minimise reliance on
+public support, implying that mobility into or out of these advantaged
+positions should alter individuals' stakes in pension commodification.
+
+Research on redistribution and inequality offers additional, indirect
+support for this self-interest mechanism. Individuals with pessimistic
+mobility expectations---especially those anticipating downward
+movement---display stronger support for redistributive policies
+(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018}).
+Studies tracking realised mobility show that, compared to immobility,
+upward moves are associated with more economically conservative, less
+redistributive attitudes, whereas downward moves strengthen demand for
+state intervention and redistribution
+(\citeproc{ref-ares_changing_2020}{Ares, 2020};
+\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019};
+\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022};
+\citeproc{ref-rodriguez_intergenerational_2024}{Rodriguez \&
+Matilla-Garcia, 2024}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
+2011}). Over the life course, cumulative exposure to higher-class
+positions---whether by staying or moving up---is linked to more
+market-liberal economic values, while sustained exposure to lower-class
+positions---through stability or downward mobility---reinforces
+pro-redistributive orientations
+(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023},
+\citeproc{ref-helgason_class_2025}{2025}). Domain-specific evidence also
+aligns with changing material stakes: Gugushvili
+(\citeproc{ref-gugushvili_subjective_2017}{2017}) finds that upward
+mobility is associated with lower support for government spending on
+housing and pensions, whereas downward mobility reduces support for
+healthcare and education but increases support for housing and pensions,
+reflecting the material nature of these domains. Beyond redistribution,
+upward mobility fosters more individualistic attributions of poverty and
+wealth and greater legitimation of income inequality
+(\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
+\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b},
+\citeproc{ref-gugushvili_intergenerational_2016c}{2016a}), and societies
+with higher economic mobility display more tolerance of inequality
+(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). Taken
+together, this evidence suggests that social mobility realigns attitudes
+with changing material stakes and distributive interests. Although the
+present study cannot directly test the underlying mechanism, it treats
+myopic self-interest as the central interpretive framework for the
+expected effects of intergenerational mobility on support for pension
+commodification.
+
+Consequently, the main hypothesis of this research is:
+
+\(H_1\): Compared to immobility, intergenerational upward mobility is
+associated with stronger, and downward mobility with weaker, support for
+the commodification of pensions.
+
+While material self-interest is the main lens, mobility also reshapes
+norms and interpretive frames. Mobile individuals move across social
+milieus and networks and are exposed to new standards and beliefs about
+fairness and deservingness
+(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}).
+Acculturation accounts hold that people gradually adopt the attitudes of
+their destination class
+(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}). Mobility can thus reshape how individuals
+interpret success, failure and inequality, with upward moves fostering
+more individualistic and meritocratic understandings and downward moves
+encouraging structural accounts of disadvantage
+(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015};
+\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}) Applied to pensions,
+those who rise may come to see market-based schemes as fair systems that
+reward effort and foresight. In contrast, those who fall may view
+commodified pensions as unjust and insufficiently protective. Therefore,
+shifts in meritocratic norms can be a second channel through which the
+mobility effect operates, serving as a complementary acculturation
+mechanism that may influence support for pension commodification, as
+examined in the next section.
 
@@
 
-Meritocracy constitutes a central ideological framework for legitimizing
-different types of social inequality, for instance through market
-justice beliefs. Rooted in the belief that rewards and positions should
-be allocated based on individual effort and talent, meritocracy operates
-as a normative ideal and a descriptive belief about how society
-functions. As initially conceptualized by Michael Young
-(\citeproc{ref-young_rise_1958}{1958}), the term was meant to critique a
-system in which merit-based stratification becomes a new form of
-inequality. However, over time, meritocracy has been widely supported in
-many societies as a fair and desirable principle of distribution,
-particularly within liberal democracies and market-oriented economies
-(\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
-\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
-perspective, the belief in meritocracy is more than a cognitive
-assessment; it reflects a moral lens through which individuals interpret
-inequality. People who believe that success results from hard work and
-talent are more likely to view social and economic disparities as
-legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
-2019}). Conversely, if they see outcomes as driven by luck, social
-origin, or systemic barriers, inequality is more likely to be perceived
-as unjust. This distinction becomes crucial in societies with persistent
-structural inequality, where public narratives often emphasize personal
-responsibility and merit while overlooking entrenched disadvantages.
-
-I adopt a multidimensional perspective on meritocracy, distinguishing
-between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through
-which individuals justify inequality
-(\citeproc{ref-young_rise_1958}{Young, 1958}). Effort-based meritocracy
-emphasizes hard work and perseverance as the basis for social rewards,
-aligning closely with cultural narratives of personal responsibility. A
-talent-based meritocracy, by contrast, emphasizes intelligence and
-innate abilities, which are often perceived as less malleable and more
-unequally distributed. Both dimensions have been shown to correlate with
-acceptance of inequality, but they may carry distinct implications for
-how inequality is justified in specific domains
+Meritocracy refers to a distributive system in which individual
+merit---typically defined as effort and talent---is the primary
+criterion for allocating resources and rewards, rather than social
+origins or inherited privilege (\citeproc{ref-sen_merit_2000}{Sen,
+2000}; \citeproc{ref-young_rise_1958}{Young, 1958}). Originally coined
+by Young in a dystopian critique of a society where power and status are
+justified by achievement and mobility only driven by ``merit'', the term
+has been re-appropriated as a positive ideal of fairness, especially in
+liberal and market-oriented societies
+(\citeproc{ref-dubet_repensar_2011}{Dubet, 2011};
+\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
+\citeproc{ref-vandewerfhorst_meritocracy_2024}{Van De Werfhorst, 2024}).
+From a sociological perspective, belief in meritocracy is not just a
+cognitive judgement; it constitutes a moral lens through which
+individuals interpret social and economic disparities
 (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
-The relevance of this distinction is supported by recent studies, which
-show that individuals respond differently to these dimensions. For
-instance, perceptions that effort is rewarded in society are more
-strongly associated with positive evaluations of fairness and acceptance
-of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}; \citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al., 2015};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}). This may be because
-effort is seen as a controllable and morally virtuous trait, whereas
-talent is often perceived as a natural advantage. Consequently,
-effort-based meritocracy is likely more potent in legitimizing
-inequality, particularly in neoliberal contexts.
-
-These dimensions of meritocracy reflect how respondents perceive
-society's distributive logic, regardless of whether they endorse
-meritocratic principles. This distinction aligns with recent findings
-indicating that individuals distinguish between how merit is perceived
-in society and how it should ideally operate, which in turn shapes their
-preferences for redistribution and justice
-(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
-2025}). Meritocratic beliefs serve as symbolic justifications for
-unequal outcomes, particularly when access is stratified by income or
-social opportunity. Prior studies in Chile have shown that individuals
-who perceive higher levels of meritocracy tend to express stronger
-support for unequal distributions that reflect market outcomes in
-healthcare, education and pensions
-(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
-2024}).
-
-In addition to influencing individual attitudes toward inequality,
-meritocratic beliefs can contribute to social division and the
-stigmatization of disadvantaged groups. Recent research has demonstrated
-that exposure to meritocratic narratives can reinforce the belief that
-poverty results from individual failings rather than systemic
-conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
-et al., 2023}). This reinforces negative stereotypes and reduces empathy
-toward individuals from lower socioeconomic backgrounds. Moreover,
-Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
-that meritocratic narratives can serve as feedback mechanisms that shape
-public opinion and well-being by framing individuals' understanding of
-welfare outcomes as deserved or undeserved within existing institutional
-structures. This psychological mechanism highlights the normative power
-of meritocracy in stabilizing unequal systems by shaping political
-attitudes and personal perceptions of success and failure.
-
-Importantly, recent research has explored how meritocratic beliefs
-interact with experiences of intergenerational mobility to shape
-distributive attitudes. The belief that one's success is earned can lead
-upwardly mobile individuals to internalize meritocratic narratives and
-justify existing inequalities, reinforcing support for market justice
-(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
-\citeproc{ref-molina_its_2019}{Molina et al., 2019}). Conversely,
-downwardly mobile individuals who maintain strong meritocratic beliefs
-may interpret their status as a personal failure, reducing their support
-for redistribution (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017}).
-At the macro level, Shariff et al.
-(\citeproc{ref-shariff_income_2016}{2016}) show that higher perceived
-mobility increases tolerance for inequality, suggesting that meritocracy
-and mobility are mutually reinforcing.
-
-Taken together, this literature supports the idea that meritocratic
-beliefs can moderate the relationship between mobility and market
-justice preferences. Individuals who experience mobility---especially
-upward---may draw on meritocratic narratives to legitimize both their
-own status and broader inequalities, thereby strengthening their support
-for market-based distribution. Accordingly, the second hypothesis of
-this study is:
-
-\(H2\): The positive (negative) relationship between upward (downward)
-social mobility and support for market justice in healthcare, pensions,
-and education is moderated by meritocratic beliefs; specifically, this
-association is stronger (weaker) among individuals with higher
-perceptions of meritocracy.
-
-\subsection{The Chilean context}\label{the-chilean-context}
-
-Chile offers a compelling case for examining individual preferences for
-market justice in the access to core social services. Since the early
-1990s, the country experienced a long period of sustained economic
-growth, with per capita GDP increasing by nearly 4\% annually until the
-mid-2010s (\citeproc{ref-llorca-jana_historia_2021}{Llorca-Jaña \&
-Miller, 2021}). This growth brought notable improvements in living
-standards, including a strong decline in income poverty---from 38.6\% in
-1990 to 6.5\% in 2022 (\citeproc{ref-mideplan_evolucion_2017}{MIDEPLAN,
-2017}). However, this success story has been tempered by growing
-concerns over the unequal distribution of its benefits. In recent years,
-economic growth has slowed---exacerbated by the global impact of the
-COVID-19 pandemic---and structural inequalities have become more visible
-(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). Despite its
-economic achievements, Chile remains one of the most unequal countries
-in Latin America and the OCDE. The poorest 50\% of the population
-captures just 10\% of total income and holds negative net wealth, while
-the wealthiest 1\% receives nearly 27\% of all income and controls
-almost half (49.6\%) of the country's wealth
-(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a figure that
-has seen little change since the 1990s
-(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). These stark
-disparities extend beyond income, reflecting deep inequalities in access
-to essential social services (\citeproc{ref-pnud_desiguales_2017}{PNUD,
-2017}).
-
-A significant share of Chile's inequality stems from neoliberal reforms
-that privatized and commodified key domains of social reproduction
-(\citeproc{ref-arrizabalo_milagro_1995}{Arrizabalo, 1995};
-\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Introduced during the
-military dictatorship (1973--1989) and expanded under democratic
-governments, these reforms embedded market logic into public services
-through concessions, subsidies, demand-side vouchers, and pro-private
-regulatory frameworks (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
-\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This
-``crowded-out'' welfare model benefits higher-income groups, leaving
-lower-income individuals to rely on limited public options. Scholars
-argue that this neoliberal shift repurposed the state to create niches
-of capitalist accumulation -from water to education-, giving rise to a
-model of ``public-service capitalism'' heavily reliant on state funding
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In healthcare, around
-79\% of the population is covered by the public insurer FONASA, while
-15\% are enrolled in private ISAPREs, which offer faster and
-higher-quality services to wealthier groups
-(\citeproc{ref-mideplan_estadisticas_2024}{MIDEPLAN, 2024}). The pension
-system is based on individual capitalization, with mandatory
-contributions managed by private administrators investing in the
-financial market, currently involving 11 million contributors through
-private fund administrators (AFPs), yet 27\% of the labor force remains
-outside the system due to informality
-(\citeproc{ref-superintendenciadepensiones_estadisticas_2024}{Superintendencia
-de Pensiones, 2024}). Education is similarly stratified: only 30.6\% of
-students attend fully public schools, while 54\% are in voucher-funded
-private institutions and 9.3\% in fully private schools that largely
-serve affluent families
-(\citeproc{ref-ministeriodeeducacion_resumen_2023}{Ministerio de
-Educación, 2023}). This arrangement has created a structurally segmented
-system, where access and quality of services depend on ability to pay,
-reinforced by state subsidies to private providers
-(\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).
-
-Over recent decades, Chile has undergone a significant transformation in
-its class structure, shaped by sustained economic growth, the expansion
-of the service sector, the massification of education, and major changes
-in the role of the state and social policy
-(\citeproc{ref-ruiz_chilenos_2014}{Ruiz \& Boccardo, 2014}). These
-changes contributed to the growth of the middle sectors, a phenomena
-labelled as ``mesocratrization'' of society
-(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013}).
-However, this expansion did not reduce structural inequalities nor
-facilitate upward mobility into elite positions
-(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
-\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}). While Chile shows high absolute mobility, most flows occur
-between adjacent classes---particularly from working classes into lower
-service-class positions---while access to upper classes remains limited,
-indicating strong elite closure and reproduction
-(\citeproc{ref-espinoza_estratificacion_2013}{Espinoza et al., 2013};
-\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014};
-\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
-2021}; \citeproc{ref-perez-ahumada_clases_2019}{Pérez-Ahumada, 2019};
-\citeproc{ref-torche_unequal_2005}{Torche, 2005}). Though often
-described as ``unequal but fluid'', recent evidence points to growing
-rigidity in relative mobility, especially after the economic slowdown,
-increased household debt, labor market precarity, and the COVID-19
-crisis (\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}).
-However, the majority of the population identifies themselves as members
-of the middle class (\citeproc{ref-castillo_todos_2013}{Castillo et al.,
-2013}). These dynamics have fueled a legitimacy crisis: meritocracy is
-increasingly questioned, and many perceive that advancement depends more
-on informal networks than on effort or earned credentials
-(\citeproc{ref-barozet_clases_2021}{Barozet et al., 2021}). The
-combination of constrained mobility, persistent inequality, and
-concentrated opportunity has eroded expectations of upward mobility and
-intensified social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD,
-2017}).
-
-Despite recurrent social unrest, Chile presents a paradoxical
-coexistence between strong conflict over inequality and widespread
-public legitimation of it. The October 2019 ``social outburst'', a
-period of mass protests and severe political repression throughout the
-country, crystallized discontent over the privatization and
-commodification of social services, alongside a crisis of political
-legitimacy (\citeproc{ref-somma_no_2021}{Somma et al., 2021}). A survey
-during the protests identified pensions, healthcare, and education as
-the top demands
-(\citeproc{ref-nucleodesociologiacontingente_informe_2020}{Núcleo de
-Sociología Contingente, 2020}), reflecting dissatisfaction with Chile's
-stratified welfare regime. Yet, this unrest coexists with strong
-meritocratic beliefs and high tolerance for income inequality. Castillo
-(\citeproc{ref-castillo_legitimacy_2011}{2011}) shows that wage gaps
-between occupations are consistently justified, and the larger the
-perceived gap, the greater its legitimation, an effect amplified by
-Chile's structural inequality. Moreover, perceptions of unfairness due
-to non-meritocratic factors often reinforce belief in effort and talent
-as legitimate bases for success
-(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}),
-which are associated with lower perceived injustice
+The conviction that economic inequalities are justified because they
+reflect differential merit has been identified as a key mechanism behind
+the persistence of inequality, since it recasts structural advantages
+and disadvantages as outcomes of individual performance and encourages
+``winners'' to see their position as earned and ``losers'' to
+internalise blame rather than question the underlying structure of
+opportunity (\citeproc{ref-mijs_unfulfillable_2016}{Mijs, 2016};
+\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020};
+\citeproc{ref-wilson_role_2003}{Wilson, 2003}). Understanding how these
+beliefs in meritocracy are structured is therefore crucial for analysing
+attitudes toward inequality.
+
+Recent research has emphasised the need to decompose the term
+``meritocratic beliefs'' and to distinguish between meritocratic
+preferences and meritocratic perceptions
+(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023};
+\citeproc{ref-zhu_meritocratic_2025}{Li Zhu, 2025}). Preferences refer
+to normative ideas about how rewards should be allocated---whether
+people think effort and ability ought to determine life
+chances---whereas perceptions capture evaluations of how meritocracy
+actually operates in society, that is, whether people believe that
+existing inequalities reflect merit-based processes
+(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This
+distinction is crucial for understanding how people interpret inequality
+in stratified societies because it clarifies the (mis)alignment between
+normative support for meritocracy and evaluations about its actual
+implementation (\citeproc{ref-castillo_multidimensional_2023}{Castillo
+et al., 2023}; \citeproc{ref-lindner_does_2024}{Lindner et al., 2024};
+\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
+\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022};
+\citeproc{ref-newman_economic_2023}{Newman, 2023}). In this study, I
+focus on meritocratic perceptions of effort and talent as key dimensions
+of broader meritocratic beliefs, given their central role in shaping
+attitudes toward inequality and welfare.
+
+Several empirical studies have examined the social foundations and
+consequences of meritocratic perceptions. Individuals with higher levels
+of education, income and occupational prestige are more likely to
+endorse merit-based explanations for social outcomes
+(\citeproc{ref-duru-bellat_who_2012}{Duru-Bellat \& Tenret, 2012};
+\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
+2018}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019}). These perceptions
+function as a normative framework that legitimises several unequal
+outcomes, especially when access to social goods is stratified by income
+or social background. Individuals who see their society as meritocratic
+tend to show lower support for redistribution
+(\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023};
+\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
+2025}), greater legitimisation of class inequality
+(\citeproc{ref-darnon_where_2018}{Darnon et al., 2018}), increased
+tolerance of income gaps (\citeproc{ref-batruch_belief_2023}{Batruch et
+al., 2023}), more support for system-justifying ideologies
+(\citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al., 2015}), and
+even lower perceived levels of economic inequality itself
 (\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
-2019}). Mac-Clure et al.
-(\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}) find
-that low and lower-middle status individuals are less likely than
-upper-status groups to view educational income differences as unfair.
-Qualitative research further reveals the internalization of the
-enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
-understood as a way in which the subjectivization process arises in line
-with positioning the self in the market, generating its own market value
-and subjecting it to competition. Canales et al.
-(\citeproc{ref-canalesceron_sujeto_2021}{2021}) describes how families
-navigate school choice as investment decisions, while Panes
-(\citeproc{ref-panes_criticas_2020}{2020}) finds that many workers frame
-pension contributions as individual investments aimed at maximizing
-future returns. These representations suggest that market-based moral
-orientations are deeply embedded in Chilean subjectivities, offering
-fertile ground to study how support for market justice varies across
-welfare domains.
-
-Although still limited, recent research on market justice preferences in
-Chile indicates that support for these beliefs has increased in recent
-years across domains such as healthcare, pensions, and education, and
-that both objective and subjective dimensions of inequality shape them.
-Castillo et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
-that agreement with the notion that it is fair for higher-income
-individuals to access better services was relatively low in 2016, but
-rose significantly by 2023---especially in the pension domain. Otero and
-Mendoza (\citeproc{ref-otero_power_2024}{2024}) provide evidence that
-social class, measured using the EGP scheme, influences these
-preferences: individuals in lower or subordinate class positions are
-less supportive of market-based access to welfare services compared to
-those in higher or privileged positions. Additionally, greater diversity
-in social class networks is associated with lower support for market
-justice. This is particularly relevant in the Chilean context, where one
-of the main determinants of class-diverse networks is intergenerational
-social mobility, which exposes individuals to relationships with people
-from different class backgrounds (\citeproc{ref-otero_lives_2022}{Otero
-et al., 2022}). Regarding subjective factors, Castillo et al.
-(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
-2024}) find that individuals who endorse meritocratic beliefs---such as
-the idea that effort and talent are rewarded in Chile---are more likely
-to support market justice across welfare domains.
-
-\begin{table}[htbp]
+2019}). This evidence provides a basis for extending the study of the
+attitudinal consequences of meritocracy to preferences for market-based
+welfare.
+
+Preferences for the commodification of social welfare services are
+anchored in normative frames. Studies show that support for market-based
+welfare is higher among individuals who endorse meritocratic and liberal
+principles---that benefits should depend on individual effort and
+contributions---and among those with economically conservative
+orientations, whereas egalitarian views reduce support for market
+allocation (\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo,
+2013}; \citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
+\citeproc{ref-quadagno_has_2012}{Quadagno \& Pederson, 2012}). In the
+pension field, factorial evidence from Chile indicates that merit-based
+criteria such as educational attainment and years in the labour force
+weigh more heavily than need-based attributes in judgments of just
+pension levels, with respondents willing to accept very low pensions for
+those perceived as low achievers
+(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al.,
+2019}). Recently, Castillo, Laffert, et al.
+(\citeproc{ref-castillo_perceptions_2025}{2025}) show that agreement
+with the notion that it is fair for higher-income individuals to access
+better welfare services was relatively low in 2016 but rose
+significantly by 2023, especially in the pension domain. They also find
+that individuals who endorse more meritocratic perceptions--specifically
+the perception that effort is rewarded-- are more likely to support
+market-based access for healthcare, education and pensions. All in all,
+meritocratic perceptions are a powerful source of inequality
+legitimation, anchoring who is seen as deserving to be better or worse
+off, and are strongly linked to how individuals understand unequal life
+chances and movements up and down the social hierarchy.
+
+Social mobility and meritocratic perceptions are closely intertwined in
+the acculturation mechanism, although the direction of this link can
+vary across trajectories and contexts. A large body of work consistently
+shows that upwardly mobile individuals tend to interpret their
+trajectory through a meritocratic lens, reading their success as
+evidence that effort and ability are rewarded and reinforcing
+individualistic attributions of poverty and wealth
+(\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
+\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
+\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022};
+\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), even in a
+highly unequal and low mobility setting were the narratives of
+individual success reinforces (\citeproc{ref-deng_its_2025}{Deng \&
+Wang, 2025}). Studies of economic elites in Chile similarly report
+strong endorsement of meritocracy, with respondents attributing their
+ascent mainly to talent and business or leadership skills
+(\citeproc{ref-atria_economic_2020}{Atria et al., 2020}), in line with
+self-serving attribution patterns
+(\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981};
+\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}).Evidence
+on downward or blocked mobility is more ambivalent. Exposure to
+low-mobility contexts and experiences of downward mobility are
+associated with weaker beliefs that effort is rewarded, lower just-world
+and opportunity beliefs, and stronger structural attributions for
+success and failure (\citeproc{ref-davidai_why_2018}{Davidai, 2018};
+\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
+\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
+\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022}). Yet, under strong
+meritocratic convictions, status loss can also be internalised as
+personal failure, sustaining system-justifying attitudes and the
+legitimation of existing hierarchies as a psychological defence strategy
+(\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
+\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
+\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In sum, upward mobility
+is generally associated with stronger meritocratic perceptions, whereas
+downward mobility is more closely linked to structural explanations of
+inequality, without necessarily eroding meritocratic convictions.
+
+The ways in which social mobility trajectories shape interpretations of
+meritocracy can determine whether mobility translates into stronger or
+weaker support for market-based pension welfare. Following Mau's
+(\citeproc{ref-mau_inequality_2015}{2015}) account of the ``majority
+class'', upward mobility is expected to strengthen market-based welfare
+not only because material interests change, but because those who rise
+come to read their story as one of individual effort and talent
+rewarded. In this view, markets are seen as fair arenas that properly
+allocate pensions and other welfare benefits according to merit, so
+upwardly mobile individuals with strong meritocratic convictions should
+be especially prone to endorse commodified pensions. In contrast,
+downward mobility typically fosters de-commodifying demands, yet under
+strong meritocratic beliefs, status loss may be read as individually
+deserved, weakening---or even reversing---its usual de-commodifying
+pattern, as market outcomes are still viewed as broadly fair regardless
+of direction. Empirical research on redistribution and attitudes toward
+inequality supports this mechanism. Survey studies suggest that
+subjective social mobility shapes support for redistribution and pension
+spending through self-serving bias
+(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
+\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}), although recent
+work indicates that this mechanism is evident only under an explicit
+experimental design (\citeproc{ref-molina_its_2019}{Molina et al.,
+2019}). In such designs, exposure to upward-mobility information or
+self-made narratives reduces support for redistribution, increases
+tolerance of inequality and system justification, and even legitimises
+exploitative work conditions, whereas exposure to low or downward
+mobility produces the opposite pattern; crucially, these effects are
+partly mediated by stronger meritocratic beliefs
+(\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
+\citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al., 2025};
+\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In this
+framework, meritocratic beliefs condition the attitudinal consequences
+of mobility by amplifying the pro-market effects of upward mobility and
+muting the critical potential of downward mobility; thus, I treat
+meritocratic perceptions as a moderating mechanism linking
+intergenerational mobility to preferences for pension commodification.
+
+Accordingly, the second set of hypotheses of this study is:
+
+\(H_{2a}\): Under high meritocracy, upward mobility should generate a
+stronger increase in support for market-based pension access than under
+low meritocracy.
+
+\(H_{2b}\): Under low meritocracy, downward mobility should moderately
+reduce support for market-based pension access, whereas under high
+meritocracy, this negative effect should be attenuated toward neutral or
+slightly positive attitudes.
+
+A summary of the hypotheses and their expected directional patterns is
+presented in Table~\ref{tbl-hipo}.
+
+\begin{table}
+
+\caption{\label{tbl-hipo}Overview of hypotheses with expected directional patterns }
+
+\centering{
+
 \centering
-\caption{Overview of hypotheses with expected directional patterns}
+
 \begin{tabular}{p{3.2cm} p{3.2cm} p{4cm} p{5cm}}
@@
 \end{tabular}
-\end{table}
+
+}
+
+\end{table}%
+
+\subsection{The Chilean context}\label{the-chilean-context}
+
+Chile offers a compelling case for examining preferences for
+market-based pension welfare. Despite sustained economic growth, it
+remains one of the most unequal countries in Latin America and the OECD.
+The poorest 50\% of the population captures just 10\% of total income
+and holds negative net wealth, while the wealthiest 1\% receives nearly
+27\% of all income and controls almost half of the country's wealth
+(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}), a pattern
+that has changed little since the 1990s
+(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Research on
+intergenerational mobility shows that Chile combines high absolute
+mobility with mostly short-range moves---mainly from working classes
+into lower service-class positions---while access to upper and elite
+classes remains tightly restricted, and that despite the expansion of
+higher education, relative mobility remains rigid and intergenerational
+inequalities persist (\citeproc{ref-espinoza_movilidad_2014}{Espinoza \&
+Núñez, 2014}; \citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán
+\& Fachelli, 2021}; \citeproc{ref-salgado_uplifting_2025}{Salgado et
+al., 2025}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}).
+Crucially, a significant share of Chile's inequality is rooted in
+neoliberal reforms that privatised and commodified key social services,
+stratifying and segmenting access to high-quality education, healthcare
+and housing between classes (\citeproc{ref-ferre_welfare_2023}{Ferre,
+2023}; \citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}).
+
+Introduced under the military dictatorship (1973--1989) and expanded by
+democratic governments, neoliberal reforms in Chile embedded market
+logic into public services through concessions, subsidies, vouchers and
+pro-private regulation (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This
+``crowded-out'' welfare model benefits higher-income groups, leaving
+lower-income individuals to rely on limited public options. Scholars
+argue that this neoliberal shift repurposed the state to create niches
+of capitalist accumulation---from water to education---giving rise to a
+model of ``public-service capitalism'' heavily reliant on state funding
+(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In this setting,
+class trajectories unfold under conditions of high inequality and
+market-based provision in core welfare domains, such as pensions, making
+Chile a beneficial context for studying support for pension
+commodification.
+
+In pensions, Chile was the first country to fully replace a public
+pay-as-you-go social insurance system with mandatory individual
+capitalisation: in 1981, the tripartite cajas de previsión were replaced
+by a fully funded defined-contribution scheme managed by private Pension
+Fund Administrators (AFP) (\citeproc{ref-arenas_sistemas_2019}{Arenas,
+2019}; \citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
+Framed by its principal architect, José Piñera, as ``the mother of all
+battles'', the reform was a way to roll back the redistributive role of
+the state, shift ``freedom of choice'' to individuals and create a
+``society of worker-capitalists''
+(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
+\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). Dependent workers
+contribute 10\% of their wages to individual accounts, plus an
+administrative fee, while employers contribute nothing; AFPs invest
+these mandatory savings in domestic and international financial markets,
+and since the 2000s affiliates choose (or are defaulted into)
+``multifunds'' (A--E) with different risk profiles, directly exposing
+pensions to market volatility
+(\citeproc{ref-barriga_pensiones_2024}{Barriga \& Kremerman, 2024};
+\citeproc{ref-hyde_chiles_2015}{Hyde \& Borzutzky, 2015}). After more
+than four decades, this architecture has produced an open pension
+crisis: for most retirees, AFP-financed pensions fall below the minimum
+wage and are clearly insufficient to avoid old-age poverty, with
+replacement rates around 30\% even among cohorts with long contribution
+histories (\citeproc{ref-galvez_pensiones_2024}{Gálvez et al., 2024};
+\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019}). To
+prevent destitution, the state has been forced to re-enter through
+tax-funded solidarity pillars ---the Nuevo Pilar Solidario (2008) and
+the Pensión Garantizada Universal (PGU, 2022)--- that partially repair
+these failures (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-solimano_rise_2021}{Solimano, 2021}) by topping up low or
+non-existent contributory pensions for the poorest and lower-middle
+older adults, so that for many low earners these tax-funded
+contributions are the main component of their total pension income
+(\citeproc{ref-galvez_pensiones_2024}{Gálvez et al., 2024}). Meanwhile,
+the AFP industry has remained highly concentrated
+(\citeproc{ref-hyde_chiles_2015}{Hyde \& Borzutzky, 2015}), posting very
+high returns on equity (around US\$ 613 million in 2024) and fostering
+powerful national economic actors
+(\citeproc{ref-ruiz_formacion_2020}{Ruiz, 2020}), leaving Chile with a
+structurally fragile, segmented and regressive pension system that
+combines poor pensions, heavy reliance on public subsidies and high,
+sustained profits for private financial intermediaries. Chile's
+combination of constrained mobility, persistent inequality and deep
+privatisation has eroded expectations of upward mobility and fuelled
+social discontent (\citeproc{ref-pnud_desiguales_2017}{PNUD, 2017}),
+with pensions at the centre of this conflict.
+
+Despite recurrent social unrest, Chile exhibits a paradoxical
+coexistence of intense conflict over inequality and its widespread
+legitimation, vividly illustrated by pensions. The October 2019 ``social
+outburst'', a period of mass protests and severe political repression
+throughout the country, crystallised discontent over the privatisation
+and commodification of social services, alongside a crisis of political
+legitimacy (\citeproc{ref-somma_no_2021}{Somma et al., 2021}). One of
+the main conflicts in this period centred on grievances over low
+pensions and the profits of private pension funds, building on earlier
+mobilisation by the No+AFP movement since 2016; a survey during the 2019
+protests identified pensions as the top public demand
+(\citeproc{ref-nudesoc_informe_2020}{NUDESOC, 2020}). This conflict
+helped place comprehensive pension reform on the political agenda after
+2022. However, contestation coexists with enduring market-conforming
+orientations. Research documents strong meritocratic beliefs and high
+tolerance for inequality
+(\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
+2019}; \citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}),
+and the COVID-19 emergency withdrawals of pension savings gave rise to
+the slogan ``with my money, no'', expressing opposition both to greater
+pay-as-you-go elements and to continued AFP control
+(\citeproc{ref-galvez_con_2023}{Gálvez, 2023}). This stance reflects the
+internalisation of the ``enterprising self''
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}): families navigate
+school choice as investment strategies
+(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}),
+and workers denounce the pension system as illegitimate in its design
+and outcomes while simultaneously treating their contributions as
+personal investments aimed at maximising returns and asserting that they
+themselves ---rather than AFPs, the state or ``society''--- are the
+legitimate owners of their pension funds
+(\citeproc{ref-panes_criticas_2020}{Panes, 2020}). These ambivalent
+representations indicate that market-based moral orientations are deeply
+embedded, making Chile a fertile setting to study how support for
+pension commodification varies across social groups and meritocratic
+beliefs.
 
@@
 
-\subsection{Descriptive statistics}\label{descriptive-statistics}
+\subsection{Descriptive analysis}\label{descriptive-analysis}
 
@@
 mobility treatment at low and high meritocratic beliefs. Across
-trajectories, mobility effects for low-merit respondents are uniformly
-small and imprecise, hovering around zero for both upward and downward
-moves. Among high-merit respondents, the pattern aligns only partially
-with the proposed mechanism: upward mobility from middle to high status
-shows a clear and statistically significant pro-commodification effect
-(\(\beta\) = 0.22, 95\% CI {[}0.05, 0.38{]}), whereas all other
+trajectories, mobility effects among low-merit respondents are uniformly
+small, imprecise, and centered around zero for both upward and downward
+moves. Among high-merit respondents, only upward mobility from middle to
+high status shows a clear pro-commodification pattern, with a positive
+and statistically significant effect on MJP for pensions
+(\(\beta = 0.22\), 95\% CI {[}0.05, 0.38{]}). All other
 trajectories---Low→Middle, Low→High, Middle→Low, High→Middle, and
-High→Low---remain statistically indistinguishable from zero or display
-negative estimates that do not systematically weaken under high merit.
-Formal interaction tests also fail to detect strong heterogeneity by
-meritocratic beliefs (see \hyperref[anexo]{Supplementary Material} for
-interaction models). Taken together, the results provide only limited
-support for the mechanism hypothesis: meritocracy appears to amplify the
-pro-market consequences of upward mobility in one specific trajectory,
-but there is no general or consistent evidence that meritocratic beliefs
-constitute a stable channel through which intergenerational mobility
-shapes preferences for pension commodification.
+High→Low---remain statistically indistinguishable from zero.
+Nonetheless, the point estimates broadly align with the theoretical
+expectations for upward moves: upwardly mobile individuals with stronger
+meritocratic beliefs tend to display higher support for pension
+commodification than their low-merit counterparts. By contrast, downward
+trajectories exhibit a pattern opposite to the mechanism hypothesis,
+with high-merit respondents generally showing more de-commodifying
+attitudes than those with weaker meritocratic beliefs. However, formal
+interaction tests do not detect robust heterogeneity in mobility effects
+by meritocratic beliefs (see \hyperref[anexo]{Supplementary Material}
+for full interaction models): even for the Middle→High trajectory, the
+simple effect under high meritocracy does not differ significantly from
+its low-merit counterpart. Taken together, these results do not provide
+convincing support for the mechanism hypothesis: meritocratic beliefs do
+not operate as a consistent or general channel through which
+intergenerational mobility shapes preferences for the commodification of
+pensions.
 
@@
 
-\bibitem[\citeproctext]{ref-arrizabalomontoro_debate_2019}
-Arrizabalo Montoro, X., Del Rosal, M., \& Javier Murillo Arroyo, F.
-(2019). The {Debate} on {Pension Systems}: {The Paradigmatic Cases} of
-{Chile} and {Spain}. \emph{The American Journal of Economics and
-Sociology}, \emph{78}(1), 195--223.
-\url{https://doi.org/10.1111/ajes.12262}
-
-\bibitem[\citeproctext]{ref-arrizabalo_milagro_1995}
-Arrizabalo, X. (1995). \emph{{Milagro o quimera: la economía chilena
-durante la dictadura}}. Libros de la Catarata.
+\bibitem[\citeproctext]{ref-arza_pension_2008}
+Arza, C. (2008). Pension {Reform} in {Latin America}: {Distributional
+Principles}, {Inequalities} and {Alternative Policy Options}.
+\emph{Journal of Latin American Studies}, \emph{40}(1), 1--28.
+\url{https://doi.org/10.1017/S0022216X07003616}
+
+\bibitem[\citeproctext]{ref-atria_economic_2020}
+Atria, J., Castillo, J., Maldonado, L., \& Ramirez, S. (2020). Economic
+{Elites}' {Attitudes Toward Meritocracy} in {Chile}: {A Moral Economy
+Perspective}. \emph{American Behavioral Scientist}, \emph{64}(9),
+1219--1241. \url{https://doi.org/10.1177/0002764220941214}
 
@@
 
-\bibitem[\citeproctext]{ref-barozet_clases_2021}
-Barozet, E., Contreras, D., Espinoza, V., Gayo, M., \& Méndez, M. L.
-(2021). \emph{{Clases medias en tiempos de crisis: vulnerabilidad
-persistente, desafíos para la cohesión y un nuevo pacto social en
-Chile}}. Santiago: Comisión Económica para América Latina y el Caribe
-(CEPAL).
+\bibitem[\citeproctext]{ref-barriga_pensiones_2024}
+Barriga, F., \& Kremerman, M. (2024). {Pensiones sin Seguridad Social:
+\textquestiondown Cómo se calcula el monto de las pensiones en Chile?}
+\emph{Fundación Sol}.
 
@@
 
+\bibitem[\citeproctext]{ref-borzutzky_chiles_2016}
+Borzutzky, S., \& Hyde, M. (2016). Chile's private pension system at 35:
+Impact and lessons. \emph{Journal of International and Comparative
+Social Policy}, \emph{32}(1), 57--73.
+\url{https://doi.org/10.1080/21699763.2016.1148623}
+
 \bibitem[\citeproctext]{ref-breen_effects_2024}
@@
 \url{https://doi.org/10.1086/706980}
-
-\bibitem[\citeproctext]{ref-campbell_institutional_2020}
-Campbell, J. L. (2020). \emph{Institutional {Change} and
-{Globalization}}. Princeton University Press.
-\url{https://doi.org/10.2307/j.ctv131bw68}
 
@@
 
+\bibitem[\citeproctext]{ref-carranza_what_2024}
+Carranza, R., Contreras, D., \& Otero, G. (2024). What makes elites more
+or less egalitarian? {Variations} in attitudes towards inequality within
+the economic, political and cultural elites in {Chile}.
+\emph{Socio-Economic Review}, \emph{22}(3), 1141--1167.
+\url{https://doi.org/10.1093/ser/mwae008}
+
 \bibitem[\citeproctext]{ref-castillo_legitimacy_2011}
@@
 
-\bibitem[\citeproctext]{ref-castillo_todos_2013}
-Castillo, J. C., Miranda, D., \& Cabib, I. M. (2013). Todos somos de
-clase media: {Sobre} el estatus social subjetivo en {Chile}. \emph{Latin
-American Research Review}, \emph{48}(1), 155--173.
-\url{https://doi.org/10.1353/lar.2013.0006}
-
 \bibitem[\citeproctext]{ref-castillo_deserving_2019}
@@
 
+\bibitem[\citeproctext]{ref-cavaille_fair_2025}
+Cavaillé, C. (2025). \emph{Fair {Enough}?: {Support} for
+{Redistribution} in the {Age} of {Inequality}} (1st ed.). Cambridge
+University Press. \url{https://doi.org/10.1017/9781009366038}
+
 \bibitem[\citeproctext]{ref-chancel_world_2022}
@@
 
+\bibitem[\citeproctext]{ref-darnon_where_2018}
+Darnon, C., Wiederkehr, V., Dompnier, B., \& Martinot, D. (2018).
+{``{Where} there is a will, there is a way''}: {Belief} in school
+meritocracy and the social-class achievement gap. \emph{British Journal
+of Social Psychology}, \emph{57}(1), 250--262.
+\url{https://doi.org/10.1111/bjso.12214}
+
+\bibitem[\citeproctext]{ref-davidai_why_2018}
+Davidai, S. (2018). Why do {Americans} believe in economic mobility?
+{Economic} inequality, external attributions of wealth and poverty, and
+the belief in economic mobility. \emph{Journal of Experimental Social
+Psychology}, \emph{79}, 138--148.
+\url{https://doi.org/10.1016/j.jesp.2018.07.012}
+
 \bibitem[\citeproctext]{ref-day_movin_2017}
@@
 
-\bibitem[\citeproctext]{ref-espinoza_estratificacion_2013}
-Espinoza, V., Barozet, E., \& Méndez, M. L. (2013). {Estratificación y
-movilidad social bajo un modelo neoliberal: El caso de Chile}.
-\emph{Lavboratorio}, (25).
+\bibitem[\citeproctext]{ref-deng_its_2025}
+Deng, Y., \& Wang, F. (2025). It's my fault, {I} should try harder!
+{The} narratives of self-made upward mobility sustain belief in
+meritocracy in low social mobility context. \emph{British Journal of
+Psychology}, bjop.70015. \url{https://doi.org/10.1111/bjop.70015}
+
+\bibitem[\citeproctext]{ref-dubet_repensar_2011}
+Dubet, F. (2011). \emph{{Repensar la justicia social}} (Sexta Edición).
+Siglo XXI.
+
+\bibitem[\citeproctext]{ref-duru-bellat_who_2012}
+Duru-Bellat, M., \& Tenret, E. (2012). Who's for {Meritocracy}?
+{Individual} and {Contextual Variations} in the {Faith}.
+\emph{Comparative Education Review}, \emph{56}(2), 223--247.
+\url{https://doi.org/10.1086/661290}
+
+\bibitem[\citeproctext]{ref-edlundDemocraticClassStruggle2015a}
+Edlund, J., \& Lindh, A. (2015). The {Democratic Class Struggle
+Revisited}: {The Welfare State}, {Social Cohesion} and {Political
+Conflict}. \emph{Acta Sociologica}, \emph{58}(4), 311--328.
+\url{https://doi.org/10.1177/0001699315610176}
 
@@
 
+\bibitem[\citeproctext]{ref-fourcade_moral_2007}
+Fourcade, M., \& Healy, K. (2007). Moral {Views} of {Market Society}.
+\emph{Annual Review of Sociology}, \emph{33}(1), 285--311.
+\url{https://doi.org/10.1146/annurev.soc.33.040406.131642}
+
 \bibitem[\citeproctext]{ref-galvez_con_2023}
 Gálvez, R. (2023). Con mi plata no\ldots{} alcanza. \emph{CNN Chile}.
+
+\bibitem[\citeproctext]{ref-galvez_pensiones_2024}
+Gálvez, R., Kremerman, M., \& Palacios, V. R. (2024). {Pensiones bajo el
+mínimo: Los montos de las pensiones que paga el sistema de
+capitalización individual en Chile}. \emph{Fundación Sol}.
 
@@
 
+\bibitem[\citeproctext]{ref-garcia-sanchez_perceptions_2018}
+García-Sánchez, E., Willis, G. B., Rodríguez-Bailón, R., Palacio Sañudo,
+J., David Polo, J., \& Rentería Pérez, E. (2018). Perceptions of
+{Economic Inequality} and {Support} for {Redistribution}: {The} role of
+{Existential} and {Utopian Standards}. \emph{Social Justice Research},
+\emph{31}(4), 335--354. \url{https://doi.org/10.1007/s11211-018-0317-6}
+
 \bibitem[\citeproctext]{ref-gelman_regression_2020}
@@
 
-\bibitem[\citeproctext]{ref-gugushvili_heterogeneous_2024}
-Gugushvili, A. (2024). The heterogeneous well-being effects of
-intergenerational mobility perceptions. \emph{Journal of Health
-Psychology}, \emph{29}(2), 99--112.
-\url{https://doi.org/10.1177/13591053231187345}
-
-\bibitem[\citeproctext]{ref-hadjar_does_2015}
-Hadjar, A., \& Samuel, R. (2015). Does upward social mobility increase
-life satisfaction? {A} longitudinal analysis using {British} and {Swiss}
-panel data. \emph{Research in Social Stratification and Mobility},
-\emph{39}, 48--58. \url{https://doi.org/10.1016/j.rssm.2014.12.002}
-
 \bibitem[\citeproctext]{ref-hainmueller_entropy_2012}
@@
 
-\bibitem[\citeproctext]{ref-immergut_theoretical_1998}
-Immergut, E. M. (1998). The {Theoretical Core} of the {New
-Institutionalism}. \emph{Politics \& Society}, \emph{26}(1), 5--34.
-\url{https://doi.org/10.1177/0032329298026001002}
+\bibitem[\citeproctext]{ref-hyde_chiles_2015}
+Hyde, M., \& Borzutzky, S. (2015). Chile's {``{Neoliberal}''}
+{Retirement System}? {Concentration}, {Competition}, and {Economic
+Predation} in {``{Private}''} {Pensions}: {Chile}'s {``{Neoliberal}''}
+{Retirement System}? \emph{Poverty \& Public Policy}, \emph{7}(2),
+123--157. \url{https://doi.org/10.1002/pop4.98}
 
@@
 
-\bibitem[\citeproctext]{ref-jasso_justice_1978}
-Jasso, G. (1978). On the {Justice} of {Earnings}: {A New Specification}
-of the {Justice Evaluation Function}. \emph{American Journal of
-Sociology}, \emph{83}(6), 1398--1419.
-\url{https://doi.org/10.1086/226706}
-
-\bibitem[\citeproctext]{ref-jasso_distributive_2016}
-Jasso, G., Törnblom, K. Y., \& Sabbagh, C. (2016). Distributive
-{Justice}. In C. Sabbagh \& M. Schmitt (Eds.), \emph{Handbook of {Social
-Justice Theory} and {Research}} (pp. 201--218). New York, NY: Springer.
-\url{https://doi.org/10.1007/978-1-4939-3216-0_11}
-
 \bibitem[\citeproctext]{ref-jasso_gender_1999}
@@
 
+\bibitem[\citeproctext]{ref-kulinClassValuesAttitudes2013}
+Kulin, J., \& Svallfors, S. (2013). Class, {Values}, and {Attitudes
+Towards Redistribution}: {A European Comparison}. \emph{European
+Sociological Review}, \emph{29}(2), 155--167.
+\url{https://doi.org/10.1093/esr/jcr046}
+
 \bibitem[\citeproctext]{ref-lane_market_1986}
@@
 
-\bibitem[\citeproctext]{ref-lenski_status_1954}
-Lenski, G. E. (1954). Status {Crystallization}: {A Non-Vertical
-Dimension} of {Social Status}. \emph{American Sociological Review},
-\emph{19}(4), 405. \url{https://doi.org/10.2307/2087459}
-
 \bibitem[\citeproctext]{ref-lindh_public_2015}
@@
 
-\bibitem[\citeproctext]{ref-llorca-jana_historia_2021}
-Llorca-Jaña, M., \& Miller, R. M. D. (2021). \emph{{Historia económica
-de Chile desde la independencia}}. Santiago de Chile: RIL editores.
+\bibitem[\citeproctext]{ref-lindner_does_2024}
+Lindner, T., Mijs, J. J. B., De Koster, W., \& Van Der Waal, J. (2024).
+Does informing citizens about the non-meritocratic nature of inequality
+bolster support for a universal basic income? {Evidence} from a
+population-based survey experiment. \emph{European Societies},
+\emph{26}(3), 575--601.
+\url{https://doi.org/10.1080/14616696.2023.2272263}
 
@@
 
+\bibitem[\citeproctext]{ref-matamoros-lima_social_2025}
+Matamoros-Lima, J., Galdi, S., Moya, M., \& Willis, G. B. (2025). Social
+mobility beliefs and attitudes toward redistribution: {Potential}
+explanatory mechanisms. \emph{Political Psychology}, \emph{46}(4),
+884--902. \url{https://doi.org/10.1111/pops.13042}
+
 \bibitem[\citeproctext]{ref-mau_inequality_2015}
@@
 
-\bibitem[\citeproctext]{ref-mideplan_evolucion_2017}
-MIDEPLAN. (2017). \emph{Evolución de la pobreza 1990-2017}. Ministerio
-de Desarrollo Social y Familia - PNUD.
-
-\bibitem[\citeproctext]{ref-mideplan_estadisticas_2024}
-MIDEPLAN. (2024). \emph{Estadísticas {Encuesta Casen} 2022, sección
-{Salud}}. Observatorio social.
+\bibitem[\citeproctext]{ref-mijs_unfulfillable_2016}
+Mijs, J. (2016). The {Unfulfillable Promise} of {Meritocracy}: {Three
+Lessons} and {Their Implications} for {Justice} in {Education}.
+\emph{Social Justice Research}, \emph{29}(1), 14--34.
+\url{https://doi.org/10.1007/s11211-014-0228-0}
 
@@
 
+\bibitem[\citeproctext]{ref-mijs_how_2022}
+Mijs, J., \& Hoy, C. (2022). How {Information} about {Inequality Impacts
+Belief} in {Meritocracy}: {Evidence} from a {Randomized Survey
+Experiment} in {Australia}, {Indonesia} and {Mexico}. \emph{Social
+Problems}, \emph{69}(1), 91--122.
+\url{https://doi.org/10.1093/socpro/spaa059}
+
 \bibitem[\citeproctext]{ref-miller_selfserving_1975}
@@
 
-\bibitem[\citeproctext]{ref-ministeriodeeducacion_resumen_2023}
-Ministerio de Educación. (2023). \emph{Resumen de matrícula por
-establecimiento educacional}. Datos abiertos Ministerio de Educación.
-
 \bibitem[\citeproctext]{ref-molina_its_2019}
@@
 
-\bibitem[\citeproctext]{ref-nucleodesociologiacontingente_informe_2020}
-Núcleo de Sociología Contingente, {[}NUDESOC{]}. (2020). \emph{Informe
-de resultados oficial {Encuesta Zona Cero}}. Santiago de Chile.
+\bibitem[\citeproctext]{ref-newman_economic_2023}
+Newman, B. J. (2023). Economic {Inequality}, the {Working Poor}, and
+{Belief} in the {American Dream}. \emph{Public Opinion Quarterly},
+\emph{86}(4), 944--954. \url{https://doi.org/10.1093/poq/nfac043}
+
+\bibitem[\citeproctext]{ref-nudesoc_informe_2020}
+NUDESOC, N. de S. C. (2020). \emph{Informe de resultados oficial
+{Encuesta Zona Cero}}. Santiago de Chile.
 
@@
 
-\bibitem[\citeproctext]{ref-otero_lives_2022}
-Otero, G., Volker, B., Rözer, J., \& Mollenhorst, G. (2022). The lives
-of others: {Class} divisions, network segregation, and attachment to
-society in {Chile}. \emph{The British Journal of Sociology},
-\emph{73}(4), 754--785. \url{https://doi.org/10.1111/1468-4446.12966}
-
 \bibitem[\citeproctext]{ref-panes_criticas_2020}
@@
 
-\bibitem[\citeproctext]{ref-perez-ahumada_clases_2019}
-Pérez-Ahumada, P. (2019). {Clases sociales, sectores económicos y
-cambios en la estructura social chilena entre 1992 y 2013}.
-\emph{Revista de la CEPAL}, \emph{2018}(126), 171--192.
-\url{https://doi.org/10.18356/0ce0faed-es}
-
 \bibitem[\citeproctext]{ref-pnud_desiguales_2017}
@@
 
-\bibitem[\citeproctext]{ref-polanyi_great_1975}
-Polanyi, K. (1975). \emph{The great transformation} (Repr). New York,
-NY: Octagon Books.
-
-\bibitem[\citeproctext]{ref-prag_subjective_2021}
-Präg, P., \& Gugushvili, A. (2021). Subjective social mobility and
-health in {Germany}. \emph{European Societies}, \emph{23}(4), 464--486.
-\url{https://doi.org/10.1080/14616696.2021.1887916}
-
-\bibitem[\citeproctext]{ref-ruiz_chilenos_2014}
-Ruiz, C., \& Boccardo, G. (2014). \emph{Los chilenos bajo el
-neoliberalismo: Clases y conflicto social} (Primera edición). Santiago
-de Chile: Ediciones y Publicaciones El Buen Aire.
+\bibitem[\citeproctext]{ref-quadagno_has_2012}
+Quadagno, J., \& Pederson, J. (2012). Has support for {Social Security}
+declined? {Attitudes} toward the public pension scheme in the {USA},
+2000 and 2010. \emph{International Journal of Social Welfare},
+\emph{21}(s1). \url{https://doi.org/10.1111/j.1468-2397.2012.00877.x}
+
+\bibitem[\citeproctext]{ref-rodriguez_intergenerational_2024}
+Rodriguez, I., \& Matilla-Garcia, M. (2024). Intergenerational
+{Mobility} and {Preferences} for {Redistribution}: {Empirical Evidence}
+from {Spain}. SSRN. \url{https://doi.org/10.2139/ssrn.5045536}
+
+\bibitem[\citeproctext]{ref-ruiz_formacion_2020}
+Ruiz, F. (2020). {Formación de clases y conflicto social en el sistema
+previsional chileno: mecanismos de acumulación capitalista y desafíos
+para la democracia}. \emph{REVISTA ENCUENTROS}, \emph{18}(3), 88--99.
+\url{https://doi.org/10.15665/encuent.v18i3.2134}
 
@@
 
+\bibitem[\citeproctext]{ref-sen_merit_2000}
+Sen, A. (2000). Merit and {Justice}. In K. Arrow, S. Bowles, \& S. N.
+Durlauf (Eds.), \emph{Meritocracy and {Economic Inequality}} (pp.
+5--16). Princeton University Press.
+\url{https://doi.org/10.1515/9780691190334-003}
+
 \bibitem[\citeproctext]{ref-shariff_income_2016}
@@
 
-\bibitem[\citeproctext]{ref-sorokin_social_1927}
-Sorokin, P. A. (1927). \emph{Social {Mobility}}. Harper \& Brothers.
-
 \bibitem[\citeproctext]{ref-streeck_how_2016}
@@
 system}. London: Verso.
-
-\bibitem[\citeproctext]{ref-superintendenciadepensiones_estadisticas_2024}
-Superintendencia de Pensiones. (2024). \emph{Estadísticas de {Afiliados}
-en {AFP} en {Sistema} de {Pensiones}. {Abril} 2024}. Superintendencia de
-Pensiones.
 
@@
 
+\bibitem[\citeproctext]{ref-vandewerfhorst_meritocracy_2024}
+Van De Werfhorst, H. G. (2024). Is {Meritocracy Not So Bad After All}?
+{Educational Expansion} and {Intergenerational Mobility} in 40
+{Countries}. \emph{American Sociological Review}, \emph{89}(6),
+1181--1213. \url{https://doi.org/10.1177/00031224241292352}
+
 \bibitem[\citeproctext]{ref-verbic_political_2019}
@@
 
+\bibitem[\citeproctext]{ref-wrightClassCountsComparative1997a}
+Wright, E. O. (1997). \emph{Class {Counts}: {Comparative Studies} in
+{Class Analysis}}. Cambridge University Press.
+
 \bibitem[\citeproctext]{ref-wrightUnderstandingClass2015}
@@
 
+\bibitem[\citeproctext]{ref-zhu_meritocratic_2025}
+Zhu, Li. (2025). Meritocratic beliefs in the {United States}, {Finland},
+and {China}: {A} multidimensional approach using latent class analysis.
+\emph{The British Journal of Sociology}, \emph{76}(1), 153--172.
+\url{https://doi.org/10.1111/1468-4446.13152}
+
+\bibitem[\citeproctext]{ref-zhu_policy_2015}
+Zhu, Ling, \& Lipsmeyer, C. S. (2015). Policy feedback and economic
+risk: The influence of privatization on social policy preferences.
+\emph{Journal of European Public Policy}, \emph{22}(10), 1489--1511.
+\url{https://doi.org/10.1080/13501763.2015.1031159}
+
 \end{CSLReferences}
@@
 
-\subsection{Descriptive statistics}\label{descriptive-statistics-1}
+\subsection{Descriptive statistics}\label{descriptive-statistics}
 

### 2025-11-30 processing/path_merit.R

@@ -1,0 +1,181 @@
+### Low to middle
+
+data_lm <- datos1 %>% 
+  select(idencuesta, 
+         ola, 
+         y = merit_i, 
+         t = moblm, 
+         w = w_low_lm_t, 
+         edad = m0_edad_f, 
+         sexo = sexo_f, 
+         nac = nacionalidad_f,
+         etnia = etnia_f,
+         hogar = hogar_bip_f, 
+         edu0 = educ_orig_f) %>% 
+  filter(!is.na(t), !is.na(w)) %>% 
+  mutate(ola = factor(ola, levels = c("1", "2", "3"),
+                      labels = c("2016", "2018", "2023")),
+         y = as.numeric(y))
+
+mp_lm <-
+  lm_robust(y ~ t + ola,
+            weights = w,
+            se_type = "CR2",
+            clusters = idencuesta,
+            data = data_lm) 
+
+### Low to high
+
+data_lh <- datos1 %>% 
+  select(idencuesta, 
+         ola, 
+         y = merit_i, 
+         t = moblh, 
+         w = w_low_lh_t, 
+         edad = m0_edad_f, 
+         sexo = sexo_f, 
+         nac = nacionalidad_f,
+         etnia = etnia_f,
+         hogar = hogar_bip_f, 
+         edu0 = educ_orig_f) %>% 
+  filter(!is.na(t), !is.na(w)) %>% 
+  mutate(ola = factor(ola, levels = c("1", "2", "3"),
+                      labels = c("2016", "2018", "2023")),
+         y = as.numeric(y))
+
+
+mp_lh <- 
+  lm_robust(y ~ t + ola,
+            weights = w,
+            se_type = "CR2",
+            clusters = idencuesta,
+            data = data_lh) 
+
+### middle to high
+
+data_mh <- datos2 %>% 
+  select(idencuesta, 
+         ola, 
+         y = merit_i, 
+         t = mobmh, 
+         w = w_mid_mh_t, 
+         edad = m0_edad_f, 
+         sexo = sexo_f, 
+         nac = nacionalidad_f,
+         etnia = etnia_f,
+         hogar = hogar_bip_f, 
+         edu0 = educ_orig_f) %>% 
+  filter(!is.na(t), !is.na(w)) %>% 
+  mutate(ola = factor(ola, levels = c("1", "2", "3"),
+                      labels = c("2016", "2018", "2023")),
+         y = as.numeric(y))
+
+
[…]

### 2025-12-01 paper/01-introduction.qmd
Up borrador VF

@@
+---
+bibliography: "../input/bib/tesis-magister.bib"
+csl: "../input/bib/apa6.csl"
+---
+
 # Introduction
@@
 From a moral-economy perspective, this diffusion of market rules has reconfigured the normative order: through policy feedback, welfare institutions embed market-based criteria of “fair” allocation and shape how citizens understand deservingness and the balance between state and market [@svallforsMoralEconomyClass2006a; @koos_moral_2019].
-Against this backdrop, growing scholarship examines *market justice preferences*—the extent to which citizens regard market-based criteria as a fair basis for the allocation of essential services such as healthcare, education and pensions [@castillo_perceptions_2025; @svallfors_political_2007; @lindh_public_2015].
+Against this backdrop, growing scholarship examines *market justice preferences*—the extent to which citizens regard market-based criteria as a fair basis for the allocation of essential services such as healthcare, education, and pensions [@immergut_it_2020; @castillo_changes_2025; @castillo_perceptions_2025; @svallfors_political_2007; @lindh_public_2015].
 These orientations matter because they legitimize unequal outcomes as the product of individual responsibility and promote the conception of basic social services as commodities.
@@
 Pension privatisation lies at the core of Latin America’s trajectory of commodification [@huber_political_2000].
-As many countries expanded mandatory or voluntary individual-capitalisation schemes—most radically in Chile—pensions became organised around contribution-based entitlements and market performance [@arenas_sistemas_2019; @verbic_political_2019].
+As many countries expanded mandatory or voluntary individual-capitalisation schemes—most radically in Chile—pensions became organised around funded contribution-based entitlements and market performance [@arenas_sistemas_2019; @verbic_political_2019].
 This shift reshapes not only how pension institutions allocate old-age risk and tie benefit levels to lifetime labour-market trajectories [@oecd_pensions_2023; @madero-cabib_private_2019], but also transforms the justice principles citizens associate with old-age security: they must judge whether higher benefits should follow market-derived criteria or solidaristic principles [@mau_inequality_2015; @borzutzky_pension_2012].
@@
 Research shows that support for market-based access to welfare services is strongly stratified: people with higher socioeconomic status are more likely to support market-based criteria than disadvantaged groups [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015; @svallfors_political_2007].
-Evidence for pensions points in the same direction: higher income and greater financial gains in funded schemes correlate with stronger support for market-oriented pension policies [@busemeyer_welfare_2020; @kerner_pension_2020].
+Evidence for pensions points in the same direction: higher income and greater financial returns in funded schemes correlate with stronger support for market-oriented pension provision [@busemeyer_welfare_2020; @kerner_pension_2020].
 Yet we know very little about how movement within the class structure shapes preferences for pension commodification.
 This gap is especially salient in Latin America, where mobility has unfolded under high inequality and deeply privatised welfare regimes [@lopez-roldan_comparative_2021].
-Social origins and destinations anchor material interests and justice principles [@alesina_intergenerational_2018; @langsaether_explaining_2022; @gugushvili_subjective_2017], while movement between these positions exposes individuals to new risks, resources and normative environments that reshape fairness judgements [@gugushvili_trends_2014; @helgason_longterm_2023].
+Social origins and destinations anchor material interests and justice principles [@alesina_intergenerational_2018; @langsaether_explaining_2022; @gugushvili_subjective_2017], while movement between these positions exposes individuals to new risks, resources, and normative environments that reshape fairness judgements [@jaime-castillo_social_2019; @helgason_longterm_2023].
 Social mobility can thus foster market-justice orientations, helping explain how inequalities in access to welfare services become normatively sustained [@mau_inequality_2015].
 
-Existing accounts identify several mechanisms through which intergenerational mobility may shape distributive justice preferences but converge on two channels especially relevant for market-based welfare: changes in material interests and shifts in normative frames [@ares_changing_2020; @jaime-castillo_social_2019; @helgason_longterm_2023; @gugushvili_intergenerational_2016c].
+Two channels are particularly relevant to how intergenerational mobility can shape preferences for market-based welfare: changes in material interests and shifts in normative frames [@ares_changing_2020; @jaime-castillo_social_2019; @helgason_longterm_2023; @gugushvili_intergenerational_2016c].
 Mobility alters individuals’ material stakes by changing their occupational status, income, and risk exposure.
-At the same time, it reshapes their fairness judgments.
+At the same time, it reshapes their fairness judgments through acculturation and learning experiences about the norms and values of the destination class.
 Empirically, upward mobility is often associated with stronger meritocratic interpretations of success and fairness, whereas downward mobility is more closely linked to structural explanations of inequality without necessarily eroding meritocratic convictions [@mijs_belief_2022; @bucca_merit_2016; @deng_its_2025].
-Regarding market-based preferences, recent research shows that meritocratic beliefs – the conviction that effort and ability are rewarded – are strongly associated with support for market allocation [@castillo_perceptions_2025].
-In such contexts, the upwardly mobile are especially likely to view markets as just distributors, since their own trajectories appear to confirm that markets appropriately reward effort; those who experience downward mobility may still interpret outcomes through a meritocratic lens, often reading losses as personal failure rather than as evidence against market fairness.
+Regarding market-based preferences, recent research shows that meritocratic perceptions – the conviction that effort is rewarded – are strongly associated with support for market allocation [@castillo_perceptions_2025].
+In such contexts, it seems plausible that social mobility shapes preferences for pension commodification through changes in their material interests and in their meritocratic interpretations of mobility.
 Thus, mobility and meritocracy evolve on partly separate tracks yet intersect systematically [@mau_inequality_2015].
@@
 Meritocratic beliefs are treated not as mediators but as moderators of these effects: among strong meritocrats, upward mobility should amplify pro-commodification attitudes, whereas downward mobility should attenuate – or even reverse – its expected de-commodifying impact, as meritocratic interpretations of effort lead individuals to view market outcomes as fair regardless of direction.
-This inquiry contributes to research on market justice by identifying the causal effects of mobility and using variation in meritocratic beliefs to assess whether these effects follow a meritocratic pathway.
+This inquiry contributes to research on market justice by identifying the causal effects of social mobility and the role of meritocratic beliefs.
 
@@
 It was the first country worldwide in 1981 to fully replace a public pay-as-you-go system with a mandatory, privately administered defined-contribution scheme, while the state has progressively expanded a segmented solidarity pillar to compensate for insufficient benefits [@madariaga_three_2020; @boccardo_30_2020; @solimano_rise_2021].
-In parallel—and despite waves of protest from 2006 to 2019 against the private pension system [@somma_no_2021]—Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing attitudes toward pension distribution and inequality [@araujo_desafios_2012; @galvez_con_2023].
-The combination of profound inequality, market-driven old-age security and contested legitimacy makes Chile a critical case for analysing how mobility and meritocratic beliefs structure support for pension commodification.
+In parallel—and despite waves of protest from 2016 to 2019 against the private pension system [@somma_no_2021]—Chilean subjectivities have been increasingly shaped by neoliberal discourses and market logics, influencing attitudes toward pension distribution and inequality [@araujo_desafios_2012; @galvez_con_2023].
+The combination of profound inequality, market-driven old-age security, and contested legitimacy makes Chile a critical case for analysing how mobility and meritocratic beliefs structure support for pension commodification.
 
@@
 (1) How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile?
-(2) How do meritocratic beliefs condition or moderate this relationship?
+(2) How do meritocratic beliefs condition this relationship?
 
-To address these questions, the study draws on large-scale, representative survey data for the urban Chilean population collected in 2016, 2018 and 2023.
+To address these questions, the study draws on large-scale, representative survey data for the urban Chilean population collected in 2016, 2018, and 2023.
 It adopts a causal framework of mobility effects inspired by @breen_effects_2024.
-The next section outlines the theoretical framework linking preferences for market-based welfare, social mobility and expected heterogeneity by meritocratic beliefs and time, and derives hypotheses.
-A subsequent section presents the causal identification strategy, the data and the analytical approach.
+The following section outlines the theoretical framework linking preferences for pension welfare commodification, social mobility, and expected heterogeneity by meritocratic beliefs and derives hypotheses.
+A subsequent section presents the causal identification strategy, the data, and the analytical approach.
 The empirical sections report the findings, and the article concludes by discussing what these results reveal about the politics of market-based pension welfare and social mobility.

### 2025-12-01 paper/02-antecedents.qmd
Up borrador VF

@@
 In an era of privatisation and commodification, market logic has expanded into core areas of social reproduction—childcare, healthcare, education and pensions—stratifying access and quality [@ferre_welfare_2023; @gingrich_making_2011] and generating feedback effects on public beliefs about welfare and market inequality [@lindh_public_2015].
-In this context, support for market-based welfare provision has grown worldwide, particularly among higher-income groups who view private alternatives as more efficient or higher-quality [@busemeyer_welfare_2020; @lindh_public_2015].
+In this context, support for market-based welfare services provision has grown worldwide, particularly among higher-income groups who view private alternatives as more efficient or higher-quality [@busemeyer_welfare_2020; @lindh_public_2015].
 These institutional transformations provide the backdrop for analysing how far citizens accept the commodification of social protection [@satz_por_2019].
@@
 Situated at this intersection, the emerging literature on market justice preferences conceptualises the market itself as a site of redistribution and asks whether individuals regard market criteria as fair bases for allocating wages and access to core welfare services [@lindh_public_2015; @lindh_bringing_2023; @castillo_perceptions_2025].
-In this perspective, the expansion of market logic into welfare domains implies that social services are treated as legitimate commodities that can be priced and stratified [@busemeyer_welfare_2020], and empirical work suggests that such institutional changes feed back into attitudes, reinforcing market-conforming understandings of fairness and responsibility [@zhu_policy_2015].
+In this perspective, the expansion of market logic into welfare domains implies that social services are treated as legitimate commodities that can be priced and stratified [@busemeyer_welfare_2020], and empirical work suggests that such institutional changes feed back into attitudes, reinforcing market-conforming understandings of fairness and welfare [@zhu_policy_2015].
 
@@
 In this study, I treat market justice less as an abstract doctrine and more concretely as preferences for commodification: support for treating welfare services as commodities, so that better services follow from higher income or ability to pay, and the resulting stratified access is seen as fair [@lindh_public_2015].
-Empirically, such orientations are typically measured with items asking whether it is fair that access to services depends on income, a strategy rooted in research on justifications of capitalist inequality [@kluegel_legitimation_1999] and applied to healthcare and education [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_changes_2025].
+Empirically, such orientations are typically measured with survey items asking whether it is fair that access to better services depends on income, a strategy rooted in research on justifications of capitalist inequality [@kluegel_legitimation_1999] and applied to healthcare and education [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_changes_2025].
 Moving beyond single domains, @lindh_public_2015 develops a comparative index of income-based access to healthcare and education, while @castillo_perceptions_2025 use the same single-item measure but add the pensions domain in Chile.
@@
 This study examines support for the commodification of pensions, defined as normative judgements on whether better pensions should depend on income or individual contributions.
-I treat this as a domain-specific form of market justice: pensions are conceived as commodities whose allocation should follow contribution- and return-based criteria rather than principles of need [@lane_market_1986; @mau_inequality_2015].
-The spread of funded, privately managed schemes built on individual accounts has transformed both the distributive and normative structure of old-age security [@huber_political_2000; @arenas_sistemas_2019].
+I treat this as a domain-specific form of market justice: pensions are conceived as commodities whose allocation should follow contribution- and return-based,  and whose outcome is seen as fair [@lane_market_1986; @mau_inequality_2015].
+The spread of funded, privately managed schemes built on individual capitalisation accounts has transformed both the distributive and normative structure of old-age security [@huber_political_2000; @arenas_sistemas_2019].
 Distributively, these schemes reassign risk from the collective to the individual by tying benefits to formal employment, contribution density, earnings and financial returns, making retirement income heavily dependent on labour-market trajectories and capital-market success [@madero-cabib_private_2019; @oecd_pensions_2023].
@@
 When citizens are asked whether higher pensions should be allocated to those who earn and contribute more, they reflect support for the market allocation of a critical social service and the legitimacy of the resulting risks and inequalities in old age [@busemeyer_welfare_2020; @lindh_public_2015].
+
+Most empirical work on market-based welfare preferences is cross-national and shows that support for income-based access to services is closely tied to countries’ economic and institutional contexts.
+Where public spending and social investment are higher, support for market allocation of health and education tends to be weaker, whereas in regimes with more prominent private providers and heavily privatized services, it is stronger [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015].
+This literature suggests that market justice preferences align with existing institutional designs and outputs, echoing welfare-attitudes and policy-feedback research, which argues that institutions not only redistribute resources but also shape the categories through which people judge deservingness [@campbell_institutional_2020].
+Dominant values and preferences are thus embedded in institutional configurations [@busemeyer_skills_2014], which actively produce moral frameworks that legitimize or contest inequality.
+Longitudinal evidence from Chile shows that agreement that higher-income groups should have access to better welfare services rose markedly between 2016 and 2023, especially for pensions [@castillo_perceptions_2025], plausibly reflecting mixed policy feedback effects of the private pension system [@busemeyer_positive_2021].
+Comparative research also shows that market justice preferences are structured by individuals’ socioeconomic positions and broader attitudes to inequality.
 
@@
 Helgason and Rehm [-@helgason_longterm_2023;-@helgason_class_2025] refine these ideas in a learning framework, in which attitudes evolve slowly through cumulative class experiences that combine economic resources, socialisation into class-specific values, information, and networks.
-Finally, attributional mechanisms highlight self-serving biases, suggesting that individuals explain their mobility trajectories in ways that legitimise their current position, thereby shaping beliefs about effort, deservingness and inequality [@gugushvili_intergenerational_2016; @molina_its_2019; @schmidt_experience_2011].
+Finally, cognitive mechanisms highlight the self-serving bias in causal attribution: individuals interpret their mobility trajectories through internal or external causes in ways that legitimise their current socioeconomic position, thereby shaping their beliefs about effort, deservingness and inequality[@gugushvili_intergenerational_2016; @molina_its_2019; @miller_selfserving_1975].
 In what follows, I focus on two mechanisms that are particularly relevant for market-based pensions: material self-interest and acculturation into meritocratic distributive norms.
@@
 A plausible mechanism runs through changes in material self-interest.
-Social mobility alters individuals’ incomes, job security and access to economic resources [@helgason_longterm_2023; @mau_inequality_2015], thereby modifying the costs and benefits they face in market-based pension systems.
+Social mobility alters individuals’ incomes, job security and access to economic resources [@helgason_longterm_2023; @mau_inequality_2015], thereby modifying the costs and benefits they face in a private individual pension capitalization system.
 Material interests can be understood as the pursuit of economic well-being under the constraints and trade-offs implied by one’s class position [@wrightClassCountsComparative1997a].
@@
 Mobility can thus reshape how individuals interpret success, failure and inequality, with upward moves fostering more individualistic and meritocratic understandings and downward moves encouraging structural accounts of disadvantage [@gugushvili_subjective_2017; @mau_inequality_2015; @mijs_belief_2022] Applied to pensions, those who rise may come to see market-based schemes as fair systems that reward effort and foresight.
-In contrast, those who fall may view commodified pensions as unjust and insufficiently protective.
+In contrast, those who fall may view commodified pensions as unjust and  and that the pension system does not reward merit.
 Therefore, shifts in meritocratic norms can be a second channel through which the mobility effect operates, serving as a complementary acculturation mechanism that may influence support for pension commodification, as examined in the next section.
@@
 They also find that individuals who endorse more meritocratic perceptions–specifically the perception that effort is rewarded– are more likely to support market-based access for healthcare, education and pensions.
-All in all, meritocratic perceptions are a powerful source of inequality legitimation, anchoring who is seen as deserving to be better or worse off, and are strongly linked to how individuals understand unequal life chances and movements up and down the social hierarchy.
+All in all, meritocratic perceptions are a powerful source of inequality legitimation, anchoring who is seen as deserving to be better or worse off, and are strongly linked to how individuals understand unequal life chances and movements within the social hierarchy.
 
@@
 A large body of work consistently shows that upwardly mobile individuals tend to interpret their trajectory through a meritocratic lens, reading their success as evidence that effort and ability are rewarded and reinforcing individualistic attributions of poverty and wealth [@bucca_merit_2016; @gugushvili_intergenerational_2016; @shariff_income_2016; @mijs_belief_2022], even in a highly unequal and low mobility setting were the narratives of individual success reinforces [@deng_its_2025].
-Studies of economic elites in Chile similarly report strong endorsement of meritocracy, with respondents attributing their ascent mainly to talent and business or leadership skills [@atria_economic_2020], in line with self-serving attribution patterns [@kluegel_beliefs_1981; @miller_selfserving_1975].Evidence on downward or blocked mobility is more ambivalent.
+Studies of economic elites in Chile similarly report strong endorsement of meritocracy, with respondents attributing their ascent mainly to talent and business or leadership skills [@atria_economic_2020].
+Evidence on downward or blocked mobility is more ambivalent.
 Exposure to low-mobility contexts and experiences of downward mobility are associated with weaker beliefs that effort is rewarded, lower just-world and opportunity beliefs, and stronger structural attributions for success and failure [@gugushvili_intergenerational_2016; @davidai_why_2018; @day_movin_2017; @mijs_how_2022].
@@
 In this view, markets are seen as fair arenas that properly allocate pensions and other welfare benefits according to merit, so upwardly mobile individuals with strong meritocratic convictions should be especially prone to endorse commodified pensions.
-In contrast, downward mobility typically fosters de-commodifying demands, yet under strong meritocratic beliefs, status loss may be read as individually deserved, weakening—or even reversing—its usual de-commodifying pattern, as market outcomes are still viewed as broadly fair regardless of direction.
+In contrast, downward mobility typically fosters de-commodifying demands, yet under strong meritocratic perceptions, status loss may be read as individually deserved, weakening—or even reversing—its usual de-commodifying pattern, as market outcomes are still viewed as broadly fair regardless of direction.
 Empirical research on redistribution and attitudes toward inequality supports this mechanism.
-Survey studies suggest that subjective social mobility shapes support for redistribution and pension spending through self-serving bias [@schmidt_experience_2011; @gugushvili_subjective_2017], although recent work indicates that this mechanism is evident only under an explicit experimental design [@molina_its_2019].
-In such designs, exposure to upward-mobility information or self-made narratives reduces support for redistribution, increases tolerance of inequality and system justification, and even legitimises exploitative work conditions, whereas exposure to low or downward mobility produces the opposite pattern; crucially, these effects are partly mediated by stronger meritocratic beliefs [@matamoros-lima_social_2025; @shariff_income_2016; @deng_its_2025].
-In this framework, meritocratic beliefs condition the attitudinal consequences of mobility by amplifying the pro-market effects of upward mobility and muting the critical potential of downward mobility; thus, I treat meritocratic perceptions as a moderating mechanism linking intergenerational mobility to preferences for pension commodification.
+Survey studies suggest that subjective social mobility shapes support for redistribution and pension spending through self-serving bias [@schmidt_experience_2011; @gugushvili_subjective_2017], although recent work indicates that this mechanism is evident only under explicit constraints in experimental designs [@molina_its_2019].
+More in line with the acculturation theories, experimental reserach has shown that exposure to upward-mobility information or self-made narratives reduces support for redistribution, increases tolerance of inequality and system justification, and even legitimises exploitative work conditions, whereas exposure to low or downward mobility produces the opposite pattern; crucially, these effects are partly mediated by stronger meritocratic perceptions [@matamoros-lima_social_2025; @shariff_income_2016; @deng_its_2025].
+In this framework, meritocratic perceptions condition the attitudinal consequences of mobility by amplifying the pro-market effects of upward mobility and muting the critical potential of downward mobility; thus, I treat meritocratic perceptions as a moderating mechanism linking intergenerational mobility to preferences for pension commodification.
 
@@
 Scholars argue that this neoliberal shift repurposed the state to create niches of capitalist accumulation—from water to education—giving rise to a model of “public-service capitalism” heavily reliant on state funding [@boccardo_30_2020].
-In this setting, class trajectories unfold under conditions of high inequality and market-based provision in core welfare domains, such as pensions, making Chile a beneficial context for studying support for pension commodification.
+In this setting, class trajectories unfold under conditions of high inequality and market-based provision in core welfare domains, such as pensions.
 
-In pensions, Chile was the first country to fully replace a public pay-as-you-go social insurance system with mandatory individual capitalisation: in 1981, the tripartite cajas de previsión were replaced by a fully funded defined-contribution scheme managed by private Pension Fund Administrators (AFP) [@arenas_sistemas_2019; @huber_political_2000].
+Chile was the first country to fully replace a public pay-as-you-go social insurance system with mandatory individual capitalisation: in 1981, the tripartite *"cajas de previsión"*(insurance schemes) were replaced by a fully funded defined-contribution scheme managed by private Pension Fund Administrators (AFP) [@arenas_sistemas_2019; @huber_political_2000].
 Framed by its principal architect, José Piñera, as “the mother of all battles”, the reform was a way to roll back the redistributive role of the state, shift “freedom of choice” to individuals and create a “society of worker-capitalists” [@borzutzky_pension_2012; @solimano_rise_2021].
@@
 The October 2019 “social outburst”, a period of mass protests and severe political repression throughout the country, crystallised discontent over the privatisation and commodification of social services, alongside a crisis of political legitimacy [@somma_no_2021].
-One of the main conflicts in this period centred on grievances over low pensions and the profits of private pension funds, building on earlier mobilisation by the No+AFP movement since 2016; a survey during the 2019 protests identified pensions as the top public demand [@nudesoc_informe_2020].
+One of the main conflicts in this period centred on grievances over low pensions and the profits of private pension funds, building on earlier mobilisation by the *"No+AFP"* movement since 2016; a survey during the 2019 protests identified pensions as the top public demand [@nudesoc_informe_2020].
 This conflict helped place comprehensive pension reform on the political agenda after 2022.
 However, contestation coexists with enduring market-conforming orientations.
-Research documents strong meritocratic beliefs and high tolerance for inequality [@castillo_meritocracia_2019; @mac-clure_justicia_2024], and the COVID-19 emergency withdrawals of pension savings gave rise to the slogan “with my money, no”, expressing opposition both to greater pay-as-you-go elements and to continued AFP control [@galvez_con_2023].
+Research documents strong meritocratic beliefs and high tolerance for inequality [@castillo_meritocracia_2019; @mac-clure_justicia_2024], and the COVID-19 emergency withdrawals of pension savings gave rise to the slogan *“Con mi plata NO”*, expressing opposition both to greater pay-as-you-go elements and to continued AFP control [@galvez_con_2023].
 This stance reflects the internalisation of the “enterprising self” [@mau_inequality_2015]: families navigate school choice as investment strategies [@canalesceron_sujeto_2021], and workers denounce the pension system as illegitimate in its design and outcomes while simultaneously treating their contributions as personal investments aimed at maximising returns and asserting that they themselves —rather than AFPs, the state or “society”— are the legitimate owners of their pension funds [@panes_criticas_2020].

### 2025-12-01 paper/03-methods.qmd
Up borrador VF

@@
 The analytic sample is restricted to individuals sharing the same origin $(O=j)$.
-Therefore, estimation uses pooled WLS with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.
+Therefore, estimation uses pooled WLS regression with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.
 

### 2025-12-01 paper/04-analysis.qmd
Up borrador VF

@@
 
-@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects.
+@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects (see [Supplementary Material](#anexo) for complete models).
 Results reveal a clear directional asymmetry between upward and downward mobility trajectories.
@@
 The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts.
-These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them.
+These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them, providing favourable evidence for the first hypothesis ($H_1$).
 
@@
 
-To probe whether meritocratic beliefs channel the impact of mobility on preferences for pension commodification, I estimate interaction models between mobility and meritocratic beliefs and examine the conditional treatment effects of mobility at low and high levels of meritocracy (see [Supplementary Material](#anexo) for complete models).
+To probe whether meritocratic perceptions channel the impact of mobility on preferences for pension commodification, I estimate interaction models between mobility and meritocratic perceptions and examine the conditional treatment effects of mobility at low and high levels of meritocracy (see [Supplementary Material](#anexo) for complete models).
 
@@
 
-@fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic beliefs.
+@fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic perceptions.
 Across trajectories, mobility effects among low-merit respondents are uniformly small, imprecise, and centered around zero for both upward and downward moves.
-Among high-merit respondents, only upward mobility from middle to high status shows a clear pro-commodification pattern, with a positive and statistically significant effect on MJP for pensions ($\beta = 0.22$, 95% CI [0.05, 0.38]).
+Among high-merit respondents, only upward mobility from middle to high status shows a clear pro-commodification pattern, with a positive and statistically significant effect on preferences for pension commodification ($\beta = 0.22$, 95% CI [0.05, 0.38]).
 All other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero.
-Nonetheless, the point estimates broadly align with the theoretical expectations for upward moves: upwardly mobile individuals with stronger meritocratic beliefs tend to display higher support for pension commodification than their low-merit counterparts.
-By contrast, downward trajectories exhibit a pattern opposite to the mechanism hypothesis, with high-merit respondents generally showing more de-commodifying attitudes than those with weaker meritocratic beliefs.
-However, formal interaction tests do not detect robust heterogeneity in mobility effects by meritocratic beliefs (see [Supplementary Material](#anexo) for full interaction models): even for the Middle→High trajectory, the simple effect under high meritocracy does not differ significantly from its low-merit counterpart.
-Taken together, these results do not provide convincing support for the mechanism hypothesis: meritocratic beliefs do not operate as a consistent or general channel through which intergenerational mobility shapes preferences for the commodification of pensions.
+Nonetheless, the point estimates broadly align with the theoretical expectations for upward moves: upwardly mobile individuals with stronger meritocratic perceptions tend to display higher support for pension commodification than their low-merit counterparts.
+By contrast, downward trajectories exhibit a pattern opposite to the mechanism hypothesis, with high-merit respondents generally showing more de-commodifying attitudes than those with weaker meritocratic perceptions.
+However, formal interaction tests do not detect robust heterogeneity in mobility effects by meritocratic perceptions (see [Supplementary Material](#anexo) for full interaction models): even for the Middle→High trajectory, the simple effect under high meritocracy does not differ significantly from its low-merit counterpart.
+Taken together, these results provide no convincing evidence that meritocratic perceptions moderate mobility effects on preferences for pension commodification.
+Accordingly, hypotheses $H_{2a}$ and $H_{2b}$—which posit stronger pro-market reactions to upward mobility and attenuated de-commodifying reactions to downward mobility under high meritocracy—are not supported.
 

### 2025-12-01 paper/05-discussion.qmd
Up borrador VF

---
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
---

# Discussion and conclusions

Pension privatisation lies at the heart of Latin America’s trajectory of commodification since the 1980s, reshaping how societies organise old-age security [@ferre_welfare_2023]. By replacing public pay-as-you-go pillars with individual capitalisation schemes—privately managed defined-contribution accounts—pensions became governed by contribution-based entitlements and market performance [@huber_political_2000; @oecd_pensions_2023], with Chile as a compelling example. This shift both restructured who bears old-age risks and redefined the principles deemed legitimate for allocating pension resources [@borzutzky_pension_2012; @madero-cabib_private_2019; @solimano_rise_2021]. From a moral-economy perspective, these agreements generate feedback on public beliefs, fostering opinions that access to core social services should reflect income and individual responsibility, extending the principles of market justice [@mau_inequality_2015]. Existing research shows that socioeconomic position and beliefs about inequality shape support for market-based access to welfare services [@busemeyer_skills_2014; @castillo_perceptions_2025; @immergut_it_2020; @lindh_public_2015; @svallfors_political_2007]. However, we still know little about how citizens assess the fairness of pension commodification in such settings, and how social mobility and meritocratic perceptions shape these judgements. This article addresses this gap by examining whether intergenerational mobility causally affects support for pension commodification in Chile and whether this effect varies by meritocratic perceptions. Drawing on survey data and under explicit identification assumptions, I apply Breen and Ermisch’s [-@breen_effects_2024] causal framework of mobility effects, treating destination class as a treatment, conditioning on origin class, and estimating heterogeneous destination effects.

The findings reveal a tension in Chileans’ judgements of market-based pension welfare. Across 2016, 2018 and 2023, large majorities reject the claim that it is fair for higher-income people to obtain better pensions than those with lower incomes. However, a non-trivial—and growing—minority endorses this principle, with the sharpest increase between 2018 and 2023. Market-based pension allocation thus remains a minority view. However, it is becoming more common, in line with longitudinal evidence of rising support for income-based access to welfare services in Chile [@castillo_perceptions_2025]. This coexistence unfolds in a system of mandatory, privately managed, defined-contribution accounts that tightly link old-age income to earnings, contribution density and financial returns. This design has produced low and unequal pensions and state interventions against old-age poverty [@madero-cabib_private_2019; @galvez_pensiones_2024], while framing pensions as personal investments rather than solidarity-based rights and promoting individual responsibility and contribution-based desert [@borzutzky_pension_2012]. The Chilean case fits evidence that supports the claim that income-based welfare access is stronger in liberal, marketised regimes [@lindh_public_2015]. This coexistence also echoes recent pension conflict—from the “No+AFP” protests to post-pandemic withdrawals and the “Con mi plata NO” mobilisation [@galvez_con_2023]—and illustrates mixed policy feedback: the same regime generates rights-based critiques and proprietarian defences of pension commodification. Together, these patterns show that, despite discontent and a legitimacy crisis, support for pension commodification remains salient, raising the question of which social trajectories and beliefs underpin it.

Turning to the social mobility hypothesis, the results support a causal role of intergenerational mobility in shaping preferences for pension commodification. Rise from middle to high occupational status is causally related to significantly greater support for pension commodification, whereas the opposite holds for descent from high to middle and from high to low status, relative to immobility in the origin class. These patterns are consistent with a myopic material self-interest mechanism [@helgason_longterm_2023] in Chile’s pension system, where benefits mainly depend on individual contributions: upwardly mobile middle→high individuals gain income, employment stability and contribution capacity, and can better bear investment risk, making a contribution-linked, market-based pension regime more attractive. By contrast, those moving down from high positions lose resources and security, become more exposed to labour-market risks and depend more on public or solidarity-based pillars, which helps explain their lower support for income-based allocation [@busemeyer_welfare_2020; @lindh_public_2015]. The fact that significant effects emerge only for trajectories into and out of the upper occupational status resonates with research on static social class and pensions, which shows that higher socioeconomic groups and those obtaining higher financial returns are more supportive of market-based welfare and private pension arrangements [@busemeyer_welfare_2020; @kerner_pension_2020; @castillo_socialization_2024; @otero_power_2024]. It also aligns with evidence on social mobility and redistribution, where upward mobility is linked to less redistributive attitudes and lower support for public spending on pensions, while downward mobility has the opposite effect [@ares_changing_2020; @gugushvili_subjective_2017; @schmidt_experience_2011; @rodriguez_intergenerational_2024; @alesina_intergenerational_2018]. Still, material self-interest may not be the only channel through which mobility affects these judgements, raising the question of whether meritocratic perceptions condition these effects by providing a normative lens to interpret mobility experiences.

The second set of hypotheses examined whether meritocratic perceptions condition the effects of intergenerational mobility. Building on acculturation accounts and self-serving bias arguments [@jaime-castillo_social_2019; @gugushvili_subjective_2017], I expected mobility to affect preferences for pension commodification not only through material self-interest but also through shifts in normative frames of distributive justice. Prior research shows that upwardly mobile individuals tend to interpret their trajectory through a meritocratic lens, reading their success as evidence that effort and ability are rewarded and reinforcing individualistic attributions of poverty and wealth. In contrast, downward or blocked mobility is more often associated with structural explanations of inequality and more ambivalent meritocratic beliefs, sometimes including the internalisation of status loss as personal failure [@atria_economic_2020; @bucca_merit_2016; @gugushvili_intergenerational_2016; @davidai_why_2018; @day_movin_2017; @mijs_how_2022; @deng_its_2025]. Following Mau’s [-@mau_inequality_2015] account of the “majority class”, I therefore expected upward mobility under strong meritocratic convictions to be especially conducive to endorsing commodified pensions, and strong meritocracy to dampen the usual de-commodifying effect of downward mobility. Empirically, however, the interaction models offer little support for this mechanism. Meritocratic perceptions do not systematically moderate the mobility effects identified in the primary analysis, and the few trajectory-specific patterns that tentatively point in the expected direction are small and imprecisely estimated. This result contrasts with survey and experimental studies showing that upward mobility experiences or exposure to self-made narratives reduce support for redistribution and public pension spending partly by reinforcing meritocratic convictions [@schmidt_experience_2011; @gugushvili_subjective_2017; @molina_its_2019; @shariff_income_2016; @matamoros-lima_social_2025; @deng_its_2025]. One possibility is that, in this context, intergenerational occupational mobility—as operationalised here—has limited impact on meritocratic perceptions, which appear more weakly tied to mobility trajectories than previous research suggests and more strongly anchored in broader socialisation processes and class-specific cultures.

This study makes several contributions. Substantively, it offers one of the first systematic analyses of preferences for pension commodification in Chile’s highly unequal and marketised pension regime, addressing Mau’s [-@mau_inequality_2015] question of why core elements of neoliberal welfare arrangements gain public acceptance. By showing that upward and downward intergenerational mobility into and out of upper-class positions causally shape support for market-based pension allocation, the findings highlight social mobility as a key source of legitimation for commodified pensions. Second, the article advances research on market justice and market-based welfare preferences by moving beyond static class locations and showing how intergenerational occupational mobility structures support for pension commodification in ways that partly echo, but also refine, existing evidence on class, income and financial returns. Third, the null results for the moderating role of meritocratic perceptions nuance accounts that tightly link mobility, meritocracy and the embrace of neoliberal welfare: in this context, mobility effects on pension commodification preferences appear to operate independently of meritocratic beliefs. Finally, the study contributes methodologically by applying Breen and Ermisch’s [-@breen_effects_2024] causal framework of mobility effects to survey data, formalising mobility as origin-specific treatment effects and estimating these using linear probability models with inverse probability weighting.

Despite these contributions, the study has limitations that should be acknowledged and read as avenues for future improvement. First, the causal estimates hinge on the construction of inverse probability weighting within Breen and Ermisch’s [-@breen_effects_2024] framework. Identification relies on conditional independence given a relatively limited set of origin-side covariates, together with positivity and SUTVA. Entropy balancing helps to enforce covariate balance and mitigate selection on observables, but unmeasured confounding cannot be ruled out. Sensitivity analyses suggest moderate robustness to omitted variables; therefore, the estimates should be interpreted with caution. Second, while the analysis identifies mobility effects on preferences, it does not explain why or how they arise. The results are analysed through material self-interest and acculturation mechanisms. Still, the study does not implement formal causal mediation analysis, nor can it determine whether meritocratic perceptions function as mediators rather than mere correlates. Third, the findings are context-specific: they are derived from Chile, a highly unequal and strongly marketised pension regime, which enhances internal validity but limits external validity. The patterns documented here, therefore, cannot be straightforwardly generalised to countries with different welfare institutions, mobility regimes or levels of inequality, and should be seen as a theoretically informed case study rather than a cross-national pattern.

Future research could extend these findings in several directions. First, there is a need for richer measures of preferences for pension commodification. This study relies on a single item that conflates income-based allocation and fairness judgments; better items that separate justice principles and dimensions of commodification would allow a finer assessment of how people evaluate pensions as income-conditioned commodities. Second, methodological limitations could be addressed by including more detailed origin-side covariates and life-course data, repeated measures of meritocratic perceptions, and causal mediation analyses that unpack mechanisms such as material self-interest, risk exposure, and meritocratic frames, alongside alternative measures of mobility based on class schemas, income trajectories, or wealth. Third, comparative designs would clarify how institutional context shapes the attitudinal consequences of intergenerational mobility. Applying similar causal strategies in countries with different welfare regimes, levels of inequality, mobility patterns, and degrees of pension marketisation would show whether the Chilean pattern is context-specific or part of a broader configuration linking mobility, merit and support for market-based welfare. Ultimately, clarifying when and for whom market allocation of pensions is judged fair is crucial to understanding the legitimate reach of welfare commodification in unequal, marketised societies.

### 2025-12-01 paper/06-supplementary-material.qmd
Up borrador VF

--- 2025-11-17 paper/07-supplementary-material.qmd
@@
 
+```{r}
+#| label: tbl-complete4
+#| tbl-cap: "Marginal effects of mobility by meritocratic perception (Low vs High Meritocracy)."
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+df_tab <- df_pension_lab_i %>%
+  mutate(
+    Trajectory = sub("\\n.*", "", name),
+    Merit = term,
+    Estimate = round(estimate, 3),
+    `Std. Error` = round(std.error, 3),
+    `CI Low` = round(conf.low, 3),
+    `CI High` = round(conf.high, 3),
+  ) %>%
+  select(Trajectory, Merit, Estimate, `Std. Error`, `CI Low`, `CI High`)
+
+kableExtra::kable(df_tab,
+      format = table_format,
+      caption = NULL,
+      align = "lcccccc") %>%
+kableExtra::kable_classic(full_width = F, html_font = "Cambria")
+```
+
 {{< pagebreak >}}

### 2025-12-01 paper/paper.tex
Up borrador VF

@@
 \setstretch{1.5}
-This document was last modified at 2025-11-29 23:44:43 and it was last
-rendered at 2025-11-29 23:44:43.
+This document was last modified at 2025-12-01 02:26:24 and it was last
+rendered at 2025-12-01 02:26:24.
 
@@
 as a fair basis for the allocation of essential services such as
-healthcare, education and pensions
-(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
+healthcare, education, and pensions
+(\citeproc{ref-castillo_changes_2025}{Castillo, Iturra, et al., 2025};
+\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015};
 \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
@@
 individual-capitalisation schemes---most radically in Chile---pensions
-became organised around contribution-based entitlements and market
-performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
+became organised around funded contribution-based entitlements and
+market performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
 \citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
@@
 pensions points in the same direction: higher income and greater
-financial gains in funded schemes correlate with stronger support for
-market-oriented pension policies
+financial returns in funded schemes correlate with stronger support for
+market-oriented pension provision
 (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
@@
 while movement between these positions exposes individuals to new risks,
-resources and normative environments that reshape fairness judgements
-(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
-\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). Social
-mobility can thus foster market-justice orientations, helping explain
-how inequalities in access to welfare services become normatively
-sustained (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
-
-Existing accounts identify several mechanisms through which
-intergenerational mobility may shape distributive justice preferences
-but converge on two channels especially relevant for market-based
-welfare: changes in material interests and shifts in normative frames
+resources, and normative environments that reshape fairness judgements
+(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
+\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}). Social mobility can thus foster market-justice
+orientations, helping explain how inequalities in access to welfare
+services become normatively sustained
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
+
+Two channels are particularly relevant to how intergenerational mobility
+can shape preferences for market-based welfare: changes in material
+interests and shifts in normative frames
 (\citeproc{ref-ares_changing_2020}{Ares, 2020};
@@
 changing their occupational status, income, and risk exposure. At the
-same time, it reshapes their fairness judgments. Empirically, upward
-mobility is often associated with stronger meritocratic interpretations
-of success and fairness, whereas downward mobility is more closely
-linked to structural explanations of inequality without necessarily
-eroding meritocratic convictions (\citeproc{ref-bucca_merit_2016}{Bucca,
-2016}; \citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
+same time, it reshapes their fairness judgments through acculturation
+and learning experiences about the norms and values of the destination
+class. Empirically, upward mobility is often associated with stronger
+meritocratic interpretations of success and fairness, whereas downward
+mobility is more closely linked to structural explanations of inequality
+without necessarily eroding meritocratic convictions
+(\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
+\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
 \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). Regarding
 market-based preferences, recent research shows that meritocratic
-beliefs -- the conviction that effort and ability are rewarded -- are
-strongly associated with support for market allocation
+perceptions -- the conviction that effort is rewarded -- are strongly
+associated with support for market allocation
 (\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}). In such contexts, the upwardly mobile are especially likely to
-view markets as just distributors, since their own trajectories appear
-to confirm that markets appropriately reward effort; those who
-experience downward mobility may still interpret outcomes through a
-meritocratic lens, often reading losses as personal failure rather than
-as evidence against market fairness. Thus, mobility and meritocracy
-evolve on partly separate tracks yet intersect systematically
+2025}). In such contexts, it seems plausible that social mobility shapes
+preferences for pension commodification through changes in their
+material interests and in their meritocratic interpretations of
+mobility. Thus, mobility and meritocracy evolve on partly separate
+tracks yet intersect systematically
 (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).
@@
 direction. This inquiry contributes to research on market justice by
-identifying the causal effects of mobility and using variation in
-meritocratic beliefs to assess whether these effects follow a
-meritocratic pathway.
+identifying the causal effects of social mobility and the role of
+meritocratic beliefs.
 
@@
 \citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
-despite waves of protest from 2006 to 2019 against the private pension
+despite waves of protest from 2016 to 2019 against the private pension
 system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
@@
 2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
-profound inequality, market-driven old-age security and contested
+profound inequality, market-driven old-age security, and contested
 legitimacy makes Chile a critical case for analysing how mobility and
@@
 \item
-  How do meritocratic beliefs condition or moderate this relationship?
+  How do meritocratic beliefs condition this relationship?
 \end{enumerate}
@@
 representative survey data for the urban Chilean population collected in
-2016, 2018 and 2023. It adopts a causal framework of mobility effects
+2016, 2018, and 2023. It adopts a causal framework of mobility effects
 inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
-The next section outlines the theoretical framework linking preferences
-for market-based welfare, social mobility and expected heterogeneity by
-meritocratic beliefs and time, and derives hypotheses. A subsequent
-section presents the causal identification strategy, the data and the
-analytical approach. The empirical sections report the findings, and the
-article concludes by discussing what these results reveal about the
-politics of market-based pension welfare and social mobility.
+The following section outlines the theoretical framework linking
+preferences for pension welfare commodification, social mobility, and
+expected heterogeneity by meritocratic beliefs and derives hypotheses. A
+subsequent section presents the causal identification strategy, the
+data, and the analytical approach. The empirical sections report the
+findings, and the article concludes by discussing what these results
+reveal about the politics of market-based pension welfare and social
+mobility.
 
@@
 (\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In this context,
-support for market-based welfare provision has grown worldwide,
+support for market-based welfare services provision has grown worldwide,
 particularly among higher-income groups who view private alternatives as
@@
 changes feed back into attitudes, reinforcing market-conforming
-understandings of fairness and responsibility
+understandings of fairness and welfare
 (\citeproc{ref-zhu_policy_2015}{Ling Zhu \& Lipsmeyer, 2015}).
@@
 (\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Empirically, such
-orientations are typically measured with items asking whether it is fair
-that access to services depends on income, a strategy rooted in research
-on justifications of capitalist inequality
+orientations are typically measured with survey items asking whether it
+is fair that access to better services depends on income, a strategy
+rooted in research on justifications of capitalist inequality
 (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999}) and
@@
 form of market justice: pensions are conceived as commodities whose
-allocation should follow contribution- and return-based criteria rather
-than principles of need (\citeproc{ref-lane_market_1986}{Lane, 1986};
+allocation should follow contribution- and return-based, and whose
+outcome is seen as fair (\citeproc{ref-lane_market_1986}{Lane, 1986};
 \citeproc{ref-mau_inequality_2015}{Mau, 2015}). The spread of funded,
-privately managed schemes built on individual accounts has transformed
-both the distributive and normative structure of old-age security
-(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
+privately managed schemes built on individual capitalisation accounts
+has transformed both the distributive and normative structure of old-age
+security (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
 \citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
@@
 \& Iversen, 2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).
+
+Most empirical work on market-based welfare preferences is
+cross-national and shows that support for income-based access to
+services is closely tied to countries' economic and institutional
+contexts. Where public spending and social investment are higher,
+support for market allocation of health and education tends to be
+weaker, whereas in regimes with more prominent private providers and
+heavily privatized services, it is stronger
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). This literature suggests
+that market justice preferences align with existing institutional
+designs and outputs, echoing welfare-attitudes and policy-feedback
+research, which argues that institutions not only redistribute resources
+but also shape the categories through which people judge deservingness
+(\citeproc{ref-campbell_institutional_2020}{Campbell, 2020}). Dominant
+values and preferences are thus embedded in institutional configurations
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014}), which actively
+produce moral frameworks that legitimize or contest inequality.
+Longitudinal evidence from Chile shows that agreement that higher-income
+groups should have access to better welfare services rose markedly
+between 2016 and 2023, especially for pensions
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}), plausibly reflecting mixed policy feedback effects of the
+private pension system (\citeproc{ref-busemeyer_positive_2021}{Busemeyer
+et al., 2021}). Comparative research also shows that market justice
+preferences are structured by individuals' socioeconomic positions and
+broader attitudes to inequality.
 
@@
 class experiences that combine economic resources, socialisation into
-class-specific values, information, and networks. Finally, attributional
-mechanisms highlight self-serving biases, suggesting that individuals
-explain their mobility trajectories in ways that legitimise their
-current position, thereby shaping beliefs about effort, deservingness
-and inequality
-(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
-\citeproc{ref-molina_its_2019}{Molina et al., 2019};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). In what follows,
-I focus on two mechanisms that are particularly relevant for
-market-based pensions: material self-interest and acculturation into
-meritocratic distributive norms.
+class-specific values, information, and networks. Finally, cognitive
+mechanisms highlight the self-serving bias in causal attribution:
+individuals interpret their mobility trajectories through internal or
+external causes in ways that legitimise their current socioeconomic
+position, thereby shaping their beliefs about effort, deservingness and
+inequality(\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili,
+2016b}; \citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975};
+\citeproc{ref-molina_its_2019}{Molina et al., 2019}). In what follows, I
+focus on two mechanisms that are particularly relevant for market-based
+pensions: material self-interest and acculturation into meritocratic
+distributive norms.
 
@@
 \citeproc{ref-mau_inequality_2015}{Mau, 2015}), thereby modifying the
-costs and benefits they face in market-based pension systems. Material
-interests can be understood as the pursuit of economic well-being under
-the constraints and trade-offs implied by one's class position
+costs and benefits they face in a private individual pension
+capitalization system. Material interests can be understood as the
+pursuit of economic well-being under the constraints and trade-offs
+implied by one's class position
 (\citeproc{ref-wrightClassCountsComparative1997a}{Wright, 1997}). Upward
@@
 reward effort and foresight. In contrast, those who fall may view
-commodified pensions as unjust and insufficiently protective. Therefore,
-shifts in meritocratic norms can be a second channel through which the
-mobility effect operates, serving as a complementary acculturation
-mechanism that may influence support for pension commodification, as
-examined in the next section.
+commodified pensions as unjust and and that the pension system does not
+reward merit. Therefore, shifts in meritocratic norms can be a second
+channel through which the mobility effect operates, serving as a
+complementary acculturation mechanism that may influence support for
+pension commodification, as examined in the next section.
 
@@
 off, and are strongly linked to how individuals understand unequal life
-chances and movements up and down the social hierarchy.
+chances and movements within the social hierarchy.
 
@@
 ascent mainly to talent and business or leadership skills
-(\citeproc{ref-atria_economic_2020}{Atria et al., 2020}), in line with
-self-serving attribution patterns
-(\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981};
-\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}).Evidence
-on downward or blocked mobility is more ambivalent. Exposure to
+(\citeproc{ref-atria_economic_2020}{Atria et al., 2020}). Evidence on
+downward or blocked mobility is more ambivalent. Exposure to
 low-mobility contexts and experiences of downward mobility are
@@
 downward mobility typically fosters de-commodifying demands, yet under
-strong meritocratic beliefs, status loss may be read as individually
+strong meritocratic perceptions, status loss may be read as individually
 deserved, weakening---or even reversing---its usual de-commodifying
@@
 \citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}), although recent
-work indicates that this mechanism is evident only under an explicit
-experimental design (\citeproc{ref-molina_its_2019}{Molina et al.,
-2019}). In such designs, exposure to upward-mobility information or
-self-made narratives reduces support for redistribution, increases
-tolerance of inequality and system justification, and even legitimises
-exploitative work conditions, whereas exposure to low or downward
-mobility produces the opposite pattern; crucially, these effects are
-partly mediated by stronger meritocratic beliefs
-(\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
-\citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al., 2025};
-\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In this
-framework, meritocratic beliefs condition the attitudinal consequences
-of mobility by amplifying the pro-market effects of upward mobility and
-muting the critical potential of downward mobility; thus, I treat
-meritocratic perceptions as a moderating mechanism linking
+work indicates that this mechanism is evident only under explicit
+constraints in experimental designs
+(\citeproc{ref-molina_its_2019}{Molina et al., 2019}). More in line with
+the acculturation theories, experimental reserach has shown that
+exposure to upward-mobility information or self-made narratives reduces
+support for redistribution, increases tolerance of inequality and system
+justification, and even legitimises exploitative work conditions,
+whereas exposure to low or downward mobility produces the opposite
+pattern; crucially, these effects are partly mediated by stronger
+meritocratic perceptions (\citeproc{ref-deng_its_2025}{Deng \& Wang,
+2025}; \citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al.,
+2025}; \citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In
+this framework, meritocratic perceptions condition the attitudinal
+consequences of mobility by amplifying the pro-market effects of upward
+mobility and muting the critical potential of downward mobility; thus, I
+treat meritocratic perceptions as a moderating mechanism linking
 intergenerational mobility to preferences for pension commodification.
@@
 class trajectories unfold under conditions of high inequality and
-market-based provision in core welfare domains, such as pensions, making
-Chile a beneficial context for studying support for pension
-commodification.
-
-In pensions, Chile was the first country to fully replace a public
-pay-as-you-go social insurance system with mandatory individual
-capitalisation: in 1981, the tripartite cajas de previsión were replaced
-by a fully funded defined-contribution scheme managed by private Pension
-Fund Administrators (AFP) (\citeproc{ref-arenas_sistemas_2019}{Arenas,
-2019}; \citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
-Framed by its principal architect, José Piñera, as ``the mother of all
-battles'', the reform was a way to roll back the redistributive role of
-the state, shift ``freedom of choice'' to individuals and create a
-``society of worker-capitalists''
-(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
-\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). Dependent workers
-contribute 10\% of their wages to individual accounts, plus an
+market-based provision in core welfare domains, such as pensions.
+
+Chile was the first country to fully replace a public pay-as-you-go
+social insurance system with mandatory individual capitalisation: in
+1981, the tripartite \emph{``cajas de previsión''}(insurance schemes)
+were replaced by a fully funded defined-contribution scheme managed by
+private Pension Fund Administrators (AFP)
+(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
+\citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}). Framed by
+its principal architect, José Piñera, as ``the mother of all battles'',
+the reform was a way to roll back the redistributive role of the state,
+shift ``freedom of choice'' to individuals and create a ``society of
+worker-capitalists'' (\citeproc{ref-borzutzky_pension_2012}{Borzutzky,
+2012}; \citeproc{ref-solimano_rise_2021}{Solimano, 2021}). Dependent
+workers contribute 10\% of their wages to individual accounts, plus an
 administrative fee, while employers contribute nothing; AFPs invest
@@
 pensions and the profits of private pension funds, building on earlier
-mobilisation by the No+AFP movement since 2016; a survey during the 2019
-protests identified pensions as the top public demand
+mobilisation by the \emph{``No+AFP''} movement since 2016; a survey
+during the 2019 protests identified pensions as the top public demand
 (\citeproc{ref-nudesoc_informe_2020}{NUDESOC, 2020}). This conflict
@@
 and the COVID-19 emergency withdrawals of pension savings gave rise to
-the slogan ``with my money, no'', expressing opposition both to greater
-pay-as-you-go elements and to continued AFP control
+the slogan \emph{``Con mi plata NO''}, expressing opposition both to
+greater pay-as-you-go elements and to continued AFP control
 (\citeproc{ref-galvez_con_2023}{Gálvez, 2023}). This stance reflects the
@@
 sample is restricted to individuals sharing the same origin \((O=j)\).
-Therefore, estimation uses pooled WLS with stabilized IPW--ATT from
-entropy balancing and CR2 standard errors clustered by individual to
-address heteroskedasticity and within-person dependence.
+Therefore, estimation uses pooled WLS regression with stabilized
+IPW--ATT from entropy balancing and CR2 standard errors clustered by
+individual to address heteroskedasticity and within-person dependence.
 
@@
 commodification, obtained from weighted models that combine stabilized
-IPW with covariate adjustment and fixed wave effects. Results reveal a
-clear directional asymmetry between upward and downward mobility
-trajectories. Among upwardly mobile respondents, only the Middle→High
-trajectory exhibits a significant positive effect on support for pension
-commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
-Interpreted as an origin-specific ATT, this estimate compares---for
-individuals who actually moved from the middle to the high socioeconomic
-status---their observed endorsement with the counterfactual endorsement
-they would have expressed had they remained immobile in the middle
-socioeconomic status. Holding all other predictors constant, the point
-estimate implies a 9 percentage point higher probability of endorsing
-market-based access to pension benefits for Middle→High movers relative
-to their own immobility counterfactual, statistically significant at the
-95\% confidence level. In contrast, two downward trajectories:
-High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
-(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
-effects, suggesting that individuals who experience downward mobility
-from higher-status origins express weaker preferences for market-based
-pension benefits. The remaining pathways (Low→Middle, Low→High,
-Middle→Low) display no significant differences relative to their
-non-mobile counterparts. These patterns indicate that upward movement
-within the upper segment (Middle→High) reinforces pro-commodification
-attitudes, whereas downward movement from privileged origins reduces
-them.
+IPW with covariate adjustment and fixed wave effects (see
+\hyperref[anexo]{Supplementary Material} for complete models). Results
+reveal a clear directional asymmetry between upward and downward
+mobility trajectories. Among upwardly mobile respondents, only the
+Middle→High trajectory exhibits a significant positive effect on support
+for pension commodification (\(\beta\) = 0.09, 95\% CI {[}0.02,
+0.16{]}). Interpreted as an origin-specific ATT, this estimate
+compares---for individuals who actually moved from the middle to the
+high socioeconomic status---their observed endorsement with the
+counterfactual endorsement they would have expressed had they remained
+immobile in the middle socioeconomic status. Holding all other
+predictors constant, the point estimate implies a 9 percentage point
+higher probability of endorsing market-based access to pension benefits
+for Middle→High movers relative to their own immobility counterfactual,
+statistically significant at the 95\% confidence level. In contrast, two
+downward trajectories: High→Middle (\(\beta\) = −0.07, CI {[}−0.15,
+−0.00{]}) and High→Low (\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show
+significant negative effects, suggesting that individuals who experience
+downward mobility from higher-status origins express weaker preferences
+for market-based pension benefits. The remaining pathways (Low→Middle,
+Low→High, Middle→Low) display no significant differences relative to
+their non-mobile counterparts. These patterns indicate that upward
+movement within the upper segment (Middle→High) reinforces
+pro-commodification attitudes, whereas downward movement from privileged
+origins reduces them, providing favourable evidence for the first
+hypothesis (\(H_1\)).
 
@@
 
-To probe whether meritocratic beliefs channel the impact of mobility on
-preferences for pension commodification, I estimate interaction models
-between mobility and meritocratic beliefs and examine the conditional
-treatment effects of mobility at low and high levels of meritocracy (see
-\hyperref[anexo]{Supplementary Material} for complete models).
+To probe whether meritocratic perceptions channel the impact of mobility
+on preferences for pension commodification, I estimate interaction
+models between mobility and meritocratic perceptions and examine the
+conditional treatment effects of mobility at low and high levels of
+meritocracy (see \hyperref[anexo]{Supplementary Material} for complete
+models).
 
@@
 from the meritocracy interaction models as marginal effects of the
-mobility treatment at low and high meritocratic beliefs. Across
+mobility treatment at low and high meritocratic perceptions. Across
 trajectories, mobility effects among low-merit respondents are uniformly
@@
 high status shows a clear pro-commodification pattern, with a positive
-and statistically significant effect on MJP for pensions
-(\(\beta = 0.22\), 95\% CI {[}0.05, 0.38{]}). All other
+and statistically significant effect on preferences for pension
+commodification (\(\beta = 0.22\), 95\% CI {[}0.05, 0.38{]}). All other
 trajectories---Low→Middle, Low→High, Middle→Low, High→Middle, and
@@
 expectations for upward moves: upwardly mobile individuals with stronger
-meritocratic beliefs tend to display higher support for pension
+meritocratic perceptions tend to display higher support for pension
 commodification than their low-merit counterparts. By contrast, downward
@@
 with high-merit respondents generally showing more de-commodifying
-attitudes than those with weaker meritocratic beliefs. However, formal
-interaction tests do not detect robust heterogeneity in mobility effects
-by meritocratic beliefs (see \hyperref[anexo]{Supplementary Material}
-for full interaction models): even for the Middle→High trajectory, the
-simple effect under high meritocracy does not differ significantly from
-its low-merit counterpart. Taken together, these results do not provide
-convincing support for the mechanism hypothesis: meritocratic beliefs do
-not operate as a consistent or general channel through which
-intergenerational mobility shapes preferences for the commodification of
-pensions.
+attitudes than those with weaker meritocratic perceptions. However,
+formal interaction tests do not detect robust heterogeneity in mobility
+effects by meritocratic perceptions (see \hyperref[anexo]{Supplementary
+Material} for full interaction models): even for the Middle→High
+trajectory, the simple effect under high meritocracy does not differ
+significantly from its low-merit counterpart. Taken together, these
+results provide no convincing evidence that meritocratic perceptions
+moderate mobility effects on preferences for pension commodification.
+Accordingly, hypotheses \(H_{2a}\) and \(H_{2b}\)---which posit stronger
+pro-market reactions to upward mobility and attenuated de-commodifying
+reactions to downward mobility under high meritocracy---are not
+supported.
 
@@
 
-\section{Discussion}\label{discussion}
-
-\section{Conclusion}\label{conclusion}
+\section{Discussion and conclusions}\label{discussion-and-conclusions}
+
+Pension privatisation lies at the heart of Latin America's trajectory of
+commodification since the 1980s, reshaping how societies organise
+old-age security (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). By
+replacing public pay-as-you-go pillars with individual capitalisation
+schemes---privately managed defined-contribution accounts---pensions
+became governed by contribution-based entitlements and market
+performance (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
+2000}; \citeproc{ref-oecd_pensions_2023}{OECD, 2023}), with Chile as a
+compelling example. This shift both restructured who bears old-age risks
+and redefined the principles deemed legitimate for allocating pension
+resources (\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
+\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
+\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). From a moral-economy
+perspective, these agreements generate feedback on public beliefs,
+fostering opinions that access to core social services should reflect
+income and individual responsibility, extending the principles of market
+justice (\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Existing
+research shows that socioeconomic position and beliefs about inequality
+shape support for market-based access to welfare services
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). However, we
+still know little about how citizens assess the fairness of pension
+commodification in such settings, and how social mobility and
+meritocratic perceptions shape these judgements. This article addresses
+this gap by examining whether intergenerational mobility causally
+affects support for pension commodification in Chile and whether this
+effect varies by meritocratic perceptions. Drawing on survey data and
+under explicit identification assumptions, I apply Breen and Ermisch's
+(\citeproc{ref-breen_effects_2024}{2024}) causal framework of mobility
+effects, treating destination class as a treatment, conditioning on
+origin class, and estimating heterogeneous destination effects.
+
+The findings reveal a tension in Chileans' judgements of market-based
+pension welfare. Across 2016, 2018 and 2023, large majorities reject the
+claim that it is fair for higher-income people to obtain better pensions
+than those with lower incomes. However, a non-trivial---and
+growing---minority endorses this principle, with the sharpest increase
+between 2018 and 2023. Market-based pension allocation thus remains a
+minority view. However, it is becoming more common, in line with
+longitudinal evidence of rising support for income-based access to
+welfare services in Chile
+(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
+2025}). This coexistence unfolds in a system of mandatory, privately
+managed, defined-contribution accounts that tightly link old-age income
+to earnings, contribution density and financial returns. This design has
+produced low and unequal pensions and state interventions against
+old-age poverty (\citeproc{ref-galvez_pensiones_2024}{Gálvez et al.,
+2024}; \citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al.,
+2019}), while framing pensions as personal investments rather than
+solidarity-based rights and promoting individual responsibility and
+contribution-based desert
+(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). The Chilean
+case fits evidence that supports the claim that income-based welfare
+access is stronger in liberal, marketised regimes
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). This coexistence also
+echoes recent pension conflict---from the ``No+AFP'' protests to
+post-pandemic withdrawals and the ``Con mi plata NO'' mobilisation
+(\citeproc{ref-galvez_con_2023}{Gálvez, 2023})---and illustrates mixed
+policy feedback: the same regime generates rights-based critiques and
+proprietarian defences of pension commodification. Together, these
+patterns show that, despite discontent and a legitimacy crisis, support
+for pension commodification remains salient, raising the question of
+which social trajectories and beliefs underpin it.
+
+Turning to the social mobility hypothesis, the results support a causal
+role of intergenerational mobility in shaping preferences for pension
+commodification. Rise from middle to high occupational status is
+causally related to significantly greater support for pension
+commodification, whereas the opposite holds for descent from high to
+middle and from high to low status, relative to immobility in the origin
+class. These patterns are consistent with a myopic material
+self-interest mechanism (\citeproc{ref-helgason_longterm_2023}{Helgason
+\& Rehm, 2023}) in Chile's pension system, where benefits mainly depend
+on individual contributions: upwardly mobile middle→high individuals
+gain income, employment stability and contribution capacity, and can
+better bear investment risk, making a contribution-linked, market-based
+pension regime more attractive. By contrast, those moving down from high
+positions lose resources and security, become more exposed to
+labour-market risks and depend more on public or solidarity-based
+pillars, which helps explain their lower support for income-based
+allocation (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen,
+2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). The fact that
+significant effects emerge only for trajectories into and out of the
+upper occupational status resonates with research on static social class
+and pensions, which shows that higher socioeconomic groups and those
+obtaining higher financial returns are more supportive of market-based
+welfare and private pension arrangements
+(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
+\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
+\citeproc{ref-kerner_pension_2020}{Kerner, 2020};
+\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024}). It also aligns
+with evidence on social mobility and redistribution, where upward
+mobility is linked to less redistributive attitudes and lower support
+for public spending on pensions, while downward mobility has the
+opposite effect (\citeproc{ref-alesina_intergenerational_2018}{Alesina
+et al., 2018}; \citeproc{ref-ares_changing_2020}{Ares, 2020};
+\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
+\citeproc{ref-rodriguez_intergenerational_2024}{Rodriguez \&
+Matilla-Garcia, 2024}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
+2011}). Still, material self-interest may not be the only channel
+through which mobility affects these judgements, raising the question of
+whether meritocratic perceptions condition these effects by providing a
+normative lens to interpret mobility experiences.
+
+The second set of hypotheses examined whether meritocratic perceptions
+condition the effects of intergenerational mobility. Building on
+acculturation accounts and self-serving bias arguments
+(\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
+\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
+Marqués-Perales, 2019}), I expected mobility to affect preferences for
+pension commodification not only through material self-interest but also
+through shifts in normative frames of distributive justice. Prior
+research shows that upwardly mobile individuals tend to interpret their
+trajectory through a meritocratic lens, reading their success as
+evidence that effort and ability are rewarded and reinforcing
+individualistic attributions of poverty and wealth. In contrast,
+downward or blocked mobility is more often associated with structural
+explanations of inequality and more ambivalent meritocratic beliefs,
+sometimes including the internalisation of status loss as personal
+failure (\citeproc{ref-atria_economic_2020}{Atria et al., 2020};
+\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
+\citeproc{ref-davidai_why_2018}{Davidai, 2018};
+\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
+\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
+\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
+\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022}). Following Mau's
+(\citeproc{ref-mau_inequality_2015}{2015}) account of the ``majority
+class'', I therefore expected upward mobility under strong meritocratic
+convictions to be especially conducive to endorsing commodified
+pensions, and strong meritocracy to dampen the usual de-commodifying
+effect of downward mobility. Empirically, however, the interaction
+models offer little support for this mechanism. Meritocratic perceptions
+do not systematically moderate the mobility effects identified in the
+primary analysis, and the few trajectory-specific patterns that
+tentatively point in the expected direction are small and imprecisely
+estimated. This result contrasts with survey and experimental studies
+showing that upward mobility experiences or exposure to self-made
+narratives reduce support for redistribution and public pension spending
+partly by reinforcing meritocratic convictions
+(\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
+\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
+\citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al., 2025};
+\citeproc{ref-molina_its_2019}{Molina et al., 2019};
+\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011};
+\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). One
+possibility is that, in this context, intergenerational occupational
+mobility---as operationalised here---has limited impact on meritocratic
+perceptions, which appear more weakly tied to mobility trajectories than
+previous research suggests and more strongly anchored in broader
+socialisation processes and class-specific cultures.
+
+This study makes several contributions. Substantively, it offers one of
+the first systematic analyses of preferences for pension commodification
+in Chile's highly unequal and marketised pension regime, addressing
+Mau's (\citeproc{ref-mau_inequality_2015}{2015}) question of why core
+elements of neoliberal welfare arrangements gain public acceptance. By
+showing that upward and downward intergenerational mobility into and out
+of upper-class positions causally shape support for market-based pension
+allocation, the findings highlight social mobility as a key source of
+legitimation for commodified pensions. Second, the article advances
+research on market justice and market-based welfare preferences by
+moving beyond static class locations and showing how intergenerational
+occupational mobility structures support for pension commodification in
+ways that partly echo, but also refine, existing evidence on class,
+income and financial returns. Third, the null results for the moderating
+role of meritocratic perceptions nuance accounts that tightly link
+mobility, meritocracy and the embrace of neoliberal welfare: in this
+context, mobility effects on pension commodification preferences appear
+to operate independently of meritocratic beliefs. Finally, the study
+contributes methodologically by applying Breen and Ermisch's
+(\citeproc{ref-breen_effects_2024}{2024}) causal framework of mobility
+effects to survey data, formalising mobility as origin-specific
+treatment effects and estimating these using linear probability models
+with inverse probability weighting.
+
+Despite these contributions, the study has limitations that should be
+acknowledged and read as avenues for future improvement. First, the
+causal estimates hinge on the construction of inverse probability
+weighting within Breen and Ermisch's
+(\citeproc{ref-breen_effects_2024}{2024}) framework. Identification
+relies on conditional independence given a relatively limited set of
+origin-side covariates, together with positivity and SUTVA. Entropy
+balancing helps to enforce covariate balance and mitigate selection on
+observables, but unmeasured confounding cannot be ruled out. Sensitivity
+analyses suggest moderate robustness to omitted variables; therefore,
+the estimates should be interpreted with caution. Second, while the
+analysis identifies mobility effects on preferences, it does not explain
+why or how they arise. The results are analysed through material
+self-interest and acculturation mechanisms. Still, the study does not
+implement formal causal mediation analysis, nor can it determine whether
+meritocratic perceptions function as mediators rather than mere
+correlates. Third, the findings are context-specific: they are derived
+from Chile, a highly unequal and strongly marketised pension regime,
+which enhances internal validity but limits external validity. The
+patterns documented here, therefore, cannot be straightforwardly
+generalised to countries with different welfare institutions, mobility
+regimes or levels of inequality, and should be seen as a theoretically
+informed case study rather than a cross-national pattern.
+
+Future research could extend these findings in several directions.
+First, there is a need for richer measures of preferences for pension
+commodification. This study relies on a single item that conflates
+income-based allocation and fairness judgments; better items that
+separate justice principles and dimensions of commodification would
+allow a finer assessment of how people evaluate pensions as
+income-conditioned commodities. Second, methodological limitations could
+be addressed by including more detailed origin-side covariates and
+life-course data, repeated measures of meritocratic perceptions, and
+causal mediation analyses that unpack mechanisms such as material
+self-interest, risk exposure, and meritocratic frames, alongside
+alternative measures of mobility based on class schemas, income
+trajectories, or wealth. Third, comparative designs would clarify how
+institutional context shapes the attitudinal consequences of
+intergenerational mobility. Applying similar causal strategies in
+countries with different welfare regimes, levels of inequality, mobility
+patterns, and degrees of pension marketisation would show whether the
+Chilean pattern is context-specific or part of a broader configuration
+linking mobility, merit and support for market-based welfare.
+Ultimately, clarifying when and for whom market allocation of pensions
+is judged fair is crucial to understanding the legitimate reach of
+welfare commodification in unequal, marketised societies.
 
@@
 
+\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
+Busemeyer, M., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive} and
+{Negative}: {New Perspectives} on {Feedback Effects} in {Public Opinion}
+on the {Welfare State}. \emph{British Journal of Political Science},
+\emph{51}(1), 137--162. \url{https://doi.org/10.1017/S0007123418000534}
+
 \bibitem[\citeproctext]{ref-busemeyer_welfare_2020}
@@
 \url{https://doi.org/10.1086/706980}
+
+\bibitem[\citeproctext]{ref-campbell_institutional_2020}
+Campbell, J. L. (2020). \emph{Institutional {Change} and
+{Globalization}}. Princeton University Press.
+\url{https://doi.org/10.2307/j.ctv131bw68}
 
@@
 University Press. \url{https://doi.org/10.1017/CBO9780511791529}
-
-\bibitem[\citeproctext]{ref-gugushvili_trends_2014}
-Gugushvili, A. (2014). Trends, {Covariates} and {Consequences} of
-{Intergenerational Social Mobility} in {Post- Socialist Societies}.
 

### 2025-12-01 paper/paper.tex
Up V1 borrador final
Up borrador VF1

@@
 
-% Colores extra por nombre
-\usepackage[dvipsnames]{xcolor}
+% Paquete para colores
+\usepackage{xcolor}
+
+% Definir color en formato HEX
+\definecolor{mylinkcolor}{HTML}{3827a1}
+
 % Ajustes de hipervínculos
-% \hypersetup{
-%   colorlinks = true,
-%   citecolor  = DarkSlateBlue,     % << citas
-%   urlcolor   = DarkSlateBlue  % URLs/DOIs
-% }
-% 
+\hypersetup{
+  colorlinks = true,       % activar links coloreados
+  linkcolor  = mylinkcolor,
+  citecolor  = mylinkcolor,
+  urlcolor   = mylinkcolor
+}
 \begin{document}
@@
 \setstretch{1.5}
-This document was last modified at 2025-12-01 02:26:24 and it was last
-rendered at 2025-12-01 02:26:24.
+This document was last modified at 2025-12-01 02:47:28 and it was last
+rendered at 2025-12-01 02:47:28.
 

### 2025-12-03 paper/02-antecedents.qmd
Up draft presentation

@@
 Social mobility and meritocratic perceptions are closely intertwined in the acculturation mechanism, although the direction of this link can vary across trajectories and contexts.
-A large body of work consistently shows that upwardly mobile individuals tend to interpret their trajectory through a meritocratic lens, reading their success as evidence that effort and ability are rewarded and reinforcing individualistic attributions of poverty and wealth [@bucca_merit_2016; @gugushvili_intergenerational_2016; @shariff_income_2016; @mijs_belief_2022], even in a highly unequal and low mobility setting were the narratives of individual success reinforces [@deng_its_2025].
+A large body of work consistently shows that upwardly mobile individuals tend to interpret their trajectory through a meritocratic lens, reading their success as evidence that effort and ability are rewarded and reinforcing individualistic attributions of poverty and wealth [@bucca_merit_2016; @gugushvili_intergenerational_2016; @traini_climbing_2025; @mijs_belief_2022], even in a highly unequal and low mobility setting were the narratives of individual success reinforces [@deng_its_2025].
 Studies of economic elites in Chile similarly report strong endorsement of meritocracy, with respondents attributing their ascent mainly to talent and business or leadership skills [@atria_economic_2020].
@@
 Exposure to low-mobility contexts and experiences of downward mobility are associated with weaker beliefs that effort is rewarded, lower just-world and opportunity beliefs, and stronger structural attributions for success and failure [@gugushvili_intergenerational_2016; @davidai_why_2018; @day_movin_2017; @mijs_how_2022].
-Yet, under strong meritocratic convictions, status loss can also be internalised as personal failure, sustaining system-justifying attitudes and the legitimation of existing hierarchies as a psychological defence strategy [@day_movin_2017; @mau_inequality_2015; @deng_its_2025].
+Yet, under strong meritocratic convictions, status loss can also be internalised as personal failure, sustaining system-justifying attitudes and the legitimation of existing hierarchies as a psychological defence strategy [@day_movin_2017; @mau_inequality_2015; @traini_climbing_2025].
 In sum, upward mobility is generally associated with stronger meritocratic perceptions, whereas downward mobility is more closely linked to structural explanations of inequality, without necessarily eroding meritocratic convictions.
@@
 Empirical research on redistribution and attitudes toward inequality supports this mechanism.
-Survey studies suggest that subjective social mobility shapes support for redistribution and pension spending through self-serving bias [@schmidt_experience_2011; @gugushvili_subjective_2017], although recent work indicates that this mechanism is evident only under explicit constraints in experimental designs [@molina_its_2019].
+Survey studies suggest that subjective social mobility shapes support for redistribution and pension spending through self-serving bias [@schmidt_experience_2011; @gugushvili_subjective_2017; @traini_climbing_2025], although recent work indicates that this mechanism is evident only under explicit constraints in experimental designs [@molina_its_2019].
 More in line with the acculturation theories, experimental reserach has shown that exposure to upward-mobility information or self-made narratives reduces support for redistribution, increases tolerance of inequality and system justification, and even legitimises exploitative work conditions, whereas exposure to low or downward mobility produces the opposite pattern; crucially, these effects are partly mediated by stronger meritocratic perceptions [@matamoros-lima_social_2025; @shariff_income_2016; @deng_its_2025].
+However, recent research has tested Piketty's (1995) hypothesis that meritocracy would mediate the effect of upward social mobility on redistributive preferences, without finding support for it [@rodriguez_intergenerational_2024].
 In this framework, meritocratic perceptions condition the attitudinal consequences of mobility by amplifying the pro-market effects of upward mobility and muting the critical potential of downward mobility; thus, I treat meritocratic perceptions as a moderating mechanism linking intergenerational mobility to preferences for pension commodification.

### 2025-12-03 paper/05-discussion.qmd
Up draft presentation

@@
 However, it is becoming more common, in line with longitudinal evidence of rising support for income-based access to welfare services in Chile [@castillo_perceptions_2025].
-This coexistence unfolds in a system of mandatory, privately managed, defined-contribution accounts that tightly link old-age income to earnings, contribution density and financial returns.
-This design has produced low and unequal pensions and state interventions against old-age poverty [@madero-cabib_private_2019; @galvez_pensiones_2024], while framing pensions as personal investments rather than solidarity-based rights and promoting individual responsibility and contribution-based desert [@borzutzky_pension_2012].
-The Chilean case fits evidence that supports the claim that income-based welfare access is stronger in liberal, marketised regimes [@lindh_public_2015].
-This coexistence also echoes recent pension conflict—from the “No+AFP” protests to post-pandemic withdrawals and the “Con mi plata NO” mobilisation [@galvez_con_2023]—and illustrates mixed policy feedback: the same regime generates rights-based critiques and proprietarian defences of pension commodification.
-Together, these patterns show that, despite discontent and a legitimacy crisis, support for pension commodification remains salient, raising the question of which social trajectories and beliefs underpin it.
+This coexistence unfolds within a system of mandatory, privately managed, defined-contribution accounts that tightly link old-age income to earnings, contribution density and financial returns [@madero-cabib_private_2019; @galvez_pensiones_2024], and that frame pensions as personal investments rather than solidarity-based rights, promoting individual responsibility, self-economisation, contribution-based desert and casting workers as employee-entrepreneurs who must navigate financial risks and “make their money work” [@borzutzky_pension_2012].
+The Chilean case thus fits evidence that income-based welfare access is more strongly endorsed in liberal, marketised regimes [@lindh_public_2015].This coexistence also echoes recent pension conflict—from the “No+AFP” protests to post-pandemic withdrawals and the “Con mi plata NO” mobilisation [@galvez_con_2023]—and illustrates mixed policy feedback: the same regime generates rights-based critiques and proprietarian defences of pension commodification.
+Together, these patterns suggest that, despite discontent and a legitimacy crisis, support for pension commodification remains a salient and contested orientation, raising the question of which social trajectories and beliefs underpin it.
 
@@
 Prior research shows that upwardly mobile individuals tend to interpret their trajectory through a meritocratic lens, reading their success as evidence that effort and ability are rewarded and reinforcing individualistic attributions of poverty and wealth.
-In contrast, downward or blocked mobility is more often associated with structural explanations of inequality and more ambivalent meritocratic beliefs, sometimes including the internalisation of status loss as personal failure [@atria_economic_2020; @bucca_merit_2016; @gugushvili_intergenerational_2016; @davidai_why_2018; @day_movin_2017; @mijs_how_2022; @deng_its_2025].
+In contrast, downward or blocked mobility is more often associated with structural explanations of inequality and more ambivalent meritocratic beliefs, sometimes including the internalisation of status loss as personal failure [@traini_climbing_2025; @gugushvili_intergenerational_2016; @davidai_why_2018; @day_movin_2017; @mijs_how_2022; @deng_its_2025].
 Following Mau’s [-@mau_inequality_2015] account of the “majority class”, I therefore expected upward mobility under strong meritocratic convictions to be especially conducive to endorsing commodified pensions, and strong meritocracy to dampen the usual de-commodifying effect of downward mobility.
@@
 Meritocratic perceptions do not systematically moderate the mobility effects identified in the primary analysis, and the few trajectory-specific patterns that tentatively point in the expected direction are small and imprecisely estimated.
-This result contrasts with survey and experimental studies showing that upward mobility experiences or exposure to self-made narratives reduce support for redistribution and public pension spending partly by reinforcing meritocratic convictions [@schmidt_experience_2011; @gugushvili_subjective_2017; @molina_its_2019; @shariff_income_2016; @matamoros-lima_social_2025; @deng_its_2025].
+This result contrasts with survey and experimental studies showing that upward mobility experiences or exposure to self-made narratives reduce support for redistribution and public pension spending partly by reinforcing meritocratic convictions [@schmidt_experience_2011; @gugushvili_subjective_2017; @molina_its_2019; @shariff_income_2016; @matamoros-lima_social_2025; @deng_its_2025], and aligns with recent research that has shown that meritocracy does not mediate the effect of social mobility on redistributive preferences [@rodriguez_intergenerational_2024].
 One possibility is that, in this context, intergenerational occupational mobility—as operationalised here—has limited impact on meritocratic perceptions, which appear more weakly tied to mobility trajectories than previous research suggests and more strongly anchored in broader socialisation processes and class-specific cultures.
 
-This study makes several contributions.
+The contributions of this study can be summarized as follows.
 Substantively, it offers one of the first systematic analyses of preferences for pension commodification in Chile’s highly unequal and marketised pension regime, addressing Mau’s [-@mau_inequality_2015] question of why core elements of neoliberal welfare arrangements gain public acceptance.
@@
 
-Despite these contributions, the study has limitations that should be acknowledged and read as avenues for future improvement.
+Despite its merits, the study has limitations that should be acknowledged and read as avenues for future improvement.
 First, the causal estimates hinge on the construction of inverse probability weighting within Breen and Ermisch’s [-@breen_effects_2024] framework.
-Identification relies on conditional independence given a relatively limited set of origin-side covariates, together with positivity and SUTVA.
+Identification relies on conditional independence given a relatively limited but relevant set of origin-side covariates, together with positivity and SUTVA.
 Entropy balancing helps to enforce covariate balance and mitigate selection on observables, but unmeasured confounding cannot be ruled out.
@@
 Still, the study does not implement formal causal mediation analysis, nor can it determine whether meritocratic perceptions function as mediators rather than mere correlates.
-Third, the findings are context-specific: they are derived from Chile, a highly unequal and strongly marketised pension regime, which enhances internal validity but limits external validity.
+Third, the findings are context-specific: they are derived from Chile, a highly unequal country with a strongly marketised pension regime, which enhances internal validity but limits external validity.
 The patterns documented here, therefore, cannot be straightforwardly generalised to countries with different welfare institutions, mobility regimes or levels of inequality, and should be seen as a theoretically informed case study rather than a cross-national pattern.
@@
 This study relies on a single item that conflates income-based allocation and fairness judgments; better items that separate justice principles and dimensions of commodification would allow a finer assessment of how people evaluate pensions as income-conditioned commodities.
-Second, methodological limitations could be addressed by including more detailed origin-side covariates and life-course data, repeated measures of meritocratic perceptions, and causal mediation analyses that unpack mechanisms such as material self-interest, risk exposure, and meritocratic frames, alongside alternative measures of mobility based on class schemas, income trajectories, or wealth.
+Second, methodological limitations could be addressed by including more detailed origin-side covariates and life-course data, repeated measures of meritocratic perceptions, and causal mediation analyses that unpack mechanisms such as material self-interest, risk exposure, and meritocratic frames, alongside alternative mobility measures based on class schemas, income trajectories, or wealth.
 Third, comparative designs would clarify how institutional context shapes the attitudinal consequences of intergenerational mobility.
-Applying similar causal strategies in countries with different welfare regimes, levels of inequality, mobility patterns, and degrees of pension marketisation would show whether the Chilean pattern is context-specific or part of a broader configuration linking mobility, merit and support for market-based welfare.
+Applying similar causal strategies in countries with different welfare regimes, levels of inequality, mobility patterns, and degrees of pension marketisation would show whether the Chilean pattern is context-specific or part of a broader configuration linking mobility, merit and support for market-based welfare [@mau_inequality_2015].
 Ultimately, clarifying when and for whom market allocation of pensions is judged fair is crucial to understanding the legitimate reach of welfare commodification in unequal, marketised societies.

### 2025-12-03 paper/paper.qmd
Up borrador VF
Up draft presentation

--- 2025-11-17 paper/paper.qmd
@@
 title: "Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility"
+date: today
+date-format: "MMMM, YYYY"
 format:
   html: 
+    latex-environment: true
     code-fold: true
@@
 author:
-  - name: Andreas Laffert Tamayo
+  - name: 'Andreas Laffert Tamayo\thanks{This M.A.
+thesis in Sociology is one of my first achievements on the long path toward becoming an academic committed to the study of inequalities and to a public sociology.
+This journey would not have been possible without my relentless effort and the support of my family, friends and colleagues.
+Last but certainly not least, I want to acknowledge the motivation behind my research: the everyday processes through which neoliberalism is legitimised by ordinary people.
+This project was made possible through the financial support of ANID (BECAS/MAGÍSTER NACIONAL 2024) and the intellectual support and academic environment provided by the FONDECYT Regular project Nº 1250518.
+Any errors or omissions in this work are my sole responsibility.
+Comments are welcome at: andreas.laffert@uchile.cl}'
     orcid: 0000-0002-9008-2454
@@
 abstract: | 
- My abstract
+ The privatisation and marketisation of social services have been linked to feedback effects on public attitudes towards welfare and market inequality.
+From a moral-economy perspective, prior research shows that support for market-based access to core services depends on socio-structural position, inequality beliefs and institutional context.
+Yet, we know little about how citizens judge the fairness of pension commodification, and in particular how intergenerational mobility shapes these judgements and whether meritocratic perceptions condition this relationship.
+I address this gap by analysing preferences for market-based pension access in Chile, a highly unequal country with constrained mobility and a private pension system based on mandatory individual capitalisation that depends heavily on contribution histories.
+Drawing on three waves of the Chilean Longitudinal Social Survey (ELSOC; 3,435 observations on 1,787 individuals) and under explicit identification assumptions, I apply a recent causal framework for mobility effects that treats destination class as a treatment conditional on origin class, and estimate heterogeneous destination effects on support for pension commodification.
+Results show that upward mobility from middle to upper occupational status increases support for market-based pension access relative to immobility.
+In contrast, downward mobility from upper to middle or lower status reduces it.
+Meritocratic perceptions do not significantly moderate these effects, suggesting that mobility and meritocracy operate through largely separate pathways.
+These findings advance research on welfare marketisation by providing causal evidence on how intergenerational mobility shapes preferences for market-based welfare, and by assessing the potential moderating role of meritocratic perceptions in a highly commodified regime.
   \newline
@@
 
-This document was last modified at {{< modified >}}
-and it was last rendered at {{< now >}}.
-
 :::: {.justify}
@@
 
-{{< include 06-conclusion.qmd >}}
-
 # References
@@
 
-{{< include 07-supplementary-material.qmd >}}
+{{< include 06-supplementary-material.qmd >}}

### 2025-12-03 paper/paper.tex
Up draft presentation

--- 2025-12-01 paper/paper.tex
@@
 
-  \author{Andreas Laffert Tamayo}
+  \author{Andreas Laffert
+Tamayo\thanks{This M.A.
+thesis in Sociology is one of my first achievements on the long path toward becoming an academic committed to the study of inequalities and to a public sociology.
+This journey would not have been possible without my relentless effort and the support of my family, friends and colleagues.
+Last but certainly not least, I want to acknowledge the motivation behind my research: the everyday processes through which neoliberalism is legitimised by ordinary people.
+This project was made possible through the financial support of ANID (BECAS/MAGÍSTER NACIONAL 2024) and the intellectual support and academic environment provided by the FONDECYT Regular project Nº 1250518.
+Any errors or omissions in this work are my sole responsibility.
+Comments are welcome at: andreas.laffert@uchile.cl}}
             \affil{%
@@
 
-\date{}
-
-% Colores extra por nombre
-\usepackage[dvipsnames]{xcolor}
-% Ajustes de hipervínculos
-% \hypersetup{
-%   colorlinks = true,
-%   citecolor  = DarkSlateBlue,     % << citas
-%   urlcolor   = DarkSlateBlue  % URLs/DOIs
-% }
-% 
+\date{December, 2025}
+
+\usepackage{xcolor}
+\definecolor{mylinkcolor}{HTML}{413395}
+\hypersetup{
+  colorlinks = true,
+  linkcolor  = mylinkcolor,
+  citecolor  = mylinkcolor,
+  urlcolor   = mylinkcolor
+}
 \begin{document}
@@
 \begin{abstract}
-My abstract \newline \textbf{Keywords}: Pension commodification ·
-Intergenerational mobility · Causal inference · Meritocracy · Chile
+The privatisation and marketisation of social services have been linked
+to feedback effects on public attitudes towards welfare and market
+inequality. From a moral-economy perspective, prior research shows that
+support for market-based access to core services depends on
+socio-structural position, inequality beliefs and institutional context.
+Yet, we know little about how citizens judge the fairness of pension
+commodification, and in particular how intergenerational mobility shapes
+these judgements and whether meritocratic perceptions condition this
+relationship. I address this gap by analysing preferences for
+market-based pension access in Chile, a highly unequal country with
+constrained mobility and a private pension system based on mandatory
+individual capitalisation that depends heavily on contribution
+histories. Drawing on three waves of the Chilean Longitudinal Social
+Survey (ELSOC; 3,435 observations on 1,787 individuals) and under
+explicit identification assumptions, I apply a recent causal framework
+for mobility effects that treats destination class as a treatment
+conditional on origin class, and estimate heterogeneous destination
+effects on support for pension commodification. Results show that upward
+mobility from middle to upper occupational status increases support for
+market-based pension access relative to immobility. In contrast,
+downward mobility from upper to middle or lower status reduces it.
+Meritocratic perceptions do not significantly moderate these effects,
+suggesting that mobility and meritocracy operate through largely
+separate pathways. These findings advance research on welfare
+marketisation by providing causal evidence on how intergenerational
+mobility shapes preferences for market-based welfare, and by assessing
+the potential moderating role of meritocratic perceptions in a highly
+commodified regime. \newline \textbf{Keywords}: Pension commodification
+· Intergenerational mobility · Causal inference · Meritocracy · Chile
 \end{abstract}
@@
 \setstretch{1.5}
-This document was last modified at 2025-12-01 02:26:24 and it was last
-rendered at 2025-12-01 02:26:24.
-
 \section{Introduction}\label{introduction}
@@
 \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022};
-\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), even in a
+\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}), even in a
 highly unequal and low mobility setting were the narratives of
@@
 (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
-\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
-\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In sum, upward mobility
-is generally associated with stronger meritocratic perceptions, whereas
-downward mobility is more closely linked to structural explanations of
-inequality, without necessarily eroding meritocratic convictions.
+\citeproc{ref-mau_inequality_2015}{Mau, 2015};
+\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}). In sum,
+upward mobility is generally associated with stronger meritocratic
+perceptions, whereas downward mobility is more closely linked to
+structural explanations of inequality, without necessarily eroding
+meritocratic convictions.
 
@@
 (\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
-\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}), although recent
-work indicates that this mechanism is evident only under explicit
+\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011};
+\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}), although
+recent work indicates that this mechanism is evident only under explicit
 constraints in experimental designs
@@
 2025}; \citeproc{ref-matamoros-lima_social_2025}{Matamoros-Lima et al.,
-2025}; \citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In
-this framework, meritocratic perceptions condition the attitudinal
-consequences of mobility by amplifying the pro-market effects of upward
-mobility and muting the critical potential of downward mobility; thus, I
-treat meritocratic perceptions as a moderating mechanism linking
-intergenerational mobility to preferences for pension commodification.
+2025}; \citeproc{ref-shariff_income_2016}{Shariff et al., 2016}).
+However, recent research has tested Piketty's (1995) hypothesis that
+meritocracy would mediate the effect of upward social mobility on
+redistributive preferences, without finding support for it
+(\citeproc{ref-rodriguez_intergenerational_2024}{Rodriguez \&
+Matilla-Garcia, 2024}). In this framework, meritocratic perceptions
+condition the attitudinal consequences of mobility by amplifying the
+pro-market effects of upward mobility and muting the critical potential
+of downward mobility; thus, I treat meritocratic perceptions as a
+moderating mechanism linking intergenerational mobility to preferences
+for pension commodification.
 
@@
 (\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
-2025}). This coexistence unfolds in a system of mandatory, privately
+2025}). This coexistence unfolds within a system of mandatory, privately
 managed, defined-contribution accounts that tightly link old-age income
-to earnings, contribution density and financial returns. This design has
-produced low and unequal pensions and state interventions against
-old-age poverty (\citeproc{ref-galvez_pensiones_2024}{Gálvez et al.,
-2024}; \citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al.,
-2019}), while framing pensions as personal investments rather than
-solidarity-based rights and promoting individual responsibility and
-contribution-based desert
-(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). The Chilean
-case fits evidence that supports the claim that income-based welfare
-access is stronger in liberal, marketised regimes
-(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). This coexistence also
+to earnings, contribution density and financial returns
+(\citeproc{ref-galvez_pensiones_2024}{Gálvez et al., 2024};
+\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019}),
+and that frame pensions as personal investments rather than
+solidarity-based rights, promoting individual responsibility,
+self-economisation, contribution-based desert and casting workers as
+employee-entrepreneurs who must navigate financial risks and ``make
+their money work'' (\citeproc{ref-borzutzky_pension_2012}{Borzutzky,
+2012}). The Chilean case thus fits evidence that income-based welfare
+access is more strongly endorsed in liberal, marketised regimes
+(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).This coexistence also
 echoes recent pension conflict---from the ``No+AFP'' protests to
@@
 proprietarian defences of pension commodification. Together, these
-patterns show that, despite discontent and a legitimacy crisis, support
-for pension commodification remains salient, raising the question of
-which social trajectories and beliefs underpin it.
+patterns suggest that, despite discontent and a legitimacy crisis,
+support for pension commodification remains a salient and contested
+orientation, raising the question of which social trajectories and
+beliefs underpin it.
 
@@
 sometimes including the internalisation of status loss as personal
-failure (\citeproc{ref-atria_economic_2020}{Atria et al., 2020};
-\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
-\citeproc{ref-davidai_why_2018}{Davidai, 2018};
+failure (\citeproc{ref-davidai_why_2018}{Davidai, 2018};
 \citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
@@
 \citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b};
-\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022}). Following Mau's
-(\citeproc{ref-mau_inequality_2015}{2015}) account of the ``majority
-class'', I therefore expected upward mobility under strong meritocratic
-convictions to be especially conducive to endorsing commodified
-pensions, and strong meritocracy to dampen the usual de-commodifying
-effect of downward mobility. Empirically, however, the interaction
-models offer little support for this mechanism. Meritocratic perceptions
-do not systematically moderate the mobility effects identified in the
-primary analysis, and the few trajectory-specific patterns that
-tentatively point in the expected direction are small and imprecisely
-estimated. This result contrasts with survey and experimental studies
-showing that upward mobility experiences or exposure to self-made
-narratives reduce support for redistribution and public pension spending
-partly by reinforcing meritocratic convictions
+\citeproc{ref-mijs_how_2022}{Mijs \& Hoy, 2022};
+\citeproc{ref-traini_climbing_2025}{Traini et al., 2025}). Following
+Mau's (\citeproc{ref-mau_inequality_2015}{2015}) account of the
+``majority class'', I therefore expected upward mobility under strong
+meritocratic convictions to be especially conducive to endorsing
+commodified pensions, and strong meritocracy to dampen the usual
+de-commodifying effect of downward mobility. Empirically, however, the
+interaction models offer little support for this mechanism. Meritocratic
+perceptions do not systematically moderate the mobility effects
+identified in the primary analysis, and the few trajectory-specific
+patterns that tentatively point in the expected direction are small and
+imprecisely estimated. This result contrasts with survey and
+experimental studies showing that upward mobility experiences or
+exposure to self-made narratives reduce support for redistribution and
+public pension spending partly by reinforcing meritocratic convictions
 (\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
@@
 \citeproc{ref-schmidt_experience_2011}{Schmidt, 2011};
-\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). One
-possibility is that, in this context, intergenerational occupational
-mobility---as operationalised here---has limited impact on meritocratic
-perceptions, which appear more weakly tied to mobility trajectories than
-previous research suggests and more strongly anchored in broader
-socialisation processes and class-specific cultures.
-
-This study makes several contributions. Substantively, it offers one of
-the first systematic analyses of preferences for pension commodification
-in Chile's highly unequal and marketised pension regime, addressing
-Mau's (\citeproc{ref-mau_inequality_2015}{2015}) question of why core
-elements of neoliberal welfare arrangements gain public acceptance. By
-showing that upward and downward intergenerational mobility into and out
-of upper-class positions causally shape support for market-based pension
+\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}), and aligns
+with recent research that has shown that meritocracy does not mediate
+the effect of social mobility on redistributive preferences
+(\citeproc{ref-rodriguez_intergenerational_2024}{Rodriguez \&
+Matilla-Garcia, 2024}). One possibility is that, in this context,
+intergenerational occupational mobility---as operationalised here---has
+limited impact on meritocratic perceptions, which appear more weakly
+tied to mobility trajectories than previous research suggests and more
+strongly anchored in broader socialisation processes and class-specific
+cultures.
+
+The contributions of this study can be summarized as follows.
+Substantively, it offers one of the first systematic analyses of
+preferences for pension commodification in Chile's highly unequal and
+marketised pension regime, addressing Mau's
+(\citeproc{ref-mau_inequality_2015}{2015}) question of why core elements
+of neoliberal welfare arrangements gain public acceptance. By showing
+that upward and downward intergenerational mobility into and out of
+upper-class positions causally shape support for market-based pension
 allocation, the findings highlight social mobility as a key source of
@@
 
-Despite these contributions, the study has limitations that should be
+Despite its merits, the study has limitations that should be
 acknowledged and read as avenues for future improvement. First, the
@@
 (\citeproc{ref-breen_effects_2024}{2024}) framework. Identification
-relies on conditional independence given a relatively limited set of
-origin-side covariates, together with positivity and SUTVA. Entropy
-balancing helps to enforce covariate balance and mitigate selection on
-observables, but unmeasured confounding cannot be ruled out. Sensitivity
-analyses suggest moderate robustness to omitted variables; therefore,
-the estimates should be interpreted with caution. Second, while the
-analysis identifies mobility effects on preferences, it does not explain
-why or how they arise. The results are analysed through material
-self-interest and acculturation mechanisms. Still, the study does not
-implement formal causal mediation analysis, nor can it determine whether
-meritocratic perceptions function as mediators rather than mere
-correlates. Third, the findings are context-specific: they are derived
-from Chile, a highly unequal and strongly marketised pension regime,
-which enhances internal validity but limits external validity. The
-patterns documented here, therefore, cannot be straightforwardly
-generalised to countries with different welfare institutions, mobility
-regimes or levels of inequality, and should be seen as a theoretically
-informed case study rather than a cross-national pattern.
+relies on conditional independence given a relatively limited but
+relevant set of origin-side covariates, together with positivity and
+SUTVA. Entropy balancing helps to enforce covariate balance and mitigate
+selection on observables, but unmeasured confounding cannot be ruled
+out. Sensitivity analyses suggest moderate robustness to omitted
+variables; therefore, the estimates should be interpreted with caution.
+Second, while the analysis identifies mobility effects on preferences,
+it does not explain why or how they arise. The results are analysed
+through material self-interest and acculturation mechanisms. Still, the
+study does not implement formal causal mediation analysis, nor can it
+determine whether meritocratic perceptions function as mediators rather
+than mere correlates. Third, the findings are context-specific: they are
+derived from Chile, a highly unequal country with a strongly marketised
+pension regime, which enhances internal validity but limits external
+validity. The patterns documented here, therefore, cannot be
+straightforwardly generalised to countries with different welfare
+institutions, mobility regimes or levels of inequality, and should be
+seen as a theoretically informed case study rather than a cross-national
+pattern.
 
@@
 self-interest, risk exposure, and meritocratic frames, alongside
-alternative measures of mobility based on class schemas, income
+alternative mobility measures based on class schemas, income
 trajectories, or wealth. Third, comparative designs would clarify how
@@
 Chilean pattern is context-specific or part of a broader configuration
-linking mobility, merit and support for market-based welfare.
-Ultimately, clarifying when and for whom market allocation of pensions
-is judged fair is crucial to understanding the legitimate reach of
-welfare commodification in unequal, marketised societies.
+linking mobility, merit and support for market-based welfare
+(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Ultimately, clarifying
+when and for whom market allocation of pensions is judged fair is
+crucial to understanding the legitimate reach of welfare commodification
+in unequal, marketised societies.
 
@@
 619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
+
+\bibitem[\citeproctext]{ref-traini_climbing_2025}
+Traini, C., Ehmes, S., \& Gangl, M. (2025). Climbing the ladder or
+falling behind: {How} social mobility shapes perceptions of meritocracy
+in the wake of rising inequality. \emph{PO- LAR Working Paper \#7.
+Frankfurt: Goethe University. Retrieved from Www.polar-Project.org,
+Version Dated 12 September 2025}.
 

### 2025-12-03 presentations/social_mobility_2025/borrador.qmd
Up draft presentation

---
bibliography: "bib/tesis-magister.bib"
csl: "bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
editor_options: 
  chunk_output_type: console
---

```{r setup2, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")

 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }

 library(conflicted)

conflicts_prefer(dplyr::filter)
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(here,
               kableExtra)

options(scipen=999)
rm(list = ls())

```

```{r}
#| label: data
#| include: false

source(file = here("processing/prod_analysis_multinom.R"))
library(here)
library(janitor)
library(dagitty)
library(ggdag)
library(tidyverse)

```

::: columns

::: {.column width="20%"}

![](images/logo_isuc.png)

![](images/jusmer_trans.png)
</br>

:::

::: {.column .column-right width="80%"}

# **Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility**

------------------------------------------------------------------------

**Andreas Laffert^1**

::: {.blue2 .medium}

**^1^Institute of Sociology, Pontifical Catholic University of Chile**

:::

International Conference: The Future of Social Mobility

3-4 December 2025, Chile

:::
:::

# Context and motivation {.xlarge data-background-color="#523870"}

## Context: marketisation & pensions

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @ferre_welfare_2023]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Pension privatization and individual capitalization are central to this trajectory in Latin America [@huber_political_2000; @verbic_political_2019]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

These reforms redistribute old-age risks and also reshape norms about what is fair in welfare: feedback effects on public attitudes [@busemeyer_welfare_2020; @koos_moral_2019; @lindh_bringing_2023]

:::

::: {.box-inv-5 .sp-after-half .fragment style="font-size: 110%;"}

Preferences for the market-based access to social welfare [@busemeyer_skills_2014; @castillo_perceptions_2025; @lindh_public_2015; @immergut_it_2020; @svallfors_political_2007]

:::

## Chile and the research gap

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Chile combines high inequality, deep welfare privatisation and contested private pensions [@flores_top_2020; @boccardo_30_2020; @somma_no_2021; @galvez_con_2023; @madariaga_three_2020]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

High absolute mobility and educational expansion coexist with short-range class moves, restricted access to elite positions, and persistent intergenerational inequalities [@lopez-roldan_comparative_2021; @salgado_uplifting_2025; @torche_intergenerational_2014]
:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

We know that support for market-based services is stratified [@lindh_public_2015; @busemeyer_welfare_2020; @kerner_pension_2020; @otero_power_2024] and linked to meritocracy [@castillo_perceptions_2025], but less about how mobility trajectories and meritocratic beliefs shape these preferences in the pension domain in Chile.
:::

::: {.incremental .highlight-last style="font-size: 110%;"}

- Privatización de los sistemas de pensiones están en el corazón de la trayectoria de mercantilización en AL [@huber_political_2000; @verbic_political_2019]

- The expandension of mandatory or voluntary privately managed schemes built on individual capitalisation accounts has transformed both the distributive and normative structure of old-age security:

  * how allocate old-age risk and tie benefit levels to lifetime labour-market trajectories, erarnings and capital-market success [@madero-cabib_private_2019; @oecd_pensions_2023; @arenas_sistemas_2019]
  * personal investment decisions rather than collective obligations, promoting contribution-based desert and workers have to make their one money work [@borzutzky_pension_2012; @arza_pension_2008]

- No hay estudio sistemático sobre las preferencias por comodificación de pensiones y de los motivos que llevan a abrazar estos principios, menos en AL [@mau_inequality_2015]

:::

## The Chilean case

::: {.incremental .highlight-last style="font-size: 120%;"}

- Despite sustanied economic growth, one of the most eocnomic inequal contries [@chancel_world_2022; @flores_top_2020]
- Profunda privatización y comodificación de áreas de reproducción social con fuerte dependencia Estatal; croweded welfare model [@madariaga_three_2020; @boccardo_30_2020]
- Research on intergenerational mobility shows that Chile combines high absolute mobility with mostly short-range
moves—mainly from working classes into lower service-class positions—while access to upper and elite classes remains tightly
restricted, and that despite the expansion of higher education, relative mobility remains rigid and intergenerational inequalities
persist (Espinoza & Núñez, 2014; López-Roldán & Fachelli, 2021; Salgado et al., 2025; Torche, 2014).
- Pension system: fully funded defined-contribution scheme (individual capitalization) managed by private Pension Fund Administrators (AFP)
- Legitimidad previsional contestada: conflictos [@somma_no_2021] y legitimidad del sistema ("con mi plata no") [@galvez_con_2023]
:::

## This study

::: {.incremental .highlight-last style="font-size: 120%;"}

- La mayoria de las investigaciones sobre preferencias por el bienestar social en base al mercado son comparativas: importa el entorno eocnomico e sintitcuonal
- Sabmos que las preferencias por las pensiones en base al mercado estan estratificadas: indivdouos de mayor estatus socecionmico prefieren mas estas soluciones de mercado o mas market-based welfare en servicios como salud y educacion, y tambien qye las actitudes hacua la desigualdad como las creencias meritocraticas las incrementan
- Sin embargo, no sabemos como el movimiento entre posiciones dentro de la estructura social afecta estas preferencias, lo que es especialmente relevante en piases como CHile donde la mov ha ocirrodo en contextos muy desigualdes y privatizados
- Tampoco sabemos en qué medida el efecto de la mov puede ser explicado por variaciones en las creencias meritocraticas
:::

#  {data-background-color="#523870"}

:::: {style="font-size: 170%;"}

(1) How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile?
(2) How do meritocratic beliefs condition this relationship?

::::

## Hypothesis

::: {.incremental .highlight-last style="font-size: 120%;"}

- Social mobility alters individuals’ incomes, job security and access to economic
resources (Helgason & Rehm, 2023; Mau, 2015), thereby modifying the costs and benefits they face in a private individual
pension capitalization system.

- Social mobility and preferences for redistribution and social spending [@alesina_intergenerational_2018, @rodriguez_intergenerational_2024; @langsaether_explaining_2022; @ares_changing_2020; @gugushvili_subjective_2017]

- $H_1$: Compared to immobility, intergenerational upward mobility is associated with stronger, and downward mobility with weaker, support for the commodification of pensions

:::

## Hypothesis

::: {.incremental .highlight-last style="font-size: 120%;"}

- Social mobility also entails changes in the justice principles and normative beleifs [@helgason_longterm_2023; @jaime-castillo_social_2019; @gugushvili_intergenerational_2016]

- upward mobility
is generally associated with stronger meritocratic perceptions, whereas downward mobility is more closely linked to structural
explanations of inequality, without necessarily eroding meritocratic convictions

- Parte del efecto de la movilidad social en distintas actitudes hacia la deisgaldad pasa por creer fuertemente en la meritocracia

- $H_{2a}$: Under high meritocracy, upward mobility should generate a stronger increase in support for market-based pension access than under low meritocracy.

- $H_{2b}$: Under low meritocracy, downward mobility should moderately reduce support for market-based pension access, whereas under high meritocracy, this negative effect should be attenuated toward neutral or slightly positive attitudes.

:::

# Método {.xlarge data-background-color="#523870"}

(M)odel, (I)nquiry, (D)ata, (A)nswer estrategy [@blair_research_2023]

## Un modelo causal para los efectos de la movilidad (M)

::::: columns
::: {.column width="50%"}
::: {.fragment}

![](images/breen.png)

:::

:::

::: {.column width="50%" .incremental .highlight-last style="font-size: 100%;"}

- @breen_effects_2024 proponen un método para estimar el efecto (causal) de la movilidad social intergernacional sobre un outcome
- ¿En qué habría cambiado el resultado _Y_ entre las personas de origen _j_ que llegaron al destino _k_ si, hipotéticamente, hubieran llegado al destino _k'_?

:::
:::::

## A Causal Model for Mobility Effects

```{r}
#| label: fig-dag
#| out-width: '85%'
#| fig-cap: "Causal graph of the effect of intergenerational social mobility on preferences for commodification. Y = preferences for pension commodification, D = an individual’s class of destination, O = an individual’s class of origin, C = different attributes determined in childhood or earlier that affects an individual’s class of origin and destination. Finally, U = unobserved factors influencing origin attributes."
#| fig-cap-location: bottom
#| echo: false
#| warning: false
#| message: false

dag_mjp <- dagify(
  Y ~ D + C,
  D ~ O + C,
  O ~ C,
  C ~ U,
  coords = time_ordered_coords(
    list(
      "U",
      c("C", "O"),
      "D",
      "Y"
    )
  ),
  exposure = "D",
  outcome = "Y",
  labels = c(
    Y = "Y",
    D = "D",
    O = "O",
    C = "C",
    U = "U"
  )
)

curvatures <- rep(0, 8)
curvatures[3] <- -.25

dag_mjp |>
  tidy_dagitty() |>
  ggplot(aes(x, y, xend = xend, yend = yend)) +
  geom_dag_point(shape = 21, fill = "black", size = 14, stroke = 0.8) +
  geom_dag_edges_arc(curvature = curvatures) +
  geom_dag_text(aes(label = label), color = "white", fontface = "bold", size = 4) +
  theme_dag()

```

## Inquiry and estimand

- How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile?

- @breen_effects_2024: Efecto causal promedio de pasar de una clase de origen *j* a una clase de destino *k* comparando a las personas que realmente alcanzan *k* con su resultado contrafactual si se hubieran mantenido en su origen *j* (@eq-att).

Formally,  for individuals with origin $(O=j)$ who attain destination $(D=k)$, the estimand is:

$$ 
ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
$${#eq-att}

## Datos y variables (D)

::: {.incremental .highlight-last style="font-size: 120%;"}

- [(ELSOC)](https://coes.cl/elsoc/) (COES): panel urbano nacional 2016–2023; muestreo probabilístico estratificado, por conglomerados y multietápico; población objetivo 18–75 años residentes habituales

- Analizo 2016, 2018 y 2023 por disponibilidad de variables ocupacionales

- Muestra analítica: *N*=3.435 observaciones anidadas en *N*=1.787 individuos (2016: 914; 2018: 1.377; 2023: 1.144)

- Submuestras por trayectorias de movilidad (p. ej., baja→baja, baja→media), reteniendo movers y no-movers apareados desde la misma clase de origen

:::

## Outcome

```{r}
#| label: fig-desc
#| fig-cap-location: top
#| fig-cap: "Distribution of responses for pension commodification preferences in Chile (2016-2023)"
#| fig-align: center
#| out-width: '130%'

g1 +
  labs(title = "It is fair that people with higher incomes have better pensions than people\nwith lower incomes?",
       subtitle = "Likert: 1 (Strongly disagree) - 5 (Strongly agree)") +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"),
        text = element_text(size = 16))

```

## Tratamiento

::: {.incremental .highlight-last style="font-size: 120%;"}

- **Class assignment (O and D):** Using ISCO-08/ISEI, I define origin (O ∈ {low, medium, high}) and destination (D) based on tertiles.

- For each origin \(O=j\), the treatment is to reach \(D=k\) (mobile) vs. immobile \(D=j\).

```{r}
#| label: tbl-matrix
#| tbl-cap: "Movilidad ocupacional por grupos ocupacionales"
#| tbl-cap-location: top
#| results: asis
#| echo: false
library(janitor)
db %>%
  tabyl(estrato_orig, estrato_ocupa) %>%
  adorn_totals(where = c("row", "col")) %>%     # agrega totales
  adorn_percentages("row") %>%                  # porcentajes por fila
  adorn_pct_formatting(digits = 1) %>%          # formato %
  adorn_ns() %>% 
  as_tibble() %>% 
  mutate(Offspring = NA) %>% 
  select(estrato_orig, Offspring, everything()) %>% 
  kableExtra::kable(format = "html", 
                    booktabs= T,
                    align = 'c', 
                    col.names = c("Father↓", "Offspring→", "Low", "Middle", "High", "Total"),
                    caption = NULL) %>%
  kableExtra::kable_styling(latex_options = "hold_position", 
                            bootstrap_options = "striped",
                            full_width = F)

```

:::

## Tratamiento

::: {.incremental .highlight-last style="font-size: 120%;"}

2. **ATT por entropy balancing:**  
   - $C$: (a) parental education (highest of father/mother), in 10 ordinal categories from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
   - En cada estrato \(O=j\), uso `ebal` para **reponderar solo a los controles** (inmóviles) e igualar sus valores de los confusores pre-tratamiento \(C\) a los de los tratados (móviles).  
   - Obtengo **IPW-ATT estabilizados** (tratados=1; controles = pesos de `ebal`).

:::

## Heterogenidad causal

::: {.incremental .highlight-last style="font-size: 120%;"}

- Meritocracia [@young_rise_1958]: dos ítems (“recompensa al esfuerzo” y “a la inteligencia/habilidades”), Likert 1–5; se promedian en un índice y se dicotomiza: baja ≤3 vs. alta ≥4.

- Controls: los mismos pretratamientos $C$ (block backdoor) and wave fixed effects

:::

## Controles
::: {.incremental .highlight-last style="font-size: 120%;"}

- Incluyo los mismos pretratamientos $C$ que en IPW no para re-balancear, sino para bloquear backdoors (según el DAG) y 
- Lograr doble robustez: consistencia si el modelo de pesos o el de resultados está bien especificado [@wooldridge_introductory_2009]
:::

## Estrategia analítica (A)

$$
Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
$$ {#eq-final}

::: {.incremental style="font-size:110%;"}

- **Muestra:** individuos con mismo origen de clase $O=j$  
-  **Outcome:** $Y_{it}$ preferencia por comodificación de pensiones 
- **Tratamiento:** $T_i=1$ si $D_i=k$ (*móvil*), $T_i=0$ si $D_i=j$ (*inmóvil*)  
- **Pesos:** IPW–ATT *estabilizados* con `ebal`  
- **Errores estándar:** CR2 *clusterizados por id*  
- **Efectos fijos:** $\lambda_{t(i)}$ por ola  
- **Controles:** $X_i'\gamma$ covariables propensity score movilidad
- **Interpretación:** $\beta = \widehat{ATT}_{j,k,j}$  
  → efecto causal de alcanzar destino $k$ respecto a permanecer en $j$

:::

# Resultados {.xlarge data-background-color="#523870"}

## Mobility effects

::: {.fragment style="font-size: 110%;"}

```{r}
#| label: fig-mp
#| fig-cap-location: top
#| fig-cap: "Efectos de movilidad en preferencias por la mercantilización de las pensiones"
#| fig-align: center
#| out-width: '180%'
#| fig-asp: 0.55

g3 +
  theme(text = element_text(size = 14))
```

:::

## Heterogenidad

::: {.fragment style="font-size: 110%;"}

```{r}
#| label: fig-mp_i
#| fig-cap-location: top
#| fig-cap: "Efectos de movilidad en preferencias por la mercantilización de las pensiones, según  percepción meritocrática"
#| fig-align: center
#| out-width: '160%'
#| fig-asp: 0.55

g4+
  theme(text = element_text(size = 14))
```

:::

## Robustness check and sensitivity analysis

::: {.incremental .highlight-last style="font-size: 120%;"}

- **E-values** [@cinelli_making_2020]: **robustez moderada** en Middle→High y High→Low; borderline en High→Middle.

- **Check robustez**: I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs. 1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification. Se mantienen e incluso mejoran los resultados en (ii)
:::

# Discusión y conclusiones {data-background-color="#523870"}

## Discusión y conclusiones

::: {.incremental style="font-size: 120%;"}

1) Mark-based pension preferences in Chile: the mayortity rejects pension comodication but recent waves muestran un crecimiento considerable --> contextos altamente merkantilizado y desigual [@lindh_public_2015] y legitimidad controvertida (conflicto y con mi plata no) [@bor]

2) Hypothesis: upward mobility from Middle-High increase support while donward High-Middle y High-Low decrease, significant effects emerge only for trajectories into and out of the upper occupational status resonates with research on static social class and pensions. Meritocracy doosent moderate this effect: (no hay efecto de movilidad en meritocracia), y por ende operan por caminos paralelos.

3) Proyecciones: mejorar $C$, inspeccionar mas a detalle y ocupar otras claisifcaciones ocupacionales, incluir rol del tiempo y mediaciones

:::

# Gracias por su atención!

-   **Github del proyecto:** <https://github.com/Andreas-Lafferte/mobility-market-justice>

## Referencias

### 2025-12-03 presentations/social_mobility_2025/mjp-mobility-conf-2025.qmd
Up draft presentation

---
bibliography: "bib/tesis-magister.bib"
csl: "bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
editor_options: 
  chunk_output_type: console
---

```{r setup2, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")

 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }

 library(conflicted)

conflicts_prefer(dplyr::filter)
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(here,
               kableExtra)

options(scipen=999)
rm(list = ls())

```

```{r}
#| label: data
#| include: false

source(file = here("processing/prod_analysis_multinom.R"))
library(here)
library(janitor)
library(dagitty)
library(ggdag)
library(tidyverse)

```

::: columns

::: {.column width="20%"}

![](images/logo_isuc.png)

</br>

![](images/jusmer_trans.png)

:::

::: {.column .column-right width="80%"}

# **Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility**

------------------------------------------------------------------------

**Andreas Laffert^1^**

::: {.blue2 .medium}

**^1^Institute of Sociology, Pontifical Catholic University of Chile**

:::

International Conference: The Future of Social Mobility

3-4 December 2025, Chile

:::
:::

# Context and motivation {.xlarge data-background-color="#523870"}

## Context: marketisation and pensions

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Privatization and commodification of public goods and social services [@gingrich_making_2011; @ferre_welfare_2023]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

In Latin America, pension privatization and individual capitalization have been pivotal [@huber_political_2000; @verbic_political_2019]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Reallocate old-age risks and reshape public norms about what is fair in welfare [@busemeyer_welfare_2020; @madero-cabib_private_2019; @borzutzky_pension_2012]

:::

::: {.box-inv-5 .sp-after-half .fragment style="font-size: 110%;"}

Preferences for the market-based access to social welfare [@busemeyer_skills_2014; @castillo_perceptions_2025; @lindh_public_2015; @immergut_it_2020; @svallfors_political_2007]

:::

## Chile and the research gap

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Chile combines high inequality, deep welfare privatisation and contested private pensions [@flores_top_2020; @somma_no_2021; @madariaga_three_2020]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

High absolute mobility coexist with short-range class moves, restricted access to elite positions, and persistent intergenerational inequalities [@lopez-roldan_comparative_2021; @salgado_uplifting_2025; @torche_intergenerational_2014]
:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

We know that support for market-based services is stratified [@lindh_public_2015; @busemeyer_welfare_2020; @kerner_pension_2020; @otero_power_2024] and linked to meritocracy [@castillo_perceptions_2025], but less about how social mobility and meritocratic beliefs shape these preferences in the pension domain.
:::

## Research questions and H1

::: {.incremental .highlight-last style="font-size: 120%;"}

- RQ1: How does intergenerational social mobility affect preferences for pension commodification in Chile?

- Social mobility changes incomes, security and risk exposure, altering material interests in a funded DC system [@busemeyer_welfare_2020; @helgason_longterm_2023; @mau_inequality_2015]

- Social mobility and preferences for redistribution and social spending [@alesina_intergenerational_2018; @rodriguez_intergenerational_2024; @langsaether_explaining_2022; @ares_changing_2020; @gugushvili_subjective_2017]

- $H_1$: Compared to immobility, intergenerational upward mobility is associated with stronger, and downward mobility with weaker, support for the commodification of pensions

:::

## Research questions and H2

::: {.incremental .highlight-last style="font-size: 120%;"}

- RQ2: How do meritocratic beliefs condition the mobility–commodification relationship?

- Social mobility reshapes inequality attitudes by pushing the upwardly mobile toward meritocratic views and the downwardly mobile toward more structural accounts, with strong meritocratic beliefs acting as a key channel of these effects [@mijs_belief_2022; @gugushvili_intergenerational_2016; @bucca_merit_2016; @traini_climbing_2025; @matamoros-lima_social_2025; @deng_its_2025]

- $H_{2}$: Under high meritocracy, upward mobility should be more pro-commodification and downward mobility less de-commodifying than under low meritocracy

:::

# Method {.xlarge data-background-color="#523870"}

(M)odel, (I)nquiry, (D)ata, (A)nswer estrategy [@blair_research_2023]

## Causal model (M): @breen_effects_2024

::::: columns
::: {.column width="50%"}
::: {.fragment}

![](images/breen.png)

:::

:::

::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}

- I follow @breen_effects_2024 to estimate causal effects of intergenerational mobility.
- For each origin class $O=j$, I compare movers from $j$ to $k$ with their counterfactual outcome if they had remained in $j$

:::
:::::

## A Causal Model for Mobility Effects

```{r}
#| label: fig-dag
#| out-width: '95%'
#| fig-cap: "Causal graph of the effect of intergenerational social mobility on preferences for commodification. Y = preferences for pension commodification, D = an individual’s class of destination, O = an individual’s class of origin, C = different attributes determined in childhood or earlier that affects an individual’s class of origin and destination. Finally, U = unobserved factors influencing origin attributes."
#| fig-cap-location: bottom
#| echo: false
#| warning: false
#| message: false

dag_mjp <- dagify(
  Y ~ D + C,
  D ~ O + C,
  O ~ C,
  C ~ U,
  coords = time_ordered_coords(
    list(
      "U",
      c("C", "O"),
      "D",
      "Y"
    )
  ),
  exposure = "D",
  outcome = "Y",
  labels = c(
    Y = "Y",
    D = "D",
    O = "O",
    C = "C",
    U = "U"
  )
)

curvatures <- rep(0, 8)
curvatures[3] <- -.25

dag_mjp |>
  tidy_dagitty() |>
  ggplot(aes(x, y, xend = xend, yend = yend)) +
  geom_dag_point(shape = 21, fill = "black", size = 14, stroke = 0.8) +
  geom_dag_edges_arc(curvature = curvatures) +
  geom_dag_text(aes(label = label), color = "white", fontface = "bold", size = 4) +
  theme_dag()

```

## Inquiry and estimand
::: {.incremental .highlight-last style="font-size: 120%;"}

- How does intergenerational social mobility affect individuals’ preferences for the commodification of pensions in Chile?

- @breen_effects_2024: Average causal effect of moving from origin class *j* to destination class *k*, comparing individuals who actually reach *k* with their counterfactual outcome had they remained in their origin *j* (@eq-att).

- Formally,  for individuals with origin $(O=j)$ who attain destination $(D=k)$, the estimand is:

:::

::: {.fragment}
$$ 
ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
$${#eq-att}
:::

## Data (D)

::: {.incremental .highlight-last style="font-size: 120%;"}

- [ELSOC](https://coes.cl/elsoc/) (COES): national urban panel 2016–2023; stratified, cluster, multistage probability sampling; target population: regular residents aged 18–75

- I analyze 2016, 2018, and 2023 based on the availability of occupational variables

- Analytical sample: *N*=3,435 observations nested in *N*=1,787 individuals (2016: 914; 2018: 1,377; 2023: 1,144)

- Subsamples by mobility trajectories (e.g., low→low, low→medium), retaining movers and non-movers paired from the same class of origin
:::

## Outcome

```{r}
#| label: fig-desc
#| fig-cap-location: top
#| fig-cap: "Distribution of responses for pension commodification preferences in Chile (2016-2023)"
#| fig-align: center
#| out-width: '130%'

g1 +
  labs(title = "It is fair that people with higher incomes have better pensions than people\nwith lower incomes?",
       subtitle = "Likert: 1 (Strongly disagree) - 5 (Strongly agree)") +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"),
        text = element_text(size = 16))

```

## Treatment: mobility and class matrix (D)

::: {.incremental .highlight-last style="font-size: 120%;"}

- **Class assignment (O and D):** Using ISCO-08/ISEI, I define origin $O$ and destination $D$ based on tertiles ∈ {low, medium, high}
- For each origin $(O=j)$, the treatment is to reach $(D=k)$ (mobile) vs. immobile $(D=j)$

:::

::: {.fragment}

```{r}
#| label: tbl-matrix
#| tbl-cap: "Occupational mobility by occupational groups"
#| tbl-cap-location: top
#| results: asis
#| echo: false
library(janitor)
db %>%
  tabyl(estrato_orig, estrato_ocupa) %>%
  adorn_totals(where = c("row", "col")) %>%     # agrega totales
  adorn_percentages("row") %>%                  # porcentajes por fila
  adorn_pct_formatting(digits = 1) %>%          # formato %
  adorn_ns() %>% 
  as_tibble() %>% 
  mutate(Offspring = NA) %>% 
  select(estrato_orig, Offspring, everything()) %>% 
  kableExtra::kable(format = "html", 
                    booktabs= T,
                    align = 'c', 
                    col.names = c("Father↓", "Offspring→", "Low", "Middle", "High", "Total"),
                    caption = NULL) %>%
  kableExtra::kable_styling(latex_options = "hold_position", 
                            bootstrap_options = "striped",
                            full_width = F)

```

:::

## Weights and covariates (D → A)

::: {.incremental .highlight-last style="font-size: 120%;"}

- Pre-treatment covariates $C$: (a) parental education (highest of father/mother), in 10 ordinal categories from no schooling to postgraduate studies; (b) co-residence with both parents at age 15 (0 = no, 1 = yes); (c) nationality (0 = non-Chilean, 1 = Chilean); (d) age in years; (e) sex (0 = male, 1 = female); and (f) indigenous ethnicity (0 = no, 1 = yes).
- Within each origin, entropy balancing reweights non-movers so their covariate distribution matches that of movers → stabilised IPW–ATT.
- These weights target the ATT and help adjust for selection into mobility.

:::

## Heterogeneity and controls (D → A)

::: {.incremental .highlight-last style="font-size: 120%;"}

- Meritocracy: two items (effort and intelligence/skills rewarded), averaged and dichotomised into low vs. high meritocracy.

- I estimate models with mobility × meritocracy interactions to study heterogeneous treatment effects.

- The same pre-treatment covariates $C$ enter the outcome model to block backdoor paths and achieve double robustness.

:::

## Analytic strategy (A)

::: {.incremental .highlight-last style="font-size: 120%;"}

- For each origin class, I estimate weighted linear probability models:

:::

::: {.fragment style="font-size: 120%;"}
$$
Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
$$ {#eq-final}
:::

::: {.incremental style="font-size:120%;"}

- Models use stabilised IPW–ATT weights, wave fixed effects and CR2 cluster-robust SEs by individual.
- $\beta$ is interpreted as $\widehat{ATT}_{j,k,j}$: the causal effect of reaching destination $k$ versus staying in origin $j$

:::

# Results {.xlarge data-background-color="#523870"}

## Mobility effects

::: {.fragment style="font-size: 120%;"}

```{r}
#| label: fig-mp
#| fig-cap-location: top
#| fig-cap: "Mobility effects on preferences for pension comodification"
#| fig-align: center
#| out-width: '180%'
#| fig-asp: 0.6

g3 +
  theme(text = element_text(size = 14))
```

:::

## The rol of meritocracy

::: {.fragment style="font-size: 120%;"}

```{r}
#| label: fig-mp_i
#| fig-cap-location: top
#| fig-cap: "Mobility effects on preferences for pension comodification, by meritocratic perceptions"
#| fig-align: center
#| out-width: '160%'
#| fig-asp: 0.6

g4+
  theme(text = element_text(size = 14))
```

:::

## Robustness check and sensitivity analysis

::: {.incremental .highlight-last style="font-size: 120%;"}

- **Unmeasured confounding** [@cinelli_making_2020]: **Moderate robustness** in Middle→High and High→Low; borderline in High→Middle.

- **Robustness check**: I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs. 1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification. The results are maintained and even improved (only in ii).
:::

# Discussion and conclusions {data-background-color="#523870"}

## Discussion and conclusions

::: {.incremental style="font-size: 120%;"}

1) Most Chileans reject income-based pension allocation, yet support for this market principle has grown in a highly marketised and unequal setting [@lindh_public_2015; @castillo_perceptions_2025].

2) Intergenerational moves into and out of upper-status positions causally shape support for pension commodification, while donward High-Middle y High-Low decrease it [@gugushvili_subjective_2017; @rodriguez_intergenerational_2024; @ares_changing_2020]; meritocratic beliefs correlate with these preferences but seem to operate on a parallel track.

3) Future work: better origin-side measures, richer batteries of market-justice items and comparative applications beyond Chile.

:::

# Thanks!

-   **Github project:** <https://github.com/Andreas-Lafferte/mobility-market-justice>

## References

### 2025-12-03 processing/prod_prep_multinom.R

@@ -197,13 +197,25 @@ db <- elsoc_long_2016_2023 %>%
 
 db <- left_join(db, ola6, by = "idencuesta")
 
+library(DIGCLASS)
+
+
 db$isco88to08 <- occupar::isco88to08(db$ciuo88_m03)
+db$isco88to08_dig <- DIGCLASS::isco88_to_isco08(as.character(db$ciuo08_m03))
+
+frq(db$isco88to08)
+frq(db$isco88to08_dig)
+
 db$isco08 <- NA
 db$isco08[db$ola %in% c(3,7)] <- db$ciuo08_m03[db$ola %in% c(3,7)]
 db$isco08[db$ola %in% c(1)] <- db$isco88to08[db$ola %in% c(1)]
 
 db <- db %>% 
-  mutate(isei08_ocupa = occupar::isco08toISEI08(isco08))
+  mutate(isei08_ocupa = occupar::isco08toISEI08(isco08),
+         isei08_ocupa_dig = DIGCLASS::isco08_to_isei(as.character(isco08)))
+
+frq(db$isei08_ocupa)
+frq(db$isei08_ocupa_dig)
 
 ## Estimar IPW
 

### 2025-12-04 presentations/social_mobility_2025/mjp-mobility-conf-2025.qmd
Up ahora si
up

@@
 
-## Causal model (M): @breen_effects_2024
+## Mobility effects approach
 
@@
 
-## A Causal Model for Mobility Effects
+## Causal model (M)
 
@@
 
-## Inquiry and estimand
+## Inquiry and estimand (I)
 ::: {.incremental .highlight-last style="font-size: 120%;"}
@@
 
-## Treatment: mobility and class matrix (D)
+## Treatment: mobility and class matrix
 
@@
 
-## Weights and covariates (D → A)
+## Weights and covariates
 
@@
 
-## Heterogeneity and controls (D → A)
+## Heterogeneity and controls
 
@@
 
-## Robustness check and sensitivity analysis
-
-::: {.incremental .highlight-last style="font-size: 120%;"}
-
-- **Unmeasured confounding** [@cinelli_making_2020]: **Moderate robustness** in Middle→High and High→Low; borderline in High→Middle.
-
-- **Robustness check**: I re-estimate the models under three alternative codings of the outcome: (i) recoding the neutral category as agreement (3–5 vs.
-1–2), (ii) excluding neutral responses from the analysis, and (iii) using the original 5-point item in a linear specification.
-The results are maintained and even improved (only in ii).
-:::
-
 # Discussion and conclusions {data-background-color="#523870"}
@@
 
-1) Most Chileans reject income-based pension allocation, yet support for this market principle has grown in a highly marketised and unequal setting [@lindh_public_2015; @castillo_perceptions_2025].
+1) Intergenerational moves into and out of upper-status positions causally shape support for pension commodification, while donward High-Middle y High-Low decrease it [@gugushvili_subjective_2017; @helgason_longterm_2023; @busemeyer_welfare_2020; @ares_changing_2020]
 
-2) Intergenerational moves into and out of upper-status positions causally shape support for pension commodification, while donward High-Middle y High-Low decrease it [@gugushvili_subjective_2017; @rodriguez_intergenerational_2024; @ares_changing_2020]; meritocratic beliefs correlate with these preferences but seem to operate on a parallel track.
+2) Meritocratic beliefs correlate with these preferences but seem to operate on a parallel track [@@rodriguez_intergenerational_2024]
 

### 2025-12-10 presentations/seminario_clase_redes/mjp-mobility-redes-2025.qmd
Up presentación seminario redes

---
bibliography: "bib/tesis-magister.bib"
csl: "bib/apa6.csl"
lang: es
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:

      knitr.graphics.auto_pdf: true
format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
    mathjax: "default"
editor_options: 
  chunk_output_type: console
---

```{r setup2, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")

 table_format = if (knitr::is_html_output()) {
   #conditional instructions for kable
   "html" #if html set "html" in format
 } else if (knitr::is_latex_output()) {
   "latex"#if latex set "latex" in format
 }

 library(conflicted)

conflicts_prefer(dplyr::filter)
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(here,
               kableExtra)

options(scipen=999)
rm(list = ls())

```

```{r}
#| label: data
#| include: false

source(file = here("processing/prod_analysis_multinom.R"))
library(here)
library(janitor)
library(dagitty)
library(ggdag)
library(tidyverse)

```

::: columns

::: {.column width="20%"}

![](images/logo_isuc.png)

</br>

![](images/jusmer_trans.png)

:::

::: {.column .column-right width="80%"}

# **Preferencias por la comodificación de las pensiones en Chile: el rol de la movilidad social intergeneracional**

------------------------------------------------------------------------

**Andreas Laffert^1^**

::: {.blue2 .medium}

**^1^Instituto de Sociología, Pontificia Universidad Católica de Chile**

:::

Seminario: Clase, redes personales y actitudes políticas

11 Diciembre 2025, Chile

:::
:::

# Contexto y motivación {.xlarge data-background-color="#523870"}

## Contexto: mercantilización y pensiones

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Privatización y mercantilización de diversos bienes públicos y servicios sociales. [@gingrich_making_2011; @ferre_welfare_2023]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

En América Latina, la privatización de las pensiones y la capitalización individual han sido fundamentales[@huber_political_2000; @verbic_political_2019]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Reasigna los riesgos de la vejez y cambia las normas públicas sobre lo que es justo en materia de bienestar social [@busemeyer_welfare_2020; @madero-cabib_private_2019; @borzutzky_pension_2012]

:::

::: {.box-inv-5 .sp-after-half .fragment style="font-size: 110%;"}

Preferencias por el acceso al bienestar social basado en el mercado [@busemeyer_skills_2014; @castillo_perceptions_2025; @lindh_public_2015; @immergut_it_2020; @svallfors_political_2007]

:::

## Chile y la brecha en la investigación

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Chile combina una elevada desigualdad, una profunda privatización del bienestar y un sistema de pensiones privado tensionado [@flores_top_2020; @somma_no_2021; @madariaga_three_2020]

:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Alta movilidad absoluta coexiste con movimientos de clase de corto alcance, acceso restringido a puestos de élite y desigualdades intergeneracionales persistentes. [@lopez-roldan_comparative_2021; @salgado_uplifting_2025; @torche_intergenerational_2014]
:::

::: {.box-inv-4 .sp-after .fragment style="font-size: 110%;"}

Sabemos que el apoyo a los servicios basados en el mercado está estratificado [@lindh_public_2015; @busemeyer_welfare_2020; @kerner_pension_2020; @otero_power_2024] y vinculado a la meritocracia [@castillo_perceptions_2025], pero sabemos menos sobre cómo la movilidad social y las creencias meritocráticas configuran estas preferencias en el ámbito de las pensiones
:::

## Preguntas de investigación y H1

::: {.incremental .highlight-last style="font-size: 120%;"}

- RQ1: ¿Cómo afecta la movilidad social intergeneracional a las preferencias por la mercantilización de las pensiones en Chile?

- La movilidad social modifica los ingresos, la seguridad y la exposición al riesgo, alterando los intereses materiales en un sistema de capitalización individual de contribución definida[@busemeyer_welfare_2020; @helgason_longterm_2023; @mau_inequality_2015]

- Movilidad social y preferencias por la redistribución y el gasto social [@alesina_intergenerational_2018; @rodriguez_intergenerational_2024; @langsaether_explaining_2022; @ares_changing_2020; @gugushvili_subjective_2017]

- $H_1$: En comparación con la inmovilidad, la movilidad social intergeneracional ascendente se asocia con un mayor apoyo, y la movilidad descendente con un menor apoyo, a la mercantilización de las pensiones

:::

## Preguntas de investigación y H2

::: {.incremental .highlight-last style="font-size: 120%;"}

- RQ2: ¿Cómo las creencias meritocráticas condicionan la relación entre movilidad social y mercantilización?

- La movilidad social reconfigura las actitudes frente a la desigualdad al empujar a quienes experimentan movilidad ascendente hacia visiones más meritocráticas y a quienes experimentan movilidad descendente hacia explicaciones más estructurales, siendo las fuertes creencias meritocráticas un canal clave de estos efectos [@mijs_belief_2022; @gugushvili_intergenerational_2016; @bucca_merit_2016; @traini_climbing_2025; @matamoros-lima_social_2025; @deng_its_2025]

- $H_{2}$: Bajo altos niveles de creencias meritocráticas, la movilidad ascendente debería asociarse con actitudes más pro-mercantilización y la movilidad descendente con actitudes menos des-mercantilizadoras que bajo bajos niveles de meritocracia

:::

# Método {.xlarge data-background-color="#523870"}

(M)odel, (I)nquiry, (D)ata, (A)nswer estrategy [@blair_research_2023]

## Enfoque: Mobility effects

::::: columns
::: {.column width="50%"}
::: {.fragment}

![](images/breen.png)

:::

:::

::: {.column width="50%" .incremental .highlight-last style="font-size: 110%;"}

- Sigo a @breen_effects_2024 para estimar efectos causales de la movilidad social intergeneracional.
- Para cada clase de origen $O=j$, comparo a quienes se desplazan desde $j$ hasta $k$ con su resultado contrafactual si se hubiesen mantenido en $j$.

:::
:::::

## Modelo causal (M)

```{r}
#| label: fig-dag
#| out-width: '95%'
#| fig-cap: "Gráfico causal del efecto de la movilidad social intergeneracional sobre las preferencias por la mercantilización. (Y) denota las preferencias por la mercantilización de las pensiones, (D) la clase de destino del individuo, (O) la clase de origen, (C) diferentes atributos determinados en la infancia o en etapas tempranas que afectan la clase de origen y de destino del individuo y, finalmente, (U) factores no observados que influyen en dichos atributos de origen"
#| fig-cap-location: bottom
#| echo: false
#| warning: false
#| message: false

dag_mjp <- dagify(
  Y ~ D + C,
  D ~ O + C,
  O ~ C,
  C ~ U,
  coords = time_ordered_coords(
    list(
      "U",
      c("C", "O"),
      "D",
      "Y"
    )
  ),
  exposure = "D",
  outcome = "Y",
  labels = c(
    Y = "Y",
    D = "D",
    O = "O",
    C = "C",
    U = "U"
  )
)

curvatures <- rep(0, 8)
curvatures[3] <- -.25

dag_mjp |>
  tidy_dagitty() |>
  ggplot(aes(x, y, xend = xend, yend = yend)) +
  geom_dag_point(shape = 21, fill = "black", size = 14, stroke = 0.8) +
  geom_dag_edges_arc(curvature = curvatures) +
  geom_dag_text(aes(label = label), color = "white", fontface = "bold", size = 4) +
  theme_dag()

```

## Pregunta de investigación y estimando (I)

::: {.incremental .highlight-last style="font-size: 120%;"}

- ¿Cómo afecta la movilidad social intergeneracional las preferencias de las personas por la mercantilización de las pensiones en Chile?

- @breen_effects_2024: efecto causal promedio de pasar desde una clase de origen *j* a una clase de destino *k*, comparando a los individuos que efectivamente alcanzan *k* con su resultado contrafactual si hubiesen permanecido en su clase de origen *j* (@eq-att).

- Formalmente, para individuos con origen ((O = j)) que alcanzan el destino ((D = k)), el estimando es:

:::

::: {.fragment}
$$ 
ATT_{j,k,j} = E\big[Y(D = k) \mid O = j, D = k \big] - E\big[Y(D = j) \mid O = j, D = k \big]
$${#eq-att}
:::

## Datos (D)

::: {.incremental .highlight-last style="font-size: 120%;"}

- [ELSOC](https://coes.cl/elsoc/) (COES): panel urbano nacional 2016–2023; muestreo probabilístico estratificado, por conglomerados y multietápico; población objetivo: residentes habituales de 18 a 75 años.

- Analizo los años 2016, 2018 y 2023 en función de la disponibilidad de variables ocupacionales.

- Muestra analítica: *N* = 3.435 observaciones anidadas en *N* = 1.787 individuos (2016: 914; 2018: 1.377; 2023: 1.144).

- Submuestras por trayectorias de movilidad (por ejemplo, baja→baja, baja→media), reteniendo móviles y no móviles emparejados dentro de la misma clase de origen.
:::

## Outcome

```{r}
#| label: fig-desc
#| fig-cap-location: top
#| fig-cap: "Distribución de las respuestas sobre las preferencias por la mercantilización de las pensiones en Chile (2016–2023)"
#| fig-align: center
#| out-width: '130%'

g1 +
  labs(title = "It is fair that people with higher incomes have better pensions than people\nwith lower incomes?",
       subtitle = "Likert: 1 (Strongly disagree) - 5 (Strongly agree)") +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"),
        text = element_text(size = 16))

```

## Tratamiento: matriz de movilidad y clase

::: {.incremental .highlight-last style="font-size: 120%;"}

- **Asignación de clase (O y D):** A partir de ISCO-08/ISEI, defino el origen ($O$) y el destino ($D$) según terciles ∈ {baja, media, alta}
- Para cada origen $(O = j)$, el tratamiento consiste en alcanzar $(D = k)$ (móvil) versus permanecer inmóvil $(D = j)$

:::

::: {.fragment}

```{r}
#| label: tbl-matrix
#| tbl-cap: "Movilidad ocupacional según grupos ocupacionales"
#| tbl-cap-location: top
#| results: asis
#| echo: false
library(janitor)
db %>%
  tabyl(estrato_orig, estrato_ocupa) %>%
  adorn_totals(where = c("row", "col")) %>%     # agrega totales
  adorn_percentages("row") %>%                  # porcentajes por fila
  adorn_pct_formatting(digits = 1) %>%          # formato %
  adorn_ns() %>% 
  as_tibble() %>% 
  mutate(Offspring = NA) %>% 
  select(estrato_orig, Offspring, everything()) %>% 
  kableExtra::kable(format = "html", 
                    booktabs= T,
                    align = 'c', 
                    col.names = c("Father↓", "Offspring→", "Low", "Middle", "High", "Total"),
                    caption = NULL) %>%
  kableExtra::kable_styling(latex_options = "hold_position", 
                            bootstrap_options = "striped",
                            full_width = F)

```

:::

## Pesos y covariables

::: {.incremental .highlight-last style="font-size: 120%;"}

- Covariables **pretratamiento** $C$: (a) educación parental (máximo nivel educativo de padre/madre), en 10 categorías ordinales desde sin escolaridad hasta estudios de posgrado; (b) corresidencia con ambos padres a los 15 años (0 = no, 1 = sí); (c) nacionalidad (0 = no chilena, 1 = chilena); (d) edad en años; (e) sexo (0 = hombre, 1 = mujer); y (f) pertenencia a pueblo indígena (0 = no, 1 = sí).
- Dentro de cada clase de origen, el *entropy balancing* repondera a los no móviles de modo que su distribución de covariables coincida con la de los móviles → IPW–ATT estabilizado.
- Estos pesos identifican el ATT y contribuyen a ajustar por la selección en la movilidad.

:::

## Heterogeneidad y controles

::: {.incremental .highlight-last style="font-size: 120%;"}

- Meritocracia: dos ítems (recompensa al esfuerzo y a la inteligencia/habilidades), promediados y dicotomizados en baja vs. alta meritocracia.

- Estimo modelos con interacciones movilidad × meritocracia para analizar efectos de tratamiento heterogéneos.

- Las mismas covariables pretratamiento $(C)$ se incluyen en el modelo de desenlace para bloquear caminos de backdoor y lograr doble robustez.

:::

## Estrategia analítica (A)

::: {.incremental .highlight-last style="font-size: 120%;"}

- Para cada clase de origen, estimo modelos lineales de probabilidad ponderados:

:::

::: {.fragment style="font-size: 120%;"}
$$
Y_{it}=\alpha+\beta T_i+X_i'\gamma+\lambda_t+\varepsilon_{it}
$$ {#eq-final}
:::

::: {.incremental style="font-size:120%;"}

- Los modelos utilizan pesos IPW–ATT estabilizados, efectos fijos de ola y errores estándar robustos agrupados (CR2) a nivel de individuo.
- $\beta$ se interpreta como $\widehat{ATT}_{j,k,j}$: el efecto causal de alcanzar el destino $k$ en comparación con permanecer en el origen $j$.

:::

# Resultados {.xlarge data-background-color="#523870"}

## Efectos de movilidad

::: {.fragment style="font-size: 120%;"}

```{r}
#| label: fig-mp
#| fig-cap-location: top
#| fig-cap: "Efectos de movilidad en las preferencias por comodificación de las pensiones"
#| fig-align: center
#| out-width: '180%'
#| fig-asp: 0.6

g3 +
  theme(text = element_text(size = 14))
```

:::

## El rol de la meritocracia

::: {.fragment style="font-size: 120%;"}

```{r}
#| label: fig-mp_i
#| fig-cap-location: top
#| fig-cap: "Efectos de movilidad en las preferencias por comodificación de las pensiones, según percepciones meritocráticas"
#| fig-align: center
#| out-width: '160%'
#| fig-asp: 0.6

g4+
  theme(text = element_text(size = 14))
```

:::

# Discusión y conclusiones {data-background-color="#523870"}

## Discusión y conclusiones

::: {.incremental style="font-size: 120%;"}

1) Los desplazamientos intergeneracionales hacia y desde posiciones de estatus alto moldean causalmente el apoyo a la mercantilización de las pensiones, mientras que las trayectorias descendentes de alta a media y de alta a baja lo reducen [@gugushvili_subjective_2017; @helgason_longterm_2023; @busemeyer_welfare_2020; @ares_changing_2020].

2) Las creencias meritocráticas se correlacionan con estas preferencias, pero parecen operar en una vía paralela [@rodriguez_intergenerational_2024].

3) Trabajo futuro: mejorar las medidas del lado del origen, incorporar baterías más ricas de ítems de justicia de mercado y desarrollar aplicaciones comparativas más allá del caso chileno.

:::

# Muchas gracias!

-   **Github project:** <https://github.com/Andreas-Lafferte/mobility-market-justice>

## Referencias

::: {#refs}
:::

# Material suplementario

## Chequeos de robustez y análisis de sensibilidad

::: {.incremental .highlight-last style="font-size: 120%;"}

- **Chequeo de robustez:** Recodificar la VD de tres formas (3–5 vs. 1–2, excluir neutros y escala 5 puntos lineal) preserva el patrón sustantivo: la trayectoria Middle→High sigue asociada positivamente y High→Low negativamente con la mercantilización, mientras que el efecto High→Middle deja de ser significativo y se muestra sensible a la codificación del desenlace.

- **E-values / análisis de sensibilidad (Cinelli & Hazlett):** Los efectos Middle→High y High→Low son moderadamente robustos, pues requerirían un confusor no medido más fuerte que el sexo (~10–11% de la varianza residual conjunta) para anularlos, mientras que el efecto High→Middle podría desaparecer con confusores de magnitud más modesta.

:::

## El efecto de la edad

```{r}
#| echo: false

# -----------
# Age 
# -----------

### Interacciones

### Low to middle

data_lm_e <- datos1 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = moblm, 
         w = w_low_lm_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")),
         edad = case_when(
           edad >= 18 & edad <= 29 ~ '18-29',
           edad >= 30 & edad <= 49 ~ '30-49',
           edad >= 50 & edad <= 64 ~ '50-64',
           edad >= 65 ~ '65 or more'),
         edad = factor(edad, levels = c('18-29', '30-49', '50-64', '65 or more')))

mp_lm_e <-
  lm_robust(y ~ t*edad + sexo + nac + etnia + hogar + edu0 + ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_lm_e)

### Low to high

data_lh_e <- datos1 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = moblh, 
         w = w_low_lh_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")),
         edad = case_when(
           edad >= 18 & edad <= 29 ~ '18-29',
           edad >= 30 & edad <= 49 ~ '30-49',
           edad >= 50 & edad <= 64 ~ '50-64',
           edad >= 65 ~ '65 or more'),
         edad = factor(edad, levels = c('18-29', '30-49', '50-64', '65 or more')))

mp_lh_e <- 
  lm_robust(y ~ t*edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_lh_e)

### middle to high

data_mh_e <- datos2 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobmh, 
         w = w_mid_mh_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")),
         edad = case_when(
           edad >= 18 & edad <= 29 ~ '18-29',
           edad >= 30 & edad <= 49 ~ '30-49',
           edad >= 50 & edad <= 64 ~ '50-64',
           edad >= 65 ~ '65 or more'),
         edad = factor(edad, levels = c('18-29', '30-49', '50-64', '65 or more')))

mp_mh_e <-  
  lm_robust(y ~ t*edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_mh_e)

### middle to low

data_ml_e <- datos2 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobml, 
         w = w_mid_ml_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")),
         edad = case_when(
           edad >= 18 & edad <= 29 ~ '18-29',
           edad >= 30 & edad <= 49 ~ '30-49',
           edad >= 50 & edad <= 64 ~ '50-64',
           edad >= 65 ~ '65 or more'),
         edad = factor(edad, levels = c('18-29', '30-49', '50-64', '65 or more')))

mp_ml_e <- 
  lm_robust(y ~ t*edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_ml_e)

### High to middle

data_hm_e <- datos3 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobhm, 
         w = w_hig_hm_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")),
         edad = case_when(
           edad >= 18 & edad <= 29 ~ '18-29',
           edad >= 30 & edad <= 49 ~ '30-49',
           edad >= 50 & edad <= 64 ~ '50-64',
           edad >= 65 ~ '65 or more'),
         edad = factor(edad, levels = c('18-29', '30-49', '50-64', '65 or more')))

mp_hm_e <- 
  lm_robust(y ~ t*edad + sexo + nac + etnia + hogar + edu0+ ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_hm_e)

### High to low

data_hl_e <- datos3 %>% 
  select(idencuesta, 
         ola, 
         y = just_pension_f, 
         t = mobhl, 
         w = w_hig_hl_t, 
         edad = m0_edad_f, 
         sexo = sexo_f, 
         nac = nacionalidad_f,
         etnia = etnia_f,
         hogar = hogar_bip_f, 
         edu0 = educ_orig_f) %>% 
  filter(!is.na(t), !is.na(w)) %>% 
  mutate(ola = factor(ola, levels = c("1", "2", "3"),
                      labels = c("2016", "2018", "2023")),
         edad = case_when(
           edad >= 18 & edad <= 29 ~ '18-29',
           edad >= 30 & edad <= 49 ~ '30-49',
           edad >= 50 & edad <= 64 ~ '50-64',
           edad >= 65 ~ '65 or more'),
         edad = factor(edad, levels = c('18-29', '30-49', '50-64', '65 or more')))

mp_hl_e <- 
  lm_robust(y ~ t*edad + sexo + nac + etnia + hogar + edu0 + ola,
            weights = w,
            se_type = "CR2",
            clusters = idencuesta,
            data = data_hl_e)

```

::: {style="font-size: 50%;"}

```{r}
#| label: tbl-inttime
#| tbl-cap: "Efectos de movilidad en preferencias por la mercantilización de las pensiones, por edad"
#| tbl-cap-location: top
#| results: asis
#| echo: false

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment",
  "edad30-49" = "30-49",
  "edad50-64" = "50-64",
  "edad65 or more" = "65 or more",
  "t:edad30-49" = "Mobility treatment x Age 30-49",
  "t:edad50-64" = "Mobility treatment x Age 50-64",
  "t:edad65 or more" = "Mobility treatment x Age 65 or more"
  )

texreg::htmlreg(list(mp_lm_e,
               mp_lh_e,
               mp_mh_e,
               mp_ml_e,
               mp_hm_e,
               mp_hl_e),
          custom.model.names = c(
            "Low-Middle",
            "Low-High",
            "Middle-High",            
            "Middle-Low",
            "High-Middle",
            "High-Low"
          ),   caption.above = NULL,
               caption = NULL,
               custom.coef.map = ccoef,
               digits = 2,
               groups = list("Age (Ref.= 18-29)" = 3:5, "Mobility treatment x Age (Ref.= 18-29)" = 6:8),
               custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.5,
               center = T,float.pos = "H!",
               custom.gof.rows = list("Controls"=rep("Yes",6)))

```
:::

### 2025-12-11 presentations/seminario_clase_redes/mjp-mobility-redes-2025.qmd
Up presentacion jueves 11

@@
 
-Privatización y mercantilización de diversos bienes públicos y servicios sociales. [@gingrich_making_2011; @ferre_welfare_2023]
+Privatización y mercantilización de bienes públicos y servicios sociales [@gingrich_making_2011; @ferre_welfare_2023]
 
@@
 
-En América Latina, la privatización de las pensiones y la capitalización individual han sido fundamentales[@huber_political_2000; @verbic_political_2019]
+En América Latina, la privatización de las pensiones y la capitalización individual han sido claves [@huber_political_2000; @verbic_political_2019]
 
@@
 
-Reasigna los riesgos de la vejez y cambia las normas públicas sobre lo que es justo en materia de bienestar social [@busemeyer_welfare_2020; @madero-cabib_private_2019; @borzutzky_pension_2012]
+Reasigna los riesgos de la vejez y moldea las normas públicas sobre lo que es justo en materia de bienestar social [@busemeyer_welfare_2020; @madero-cabib_private_2019; @borzutzky_pension_2012]
 
@@
 
-Chile combina una elevada desigualdad, una profunda privatización del bienestar y un sistema de pensiones privado tensionado [@flores_top_2020; @somma_no_2021; @madariaga_three_2020]
+Chile combina una elevada desigualdad, una profunda privatización del bienestar y un sistema de pensiones privado [@flores_top_2020; @somma_no_2021; @madariaga_three_2020]
 
@@
 
-Alta movilidad absoluta coexiste con movimientos de clase de corto alcance, acceso restringido a puestos de élite y desigualdades intergeneracionales persistentes.
-[@lopez-roldan_comparative_2021; @salgado_uplifting_2025; @torche_intergenerational_2014]
+Alta movilidad absoluta coexiste con trayectorias cortas, élites cerradas y desigualdad intergeneracional persistente [@lopez-roldan_comparative_2021; @salgado_uplifting_2025; @torche_intergenerational_2014]
+
 :::
@@
 
-Sabemos que el apoyo a los servicios basados en el mercado está estratificado [@lindh_public_2015; @busemeyer_welfare_2020; @kerner_pension_2020; @otero_power_2024] y vinculado a la meritocracia [@castillo_perceptions_2025], pero sabemos menos sobre cómo la movilidad social y las creencias meritocráticas configuran estas preferencias en el ámbito de las pensiones
+Sabemos que el apoyo a los servicios basados en el mercado está estratificado y asociado a la meritocracia [@lindh_public_2015; @busemeyer_welfare_2020; @kerner_pension_2020; @otero_power_2024; @castillo_perceptions_2025], pero mucho menos sobre cómo la movilidad social y estas creencias moldean las preferencias de justicia de mercado en pensiones
+
 :::
@@
 
-- La movilidad social modifica los ingresos, la seguridad y la exposición al riesgo, alterando los intereses materiales en un sistema de capitalización individual de contribución definida[@busemeyer_welfare_2020; @helgason_longterm_2023; @mau_inequality_2015]
+- La movilidad social modifica los ingresos, la seguridad y la exposición al riesgo, alterando los intereses materiales en un sistema de capitalización individual de contribución definida [@busemeyer_welfare_2020; @helgason_longterm_2023; @mau_inequality_2015]
 
@@
 
-- La movilidad social reconfigura las actitudes frente a la desigualdad al empujar a quienes experimentan movilidad ascendente hacia visiones más meritocráticas y a quienes experimentan movilidad descendente hacia explicaciones más estructurales, siendo las fuertes creencias meritocráticas un canal clave de estos efectos [@mijs_belief_2022; @gugushvili_intergenerational_2016; @bucca_merit_2016; @traini_climbing_2025; @matamoros-lima_social_2025; @deng_its_2025]
+- La movilidad social reconfigura las actitudes frente a la desigualdad: la ascendente impulsa visiones más meritocráticas, la descendente explicaciones más estructurales, con las creencias meritocráticas como canal central [@mijs_belief_2022; @gugushvili_intergenerational_2016; @bucca_merit_2016; @traini_climbing_2025; @matamoros-lima_social_2025; @deng_its_2025]
 
@@
 
-- Sigo a @breen_effects_2024 para estimar efectos causales de la movilidad social intergeneracional.
-- Para cada clase de origen $O=j$, comparo a quienes se desplazan desde $j$ hasta $k$ con su resultado contrafactual si se hubiesen mantenido en $j$.
+- Sigo a @breen_effects_2024 para estimar efectos causales de la movilidad social intergeneracional
+- Para cada clase de origen $O=j$, comparo a quienes se desplazan desde $j$ hasta $k$ con su resultado contrafactual si se hubiesen mantenido en $j$
 
@@
 
-- ¿Cómo afecta la movilidad social intergeneracional las preferencias de las personas por la mercantilización de las pensiones en Chile?
+- ¿Cómo afecta la movilidad social intergeneracional las preferencias por la comodificación de las pensiones en Chile?
 
-- @breen_effects_2024: efecto causal promedio de pasar desde una clase de origen *j* a una clase de destino *k*, comparando a los individuos que efectivamente alcanzan *k* con su resultado contrafactual si hubiesen permanecido en su clase de origen *j* (@eq-att).
+- @breen_effects_2024: efecto causal promedio de pasar desde una clase de origen *j* a una clase de destino *k*, comparando a los individuos que efectivamente alcanzan *k* con su resultado contrafactual si hubiesen permanecido en su clase de origen *j* (@eq-att)
 
@@
 
-- [ELSOC](https://coes.cl/elsoc/) (COES): panel urbano nacional 2016–2023; muestreo probabilístico estratificado, por conglomerados y multietápico; población objetivo: residentes habituales de 18 a 75 años.
+- [ELSOC](https://coes.cl/elsoc/) (COES): panel urbano nacional 2016–2023; muestreo probabilístico estratificado, por conglomerados y multietápico; población objetivo: residentes habituales de 18 a 75 años
 
-- Analizo los años 2016, 2018 y 2023 en función de la disponibilidad de variables ocupacionales.
+- Analizo los años 2016, 2018 y 2023 en función de la disponibilidad de variables ocupacionales
 
-- Muestra analítica: *N* = 3.435 observaciones anidadas en *N* = 1.787 individuos (2016: 914; 2018: 1.377; 2023: 1.144).
+- Muestra analítica: *N* = 3.435 observaciones anidadas en *N* = 1.787 individuos (2016: 914; 2018: 1.377; 2023: 1.144)
 
-- Submuestras por trayectorias de movilidad (por ejemplo, baja→baja, baja→media), reteniendo móviles y no móviles emparejados dentro de la misma clase de origen.
+- Submuestras por trayectorias de movilidad (por ejemplo, baja→baja, baja→media), reteniendo móviles y no móviles emparejados dentro de la misma clase de origen
+
 :::
@@
 
-- Covariables **pretratamiento** $C$: (a) educación parental (máximo nivel educativo de padre/madre), en 10 categorías ordinales desde sin escolaridad hasta estudios de posgrado; (b) corresidencia con ambos padres a los 15 años (0 = no, 1 = sí); (c) nacionalidad (0 = no chilena, 1 = chilena); (d) edad en años; (e) sexo (0 = hombre, 1 = mujer); y (f) pertenencia a pueblo indígena (0 = no, 1 = sí).
-- Dentro de cada clase de origen, el *entropy balancing* repondera a los no móviles de modo que su distribución de covariables coincida con la de los móviles → IPW–ATT estabilizado.
-- Estos pesos identifican el ATT y contribuyen a ajustar por la selección en la movilidad.
+- Covariables **pretratamiento** $C$: (a) educación parental (máximo nivel educativo de padre/madre), en 10 categorías ordinales desde sin escolaridad hasta estudios de posgrado; (b) corresidencia con ambos padres a los 15 años (0 = no, 1 = sí); (c) nacionalidad (0 = no chilena, 1 = chilena); (d) edad en años; (e) sexo (0 = hombre, 1 = mujer); y (f) pertenencia a pueblo indígena (0 = no, 1 = sí)
+- Dentro de cada clase de origen, el *entropy balancing* repondera a los no móviles de modo que su distribución de covariables coincida con la de los móviles → IPW–ATT estabilizado
+- Estos pesos identifican el ATT y contribuyen a ajustar por la selección en la movilidad
 
@@
 
-- Meritocracia: dos ítems (recompensa al esfuerzo y a la inteligencia/habilidades), promediados y dicotomizados en baja vs. alta meritocracia.
+- Meritocracia: dos ítems (recompensa al esfuerzo y a la inteligencia/habilidades), promediados y dicotomizados en baja vs. alta meritocracia
 
-- Estimo modelos con interacciones movilidad × meritocracia para analizar efectos de tratamiento heterogéneos.
+- Estimo modelos con interacciones movilidad × meritocracia para analizar efectos de tratamiento heterogéneos
 
-- Las mismas covariables pretratamiento $(C)$ se incluyen en el modelo de desenlace para bloquear caminos de backdoor y lograr doble robustez.
+- Las mismas covariables pretratamiento $(C)$ se incluyen en el modelo de desenlace para bloquear caminos de backdoor y lograr doble robustez
 
@@
 
-- Los modelos utilizan pesos IPW–ATT estabilizados, efectos fijos de ola y errores estándar robustos agrupados (CR2) a nivel de individuo.
-- $\beta$ se interpreta como $\widehat{ATT}_{j,k,j}$: el efecto causal de alcanzar el destino $k$ en comparación con permanecer en el origen $j$.
+- Los modelos utilizan pesos IPW–ATT estabilizados, efectos fijos de ola y errores estándar robustos agrupados (CR2) a nivel de individuo
+- $\beta$ se interpreta como $\widehat{ATT}_{j,k,j}$: el efecto causal de alcanzar el destino $k$ en comparación con permanecer en el origen $j$
 
@@
 
-1) Los desplazamientos intergeneracionales hacia y desde posiciones de estatus alto moldean causalmente el apoyo a la mercantilización de las pensiones, mientras que las trayectorias descendentes de alta a media y de alta a baja lo reducen [@gugushvili_subjective_2017; @helgason_longterm_2023; @busemeyer_welfare_2020; @ares_changing_2020].
+1) La movilidad intergeneracional hacia y desde posiciones de alto estatus afectan causalmente el apoyo a la mercantilización de las pensiones, mientras que las trayectorias descendentes de alta a media y de alta a baja lo reducen [@gugushvili_subjective_2017; @helgason_longterm_2023; @busemeyer_welfare_2020; @ares_changing_2020]
 
-2) Las creencias meritocráticas se correlacionan con estas preferencias, pero parecen operar en una vía paralela [@rodriguez_intergenerational_2024].
+2) Las creencias meritocráticas se correlacionan con estas preferencias, pero parecen operar en una vía paralela a la movilidad [@rodriguez_intergenerational_2024]
 
-3) Trabajo futuro: mejorar las medidas del lado del origen, incorporar baterías más ricas de ítems de justicia de mercado y desarrollar aplicaciones comparativas más allá del caso chileno.
+3) Trabajo futuro: ajustar enfoque y enviarlo, mejorar las medidas del lado del origen ($C$) e incorporar baterías más ricas de ítems de justicia de mercado
 
@@
 - **Chequeo de robustez:** Recodificar la VD de tres formas (3–5 vs.
-1–2, excluir neutros y escala 5 puntos lineal) preserva el patrón sustantivo: la trayectoria Middle→High sigue asociada positivamente y High→Low negativamente con la mercantilización, mientras que el efecto High→Middle deja de ser significativo y se muestra sensible a la codificación del desenlace.
+1–2, excluir neutros y escala 5 puntos lineal) preserva el patrón sustantivo: la trayectoria Middle→High sigue asociada positivamente y High→Low negativamente con la mercantilización, mientras que el efecto High→Middle deja de ser significativo y se muestra sensible a la codificación del desenlace
 
-- **E-values / análisis de sensibilidad (Cinelli & Hazlett):** Los efectos Middle→High y High→Low son moderadamente robustos, pues requerirían un confusor no medido más fuerte que el sexo (~10–11% de la varianza residual conjunta) para anularlos, mientras que el efecto High→Middle podría desaparecer con confusores de magnitud más modesta.
+- **E-values / análisis de sensibilidad (Cinelli & Hazlett):** Los efectos Middle→High y High→Low son moderadamente robustos, pues requerirían un confusor no medido más fuerte que el sexo (~10–11% de la varianza residual conjunta) para anularlos, mientras que el efecto High→Middle podría desaparecer con confusores de magnitud más modesta
 
